---
title: "JMJ Cluster BLAST Visualization"
author: "Sawers-Rellan Labs"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "shared_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8
)

# Setup project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("shared_paper")
```

# Overview

This notebook visualizes the JMJ cluster region in the B73 maize genome by
mapping cDNA sequences back to the genomic region. The visualization shows:

1. **Gene annotation**: Exon structure of the 5 JMJ paralogs in the cluster
2. **BLAST hits**: Alignment of cDNA sequences to the genomic region

The JMJ cluster spans approximately 190 kb on chromosome 4 (177,400,000 - 177,590,000)
and contains five jumonji-domain containing genes:

| Gene | Transcript | Description |
|------|------------|-------------|
| jmj9 | Zm00001eb191790_T001 | Full paralog |
| psi | Zm00001eb191790_T006 | Pseudogene |
| jmj6 | Zm00001eb191790_T013 | Full paralog |
| jmj4 | Zm00001eb191820_T001 | Full paralog |
| jmj2 | Zm00001d051961_T002 | Full paralog (V4 annotation) |

# Load Libraries

```{r libraries}
library(GenomicRanges)
library(ggbio)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(knitr)
```

***

# Define File Paths

```{r file-paths}
# Input files
annotation_file <- file.path(paths$data, "jmj_annotation_5_genes.csv")
blast_file <- file.path(paths$intermediate, "B73_jmj_cluster_genomic_vs_candidates_v5_cDNA.blast")

# Check files exist
if (!file.exists(annotation_file)) {
  stop("Annotation file not found: ", annotation_file,
       "\nPlease ensure jmj_annotation_5_genes.csv is in the data directory.")
}

if (!file.exists(blast_file)) {
  stop("BLAST file not found: ", blast_file,
       "\nPlease run scripts/02_genomics_foundation/generate_jmj_blast_data.sh first.",
       "\nOutput should be saved to results/shared_paper/intermediate/")
}

cat("Input files:\n")
cat("  Annotation:", annotation_file, "\n")
cat("  BLAST:", blast_file, "\n")
```

***
# Define Constants

```{r constants}
# JMJ cluster genomic coordinates (B73 NAM5)
CLUSTER_CHROM <- "4"
CLUSTER_START <- 177400000
CLUSTER_END <- 177590000

# The BLAST query coordinates are relative to the extracted region (0-based)
# We need to add the cluster start position to convert to genomic coordinates
COORD_OFFSET <- CLUSTER_START + 1  # +1 for 1-based coordinates

# Transcript display order (5' to 3' in cluster)
TRANSCRIPT_ORDER <- c(
  "Zm00001eb191790_T001",  # jmj9
  "Zm00001d051961_T002",   # jmj2 (V4)
  "Zm00001eb191790_T006",  # psi
  "Zm00001eb191790_T013",  # jmj6
  "Zm00001eb191820_T001"   # jmj4
)

# Gene name mapping for labels
GENE_NAMES <- c(
  "Zm00001eb191790_T001" = "jmj9",
  "Zm00001d051961_T002" = "jmj2",
  "Zm00001eb191790_T006" = "psi",
  "Zm00001eb191790_T013" = "jmj6",
  "Zm00001eb191820_T001" = "jmj4"
)
```

***

# Load Gene Annotation

```{r load-annotation}
# Load annotation table
jmj_annotation_table <- read.csv(annotation_file)

cat("Annotation table dimensions:", dim(jmj_annotation_table), "\n")
cat("Feature types:", unique(jmj_annotation_table$type), "\n")

# Filter to exons only
exon_table <- jmj_annotation_table %>%
  filter(type == "exon")

cat("Number of exons:", nrow(exon_table), "\n")
```

```{r create-exon-ranges}
# Create GenomicRanges object for exons
exon_ranges <- with(exon_table,
  IRanges(start = start, end = end)
)

exons <- with(exon_table,
  GRanges(
    seqnames = seqnames,
    ranges = exon_ranges,
    transcript = transcript_id,
    type = "exon",
    track = "jmj_annotation"
  )
)

# Add 100% identity for consistent coloring
exons$identity <- 100

cat("Created GRanges object with", length(exons), "exons\n")
```

***

# Load BLAST Results

```{r load-blast}
# BLAST outfmt 6 column names
blast_cols <- c(
  "qseqid",    # Query sequence ID
  "sseqid",    # Subject sequence ID
  "identity",  # Percent identity
  "length",    # Alignment length
  "mismatch",  # Number of mismatches
  "gapopen",   # Number of gap opens
  "qstart",    # Query start
  "qend",      # Query end
  "sstart",    # Subject start
  "send",      # Subject end
  "evalue",    # E-value
  "bitscore"   # Bit score
)

# Load BLAST results
blast <- read.table(blast_file, header = FALSE, sep = "\t")
colnames(blast) <- blast_cols

cat("BLAST results dimensions:", dim(blast), "\n")
cat("Unique subjects:", length(unique(blast$sseqid)), "\n")
```

```{r process-blast}
# Convert coordinates from query-relative to genomic
blast <- blast %>%
  mutate(
    qseqid = CLUSTER_CHROM,  # Chromosome 4
    qstart = qstart + COORD_OFFSET,
    qend = qend + COORD_OFFSET
  )

# Create GenomicRanges object for BLAST hits
hit_ranges <- with(blast,
  IRanges(start = qstart, end = qend)
)

hits <- with(blast,
  GRanges(
    type = "blast_hit",
    seqnames = qseqid,
    ranges = hit_ranges,
    transcript = sseqid,
    score = bitscore,
    identity = identity,
    track = sseqid
  )
)

# Order transcripts for display
hits$transcript <- factor(hits$transcript, levels = TRANSCRIPT_ORDER)

cat("Created GRanges object with", length(hits), "BLAST hits\n")
```

***

# Summary Statistics

```{r blast-summary}
# Summarize BLAST hits by transcript
blast_summary <- blast %>%
  group_by(sseqid) %>%
  summarize(
    n_hits = n(),
    total_length = sum(length),
    mean_identity = mean(identity),
    max_identity = max(identity),
    min_evalue = min(evalue),
    .groups = "drop"
  ) %>%
  mutate(gene = GENE_NAMES[sseqid]) %>%
  arrange(factor(sseqid, levels = TRANSCRIPT_ORDER))

kable(blast_summary,
      col.names = c("Transcript", "Hits", "Total bp", "Mean %ID", "Max %ID", "Min E-value", "Gene"),
      digits = 2,
      caption = "BLAST hit summary by transcript")
```

***

# Visualization

## BLAST Hits Plot

```{r blast-plot, fig.height=8, fig.width=12}
# Split hits by transcript for faceted plotting
hits_list <- split(hits, hits$transcript)

# Create individual plots for each transcript
blast_plots <- lapply(names(hits_list), function(tx) {
  gr <- hits_list[[tx]]
  gene_name <- GENE_NAMES[tx]

  autoplot(gr, geom = "rect", aes(fill = identity)) +
    ggplot2::xlim(CLUSTER_START, CLUSTER_END) +
    scale_fill_viridis_c(name = "% Identity", option = "plasma", limits = c(85, 100)) +
    ggplot2::ggtitle(paste0(gene_name, " (", tx, ")")) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.title = element_text(size = 10, hjust = 1)
    )
})

# Combine with shared legend
blast_plot <- ggarrange(
  plotlist = lapply(blast_plots, function(p) p@ggplot),
  ncol = 1,
  align = "v",
  common.legend = TRUE,
  legend = "right"
)

print(blast_plot)
```

## Gene Annotation Plot

```{r gene-plot, fig.height=3, fig.width=12}
# Create gene annotation plot using autoplot
gene_plot <- autoplot(exons, geom = "rect", aes(fill = transcript, color = transcript)) +
  ggplot2::xlim(CLUSTER_START, CLUSTER_END) +
  scale_fill_brewer(palette = "Set1", name = "Transcript") +
  scale_color_brewer(palette = "Set1", name = "Transcript") +
  ggplot2::labs(x = "Chromosome 4 Position (bp)") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

print(gene_plot@ggplot)
```

## Combined Figure

```{r combined-figure, fig.height=12, fig.width=12}
# Combine BLAST and gene annotation plots
combined_plot <- ggarrange(
  blast_plot,
  gene_plot@ggplot,
  ncol = 1,
  heights = c(3, 1),
  align = "v"
)

print(combined_plot)
```

***

# Export Figure

```{r export}
# Export combined figure
output_pdf <- file.path(paths$figures, "jmj_cluster_blast.pdf")
output_png <- file.path(paths$figures, "jmj_cluster_blast.png")

ggplot2::ggsave(
  output_pdf,
  plot = combined_plot,
  width = 12,
  height = 12,
  device = "pdf"
)

ggplot2::ggsave(
  output_png,
  plot = combined_plot,
  width = 12,
  height = 12,
  dpi = 300
)

cat("Exported figures:\n")
cat("  PDF:", output_pdf, "\n")
cat("  PNG:", output_png, "\n")
```

***

# Session Info

```{r session-info}
sessionInfo()
```
