---
title: "Field-Only Perturbation Analysis: Data Preparation"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/field_perturbation_test"
  n_permutations: 200
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/01_data_prep.Rmd', output_dir='output/field_perturbation_YYYYMMDD/01_data_prep', params=list(run_dir='output/field_perturbation_YYYYMMDD'))"

Note: This script is typically run via run_field_perturbation.sh which handles
output directories automatically.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Load PSU 2022 field data only (no Crow 2020 data), prepare for
network construction without batch correction.

**Approach**:
1. Load pre-normalized voom object (already TMM + voom normalized)
2. Extract expression matrix and sample metadata
3. Recode genotypes (B73 = CTRL, Inv4m = Inv4m)
4. QC with PCA
5. No batch correction needed (single experiment)

**Expected outcome**: Expression matrix (43 samples, ~20k genes) ready for
gene filtering and network construction.

**Key Difference from Joint Pipeline**: Uses only PSU 2022 field data and
skips batch correction since there is only one experiment.

---

# Setup and Data Loading

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(patchwork)
})

set.seed(123)
```

```{r setup_output}
# Output directory
OUTPUT_DIR <- here::here(params$run_dir, "01_data_prep")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Analysis type: PSU 2022 field-only (no batch correction)\n\n")
}
```

## Load PSU 2022 Normalized Data

```{r load_psu_data}
# Load pre-normalized voom object (43 QC'd samples)
psu_voom <- readRDS(file.path(paths$intermediate, "normalized_expression_voom_object_leaf_trt.rds"))

# Extract expression matrix (genes x samples)
expr_matrix <- psu_voom$E

# Extract sample metadata
sample_info <- psu_voom$targets %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  mutate(
    # Recode B73 to CTRL for consistency
    Genotype = recode(Genotype, "B73" = "CTRL"),
    Tissue = "Leaf",
    Experiment = "PSU2022"
  ) %>%
  dplyr::select(sample_id, Genotype, Tissue, Experiment, Treatment, lib.size)

{
  cat("=== PSU 2022 DATA ===\n")
  cat("Expression matrix:", nrow(expr_matrix), "genes x",
      ncol(expr_matrix), "samples\n")
  cat("\nGenotype breakdown:\n")
  print(table(sample_info$Genotype))
  cat("\nTreatment breakdown:\n")
  print(table(sample_info$Treatment))
  cat("\nLibrary size range (M):", round(range(sample_info$lib.size) / 1e6, 1), "\n\n")
}
```

---

# Quality Control: PCA

```{r pca_qc, fig.width=12, fig.height=5}
# PCA
pca_result <- prcomp(t(expr_matrix), scale. = TRUE)
var_explained <- summary(pca_result)$importance[2, 1:2] * 100

pca_df <- tibble(
  sample_id = sample_info$sample_id,
  Genotype = sample_info$Genotype,
  Treatment = sample_info$Treatment,
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2]
)

# PCA by Genotype
p1 <- ggplot(pca_df, aes(PC1, PC2, color = Genotype, shape = Treatment)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("CTRL" = "#009E73", "Inv4m" = "#D55E00")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "PSU 2022 Field Data (by Genotype)",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  )

# PCA by Treatment
p2 <- ggplot(pca_df, aes(PC1, PC2, color = Treatment, shape = Genotype)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal(base_size = 12) +
  labs(
    title = "PSU 2022 Field Data (by Treatment)",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  )

p1 + p2

ggsave(
  file.path(OUTPUT_DIR, "pca_qc.pdf"),
  p1 + p2,
  width = 12, height = 5
)
```

## Quantify Genotype Effect

```{r genotype_effect}
# R-squared for PC1 ~ Genotype
r2_geno <- summary(lm(PC1 ~ Genotype, data = pca_df))$r.squared

{
  cat("=== GENOTYPE EFFECT ===\n")
  cat("PC1 ~ Genotype R-squared:", round(r2_geno, 3), "\n")
  cat("This represents the variance explained by Inv4m genotype.\n\n")
}
```

---

# Save Outputs

```{r save_outputs}
# Save expression matrix (genes x samples, as in joint pipeline)
saveRDS(
  expr_matrix,
  file.path(OUTPUT_DIR, "combined_expression.rds")
)

# Save sample metadata
write_csv(
  sample_info,
  file.path(OUTPUT_DIR, "sample_metadata.csv")
)

# Save summary statistics
summary_stats <- tibble(
  metric = c(
    "Total samples",
    "CTRL samples",
    "Inv4m samples",
    "Total genes",
    "R2 Genotype (PC1)"
  ),
  value = c(
    nrow(sample_info),
    sum(sample_info$Genotype == "CTRL"),
    sum(sample_info$Genotype == "Inv4m"),
    nrow(expr_matrix),
    round(r2_geno, 3)
  )
)

write_csv(
  summary_stats,
  file.path(OUTPUT_DIR, "data_prep_summary.csv")
)

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("combined_expression.rds:", nrow(expr_matrix), "genes x",
      ncol(expr_matrix), "samples\n")
  cat("sample_metadata.csv:", nrow(sample_info), "samples\n")
  cat("pca_qc.pdf: QC plot\n")
  cat("data_prep_summary.csv: summary statistics\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Sample counts
  cat("Check 1: Sample counts\n")
  cat("  Total:", nrow(sample_info), "(expected: 43)\n")
  stopifnot(nrow(sample_info) == 43)

  # Check 2: Genotype balance
  cat("\nCheck 2: Genotype balance\n")
  geno_table <- table(sample_info$Genotype)
  print(geno_table)
  stopifnot(geno_table["CTRL"] > 15)
  stopifnot(geno_table["Inv4m"] > 15)

  # Check 3: Expression matrix dimensions
  cat("\nCheck 3: Expression dimensions\n")
  cat("  Genes:", nrow(expr_matrix), "\n")
  cat("  Samples:", ncol(expr_matrix), "\n")
  stopifnot(ncol(expr_matrix) == nrow(sample_info))

  # Check 4: No missing values
  cat("\nCheck 4: Missing values\n")
  n_na <- sum(is.na(expr_matrix))
  cat("  NA values:", n_na, "\n")
  stopifnot(n_na == 0)

  cat("\n=== ALL CHECKS PASSED ===\n")
}
```

# Session Info

```{r session_info}
sessionInfo()
```
