---
title: "Field-Only Perturbation Analysis: Gene Filtering"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/field_perturbation_test"
  n_permutations: 200
  gene_selection: "fdr"
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/02_gene_filter.Rmd', output_dir='output/field_perturbation_YYYYMMDD/02_gene_filter', params=list(run_dir='output/field_perturbation_YYYYMMDD'))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Filter the expression matrix to differentially expressed genes.

**Gene Selection Mode**: `r params$gene_selection`

**Modes**:
- `fdr`: Confirmed DEGs (FDR < 0.05) - ~465 genes
- `pvalue`: DEG neighborhood (P < 0.005) - ~851 genes

**Approach**: Use the DEG effects from the PA2022 interaction model to select genes
based on the specified threshold for the Inv4m predictor.

**Expected outcome**: Filtered expression matrix with differentially expressed genes.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

set.seed(123)
```

```{r setup_directories}
# Input/output directories
INPUT_DIR <- here::here(params$run_dir, "01_data_prep")
OUTPUT_DIR <- here::here(params$run_dir, "02_gene_filter")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Gene selection mode
GENE_SELECTION <- params$gene_selection

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Input directory:", INPUT_DIR, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Gene selection:", GENE_SELECTION, "\n")
  if (GENE_SELECTION == "fdr") {
    cat("Filtering: FDR < 0.05 (confirmed DEGs only)\n\n")
  } else {
    cat("Filtering: P < 0.005 (DEG neighborhood)\n\n")
  }
}
```

---

# Load Data

## Expression Data

```{r load_expression}
# Load expression from data prep step
expr_matrix <- readRDS(file.path(INPUT_DIR, "combined_expression.rds"))

# Load metadata
sample_meta <- read_csv(
  file.path(INPUT_DIR, "sample_metadata.csv"),
  show_col_types = FALSE
)

{
  cat("=== EXPRESSION DATA ===\n")
  cat("Genes:", nrow(expr_matrix), "\n")
  cat("Samples:", ncol(expr_matrix), "\n")
  cat("Sample breakdown:\n")
  print(table(sample_meta$Genotype))
}
```

## DEG Effects from PA2022 Analysis

```{r load_deg_effects}
# Load DEG effects from PA2022 interaction model
effects_df <- read_csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  show_col_types = FALSE
)

{
  cat("\n=== DEG EFFECTS FILE ===\n")
  cat("Total rows:", nrow(effects_df), "\n")
  cat("Predictors:\n")
  print(table(effects_df$predictor))
}
```

---

# Define DEG Set

Gene selection based on the specified mode (`r GENE_SELECTION`).

```{r define_deg}
# Thresholds
FDR_THRESHOLD <- 0.05
PVALUE_THRESHOLD <- 0.005

# Get both gene sets for comparison
fdr_degs <- effects_df %>%
  filter(predictor == "Inv4m", adj.P.Val < FDR_THRESHOLD) %>%
  pull(gene) %>%
  unique()

pvalue_degs <- effects_df %>%
  filter(predictor == "Inv4m", P.Value < PVALUE_THRESHOLD) %>%
  pull(gene) %>%
  unique()

# Select genes based on mode
if (GENE_SELECTION == "fdr") {
  selected_genes <- fdr_degs
  threshold_label <- paste0("FDR < ", FDR_THRESHOLD)
} else {
  selected_genes <- pvalue_degs
  threshold_label <- paste0("P < ", PVALUE_THRESHOLD)
}

{
  cat("=== GENE SELECTION ===\n")
  cat("Mode:", GENE_SELECTION, "\n")
  cat("Threshold:", threshold_label, "\n")
  cat("Selected genes:", length(selected_genes), "\n\n")
  cat("Comparison:\n")
  cat("  FDR < 0.05:", length(fdr_degs), "genes\n")
  cat("  P < 0.005:", length(pvalue_degs), "genes\n")
  cat("  Neighborhood-only genes:", length(pvalue_degs) - length(fdr_degs), "\n")
}
```

---

# Filter Expression to Selected DEGs

```{r filter_expression}
# Find genes present in expression data
genes_in_expr <- intersect(selected_genes, rownames(expr_matrix))
genes_missing <- setdiff(selected_genes, rownames(expr_matrix))

# Filter expression matrix
expr_filtered <- expr_matrix[genes_in_expr, ]

{
  cat("\n=== GENE FILTERING ===\n")
  cat("Selected genes:", length(selected_genes), "\n")
  cat("Present in expression data:", length(genes_in_expr), "\n")
  cat("Missing from expression data:", length(genes_missing), "\n")

  if (length(genes_missing) > 0 && length(genes_missing) <= 20) {
    cat("Missing genes:", paste(genes_missing, collapse = ", "), "\n")
  }

  cat("\nFiltered expression matrix:", nrow(expr_filtered), "genes x",
      ncol(expr_filtered), "samples\n")
}
```

---

# Prepare Gene Annotation

```{r annotate_genes}
# Get annotation info for filtered genes
gene_info <- effects_df %>%
  filter(predictor == "Inv4m", gene %in% genes_in_expr) %>%
  dplyr::select(
    gene, logFC, P.Value, adj.P.Val,
    regulation, in_Inv4m, in_cis
  ) %>%
  mutate(
    is_confirmed_DEG = adj.P.Val < FDR_THRESHOLD,
    is_neighborhood = P.Value < PVALUE_THRESHOLD & !is_confirmed_DEG,
    selection_mode = GENE_SELECTION
  ) %>%
  arrange(adj.P.Val)

{
  cat("\n=== GENE ANNOTATION ===\n")
  cat("Genes with annotation:", nrow(gene_info), "\n")
  cat("Confirmed DEGs (FDR < 0.05):", sum(gene_info$is_confirmed_DEG), "\n")
  cat("Neighborhood-only genes:", sum(gene_info$is_neighborhood), "\n")
  cat("\nRegulation breakdown:\n")
  print(table(gene_info$regulation))
  cat("\nCis-regulated:", sum(gene_info$in_cis, na.rm = TRUE), "\n")
  cat("In Inv4m region:", sum(gene_info$in_Inv4m, na.rm = TRUE), "\n")
}
```

---

# Visualize Gene Selection

```{r volcano_plot, fig.width=10, fig.height=6}
# Prepare data for volcano plot
volcano_df <- effects_df %>%
  filter(predictor == "Inv4m") %>%
  mutate(
    is_confirmed = adj.P.Val < FDR_THRESHOLD,
    is_neighborhood = P.Value < PVALUE_THRESHOLD & !is_confirmed,
    category = case_when(
      GENE_SELECTION == "fdr" & is_confirmed ~ "Selected (FDR < 0.05)",
      GENE_SELECTION == "pvalue" & (is_confirmed | is_neighborhood) ~ "Selected (P < 0.005)",
      GENE_SELECTION == "fdr" & is_neighborhood ~ "Neighborhood (excluded)",
      TRUE ~ "Not selected"
    )
  )

# Set color scheme based on mode
if (GENE_SELECTION == "fdr") {
  color_values <- c(
    "Selected (FDR < 0.05)" = "#D55E00",
    "Neighborhood (excluded)" = "#56B4E9",
    "Not selected" = "gray70"
  )
  title_text <- "Field-Only Analysis: Confirmed DEGs (FDR < 0.05)"
  subtitle_text <- sprintf(
    "%d genes selected | %d neighborhood genes EXCLUDED",
    length(genes_in_expr),
    sum(volcano_df$is_neighborhood)
  )
} else {
  color_values <- c(
    "Selected (P < 0.005)" = "#D55E00",
    "Not selected" = "gray70"
  )
  title_text <- "Field-Only Analysis: DEG Neighborhood (P < 0.005)"
  subtitle_text <- sprintf(
    "%d genes selected (includes %d confirmed + %d neighborhood)",
    length(genes_in_expr),
    sum(gene_info$is_confirmed_DEG),
    sum(gene_info$is_neighborhood)
  )
}

# Volcano plot
ggplot(volcano_df, aes(x = logFC, y = -log10(P.Value), color = category)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(
    yintercept = -log10(PVALUE_THRESHOLD),
    linetype = "dashed",
    color = "gray50"
  ) +
  scale_color_manual(values = color_values, name = "Gene Category") +
  annotate(
    "text",
    x = max(volcano_df$logFC) * 0.8,
    y = -log10(PVALUE_THRESHOLD) + 0.3,
    label = paste("P =", PVALUE_THRESHOLD),
    color = "gray50",
    size = 3
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = title_text,
    subtitle = subtitle_text,
    x = "log2 Fold Change (Inv4m vs CTRL)",
    y = "-log10(P-value)"
  )

ggsave(
  file.path(OUTPUT_DIR, "deg_selection_volcano.pdf"),
  width = 10, height = 6
)
```

```{r logfc_distribution, fig.width=10, fig.height=5}
# Distribution of logFC in selected genes
ggplot(gene_info, aes(x = logFC, fill = regulation)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  scale_fill_manual(
    values = c(
      "Upregulated" = "#D55E00",
      "Downregulated" = "#0072B2",
      "Unregulated" = "gray50"
    )
  ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_minimal(base_size = 12) +
  labs(
    title = "log2 Fold Change Distribution in Confirmed DEGs",
    x = "log2 Fold Change (Inv4m vs CTRL)",
    y = "Count",
    fill = "Regulation"
  )

ggsave(
  file.path(OUTPUT_DIR, "deg_logfc_distribution.pdf"),
  width = 10, height = 5
)
```

---

# Save Outputs

```{r save_outputs}
# Save filtered expression
saveRDS(
  expr_filtered,
  file.path(OUTPUT_DIR, "filtered_expression.rds")
)

# Save gene list
write_csv(
  gene_info,
  file.path(OUTPUT_DIR, "deg_neighborhood_genes.csv")
)

# Save summary
summary_stats <- tibble(
  metric = c(
    "Gene selection mode",
    "Threshold",
    "Selected genes",
    "Genes in expression data",
    "Genes missing",
    "Confirmed DEGs (FDR < 0.05)",
    "Neighborhood-only genes",
    "Upregulated",
    "Downregulated",
    "Cis-regulated",
    "In Inv4m region"
  ),
  value = c(
    GENE_SELECTION,
    threshold_label,
    length(selected_genes),
    length(genes_in_expr),
    length(genes_missing),
    sum(gene_info$is_confirmed_DEG),
    sum(gene_info$is_neighborhood),
    sum(gene_info$regulation == "Upregulated"),
    sum(gene_info$regulation == "Downregulated"),
    sum(gene_info$in_cis, na.rm = TRUE),
    sum(gene_info$in_Inv4m, na.rm = TRUE)
  )
)

write_csv(
  summary_stats,
  file.path(OUTPUT_DIR, "gene_filter_summary.csv")
)

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("filtered_expression.rds:", nrow(expr_filtered), "genes x",
      ncol(expr_filtered), "samples\n")
  cat("deg_neighborhood_genes.csv:", nrow(gene_info), "genes\n")
  cat("gene_filter_summary.csv: summary statistics\n")
  cat("deg_selection_volcano.pdf: selection visualization\n")
  cat("deg_logfc_distribution.pdf: logFC distribution\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Gene count in expected range (depends on mode)
  cat("Check 1: Gene count in expected range\n")
  cat("  Filtered genes:", nrow(expr_filtered), "\n")
  if (GENE_SELECTION == "fdr") {
    cat("  Expected (FDR mode): ~400-500\n")
    stopifnot(nrow(expr_filtered) > 300 && nrow(expr_filtered) < 600)
  } else {
    cat("  Expected (pvalue mode): ~800-900\n")
    stopifnot(nrow(expr_filtered) > 700 && nrow(expr_filtered) < 1000)
  }

  # Check 2: Sample count preserved
  cat("\nCheck 2: Sample count preserved\n")
  cat("  Samples before:", ncol(expr_matrix), "\n")
  cat("  Samples after:", ncol(expr_filtered), "\n")
  stopifnot(ncol(expr_filtered) == ncol(expr_matrix))

  # Check 3: All genes have annotation
  cat("\nCheck 3: All genes annotated\n")
  cat("  Genes in expression:", nrow(expr_filtered), "\n")
  cat("  Genes in annotation:", nrow(gene_info), "\n")
  stopifnot(nrow(gene_info) == nrow(expr_filtered))

  # Check 4: No missing values
  cat("\nCheck 4: No missing values\n")
  n_na <- sum(is.na(expr_filtered))
  cat("  NA values:", n_na, "\n")
  stopifnot(n_na == 0)

  # Check 5: Gene set composition
  cat("\nCheck 5: Gene set composition\n")
  cat("  Mode:", GENE_SELECTION, "\n")
  cat("  Confirmed DEGs:", sum(gene_info$is_confirmed_DEG), "\n")
  cat("  Neighborhood-only:", sum(gene_info$is_neighborhood), "\n")
  if (GENE_SELECTION == "fdr") {
    stopifnot(sum(gene_info$is_neighborhood) == 0)
  }

  cat("\n=== ALL CHECKS PASSED ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
