---
title: "Leaf Gradient Gene Preparation for Field Perturbation Pipeline"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "results/inversion_paper/field_perturbation/run_leaf_gradient"
  source_run_dir: "results/inversion_paper/field_perturbation/run_20251231_201332"
---

<!--
Replacement for 02_gene_filter.Rmd for the leaf gradient gene set (Ojeda 2026).
Instead of filtering by FDR, this script selects photosynthesis and senescence
hub genes from two pre-defined lists and subsets the expression matrix.

Run this notebook first to populate the new run directory, then run steps 3-7
via the shell script:

  Rscript -e "rmarkdown::render(
    'scripts/inversion_paper/field_perturbation/02_leaf_gradient_gene_prep.Rmd',
    output_dir = '{run_dir}/02_gene_filter',
    params = list(
      run_dir        = '{run_dir}',
      source_run_dir = 'results/inversion_paper/field_perturbation/run_20251231_201332'
    )
  )"

  ./scripts/inversion_paper/run_field_perturbation.sh \
    --resume {run_dir} \
    --start 3 --end 7 --yes
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = here::here())

source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Prepare the expression matrix and directory structure for the
leaf gradient gene consensus pipeline (steps 3–7).

Leaf gradient genes are photosynthesis and senescence hub genes from Ojeda 2026,
identified as co-expressed along the leaf developmental gradient (young base =
photosynthetic, old tip = senescing). These genes are NOT differentially expressed
between CTRL and Inv4m genotypes, making them suitable as a topology control for
the genotype-response module perturbation analysis.

This script:
1. Creates the new run directory structure
2. Copies `sample_metadata.csv` and `combined_expression.rds` from the source
   DEG run into `01_data_prep/` (these are identical — same experiment)
3. Loads the Ojeda 2026 gene lists, excludes any overlap with Inv4m DEGs,
   subsets the expression matrix, and saves `filtered_expression.rds` to
   `02_gene_filter/`

After this script completes, run steps 3–7 via `run_field_perturbation.sh`.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})
```

```{r setup_directories}
RUN_DIR    <- here(params$run_dir)
SOURCE_DIR <- here(params$source_run_dir)

SOURCE_DATA_PREP   <- file.path(SOURCE_DIR, "01_data_prep")
SOURCE_GENE_FILTER <- file.path(SOURCE_DIR, "02_gene_filter")

TARGET_DATA_PREP   <- file.path(RUN_DIR, "01_data_prep")
TARGET_GENE_FILTER <- file.path(RUN_DIR, "02_gene_filter")

# Create run directory structure (mirrors what run_field_perturbation.sh creates)
for (d in c(
  "01_data_prep", "02_gene_filter",
  "03_reference_network/ctrl", "03_reference_network/inv4m",
  "04_consensus_networks/ctrl", "04_consensus_networks/inv4m",
  "05_bootstrap_support/ctrl", "05_bootstrap_support/inv4m",
  "06_preservation", "07_module_annotation", "logs"
)) {
  dir.create(file.path(RUN_DIR, d), recursive = TRUE, showWarnings = FALSE)
}

cat("Run directory:    ", RUN_DIR, "\n")
cat("Source directory: ", SOURCE_DIR, "\n")

# Verify source exists
stopifnot(
  "Source 01_data_prep not found" = dir.exists(SOURCE_DATA_PREP),
  "Source 02_gene_filter not found" = dir.exists(SOURCE_GENE_FILTER)
)
```

---

# Copy Data Prep Outputs

`sample_metadata.csv` and `combined_expression.rds` are identical to the DEG run
— same PA2022 experiment, same 43 samples.

```{r copy_data_prep}
files_to_copy <- c("sample_metadata.csv", "combined_expression.rds")

for (f in files_to_copy) {
  src  <- file.path(SOURCE_DATA_PREP, f)
  dest <- file.path(TARGET_DATA_PREP, f)
  if (!file.exists(src)) {
    stop("Source file not found: ", src)
  }
  file.copy(src, dest, overwrite = TRUE)
  cat("Copied:", f, "\n")
}
```

---

# Load Expression Data

```{r load_expression}
sample_meta <- read_csv(
  file.path(TARGET_DATA_PREP, "sample_metadata.csv"),
  show_col_types = FALSE
)

full_expr_raw <- readRDS(file.path(TARGET_DATA_PREP, "combined_expression.rds"))
full_expr_mat <- if (inherits(full_expr_raw, "EList")) full_expr_raw$E else as.matrix(full_expr_raw)

cat("Expression matrix:", nrow(full_expr_mat), "genes x", ncol(full_expr_mat), "samples\n")
cat("Samples:", nrow(sample_meta), "\n")
print(table(sample_meta$Genotype))
```

---

# Load Ojeda 2026 Gene Lists

```{r load_ojeda_genes}
photo_genes <- read.table(
  file.path(paths$data, "photosynthesis_expression_index_genes_ojeda2026.txt"),
  header = FALSE, stringsAsFactors = FALSE
)$V1

sag_genes <- read.table(
  file.path(paths$data, "senesence_hubs_ojeda2026.tab"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)$gene

cat("Photosynthesis genes (Ojeda 2026):", length(photo_genes), "\n")
cat("Senescence genes (Ojeda 2026):    ", length(sag_genes), "\n")
cat("Overlap between lists:             ",
    length(intersect(photo_genes, sag_genes)), "\n")
```

---

# Build Leaf Gradient Gene Set

Exclude genes not present in the expression matrix and any overlap with Inv4m DEGs.

```{r build_gene_set}
# Load DEG list from source run to exclude overlaps
deg_modules <- read_csv(
  file.path(SOURCE_GENE_FILTER, "filtered_expression.rds") |>
    (\(x) NULL)(),   # just for path check — load actual DEG list below
  show_col_types = FALSE
) |> invisible()

# DEG gene names from source filtered expression
deg_expr <- readRDS(file.path(SOURCE_GENE_FILTER, "filtered_expression.rds"))
deg_genes <- rownames(deg_expr)

cat("Inv4m DEGs (source run):", length(deg_genes), "\n\n")

# Subset to genes present in expression matrix, exclude DEGs
photo_in_matrix <- intersect(photo_genes, rownames(full_expr_mat))
sag_in_matrix   <- intersect(sag_genes,   rownames(full_expr_mat))

photo_control <- setdiff(photo_in_matrix, deg_genes)
sag_control   <- setdiff(sag_in_matrix,   deg_genes)
leaf_genes    <- unique(c(photo_control, sag_control))

cat("=== GENE SET SUMMARY ===\n")
cat("Photosynthesis — in matrix:", length(photo_in_matrix),
    "| after DEG exclusion:", length(photo_control), "\n")
cat("Senescence     — in matrix:", length(sag_in_matrix),
    "| after DEG exclusion:", length(sag_control), "\n")
cat("Combined leaf gradient genes:", length(leaf_genes), "\n")

if (length(photo_genes) - length(photo_in_matrix) > 0)
  cat("WARNING:", length(photo_genes) - length(photo_in_matrix),
      "photosynthesis genes not found in expression matrix\n")
if (length(sag_genes) - length(sag_in_matrix) > 0)
  cat("WARNING:", length(sag_genes) - length(sag_in_matrix),
      "senescence genes not found in expression matrix\n")
```

---

# QC: Expression Distributions

Check that the leaf gradient genes have reasonable expression levels and that
the two programs are distinguishable.

```{r expression_qc, fig.width=9, fig.height=4}
leaf_gene_type <- c(
  setNames(rep("photosynthesis", length(photo_control)), photo_control),
  setNames(rep("senescence",     length(sag_control)),   sag_control)
)

# Mean expression across all samples
mean_expr <- rowMeans(full_expr_mat[leaf_genes, ])

tibble(
  gene      = leaf_genes,
  mean_expr = mean_expr,
  program   = leaf_gene_type[leaf_genes]
) |>
  ggplot(aes(x = mean_expr, fill = program)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c(photosynthesis = "forestgreen",
                                senescence    = "goldenrod3")) +
  theme_minimal(base_size = 13) +
  labs(x = "Mean log-CPM across samples", y = "Gene count", fill = NULL,
       title = "Leaf gradient gene expression levels")
```

```{r correlation_check, fig.width=5, fig.height=4}
# Mean expression index per sample for each program
photo_index <- colMeans(full_expr_mat[photo_control, sample_meta$sample_id])
sag_index   <- colMeans(full_expr_mat[sag_control,   sample_meta$sample_id])

cor_ps <- cor(photo_index, sag_index, method = "pearson")
cat("Pearson correlation photosynthesis vs senescence indices:",
    round(cor_ps, 3), "\n")
cat("(Expected: negative — anti-correlated along leaf gradient)\n")

tibble(photo = photo_index, senescence = sag_index) |>
  ggplot(aes(x = photo, y = senescence)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40", linewidth = 0.7) +
  theme_minimal(base_size = 13) +
  labs(x = "Mean photosynthesis expression",
       y = "Mean senescence expression",
       title = sprintf("Photo vs senescence (r = %.3f)", cor_ps))
```

---

# Save filtered_expression.rds

```{r save_filtered_expression}
leaf_expr <- full_expr_mat[leaf_genes, sample_meta$sample_id]

cat("Saving filtered_expression.rds\n")
cat("  Dimensions:", nrow(leaf_expr), "genes x", ncol(leaf_expr), "samples\n")

saveRDS(leaf_expr,
        file.path(TARGET_GENE_FILTER, "filtered_expression.rds"))

# Save gene list with program annotation for reference
gene_list_df <- tibble(
  gene    = leaf_genes,
  program = leaf_gene_type[leaf_genes]
)
write_csv(gene_list_df,
          file.path(TARGET_GENE_FILTER, "leaf_gradient_gene_list.csv"))

cat("  Saved to:", file.path(TARGET_GENE_FILTER, "filtered_expression.rds"), "\n")
cat("  Gene list saved to:", file.path(TARGET_GENE_FILTER, "leaf_gradient_gene_list.csv"), "\n")
```

---

# Summary

```{r summary}
cat("=== LEAF GRADIENT GENE PREP COMPLETE ===\n\n")
cat("Run directory:", RUN_DIR, "\n\n")
cat("Gene set:\n")
cat("  Photosynthesis:", length(photo_control), "genes\n")
cat("  Senescence:    ", length(sag_control),   "genes\n")
cat("  Total:         ", length(leaf_genes),    "genes\n\n")
cat("Photo-senescence correlation:", round(cor_ps, 3), "\n\n")
cat("Next: run steps 3-7 via the pipeline shell script:\n\n")
cat("  ./scripts/inversion_paper/run_field_perturbation.sh \\\n")
cat("    --resume", RUN_DIR, "\\\n")
cat("    --start 3 --end 7 --yes\n")
```
