---
title: "Field-Only Perturbation Analysis: Reference Network Construction"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  min_module_size: 25
  deep_split: 4
  network_type: "signed"
  forced_power: 0
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/03_reference_network.Rmd', output_dir='output/perturbation_YYYYMMDD/03_reference_network', params=list(run_dir='output/perturbation_YYYYMMDD', min_module_size=25))"

Note: This script is typically run via run_joint_perturbation.sh which handles
output directories automatically.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Build reference networks for CTRL and Inv4m conditions separately.
Select optimal soft power threshold, construct networks using `blockwiseModules()`,
and extract module assignments and connectivity. These parameters will be used
consistently in consensus and bootstrap steps.

**Approach**:
1. Load filtered expression from gene filter step
2. Split by condition (CTRL vs Inv4m)
3. For each condition:
   - Select soft power threshold
   - Build reference network with blockwiseModules
   - Calculate connectivity
   - Save parameters for downstream use

**Expected outcome**: Reference networks for both conditions with saved parameters
(power, minModuleSize) for use in consensus and bootstrap steps.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(WGCNA)
  library(here)
})

allowWGCNAThreads()
set.seed(123)
```

```{r setup_directories}
# Input/output directories
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
FILTER_DIR <- here::here(params$run_dir, "02_gene_filter")
OUTPUT_DIR <- here::here(params$run_dir, "03_reference_network")
dir.create(file.path(OUTPUT_DIR, "ctrl"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_DIR, "inv4m"), recursive = TRUE, showWarnings = FALSE)

# Parameters
MIN_MODULE_SIZE <- params$min_module_size
DEEP_SPLIT <- params$deep_split
NETWORK_TYPE <- params$network_type

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Min module size:", MIN_MODULE_SIZE, "\n")
  cat("Deep split:", DEEP_SPLIT, "\n")
  cat("Network type:", NETWORK_TYPE, "\n\n")
}
```

---

# Load Data

```{r load_data}
# Load filtered expression and sample metadata
expr_filtered <- readRDS(file.path(FILTER_DIR, "filtered_expression.rds"))
sample_meta <- read_csv(
  file.path(DATA_PREP_DIR, "sample_metadata.csv"),
  show_col_types = FALSE
)

# Split samples by condition
ctrl_samples <- sample_meta %>% filter(Genotype == "CTRL") %>% pull(sample_id)
inv4m_samples <- sample_meta %>% filter(Genotype == "Inv4m") %>% pull(sample_id)

# Create expression matrices (samples x genes for WGCNA)
datExpr_CTRL <- t(expr_filtered[, ctrl_samples])
datExpr_Inv4m <- t(expr_filtered[, inv4m_samples])

{
  cat("=== DATA LOADED ===\n")
  cat("Total genes:", nrow(expr_filtered), "\n")
  cat("CTRL samples:", length(ctrl_samples), "\n")
  cat("Inv4m samples:", length(inv4m_samples), "\n")
  cat("datExpr_CTRL:", nrow(datExpr_CTRL), "samples x", ncol(datExpr_CTRL), "genes\n")
  cat("datExpr_Inv4m:", nrow(datExpr_Inv4m), "samples x", ncol(datExpr_Inv4m), "genes\n\n")
}
```

---

# CTRL Reference Network

## Soft Power Selection

```{r ctrl_soft_power}
powers <- 1:20

sft_ctrl <- pickSoftThreshold(
  datExpr_CTRL,
  powerVector = powers,
  networkType = NETWORK_TYPE,
  verbose = 3
)

saveRDS(sft_ctrl, file.path(OUTPUT_DIR, "ctrl", "sft_results.rds"))
```

```{r ctrl_soft_power_plot, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

# Scale-free topology fit
plot(
  sft_ctrl$fitIndices[, 1],
  -sign(sft_ctrl$fitIndices[, 3]) * sft_ctrl$fitIndices[, 2],
  xlab = "Soft Threshold (power)",
  ylab = "Scale Free Topology Model Fit, signed R²",
  type = "n",
  main = "CTRL: Scale Independence"
)
text(
  sft_ctrl$fitIndices[, 1],
  -sign(sft_ctrl$fitIndices[, 3]) * sft_ctrl$fitIndices[, 2],
  labels = powers,
  cex = 0.9,
  col = "red"
)
abline(h = 0.80, col = "blue", lty = 2)

# Mean connectivity
plot(
  sft_ctrl$fitIndices[, 1],
  sft_ctrl$fitIndices[, 5],
  xlab = "Soft Threshold (power)",
  ylab = "Mean Connectivity",
  type = "n",
  main = "CTRL: Mean Connectivity"
)
text(
  sft_ctrl$fitIndices[, 1],
  sft_ctrl$fitIndices[, 5],
  labels = powers,
  cex = 0.9,
  col = "red"
)

par(mfrow = c(1, 1))
```

```{r ctrl_select_power}
# Select power: forced_power param overrides auto-selection when > 0
if (!is.null(params$forced_power) && params$forced_power > 0) {
  power_ctrl <- params$forced_power
  cat("NOTE: Using forced soft power:", power_ctrl, "\n")
} else if (!is.na(sft_ctrl$powerEstimate)) {
  power_ctrl <- sft_ctrl$powerEstimate
} else {
  r2_idx <- which(sft_ctrl$fitIndices$SFT.R.sq > 0.80)
  if (length(r2_idx) > 0) {
    power_ctrl <- sft_ctrl$fitIndices$Power[min(r2_idx)]
  } else {
    power_ctrl <- 12
  }
}

saveRDS(power_ctrl, file.path(OUTPUT_DIR, "ctrl", "selected_power.rds"))

{
  cat("=== CTRL SOFT POWER ===\n")
  cat("Selected power:", power_ctrl, "\n")
  r2_at_power <- sft_ctrl$fitIndices$SFT.R.sq[power_ctrl]
  cat("R² at power:", round(r2_at_power, 3), "\n\n")
}
```

## Network Construction

```{r ctrl_network}
net_ctrl <- blockwiseModules(
  datExpr_CTRL,
  power = power_ctrl,
  networkType = NETWORK_TYPE,
  TOMType = NETWORK_TYPE,
  minModuleSize = MIN_MODULE_SIZE,
  maxBlockSize = 25000,
  reassignThreshold = 0,
  mergeCutHeight = 0,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  deepSplit = DEEP_SPLIT,
  verbose = 3
)

saveRDS(net_ctrl, file.path(OUTPUT_DIR, "ctrl", "net_reference.rds"))
saveRDS(net_ctrl$MEs, file.path(OUTPUT_DIR, "ctrl", "module_eigengenes.rds"))

# Module assignments
ctrl_modules <- data.frame(
  gene = names(net_ctrl$colors),
  module = net_ctrl$colors,
  color = labels2colors(net_ctrl$colors),
  stringsAsFactors = FALSE
)

saveRDS(ctrl_modules, file.path(OUTPUT_DIR, "ctrl", "reference_modules.rds"))
write_csv(ctrl_modules, file.path(OUTPUT_DIR, "ctrl", "reference_modules.csv"))

{
  cat("=== CTRL NETWORK ===\n")
  cat("Total modules:", length(unique(net_ctrl$colors)), "\n")
  cat("Module sizes:\n")
  print(sort(table(net_ctrl$colors), decreasing = TRUE))
  cat("\n")
}
```

## Connectivity

```{r ctrl_connectivity}
adj_ctrl <- adjacency(datExpr_CTRL, power = power_ctrl, type = NETWORK_TYPE)
connectivity_ctrl <- intramodularConnectivity(adj_ctrl, net_ctrl$colors)

saveRDS(adj_ctrl, file.path(OUTPUT_DIR, "ctrl", "adjacency_matrix.rds"))
saveRDS(connectivity_ctrl, file.path(OUTPUT_DIR, "ctrl", "connectivity.rds"))

{
  cat("=== CTRL CONNECTIVITY ===\n")
  cat("Mean kWithin:", round(mean(connectivity_ctrl$kWithin), 2), "\n\n")
}
```

## Save Parameters

```{r ctrl_params}
ctrl_params <- list(
  power = power_ctrl,
  networkType = NETWORK_TYPE,
  TOMType = NETWORK_TYPE,
  minModuleSize = MIN_MODULE_SIZE,
  maxBlockSize = 25000,
  reassignThreshold = 0,
  mergeCutHeight = 0,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  deepSplit = DEEP_SPLIT
)

saveRDS(ctrl_params, file.path(OUTPUT_DIR, "ctrl", "reference_params.rds"))
```

---

# Inv4m Reference Network

## Soft Power Selection

```{r inv4m_soft_power}
sft_inv4m <- pickSoftThreshold(
  datExpr_Inv4m,
  powerVector = powers,
  networkType = NETWORK_TYPE,
  verbose = 3
)

saveRDS(sft_inv4m, file.path(OUTPUT_DIR, "inv4m", "sft_results.rds"))
```

```{r inv4m_soft_power_plot, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

plot(
  sft_inv4m$fitIndices[, 1],
  -sign(sft_inv4m$fitIndices[, 3]) * sft_inv4m$fitIndices[, 2],
  xlab = "Soft Threshold (power)",
  ylab = "Scale Free Topology Model Fit, signed R²",
  type = "n",
  main = "Inv4m: Scale Independence"
)
text(
  sft_inv4m$fitIndices[, 1],
  -sign(sft_inv4m$fitIndices[, 3]) * sft_inv4m$fitIndices[, 2],
  labels = powers,
  cex = 0.9,
  col = "red"
)
abline(h = 0.80, col = "blue", lty = 2)

plot(
  sft_inv4m$fitIndices[, 1],
  sft_inv4m$fitIndices[, 5],
  xlab = "Soft Threshold (power)",
  ylab = "Mean Connectivity",
  type = "n",
  main = "Inv4m: Mean Connectivity"
)
text(
  sft_inv4m$fitIndices[, 1],
  sft_inv4m$fitIndices[, 5],
  labels = powers,
  cex = 0.9,
  col = "red"
)

par(mfrow = c(1, 1))
```

```{r inv4m_select_power}
if (!is.null(params$forced_power) && params$forced_power > 0) {
  power_inv4m <- params$forced_power
  cat("NOTE: Using forced soft power:", power_inv4m, "\n")
} else if (!is.na(sft_inv4m$powerEstimate)) {
  power_inv4m <- sft_inv4m$powerEstimate
} else {
  r2_idx <- which(sft_inv4m$fitIndices$SFT.R.sq > 0.80)
  if (length(r2_idx) > 0) {
    power_inv4m <- sft_inv4m$fitIndices$Power[min(r2_idx)]
  } else {
    power_inv4m <- 12
  }
}

saveRDS(power_inv4m, file.path(OUTPUT_DIR, "inv4m", "selected_power.rds"))

{
  cat("=== INV4M SOFT POWER ===\n")
  cat("Selected power:", power_inv4m, "\n")
  r2_at_power <- sft_inv4m$fitIndices$SFT.R.sq[power_inv4m]
  cat("R² at power:", round(r2_at_power, 3), "\n\n")
}
```

## Network Construction

```{r inv4m_network}
net_inv4m <- blockwiseModules(
  datExpr_Inv4m,
  power = power_inv4m,
  networkType = NETWORK_TYPE,
  TOMType = NETWORK_TYPE,
  minModuleSize = MIN_MODULE_SIZE,
  maxBlockSize = 25000,
  reassignThreshold = 0,
  mergeCutHeight = 0,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  deepSplit = DEEP_SPLIT,
  verbose = 3
)

saveRDS(net_inv4m, file.path(OUTPUT_DIR, "inv4m", "net_reference.rds"))
saveRDS(net_inv4m$MEs, file.path(OUTPUT_DIR, "inv4m", "module_eigengenes.rds"))

inv4m_modules <- data.frame(
  gene = names(net_inv4m$colors),
  module = net_inv4m$colors,
  color = labels2colors(net_inv4m$colors),
  stringsAsFactors = FALSE
)

saveRDS(inv4m_modules, file.path(OUTPUT_DIR, "inv4m", "reference_modules.rds"))
write_csv(inv4m_modules, file.path(OUTPUT_DIR, "inv4m", "reference_modules.csv"))

{
  cat("=== INV4M NETWORK ===\n")
  cat("Total modules:", length(unique(net_inv4m$colors)), "\n")
  cat("Module sizes:\n")
  print(sort(table(net_inv4m$colors), decreasing = TRUE))
  cat("\n")
}
```

## Connectivity

```{r inv4m_connectivity}
adj_inv4m <- adjacency(datExpr_Inv4m, power = power_inv4m, type = NETWORK_TYPE)
connectivity_inv4m <- intramodularConnectivity(adj_inv4m, net_inv4m$colors)

saveRDS(adj_inv4m, file.path(OUTPUT_DIR, "inv4m", "adjacency_matrix.rds"))
saveRDS(connectivity_inv4m, file.path(OUTPUT_DIR, "inv4m", "connectivity.rds"))

{
  cat("=== INV4M CONNECTIVITY ===\n")
  cat("Mean kWithin:", round(mean(connectivity_inv4m$kWithin), 2), "\n\n")
}
```

## Save Parameters

```{r inv4m_params}
inv4m_params <- list(
  power = power_inv4m,
  networkType = NETWORK_TYPE,
  TOMType = NETWORK_TYPE,
  minModuleSize = MIN_MODULE_SIZE,
  maxBlockSize = 25000,
  reassignThreshold = 0,
  mergeCutHeight = 0,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  deepSplit = DEEP_SPLIT
)

saveRDS(inv4m_params, file.path(OUTPUT_DIR, "inv4m", "reference_params.rds"))
```

---

# Summary

```{r summary}
summary_df <- tibble(
  condition = c("CTRL", "Inv4m"),
  n_samples = c(nrow(datExpr_CTRL), nrow(datExpr_Inv4m)),
  n_genes = c(ncol(datExpr_CTRL), ncol(datExpr_Inv4m)),
  soft_power = c(power_ctrl, power_inv4m),
  n_modules = c(length(unique(net_ctrl$colors)), length(unique(net_inv4m$colors))),
  mean_kWithin = c(
    round(mean(connectivity_ctrl$kWithin), 2),
    round(mean(connectivity_inv4m$kWithin), 2)
  )
)

write_csv(summary_df, file.path(OUTPUT_DIR, "reference_network_summary.csv"))

{
  cat("=== REFERENCE NETWORK SUMMARY ===\n")
  print(summary_df)
  cat("\n")
  cat("Parameters saved for downstream use:\n")
  cat("  - Power: CTRL =", power_ctrl, ", Inv4m =", power_inv4m, "\n")
  cat("  - minModuleSize:", MIN_MODULE_SIZE, "\n")
  cat("  - networkType:", NETWORK_TYPE, "\n")
}
```

---

# Validation

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  cat("Check 1: Expression matrix orientation\n")
  stopifnot(nrow(datExpr_CTRL) == length(ctrl_samples))
  stopifnot(nrow(datExpr_Inv4m) == length(inv4m_samples))
  cat("  CTRL:", nrow(datExpr_CTRL), "samples x", ncol(datExpr_CTRL), "genes ✓\n")
  cat("  Inv4m:", nrow(datExpr_Inv4m), "samples x", ncol(datExpr_Inv4m), "genes ✓\n")

  cat("\nCheck 2: Soft power selected\n")
  stopifnot(power_ctrl > 0)
  stopifnot(power_inv4m > 0)
  cat("  CTRL power:", power_ctrl, "✓\n")
  cat("  Inv4m power:", power_inv4m, "✓\n")

  cat("\nCheck 3: Modules detected\n")
  stopifnot(length(unique(net_ctrl$colors)) > 1)
  stopifnot(length(unique(net_inv4m$colors)) > 1)
  cat("  CTRL modules:", length(unique(net_ctrl$colors)), "✓\n")
  cat("  Inv4m modules:", length(unique(net_inv4m$colors)), "✓\n")

  cat("\nCheck 4: Output files created\n")
  expected <- c(
    "ctrl/selected_power.rds", "ctrl/reference_params.rds",
    "inv4m/selected_power.rds", "inv4m/reference_params.rds"
  )
  for (f in expected) {
    exists <- file.exists(file.path(OUTPUT_DIR, f))
    cat("  ", f, ":", ifelse(exists, "✓", "MISSING"), "\n")
  }

  cat("\n=== ALL CHECKS PASSED ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
