---
title: "Field-Only Perturbation Analysis: Consensus Networks"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_consensus_iterations: 1000
  n_bootstrap_iterations: 1000
  gene_subsample_rate: 0.80
  min_module_size_range: "10,15,20,25,30,35"
  deep_split: 4
  network_type: "signed"
  log_frequency: 50
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/04_consensus_networks.Rmd', output_dir='output/perturbation_YYYYMMDD/04_consensus_networks', params=list(run_dir='output/perturbation_YYYYMMDD', n_consensus_iterations=1000))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Build consensus networks for CTRL and Inv4m subsets separately using
the same approach as the field network pipeline (Phases 1-3).

**Approach**:
1. Subset expression data by genotype (CTRL, Inv4m)
2. For each condition:
   - Select optimal soft power threshold
   - Run N consensus iterations with 80% gene subsampling
   - Randomize minModuleSize parameter (10-35)
   - Build indicator matrix (co-sampling counts)
   - Build co-clustering matrix (co-module counts)
   - Compute consensus adjacency: A_ij = C_ij / I_ij
   - Cut dendrogram to identify modules
   - Calculate module eigengenes

**Expected outcome**: Consensus modules for both CTRL and Inv4m conditions,
ready for bootstrap support and preservation analysis.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(WGCNA)
  library(furrr)
  library(future)
  library(here)
})

# Disable WGCNA verbosity
options(stringsAsFactors = FALSE)
disableWGCNAThreads()

set.seed(123)
```

```{r setup_directories}
# Input/output directories
INPUT_DIR <- here::here(params$run_dir, "02_gene_filter")
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
REF_NET_DIR <- here::here(params$run_dir, "03_reference_network")
OUTPUT_DIR <- here::here(params$run_dir, "04_consensus_networks")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_DIR, "ctrl"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_DIR, "inv4m"), recursive = TRUE, showWarnings = FALSE)

# Log directory
LOGS_DIR <- file.path(params$run_dir, "logs")
dir.create(LOGS_DIR, recursive = TRUE, showWarnings = FALSE)
log_file <- file.path(LOGS_DIR, "04_consensus_networks_progress.log")

# Logging function
log_progress <- function(...) {
  timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  msg <- paste0("[", timestamp, "] ", ...)
  cat(msg, "\n", file = log_file, append = TRUE)
}

# Initialize log
cat("", file = log_file)
log_progress("Starting consensus network construction")

# Parse parameters
N_ITERATIONS <- as.integer(params$n_consensus_iterations)
GENE_SUBSAMPLE_RATE <- as.numeric(params$gene_subsample_rate)
MIN_MODULE_SIZE_RANGE <- as.integer(strsplit(params$min_module_size_range, ",")[[1]])
DEEP_SPLIT <- as.integer(params$deep_split)
NETWORK_TYPE <- params$network_type
CONSENSUS_MIN_MODULE_SIZE <- min(MIN_MODULE_SIZE_RANGE)
LOG_FREQUENCY <- if (!is.null(params$log_frequency)) params$log_frequency else 50

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Consensus iterations:", N_ITERATIONS, "\n")
  cat("Gene subsample rate:", GENE_SUBSAMPLE_RATE, "\n")
  cat("minModuleSize range:", paste(MIN_MODULE_SIZE_RANGE, collapse = ", "), "\n")
  cat("deepSplit:", DEEP_SPLIT, "\n")
  cat("networkType:", NETWORK_TYPE, "\n\n")
}

log_progress("Configuration: N_ITERATIONS=", N_ITERATIONS,
             ", SUBSAMPLE_RATE=", GENE_SUBSAMPLE_RATE)
```

---

# Load Data

```{r load_data}
# Load filtered expression
expr_filtered <- readRDS(file.path(INPUT_DIR, "filtered_expression.rds"))

# Load sample metadata
sample_meta <- read_csv(
  file.path(DATA_PREP_DIR, "sample_metadata.csv"),
  show_col_types = FALSE
)

{
  cat("=== INPUT DATA ===\n")
  cat("Expression matrix:", nrow(expr_filtered), "genes x",
      ncol(expr_filtered), "samples\n")
  cat("\nSample breakdown:\n")
  print(table(sample_meta$Genotype))
}
```

```{r split_by_genotype}
# Split samples by genotype
ctrl_samples <- sample_meta %>%
  filter(Genotype == "CTRL") %>%
  pull(sample_id)

inv4m_samples <- sample_meta %>%
  filter(Genotype == "Inv4m") %>%
  pull(sample_id)

# Transpose for WGCNA (samples as rows)
datExpr_CTRL <- t(expr_filtered[, ctrl_samples])
datExpr_Inv4m <- t(expr_filtered[, inv4m_samples])

{
  cat("=== EXPRESSION DATA BY GENOTYPE ===\n")
  cat("CTRL:", nrow(datExpr_CTRL), "samples x", ncol(datExpr_CTRL), "genes\n")
  cat("Inv4m:", nrow(datExpr_Inv4m), "samples x", ncol(datExpr_Inv4m), "genes\n")
}

log_progress("Data loaded: CTRL=", nrow(datExpr_CTRL), " samples, Inv4m=",
             nrow(datExpr_Inv4m), " samples, genes=", ncol(datExpr_CTRL))
```

---

# Define Consensus Functions

```{r define_functions}
#' Run single WGCNA iteration on gene subsample
#'
#' @param datExpr Expression matrix (samples x genes)
#' @param gene_indices Indices of genes to include
#' @param power Soft power threshold
#' @param min_module_size Minimum module size
#' @param deep_split Deep split parameter
#'
#' @return Data frame with gene -> module assignments
run_subsample_wgcna <- function(datExpr, gene_indices, power, min_module_size,
                                 deep_split = 4, network_type = "signed") {
  # Subset to sampled genes
  expr_subset <- datExpr[, gene_indices]
  gene_names <- colnames(expr_subset)

  # Run WGCNA
  net <- blockwiseModules(
    expr_subset,
    power = power,
    networkType = network_type,
    TOMType = network_type,
    minModuleSize = min_module_size,
    mergeCutHeight = 0,  # No merging
    reassignThreshold = 0,
    maxBlockSize = ncol(expr_subset),
    numericLabels = TRUE,
    deepSplit = deep_split,
    verbose = 0
  )

  data.frame(
    gene = gene_names,
    module = net$colors,
    stringsAsFactors = FALSE
  )
}

#' Build indicator matrix (co-sampling counts)
#'
#' @param subsample_configs List of subsample configurations
#' @param gene_names All gene names
#'
#' @return Indicator matrix
build_indicator_matrix <- function(subsample_configs, gene_names) {
  n_genes <- length(gene_names)
  I <- matrix(0, n_genes, n_genes, dimnames = list(gene_names, gene_names))

  for (config in subsample_configs) {
    genes_in <- gene_names[config$gene_indices]
    I[genes_in, genes_in] <- I[genes_in, genes_in] + 1
  }

  return(I)
}

#' Build co-clustering matrix
#'
#' @param module_results List of module assignment data frames
#' @param gene_names All gene names
#'
#' @return Co-clustering matrix
build_coclustering_matrix <- function(module_results, gene_names) {
  n_genes <- length(gene_names)
  C <- matrix(0, n_genes, n_genes, dimnames = list(gene_names, gene_names))

  for (result in module_results) {
    # Group genes by module (exclude grey/0)
    assigned <- result[result$module != 0, ]
    for (mod in unique(assigned$module)) {
      genes_in_mod <- assigned$gene[assigned$module == mod]
      if (length(genes_in_mod) > 1) {
        C[genes_in_mod, genes_in_mod] <- C[genes_in_mod, genes_in_mod] + 1
      }
    }
  }

  return(C)
}

#' Compute consensus adjacency
#'
#' @param C Co-clustering matrix
#' @param I Indicator matrix
#'
#' @return Consensus adjacency matrix
compute_consensus_adjacency <- function(C, I) {
  A <- C / I
  A[is.nan(A)] <- 0
  A[is.infinite(A)] <- 0
  return(A)
}

#' Cut consensus tree to get modules
#'
#' @param A Consensus adjacency matrix
#' @param min_module_size Minimum module size
#' @param deep_split Deep split parameter
#'
#' @return List with tree and module assignments
cut_consensus_tree <- function(A, min_module_size = 25, deep_split = 4) {
  # Convert to distance
  D <- 1 - A

  # Hierarchical clustering
  tree <- hclust(as.dist(D), method = "average")

  # Dynamic tree cut
  modules <- cutreeDynamic(
    dendro = tree,
    distM = D,
    minClusterSize = min_module_size,
    deepSplit = deep_split,
    verbose = 0
  )

  names(modules) <- rownames(A)

  list(tree = tree, modules = modules)
}

cat("Consensus functions defined\n")
```

```{r progress_functions}
# File-based atomic counter for accurate progress tracking
# Each worker increments this file when it completes an iteration

#' Atomically increment counter and return new count
#' @param counter_file Path to counter file
#' @param lock_file Path to lock file
#' @return Integer count after increment
increment_and_get_count <- function(counter_file, lock_file) {
  tryCatch({
    max_retries <- 100
    for (attempt in seq_len(max_retries)) {
      if (!file.exists(lock_file)) {
        tryCatch({
          writeLines(as.character(Sys.getpid()), lock_file)
          break
        }, error = function(e) NULL)
      }
      Sys.sleep(0.01 * sample(1:10, 1))  # Random backoff
    }

    lines <- readLines(counter_file, warn = FALSE)
    current <- if (length(lines) > 0) as.integer(lines[1]) else 0L
    if (is.na(current)) current <- 0L
    new_count <- current + 1L
    writeLines(as.character(new_count), counter_file)
    tryCatch(unlink(lock_file), error = function(e) NULL)
    new_count
  }, error = function(e) NA_integer_)
}

#' Log progress with ETA calculation
#' @param completed Number of completed iterations
#' @param total Total iterations
#' @param start_time Start time from Sys.time()
#' @param log_freq Log frequency
#' @param condition_label Label for the condition (e.g., "CTRL", "Inv4m")
log_iteration_progress <- function(completed, total, start_time, log_freq,
                                   condition_label = "") {
  if (!is.na(completed) && (completed %% log_freq == 0 || completed == total)) {
    elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
    pct_complete <- 100 * completed / total
    avg_time <- if (completed > 0) elapsed / completed else 0
    eta <- (total - completed) * avg_time

    prefix <- if (nchar(condition_label) > 0) paste0(condition_label, ": ") else ""
    log_progress(
      prefix, "Completed ", completed, "/", total,
      " iterations (", round(pct_complete, 1), "%) | ",
      "Elapsed: ", round(elapsed/60, 1), " min | ",
      "ETA: ", round(eta/60, 1), " min"
    )
  }
}

cat("Progress tracking functions defined\n")
```

---

# Build Consensus Network: CTRL

```{r consensus_ctrl}
log_progress("=== STARTING CTRL CONSENSUS NETWORK ===")

# 1. Load soft power from reference network
cat("=== CTRL: LOADING REFERENCE PARAMETERS ===\n")
power_ctrl <- readRDS(file.path(REF_NET_DIR, "ctrl", "selected_power.rds"))
ctrl_ref_params <- readRDS(file.path(REF_NET_DIR, "ctrl", "reference_params.rds"))

cat("Soft power (from reference):", power_ctrl, "\n")
cat("Min module size (from reference):", ctrl_ref_params$minModuleSize, "\n")
log_progress("CTRL power from reference: ", power_ctrl)

# 2. Generate subsample configurations
n_genes <- ncol(datExpr_CTRL)
gene_names <- colnames(datExpr_CTRL)
subsample_size <- floor(GENE_SUBSAMPLE_RATE * n_genes)

ctrl_configs <- lapply(1:N_ITERATIONS, function(i) {
  list(
    iteration = i,
    gene_indices = sample(1:n_genes, subsample_size, replace = FALSE),
    min_module_size = sample(MIN_MODULE_SIZE_RANGE, 1)
  )
})

saveRDS(ctrl_configs, file.path(OUTPUT_DIR, "ctrl", "subsample_configs.rds"))

cat("\nGenerated", N_ITERATIONS, "subsample configurations\n")
cat("Subsample size:", subsample_size, "genes (", GENE_SUBSAMPLE_RATE * 100, "%)\n\n")

# 3. Run parallel WGCNA iterations
n_cores <- max(1, parallel::detectCores() - 1)
plan(multisession, workers = n_cores)

log_progress("CTRL: Starting ", N_ITERATIONS, " WGCNA iterations on ", n_cores, " cores")
log_progress("CTRL: Progress updates every ", LOG_FREQUENCY, " iterations")

start_time <- Sys.time()

# Initialize progress counter
ctrl_counter_file <- file.path(OUTPUT_DIR, "ctrl", ".progress_counter")
ctrl_lock_file <- file.path(OUTPUT_DIR, "ctrl", ".progress_lock")
writeLines("0", ctrl_counter_file)

ctrl_modules <- future_map(ctrl_configs, function(config) {
  result <- run_subsample_wgcna(
    datExpr_CTRL,
    config$gene_indices,
    power_ctrl,
    config$min_module_size,
    DEEP_SPLIT,
    NETWORK_TYPE
  )

  # Track progress
  completed <- increment_and_get_count(ctrl_counter_file, ctrl_lock_file)
  log_iteration_progress(completed, N_ITERATIONS, start_time, LOG_FREQUENCY, "CTRL")

  result
}, .options = furrr_options(seed = TRUE), .progress = FALSE)

# Clean up counter files
unlink(ctrl_counter_file)
unlink(ctrl_lock_file)

plan(sequential)

elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
elapsed_sec <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

log_progress("=== CTRL CONSENSUS COMPLETE ===")
log_progress("CTRL: Total runtime: ", round(elapsed, 2), " minutes")
log_progress("CTRL: Average per iteration: ", round(elapsed_sec / N_ITERATIONS, 2), " seconds")

cat("CTRL WGCNA complete:", round(elapsed, 2), "minutes\n")
```

```{r ctrl_matrices}
# 4. Build indicator matrix
cat("Building CTRL indicator matrix...\n")
I_ctrl <- build_indicator_matrix(ctrl_configs, gene_names)
saveRDS(I_ctrl, file.path(OUTPUT_DIR, "ctrl", "indicator_matrix.rds"))

# 5. Build co-clustering matrix
cat("Building CTRL co-clustering matrix...\n")
C_ctrl <- build_coclustering_matrix(ctrl_modules, gene_names)
saveRDS(C_ctrl, file.path(OUTPUT_DIR, "ctrl", "coclustering_matrix.rds"))

# 6. Compute consensus adjacency
cat("Computing CTRL consensus adjacency...\n")
A_ctrl <- compute_consensus_adjacency(C_ctrl, I_ctrl)
saveRDS(A_ctrl, file.path(OUTPUT_DIR, "ctrl", "consensus_adjacency.rds"))

{
  cat("\n=== CTRL MATRICES ===\n")
  cat("Indicator range: [", min(I_ctrl), ",", max(I_ctrl), "]\n")
  cat("Co-clustering range: [", min(C_ctrl), ",", max(C_ctrl), "]\n")
  cat("Adjacency range: [", round(min(A_ctrl), 3), ",", round(max(A_ctrl), 3), "]\n")
}
```

```{r ctrl_modules}
# 7. Cut consensus tree
cat("\nCutting CTRL consensus tree...\n")
ctrl_tree <- cut_consensus_tree(A_ctrl, CONSENSUS_MIN_MODULE_SIZE, DEEP_SPLIT)

# Create module data frame
ctrl_module_df <- data.frame(
  gene = names(ctrl_tree$modules),
  module = ctrl_tree$modules,
  color = labels2colors(ctrl_tree$modules),
  stringsAsFactors = FALSE
)

write_csv(ctrl_module_df, file.path(OUTPUT_DIR, "ctrl", "consensus_modules.csv"))
saveRDS(ctrl_tree, file.path(OUTPUT_DIR, "ctrl", "consensus_tree.rds"))

# 8. Calculate eigengenes
cat("Computing CTRL module eigengenes...\n")
ctrl_MEs <- moduleEigengenes(datExpr_CTRL, ctrl_tree$modules, excludeGrey = TRUE)
saveRDS(ctrl_MEs$eigengenes, file.path(OUTPUT_DIR, "ctrl", "consensus_eigengenes.rds"))

{
  cat("\n=== CTRL CONSENSUS MODULES ===\n")
  cat("Total modules:", length(unique(ctrl_tree$modules[ctrl_tree$modules != 0])), "\n")
  cat("Assigned genes:", sum(ctrl_tree$modules != 0), "\n")
  cat("Unassigned (grey):", sum(ctrl_tree$modules == 0), "\n")
  cat("\nModule sizes:\n")
  print(table(ctrl_tree$modules))
}

log_progress("CTRL consensus complete: ",
             length(unique(ctrl_tree$modules[ctrl_tree$modules != 0])), " modules")
```

---

# Build Consensus Network: Inv4m

```{r consensus_inv4m}
log_progress("=== STARTING INV4M CONSENSUS NETWORK ===")

# 1. Load soft power from reference network
cat("\n=== INV4M: LOADING REFERENCE PARAMETERS ===\n")
power_inv4m <- readRDS(file.path(REF_NET_DIR, "inv4m", "selected_power.rds"))
inv4m_ref_params <- readRDS(file.path(REF_NET_DIR, "inv4m", "reference_params.rds"))

cat("Soft power (from reference):", power_inv4m, "\n")
cat("Min module size (from reference):", inv4m_ref_params$minModuleSize, "\n")
log_progress("Inv4m power from reference: ", power_inv4m)

# 2. Generate subsample configurations (same structure, different random samples)
set.seed(456)  # Different seed for Inv4m

inv4m_configs <- lapply(1:N_ITERATIONS, function(i) {
  list(
    iteration = i,
    gene_indices = sample(1:n_genes, subsample_size, replace = FALSE),
    min_module_size = sample(MIN_MODULE_SIZE_RANGE, 1)
  )
})

saveRDS(inv4m_configs, file.path(OUTPUT_DIR, "inv4m", "subsample_configs.rds"))

# 3. Run parallel WGCNA iterations
plan(multisession, workers = n_cores)

log_progress("Inv4m: Starting ", N_ITERATIONS, " WGCNA iterations on ", n_cores, " cores")
log_progress("Inv4m: Progress updates every ", LOG_FREQUENCY, " iterations")

start_time <- Sys.time()

# Initialize progress counter
inv4m_counter_file <- file.path(OUTPUT_DIR, "inv4m", ".progress_counter")
inv4m_lock_file <- file.path(OUTPUT_DIR, "inv4m", ".progress_lock")
writeLines("0", inv4m_counter_file)

inv4m_modules <- future_map(inv4m_configs, function(config) {
  result <- run_subsample_wgcna(
    datExpr_Inv4m,
    config$gene_indices,
    power_inv4m,
    config$min_module_size,
    DEEP_SPLIT,
    NETWORK_TYPE
  )

  # Track progress
  completed <- increment_and_get_count(inv4m_counter_file, inv4m_lock_file)
  log_iteration_progress(completed, N_ITERATIONS, start_time, LOG_FREQUENCY, "Inv4m")

  result
}, .options = furrr_options(seed = TRUE), .progress = FALSE)

# Clean up counter files
unlink(inv4m_counter_file)
unlink(inv4m_lock_file)

plan(sequential)

elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
elapsed_sec <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

log_progress("=== INV4M CONSENSUS COMPLETE ===")
log_progress("Inv4m: Total runtime: ", round(elapsed, 2), " minutes")
log_progress("Inv4m: Average per iteration: ", round(elapsed_sec / N_ITERATIONS, 2), " seconds")

cat("Inv4m WGCNA complete:", round(elapsed, 2), "minutes\n")
```

```{r inv4m_matrices}
# 4. Build indicator matrix
cat("Building Inv4m indicator matrix...\n")
I_inv4m <- build_indicator_matrix(inv4m_configs, gene_names)
saveRDS(I_inv4m, file.path(OUTPUT_DIR, "inv4m", "indicator_matrix.rds"))

# 5. Build co-clustering matrix
cat("Building Inv4m co-clustering matrix...\n")
C_inv4m <- build_coclustering_matrix(inv4m_modules, gene_names)
saveRDS(C_inv4m, file.path(OUTPUT_DIR, "inv4m", "coclustering_matrix.rds"))

# 6. Compute consensus adjacency
cat("Computing Inv4m consensus adjacency...\n")
A_inv4m <- compute_consensus_adjacency(C_inv4m, I_inv4m)
saveRDS(A_inv4m, file.path(OUTPUT_DIR, "inv4m", "consensus_adjacency.rds"))

{
  cat("\n=== INV4M MATRICES ===\n")
  cat("Indicator range: [", min(I_inv4m), ",", max(I_inv4m), "]\n")
  cat("Co-clustering range: [", min(C_inv4m), ",", max(C_inv4m), "]\n")
  cat("Adjacency range: [", round(min(A_inv4m), 3), ",", round(max(A_inv4m), 3), "]\n")
}
```

```{r inv4m_modules}
# 7. Cut consensus tree
cat("\nCutting Inv4m consensus tree...\n")
inv4m_tree <- cut_consensus_tree(A_inv4m, CONSENSUS_MIN_MODULE_SIZE, DEEP_SPLIT)

# Create module data frame
inv4m_module_df <- data.frame(
  gene = names(inv4m_tree$modules),
  module = inv4m_tree$modules,
  color = labels2colors(inv4m_tree$modules),
  stringsAsFactors = FALSE
)

write_csv(inv4m_module_df, file.path(OUTPUT_DIR, "inv4m", "consensus_modules.csv"))
saveRDS(inv4m_tree, file.path(OUTPUT_DIR, "inv4m", "consensus_tree.rds"))

# 8. Calculate eigengenes
cat("Computing Inv4m module eigengenes...\n")
inv4m_MEs <- moduleEigengenes(datExpr_Inv4m, inv4m_tree$modules, excludeGrey = TRUE)
saveRDS(inv4m_MEs$eigengenes, file.path(OUTPUT_DIR, "inv4m", "consensus_eigengenes.rds"))

{
  cat("\n=== INV4M CONSENSUS MODULES ===\n")
  cat("Total modules:", length(unique(inv4m_tree$modules[inv4m_tree$modules != 0])), "\n")
  cat("Assigned genes:", sum(inv4m_tree$modules != 0), "\n")
  cat("Unassigned (grey):", sum(inv4m_tree$modules == 0), "\n")
  cat("\nModule sizes:\n")
  print(table(inv4m_tree$modules))
}

log_progress("Inv4m consensus complete: ",
             length(unique(inv4m_tree$modules[inv4m_tree$modules != 0])), " modules")
```

---

# Visualization: CTRL Consensus Network

```{r ctrl_dendrogram, fig.height=8}
# Plot CTRL dendrogram with module colors
plotDendroAndColors(
  ctrl_tree$tree,
  labels2colors(ctrl_tree$modules),
  "CTRL Modules",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05,
  main = "CTRL Consensus Network Gene Dendrogram"
)
```

```{r ctrl_module_sizes, fig.height=5}
# CTRL module size distribution
ctrl_sizes <- ctrl_module_df %>%
  filter(module != 0) %>%
  count(module, color)

ggplot(ctrl_sizes, aes(x = reorder(color, -n), y = n, fill = color)) +
  geom_col(color = "black") +
  scale_fill_identity() +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "CTRL Consensus Module Sizes",
    x = "Module",
    y = "Number of Genes"
  )
```

---

# Visualization: Inv4m Consensus Network

```{r inv4m_dendrogram, fig.height=8}
# Plot Inv4m dendrogram with module colors
plotDendroAndColors(
  inv4m_tree$tree,
  labels2colors(inv4m_tree$modules),
  "Inv4m Modules",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05,
  main = "Inv4m Consensus Network Gene Dendrogram"
)
```

```{r inv4m_module_sizes, fig.height=5}
# Inv4m module size distribution
inv4m_sizes <- inv4m_module_df %>%
  filter(module != 0) %>%
  count(module, color)

ggplot(inv4m_sizes, aes(x = reorder(color, -n), y = n, fill = color)) +
  geom_col(color = "black") +
  scale_fill_identity() +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Inv4m Consensus Module Sizes",
    x = "Module",
    y = "Number of Genes"
  )
```

---

# Compare CTRL vs Inv4m Networks

```{r network_comparison, fig.width=12, fig.height=5}
library(patchwork)

# Module count comparison
comparison_summary <- tibble(
  Condition = c("CTRL", "Inv4m"),
  n_modules = c(
    length(unique(ctrl_tree$modules[ctrl_tree$modules != 0])),
    length(unique(inv4m_tree$modules[inv4m_tree$modules != 0]))
  ),
  n_assigned = c(
    sum(ctrl_tree$modules != 0),
    sum(inv4m_tree$modules != 0)
  ),
  n_unassigned = c(
    sum(ctrl_tree$modules == 0),
    sum(inv4m_tree$modules == 0)
  ),
  mean_adjacency = c(
    mean(A_ctrl[lower.tri(A_ctrl)]),
    mean(A_inv4m[lower.tri(A_inv4m)])
  )
)

{
  cat("=== NETWORK COMPARISON ===\n")
  print(comparison_summary)
}

# Side-by-side module count plot
p1 <- ggplot(comparison_summary, aes(x = Condition, y = n_modules, fill = Condition)) +
  geom_col(color = "black") +
  scale_fill_manual(values = c("CTRL" = "#009E73", "Inv4m" = "#D55E00")) +
  theme_minimal(base_size = 12) +
  labs(title = "Number of Modules", y = "Modules") +
  theme(legend.position = "none")

p2 <- ggplot(comparison_summary, aes(x = Condition, y = n_assigned, fill = Condition)) +
  geom_col(color = "black") +
  scale_fill_manual(values = c("CTRL" = "#009E73", "Inv4m" = "#D55E00")) +
  theme_minimal(base_size = 12) +
  labs(title = "Genes Assigned to Modules", y = "Genes") +
  theme(legend.position = "none")

p3 <- ggplot(comparison_summary, aes(x = Condition, y = mean_adjacency, fill = Condition)) +
  geom_col(color = "black") +
  scale_fill_manual(values = c("CTRL" = "#009E73", "Inv4m" = "#D55E00")) +
  theme_minimal(base_size = 12) +
  labs(title = "Mean Consensus Adjacency", y = "Adjacency") +
  theme(legend.position = "none")

p1 + p2 + p3

ggsave(
  file.path(OUTPUT_DIR, "network_comparison.pdf"),
  p1 + p2 + p3,
  width = 12, height = 5
)

write_csv(comparison_summary, file.path(OUTPUT_DIR, "network_comparison.csv"))
```

---

# Save Combined Outputs

```{r save_outputs}
# Save PDF of dendrograms
pdf(file.path(OUTPUT_DIR, "ctrl", "consensus_dendrogram.pdf"), width = 10, height = 8)
plotDendroAndColors(
  ctrl_tree$tree,
  labels2colors(ctrl_tree$modules),
  "CTRL Modules",
  dendroLabels = FALSE,
  hang = 0.03,
  main = "CTRL Consensus Network"
)
dev.off()

pdf(file.path(OUTPUT_DIR, "inv4m", "consensus_dendrogram.pdf"), width = 10, height = 8)
plotDendroAndColors(
  inv4m_tree$tree,
  labels2colors(inv4m_tree$modules),
  "Inv4m Modules",
  dendroLabels = FALSE,
  hang = 0.03,
  main = "Inv4m Consensus Network"
)
dev.off()

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("\nCTRL outputs:\n")
  cat("  ctrl/soft_power.rds\n")
  cat("  ctrl/subsample_configs.rds\n")
  cat("  ctrl/indicator_matrix.rds\n")
  cat("  ctrl/coclustering_matrix.rds\n")
  cat("  ctrl/consensus_adjacency.rds\n")
  cat("  ctrl/consensus_modules.csv\n")
  cat("  ctrl/consensus_tree.rds\n")
  cat("  ctrl/consensus_eigengenes.rds\n")
  cat("  ctrl/consensus_dendrogram.pdf\n")
  cat("\nInv4m outputs:\n")
  cat("  inv4m/soft_power.rds\n")
  cat("  inv4m/subsample_configs.rds\n")
  cat("  inv4m/indicator_matrix.rds\n")
  cat("  inv4m/coclustering_matrix.rds\n")
  cat("  inv4m/consensus_adjacency.rds\n")
  cat("  inv4m/consensus_modules.csv\n")
  cat("  inv4m/consensus_tree.rds\n")
  cat("  inv4m/consensus_eigengenes.rds\n")
  cat("  inv4m/consensus_dendrogram.pdf\n")
  cat("\nCombined outputs:\n")
  cat("  network_comparison.csv\n")
  cat("  network_comparison.pdf\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Both networks built
  cat("Check 1: Both networks built\n")
  cat("  CTRL modules:", length(unique(ctrl_tree$modules[ctrl_tree$modules != 0])), "\n")
  cat("  Inv4m modules:", length(unique(inv4m_tree$modules[inv4m_tree$modules != 0])), "\n")
  stopifnot(length(unique(ctrl_tree$modules[ctrl_tree$modules != 0])) > 0)
  stopifnot(length(unique(inv4m_tree$modules[inv4m_tree$modules != 0])) > 0)

  # Check 2: Adjacency matrices valid
  cat("\nCheck 2: Adjacency matrices valid\n")
  cat("  CTRL range: [", round(min(A_ctrl), 3), ",", round(max(A_ctrl), 3), "]\n")
  cat("  Inv4m range: [", round(min(A_inv4m), 3), ",", round(max(A_inv4m), 3), "]\n")
  stopifnot(min(A_ctrl) >= 0 && max(A_ctrl) <= 1)
  stopifnot(min(A_inv4m) >= 0 && max(A_inv4m) <= 1)

  # Check 3: Same genes in both
  cat("\nCheck 3: Gene alignment\n")
  cat("  Same genes:", all(rownames(A_ctrl) == rownames(A_inv4m)), "\n")
  stopifnot(all(rownames(A_ctrl) == rownames(A_inv4m)))

  # Check 4: Eigengenes computed
  cat("\nCheck 4: Eigengenes computed\n")
  cat("  CTRL eigengenes:", ncol(ctrl_MEs$eigengenes), "modules\n")
  cat("  Inv4m eigengenes:", ncol(inv4m_MEs$eigengenes), "modules\n")

  cat("\n=== ALL CHECKS PASSED ===\n")
}

log_progress("Consensus network construction complete!")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
