---
title: "Field-Only Perturbation Analysis: Bootstrap Support"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_bootstrap_iterations: 1000
  stability_threshold_stable: 0.70
  stability_threshold_moderate: 0.50
  network_type: "signed"
  log_frequency: 50
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/05_bootstrap_support.Rmd', output_dir='output/perturbation_YYYYMMDD/05_bootstrap_support', params=list(run_dir='output/perturbation_YYYYMMDD', n_bootstrap_iterations=1000))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Quantify stability of consensus modules for both CTRL and Inv4m
conditions via bootstrap resampling of samples.

**Approach**:
1. For each condition (CTRL, Inv4m):
   - Resample samples WITH replacement
   - Rebuild WGCNA network with fixed parameters
   - Compare bootstrap modules to consensus modules
   - Compute Largest Fragment Proportion (LFP) per module
2. Classify module stability: stable (>=0.70), moderate (>=0.50), unstable (<0.50)

**Expected outcome**: Stability metrics for each consensus module in both conditions.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(WGCNA)
  library(furrr)
  library(future)
  library(here)
})

# Disable WGCNA verbosity
options(stringsAsFactors = FALSE)
disableWGCNAThreads()

set.seed(123)
```

```{r setup_directories}
# Input/output directories
REF_NET_DIR <- here::here(params$run_dir, "03_reference_network")
CONSENSUS_DIR <- here::here(params$run_dir, "04_consensus_networks")
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
FILTER_DIR <- here::here(params$run_dir, "02_gene_filter")
OUTPUT_DIR <- here::here(params$run_dir, "05_bootstrap_support")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_DIR, "ctrl"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_DIR, "inv4m"), recursive = TRUE, showWarnings = FALSE)

# Log directory
LOGS_DIR <- file.path(params$run_dir, "logs")
dir.create(LOGS_DIR, recursive = TRUE, showWarnings = FALSE)
log_file <- file.path(LOGS_DIR, "05_bootstrap_support_progress.log")

# Logging function
log_progress <- function(...) {
  timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  msg <- paste0("[", timestamp, "] ", ...)
  cat(msg, "\n", file = log_file, append = TRUE)
}

# Initialize log
cat("", file = log_file)
log_progress("Starting bootstrap support analysis")

# Parameters
N_BOOTSTRAP <- as.integer(params$n_bootstrap_iterations)
STABILITY_STABLE <- as.numeric(params$stability_threshold_stable)
STABILITY_MODERATE <- as.numeric(params$stability_threshold_moderate)
NETWORK_TYPE <- params$network_type
LOG_FREQUENCY <- if (!is.null(params$log_frequency)) params$log_frequency else 50

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Bootstrap iterations:", N_BOOTSTRAP, "\n")
  cat("Stability threshold (stable):", STABILITY_STABLE, "\n")
  cat("Stability threshold (moderate):", STABILITY_MODERATE, "\n")
  cat("Network type:", NETWORK_TYPE, "\n\n")
}

log_progress("Configuration: N_BOOTSTRAP=", N_BOOTSTRAP)
```

---

# Load Data

```{r load_data}
# Load filtered expression
expr_filtered <- readRDS(file.path(FILTER_DIR, "filtered_expression.rds"))

# Load sample metadata
sample_meta <- read_csv(
  file.path(DATA_PREP_DIR, "sample_metadata.csv"),
  show_col_types = FALSE
)

# Load consensus networks
ctrl_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)
inv4m_modules <- read_csv(
  file.path(CONSENSUS_DIR, "inv4m", "consensus_modules.csv"),
  show_col_types = FALSE
)

# Load reference network parameters (power + minModuleSize)
ctrl_ref_params <- readRDS(file.path(REF_NET_DIR, "ctrl", "reference_params.rds"))
inv4m_ref_params <- readRDS(file.path(REF_NET_DIR, "inv4m", "reference_params.rds"))

power_ctrl <- ctrl_ref_params$power
power_inv4m <- inv4m_ref_params$power
min_module_size_ctrl <- ctrl_ref_params$minModuleSize
min_module_size_inv4m <- inv4m_ref_params$minModuleSize
deep_split_ctrl <- ctrl_ref_params$deepSplit
deep_split_inv4m <- inv4m_ref_params$deepSplit

{
  cat("=== INPUT DATA ===\n")
  cat("Expression matrix:", nrow(expr_filtered), "genes x",
      ncol(expr_filtered), "samples\n")
  cat("CTRL consensus modules:", length(unique(ctrl_modules$module[ctrl_modules$module != 0])), "\n")
  cat("Inv4m consensus modules:", length(unique(inv4m_modules$module[inv4m_modules$module != 0])), "\n")
  cat("\nReference network parameters:\n")
  cat("CTRL - power:", power_ctrl, ", minModuleSize:", min_module_size_ctrl,
      ", deepSplit:", deep_split_ctrl, "\n")
  cat("Inv4m - power:", power_inv4m, ", minModuleSize:", min_module_size_inv4m,
      ", deepSplit:", deep_split_inv4m, "\n")
}
```

```{r prepare_expression}
# Split samples by genotype
ctrl_samples <- sample_meta %>%
  filter(Genotype == "CTRL") %>%
  pull(sample_id)

inv4m_samples <- sample_meta %>%
  filter(Genotype == "Inv4m") %>%
  pull(sample_id)

# Transpose for WGCNA (samples as rows)
datExpr_CTRL <- t(expr_filtered[, ctrl_samples])
datExpr_Inv4m <- t(expr_filtered[, inv4m_samples])

{
  cat("\n=== EXPRESSION DATA ===\n")
  cat("CTRL:", nrow(datExpr_CTRL), "samples x", ncol(datExpr_CTRL), "genes\n")
  cat("Inv4m:", nrow(datExpr_Inv4m), "samples x", ncol(datExpr_Inv4m), "genes\n")
}
```

---

# Define Bootstrap Functions

```{r define_functions}
#' Run single bootstrap iteration of WGCNA
#'
#' @param datExpr Expression matrix (samples x genes)
#' @param power Soft thresholding power
#' @param params List of WGCNA parameters
#' @param iteration_num Iteration number
#'
#' @return Data frame with gene -> module assignments
run_bootstrap_iteration <- function(datExpr, power, min_module_size = 10,
                                     deep_split = 4, network_type = "signed",
                                     iteration_num = NULL) {
  # Resample samples WITH replacement
  n_samples <- nrow(datExpr)
  boot_indices <- sample(1:n_samples, n_samples, replace = TRUE)
  expr_boot <- datExpr[boot_indices, ]

  # Run WGCNA
  net_boot <- blockwiseModules(
    expr_boot,
    power = power,
    networkType = network_type,
    TOMType = network_type,
    minModuleSize = min_module_size,
    mergeCutHeight = 0,
    reassignThreshold = 0,
    maxBlockSize = ncol(expr_boot),
    numericLabels = TRUE,
    deepSplit = deep_split,
    verbose = 0
  )

  data.frame(
    gene = colnames(expr_boot),
    module = net_boot$colors,
    iteration = iteration_num,
    stringsAsFactors = FALSE
  )
}

#' Compute fragmentation-aware module support
#'
#' @param consensus_modules Data frame with consensus module assignments
#' @param bootstrap_results List of bootstrap iteration results
#'
#' @return Data frame with support statistics per module
compute_module_support <- function(consensus_modules, bootstrap_results) {
  # Get list of consensus modules (exclude grey/0)
  modules <- unique(consensus_modules$module[consensus_modules$module != 0])

  support <- map_dfr(modules, function(mod) {
    # Genes in this consensus module
    mod_genes <- consensus_modules$gene[consensus_modules$module == mod]
    mod_color <- consensus_modules$color[consensus_modules$module == mod][1]

    # For each bootstrap iteration, compute fragmentation metrics
    iter_stats <- map_dfr(bootstrap_results, function(boot) {
      boot_sub <- boot[boot$gene %in% mod_genes, ]
      if (nrow(boot_sub) == 0) {
        return(tibble(largest_prop = 0, entropy = NA_real_, n_fragments = 0L))
      }

      # Distribution across bootstrap modules (including grey/0 for entropy)
      counts <- table(boot_sub$module)
      props <- counts / sum(counts)

      # Largest fragment proportion (exclude grey/0)
      counts_nonzero <- counts[names(counts) != "0"]
      if (length(counts_nonzero) == 0) {
        largest_prop <- 0
      } else {
        largest_prop <- max(counts_nonzero) / length(mod_genes)
      }

      # Shannon entropy (all modules including grey)
      entropy <- -sum(props * log2(props + 1e-12))

      # Number of fragments (distinct modules genes split into)
      n_fragments <- length(counts)

      tibble(largest_prop = largest_prop, entropy = entropy, n_fragments = n_fragments)
    })

    n_iter <- nrow(iter_stats)
    sd_val <- sd(iter_stats$largest_prop)

    tibble(
      module = mod,
      color = mod_color,
      n_genes = length(mod_genes),
      mean_lfp = mean(iter_stats$largest_prop),
      median_lfp = median(iter_stats$largest_prop),
      sd_lfp = sd_val,
      se_lfp = sd_val / sqrt(n_iter),  # Standard error of the mean
      min_lfp = min(iter_stats$largest_prop),
      max_lfp = max(iter_stats$largest_prop),
      mean_entropy = mean(iter_stats$entropy, na.rm = TRUE),
      median_entropy = median(iter_stats$entropy, na.rm = TRUE),
      mean_fragments = mean(iter_stats$n_fragments)
    )
  })

  support
}

cat("Bootstrap functions defined\n")
```

```{r progress_functions}
# File-based atomic counter for accurate progress tracking
# Each worker increments this file when it completes an iteration

#' Atomically increment counter and return new count
#' @param counter_file Path to counter file
#' @param lock_file Path to lock file
#' @return Integer count after increment
increment_and_get_count <- function(counter_file, lock_file) {
  tryCatch({
    max_retries <- 100
    for (attempt in seq_len(max_retries)) {
      if (!file.exists(lock_file)) {
        tryCatch({
          writeLines(as.character(Sys.getpid()), lock_file)
          break
        }, error = function(e) NULL)
      }
      Sys.sleep(0.01 * sample(1:10, 1))  # Random backoff
    }

    lines <- readLines(counter_file, warn = FALSE)
    current <- if (length(lines) > 0) as.integer(lines[1]) else 0L
    if (is.na(current)) current <- 0L
    new_count <- current + 1L
    writeLines(as.character(new_count), counter_file)
    tryCatch(unlink(lock_file), error = function(e) NULL)
    new_count
  }, error = function(e) NA_integer_)
}

#' Log progress with ETA calculation
#' @param completed Number of completed iterations
#' @param total Total iterations
#' @param start_time Start time from Sys.time()
#' @param log_freq Log frequency
#' @param condition_label Label for the condition (e.g., "CTRL", "Inv4m")
log_iteration_progress <- function(completed, total, start_time, log_freq,
                                   condition_label = "") {
  if (!is.na(completed) && (completed %% log_freq == 0 || completed == total)) {
    elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
    pct_complete <- 100 * completed / total
    avg_time <- if (completed > 0) elapsed / completed else 0
    eta <- (total - completed) * avg_time

    prefix <- if (nchar(condition_label) > 0) paste0(condition_label, ": ") else ""
    log_progress(
      prefix, "Completed ", completed, "/", total,
      " iterations (", round(pct_complete, 1), "%) | ",
      "Elapsed: ", round(elapsed/60, 1), " min | ",
      "ETA: ", round(eta/60, 1), " min"
    )
  }
}

cat("Progress tracking functions defined\n")
```

---

# Bootstrap Support: CTRL

```{r bootstrap_ctrl}
log_progress("=== STARTING CTRL BOOTSTRAP SUPPORT ===")

# Configure parallelization
n_cores <- max(1, parallel::detectCores() - 1)
plan(multisession, workers = n_cores)

log_progress("CTRL: Starting ", N_BOOTSTRAP, " bootstrap iterations on ", n_cores, " cores")
log_progress("CTRL: Progress updates every ", LOG_FREQUENCY, " iterations")

start_time <- Sys.time()

# Initialize progress counter
ctrl_counter_file <- file.path(OUTPUT_DIR, "ctrl", ".progress_counter")
ctrl_lock_file <- file.path(OUTPUT_DIR, "ctrl", ".progress_lock")
writeLines("0", ctrl_counter_file)

# Run bootstrap iterations using reference network parameters
ctrl_bootstrap <- future_map(1:N_BOOTSTRAP, function(i) {
  result <- run_bootstrap_iteration(
    datExpr_CTRL,
    power_ctrl,
    min_module_size = min_module_size_ctrl,
    deep_split = deep_split_ctrl,
    network_type = NETWORK_TYPE,
    iteration_num = i
  )

  # Track progress
  completed <- increment_and_get_count(ctrl_counter_file, ctrl_lock_file)
  log_iteration_progress(completed, N_BOOTSTRAP, start_time, LOG_FREQUENCY, "CTRL")

  result
}, .options = furrr_options(seed = TRUE), .progress = FALSE)

# Clean up counter files
unlink(ctrl_counter_file)
unlink(ctrl_lock_file)

plan(sequential)

elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
elapsed_sec <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

log_progress("=== CTRL BOOTSTRAP COMPLETE ===")
log_progress("CTRL: Total runtime: ", round(elapsed, 2), " minutes")
log_progress("CTRL: Average per iteration: ", round(elapsed_sec / N_BOOTSTRAP, 2), " seconds")

cat("CTRL bootstrap complete:", round(elapsed, 2), "minutes\n")

# Save raw bootstrap results
saveRDS(ctrl_bootstrap, file.path(OUTPUT_DIR, "ctrl", "bootstrap_results.rds"))
```

```{r ctrl_support}
# Compute support metrics
cat("Computing CTRL module support...\n")
ctrl_support <- compute_module_support(ctrl_modules, ctrl_bootstrap)

# Add stability classification
ctrl_support <- ctrl_support %>%
  mutate(
    stability = case_when(
      mean_lfp >= STABILITY_STABLE ~ "Stable",
      mean_lfp >= STABILITY_MODERATE ~ "Moderate",
      TRUE ~ "Unstable"
    ),
    stability = factor(stability, levels = c("Stable", "Moderate", "Unstable"))
  ) %>%
  arrange(desc(mean_lfp))

# Save support metrics
write_csv(ctrl_support, file.path(OUTPUT_DIR, "ctrl", "module_support.csv"))
saveRDS(ctrl_support, file.path(OUTPUT_DIR, "ctrl", "module_support.rds"))

{
  cat("\n=== CTRL MODULE SUPPORT ===\n")
  cat("Modules analyzed:", nrow(ctrl_support), "\n")
  cat("\nStability classification:\n")
  print(table(ctrl_support$stability))
  cat("\nSupport summary:\n")
  print(summary(ctrl_support$mean_lfp))

  cat("\n=== ALL CTRL MODULES BY STABILITY ===\n")
  for (stab in c("Stable", "Moderate", "Unstable")) {
    mods <- ctrl_support %>% filter(stability == stab)
    cat("\n", stab, "(n=", nrow(mods), "):\n", sep = "")
    if (nrow(mods) > 0) {
      print(mods %>%
              dplyr::select(module, color, n_genes, mean_lfp, mean_entropy, mean_fragments) %>%
              as.data.frame())
    }
  }
}

log_progress("CTRL support complete: ", sum(ctrl_support$stability == "Stable"),
             " stable modules")
```

---

# Bootstrap Support: Inv4m

```{r bootstrap_inv4m}
log_progress("=== STARTING INV4M BOOTSTRAP SUPPORT ===")

set.seed(456)  # Different seed

plan(multisession, workers = n_cores)

log_progress("Inv4m: Starting ", N_BOOTSTRAP, " bootstrap iterations on ", n_cores, " cores")
log_progress("Inv4m: Progress updates every ", LOG_FREQUENCY, " iterations")

start_time <- Sys.time()

# Initialize progress counter
inv4m_counter_file <- file.path(OUTPUT_DIR, "inv4m", ".progress_counter")
inv4m_lock_file <- file.path(OUTPUT_DIR, "inv4m", ".progress_lock")
writeLines("0", inv4m_counter_file)

# Run bootstrap iterations using reference network parameters
inv4m_bootstrap <- future_map(1:N_BOOTSTRAP, function(i) {
  result <- run_bootstrap_iteration(
    datExpr_Inv4m,
    power_inv4m,
    min_module_size = min_module_size_inv4m,
    deep_split = deep_split_inv4m,
    network_type = NETWORK_TYPE,
    iteration_num = i
  )

  # Track progress
  completed <- increment_and_get_count(inv4m_counter_file, inv4m_lock_file)
  log_iteration_progress(completed, N_BOOTSTRAP, start_time, LOG_FREQUENCY, "Inv4m")

  result
}, .options = furrr_options(seed = TRUE), .progress = FALSE)

# Clean up counter files
unlink(inv4m_counter_file)
unlink(inv4m_lock_file)

plan(sequential)

elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
elapsed_sec <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

log_progress("=== INV4M BOOTSTRAP COMPLETE ===")
log_progress("Inv4m: Total runtime: ", round(elapsed, 2), " minutes")
log_progress("Inv4m: Average per iteration: ", round(elapsed_sec / N_BOOTSTRAP, 2), " seconds")

cat("Inv4m bootstrap complete:", round(elapsed, 2), "minutes\n")

# Save raw bootstrap results
saveRDS(inv4m_bootstrap, file.path(OUTPUT_DIR, "inv4m", "bootstrap_results.rds"))
```

```{r inv4m_support}
# Compute support metrics
cat("Computing Inv4m module support...\n")
inv4m_support <- compute_module_support(inv4m_modules, inv4m_bootstrap)

# Add stability classification
inv4m_support <- inv4m_support %>%
  mutate(
    stability = case_when(
      mean_lfp >= STABILITY_STABLE ~ "Stable",
      mean_lfp >= STABILITY_MODERATE ~ "Moderate",
      TRUE ~ "Unstable"
    ),
    stability = factor(stability, levels = c("Stable", "Moderate", "Unstable"))
  ) %>%
  arrange(desc(mean_lfp))

# Save support metrics
write_csv(inv4m_support, file.path(OUTPUT_DIR, "inv4m", "module_support.csv"))
saveRDS(inv4m_support, file.path(OUTPUT_DIR, "inv4m", "module_support.rds"))

{
  cat("\n=== INV4M MODULE SUPPORT ===\n")
  cat("Modules analyzed:", nrow(inv4m_support), "\n")
  cat("\nStability classification:\n")
  print(table(inv4m_support$stability))
  cat("\nSupport summary:\n")
  print(summary(inv4m_support$mean_lfp))

  cat("\n=== ALL INV4M MODULES BY STABILITY ===\n")
  for (stab in c("Stable", "Moderate", "Unstable")) {
    mods <- inv4m_support %>% filter(stability == stab)
    cat("\n", stab, "(n=", nrow(mods), "):\n", sep = "")
    if (nrow(mods) > 0) {
      print(mods %>%
              dplyr::select(module, color, n_genes, mean_lfp, mean_entropy, mean_fragments) %>%
              as.data.frame())
    }
  }
}

log_progress("Inv4m support complete: ", sum(inv4m_support$stability == "Stable"),
             " stable modules")
```

---

# Compare Support Between Conditions

```{r compare_support}
# Create combined summary
support_comparison <- tibble(
  Condition = c("CTRL", "Inv4m"),
  n_modules = c(nrow(ctrl_support), nrow(inv4m_support)),
  n_stable = c(
    sum(ctrl_support$stability == "Stable"),
    sum(inv4m_support$stability == "Stable")
  ),
  n_moderate = c(
    sum(ctrl_support$stability == "Moderate"),
    sum(inv4m_support$stability == "Moderate")
  ),
  n_unstable = c(
    sum(ctrl_support$stability == "Unstable"),
    sum(inv4m_support$stability == "Unstable")
  ),
  mean_lfp = c(
    mean(ctrl_support$mean_lfp),
    mean(inv4m_support$mean_lfp)
  ),
  median_lfp = c(
    median(ctrl_support$mean_lfp),
    median(inv4m_support$mean_lfp)
  )
)

write_csv(support_comparison, file.path(OUTPUT_DIR, "support_comparison.csv"))

{
  cat("\n=== SUPPORT COMPARISON ===\n")
  print(support_comparison)
}
```

---

# Visualization: CTRL Support

```{r ctrl_support_plot, fig.width=10, fig.height=6}
p_ctrl <- ggplot(
  ctrl_support,
  aes(
    x = reorder(color, mean_lfp),
    y = mean_lfp,
    fill = color
  )
) +
  geom_col(color = "black") +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_lfp - se_lfp),
      ymax = pmin(1, mean_lfp + se_lfp)
    ),
    width = 0.3
  ) +
  geom_hline(yintercept = STABILITY_STABLE, linetype = "dashed", color = "darkgreen") +
  geom_hline(yintercept = STABILITY_MODERATE, linetype = "dashed", color = "orange") +
  scale_fill_identity() +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "CTRL Consensus Module Bootstrap Support",
    subtitle = sprintf(
      "n = %d bootstrap iterations | Stable >= %.2f | Moderate >= %.2f | Error bars = SEM",
      N_BOOTSTRAP, STABILITY_STABLE, STABILITY_MODERATE
    ),
    x = "Module",
    y = "Mean Support (LFP)"
  )

print(p_ctrl)

ggsave(
  file.path(OUTPUT_DIR, "ctrl", "module_support_plot.pdf"),
  p_ctrl,
  width = 10, height = 6
)
```

---

# Visualization: Inv4m Support

```{r inv4m_support_plot, fig.width=10, fig.height=6}
p_inv4m <- ggplot(
  inv4m_support,
  aes(
    x = reorder(color, mean_lfp),
    y = mean_lfp,
    fill = color
  )
) +
  geom_col(color = "black") +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_lfp - se_lfp),
      ymax = pmin(1, mean_lfp + se_lfp)
    ),
    width = 0.3
  ) +
  geom_hline(yintercept = STABILITY_STABLE, linetype = "dashed", color = "darkgreen") +
  geom_hline(yintercept = STABILITY_MODERATE, linetype = "dashed", color = "orange") +
  scale_fill_identity() +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Inv4m Consensus Module Bootstrap Support",
    subtitle = sprintf(
      "n = %d bootstrap iterations | Stable >= %.2f | Moderate >= %.2f | Error bars = SEM",
      N_BOOTSTRAP, STABILITY_STABLE, STABILITY_MODERATE
    ),
    x = "Module",
    y = "Mean Support (LFP)"
  )

print(p_inv4m)

ggsave(
  file.path(OUTPUT_DIR, "inv4m", "module_support_plot.pdf"),
  p_inv4m,
  width = 10, height = 6
)
```

---

# Visualization: Fragmentation Entropy

```{r ctrl_entropy_plot, fig.width=10, fig.height=6}
# CTRL fragmentation entropy plot
p_ctrl_entropy <- ggplot(
  ctrl_support,
  aes(
    x = reorder(color, -mean_entropy),
    y = mean_entropy,
    fill = color
  )
) +
  geom_col(color = "black") +
  scale_fill_identity() +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "CTRL Consensus Module Fragmentation",
    subtitle = "Higher entropy = more dispersed across bootstrap modules",
    x = "Module",
    y = "Mean Fragmentation Entropy (bits)"
  )

print(p_ctrl_entropy)

ggsave(
  file.path(OUTPUT_DIR, "ctrl", "fragmentation_entropy_plot.pdf"),
  p_ctrl_entropy,
  width = 10, height = 6
)
```

```{r inv4m_entropy_plot, fig.width=10, fig.height=6}
# Inv4m fragmentation entropy plot
p_inv4m_entropy <- ggplot(
  inv4m_support,
  aes(
    x = reorder(color, -mean_entropy),
    y = mean_entropy,
    fill = color
  )
) +
  geom_col(color = "black") +
  scale_fill_identity() +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Inv4m Consensus Module Fragmentation",
    subtitle = "Higher entropy = more dispersed across bootstrap modules",
    x = "Module",
    y = "Mean Fragmentation Entropy (bits)"
  )

print(p_inv4m_entropy)

ggsave(
  file.path(OUTPUT_DIR, "inv4m", "fragmentation_entropy_plot.pdf"),
  p_inv4m_entropy,
  width = 10, height = 6
)
```

---

# Visualization: Comparison

```{r comparison_plot, fig.width=12, fig.height=5}
library(patchwork)

# Stability distribution comparison
stability_data <- bind_rows(
  ctrl_support %>% mutate(Condition = "CTRL"),
  inv4m_support %>% mutate(Condition = "Inv4m")
)

p_compare_hist <- ggplot(stability_data, aes(x = mean_lfp, fill = Condition)) +
  geom_histogram(bins = 15, alpha = 0.7, position = "dodge") +
  geom_vline(xintercept = STABILITY_STABLE, linetype = "dashed", color = "darkgreen") +
  geom_vline(xintercept = STABILITY_MODERATE, linetype = "dashed", color = "orange") +
  scale_fill_manual(values = c("CTRL" = "#009E73", "Inv4m" = "#D55E00")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Distribution of Module Support",
    x = "Mean LFP",
    y = "Number of Modules"
  )

p_compare_box <- ggplot(stability_data, aes(x = Condition, y = mean_lfp, fill = Condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = STABILITY_STABLE, linetype = "dashed", color = "darkgreen") +
  geom_hline(yintercept = STABILITY_MODERATE, linetype = "dashed", color = "orange") +
  scale_fill_manual(values = c("CTRL" = "#009E73", "Inv4m" = "#D55E00")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Module Support Comparison",
    y = "Mean LFP"
  ) +
  theme(legend.position = "none")

p_compare_hist + p_compare_box

ggsave(
  file.path(OUTPUT_DIR, "support_comparison_plot.pdf"),
  p_compare_hist + p_compare_box,
  width = 12, height = 5
)
```

---

# Save Summary Statistics

```{r save_summary}
summary_stats <- tibble(
  metric = c(
    "Bootstrap iterations",
    "CTRL modules analyzed",
    "CTRL stable modules",
    "CTRL moderate modules",
    "CTRL unstable modules",
    "CTRL mean LFP",
    "Inv4m modules analyzed",
    "Inv4m stable modules",
    "Inv4m moderate modules",
    "Inv4m unstable modules",
    "Inv4m mean LFP",
    "Stability threshold (stable)",
    "Stability threshold (moderate)"
  ),
  value = c(
    N_BOOTSTRAP,
    nrow(ctrl_support),
    sum(ctrl_support$stability == "Stable"),
    sum(ctrl_support$stability == "Moderate"),
    sum(ctrl_support$stability == "Unstable"),
    round(mean(ctrl_support$mean_lfp), 3),
    nrow(inv4m_support),
    sum(inv4m_support$stability == "Stable"),
    sum(inv4m_support$stability == "Moderate"),
    sum(inv4m_support$stability == "Unstable"),
    round(mean(inv4m_support$mean_lfp), 3),
    STABILITY_STABLE,
    STABILITY_MODERATE
  )
)

write_csv(summary_stats, file.path(OUTPUT_DIR, "bootstrap_summary.csv"))

{
  cat("\n=== SAVED OUTPUTS ===\n")
  cat("\nCTRL outputs:\n")
  cat("  ctrl/bootstrap_results.rds\n")
  cat("  ctrl/module_support.csv\n")
  cat("  ctrl/module_support.rds\n")
  cat("  ctrl/module_support_plot.pdf\n")
  cat("\nInv4m outputs:\n")
  cat("  inv4m/bootstrap_results.rds\n")
  cat("  inv4m/module_support.csv\n")
  cat("  inv4m/module_support.rds\n")
  cat("  inv4m/module_support_plot.pdf\n")
  cat("\nCombined outputs:\n")
  cat("  support_comparison.csv\n")
  cat("  support_comparison_plot.pdf\n")
  cat("  bootstrap_summary.csv\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Bootstrap iterations completed
  cat("Check 1: Bootstrap iterations completed\n")
  cat("  CTRL:", length(ctrl_bootstrap), "iterations\n")
  cat("  Inv4m:", length(inv4m_bootstrap), "iterations\n")
  stopifnot(length(ctrl_bootstrap) == N_BOOTSTRAP)
  stopifnot(length(inv4m_bootstrap) == N_BOOTSTRAP)

  # Check 2: All modules analyzed
  cat("\nCheck 2: All modules analyzed\n")
  n_ctrl_mods <- length(unique(ctrl_modules$module[ctrl_modules$module != 0]))
  n_inv4m_mods <- length(unique(inv4m_modules$module[inv4m_modules$module != 0]))
  cat("  CTRL modules:", nrow(ctrl_support), "/", n_ctrl_mods, "\n")
  cat("  Inv4m modules:", nrow(inv4m_support), "/", n_inv4m_mods, "\n")
  stopifnot(nrow(ctrl_support) == n_ctrl_mods)
  stopifnot(nrow(inv4m_support) == n_inv4m_mods)

  # Check 3: LFP values in valid range
  cat("\nCheck 3: LFP values valid\n")
  cat("  CTRL LFP range: [", round(min(ctrl_support$mean_lfp), 3), ",",
      round(max(ctrl_support$mean_lfp), 3), "]\n")
  cat("  Inv4m LFP range: [", round(min(inv4m_support$mean_lfp), 3), ",",
      round(max(inv4m_support$mean_lfp), 3), "]\n")
  stopifnot(all(ctrl_support$mean_lfp >= 0 & ctrl_support$mean_lfp <= 1))
  stopifnot(all(inv4m_support$mean_lfp >= 0 & inv4m_support$mean_lfp <= 1))

  # Check 4: Stability classification complete
  cat("\nCheck 4: Stability classification\n")
  cat("  CTRL classified:", all(!is.na(ctrl_support$stability)), "\n")
  cat("  Inv4m classified:", all(!is.na(inv4m_support$stability)), "\n")

  cat("\n=== ALL CHECKS PASSED ===\n")
}

log_progress("Bootstrap support analysis complete!")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
