---
title: "Field-Only Perturbation Analysis: Consensus-Based Preservation"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_permutations: 200
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/06_preservation.Rmd', output_dir='output/perturbation_YYYYMMDD/06_preservation', params=list(run_dir='output/perturbation_YYYYMMDD', n_permutations=200))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Compare CTRL and Inv4m consensus networks to identify modules
that are disrupted by the Inv4m inversion.

**Approach**:
1. Compare consensus adjacency matrices between conditions
2. Compute module-level adjacency preservation (correlation)
3. Calculate delta kWithin (connectivity change) from consensus adjacency
4. Map modules between conditions using gene overlap
5. Integrate with bootstrap support metrics

**Key Metrics**:
- **Adjacency Correlation**: Cor(A_ctrl, A_inv4m) within module - <0.5 = disrupted
- **Mean Adjacency Delta**: Mean(A_inv4m - A_ctrl) - negative = weakened connections
- **Delta kWithin**: kWithin_inv4m - kWithin_ctrl - negative = lost connectivity

**Expected outcome**: List of disrupted modules with preservation and support metrics.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(WGCNA)
  library(here)
  library(patchwork)
  library(ggrepel)
  library(rstatix)
})

# Enable multi-threading
allowWGCNAThreads()

set.seed(123)
```

```{r setup_directories}
# Input/output directories
CONSENSUS_DIR <- here::here(params$run_dir, "04_consensus_networks")
BOOTSTRAP_DIR <- here::here(params$run_dir, "05_bootstrap_support")
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
OUTPUT_DIR <- here::here(params$run_dir, "06_preservation")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Permutations:", params$n_permutations, "\n\n")
}
```

---

# Load Data

```{r load_consensus}
# Load consensus adjacency matrices
A_ctrl <- readRDS(file.path(CONSENSUS_DIR, "ctrl", "consensus_adjacency.rds"))
A_inv4m <- readRDS(file.path(CONSENSUS_DIR, "inv4m", "consensus_adjacency.rds"))

# Load consensus modules
ctrl_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)
inv4m_modules <- read_csv(
  file.path(CONSENSUS_DIR, "inv4m", "consensus_modules.csv"),
  show_col_types = FALSE
)

{
  cat("=== CONSENSUS NETWORKS ===\n")
  cat("CTRL adjacency:", nrow(A_ctrl), "x", ncol(A_ctrl), "\n")
  cat("Inv4m adjacency:", nrow(A_inv4m), "x", ncol(A_inv4m), "\n")
  cat("CTRL modules:", length(unique(ctrl_modules$module[ctrl_modules$module != 0])), "\n")
  cat("Inv4m modules:", length(unique(inv4m_modules$module[inv4m_modules$module != 0])), "\n")
}
```

```{r load_support}
# Load bootstrap support metrics
ctrl_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "ctrl", "module_support.csv"),
  show_col_types = FALSE
)
inv4m_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "inv4m", "module_support.csv"),
  show_col_types = FALSE
)

{
  cat("\n=== BOOTSTRAP SUPPORT ===\n")
  cat("CTRL stable modules:", sum(ctrl_support$stability == "Stable"), "\n")
  cat("Inv4m stable modules:", sum(inv4m_support$stability == "Stable"), "\n")
}
```

```{r load_go_annotations}
# Load top GO BP terms from 07_module_annotation (for boxplot annotation)
ANNOTATION_DIR <- here::here(params$run_dir, "07_module_annotation")
top_bp_file <- file.path(ANNOTATION_DIR, "top_BP_per_module_rrvgo.csv")

if (file.exists(top_bp_file)) {
  top_bp_terms <- read_csv(top_bp_file, show_col_types = FALSE)
  cat("\n=== GO BP ANNOTATIONS ===\n")
  cat("Modules with GO BP annotation:", nrow(top_bp_terms), "\n")
} else {
  top_bp_terms <- NULL
  cat("\n=== GO BP ANNOTATIONS ===\n")
  cat("Note: No GO annotations found (run 07_module_annotation.Rmd first)\n")
}
```

---

# Validate Gene Alignment

```{r validate_genes}
# Check genes are aligned
stopifnot(
  "Genes must match between adjacency matrices" =
    all(rownames(A_ctrl) == rownames(A_inv4m))
)

gene_names <- rownames(A_ctrl)
n_genes <- length(gene_names)

cat("Genes aligned:", n_genes, "\n")
```

---

# Compute Consensus kWithin

**kWithin from Consensus Adjacency**: Mean adjacency of a gene to other genes
in its module, using the consensus adjacency matrix.

```{r consensus_kwithin}
#' Compute intramodular connectivity from consensus adjacency
#'
#' @param A Consensus adjacency matrix
#' @param modules Named vector of module assignments
#'
#' @return Named vector of kWithin values
compute_consensus_kWithin <- function(A, modules) {
  genes <- names(modules)

  kWithin <- sapply(genes, function(g) {
    mod <- modules[g]
    if (mod == 0) return(NA)

    mod_genes <- names(modules)[modules == mod]
    # Mean adjacency to other genes in module (exclude self)
    other_genes <- setdiff(mod_genes, g)
    if (length(other_genes) == 0) return(0)
    mean(A[g, other_genes])
  })

  return(kWithin)
}

# Create named module vectors
ctrl_mod_vec <- setNames(ctrl_modules$module, ctrl_modules$gene)

# Compute kWithin using CTRL modules as reference
k_ctrl <- compute_consensus_kWithin(A_ctrl, ctrl_mod_vec)
k_inv4m <- compute_consensus_kWithin(A_inv4m, ctrl_mod_vec)

# Gene-level delta kWithin
delta_k <- tibble(
  gene = names(k_ctrl),
  module = ctrl_mod_vec[names(k_ctrl)],
  kWithin_ctrl = k_ctrl,
  kWithin_inv4m = k_inv4m,
  delta_kWithin = k_inv4m - k_ctrl
) %>%
  filter(!is.na(module), module != 0)

# Add module colors
delta_k <- delta_k %>%
  left_join(ctrl_modules %>% dplyr::select(gene, color), by = "gene")

write_csv(delta_k, file.path(OUTPUT_DIR, "delta_kWithin_genes.csv"))

{
  cat("=== DELTA kWITHIN (GENE-LEVEL) ===\n")
  cat("Genes analyzed:", nrow(delta_k), "\n")
  cat("Genes with decreased connectivity:", sum(delta_k$delta_kWithin < 0, na.rm = TRUE), "\n")
  cat("Genes with increased connectivity:", sum(delta_k$delta_kWithin > 0, na.rm = TRUE), "\n")
  cat("\nDelta kWithin distribution:\n")
  print(summary(delta_k$delta_kWithin))
}
```

---

# Module-Level Adjacency Preservation

**Approach**: For each CTRL module, compute:
1. Correlation of within-module adjacency values between A_ctrl and A_inv4m
2. Mean change in adjacency

```{r module_adjacency}
#' Compare module adjacency between conditions
#'
#' @param A_ctrl CTRL consensus adjacency
#' @param A_inv4m Inv4m consensus adjacency
#' @param modules_ctrl Named vector of CTRL module assignments
#'
#' @return Data frame with preservation metrics per module
compare_module_adjacency <- function(A_ctrl, A_inv4m, modules_ctrl) {
  modules <- sort(unique(as.integer(modules_ctrl[modules_ctrl != 0])))

  # Get genes present in both adjacency matrices
  shared_genes <- intersect(rownames(A_ctrl), rownames(A_inv4m))

  preservation_list <- lapply(modules, function(mod) {
    mod_genes <- names(modules_ctrl)[modules_ctrl == mod]

    # Filter to genes in both adjacency matrices
    genes <- intersect(mod_genes, shared_genes)

    if (length(genes) < 3) {
      return(data.frame(
        module = as.integer(mod),
        n_genes = as.integer(length(mod_genes)),
        n_genes_analyzed = as.integer(length(genes)),
        adjacency_correlation = NA_real_,
        mean_adjacency_delta = NA_real_,
        mean_adjacency_ctrl = NA_real_,
        mean_adjacency_inv4m = NA_real_,
        stringsAsFactors = FALSE
      ))
    }

    # Extract within-module adjacency (lower triangle only)
    A_ctrl_mod <- A_ctrl[genes, genes]
    A_inv4m_mod <- A_inv4m[genes, genes]

    ctrl_vals <- A_ctrl_mod[lower.tri(A_ctrl_mod)]
    inv4m_vals <- A_inv4m_mod[lower.tri(A_inv4m_mod)]

    # Correlation of adjacency values
    cor_adj <- as.numeric(cor(ctrl_vals, inv4m_vals, use = "pairwise.complete.obs"))

    # Mean adjacency values
    mean_ctrl <- as.numeric(mean(ctrl_vals, na.rm = TRUE))
    mean_inv4m <- as.numeric(mean(inv4m_vals, na.rm = TRUE))

    data.frame(
      module = as.integer(mod),
      n_genes = as.integer(length(mod_genes)),
      n_genes_analyzed = as.integer(length(genes)),
      adjacency_correlation = cor_adj,
      mean_adjacency_delta = as.numeric(mean_inv4m - mean_ctrl),
      mean_adjacency_ctrl = mean_ctrl,
      mean_adjacency_inv4m = mean_inv4m,
      stringsAsFactors = FALSE
    )
  })

  # Combine all results
  result <- do.call(rbind, preservation_list)

  # Report on gene coverage
  total_genes <- sum(result$n_genes)
  analyzed_genes <- sum(result$n_genes_analyzed)
  if (analyzed_genes < total_genes) {
    message(sprintf("Note: %d/%d genes analyzed (%.1f%% - some genes not in both adjacencies)",
                    analyzed_genes, total_genes, 100 * analyzed_genes / total_genes))
  }

  result
}

# Compute module-level adjacency preservation
adjacency_preservation <- compare_module_adjacency(A_ctrl, A_inv4m, ctrl_mod_vec)

# Add module colors
adjacency_preservation <- adjacency_preservation %>%
  left_join(
    ctrl_modules %>% distinct(module, color),
    by = "module"
  )

{
  cat("\n=== ADJACENCY PRESERVATION ===\n")
  cat("Modules analyzed:", nrow(adjacency_preservation), "\n")
  cat("\nAdjacency correlation distribution:\n")
  print(summary(adjacency_preservation$adjacency_correlation))
}
```

---

# Module-Level Delta kWithin

```{r module_delta_k}
# Aggregate to module level
module_delta_k <- delta_k %>%
  group_by(module, color) %>%
  summarise(
    n_genes = n(),
    median_delta_k = median(delta_kWithin, na.rm = TRUE),
    mean_delta_k = mean(delta_kWithin, na.rm = TRUE),
    sd_delta_k = sd(delta_kWithin, na.rm = TRUE),
    pct_decreased = mean(delta_kWithin < 0, na.rm = TRUE) * 100,
    .groups = "drop"
  )

write_csv(module_delta_k, file.path(OUTPUT_DIR, "delta_kWithin_modules.csv"))

{
  cat("\n=== MODULE-LEVEL DELTA kWITHIN ===\n")
  print(module_delta_k %>% arrange(median_delta_k))
}
```

## Statistical Test for Connectivity Change

**Approach**: For each module, test whether the mean Δ kWithin is significantly
different from zero using a one-sample t-test. Apply FDR correction across all
modules.

```{r delta_k_ttest}
# One-sample t-test per module: H0: mean(delta_kWithin) = 0
delta_k_tests <- delta_k %>%
  group_by(module, color) %>%
  rstatix::t_test(delta_kWithin ~ 1, mu = 0, alternative = "two.sided") %>%
  ungroup() %>%
  rstatix::adjust_pvalue(method = "fdr") %>%
  dplyr::select(module, color, statistic, df, p, p.adj) %>%
  dplyr::rename(t_statistic = statistic, p_value = p, p_adj_fdr = p.adj)

# Merge with module_delta_k
module_delta_k <- module_delta_k %>%
  left_join(
    delta_k_tests %>% dplyr::select(module, t_statistic, p_value, p_adj_fdr),
    by = "module"
  )

# Update CSV with t-test results
write_csv(module_delta_k, file.path(OUTPUT_DIR, "delta_kWithin_modules.csv"))

{
  cat("\n=== T-TEST: Δ kWithin ≠ 0 (FDR-corrected) ===\n")
  cat("Modules with significant connectivity change (FDR < 0.05):",
      sum(module_delta_k$p_adj_fdr < 0.05, na.rm = TRUE), "of",
      nrow(module_delta_k), "\n\n")
  print(
    module_delta_k %>%
      dplyr::select(module, color, n_genes, mean_delta_k, t_statistic, p_adj_fdr) %>%
      mutate(
        mean_delta_k = round(mean_delta_k, 3),
        t_statistic = round(t_statistic, 2),
        p_adj_fdr = signif(p_adj_fdr, 3),
        sig = ifelse(p_adj_fdr < 0.05, "*", "")
      ) %>%
      arrange(p_adj_fdr)
  )
}
```

---

# Module Mapping Between Conditions

**Approach**: Map CTRL modules to Inv4m modules using Jaccard overlap.

```{r module_mapping}
# Create mapping between CTRL and Inv4m modules
ctrl_mod_list <- split(ctrl_modules$gene, ctrl_modules$module)
ctrl_mod_list <- ctrl_mod_list[names(ctrl_mod_list) != "0"]

inv4m_mod_list <- split(inv4m_modules$gene, inv4m_modules$module)
inv4m_mod_list <- inv4m_mod_list[names(inv4m_mod_list) != "0"]

# Compute Jaccard overlap
module_mapping <- expand_grid(
  ctrl_module = names(ctrl_mod_list),
  inv4m_module = names(inv4m_mod_list)
) %>%
  rowwise() %>%
  mutate(
    ctrl_genes = list(ctrl_mod_list[[ctrl_module]]),
    inv4m_genes = list(inv4m_mod_list[[inv4m_module]]),
    n_overlap = length(intersect(ctrl_genes, inv4m_genes)),
    n_ctrl = length(ctrl_genes),
    n_inv4m = length(inv4m_genes),
    jaccard = n_overlap / length(union(ctrl_genes, inv4m_genes)),
    overlap_pct_ctrl = n_overlap / n_ctrl * 100
  ) %>%
  ungroup() %>%
  dplyr::select(-ctrl_genes, -inv4m_genes)

# Best match for each CTRL module
best_matches <- module_mapping %>%
  group_by(ctrl_module) %>%
  slice_max(jaccard, n = 1) %>%
  ungroup() %>%
  arrange(as.integer(ctrl_module))

write_csv(best_matches, file.path(OUTPUT_DIR, "module_mapping.csv"))

{
  cat("\n=== MODULE MAPPING (CTRL → INV4M) ===\n")
  print(best_matches)
}
```

---

# Integrate All Metrics

```{r integrate_metrics}
# Ensure consistent module types (all as integer)
adjacency_preservation <- adjacency_preservation %>%
  mutate(module = as.integer(module))

module_delta_k <- module_delta_k %>%
  mutate(module = as.integer(module))

ctrl_support <- ctrl_support %>%
  mutate(module = as.integer(module))

best_matches <- best_matches %>%
  mutate(ctrl_module = as.integer(ctrl_module))

# Combine all preservation metrics (join by module only - n_genes may differ slightly)
preservation_summary <- adjacency_preservation %>%
  left_join(
    module_delta_k %>% dplyr::select(-n_genes),
    by = c("module", "color")
  ) %>%
  left_join(
    ctrl_support %>% dplyr::select(module, ctrl_lfp = mean_lfp, ctrl_stability = stability),
    by = "module"
  ) %>%
  left_join(
    best_matches %>%
      dplyr::select(ctrl_module, best_inv4m_match = inv4m_module, jaccard),
    by = c("module" = "ctrl_module")
  ) %>%
  arrange(adjacency_correlation)

# Identify disrupted modules (statistically supported by t-test)
# Criteria: adjacency correlation < 0.5 OR significant connectivity loss
disrupted_modules <- preservation_summary %>%
  filter(
    adjacency_correlation < 0.5 |
    (median_delta_k < -0.1 & pct_decreased > 60)
  )

# Ensure all columns are simple types (no lists)
preservation_summary <- preservation_summary %>%
  mutate(
    across(where(is.list), ~ sapply(.x, function(x) if(length(x) == 0) NA else x[1])),
    ctrl_stability = as.character(ctrl_stability)
  )

# Save results
write_csv(preservation_summary, file.path(OUTPUT_DIR, "preservation_summary.csv"))
write_csv(disrupted_modules, file.path(OUTPUT_DIR, "disrupted_modules.csv"))
saveRDS(preservation_summary, file.path(OUTPUT_DIR, "preservation_summary.rds"))

# Generate LaTeX table for Overleaf (sorted by FDR p-value)
preservation_latex <- preservation_summary %>%
  arrange(p_adj_fdr) %>%
  dplyr::select(
    color, n_genes, adjacency_correlation, median_delta_k, p_adj_fdr, ctrl_lfp
  )

# Format p-values for LaTeX scientific notation
format_pval_latex <- function(p) {
  if (is.na(p)) return("---")
  if (p < 0.001) {
    exp <- floor(log10(p))
    mant <- p / 10^exp
    return(sprintf("$%.1f \\times 10^{%d}$", mant, exp))
  } else {
    return(sprintf("%.3f", p))
  }
}

# Build LaTeX table with proper p-value formatting
latex_lines <- c(
  "\\begin{table}[!ht]",
  "\\caption{\\textbf{WGCNA module preservation statistics.} Modules sorted by FDR-corrected p-value from one-sample t-test of $\\Delta k_{Within} \\neq 0$. Adj.\\ Corr.\\ = adjacency correlation between CTRL and Inv4m networks. LFP = Largest Fragment Proportion (bootstrap support in CTRL).}",
  "\\label{tab::preservation}",
  "\\centering",
  "\\small",
  "\\begin{tabular}{@{}lrrrrr@{}}",
  "\\toprule",
  "\\textbf{Module} & \\textbf{Genes} & \\textbf{Adj. Corr.} & \\textbf{$\\Delta k_{Within}$} & \\textbf{FDR} & \\textbf{LFP} \\\\",
  "\\midrule"
)

for (i in 1:nrow(preservation_latex)) {
  row <- preservation_latex[i, ]
  p_formatted <- format_pval_latex(row$p_adj_fdr)
  latex_lines <- c(latex_lines, sprintf(
    "%s & %d & %.2f & %.2f & %s & %.2f \\\\",
    row$color, row$n_genes, round(row$adjacency_correlation, 2),
    round(row$median_delta_k, 2), p_formatted, round(row$ctrl_lfp, 2)
  ))
}

latex_lines <- c(latex_lines, "\\bottomrule", "\\end{tabular}", "\\end{table}")

writeLines(latex_lines, file.path(OUTPUT_DIR, "preservation_table.tex"))
cat("Saved: preservation_table.tex\n")

{
  cat("\n=== INTEGRATED PRESERVATION SUMMARY ===\n")
  cat("Total modules:", nrow(preservation_summary), "\n")
  cat("\nAdjacency correlation range:",
      round(min(preservation_summary$adjacency_correlation), 2), "to",
      round(max(preservation_summary$adjacency_correlation), 2), "\n")
  cat("Modules with adj. correlation < 0.30:",
      sum(preservation_summary$adjacency_correlation < 0.30), "\n")

  cat("\nModules with lowest preservation (adj. corr. < 0.5):\n")
  print(
    preservation_summary %>%
      filter(adjacency_correlation < 0.5) %>%
      dplyr::select(module, color, n_genes, adjacency_correlation,
                    median_delta_k, ctrl_stability, jaccard)
  )
}
```

---

# Visualization: Adjacency Correlation

```{r adjacency_correlation_plot, fig.width=10, fig.height=6}
p_adj_cor <- ggplot(
  preservation_summary,
  aes(x = n_genes, y = adjacency_correlation, fill = color)
) +
  geom_point(size = 4, alpha = 0.8, shape = 21, color = "black") +
  geom_hline(yintercept = c(0.3, 0.5), linetype = "dashed", color = "gray50") +
  geom_text_repel(
    aes(label = color),
    size = 3,
    max.overlaps = 15
  ) +
  scale_fill_identity() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Module Preservation: Adjacency Correlation (CTRL vs Inv4m)",
    subtitle = "Based on consensus adjacency matrices",
    x = "Module Size (genes)",
    y = "Adjacency Correlation"
  )

print(p_adj_cor)

ggsave(
  file.path(OUTPUT_DIR, "adjacency_correlation_plot.pdf"),
  p_adj_cor,
  width = 10, height = 6
)
```

---

# Visualization: Delta kWithin Boxplot

```{r delta_k_boxplot, fig.width=12, fig.height=8}
# Order modules by median delta kWithin
module_order <- module_delta_k %>%
  arrange(median_delta_k) %>%
  pull(color)

# Join median_delta_k for coloring
delta_k_plot <- delta_k %>%
  left_join(
    module_delta_k %>% dplyr::select(color, median_delta_k),
    by = "color"
  ) %>%
  mutate(color = factor(color, levels = module_order))

# Prepare significance annotation from t-test results
sig_annotation <- module_delta_k %>%
  mutate(
    color = factor(color, levels = module_order),
    sig_label = case_when(
      p_adj_fdr < 0.001 ~ "***",
      p_adj_fdr < 0.01 ~ "**",
      p_adj_fdr < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  dplyr::select(color, sig_label)

# Prepare GO term annotation data (if available)
if (!is.null(top_bp_terms) && nrow(top_bp_terms) > 0) {
  go_annotation <- top_bp_terms %>%
    filter(module_color %in% module_order) %>%
    mutate(
      color = factor(module_color, levels = module_order),
      # Truncate long descriptions for display
      short_desc = str_trunc(Description, 30)
    )
  has_go_annotation <- TRUE
} else {
  has_go_annotation <- FALSE
}

# Get y position for significance markers (above each boxplot)
sig_y_pos <- delta_k_plot %>%
  group_by(color) %>%
  summarize(y_max = max(delta_kWithin, na.rm = TRUE), .groups = "drop") %>%
  left_join(sig_annotation, by = "color") %>%
  mutate(y_sig = y_max + 0.08)

# Create boxplot with module colors
p_delta_k <- ggplot(delta_k_plot, aes(x = color, y = delta_kWithin, fill = color)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_text(
    data = sig_y_pos,
    aes(x = color, y = y_sig, label = sig_label),
    inherit.aes = FALSE,
    size = 6,
    vjust = 0
  ) +
  scale_fill_identity() +
  theme_minimal(base_size = 25) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
    plot.title = element_blank(),
    plot.subtitle = element_blank()
  ) +
  labs(
    x = "Module",
    y = "Δ kWithin (Inv4m - CTRL)"
  )

# Add GO term annotation if available
if (has_go_annotation) {
  # Fixed y position for annotation - add as column to data
  go_annotation <- go_annotation %>%
    mutate(y_pos = 0.5)

  p_delta_k_annotated <- p_delta_k +
    geom_text(
      data = go_annotation,
      aes(x = color, y = y_pos, label = short_desc),
      vjust = 0,
      hjust = 0,
      size = 5,
      angle = 45,
      inherit.aes = FALSE
    ) +
    coord_cartesian(clip = "off", ylim = c(NA, 1.1)) +
    theme(plot.margin = margin(t = 120, r = 10, b = 10, l = 10))

  print(p_delta_k_annotated)

  # Save annotated version
  ggsave(
    file.path(OUTPUT_DIR, "delta_kWithin_boxplot_annotated.pdf"),
    p_delta_k_annotated,
    width = 12, height = 10
  )
  ggsave(
    file.path(OUTPUT_DIR, "delta_kWithin_boxplot_annotated.png"),
    p_delta_k_annotated,
    width = 12, height = 10, dpi = 300
  )
  saveRDS(p_delta_k_annotated, file.path(OUTPUT_DIR, "delta_kWithin_boxplot_annotated.rds"))
  cat("Saved: delta_kWithin_boxplot_annotated.pdf/png/rds\n")
} else {
  print(p_delta_k)
}

# Also save simple version (without GO annotation)
ggsave(
  file.path(OUTPUT_DIR, "delta_kWithin_boxplot.pdf"),
  p_delta_k,
  width = 12, height = 6
)
ggsave(
  file.path(OUTPUT_DIR, "delta_kWithin_boxplot.png"),
  p_delta_k,
  width = 12, height = 6, dpi = 300
)
saveRDS(p_delta_k, file.path(OUTPUT_DIR, "delta_kWithin_boxplot.rds"))
cat("Saved: delta_kWithin_boxplot.pdf/png/rds\n")
```

---

# Visualization: Correlation vs Delta kWithin

```{r correlation_vs_delta, fig.width=10, fig.height=6}
p_compare <- ggplot(
  preservation_summary,
  aes(x = median_delta_k, y = adjacency_correlation, fill = color)
) +
  geom_point(aes(size = n_genes), alpha = 0.8, shape = 21, color = "black") +
  geom_hline(yintercept = 0.3, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_text_repel(
    aes(label = color),
    size = 3,
    max.overlaps = 15
  ) +
  scale_fill_identity() +
  scale_size_continuous(range = c(3, 10)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Module Perturbation: Adjacency Correlation vs Δ kWithin",
    subtitle = "Lower values indicate greater perturbation",
    x = "Median Δ kWithin",
    y = "Adjacency Correlation",
    size = "Module Size"
  )

print(p_compare)

ggsave(
  file.path(OUTPUT_DIR, "correlation_vs_delta_k.pdf"),
  p_compare,
  width = 10, height = 6
)
```

---

# Visualization: Support vs Preservation

```{r support_vs_preservation, fig.width=10, fig.height=6}
p_support_pres <- ggplot(
  preservation_summary,
  aes(x = ctrl_lfp, y = adjacency_correlation, fill = color)
) +
  geom_point(aes(size = n_genes), alpha = 0.8, shape = 21, color = "black") +
  geom_hline(yintercept = 0.3, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "gray50") +
  geom_text_repel(
    aes(label = color),
    size = 3,
    max.overlaps = 15
  ) +
  scale_fill_identity() +
  scale_size_continuous(range = c(3, 10)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Module Bootstrap Support vs Preservation",
    subtitle = "X: CTRL module stability | Y: Adjacency correlation in Inv4m",
    x = "CTRL Bootstrap Support (LFP)",
    y = "Adjacency Correlation",
    size = "Module Size"
  )

print(p_support_pres)

ggsave(
  file.path(OUTPUT_DIR, "support_vs_preservation.pdf"),
  p_support_pres,
  width = 10, height = 6
)
```

---

# Save Summary Statistics

```{r save_summary}
summary_stats <- tibble(
  metric = c(
    "Total CTRL modules analyzed",
    "Modules with adj. correlation < 0.30",
    "Modules with adj. correlation 0.30-0.50",
    "Modules with adj. correlation >= 0.50",
    "Min adjacency correlation",
    "Max adjacency correlation",
    "Mean adjacency correlation",
    "Median adjacency correlation",
    "Genes with decreased kWithin",
    "Genes with increased kWithin",
    "Mean delta kWithin (gene-level)"
  ),
  value = c(
    nrow(preservation_summary),
    sum(preservation_summary$adjacency_correlation < 0.30),
    sum(preservation_summary$adjacency_correlation >= 0.30 & preservation_summary$adjacency_correlation < 0.50),
    sum(preservation_summary$adjacency_correlation >= 0.50),
    round(min(preservation_summary$adjacency_correlation, na.rm = TRUE), 3),
    round(max(preservation_summary$adjacency_correlation, na.rm = TRUE), 3),
    round(mean(preservation_summary$adjacency_correlation, na.rm = TRUE), 3),
    round(median(preservation_summary$adjacency_correlation, na.rm = TRUE), 3),
    sum(delta_k$delta_kWithin < 0, na.rm = TRUE),
    sum(delta_k$delta_kWithin > 0, na.rm = TRUE),
    round(mean(delta_k$delta_kWithin, na.rm = TRUE), 4)
  )
)

write_csv(summary_stats, file.path(OUTPUT_DIR, "preservation_analysis_summary.csv"))

{
  cat("\n=== SAVED OUTPUTS ===\n")
  cat("preservation_summary.csv: integrated module metrics\n")
  cat("preservation_summary.rds: R object for downstream\n")
  cat("disrupted_modules.csv: modules with adj. corr. < 0.5 or significant Δk\n")
  cat("delta_kWithin_genes.csv: gene-level connectivity change\n")
  cat("delta_kWithin_modules.csv: module-level summary\n")
  cat("module_mapping.csv: CTRL → Inv4m module correspondence\n")
  cat("adjacency_correlation_plot.pdf\n")
  cat("delta_kWithin_boxplot.pdf\n")
  cat("correlation_vs_delta_k.pdf\n")
  cat("support_vs_preservation.pdf\n")
  cat("preservation_analysis_summary.csv\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Module coverage (allow some missing due to gene overlap issues)
  cat("Check 1: Module coverage\n")
  n_ctrl_mods <- length(unique(ctrl_modules$module[ctrl_modules$module != 0]))
  n_analyzed <- nrow(preservation_summary)
  coverage_pct <- round(100 * n_analyzed / n_ctrl_mods, 1)
  cat("  CTRL modules analyzed:", n_analyzed, "/", n_ctrl_mods,
      "(", coverage_pct, "%)\n")

  # Warn if coverage is low, fail only if <80%

  if (n_analyzed < n_ctrl_mods) {
    missing_mods <- setdiff(
      unique(ctrl_modules$module[ctrl_modules$module != 0]),
      preservation_summary$module
    )
    cat("  WARNING: Missing modules:", paste(missing_mods, collapse = ", "), "\n")
  }
  if (coverage_pct < 80) {
    stop("Module coverage below 80% - check gene overlap between conditions")
  }

  # Check 2: Delta kWithin calculated
  cat("\nCheck 2: Delta kWithin coverage\n")
  n_assigned_genes <- sum(ctrl_modules$module != 0)
  cat("  Genes with delta_k:", nrow(delta_k), "/", n_assigned_genes, "\n")

  # Check 3: Adjacency correlation valid
  cat("\nCheck 3: Adjacency correlation valid\n")
  valid_cors <- preservation_summary$adjacency_correlation[
    !is.na(preservation_summary$adjacency_correlation)
  ]
  if (length(valid_cors) > 0) {
    cat("  Range: [", round(min(valid_cors), 3), ",",
        round(max(valid_cors), 3), "]\n")
    cat("  Modules with NA correlation:", sum(is.na(preservation_summary$adjacency_correlation)),
        "(likely <3 genes)\n")
  }

  # Check 4: Module mapping coverage
  cat("\nCheck 4: Module mapping\n")
  cat("  Modules mapped:", nrow(best_matches), "/", n_ctrl_mods, "\n")

  cat("\n=== VALIDATION COMPLETE ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
