---
title: "Field-Only Perturbation Analysis: Module Preservation"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_permutations: 1000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Quantify module perturbation in Inv4m samples using two
complementary metrics:

1. **Zsummary** from WGCNA `modulePreservation()` — tests whether modules
   defined in CTRL are preserved in Inv4m (permutation-based, module-level)
2. **Delta kWithin** from standard `intramodularConnectivity()` — measures
   change in connectivity magnitude (parametric t-test, gene-level)

**Interpretation**:
- **Zsummary < 2**: Module structure NOT preserved (Langfelder et al. 2011)
- **Zsummary 2-10**: Moderate preservation
- **Zsummary > 10**: Strong preservation
- **Significant negative delta kWithin**: Module connectivity decreased in Inv4m

**Inputs**: Consensus module assignments (from step 04), filtered expression
matrix (from step 02), shared soft power beta = 18 (from step 03).

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(WGCNA)
  library(here)
  library(patchwork)
  library(ggrepel)
  library(rstatix)
})

# Enable multi-threading
allowWGCNAThreads()

set.seed(123)
```

```{r setup_directories}
# Input/output directories
CONSENSUS_DIR <- here::here(params$run_dir, "04_consensus_networks")
BOOTSTRAP_DIR <- here::here(params$run_dir, "05_bootstrap_support")
REFERENCE_DIR <- here::here(params$run_dir, "03_reference_network")
FILTER_DIR <- here::here(params$run_dir, "02_gene_filter")
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
OUTPUT_DIR <- here::here(params$run_dir, "06_preservation")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n")
  cat("Permutations:", params$n_permutations, "\n\n")
}
```

---

# Load Data

```{r load_consensus_modules}
# Load consensus module assignments (from bootstrap consensus step 04)
ctrl_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)
inv4m_modules <- read_csv(
  file.path(CONSENSUS_DIR, "inv4m", "consensus_modules.csv"),
  show_col_types = FALSE
)

{
  cat("=== CONSENSUS MODULES ===\n")
  cat("CTRL modules:", length(unique(ctrl_modules$module[ctrl_modules$module != 0])), "\n")
  cat("CTRL genes:", nrow(ctrl_modules), "\n")
  cat("Inv4m modules:", length(unique(inv4m_modules$module[inv4m_modules$module != 0])), "\n")
}
```

```{r load_expression}
# Load filtered expression matrix (genes x samples)
expr_filtered <- readRDS(file.path(FILTER_DIR, "filtered_expression.rds"))

# Load sample metadata
sample_meta <- read_csv(
  file.path(DATA_PREP_DIR, "sample_metadata.csv"),
  show_col_types = FALSE
)

{
  cat("\n=== EXPRESSION DATA ===\n")
  cat("Expression matrix:", nrow(expr_filtered), "genes x",
      ncol(expr_filtered), "samples\n")
  cat("Sample breakdown:\n")
  print(table(sample_meta$Genotype))
}
```

```{r load_soft_power}
# Load empirically selected soft power (both conditions converge on same value)
power_ctrl <- readRDS(file.path(REFERENCE_DIR, "ctrl", "selected_power.rds"))
power_inv4m <- readRDS(file.path(REFERENCE_DIR, "inv4m", "selected_power.rds"))

# Use shared power for comparability
stopifnot(
  "Soft powers must match for valid comparison" = power_ctrl == power_inv4m
)
SOFT_POWER <- power_ctrl

cat("\n=== SOFT POWER ===\n")
cat("CTRL:", power_ctrl, "| Inv4m:", power_inv4m, "| Shared:", SOFT_POWER, "\n")
```

```{r load_support}
# Load bootstrap support metrics
ctrl_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "ctrl", "module_support.csv"),
  show_col_types = FALSE
)
inv4m_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "inv4m", "module_support.csv"),
  show_col_types = FALSE
)

{
  cat("\n=== BOOTSTRAP SUPPORT ===\n")
  cat("CTRL stable modules:", sum(ctrl_support$stability == "Stable"), "\n")
  cat("Inv4m stable modules:", sum(inv4m_support$stability == "Stable"), "\n")
}
```

```{r load_go_annotations}
# Load top GO BP terms from 07_module_annotation (for boxplot annotation)
ANNOTATION_DIR <- here::here(params$run_dir, "07_module_annotation")
top_bp_file <- file.path(ANNOTATION_DIR, "top_BP_per_module_rrvgo.csv")

if (file.exists(top_bp_file)) {
  top_bp_terms <- read_csv(top_bp_file, show_col_types = FALSE)
  cat("\n=== GO BP ANNOTATIONS ===\n")
  cat("Modules with GO BP annotation:", nrow(top_bp_terms), "\n")
} else {
  top_bp_terms <- NULL
  cat("\n=== GO BP ANNOTATIONS ===\n")
  cat("Note: No GO annotations found (run 07_module_annotation.Rmd first)\n")
}
```

---

# Prepare Expression Data

```{r prepare_expression}
# Split samples by genotype
ctrl_samples <- sample_meta %>%
  filter(Genotype == "CTRL") %>%
  pull(sample_id)

inv4m_samples <- sample_meta %>%
  filter(Genotype == "Inv4m") %>%
  pull(sample_id)

# Transpose to samples x genes (WGCNA convention)
datExpr_CTRL <- t(expr_filtered[, ctrl_samples])
datExpr_Inv4m <- t(expr_filtered[, inv4m_samples])

# Verify gene order matches
stopifnot(
  "Gene order mismatch" = all(colnames(datExpr_CTRL) == colnames(datExpr_Inv4m))
)

# Create module color vector from consensus assignments (CTRL as reference)
module_colors <- setNames(ctrl_modules$color, ctrl_modules$gene)
# Align to expression matrix column order
module_colors <- module_colors[colnames(datExpr_CTRL)]

{
  cat("=== PREPARED DATA ===\n")
  cat("CTRL:", nrow(datExpr_CTRL), "samples x", ncol(datExpr_CTRL), "genes\n")
  cat("Inv4m:", nrow(datExpr_Inv4m), "samples x", ncol(datExpr_Inv4m), "genes\n")
  cat("Module assignments:", sum(!is.na(module_colors)), "genes\n")
  cat("Non-grey modules:", length(unique(module_colors[module_colors != "grey"])), "\n")
}
```

---

# Module Preservation Analysis (Zsummary)

**Approach**: Use WGCNA's `modulePreservation()` to statistically test
whether modules defined in CTRL are preserved in Inv4m. This is the
standard method from Langfelder et al. (2011).

```{r module_preservation, cache=TRUE}
# Set up multiExpr format
multiExpr <- list(
  CTRL = list(data = datExpr_CTRL),
  Inv4m = list(data = datExpr_Inv4m)
)

# Module colors (reference = CTRL consensus labels)
multiColor <- list(CTRL = module_colors)

cat("Running modulePreservation with", params$n_permutations, "permutations...\n")
cat("This may take several minutes.\n\n")

mp <- modulePreservation(
  multiExpr,
  multiColor,
  referenceNetworks = 1,
  testNetworks = 2,
  networkType = "signed",
  nPermutations = params$n_permutations,
  randomSeed = 123,
  quickCor = 0,
  verbose = 3
)

saveRDS(mp, file.path(OUTPUT_DIR, "modulePreservation_results.rds"))
cat("\nModule preservation analysis complete.\n")
```

```{r extract_zsummary}
# Extract Z statistics
z_stats <- mp$preservation$Z$ref.CTRL$inColumnsAlsoPresentIn.Inv4m

# Extract observed statistics
obs_stats <- mp$preservation$observed$ref.CTRL$inColumnsAlsoPresentIn.Inv4m

# Build Zsummary table (exclude grey module = row "0")
zsummary_table <- tibble(
  module_label = rownames(z_stats),
  module_size = z_stats$moduleSize,
  Zsummary = z_stats$Zsummary.pres,
  Zdensity = z_stats$Zdensity.pres,
  Zconnectivity = z_stats$Zconnectivity.pres,
  medianRank_pres = obs_stats$medianRank.pres
) %>%
  filter(module_label != "gold") %>%  # gold = unassigned in modulePreservation
  mutate(
    preservation_status = case_when(
      Zsummary < 2 ~ "Not preserved",
      Zsummary < 10 ~ "Moderate",
      TRUE ~ "Preserved"
    )
  ) %>%
  arrange(Zsummary)

write_csv(zsummary_table, file.path(OUTPUT_DIR, "zsummary_table.csv"))

{
  cat("=== ZSUMMARY STATISTICS ===\n")
  cat("\nPreservation status:\n")
  print(table(zsummary_table$preservation_status))
  cat("\n")
  print(
    zsummary_table %>%
      dplyr::select(module_label, module_size, Zsummary, preservation_status) %>%
      mutate(Zsummary = round(Zsummary, 2))
  )
}
```

---

# Delta kWithin Analysis (Standard WGCNA)

**Approach**: Compute adjacency matrices from expression data using the
shared empirical soft power (beta = `r SOFT_POWER`), then calculate
intramodular connectivity using `intramodularConnectivity()` from WGCNA.
The CTRL consensus module assignments are used as reference for both
conditions, ensuring comparability.

```{r compute_adjacency}
# Compute signed adjacency matrices using shared soft power
cat("Computing adjacency matrices with soft power =", SOFT_POWER, "...\n")
adj_CTRL <- adjacency(datExpr_CTRL, power = SOFT_POWER, type = "signed")
adj_Inv4m <- adjacency(datExpr_Inv4m, power = SOFT_POWER, type = "signed")

cat("Adjacency matrices:", nrow(adj_CTRL), "x", ncol(adj_CTRL), "\n")
```

```{r compute_kwithin}
# Compute intramodular connectivity using CTRL module assignments for both
k_CTRL <- intramodularConnectivity(adj_CTRL, module_colors)
k_Inv4m <- intramodularConnectivity(adj_Inv4m, module_colors)

# Build gene-level delta kWithin table
delta_k <- tibble(
  gene = rownames(k_CTRL),
  module = as.integer(ctrl_modules$module[match(rownames(k_CTRL), ctrl_modules$gene)]),
  color = module_colors[rownames(k_CTRL)],
  kWithin_ctrl = k_CTRL$kWithin,
  kWithin_inv4m = k_Inv4m$kWithin,
  delta_kWithin = k_Inv4m$kWithin - k_CTRL$kWithin
) %>%
  filter(!is.na(module), module != 0)

write_csv(delta_k, file.path(OUTPUT_DIR, "delta_kWithin_genes.csv"))

{
  cat("=== DELTA kWITHIN (GENE-LEVEL) ===\n")
  cat("Genes analyzed:", nrow(delta_k), "\n")
  cat("Genes with decreased connectivity:", sum(delta_k$delta_kWithin < 0, na.rm = TRUE), "\n")
  cat("Genes with increased connectivity:", sum(delta_k$delta_kWithin > 0, na.rm = TRUE), "\n")
  cat("\nDelta kWithin distribution:\n")
  print(summary(delta_k$delta_kWithin))
}
```

---

# Module-Level Delta kWithin

```{r module_delta_k}
# Aggregate to module level
module_delta_k <- delta_k %>%
  group_by(module, color) %>%
  summarise(
    n_genes = n(),
    median_delta_k = median(delta_kWithin, na.rm = TRUE),
    mean_delta_k = mean(delta_kWithin, na.rm = TRUE),
    sd_delta_k = sd(delta_kWithin, na.rm = TRUE),
    pct_decreased = mean(delta_kWithin < 0, na.rm = TRUE) * 100,
    .groups = "drop"
  )

write_csv(module_delta_k, file.path(OUTPUT_DIR, "delta_kWithin_modules.csv"))

{
  cat("\n=== MODULE-LEVEL DELTA kWITHIN ===\n")
  print(module_delta_k %>% arrange(median_delta_k))
}
```

## Statistical Test for Connectivity Change

**Approach**: For each module, test whether the mean delta kWithin is
significantly different from zero using a one-sample t-test. Apply FDR
correction across all modules.

```{r delta_k_ttest}
# One-sample t-test per module: H0: mean(delta_kWithin) = 0
delta_k_tests <- delta_k %>%
  group_by(module, color) %>%
  rstatix::t_test(delta_kWithin ~ 1, mu = 0, alternative = "two.sided") %>%
  ungroup() %>%
  rstatix::adjust_pvalue(method = "fdr") %>%
  dplyr::select(module, color, statistic, df, p, p.adj) %>%
  dplyr::rename(t_statistic = statistic, p_value = p, p_adj_fdr = p.adj)

# Merge with module_delta_k
module_delta_k <- module_delta_k %>%
  left_join(
    delta_k_tests %>% dplyr::select(module, t_statistic, p_value, p_adj_fdr),
    by = "module"
  )

# Update CSV with t-test results
write_csv(module_delta_k, file.path(OUTPUT_DIR, "delta_kWithin_modules.csv"))

{
  cat("\n=== T-TEST: delta kWithin != 0 (FDR-corrected) ===\n")
  cat("Modules with significant connectivity change (FDR < 0.05):",
      sum(module_delta_k$p_adj_fdr < 0.05, na.rm = TRUE), "of",
      nrow(module_delta_k), "\n\n")
  print(
    module_delta_k %>%
      dplyr::select(module, color, n_genes, mean_delta_k, t_statistic, p_adj_fdr) %>%
      mutate(
        mean_delta_k = round(mean_delta_k, 3),
        t_statistic = round(t_statistic, 2),
        p_adj_fdr = signif(p_adj_fdr, 3),
        sig = ifelse(p_adj_fdr < 0.05, "*", "")
      ) %>%
      arrange(p_adj_fdr)
  )
}
```

---

# Module Mapping Between Conditions

**Approach**: Map CTRL modules to Inv4m modules using Jaccard overlap.

```{r module_mapping}
# Create mapping between CTRL and Inv4m modules
ctrl_mod_list <- split(ctrl_modules$gene, ctrl_modules$module)
ctrl_mod_list <- ctrl_mod_list[names(ctrl_mod_list) != "0"]

inv4m_mod_list <- split(inv4m_modules$gene, inv4m_modules$module)
inv4m_mod_list <- inv4m_mod_list[names(inv4m_mod_list) != "0"]

# Compute Jaccard overlap
module_mapping <- expand_grid(
  ctrl_module = names(ctrl_mod_list),
  inv4m_module = names(inv4m_mod_list)
) %>%
  rowwise() %>%
  mutate(
    ctrl_genes = list(ctrl_mod_list[[ctrl_module]]),
    inv4m_genes = list(inv4m_mod_list[[inv4m_module]]),
    n_overlap = length(intersect(ctrl_genes, inv4m_genes)),
    n_ctrl = length(ctrl_genes),
    n_inv4m = length(inv4m_genes),
    jaccard = n_overlap / length(union(ctrl_genes, inv4m_genes)),
    overlap_pct_ctrl = n_overlap / n_ctrl * 100
  ) %>%
  ungroup() %>%
  dplyr::select(-ctrl_genes, -inv4m_genes)

# Best match for each CTRL module
best_matches <- module_mapping %>%
  group_by(ctrl_module) %>%
  slice_max(jaccard, n = 1) %>%
  ungroup() %>%
  arrange(as.integer(ctrl_module))

write_csv(best_matches, file.path(OUTPUT_DIR, "module_mapping.csv"))

{
  cat("\n=== MODULE MAPPING (CTRL -> INV4M) ===\n")
  print(best_matches)
}
```

---

# Integrate All Metrics

```{r integrate_metrics}
# Ensure consistent types
module_delta_k <- module_delta_k %>%
  mutate(module = as.integer(module))

ctrl_support <- ctrl_support %>%
  mutate(module = as.integer(module))

best_matches <- best_matches %>%
  mutate(ctrl_module = as.integer(ctrl_module))

# Map module integer to color for Zsummary join
module_color_map <- ctrl_modules %>%
  distinct(module, color) %>%
  mutate(module = as.integer(module))

zsummary_for_join <- zsummary_table %>%
  left_join(module_color_map, by = c("module_label" = "color")) %>%
  dplyr::select(module, Zsummary, Zdensity, Zconnectivity,
                medianRank_pres, preservation_status)

# Combine all preservation metrics
preservation_summary <- module_delta_k %>%
  left_join(
    zsummary_for_join,
    by = "module"
  ) %>%
  left_join(
    ctrl_support %>% dplyr::select(module, ctrl_lfp = mean_lfp, ctrl_stability = stability),
    by = "module"
  ) %>%
  left_join(
    best_matches %>%
      dplyr::select(ctrl_module, best_inv4m_match = inv4m_module, jaccard),
    by = c("module" = "ctrl_module")
  ) %>%
  mutate(ctrl_stability = as.character(ctrl_stability)) %>%
  arrange(Zsummary)

# Identify disrupted modules (FDR-corrected t-test on delta kWithin)
disrupted_modules <- preservation_summary %>%
  filter(p_adj_fdr < 0.05 & mean_delta_k < 0)

# Save results
write_csv(preservation_summary, file.path(OUTPUT_DIR, "preservation_summary.csv"))
write_csv(disrupted_modules, file.path(OUTPUT_DIR, "disrupted_modules.csv"))
saveRDS(preservation_summary, file.path(OUTPUT_DIR, "preservation_summary.rds"))

{
  cat("\n=== INTEGRATED PRESERVATION SUMMARY ===\n")
  cat("Total modules:", nrow(preservation_summary), "\n")
  cat("Disrupted (FDR < 0.05, negative delta_k):", nrow(disrupted_modules), "\n")
  cat("\nZsummary range:",
      round(min(preservation_summary$Zsummary, na.rm = TRUE), 2), "to",
      round(max(preservation_summary$Zsummary, na.rm = TRUE), 2), "\n")
  cat("\n")
  print(
    preservation_summary %>%
      dplyr::select(module, color, n_genes, Zsummary, median_delta_k,
                    p_adj_fdr, ctrl_lfp) %>%
      mutate(
        Zsummary = round(Zsummary, 2),
        median_delta_k = round(median_delta_k, 3),
        p_adj_fdr = signif(p_adj_fdr, 3),
        ctrl_lfp = round(ctrl_lfp, 2)
      )
  )
}
```

---

# Generate LaTeX Table

```{r latex_table}
# Sorted by FDR p-value. Includes Z_density and Z_connectivity
# (component statistics from Langfelder & Horvath 2011) alongside Z_summary
# so the text's claims about rank-structure preservation can be verified.
#
# Detect which reference network this run represents. The Leaf Gradient
# topological control set uses a distinct color relabeling (sky, peach,
# jade, coral, lavender) to avoid confusion with the Genotype Response
# modules. Output filename and LaTeX label differ accordingly.
is_leaf_gradient <- grepl("leaf_gradient", params$run_dir, fixed = TRUE)

if (is_leaf_gradient) {
  out_filename <- "preservation_table_leaf_gradient.tex"
  latex_label <- "tab::preservation_lg"
  caption_title <- paste(
    "WGCNA module preservation statistics for the Leaf Gradient",
    "topological control set"
  )
  extra_caption <- paste(
    "The Leaf Gradient set groups genes by expression along the leaf",
    "developmental gradient and serves as an orthogonal control for",
    "the Genotype Response modules in Table~\\ref{tab::preservation}."
  )
  color_relabel <- c(
    turquoise = "sky",
    blue      = "peach",
    brown     = "jade",
    yellow    = "coral",
    green     = "lavender"
  )
} else {
  out_filename <- "preservation_table.tex"
  latex_label <- "tab::preservation"
  caption_title <- paste(
    "WGCNA module preservation statistics for the Genotype Response",
    "(Inv4m DEG) reference network"
  )
  extra_caption <- ""
  color_relabel <- NULL
}

preservation_latex <- preservation_summary %>%
  arrange(p_adj_fdr) %>%
  mutate(label_col = if (is.null(color_relabel)) {
    as.character(color)
  } else {
    unname(color_relabel[as.character(color)])
  }) %>%
  dplyr::select(label_col, n_genes, Zdensity, Zconnectivity, Zsummary,
                median_delta_k, p_adj_fdr, ctrl_lfp)

format_pval_latex <- function(p) {
  if (is.na(p)) return("---")
  if (p < 0.001) {
    exp_val <- floor(log10(p))
    mant <- p / 10^exp_val
    return(sprintf("$%.1f \\times 10^{%d}$", mant, exp_val))
  } else {
    return(sprintf("%.3f", p))
  }
}

caption_text <- paste0(
  "\\caption{\\textbf{", caption_title, ".} ",
  "Modules sorted by FDR-corrected $p$-value from one-sample ",
  "$t$-test of $\\Delta k_{Within} \\neq 0$. ",
  "$Z_{density}$, $Z_{connectivity}$, and $Z_{summary}$ from ",
  "\\texttt{modulePreservation()} with ", params$n_permutations,
  " permutations \\cite{langfelder2011}: $Z_{summary} < 2$ = not ",
  "preserved, $2$--$10$ = moderate, $> 10$ = strong. ",
  "$Z_{connectivity}$ quantifies rank preservation of intramodular ",
  "connectivity; $Z_{density}$ quantifies preservation of average ",
  "connectivity strength. LFP = Largest Fragment Proportion ",
  "(bootstrap support in CTRL).",
  if (nzchar(extra_caption)) paste0(" ", extra_caption) else "",
  "}"
)

latex_lines <- c(
  "\\begin{table}[!ht]",
  caption_text,
  paste0("\\label{", latex_label, "}"),
  "\\centering",
  "\\small",
  "\\begin{tabular}{@{}lrrrrrrr@{}}",
  "\\toprule",
  paste(
    "\\textbf{Module} & \\textbf{Genes}",
    "& \\textbf{$Z_{density}$} & \\textbf{$Z_{connectivity}$}",
    "& \\textbf{$Z_{summary}$}",
    "& \\textbf{$\\Delta k_{Within}$} & \\textbf{FDR}",
    "& \\textbf{LFP} \\\\"
  ),
  "\\midrule"
)

for (i in seq_len(nrow(preservation_latex))) {
  row <- preservation_latex[i, ]
  p_formatted <- format_pval_latex(row$p_adj_fdr)
  latex_lines <- c(latex_lines, sprintf(
    "%s & %d & %.1f & %.1f & %.1f & %.2f & %s & %.2f \\\\",
    row$label_col, row$n_genes,
    round(row$Zdensity, 1), round(row$Zconnectivity, 1),
    round(row$Zsummary, 1),
    round(row$median_delta_k, 2), p_formatted,
    round(row$ctrl_lfp, 2)
  ))
}

latex_lines <- c(latex_lines, "\\bottomrule",
                 "\\end{tabular}", "\\end{table}")

# Per project convention (CLAUDE.md), LaTeX tables live in paths$tables
out_path <- file.path(paths$tables, out_filename)
writeLines(latex_lines, out_path)
cat("Saved:", out_path, "\n")
```

---

# Visualization: Delta kWithin Boxplot (Figure 6 Panel B)

```{r delta_k_boxplot, fig.width=12, fig.height=8}
# Order modules by median delta kWithin
module_order <- module_delta_k %>%
  arrange(median_delta_k) %>%
  pull(color)

# Join median_delta_k for coloring
delta_k_plot <- delta_k %>%
  left_join(
    module_delta_k %>% dplyr::select(color, median_delta_k),
    by = "color"
  ) %>%
  mutate(color = factor(color, levels = module_order))

# Prepare significance annotation from t-test results
sig_annotation <- module_delta_k %>%
  mutate(
    color = factor(color, levels = module_order),
    sig_label = case_when(
      p_adj_fdr < 0.001 ~ "***",
      p_adj_fdr < 0.01 ~ "**",
      p_adj_fdr < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  dplyr::select(color, sig_label)

# Prepare GO term annotation data (if available)
if (!is.null(top_bp_terms) && nrow(top_bp_terms) > 0) {
  go_annotation <- top_bp_terms %>%
    filter(module_color %in% module_order) %>%
    mutate(
      color = factor(module_color, levels = module_order),
      short_desc = str_trunc(Description, 30)
    )
  has_go_annotation <- TRUE
} else {
  has_go_annotation <- FALSE
}

# Get y position for significance markers (above each boxplot)
sig_y_pos <- delta_k_plot %>%
  group_by(color) %>%
  summarize(y_max = max(delta_kWithin, na.rm = TRUE), .groups = "drop") %>%
  left_join(sig_annotation, by = "color") %>%
  mutate(y_sig = y_max + (max(y_max) - min(y_max)) * 0.05)

# Create boxplot with module colors
p_delta_k <- ggplot(delta_k_plot, aes(x = color, y = delta_kWithin, fill = color)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_text(
    data = sig_y_pos,
    aes(x = color, y = y_sig, label = sig_label),
    inherit.aes = FALSE,
    size = 6,
    vjust = 0
  ) +
  scale_fill_identity() +
  theme_minimal(base_size = 25) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
    plot.title = element_blank(),
    plot.subtitle = element_blank()
  ) +
  labs(
    x = "Module",
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)")
  )

# Add GO term annotation if available
if (has_go_annotation) {
  go_annotation <- go_annotation %>%
    mutate(y_pos = 2.5)

  p_delta_k_annotated <- p_delta_k +
    geom_text(
      data = go_annotation,
      aes(x = color, y = y_pos, label = short_desc),
      vjust = 0,
      hjust = 0,
      size = 5,
      angle = 45,
      inherit.aes = FALSE
    ) +
    coord_cartesian(clip = "off", ylim = c(NA, 4)) +
    theme(plot.margin = margin(t = 60, r = 10, b = 10, l = 10))

  print(p_delta_k_annotated)

  ggsave(
    file.path(OUTPUT_DIR, "delta_kWithin_boxplot_annotated.pdf"),
    p_delta_k_annotated,
    width = 12, height = 10
  )
  ggsave(
    file.path(OUTPUT_DIR, "delta_kWithin_boxplot_annotated.png"),
    p_delta_k_annotated,
    width = 12, height = 10, dpi = 300
  )
  saveRDS(p_delta_k_annotated, file.path(OUTPUT_DIR, "delta_kWithin_boxplot_annotated.rds"))
  cat("Saved: delta_kWithin_boxplot_annotated.pdf/png/rds\n")
} else {
  print(p_delta_k)
}

# Also save simple version
ggsave(
  file.path(OUTPUT_DIR, "delta_kWithin_boxplot.pdf"),
  p_delta_k,
  width = 12, height = 6
)
ggsave(
  file.path(OUTPUT_DIR, "delta_kWithin_boxplot.png"),
  p_delta_k,
  width = 12, height = 6, dpi = 300
)
saveRDS(p_delta_k, file.path(OUTPUT_DIR, "delta_kWithin_boxplot.rds"))
cat("Saved: delta_kWithin_boxplot.pdf/png/rds\n")
```

---

# Visualization: Zsummary (Supplementary)

```{r zsummary_plot, fig.width=10, fig.height=6}
# Zsummary vs module size with standard thresholds
p_zsummary <- ggplot(
  zsummary_table,
  aes(x = module_size, y = Zsummary, color = preservation_status)
) +
  geom_point(size = 4, alpha = 0.8) +
  geom_hline(yintercept = c(2, 10), linetype = "dashed", color = "gray50") +
  geom_text_repel(
    aes(label = module_label),
    size = 3,
    max.overlaps = 15
  ) +
  scale_color_manual(
    values = c(
      "Not preserved" = "#D55E00",
      "Moderate" = "#E69F00",
      "Preserved" = "#009E73"
    )
  ) +
  annotate("text", x = max(zsummary_table$module_size) * 0.85, y = 1,
           label = "Not preserved (Z < 2)", color = "#D55E00", size = 3.5) +
  annotate("text", x = max(zsummary_table$module_size) * 0.85, y = 6,
           label = "Moderate (2-10)", color = "#E69F00", size = 3.5) +
  annotate("text", x = max(zsummary_table$module_size) * 0.85, y = 12,
           label = "Preserved (Z > 10)", color = "#009E73", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Module Preservation: CTRL to Inv4m",
    subtitle = sprintf(
      "%d modules: %d not preserved, %d moderate, %d preserved",
      nrow(zsummary_table),
      sum(zsummary_table$preservation_status == "Not preserved"),
      sum(zsummary_table$preservation_status == "Moderate"),
      sum(zsummary_table$preservation_status == "Preserved")
    ),
    x = "Module Size (genes)",
    y = expression(Z[summary]),
    color = "Preservation"
  )

print(p_zsummary)

ggsave(
  file.path(OUTPUT_DIR, "zsummary_preservation_plot.pdf"),
  p_zsummary,
  width = 10, height = 6
)
ggsave(
  file.path(OUTPUT_DIR, "zsummary_preservation_plot.png"),
  p_zsummary,
  width = 10, height = 6, dpi = 300
)
ggsave(
  file.path(paths$figures, "zsummary_preservation_plot.png"),
  p_zsummary,
  width = 10, height = 6, dpi = 300
)
cat("Saved: zsummary_preservation_plot.pdf/png\n")
```

---

# Visualization: Zsummary vs Delta kWithin

```{r zsummary_vs_delta, fig.width=10, fig.height=6}
# Merge Zsummary into module_delta_k for comparison plot
compare_data <- module_delta_k %>%
  left_join(
    zsummary_for_join %>% dplyr::select(module, Zsummary),
    by = "module"
  ) %>%
  filter(!is.na(Zsummary))

if (nrow(compare_data) > 0) {
  p_compare <- ggplot(
    compare_data,
    aes(x = median_delta_k, y = Zsummary, fill = color)
  ) +
    geom_point(aes(size = n_genes), alpha = 0.8, shape = 21, color = "black") +
    geom_hline(yintercept = 2, linetype = "dashed", color = "gray50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_text_repel(
      aes(label = color),
      size = 3,
      max.overlaps = 15
    ) +
    scale_fill_identity() +
    scale_size_continuous(range = c(3, 10)) +
    theme_minimal(base_size = 12) +
    labs(
      title = expression(Module ~ Perturbation: ~ Z[summary] ~ vs ~ Delta ~ k[Within]),
      subtitle = "Bottom-left quadrant = most disrupted modules",
      x = expression(Median ~ Delta ~ k[Within]),
      y = expression(Z[summary]),
      size = "Module Size"
    )

  print(p_compare)

  ggsave(
    file.path(OUTPUT_DIR, "zsummary_vs_delta_k.pdf"),
    p_compare,
    width = 10, height = 6
  )
}
```

---

# Save Summary Statistics

```{r save_summary}
summary_stats <- tibble(
  metric = c(
    "Total CTRL modules analyzed",
    "Soft power (shared)",
    "Modules not preserved (Zsummary < 2)",
    "Modules moderate (2 <= Zsummary < 10)",
    "Modules preserved (Zsummary >= 10)",
    "Modules with significant delta_k (FDR < 0.05)",
    "Permutations for modulePreservation",
    "Genes with decreased kWithin",
    "Genes with increased kWithin",
    "Mean delta kWithin (gene-level)"
  ),
  value = c(
    nrow(preservation_summary),
    SOFT_POWER,
    sum(zsummary_table$preservation_status == "Not preserved"),
    sum(zsummary_table$preservation_status == "Moderate"),
    sum(zsummary_table$preservation_status == "Preserved"),
    sum(module_delta_k$p_adj_fdr < 0.05, na.rm = TRUE),
    params$n_permutations,
    sum(delta_k$delta_kWithin < 0, na.rm = TRUE),
    sum(delta_k$delta_kWithin > 0, na.rm = TRUE),
    round(mean(delta_k$delta_kWithin, na.rm = TRUE), 4)
  )
)

write_csv(summary_stats, file.path(OUTPUT_DIR, "preservation_analysis_summary.csv"))

{
  cat("\n=== SAVED OUTPUTS ===\n")
  cat("modulePreservation_results.rds: full modulePreservation output\n")
  cat("zsummary_table.csv: Zsummary statistics per module\n")
  cat("preservation_summary.csv/.rds: integrated module metrics\n")
  cat("disrupted_modules.csv: modules with FDR < 0.05 and negative delta_k\n")
  cat("delta_kWithin_genes.csv: gene-level connectivity change\n")
  cat("delta_kWithin_modules.csv: module-level summary with t-test\n")
  cat("module_mapping.csv: CTRL -> Inv4m module correspondence\n")
  cat("preservation_table.tex: LaTeX table for manuscript\n")
  cat("delta_kWithin_boxplot_annotated.rds: Panel B for Figure 6\n")
  cat("zsummary_preservation_plot.png: Supplementary figure\n")
  cat("zsummary_vs_delta_k.pdf: Diagnostic plot\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Module coverage
  cat("Check 1: Module coverage\n")
  n_ctrl_mods <- length(unique(ctrl_modules$module[ctrl_modules$module != 0]))
  n_analyzed <- nrow(preservation_summary)
  coverage_pct <- round(100 * n_analyzed / n_ctrl_mods, 1)
  cat("  CTRL modules analyzed:", n_analyzed, "/", n_ctrl_mods,
      "(", coverage_pct, "%)\n")

  if (coverage_pct < 80) {
    stop("Module coverage below 80% - check gene overlap between conditions")
  }

  # Check 2: Delta kWithin calculated for assigned genes
  cat("\nCheck 2: Delta kWithin coverage\n")
  n_assigned_genes <- sum(ctrl_modules$module != 0)
  cat("  Genes with delta_k:", nrow(delta_k), "/", n_assigned_genes, "\n")

  # Check 3: modulePreservation completed
  cat("\nCheck 3: modulePreservation results\n")
  cat("  Zsummary range: [",
      round(min(zsummary_table$Zsummary), 2), ",",
      round(max(zsummary_table$Zsummary), 2), "]\n")

  # Check 4: t-test results available
  cat("\nCheck 4: t-test results\n")
  cat("  Modules with t-test:", sum(!is.na(module_delta_k$p_adj_fdr)), "/",
      nrow(module_delta_k), "\n")
  cat("  Significant (FDR < 0.05):", sum(module_delta_k$p_adj_fdr < 0.05, na.rm = TRUE), "\n")

  # Check 5: Module mapping
  cat("\nCheck 5: Module mapping\n")
  cat("  Modules mapped:", nrow(best_matches), "/", n_ctrl_mods, "\n")

  # Check 6: Output files exist
  cat("\nCheck 6: Output files\n")
  expected_files <- c(
    "modulePreservation_results.rds", "zsummary_table.csv",
    "preservation_summary.csv", "preservation_summary.rds",
    "disrupted_modules.csv", "delta_kWithin_genes.csv",
    "delta_kWithin_modules.csv", "module_mapping.csv",
    "preservation_table.tex"
  )
  for (f in expected_files) {
    exists <- file.exists(file.path(OUTPUT_DIR, f))
    cat("  ", f, ":", ifelse(exists, "OK", "MISSING"), "\n")
  }

  cat("\n=== VALIDATION COMPLETE ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
