---
title: "Field-Only Perturbation Analysis: Functional Characterization"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_permutations: 200
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/07_field_perturbation/07_functional.Rmd', output_dir='output/perturbation_YYYYMMDD/07_functional', params=list(run_dir='output/perturbation_YYYYMMDD'))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Functionally characterize ALL CTRL consensus modules using GO
enrichment (BP only), hub gene analysis, and integration with support metrics.

**Approach**:
1. Run GO BP enrichment for ALL CTRL modules using full filtered expression as universe
2. Identify hub genes (top 5 by kWithin) for ALL modules
3. Check which hubs lost connectivity in Inv4m
4. Integrate support metrics (LFP) with preservation status
5. Summarize biological themes and disruption classification

**Expected outcome**: Functional annotation of all modules with
consensus support metrics, preservation status, and hub gene connectivity.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(clusterProfiler)
  library(here)
  library(patchwork)
})

set.seed(123)
```

```{r setup_directories}
# Input/output directories
PRESERVATION_DIR <- here::here(params$run_dir, "06_preservation")
CONSENSUS_DIR <- here::here(params$run_dir, "04_consensus_networks")
BOOTSTRAP_DIR <- here::here(params$run_dir, "05_bootstrap_support")
FILTER_DIR <- here::here(params$run_dir, "02_gene_filter")
OUTPUT_DIR <- here::here(params$run_dir, "07_functional")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n\n")
}
```

---

# Load Data

```{r load_data}
# Load preservation summary (with all metrics)
preservation_summary <- readRDS(file.path(PRESERVATION_DIR, "preservation_summary.rds"))

# Load disrupted modules
disrupted_modules <- read_csv(
  file.path(PRESERVATION_DIR, "disrupted_modules.csv"),
  show_col_types = FALSE
)

# Load CTRL consensus modules
ctrl_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)

# Load delta kWithin (gene-level)
delta_k <- read_csv(
  file.path(PRESERVATION_DIR, "delta_kWithin_genes.csv"),
  show_col_types = FALSE
)

# Load bootstrap support
ctrl_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "ctrl", "module_support.csv"),
  show_col_types = FALSE
)

# Load gene annotations from filter step
gene_info <- read_csv(
  file.path(FILTER_DIR, "deg_neighborhood_genes.csv"),
  show_col_types = FALSE
)

# Load GO annotations
TERM2GENE <- readRDS(here("data", "TERM2GENE.rds"))
TERM2NAME <- readRDS(here("data", "TERM2NAME.rds"))

{
  cat("=== INPUT DATA ===\n")
  cat("Preservation summary:", nrow(preservation_summary), "modules\n")
  cat("Disrupted modules:", nrow(disrupted_modules), "\n")
  cat("Genes with module assignments:", nrow(ctrl_modules), "\n")
  cat("Genes with delta_k:", nrow(delta_k), "\n")
  cat("GO term-gene mappings:", nrow(TERM2GENE), "\n\n")

  if (nrow(disrupted_modules) > 0) {
    cat("Disrupted modules:\n")
    print(disrupted_modules %>%
            dplyr::select(module, color, n_genes, adjacency_correlation,
                          median_delta_k, ctrl_stability))
  } else {
    cat("No disrupted modules identified.\n")
    cat("Will analyze modules with weakest preservation.\n")
  }
}
```

---

# Prepare Module Gene Lists for GO Analysis

```{r prepare_gene_lists}
# Load combined expression (merged experiments, low-expression filtered)
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
expr_combined <- readRDS(file.path(DATA_PREP_DIR, "combined_expression.rds"))

# Universe = ALL genes after merging + low-expression filtering (~18K genes)
universe <- rownames(expr_combined)

cat("=== GENE UNIVERSE ===\n")
cat("Universe genes (merged + filtered):", length(universe), "\n")

# Get ALL CTRL modules (excluding grey/0)
all_modules <- ctrl_modules %>%
  filter(module != 0) %>%
  distinct(module, color) %>%
  arrange(module)

cat("CTRL modules to analyze:", nrow(all_modules), "\n\n")

# Create named list of gene vectors for each module (by color)
module_gene_lists <- split(
  ctrl_modules %>% filter(module != 0) %>% pull(gene),
  ctrl_modules %>% filter(module != 0) %>% pull(color)
)

# Report module sizes
cat("Module sizes:\n")
print(sapply(module_gene_lists, length))
```

---

# GO Enrichment Analysis (Biological Process)

**Approach**: Run GO BP enrichment for ALL CTRL consensus modules using
`compareCluster` for comparative analysis.

```{r enrichment_function}
#' Perform GO Enrichment Analysis (BP only)
#'
#' @param x Character vector of gene IDs
#' @param ontology GO ontology (default: "BP")
#' @return enrichResult object
ego <- function(x, ontology = "BP") {
  enricher(
    gene = x,
    pvalueCutoff = 0.05,
    pAdjustMethod = "none",
    universe = universe,
    minGSSize = 10,
    maxGSSize = 500,
    qvalueCutoff = 0.2,
    TERM2GENE = TERM2GENE[TERM2GENE$GO %in% TERM2NAME[[ontology]]$go_id, ],
    TERM2NAME = TERM2NAME[[ontology]]
  )
}
```

```{r run_enrichment}
# Run comparative GO enrichment for all modules
cat("Running GO BP enrichment for all modules...\n")

compGO_BP <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego
)

# Set factor levels to preserve module order
cluster_order <- names(module_gene_lists)
levels(compGO_BP@compareClusterResult$Cluster) <- cluster_order

{
  cat("\n=== GO ENRICHMENT RESULTS ===\n")
  cat("Enriched terms per module:\n")
  print(table(compGO_BP@compareClusterResult$Cluster))
}

# Save full results
go_results <- as.data.frame(compGO_BP@compareClusterResult) %>%
  rename(module_color = Cluster) %>%
  arrange(module_color, pvalue)

write_csv(
  go_results,
  file.path(OUTPUT_DIR, "all_modules_GO_BP.csv")
)

# Save the compareCluster object for later use
saveRDS(compGO_BP, file.path(OUTPUT_DIR, "compareCluster_GO_BP.rds"))

cat("\nTotal enriched terms:", nrow(go_results), "\n")
```

---

# Top GO Terms Summary

```{r go_summary}
if (nrow(go_results) > 0) {
  # Top 5 terms per module
  go_summary <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3))

  write_csv(
    go_summary,
    file.path(OUTPUT_DIR, "go_summary_table.csv")
  )

  {
    cat("\n=== TOP GO TERMS PER MODULE ===\n")
    print(go_summary, n = 50)
  }
} else {
  go_summary <- tibble()
  cat("\nNo significant GO enrichment found.\n")
}
```

---

# Hub Gene Analysis

**Approach**: Identify hub genes using CTRL consensus kWithin, then check
connectivity change in Inv4m. Analyze ALL modules (excluding grey/0).

```{r hub_analysis}
# Merge module assignments with delta kWithin (all modules except grey)
hub_data <- ctrl_modules %>%
  filter(module != 0) %>%
  inner_join(
    delta_k %>% dplyr::select(gene, kWithin_ctrl, kWithin_inv4m, delta_kWithin),
    by = "gene"
  )

# Top 5 hub genes per module by CTRL kWithin
hub_genes <- hub_data %>%
  group_by(module, color) %>%
  slice_max(kWithin_ctrl, n = 5) %>%
  ungroup()

# Add gene annotations
hub_genes <- hub_genes %>%
  left_join(
    gene_info %>% dplyr::select(gene, logFC, regulation, in_cis, in_Inv4m),
    by = "gene"
  ) %>%
  mutate(
    hub_disrupted = delta_kWithin < -0.05,
    connectivity_change = case_when(
      delta_kWithin < -0.1 ~ "Strong loss",
      delta_kWithin < -0.02 ~ "Moderate loss",
      delta_kWithin > 0.02 ~ "Gain",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(module, desc(kWithin_ctrl))

# Save hub genes
write_csv(
  hub_genes,
  file.path(OUTPUT_DIR, "hub_genes_all_modules.csv")
)

{
  cat("\n=== HUB GENES (ALL MODULES) ===\n")
  cat("Total hub genes:", nrow(hub_genes), "\n")
  cat("Hubs with connectivity loss:", sum(hub_genes$hub_disrupted, na.rm = TRUE), "\n\n")

  print(
    hub_genes %>%
      dplyr::select(module, color, gene, kWithin_ctrl, delta_kWithin,
                    connectivity_change, regulation)
  )
}
```

---

# Hub Gene Connectivity Plot

```{r hub_connectivity, fig.width=10, fig.height=6}
if (nrow(hub_genes) > 0) {
  p_hub <- ggplot(
    hub_genes,
    aes(x = kWithin_ctrl, y = kWithin_inv4m, color = color)
  ) +
    geom_point(aes(size = abs(logFC)), alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_text(
      aes(label = gene),
      size = 2.5,
      hjust = -0.1,
      vjust = 0.5,
      check_overlap = TRUE
    ) +
    scale_color_identity() +
    theme_minimal(base_size = 12) +
    labs(
      title = "Hub Gene Connectivity Change (Consensus Adjacency)",
      subtitle = "Points below diagonal = connectivity loss in Inv4m",
      x = "kWithin (CTRL)",
      y = "kWithin (Inv4m)",
      size = "|logFC|"
    )

  print(p_hub)

  ggsave(
    file.path(OUTPUT_DIR, "hub_connectivity_plot.pdf"),
    p_hub,
    width = 10, height = 6
  )
}
```

---

# Module Theme Summary with Support Metrics

```{r biological_themes}
# Create comprehensive summary for ALL modules using preservation_summary
module_themes <- preservation_summary %>%
  dplyr::select(
    module, color, n_genes, adjacency_correlation, median_delta_k,
    pct_decreased, ctrl_lfp, ctrl_stability, jaccard, preservation_status
  )

# Add top GO term (BP only - from new structure with module_color)
if (nrow(go_results) > 0) {
  top_bp <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_BP = Description, bp_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_bp, by = "color")
}

# Add top hub gene
if (nrow(hub_genes) > 0) {
  top_hubs <- hub_genes %>%
    group_by(module) %>%
    slice_max(kWithin_ctrl, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module, top_hub = gene, hub_delta_k = delta_kWithin,
                  hub_regulation = regulation)

  module_themes <- module_themes %>%
    left_join(top_hubs, by = "module")
}

# Save theme summary
write_csv(
  module_themes,
  file.path(OUTPUT_DIR, "module_theme_summary.csv")
)

{
  cat("\n=== MODULE BIOLOGICAL THEMES WITH SUPPORT ===\n")
  print(module_themes %>%
          dplyr::select(module, color, n_genes, ctrl_lfp, ctrl_stability,
                        adjacency_correlation, preservation_status, top_BP))
}
```

---

# GO Enrichment Visualization

```{r go_visualization, fig.width=12, fig.height=10}
if (nrow(go_results) > 0) {
  n_modules <- length(unique(go_results$module_color))

  # Top 5 terms per module for plotting
  go_plot_data <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      Description = str_trunc(Description, 40),
      GeneRatio_num = sapply(strsplit(GeneRatio, "/"), function(x) {
        as.numeric(x[1]) / as.numeric(x[2])
      })
    )

  if (nrow(go_plot_data) > 0 && n_modules <= 12) {
    p_go <- ggplot(
      go_plot_data,
      aes(x = GeneRatio_num, y = reorder(Description, GeneRatio_num),
          color = -log10(pvalue), size = Count)
    ) +
      geom_point() +
      facet_wrap(~module_color, scales = "free_y", ncol = 3) +
      scale_color_viridis_c(option = "plasma", name = "-log10(p)") +
      theme_minimal(base_size = 10) +
      theme(
        strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 7)
      ) +
      labs(
        title = "GO Biological Process Enrichment (All Modules)",
        x = "Gene Ratio",
        y = NULL,
        size = "Gene Count"
      )

    print(p_go)

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_plot.pdf"),
      p_go,
      width = 12, height = max(8, n_modules * 0.8)
    )
  } else if (n_modules > 12) {
    cat("Many modules (", n_modules, "). Full plot saved to PDF.\n")
    # Save a larger plot for many modules
    p_go <- ggplot(
      go_plot_data,
      aes(x = GeneRatio_num, y = reorder(Description, GeneRatio_num),
          color = -log10(pvalue), size = Count)
    ) +
      geom_point() +
      facet_wrap(~module_color, scales = "free_y", ncol = 4) +
      scale_color_viridis_c(option = "plasma", name = "-log10(p)") +
      theme_minimal(base_size = 8) +
      theme(
        strip.text = element_text(face = "bold", size = 7),
        axis.text.y = element_text(size = 6)
      ) +
      labs(
        title = "GO Biological Process Enrichment (All Modules)",
        x = "Gene Ratio",
        y = NULL,
        size = "Count"
      )

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_plot.pdf"),
      p_go,
      width = 16, height = max(12, n_modules * 0.6)
    )
  }
}
```

---

# Integration with Inv4m Direction

```{r inv4m_direction}
# Check direction bias in ALL modules (excluding grey)
direction_summary <- ctrl_modules %>%
  filter(module != 0) %>%
  left_join(
    gene_info %>% dplyr::select(gene, regulation, in_cis, in_Inv4m),
    by = "gene"
  ) %>%
  group_by(module, color) %>%
  summarise(
    n_genes = n(),
    n_upregulated = sum(regulation == "Upregulated", na.rm = TRUE),
    n_downregulated = sum(regulation == "Downregulated", na.rm = TRUE),
    pct_upregulated = round(n_upregulated / n_genes * 100, 1),
    n_cis = sum(in_cis, na.rm = TRUE),
    pct_cis = round(n_cis / n_genes * 100, 1),
    n_in_inv4m = sum(in_Inv4m, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(
  direction_summary,
  file.path(OUTPUT_DIR, "module_direction_summary.csv")
)

{
  cat("\n=== INV4M DIRECTION IN MODULES ===\n")
  print(direction_summary)
}
```

---

# Comprehensive Module Summary Table

```{r comprehensive_summary}
# Create final comprehensive summary
comprehensive_summary <- preservation_summary %>%
  left_join(
    ctrl_support %>% dplyr::select(module, mean_entropy, n_genes_support = n_genes),
    by = "module"
  ) %>%
  left_join(
    direction_summary %>% dplyr::select(module, pct_upregulated, pct_cis),
    by = "module"
  )

# Add GO term counts (using module_color from new structure)
if (nrow(go_results) > 0) {
  go_counts <- go_results %>%
    count(module_color, name = "n_go_terms") %>%
    rename(color = module_color)

  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_go_terms = 0)
}

# Classify final disruption status
comprehensive_summary <- comprehensive_summary %>%
  mutate(
    disrupted_final = case_when(
      adjacency_correlation < 0.3 ~ "Strong",
      adjacency_correlation < 0.5 ~ "Moderate",
      adjacency_correlation < 0.7 & median_delta_k < -0.05 ~ "Weak",
      TRUE ~ "Not disrupted"
    ),
    disrupted_final = factor(
      disrupted_final,
      levels = c("Strong", "Moderate", "Weak", "Not disrupted")
    )
  )

write_csv(
  comprehensive_summary,
  file.path(OUTPUT_DIR, "comprehensive_module_summary.csv")
)

{
  cat("\n=== COMPREHENSIVE MODULE SUMMARY ===\n")
  cat("\nDisruption classification:\n")
  print(table(comprehensive_summary$disrupted_final))

  cat("\nTop disrupted modules:\n")
  print(
    comprehensive_summary %>%
      filter(disrupted_final != "Not disrupted") %>%
      dplyr::select(module, color, n_genes, ctrl_lfp, ctrl_stability,
                    adjacency_correlation, disrupted_final, pct_upregulated) %>%
      arrange(adjacency_correlation)
  )
}
```

---

# Save Final Summary Statistics

```{r save_summary}
n_modules_analyzed <- nrow(all_modules)
n_modules_with_go <- length(unique(go_results$module_color))

summary_stats <- tibble(
  metric = c(
    "Total CTRL modules analyzed",
    "GO BP terms enriched (total)",
    "Modules with GO enrichment",
    "Hub genes identified",
    "Hub genes with connectivity loss",
    "Strong disruption modules",
    "Moderate disruption modules",
    "Weak disruption modules"
  ),
  value = c(
    n_modules_analyzed,
    nrow(go_results),
    n_modules_with_go,
    nrow(hub_genes),
    sum(hub_genes$hub_disrupted, na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Strong", na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Moderate", na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Weak", na.rm = TRUE)
  )
)

write_csv(
  summary_stats,
  file.path(OUTPUT_DIR, "functional_analysis_summary.csv")
)

{
  cat("\n=== SAVED OUTPUTS ===\n")
  cat("all_modules_GO_BP.csv: GO BP enrichment for all modules\n")
  cat("compareCluster_GO_BP.rds: compareCluster object for visualization\n")
  cat("go_summary_table.csv: top GO terms per module\n")
  cat("hub_genes_all_modules.csv: hub genes with connectivity change\n")
  cat("module_theme_summary.csv: biological themes with support metrics\n")
  cat("module_direction_summary.csv: Inv4m direction summary\n")
  cat("comprehensive_module_summary.csv: all modules with all metrics\n")
  cat("go_enrichment_plot.pdf: GO enrichment visualization\n")
  cat("hub_connectivity_plot.pdf: hub gene connectivity\n")
  cat("functional_analysis_summary.csv: summary statistics\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Modules analyzed
  cat("Check 1: Modules analyzed\n")
  cat("  Modules:", nrow(all_modules), "\n")
  stopifnot(nrow(all_modules) > 0)

  # Check 2: GO enrichment ran
  cat("\nCheck 2: GO enrichment executed\n")
  cat("  Terms found:", nrow(go_results), "\n")

  # Check 3: Hub genes identified
  cat("\nCheck 3: Hub genes identified\n")
  cat("  Hub genes:", nrow(hub_genes), "\n")
  stopifnot(nrow(hub_genes) > 0)

  # Check 4: Comprehensive summary complete
  cat("\nCheck 4: Comprehensive summary\n")
  cat("  Modules in summary:", nrow(comprehensive_summary), "\n")
  stopifnot(nrow(comprehensive_summary) == nrow(preservation_summary))

  # Check 5: All outputs saved
  cat("\nCheck 5: Output files created\n")
  expected_files <- c(
    "all_modules_GO_BP.csv",
    "hub_genes_all_modules.csv",
    "functional_analysis_summary.csv",
    "comprehensive_module_summary.csv"
  )
  for (f in expected_files) {
    exists <- file.exists(file.path(OUTPUT_DIR, f))
    cat("  ", f, ":", ifelse(exists, "OK", "MISSING"), "\n")
  }

  cat("\n=== ALL CHECKS PASSED ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
