---
title: "Field-Only Perturbation Analysis: Functional Characterization"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_permutations: 200
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/inversion_paper/field_perturbation/07_module_annotation.Rmd', output_dir='output/perturbation_YYYYMMDD/07_module_annotation', params=list(run_dir='output/perturbation_YYYYMMDD'))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Functionally characterize ALL CTRL consensus modules using GO
enrichment (BP, CC, MF), hub gene analysis, and integration with support metrics.

**Approach**:
1. Run GO enrichment (BP, CC, MF) for ALL CTRL modules using full filtered expression as universe
2. Identify hub genes (top 5 by kWithin) for ALL modules
3. Check which hubs lost connectivity in Inv4m
4. Integrate support metrics (LFP) with preservation status
5. Summarize biological themes and disruption classification

**Expected outcome**: Functional annotation of all modules with
consensus support metrics, preservation status, and hub gene connectivity.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(clusterProfiler)
  library(here)
  library(patchwork)
})

set.seed(123)
```

```{r setup_directories}
# Input/output directories
PRESERVATION_DIR <- here::here(params$run_dir, "06_preservation")
CONSENSUS_DIR <- here::here(params$run_dir, "04_consensus_networks")
BOOTSTRAP_DIR <- here::here(params$run_dir, "05_bootstrap_support")
FILTER_DIR <- here::here(params$run_dir, "02_gene_filter")
OUTPUT_DIR <- here::here(params$run_dir, "07_module_annotation")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n\n")
}
```

---

# Load Data

```{r load_data}
# Load preservation summary (with all metrics)
preservation_summary <- readRDS(file.path(PRESERVATION_DIR, "preservation_summary.rds"))

# Load disrupted modules
disrupted_modules <- read_csv(
  file.path(PRESERVATION_DIR, "disrupted_modules.csv"),
  show_col_types = FALSE
)

# Load CTRL consensus modules
ctrl_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)

# Load delta kWithin (gene-level)
delta_k <- read_csv(
  file.path(PRESERVATION_DIR, "delta_kWithin_genes.csv"),
  show_col_types = FALSE
)

# Load bootstrap support
ctrl_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "ctrl", "module_support.csv"),
  show_col_types = FALSE
)

# Load gene annotations from filter step
gene_info <- read_csv(
  file.path(FILTER_DIR, "deg_neighborhood_genes.csv"),
  show_col_types = FALSE
)

# Load GO annotations
TERM2GENE <- readRDS(here("data", "TERM2GENE.rds"))
TERM2NAME <- readRDS(here("data", "TERM2NAME.rds"))

{
  cat("=== INPUT DATA ===\n")
  cat("Preservation summary:", nrow(preservation_summary), "modules\n")
  cat("Disrupted modules:", nrow(disrupted_modules), "\n")
  cat("Genes with module assignments:", nrow(ctrl_modules), "\n")
  cat("Genes with delta_k:", nrow(delta_k), "\n")
  cat("GO term-gene mappings:", nrow(TERM2GENE), "\n\n")

  if (nrow(disrupted_modules) > 0) {
    cat("Disrupted modules:\n")
    print(disrupted_modules %>%
            dplyr::select(module, color, n_genes, adjacency_correlation,
                          median_delta_k, ctrl_stability))
  } else {
    cat("No disrupted modules identified.\n")
    cat("Will analyze modules with weakest preservation.\n")
  }
}
```

---

# Prepare Module Gene Lists for GO Analysis

```{r prepare_gene_lists}
# Load combined expression (merged experiments, low-expression filtered)
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
expr_combined <- readRDS(file.path(DATA_PREP_DIR, "combined_expression.rds"))

# Universe = ALL genes after merging + low-expression filtering (~18K genes)
universe <- rownames(expr_combined)

cat("=== GENE UNIVERSE ===\n")
cat("Universe genes (merged + filtered):", length(universe), "\n")

# Get ALL CTRL modules (excluding grey/0)
all_modules <- ctrl_modules %>%
  filter(module != 0) %>%
  distinct(module, color) %>%
  arrange(module)

cat("CTRL modules to analyze:", nrow(all_modules), "\n\n")

# Create named list of gene vectors for each module (by color)
module_gene_lists <- split(
  ctrl_modules %>% filter(module != 0) %>% pull(gene),
  ctrl_modules %>% filter(module != 0) %>% pull(color)
)

# Report module sizes
cat("Module sizes:\n")
print(sapply(module_gene_lists, length))
```

---

# GO Enrichment Analysis (BP, CC, MF)

**Approach**: Run GO enrichment for all three ontologies (Biological Process,
Cellular Component, Molecular Function) for ALL CTRL consensus modules using
`compareCluster` for comparative analysis.

```{r enrichment_function}
#' Perform GO Enrichment Analysis
#'
#' @param x Character vector of gene IDs
#' @param ontology GO ontology (default: "BP")
#' @return enrichResult object
ego <- function(x, ontology = "BP") {
  enricher(
    gene = x,
    pvalueCutoff = 0.05,
    pAdjustMethod = "none",
    universe = universe,
    minGSSize = 10,
    maxGSSize = 500,
    qvalueCutoff = 0.2,
    TERM2GENE = TERM2GENE[TERM2GENE$GO %in% TERM2NAME[[ontology]]$go_id, ],
    TERM2NAME = TERM2NAME[[ontology]]
  )
}
```

```{r run_enrichment}
# Run comparative GO enrichment for all modules
cat("Running GO BP enrichment for all modules...\n")

compGO_BP <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego
)

# Set factor levels to preserve module order
cluster_order <- names(module_gene_lists)
levels(compGO_BP@compareClusterResult$Cluster) <- cluster_order

{
  cat("\n=== GO ENRICHMENT RESULTS ===\n")
  cat("Enriched terms per module:\n")
  print(table(compGO_BP@compareClusterResult$Cluster))
}

# Save full results
go_results <- as.data.frame(compGO_BP@compareClusterResult) %>%
  rename(module_color = Cluster) %>%
  arrange(module_color, pvalue)

write_csv(
  go_results,
  file.path(OUTPUT_DIR, "all_modules_GO_BP.csv")
)

# Save the compareCluster object for later use
saveRDS(compGO_BP, file.path(OUTPUT_DIR, "compareCluster_GO_BP.rds"))

cat("\nTotal enriched BP terms:", nrow(go_results), "\n")
```

## Cellular Component (CC) Enrichment

```{r run_enrichment_cc}
# Run comparative GO CC enrichment for all modules
cat("Running GO CC enrichment for all modules...\n")

# Create CC-specific enrichment function
ego_cc <- function(x) ego(x, ontology = "CC")

compGO_CC <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego_cc
)

filtered_result <- compGO_BP@compareClusterResult%>%
  filter(Count > 1) %>%
  group_by(Cluster, geneID) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  ungroup()

# Set factor levels to preserve module order
if (!is.null(compGO_CC) && nrow(compGO_CC@compareClusterResult) > 0) {
  levels(compGO_CC@compareClusterResult$Cluster) <- cluster_order
}

{
  cat("\n=== GO CC ENRICHMENT RESULTS ===\n")
  if (!is.null(compGO_CC) && nrow(compGO_CC@compareClusterResult) > 0) {
    cat("Enriched terms per module:\n")
    print(table(compGO_CC@compareClusterResult$Cluster))
  } else {
    cat("No significant CC enrichment found.\n")
  }
}

# Save full results
if (!is.null(compGO_CC) && nrow(compGO_CC@compareClusterResult) > 0) {
  go_results_cc <- as.data.frame(compGO_CC@compareClusterResult) %>%
    rename(module_color = Cluster) %>%
    arrange(module_color, pvalue)

  write_csv(
    go_results_cc,
    file.path(OUTPUT_DIR, "all_modules_GO_CC.csv")
  )

  saveRDS(compGO_CC, file.path(OUTPUT_DIR, "compareCluster_GO_CC.rds"))
  cat("\nTotal enriched CC terms:", nrow(go_results_cc), "\n")
} else {
  go_results_cc <- tibble()
  cat("\nNo significant CC enrichment found.\n")
}
```

## Molecular Function (MF) Enrichment

```{r run_enrichment_mf}
# Run comparative GO MF enrichment for all modules
cat("Running GO MF enrichment for all modules...\n")

# Create MF-specific enrichment function
ego_mf <- function(x) ego(x, ontology = "MF")

compGO_MF <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego_mf
)

# Set factor levels to preserve module order
if (!is.null(compGO_MF) && nrow(compGO_MF@compareClusterResult) > 0) {
  levels(compGO_MF@compareClusterResult$Cluster) <- cluster_order
}

{
  cat("\n=== GO MF ENRICHMENT RESULTS ===\n")
  if (!is.null(compGO_MF) && nrow(compGO_MF@compareClusterResult) > 0) {
    cat("Enriched terms per module:\n")
    print(table(compGO_MF@compareClusterResult$Cluster))
  } else {
    cat("No significant MF enrichment found.\n")
  }
}

# Save full results
if (!is.null(compGO_MF) && nrow(compGO_MF@compareClusterResult) > 0) {
  go_results_mf <- as.data.frame(compGO_MF@compareClusterResult) %>%
    rename(module_color = Cluster) %>%
    arrange(module_color, pvalue)

  write_csv(
    go_results_mf,
    file.path(OUTPUT_DIR, "all_modules_GO_MF.csv")
  )

  saveRDS(compGO_MF, file.path(OUTPUT_DIR, "compareCluster_GO_MF.rds"))
  cat("\nTotal enriched MF terms:", nrow(go_results_mf), "\n")
} else {
  go_results_mf <- tibble()
  cat("\nNo significant MF enrichment found.\n")
}
```

---

# Top GO Terms Summary

```{r go_summary}
# BP Summary
if (nrow(go_results) > 0) {
  go_summary_bp <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3), ontology = "BP")

  write_csv(
    go_summary_bp,
    file.path(OUTPUT_DIR, "go_summary_BP.csv")
  )

  cat("\n=== TOP GO BP TERMS PER MODULE ===\n")
  print(go_summary_bp, n = 30)
} else {
  go_summary_bp <- tibble()
  cat("\nNo significant BP enrichment found.\n")
}

# CC Summary
if (nrow(go_results_cc) > 0) {
  go_summary_cc <- go_results_cc %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3), ontology = "CC")

  write_csv(
    go_summary_cc,
    file.path(OUTPUT_DIR, "go_summary_CC.csv")
  )

  cat("\n=== TOP GO CC TERMS PER MODULE ===\n")
  print(go_summary_cc, n = 30)
} else {
  go_summary_cc <- tibble()
  cat("\nNo significant CC enrichment found.\n")
}

# MF Summary
if (nrow(go_results_mf) > 0) {
  go_summary_mf <- go_results_mf %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3), ontology = "MF")

  write_csv(
    go_summary_mf,
    file.path(OUTPUT_DIR, "go_summary_MF.csv")
  )

  cat("\n=== TOP GO MF TERMS PER MODULE ===\n")
  print(go_summary_mf, n = 30)
} else {
  go_summary_mf <- tibble()
  cat("\nNo significant MF enrichment found.\n")
}

# Combined summary table (top 3 per ontology per module)
go_summary_combined <- bind_rows(
  go_summary_bp %>% group_by(module_color) %>% slice_min(pvalue, n = 3, with_ties = FALSE),
  go_summary_cc %>% group_by(module_color) %>% slice_min(pvalue, n = 3, with_ties = FALSE),
  go_summary_mf %>% group_by(module_color) %>% slice_min(pvalue, n = 3, with_ties = FALSE)
) %>%
  arrange(module_color, ontology, pvalue)

write_csv(
  go_summary_combined,
  file.path(OUTPUT_DIR, "go_summary_all_ontologies.csv")
)

cat("\n=== COMBINED GO SUMMARY (all ontologies) ===\n")
cat("Total terms in combined summary:", nrow(go_summary_combined), "\n")
```

---

# Hub Gene Analysis

**Approach**: Identify hub genes using CTRL consensus kWithin, then check
connectivity change in Inv4m. Analyze ALL modules (excluding grey/0).

```{r hub_analysis}
# Merge module assignments with delta kWithin (all modules except grey)
hub_data <- ctrl_modules %>%
  filter(module != 0) %>%
  inner_join(
    delta_k %>% dplyr::select(gene, kWithin_ctrl, kWithin_inv4m, delta_kWithin),
    by = "gene"
  )

# Top 5 hub genes per module by CTRL kWithin
hub_genes <- hub_data %>%
  group_by(module, color) %>%
  slice_max(kWithin_ctrl, n = 5) %>%
  ungroup()

# Add gene annotations
hub_genes <- hub_genes %>%
  left_join(
    gene_info %>% dplyr::select(gene, logFC, regulation, in_cis, in_Inv4m),
    by = "gene"
  ) %>%
  mutate(
    hub_disrupted = delta_kWithin < -0.05,
    connectivity_change = case_when(
      delta_kWithin < -0.1 ~ "Strong loss",
      delta_kWithin < -0.02 ~ "Moderate loss",
      delta_kWithin > 0.02 ~ "Gain",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(module, desc(kWithin_ctrl))

# Save hub genes
write_csv(
  hub_genes,
  file.path(OUTPUT_DIR, "hub_genes_all_modules.csv")
)

{
  cat("\n=== HUB GENES (ALL MODULES) ===\n")
  cat("Total hub genes:", nrow(hub_genes), "\n")
  cat("Hubs with connectivity loss:", sum(hub_genes$hub_disrupted, na.rm = TRUE), "\n\n")

  print(
    hub_genes %>%
      dplyr::select(module, color, gene, kWithin_ctrl, delta_kWithin,
                    connectivity_change, regulation)
  )
}
```

---

# Hub Gene Connectivity Plot

```{r hub_connectivity, fig.width=10, fig.height=6}
if (nrow(hub_genes) > 0) {
  p_hub <- ggplot(
    hub_genes,
    aes(x = kWithin_ctrl, y = kWithin_inv4m, color = color)
  ) +
    geom_point(aes(size = abs(logFC)), alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_text(
      aes(label = gene),
      size = 2.5,
      hjust = -0.1,
      vjust = 0.5,
      check_overlap = TRUE
    ) +
    scale_color_identity() +
    theme_minimal(base_size = 12) +
    labs(
      title = "Hub Gene Connectivity Change (Consensus Adjacency)",
      subtitle = "Points below diagonal = connectivity loss in Inv4m",
      x = "kWithin (CTRL)",
      y = "kWithin (Inv4m)",
      size = "|logFC|"
    )

  print(p_hub)

  ggsave(
    file.path(OUTPUT_DIR, "hub_connectivity_plot.pdf"),
    p_hub,
    width = 10, height = 6
  )
}
```

---

# Module Theme Summary with Support Metrics

```{r biological_themes}
# Create comprehensive summary for ALL modules using preservation_summary
module_themes <- preservation_summary %>%
  dplyr::select(
    module, color, n_genes, adjacency_correlation, median_delta_k,
    pct_decreased, ctrl_lfp, ctrl_stability, jaccard, preservation_status
  )

# Add top GO terms (BP, CC, MF)
if (nrow(go_results) > 0) {
  top_bp <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_BP = Description, bp_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_bp, by = "color")
}

if (nrow(go_results_cc) > 0) {
  top_cc <- go_results_cc %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_CC = Description, cc_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_cc, by = "color")
}

if (nrow(go_results_mf) > 0) {
  top_mf <- go_results_mf %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_MF = Description, mf_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_mf, by = "color")
}

# Add top hub gene
if (nrow(hub_genes) > 0) {
  top_hubs <- hub_genes %>%
    group_by(module) %>%
    slice_max(kWithin_ctrl, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module, top_hub = gene, hub_delta_k = delta_kWithin,
                  hub_regulation = regulation)

  module_themes <- module_themes %>%
    left_join(top_hubs, by = "module")
}

# Save theme summary
write_csv(
  module_themes,
  file.path(OUTPUT_DIR, "module_theme_summary.csv")
)

{
  cat("\n=== MODULE BIOLOGICAL THEMES WITH SUPPORT ===\n")
  print(module_themes %>%
          dplyr::select(module, color, n_genes, ctrl_lfp, ctrl_stability,
                        adjacency_correlation, preservation_status, top_BP))
}
```

---

# GO Enrichment Visualization

```{r go_plot_function}
#' Create GO enrichment dot plot for a given ontology
#'
#' @param go_data Data frame with GO enrichment results
#' @param ontology_name Name of the ontology for plot title
#' @param filename Output filename
create_go_plot <- function(go_data, ontology_name, filename) {
  if (nrow(go_data) == 0) {
    cat("No significant", ontology_name, "enrichment to plot.\n")
    return(NULL)
  }

  n_modules <- length(unique(go_data$module_color))

  # Top 5 terms per module for plotting
  go_plot_data <- go_data %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      Description = str_trunc(Description, 40),
      GeneRatio_num = sapply(strsplit(GeneRatio, "/"), function(x) {
        as.numeric(x[1]) / as.numeric(x[2])
      })
    )

  if (nrow(go_plot_data) == 0) return(NULL)

  ncol_facet <- if (n_modules <= 12) 3 else 4
  base_size <- if (n_modules <= 12) 10 else 8
  text_size <- if (n_modules <= 12) 7 else 6

  p <- ggplot(
    go_plot_data,
    aes(x = GeneRatio_num, y = reorder(Description, GeneRatio_num),
        color = -log10(pvalue), size = Count)
  ) +
    geom_point() +
    facet_wrap(~module_color, scales = "free_y", ncol = ncol_facet) +
    scale_color_viridis_c(option = "plasma", name = "-log10(p)") +
    theme_minimal(base_size = base_size) +
    theme(
      strip.text = element_text(face = "bold", size = text_size),
      axis.text.y = element_text(size = text_size)
    ) +
    labs(
      title = paste("GO", ontology_name, "Enrichment (All Modules)"),
      x = "Gene Ratio",
      y = NULL,
      size = "Count"
    )

  width <- if (n_modules <= 12) 12 else 16
  height <- if (n_modules <= 12) max(8, n_modules * 0.8) else max(12, n_modules * 0.6)

  ggsave(
    file.path(OUTPUT_DIR, filename),
    p,
    width = width, height = height
  )

  return(p)
}
```

## BP Enrichment Plot

```{r go_visualization_bp, fig.width=12, fig.height=10}
p_bp <- create_go_plot(go_results, "Biological Process", "go_enrichment_BP.pdf")
if (!is.null(p_bp)) print(p_bp)
```

## CC Enrichment Plot

```{r go_visualization_cc, fig.width=12, fig.height=10}
p_cc <- create_go_plot(go_results_cc, "Cellular Component", "go_enrichment_CC.pdf")
if (!is.null(p_cc)) print(p_cc)
```

## MF Enrichment Plot

```{r go_visualization_mf, fig.width=12, fig.height=10}
p_mf <- create_go_plot(go_results_mf, "Molecular Function", "go_enrichment_MF.pdf")
if (!is.null(p_mf)) print(p_mf)
```

---

# Integration with Inv4m Direction

```{r inv4m_direction}
# Check direction bias in ALL modules (excluding grey)
direction_summary <- ctrl_modules %>%
  filter(module != 0) %>%
  left_join(
    gene_info %>% dplyr::select(gene, regulation, in_cis, in_Inv4m),
    by = "gene"
  ) %>%
  group_by(module, color) %>%
  summarise(
    n_genes = n(),
    n_upregulated = sum(regulation == "Upregulated", na.rm = TRUE),
    n_downregulated = sum(regulation == "Downregulated", na.rm = TRUE),
    pct_upregulated = round(n_upregulated / n_genes * 100, 1),
    n_cis = sum(in_cis, na.rm = TRUE),
    pct_cis = round(n_cis / n_genes * 100, 1),
    n_in_inv4m = sum(in_Inv4m, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(
  direction_summary,
  file.path(OUTPUT_DIR, "module_direction_summary.csv")
)

{
  cat("\n=== INV4M DIRECTION IN MODULES ===\n")
  print(direction_summary)
}
```

---

# Comprehensive Module Summary Table

```{r comprehensive_summary}
# Create final comprehensive summary
comprehensive_summary <- preservation_summary %>%
  left_join(
    ctrl_support %>% dplyr::select(module, mean_entropy, n_genes_support = n_genes),
    by = "module"
  ) %>%
  left_join(
    direction_summary %>% dplyr::select(module, pct_upregulated, pct_cis),
    by = "module"
  )

# Add GO term counts for all ontologies
if (nrow(go_results) > 0) {
  go_counts_bp <- go_results %>%
    count(module_color, name = "n_BP_terms") %>%
    rename(color = module_color)
  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts_bp, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_BP_terms = 0)
}

if (nrow(go_results_cc) > 0) {
  go_counts_cc <- go_results_cc %>%
    count(module_color, name = "n_CC_terms") %>%
    rename(color = module_color)
  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts_cc, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_CC_terms = 0)
}

if (nrow(go_results_mf) > 0) {
  go_counts_mf <- go_results_mf %>%
    count(module_color, name = "n_MF_terms") %>%
    rename(color = module_color)
  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts_mf, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_MF_terms = 0)
}

# Replace NA counts with 0
comprehensive_summary <- comprehensive_summary %>%
  mutate(
    n_BP_terms = replace_na(n_BP_terms, 0),
    n_CC_terms = replace_na(n_CC_terms, 0),
    n_MF_terms = replace_na(n_MF_terms, 0),
    n_GO_total = n_BP_terms + n_CC_terms + n_MF_terms
  )

# Classify final disruption status
comprehensive_summary <- comprehensive_summary %>%
  mutate(
    disrupted_final = case_when(
      adjacency_correlation < 0.3 ~ "Strong",
      adjacency_correlation < 0.5 ~ "Moderate",
      adjacency_correlation < 0.7 & median_delta_k < -0.05 ~ "Weak",
      TRUE ~ "Not disrupted"
    ),
    disrupted_final = factor(
      disrupted_final,
      levels = c("Strong", "Moderate", "Weak", "Not disrupted")
    )
  )

write_csv(
  comprehensive_summary,
  file.path(OUTPUT_DIR, "comprehensive_module_summary.csv")
)

{
  cat("\n=== COMPREHENSIVE MODULE SUMMARY ===\n")
  cat("\nDisruption classification:\n")
  print(table(comprehensive_summary$disrupted_final))

  cat("\nTop disrupted modules:\n")
  print(
    comprehensive_summary %>%
      filter(disrupted_final != "Not disrupted") %>%
      dplyr::select(module, color, n_genes, ctrl_lfp, ctrl_stability,
                    adjacency_correlation, disrupted_final, pct_upregulated) %>%
      arrange(adjacency_correlation)
  )
}
```

---

# Save Final Summary Statistics

```{r save_summary}
n_modules_analyzed <- nrow(all_modules)
n_modules_with_bp <- length(unique(go_results$module_color))
n_modules_with_cc <- length(unique(go_results_cc$module_color))
n_modules_with_mf <- length(unique(go_results_mf$module_color))

summary_stats <- tibble(
  metric = c(
    "Total CTRL modules analyzed",
    "GO BP terms enriched (total)",
    "GO CC terms enriched (total)",
    "GO MF terms enriched (total)",
    "Modules with BP enrichment",
    "Modules with CC enrichment",
    "Modules with MF enrichment",
    "Hub genes identified",
    "Hub genes with connectivity loss",
    "Strong disruption modules",
    "Moderate disruption modules",
    "Weak disruption modules"
  ),
  value = c(
    n_modules_analyzed,
    nrow(go_results),
    nrow(go_results_cc),
    nrow(go_results_mf),
    n_modules_with_bp,
    n_modules_with_cc,
    n_modules_with_mf,
    nrow(hub_genes),
    sum(hub_genes$hub_disrupted, na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Strong", na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Moderate", na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Weak", na.rm = TRUE)
  )
)

write_csv(
  summary_stats,
  file.path(OUTPUT_DIR, "functional_analysis_summary.csv")
)

{
  cat("\n=== SAVED OUTPUTS ===\n")
  cat("GO Enrichment Results:\n")
  cat("  all_modules_GO_BP.csv: GO BP enrichment for all modules\n")
  cat("  all_modules_GO_CC.csv: GO CC enrichment for all modules\n")
  cat("  all_modules_GO_MF.csv: GO MF enrichment for all modules\n")
  cat("  compareCluster_GO_BP.rds: compareCluster object (BP)\n")
  cat("  compareCluster_GO_CC.rds: compareCluster object (CC)\n")
  cat("  compareCluster_GO_MF.rds: compareCluster object (MF)\n")
  cat("  go_summary_BP.csv: top BP terms per module\n")
  cat("  go_summary_CC.csv: top CC terms per module\n")
  cat("  go_summary_MF.csv: top MF terms per module\n")
  cat("  go_summary_all_ontologies.csv: combined summary\n")
  cat("\nVisualization:\n")
  cat("  go_enrichment_BP.pdf: BP enrichment plot\n")
  cat("  go_enrichment_CC.pdf: CC enrichment plot\n")
  cat("  go_enrichment_MF.pdf: MF enrichment plot\n")
  cat("  hub_connectivity_plot.pdf: hub gene connectivity\n")
  cat("\nSummary Tables:\n")
  cat("  hub_genes_all_modules.csv: hub genes with connectivity change\n")
  cat("  module_theme_summary.csv: biological themes with support metrics\n")
  cat("  module_direction_summary.csv: Inv4m direction summary\n")
  cat("  comprehensive_module_summary.csv: all modules with all metrics\n")
  cat("  functional_analysis_summary.csv: summary statistics\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Modules analyzed
  cat("Check 1: Modules analyzed\n")
  cat("  Modules:", nrow(all_modules), "\n")
  stopifnot(nrow(all_modules) > 0)

  # Check 2: GO enrichment ran (all ontologies)
  cat("\nCheck 2: GO enrichment executed\n")
  cat("  BP terms found:", nrow(go_results), "\n")
  cat("  CC terms found:", nrow(go_results_cc), "\n")
  cat("  MF terms found:", nrow(go_results_mf), "\n")

  # Check 3: Hub genes identified
  cat("\nCheck 3: Hub genes identified\n")
  cat("  Hub genes:", nrow(hub_genes), "\n")
  stopifnot(nrow(hub_genes) > 0)

  # Check 4: Comprehensive summary complete
  cat("\nCheck 4: Comprehensive summary\n")
  cat("  Modules in summary:", nrow(comprehensive_summary), "\n")
  stopifnot(nrow(comprehensive_summary) == nrow(preservation_summary))

  # Check 5: All outputs saved
  cat("\nCheck 5: Output files created\n")
  expected_files <- c(
    "all_modules_GO_BP.csv",
    "all_modules_GO_CC.csv",
    "all_modules_GO_MF.csv",
    "go_summary_all_ontologies.csv",
    "hub_genes_all_modules.csv",
    "functional_analysis_summary.csv",
    "comprehensive_module_summary.csv"
  )
  for (f in expected_files) {
    exists <- file.exists(file.path(OUTPUT_DIR, f))
    cat("  ", f, ":", ifelse(exists, "OK", "MISSING"), "\n")
  }

  cat("\n=== ALL CHECKS PASSED ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
