---
title: "Field-Only Perturbation Analysis: Functional Characterization"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  run_dir: "output/perturbation_test"
  n_permutations: 200
---

<!--
To render this document with output in the proper directory, use:
Rscript -e "rmarkdown::render('scripts/inversion_paper/field_perturbation/07_module_annotation.Rmd', output_dir='output/perturbation_YYYYMMDD/07_module_annotation', params=list(run_dir='output/perturbation_YYYYMMDD'))"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = here::here())

# Load project paths
source(here::here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Functionally characterize ALL CTRL consensus modules using GO
enrichment (BP, CC, MF), hub gene analysis, and integration with support metrics.

**Approach**:
1. Run GO enrichment (BP, CC, MF) for ALL CTRL modules using full filtered expression as universe
2. Identify hub genes (top 5 by kWithin) for ALL modules
3. Check which hubs lost connectivity in Inv4m
4. Integrate support metrics (LFP) with preservation status
5. Summarize biological themes and disruption classification

**Expected outcome**: Functional annotation of all modules with
consensus support metrics, preservation status, and hub gene connectivity.

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(clusterProfiler)
  library(here)
  library(patchwork)
  library(rrvgo)
  library(AnnotationHub)
  library(ggrepel)
  library(DOSE)
})

set.seed(123)
```

```{r setup_directories}
# Input/output directories
PRESERVATION_DIR <- here::here(params$run_dir, "06_preservation")
CONSENSUS_DIR <- here::here(params$run_dir, "04_consensus_networks")
BOOTSTRAP_DIR <- here::here(params$run_dir, "05_bootstrap_support")
FILTER_DIR <- here::here(params$run_dir, "02_gene_filter")
OUTPUT_DIR <- here::here(params$run_dir, "07_module_annotation")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

{
  cat("=== CONFIGURATION ===\n")
  cat("Run directory:", params$run_dir, "\n")
  cat("Output directory:", OUTPUT_DIR, "\n\n")
}
```

---

# Load Data

```{r load_data}
# Load preservation summary (with all metrics)
preservation_summary <- readRDS(file.path(PRESERVATION_DIR, "preservation_summary.rds"))

# Load disrupted modules
disrupted_modules <- read_csv(
  file.path(PRESERVATION_DIR, "disrupted_modules.csv"),
  show_col_types = FALSE
)

# Load CTRL consensus modules
ctrl_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)

# Load delta kWithin (gene-level)
delta_k <- read_csv(
  file.path(PRESERVATION_DIR, "delta_kWithin_genes.csv"),
  show_col_types = FALSE
)

# Load bootstrap support
ctrl_support <- read_csv(
  file.path(BOOTSTRAP_DIR, "ctrl", "module_support.csv"),
  show_col_types = FALSE
)

# Load gene annotations from filter step
gene_info <- read_csv(
  file.path(FILTER_DIR, "deg_neighborhood_genes.csv"),
  show_col_types = FALSE
)

# Load GO annotations
TERM2GENE <- readRDS(here("data", "TERM2GENE.rds"))
TERM2NAME <- readRDS(here("data", "TERM2NAME.rds"))

# Load effects from differential expression (for locus labels in hub gene plot)
effects_df <- read_csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  show_col_types = FALSE
) %>%
  filter(predictor == "Inv4m") %>%
  group_by(gene) %>%
  arrange(adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup()

{
  cat("=== INPUT DATA ===\n")
  cat("Preservation summary:", nrow(preservation_summary), "modules\n")
  cat("Disrupted modules:", nrow(disrupted_modules), "\n")
  cat("Genes with module assignments:", nrow(ctrl_modules), "\n")
  cat("Genes with delta_k:", nrow(delta_k), "\n")
  cat("GO term-gene mappings:", nrow(TERM2GENE), "\n\n")

  if (nrow(disrupted_modules) > 0) {
    cat("Disrupted modules:\n")
    print(disrupted_modules %>%
            dplyr::select(module, color, n_genes, adjacency_correlation,
                          median_delta_k, ctrl_stability))
  } else {
    cat("No disrupted modules identified.\n")
    cat("Will analyze modules with weakest preservation.\n")
  }
}
```

---

# Prepare Module Gene Lists for GO Analysis

```{r prepare_gene_lists}
# Load combined expression (merged experiments, low-expression filtered)
DATA_PREP_DIR <- here::here(params$run_dir, "01_data_prep")
expr_combined <- readRDS(file.path(DATA_PREP_DIR, "combined_expression.rds"))

# Universe = ALL genes after merging + low-expression filtering (~18K genes)
universe <- rownames(expr_combined)

cat("=== GENE UNIVERSE ===\n")
cat("Universe genes (merged + filtered):", length(universe), "\n")

# Get ALL CTRL modules (excluding grey/0)
all_modules <- ctrl_modules %>%
  filter(module != 0) %>%
  distinct(module, color) %>%
  arrange(module)

cat("CTRL modules to analyze:", nrow(all_modules), "\n\n")

# Create named list of gene vectors for each module (by color)
module_gene_lists <- split(
  ctrl_modules %>% filter(module != 0) %>% pull(gene),
  ctrl_modules %>% filter(module != 0) %>% pull(color)
)

# Report module sizes
cat("Module sizes:\n")
print(sapply(module_gene_lists, length))
```

---

# GO Enrichment Analysis (BP, CC, MF)

**Approach**: Run GO enrichment for all three ontologies (Biological Process,
Cellular Component, Molecular Function) for ALL CTRL consensus modules using
`compareCluster` for comparative analysis.

```{r enrichment_function}
#' Perform GO Enrichment Analysis
#'
#' @param x Character vector of gene IDs
#' @param ontology GO ontology (default: "BP")
#' @return enrichResult object
ego <- function(x, ontology = "BP") {
  enricher(
    gene = x,
    pvalueCutoff = 0.05,
    pAdjustMethod = "none",
    universe = universe,
    minGSSize = 10,
    maxGSSize = 500,
    qvalueCutoff = 0.2,
    TERM2GENE = TERM2GENE[TERM2GENE$GO %in% TERM2NAME[[ontology]]$go_id, ],
    TERM2NAME = TERM2NAME[[ontology]]
  )
}
```

```{r run_enrichment}
# Run comparative GO enrichment for all modules
cat("Running GO BP enrichment for all modules...\n")

compGO_BP <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego
)

# Set factor levels to preserve module order
cluster_order <- names(module_gene_lists)
levels(compGO_BP@compareClusterResult$Cluster) <- cluster_order

{
  cat("\n=== GO ENRICHMENT RESULTS ===\n")
  cat("Enriched terms per module:\n")
  print(table(compGO_BP@compareClusterResult$Cluster))
}

# Save full results
go_results <- as.data.frame(compGO_BP@compareClusterResult) %>%
  dplyr::rename(module_color = Cluster) %>%
  arrange(module_color, pvalue)

write_csv(
  go_results,
  file.path(OUTPUT_DIR, "all_modules_GO_BP.csv")
)

# Save the compareCluster object for later use
saveRDS(compGO_BP, file.path(OUTPUT_DIR, "compareCluster_GO_BP.rds"))

cat("\nTotal enriched BP terms:", nrow(go_results), "\n")
```

## rrvgo Semantic Redundancy Reduction

**Purpose**: Reduce redundant GO terms using semantic similarity to identify
representative parent terms for each module. This provides cleaner annotations
for the boxplot visualization.

```{r rrvgo_reduction}
cat("\n=== APPLYING RRVGO SEMANTIC REDUCTION ===\n")

# Load maize annotation for semantic similarity
ah <- AnnotationHub()
org.Zm.eg.db <- ah[["AH119718"]]

# Filter low-count terms first
filtered_bp_result <- go_results %>%
  filter(Count > 1) %>%
  group_by(module_color) %>%
  arrange(pvalue, .by_group = TRUE) %>%
  ungroup()

# Calculate semantic similarity matrix across all GO terms
go_terms_all <- unique(go_results$ID)

if (length(go_terms_all) > 1) {
  cat("Calculating semantic similarity for", length(go_terms_all), "GO terms...\n")

  simMatrix <- calculateSimMatrix(
    go_terms_all,
    orgdb = org.Zm.eg.db,
    ont = "BP",
    method = "Rel"
  )

  # Reduce to representative (parent) terms
  # Use -log10(pvalue) as scores to prioritize significant terms
  scores <- setNames(
    -log10(go_results$pvalue),
    go_results$ID
  )

  reducedTerms <- reduceSimMatrix(
    simMatrix,
    scores,
    threshold = 0.7,
    orgdb = org.Zm.eg.db
  )

  # Filter to keep only parent terms
  go_results_rrvgo <- filtered_bp_result %>%
    filter(ID %in% reducedTerms$parent)

  cat("Terms after rrvgo reduction:", nrow(go_results_rrvgo),
      "(from", nrow(go_results), ")\n")

} else {
  go_results_rrvgo <- filtered_bp_result
  cat("Skipping rrvgo: only", length(go_terms_all), "GO terms\n")
}

# Extract top 1 BP term per module for boxplot annotation
top_BP_per_module <- go_results_rrvgo %>%
  group_by(module_color) %>%
  slice_min(pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(module_color, ID, Description, pvalue, Count)

# Save for use in 06_preservation.Rmd boxplot
write_csv(
  top_BP_per_module,
  file.path(OUTPUT_DIR, "top_BP_per_module_rrvgo.csv")
)

cat("\nTop BP term per module (for boxplot annotation):\n")
print(top_BP_per_module)

# Save full rrvgo-reduced results
write_csv(
  go_results_rrvgo,
  file.path(OUTPUT_DIR, "all_modules_GO_BP_rrvgo.csv")
)

cat("\nFiles saved:\n")
cat("  - top_BP_per_module_rrvgo.csv (for boxplot annotation)\n")
cat("  - all_modules_GO_BP_rrvgo.csv (full rrvgo-reduced results)\n")
```

---

## Cellular Component (CC) Enrichment

```{r run_enrichment_cc}
# Run comparative GO CC enrichment for all modules
cat("Running GO CC enrichment for all modules...\n")

# Create CC-specific enrichment function
ego_cc <- function(x) ego(x, ontology = "CC")

compGO_CC <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego_cc
)

# Set factor levels to preserve module order
if (!is.null(compGO_CC) && nrow(compGO_CC@compareClusterResult) > 0) {
  levels(compGO_CC@compareClusterResult$Cluster) <- cluster_order
}

{
  cat("\n=== GO CC ENRICHMENT RESULTS ===\n")
  if (!is.null(compGO_CC) && nrow(compGO_CC@compareClusterResult) > 0) {
    cat("Enriched terms per module:\n")
    print(table(compGO_CC@compareClusterResult$Cluster))
  } else {
    cat("No significant CC enrichment found.\n")
  }
}

# Save full results
if (!is.null(compGO_CC) && nrow(compGO_CC@compareClusterResult) > 0) {
  go_results_cc <- as.data.frame(compGO_CC@compareClusterResult) %>%
    dplyr::rename(module_color = Cluster) %>%
    arrange(module_color, pvalue)

  write_csv(
    go_results_cc,
    file.path(OUTPUT_DIR, "all_modules_GO_CC.csv")
  )

  saveRDS(compGO_CC, file.path(OUTPUT_DIR, "compareCluster_GO_CC.rds"))
  cat("\nTotal enriched CC terms:", nrow(go_results_cc), "\n")
} else {
  go_results_cc <- tibble()
  cat("\nNo significant CC enrichment found.\n")
}
```

## rrvgo Reduction for CC

```{r rrvgo_cc}
# Apply rrvgo semantic reduction to CC results
if (nrow(go_results_cc) > 0) {
  cat("\n=== APPLYING RRVGO TO CC ===\n")

  filtered_cc_result <- go_results_cc %>%
    filter(Count > 1) %>%
    group_by(module_color) %>%
    arrange(pvalue, .by_group = TRUE) %>%
    ungroup()

  go_terms_cc <- unique(go_results_cc$ID)

  if (length(go_terms_cc) > 1) {
    cat("Calculating semantic similarity for", length(go_terms_cc), "CC terms...\n")

    simMatrix_cc <- calculateSimMatrix(
      go_terms_cc,
      orgdb = org.Zm.eg.db,
      ont = "CC",
      method = "Rel"
    )

    scores_cc <- setNames(-log10(go_results_cc$pvalue), go_results_cc$ID)

    reducedTerms_cc <- reduceSimMatrix(
      simMatrix_cc,
      scores_cc,
      threshold = 0.7,
      orgdb = org.Zm.eg.db
    )

    go_results_cc_rrvgo <- filtered_cc_result %>%
      filter(ID %in% reducedTerms_cc$parent)

    cat("CC terms after rrvgo:", nrow(go_results_cc_rrvgo),
        "(from", nrow(go_results_cc), ")\n")
  } else {
    go_results_cc_rrvgo <- filtered_cc_result
  }

  write_csv(
    go_results_cc_rrvgo,
    file.path(OUTPUT_DIR, "all_modules_GO_CC_rrvgo.csv")
  )
} else {
  go_results_cc_rrvgo <- tibble()
}
```

## Molecular Function (MF) Enrichment

```{r run_enrichment_mf}
# Run comparative GO MF enrichment for all modules
cat("Running GO MF enrichment for all modules...\n")

# Create MF-specific enrichment function
ego_mf <- function(x) ego(x, ontology = "MF")

compGO_MF <- compareCluster(
  geneCluster = module_gene_lists,
  fun = ego_mf
)

# Set factor levels to preserve module order
if (!is.null(compGO_MF) && nrow(compGO_MF@compareClusterResult) > 0) {
  levels(compGO_MF@compareClusterResult$Cluster) <- cluster_order
}

{
  cat("\n=== GO MF ENRICHMENT RESULTS ===\n")
  if (!is.null(compGO_MF) && nrow(compGO_MF@compareClusterResult) > 0) {
    cat("Enriched terms per module:\n")
    print(table(compGO_MF@compareClusterResult$Cluster))
  } else {
    cat("No significant MF enrichment found.\n")
  }
}

# Save full results
if (!is.null(compGO_MF) && nrow(compGO_MF@compareClusterResult) > 0) {
  go_results_mf <- as.data.frame(compGO_MF@compareClusterResult) %>%
    dplyr::rename(module_color = Cluster) %>%
    arrange(module_color, pvalue)

  write_csv(
    go_results_mf,
    file.path(OUTPUT_DIR, "all_modules_GO_MF.csv")
  )

  saveRDS(compGO_MF, file.path(OUTPUT_DIR, "compareCluster_GO_MF.rds"))
  cat("\nTotal enriched MF terms:", nrow(go_results_mf), "\n")
} else {
  go_results_mf <- tibble()
  cat("\nNo significant MF enrichment found.\n")
}
```

## rrvgo Reduction for MF

```{r rrvgo_mf}
# Apply rrvgo semantic reduction to MF results
if (nrow(go_results_mf) > 0) {
  cat("\n=== APPLYING RRVGO TO MF ===\n")

  filtered_mf_result <- go_results_mf %>%
    filter(Count > 1) %>%
    group_by(module_color) %>%
    arrange(pvalue, .by_group = TRUE) %>%
    ungroup()

  go_terms_mf <- unique(go_results_mf$ID)

  if (length(go_terms_mf) > 1) {
    cat("Calculating semantic similarity for", length(go_terms_mf), "MF terms...\n")

    simMatrix_mf <- calculateSimMatrix(
      go_terms_mf,
      orgdb = org.Zm.eg.db,
      ont = "MF",
      method = "Rel"
    )

    scores_mf <- setNames(-log10(go_results_mf$pvalue), go_results_mf$ID)

    reducedTerms_mf <- reduceSimMatrix(
      simMatrix_mf,
      scores_mf,
      threshold = 0.7,
      orgdb = org.Zm.eg.db
    )

    go_results_mf_rrvgo <- filtered_mf_result %>%
      filter(ID %in% reducedTerms_mf$parent)

    cat("MF terms after rrvgo:", nrow(go_results_mf_rrvgo),
        "(from", nrow(go_results_mf), ")\n")
  } else {
    go_results_mf_rrvgo <- filtered_mf_result
  }

  write_csv(
    go_results_mf_rrvgo,
    file.path(OUTPUT_DIR, "all_modules_GO_MF_rrvgo.csv")
  )
} else {
  go_results_mf_rrvgo <- tibble()
}
```

---

# Top GO Terms Summary

```{r go_summary}
# BP Summary
if (nrow(go_results) > 0) {
  go_summary_bp <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3), ontology = "BP")

  write_csv(
    go_summary_bp,
    file.path(OUTPUT_DIR, "go_summary_BP.csv")
  )

  cat("\n=== TOP GO BP TERMS PER MODULE ===\n")
  print(go_summary_bp, n = 30)
} else {
  go_summary_bp <- tibble()
  cat("\nNo significant BP enrichment found.\n")
}

# CC Summary
if (nrow(go_results_cc) > 0) {
  go_summary_cc <- go_results_cc %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3), ontology = "CC")

  write_csv(
    go_summary_cc,
    file.path(OUTPUT_DIR, "go_summary_CC.csv")
  )

  cat("\n=== TOP GO CC TERMS PER MODULE ===\n")
  print(go_summary_cc, n = 30)
} else {
  go_summary_cc <- tibble()
  cat("\nNo significant CC enrichment found.\n")
}

# MF Summary
if (nrow(go_results_mf) > 0) {
  go_summary_mf <- go_results_mf %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module_color, ID, Description, GeneRatio, pvalue, Count) %>%
    mutate(pvalue = signif(pvalue, 3), ontology = "MF")

  write_csv(
    go_summary_mf,
    file.path(OUTPUT_DIR, "go_summary_MF.csv")
  )

  cat("\n=== TOP GO MF TERMS PER MODULE ===\n")
  print(go_summary_mf, n = 30)
} else {
  go_summary_mf <- tibble()
  cat("\nNo significant MF enrichment found.\n")
}

# Combined summary table (top 3 per ontology per module)
go_summary_combined <- bind_rows(
  go_summary_bp %>% group_by(module_color) %>% slice_min(pvalue, n = 3, with_ties = FALSE),
  go_summary_cc %>% group_by(module_color) %>% slice_min(pvalue, n = 3, with_ties = FALSE),
  go_summary_mf %>% group_by(module_color) %>% slice_min(pvalue, n = 3, with_ties = FALSE)
) %>%
  arrange(module_color, ontology, pvalue)

write_csv(
  go_summary_combined,
  file.path(OUTPUT_DIR, "go_summary_all_ontologies.csv")
)

cat("\n=== COMBINED GO SUMMARY (all ontologies) ===\n")
cat("Total terms in combined summary:", nrow(go_summary_combined), "\n")
```

---

# Supplementary Figure: GO Enrichment (rrvgo-reduced)

**Purpose**: Generate comprehensive GO enrichment bubble plots in Trans Network style
for supplementary material. Uses DOSE::theme_dose for visual consistency.

```{r bubble_plot_function}
#' Create bubble plot for GO enrichment (Trans Network style)
#'
#' @param go_data Data frame with GO enrichment results
#' @param ontology_name Name of the ontology for plot title
#' @param module_sizes Named vector of module sizes
#' @return ggplot object
create_bubble_plot <- function(go_data, ontology_name, module_sizes) {

  if (is.null(go_data) || nrow(go_data) == 0) {
    cat("No significant", ontology_name, "enrichment to plot.\n")
    return(NULL)
  }

  # Prepare data - top 5 per module
  to_plot <- go_data %>%
    filter(pvalue < 0.05) %>%
    mutate(neglogP = -log10(pvalue)) %>%
    group_by(module_color) %>%
    arrange(pvalue) %>%
    slice_head(n = 5) %>%
    mutate(
      label_order = 1:n(),
      Description = forcats::fct_reorder(Description, label_order)
    ) %>%
    ungroup()

  if (nrow(to_plot) == 0) {
    cat("No terms with p < 0.05 for", ontology_name, "\n")
    return(NULL)
  }

  # Module labels with gene counts
  module_order <- unique(to_plot$module_color)
  labels <- paste0(module_order, "\n(", module_sizes[module_order], ")")
  names(labels) <- module_order

  p <- to_plot %>%
    ggplot(aes(x = module_color, y = Description, size = Count, fill = neglogP)) +
    ggtitle(paste("GO", ontology_name, "Enrichment by WGCNA Module")) +
    scale_fill_gradient(
      low = "dodgerblue",
      high = "tomato",
      name = "-log10(p)",
      limits = c(1, max(to_plot$neglogP, na.rm = TRUE))
    ) +
    scale_y_discrete(
      labels = scales::wrap_format(50),
      limits = rev(levels(to_plot$Description))
    ) +
    scale_x_discrete(labels = labels, drop = FALSE) +
    geom_point(shape = 21) +
    scale_size(range = c(4, 8), name = "Count") +
    ylab(NULL) +
    xlab(NULL) +
    DOSE::theme_dose(font.size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )

  return(p)
}

# Get module sizes for labels
module_sizes <- sapply(module_gene_lists, length)
```

```{r go_supplementary_figure_bp, fig.width=14, fig.height=10}
# Create BP bubble plot (Trans Network style)
if (exists("go_results_rrvgo") && nrow(go_results_rrvgo) > 0) {

  p_go_bp <- create_bubble_plot(go_results_rrvgo, "Biological Process", module_sizes)

  if (!is.null(p_go_bp)) {
    print(p_go_bp)

    # Save supplementary figure
    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_BP_rrvgo.pdf"),
      p_go_bp,
      width = 14, height = 10
    )

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_BP_rrvgo.png"),
      p_go_bp,
      width = 14, height = 10, dpi = 300
    )

    # Save RDS for composition
    saveRDS(p_go_bp, file.path(OUTPUT_DIR, "go_enrichment_BP_rrvgo.rds"))

    cat("\nBP supplementary figure saved:\n")
    cat("  - go_enrichment_BP_rrvgo.pdf\n")
    cat("  - go_enrichment_BP_rrvgo.png\n")
    cat("  - go_enrichment_BP_rrvgo.rds\n")
  }
} else {
  cat("\nNo rrvgo-reduced GO results available for supplementary figure.\n")
}
```

```{r go_supplementary_figure_cc, fig.width=14, fig.height=10}
# Create CC bubble plot (rrvgo-reduced)
if (exists("go_results_cc_rrvgo") && nrow(go_results_cc_rrvgo) > 0) {

  p_go_cc <- create_bubble_plot(go_results_cc_rrvgo, "Cellular Component", module_sizes)

  if (!is.null(p_go_cc)) {
    print(p_go_cc)

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_CC_rrvgo.pdf"),
      p_go_cc,
      width = 14, height = 10
    )

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_CC_rrvgo.png"),
      p_go_cc,
      width = 14, height = 10, dpi = 300
    )

    saveRDS(p_go_cc, file.path(OUTPUT_DIR, "go_enrichment_CC_rrvgo.rds"))

    cat("\nCC supplementary figure saved (rrvgo-reduced):\n")
    cat("  - go_enrichment_CC_rrvgo.pdf\n")
    cat("  - go_enrichment_CC_rrvgo.png\n")
    cat("  - go_enrichment_CC_rrvgo.rds\n")
  }
}
```

```{r go_supplementary_figure_mf, fig.width=14, fig.height=10}
# Create MF bubble plot (rrvgo-reduced)
if (exists("go_results_mf_rrvgo") && nrow(go_results_mf_rrvgo) > 0) {

  p_go_mf <- create_bubble_plot(go_results_mf_rrvgo, "Molecular Function", module_sizes)

  if (!is.null(p_go_mf)) {
    print(p_go_mf)

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_MF_rrvgo.pdf"),
      p_go_mf,
      width = 14, height = 10
    )

    ggsave(
      file.path(OUTPUT_DIR, "go_enrichment_MF_rrvgo.png"),
      p_go_mf,
      width = 14, height = 10, dpi = 300
    )

    saveRDS(p_go_mf, file.path(OUTPUT_DIR, "go_enrichment_MF_rrvgo.rds"))

    cat("\nMF supplementary figure saved (rrvgo-reduced):\n")
    cat("  - go_enrichment_MF_rrvgo.pdf\n")
    cat("  - go_enrichment_MF_rrvgo.png\n")
    cat("  - go_enrichment_MF_rrvgo.rds\n")
  }
}
```

---

# Hub Gene Analysis

**Approach**: Identify hub genes using CTRL consensus kWithin, then check
connectivity change in Inv4m. Analyze ALL modules (excluding grey/0).

```{r hub_analysis}
# Merge module assignments with delta kWithin (all modules except grey)
hub_data <- ctrl_modules %>%
  filter(module != 0) %>%
  inner_join(
    delta_k %>% dplyr::select(gene, kWithin_ctrl, kWithin_inv4m, delta_kWithin),
    by = "gene"
  )

# Top 5 hub genes per module by CTRL kWithin
hub_genes <- hub_data %>%
  group_by(module, color) %>%
  slice_max(kWithin_ctrl, n = 5) %>%
  ungroup()

# Add gene annotations (including locus_label for plotting)
# Load consolidated labels from master file (includes curated + symbol + LLM-proposed)
master_labels <- read_csv(
  file.path(paths$data, "locus_labels_master.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(gene, locus_label, locus_symbol)

hub_genes <- hub_genes %>%
  left_join(
    gene_info %>% dplyr::select(gene, logFC, regulation, in_cis, in_Inv4m),
    by = "gene"
  ) %>%
  left_join(master_labels, by = "gene") %>%
  mutate(
    hub_disrupted = delta_kWithin < -0.05,
    connectivity_change = case_when(
      delta_kWithin < -0.1 ~ "Strong loss",
      delta_kWithin < -0.02 ~ "Moderate loss",
      delta_kWithin > 0.02 ~ "Gain",
      TRUE ~ "Stable"
    ),
    # Use locus_label if available, else filtered locus_symbol, else no label
    # Filter out marker names (umc, pco, etc.) from display
    display_label = case_when(
      !is.na(locus_label) & locus_label != "" ~ locus_label,
      !is.na(locus_symbol) & locus_symbol != "" &
        !grepl("^(umc|pco|bnlg|pza|gpm|csu|php|IDP|TIDP)[0-9]+|^cl[0-9]+_", locus_symbol) ~ locus_symbol,
      TRUE ~ NA_character_
    )
  ) %>%
  arrange(module, desc(kWithin_ctrl))

# Save hub genes
write_csv(
  hub_genes,
  file.path(OUTPUT_DIR, "hub_genes_all_modules.csv")
)

{
  cat("\n=== HUB GENES (ALL MODULES) ===\n")
  cat("Total hub genes:", nrow(hub_genes), "\n")
  cat("Hubs with connectivity loss:", sum(hub_genes$hub_disrupted, na.rm = TRUE), "\n\n")

  print(
    hub_genes %>%
      dplyr::select(module, color, gene, kWithin_ctrl, delta_kWithin,
                    connectivity_change, regulation)
  )
}
```

---

# Hub Gene Connectivity Plot

```{r hub_connectivity, fig.width=10, fig.height=8}
if (nrow(hub_genes) > 0) {
  p_hub <- ggplot(
    hub_genes,
    aes(x = kWithin_ctrl, y = kWithin_inv4m, color = color)
  ) +
    geom_point(aes(size = abs(logFC)), alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_text_repel(
      aes(label = display_label),
      size = 3,
      color = "black",
      max.overlaps = 25,
      box.padding = 0.35,
      point.padding = 0.3,
      segment.color = "gray70",
      na.rm = TRUE
    ) +
    scale_color_identity() +
    theme_minimal(base_size = 12) +
    labs(
      title = "Hub Gene Connectivity Change (Consensus Adjacency)",
      subtitle = "Points below diagonal = connectivity loss in Inv4m",
      x = "kWithin (CTRL)",
      y = "kWithin (Inv4m)",
      size = "|logFC|"
    )

  print(p_hub)

  ggsave(
    file.path(OUTPUT_DIR, "hub_connectivity_plot.pdf"),
    p_hub,
    width = 10, height = 8
  )

  ggsave(
    file.path(OUTPUT_DIR, "hub_connectivity_plot.png"),
    p_hub,
    width = 10, height = 8, dpi = 300
  )

  # Save as RDS for figure composition
  saveRDS(p_hub, file.path(OUTPUT_DIR, "hub_connectivity_plot.rds"))

  cat("\nHub connectivity plot saved:\n")
  cat("  - hub_connectivity_plot.pdf\n")
  cat("  - hub_connectivity_plot.png\n")
  cat("  - hub_connectivity_plot.rds\n")
}
```

---

# Module Theme Summary with Support Metrics

```{r biological_themes}
# Create comprehensive summary for ALL modules using preservation_summary
module_themes <- preservation_summary %>%
  dplyr::select(
    module, color, n_genes, adjacency_correlation, median_delta_k,
    pct_decreased, ctrl_lfp, ctrl_stability, jaccard, preservation_status
  )

# Add top GO terms (BP, CC, MF)
if (nrow(go_results) > 0) {
  top_bp <- go_results %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_BP = Description, bp_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_bp, by = "color")
}

if (nrow(go_results_cc) > 0) {
  top_cc <- go_results_cc %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_CC = Description, cc_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_cc, by = "color")
}

if (nrow(go_results_mf) > 0) {
  top_mf <- go_results_mf %>%
    group_by(module_color) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(color = module_color, top_MF = Description, mf_pval = pvalue)

  module_themes <- module_themes %>%
    left_join(top_mf, by = "color")
}

# Add top hub gene
if (nrow(hub_genes) > 0) {
  top_hubs <- hub_genes %>%
    group_by(module) %>%
    slice_max(kWithin_ctrl, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module, top_hub = gene, hub_delta_k = delta_kWithin,
                  hub_regulation = regulation)

  module_themes <- module_themes %>%
    left_join(top_hubs, by = "module")
}

# Save theme summary
write_csv(
  module_themes,
  file.path(OUTPUT_DIR, "module_theme_summary.csv")
)

{
  cat("\n=== MODULE BIOLOGICAL THEMES WITH SUPPORT ===\n")
  print(module_themes %>%
          dplyr::select(module, color, n_genes, ctrl_lfp, ctrl_stability,
                        adjacency_correlation, preservation_status, top_BP))
}
```

---

# Integration with Inv4m Direction

```{r inv4m_direction}
# Check direction bias in ALL modules (excluding grey)
direction_summary <- ctrl_modules %>%
  filter(module != 0) %>%
  left_join(
    gene_info %>% dplyr::select(gene, regulation, in_cis, in_Inv4m),
    by = "gene"
  ) %>%
  group_by(module, color) %>%
  summarise(
    n_genes = n(),
    n_upregulated = sum(regulation == "Upregulated", na.rm = TRUE),
    n_downregulated = sum(regulation == "Downregulated", na.rm = TRUE),
    pct_upregulated = round(n_upregulated / n_genes * 100, 1),
    n_cis = sum(in_cis, na.rm = TRUE),
    pct_cis = round(n_cis / n_genes * 100, 1),
    n_in_inv4m = sum(in_Inv4m, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(
  direction_summary,
  file.path(OUTPUT_DIR, "module_direction_summary.csv")
)

{
  cat("\n=== INV4M DIRECTION IN MODULES ===\n")
  print(direction_summary)
}
```

---

# Comprehensive Module Summary Table

```{r comprehensive_summary}
# Create final comprehensive summary
comprehensive_summary <- preservation_summary %>%
  left_join(
    ctrl_support %>% dplyr::select(module, mean_entropy, n_genes_support = n_genes),
    by = "module"
  ) %>%
  left_join(
    direction_summary %>% dplyr::select(module, pct_upregulated, pct_cis),
    by = "module"
  )

# Add GO term counts for all ontologies
if (nrow(go_results) > 0) {
  go_counts_bp <- go_results %>%
    count(module_color, name = "n_BP_terms") %>%
    dplyr::rename(color = module_color)
  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts_bp, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_BP_terms = 0)
}

if (nrow(go_results_cc) > 0) {
  go_counts_cc <- go_results_cc %>%
    count(module_color, name = "n_CC_terms") %>%
    dplyr::rename(color = module_color)
  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts_cc, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_CC_terms = 0)
}

if (nrow(go_results_mf) > 0) {
  go_counts_mf <- go_results_mf %>%
    count(module_color, name = "n_MF_terms") %>%
    dplyr::rename(color = module_color)
  comprehensive_summary <- comprehensive_summary %>%
    left_join(go_counts_mf, by = "color")
} else {
  comprehensive_summary <- comprehensive_summary %>%
    mutate(n_MF_terms = 0)
}

# Replace NA counts with 0
comprehensive_summary <- comprehensive_summary %>%
  mutate(
    n_BP_terms = replace_na(n_BP_terms, 0),
    n_CC_terms = replace_na(n_CC_terms, 0),
    n_MF_terms = replace_na(n_MF_terms, 0),
    n_GO_total = n_BP_terms + n_CC_terms + n_MF_terms
  )

# Classify final disruption status
comprehensive_summary <- comprehensive_summary %>%
  mutate(
    disrupted_final = case_when(
      adjacency_correlation < 0.3 ~ "Strong",
      adjacency_correlation < 0.5 ~ "Moderate",
      adjacency_correlation < 0.7 & median_delta_k < -0.05 ~ "Weak",
      TRUE ~ "Not disrupted"
    ),
    disrupted_final = factor(
      disrupted_final,
      levels = c("Strong", "Moderate", "Weak", "Not disrupted")
    )
  )

write_csv(
  comprehensive_summary,
  file.path(OUTPUT_DIR, "comprehensive_module_summary.csv")
)

{
  cat("\n=== COMPREHENSIVE MODULE SUMMARY ===\n")
  cat("\nDisruption classification:\n")
  print(table(comprehensive_summary$disrupted_final))

  cat("\nTop disrupted modules:\n")
  print(
    comprehensive_summary %>%
      filter(disrupted_final != "Not disrupted") %>%
      dplyr::select(module, color, n_genes, ctrl_lfp, ctrl_stability,
                    adjacency_correlation, disrupted_final, pct_upregulated) %>%
      arrange(adjacency_correlation)
  )
}
```

---

# Save Final Summary Statistics

```{r save_summary}
n_modules_analyzed <- nrow(all_modules)
n_modules_with_bp <- length(unique(go_results$module_color))
n_modules_with_cc <- length(unique(go_results_cc$module_color))
n_modules_with_mf <- length(unique(go_results_mf$module_color))

summary_stats <- tibble(
  metric = c(
    "Total CTRL modules analyzed",
    "GO BP terms enriched (total)",
    "GO CC terms enriched (total)",
    "GO MF terms enriched (total)",
    "Modules with BP enrichment",
    "Modules with CC enrichment",
    "Modules with MF enrichment",
    "Hub genes identified",
    "Hub genes with connectivity loss",
    "Strong disruption modules",
    "Moderate disruption modules",
    "Weak disruption modules"
  ),
  value = c(
    n_modules_analyzed,
    nrow(go_results),
    nrow(go_results_cc),
    nrow(go_results_mf),
    n_modules_with_bp,
    n_modules_with_cc,
    n_modules_with_mf,
    nrow(hub_genes),
    sum(hub_genes$hub_disrupted, na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Strong", na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Moderate", na.rm = TRUE),
    sum(comprehensive_summary$disrupted_final == "Weak", na.rm = TRUE)
  )
)

write_csv(
  summary_stats,
  file.path(OUTPUT_DIR, "functional_analysis_summary.csv")
)

{
  cat("\n=== SAVED OUTPUTS ===\n")
  cat("GO Enrichment Results:\n")
  cat("  all_modules_GO_BP.csv: GO BP enrichment for all modules\n")
  cat("  all_modules_GO_BP_rrvgo.csv: rrvgo-reduced BP terms\n")
  cat("  top_BP_per_module_rrvgo.csv: top BP term per module (for boxplot)\n")
  cat("  all_modules_GO_CC.csv: GO CC enrichment for all modules\n")
  cat("  all_modules_GO_CC_rrvgo.csv: rrvgo-reduced CC terms\n")
  cat("  all_modules_GO_MF.csv: GO MF enrichment for all modules\n")
  cat("  all_modules_GO_MF_rrvgo.csv: rrvgo-reduced MF terms\n")
  cat("  compareCluster_GO_BP/CC/MF.rds: compareCluster objects\n")
  cat("  go_summary_BP/CC/MF.csv: top terms per module\n")
  cat("  go_summary_all_ontologies.csv: combined summary\n")
  cat("\nVisualization (Trans Network style bubble plots, all rrvgo-reduced):\n")
  cat("  go_enrichment_BP_rrvgo.pdf/png/rds: BP enrichment\n")
  cat("  go_enrichment_CC_rrvgo.pdf/png/rds: CC enrichment\n")
  cat("  go_enrichment_MF_rrvgo.pdf/png/rds: MF enrichment\n")
  cat("  hub_connectivity_plot.pdf/png/rds: hub gene connectivity (ggrepel labels)\n")
  cat("\nSummary Tables:\n")
  cat("  hub_genes_all_modules.csv: hub genes with connectivity change\n")
  cat("  module_theme_summary.csv: biological themes with support metrics\n")
  cat("  module_direction_summary.csv: Inv4m direction summary\n")
  cat("  comprehensive_module_summary.csv: all modules with all metrics\n")
  cat("  functional_analysis_summary.csv: summary statistics\n")
}
```

---

# Validation Checks

```{r validation}
{
  cat("=== VALIDATION CHECKS ===\n\n")

  # Check 1: Modules analyzed
  cat("Check 1: Modules analyzed\n")
  cat("  Modules:", nrow(all_modules), "\n")
  stopifnot(nrow(all_modules) > 0)

  # Check 2: GO enrichment ran (all ontologies)
  cat("\nCheck 2: GO enrichment executed\n")
  cat("  BP terms found:", nrow(go_results), "\n")
  cat("  CC terms found:", nrow(go_results_cc), "\n")
  cat("  MF terms found:", nrow(go_results_mf), "\n")

  # Check 3: Hub genes identified
  cat("\nCheck 3: Hub genes identified\n")
  cat("  Hub genes:", nrow(hub_genes), "\n")
  stopifnot(nrow(hub_genes) > 0)

  # Check 4: Comprehensive summary complete
  cat("\nCheck 4: Comprehensive summary\n")
  cat("  Modules in summary:", nrow(comprehensive_summary), "\n")
  stopifnot(nrow(comprehensive_summary) == nrow(preservation_summary))

  # Check 5: All outputs saved
  cat("\nCheck 5: Output files created\n")
  expected_files <- c(
    "all_modules_GO_BP.csv",
    "all_modules_GO_BP_rrvgo.csv",
    "top_BP_per_module_rrvgo.csv",
    "all_modules_GO_CC.csv",
    "all_modules_GO_CC_rrvgo.csv",
    "all_modules_GO_MF.csv",
    "all_modules_GO_MF_rrvgo.csv",
    "go_summary_all_ontologies.csv",
    "hub_genes_all_modules.csv",
    "hub_connectivity_plot.rds",
    "go_enrichment_BP_rrvgo.rds",
    "go_enrichment_CC_rrvgo.rds",
    "go_enrichment_MF_rrvgo.rds",
    "functional_analysis_summary.csv",
    "comprehensive_module_summary.csv"
  )
  for (f in expected_files) {
    exists <- file.exists(file.path(OUTPUT_DIR, f))
    cat("  ", f, ":", ifelse(exists, "OK", "MISSING"), "\n")
  }

  cat("\n=== ALL CHECKS PASSED ===\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
