---
title: "Figure 1 contrast convention test (Welch t-test vs lm-emmeans, conditional vs marginal)"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "phosphorus_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source(here::here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("phosphorus_paper")
```

## Goal

Compare four conventions for the brackets in Figure 1 (Phosphorus
Starvation Response). Each convention is applied **without lm pre-selection**
and **without grouping FDR pools** beyond the natural contrast family.

| Approach | Test | Contrast level | FDR pool |
|---|---|---|---|
| **A** | Welch t-test (current Fig 1) | per-Genotype | across all (Genotype × trait) |
| **B** | emmeans on `lm(y ~ Genotype * Treatment)` | per-Genotype | across all (Genotype × trait) |
| **C** | emmeans on same lm | marginal (pooled over Genotype) | across all traits |
| **D** | emmeans on same lm | per-Genotype **AND** marginal | combined pool |

Reported: raw p, FDR-adjusted p, and survival at FDR < 0.05 for each
trait × contrast.

## Load libraries and data

```{r libraries}
library(dplyr)
library(tidyr)
library(rstatix)
library(emmeans)
library(here)
```

```{r load-data}
pheno <- read.csv(file.path(paths$intermediate,
  "PSU2022_spatially_corrected_both_treatments_with_ionome.csv")) %>%
  rename(Genotype = genotype) %>%
  filter(Genotype %in% c("CTRL", "Inv4m")) %>%
  mutate(Treatment = factor(Treatment, levels = c("+P", "-P")),
         Genotype  = factor(Genotype, levels = c("CTRL", "Inv4m"))) %>%
  droplevels()

base_phenotypes <- c("PH", "DTA", "DTS", "STW40", "STW50", "STW60", "STWHV")
ear_phenotypes  <- c("EW", "EL", "ED", "KRN", "CD", "TKW", "KW50", "TKN", "HI")
vars <- intersect(c(base_phenotypes, ear_phenotypes), colnames(pheno))
cat("Phenotypes tested:", length(vars), "\n", paste(vars, collapse = ", "), "\n")
```

## Approach A — Welch per-Genotype + FDR

```{r approach-a}
to_long <- pheno %>%
  select(Genotype, Treatment, all_of(vars)) %>%
  pivot_longer(all_of(vars), names_to = "trait", values_to = "value") %>%
  filter(!is.na(value))

A <- to_long %>%
  group_by(Genotype, trait) %>%
  filter(n() >= 4, var(value) > 0) %>%
  rstatix::t_test(value ~ Treatment, ref.group = "+P") %>%
  rstatix::adjust_pvalue(method = "fdr") %>%
  mutate(approach = "A_Welch_perGeno", contrast_level = "per_Genotype")

cat("Approach A — Welch per-Genotype, FDR across all (Genotype x trait):\n")
A %>% select(trait, Genotype, statistic, p, p.adj) %>%
  arrange(p.adj) %>% as_tibble() %>% print(n = Inf)
```

## Approach B — per-Genotype emmeans on lm

```{r approach-b}
B_raw <- lapply(vars, function(v) {
  if (sum(!is.na(pheno[[v]])) < 20) return(NULL)
  m <- tryCatch(lm(as.formula(paste(v, "~ Genotype * Treatment")),
                   data = pheno),
                error = function(e) NULL)
  if (is.null(m)) return(NULL)
  pairs(emmeans(m, ~ Treatment | Genotype), reverse = TRUE) %>%
    as_tibble() %>%
    mutate(trait = v)
}) %>% bind_rows()

B <- B_raw %>%
  rename(p = p.value, statistic = t.ratio) %>%
  mutate(approach = "B_emmeans_perGeno", contrast_level = "per_Genotype") %>%
  rstatix::adjust_pvalue(method = "fdr")

cat("Approach B — emmeans per-Genotype, FDR across all (Genotype x trait):\n")
B %>% select(trait, Genotype, statistic, p, p.adj) %>%
  arrange(p.adj) %>% as_tibble() %>% print(n = Inf)
```

## Approach C — marginal emmeans on lm

```{r approach-c}
C_raw <- lapply(vars, function(v) {
  if (sum(!is.na(pheno[[v]])) < 20) return(NULL)
  m <- tryCatch(lm(as.formula(paste(v, "~ Genotype * Treatment")),
                   data = pheno),
                error = function(e) NULL)
  if (is.null(m)) return(NULL)
  pairs(emmeans(m, ~ Treatment), reverse = TRUE) %>%
    as_tibble() %>%
    mutate(trait = v)
}) %>% bind_rows()

C <- C_raw %>%
  rename(p = p.value, statistic = t.ratio) %>%
  mutate(approach = "C_emmeans_marginal", contrast_level = "marginal") %>%
  rstatix::adjust_pvalue(method = "fdr")

cat("Approach C — emmeans marginal, FDR across traits:\n")
C %>% select(trait, statistic, p, p.adj) %>%
  arrange(p.adj) %>% as_tibble() %>% print(n = Inf)
```

## Approach D — per-Genotype + marginal combined

```{r approach-d}
D_pre <- bind_rows(
  B %>% select(trait, Genotype, statistic, p, contrast_level),
  C %>% select(trait, statistic, p, contrast_level)
)

D <- D_pre %>%
  rstatix::adjust_pvalue(method = "fdr") %>%
  mutate(approach = "D_combined")

cat("Approach D — per-Genotype + marginal combined, FDR across all:\n")
D %>%
  mutate(label = ifelse(is.na(Genotype), "marginal", as.character(Genotype))) %>%
  select(trait, label, contrast_level, statistic, p, p.adj) %>%
  arrange(p.adj) %>% as_tibble() %>% print(n = Inf)
```

## Side-by-side comparison

```{r side-by-side}
A_pivot <- A %>%
  select(trait, Genotype, p_A_Welch_perGeno = p, padj_A_Welch_perGeno = p.adj) %>%
  mutate(Genotype = as.character(Genotype))

B_pivot <- B %>%
  select(trait, Genotype, p_B_emm_perGeno = p, padj_B_emm_perGeno = p.adj) %>%
  mutate(Genotype = as.character(Genotype))

C_pivot <- C %>%
  select(trait, p_C_emm_marginal = p, padj_C_emm_marginal = p.adj) %>%
  mutate(Genotype = "marginal")

D_geno <- D %>%
  filter(contrast_level == "per_Genotype") %>%
  select(trait, Genotype, padj_D_combined = p.adj) %>%
  mutate(Genotype = as.character(Genotype))

D_marg <- D %>%
  filter(contrast_level == "marginal") %>%
  select(trait, padj_D_combined = p.adj) %>%
  mutate(Genotype = "marginal")

geno_combined <- A_pivot %>%
  full_join(B_pivot, by = c("trait", "Genotype")) %>%
  full_join(D_geno,  by = c("trait", "Genotype"))

marg_combined <- C_pivot %>%
  full_join(D_marg, by = c("trait", "Genotype"))

cat("\n=== Per-Genotype contrasts (Approaches A, B, D-conditional) ===\n")
geno_combined %>%
  arrange(trait, Genotype) %>%
  mutate(across(starts_with("p"), ~ signif(.x, 3))) %>%
  as_tibble() %>%
  print(n = Inf)

cat("\n=== Marginal contrasts (Approaches C, D-marginal) ===\n")
marg_combined %>%
  arrange(trait) %>%
  mutate(across(starts_with("p"), ~ signif(.x, 3))) %>%
  as_tibble() %>%
  print(n = Inf)
```

## Survival summary

```{r survival}
summary_table <- bind_rows(
  A %>% transmute(trait, Genotype = as.character(Genotype),
                  contrast = "per_Genotype", approach = "A_Welch", padj = p.adj),
  B %>% transmute(trait, Genotype = as.character(Genotype),
                  contrast = "per_Genotype", approach = "B_emm",   padj = p.adj),
  C %>% transmute(trait, Genotype = "marginal",
                  contrast = "marginal",     approach = "C_emm",   padj = p.adj),
  D %>% transmute(trait,
                  Genotype = ifelse(is.na(Genotype), "marginal", as.character(Genotype)),
                  contrast = contrast_level,
                  approach = "D_combined", padj = p.adj)
)

cat("\n=== Total significant brackets at FDR < 0.05 ===\n")
summary_table %>%
  group_by(approach) %>%
  summarise(n_significant = sum(padj < 0.05, na.rm = TRUE),
            total_tests   = n(),
            .groups = "drop") %>%
  print()

cat("\n=== TKN, KW50, TKW survival under each approach ===\n")
summary_table %>%
  filter(trait %in% c("TKN", "KW50", "TKW")) %>%
  arrange(trait, contrast, approach) %>%
  mutate(padj = signif(padj, 3),
         signif_at_0.05 = ifelse(padj < 0.05, "*", "ns")) %>%
  print(n = Inf)
```

## Supplementary figure: marginal Treatment effects for PH, KW50, TKN, TKW

Boxplots pool CTRL and Inv4m; bracket labels come from the marginal
emmeans contrast FDR (48-contrast D pool). PH is included as a
phenotype with a robust marginal Treatment effect and non-significant
Genotype × Treatment interaction. KW50, TKN, and TKW are included
together to address Sawers' Figure 1 comment: under $-P$ the plant
trades kernel size for kernel number (KW50 drops significantly, TKN
rises significantly in the pooled marginal contrast), leaving total
kernel weight per plant (TKW) statistically unchanged.

```{r supp-marginal-figure, fig.width=18, fig.height=6}
library(ggplot2)
library(ggpubr)
library(ggbeeswarm)
library(ggfx)
library(ggtext)

pal_trt <- c("+P" = "gold", "-P" = "purple4")

supp_traits <- c("PH", "KW50", "TKN", "TKW")

# Pull the marginal p.adj for the four traits from the 48-contrast pool
marg_stats <- D %>%
  filter(contrast_level == "marginal", trait %in% supp_traits) %>%
  transmute(trait, group1 = "+P", group2 = "-P",
            p.adj = p.adj, p.adj.signif = add_significance(tibble(p.adj),
                                                           "p.adj")$p.adj.signif)

supp_panel <- function(trait_name, ylab_text, y_limits = NULL) {
  d <- pheno %>%
    select(Treatment, value = all_of(trait_name)) %>%
    filter(!is.na(value))

  s <- marg_stats %>% filter(trait == trait_name) %>%
    mutate(y.position = max(d$value, na.rm = TRUE) +
                        0.08 * diff(range(d$value, na.rm = TRUE)))

  p <- ggplot(d, aes(x = Treatment, y = value, colour = Treatment)) +
    ggtitle(trait_name) +
    ylab(ylab_text) +
    geom_boxplot(width = 0.25, linewidth = 1, alpha = 0) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 2) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_colour_manual(values = pal_trt) +
    theme_classic2(base_size = 20) +
    theme(
      plot.title = element_markdown(hjust = 0.5, face = "bold"),
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 25, face = "bold", color = "black"),
      legend.position = "none"
    )

  if (!is.null(y_limits)) p <- p + coord_cartesian(ylim = y_limits)

  if (nrow(s) > 0) {
    p <- p + stat_pvalue_manual(
      s, label = "p.adj.signif",
      size = 10, bracket.size = 0.8,
      hide.ns = FALSE
    )
  }
  p
}

p_PH   <- supp_panel("PH",   "Plant Height (cm)")
p_KW50 <- supp_panel("KW50", "50-kernel weight (g)")
p_TKN  <- supp_panel("TKN",  "Total Kernel Number")
p_TKW  <- supp_panel("TKW",  "Total Kernel Weight (g)")

supp_marginal <- ggarrange(
  p_PH, p_KW50, p_TKN, p_TKW,
  nrow = 1, ncol = 4, align = "h",
  labels = c("A", "B", "C", "D"),
  font.label = list(size = 30)
)

print(supp_marginal)
```

## Save Approach D contrasts for figure bracket wiring

```{r save-approach-d}
D_out <- D %>%
  mutate(
    Genotype     = ifelse(is.na(Genotype), NA_character_, as.character(Genotype)),
    group1       = "+P",
    group2       = "-P",
    p.adj.signif = rstatix::add_significance(tibble(p.adj = p.adj), "p.adj")$p.adj.signif
  ) %>%
  select(trait, Genotype, contrast_level, group1, group2, statistic, p, p.adj, p.adj.signif)

write.csv(D_out,
  file.path(paths$intermediate, "PSU2022_fig1_bracket_contrasts.csv"),
  row.names = FALSE)

cat("Saved", nrow(D_out), "contrasts to PSU2022_fig1_bracket_contrasts.csv\n")
print(D_out, n = Inf)
```

## Session info

```{r session-info}
sessionInfo()
```
