---
title: "Inv4m Per-Plant Mineral Content (Biomass × Concentration)"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "phosphorus_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

source(here::here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("phosphorus_paper")
```

## Overview

Companion to `PSU2022_ionome.Rmd`. That workbook analyzes mineral
**concentration** (ppm in grain and stover). This one analyzes per-plant
mineral **content** (mg per plant):

- `grain_content_mg  = TKW_g   × conc_seed_ppm   / 1000`
- `stover_content_mg = STWHV_g × conc_stover_ppm / 1000`
- `total_content_mg  = grain_content + stover_content`
- `mineral_HI        = grain_content / total_content`

These quantities are derived at the raw plot level inside
`spatial_correction_for_INV4MXP.Rmd` and fit by SpATS as first-class
traits (same way `HI = TKW / STWHV` is handled). This workbook reads the
already-corrected content and mineral-HI columns and analyzes them;
nothing is re-derived here.

Content separates two processes the concentration-only analysis confounds:
net uptake (total content) and partitioning (mineral HI). This directly
addresses Miguel Piñeros' MP3 ask (processes behind mineral-concentration
changes) and Sawers' L176 question on per-plant flux rather than per-unit
grain.

## Load libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(ggpubr)
library(ggbeeswarm)
library(ggfx)
library(rstatix)
library(ggtext)
library(ggh4x)
```

## Load spatially corrected data

```{r load-data}
ionome <- read.csv(
  file.path(
    paths$intermediate,
    "PSU2022_spatially_corrected_both_treatments_with_ionome.csv"
  )
) %>%
  rename(Genotype = genotype) %>%
  filter(Genotype %in% c("CTRL", "Inv4m")) %>%
  droplevels()

ionome$Treatment <- factor(ionome$Treatment, levels = c("+P", "-P"))
ionome$Genotype  <- factor(ionome$Genotype,  levels = c("CTRL", "Inv4m"))

reliable <- c("Ca", "Fe", "K", "Mg", "Mn", "P", "S", "Zn")

content_vars <- c(
  paste0(reliable, "_grain_mg"),
  paste0(reliable, "_stover_mg"),
  paste0(reliable, "_HI")
)

# Alias: downstream chunks work against `content`. Nothing is derived here;
# all content + HI columns come from the upstream SpATS-corrected CSV.
content <- ionome

cat("Sample sizes:\n")
ionome %>% count(Genotype, Treatment) %>% print()
```

## Unit check: biomass and concentration ranges

Sanity-check the inputs before computing content. Expected:

- `TKW`: tens to hundreds of grams per plant (range roughly 30–300 g).
- `STWHV`: tens to hundreds of grams per plant.
- Concentrations: P/K in the thousands (ppm), Zn/Fe/Mn in the tens,
  Ca in hundreds to low thousands depending on tissue.

```{r unit-check-biomass}
ionome %>%
  select(Genotype, Treatment, TKW, STWHV) %>%
  pivot_longer(c(TKW, STWHV), names_to = "variable", values_to = "value") %>%
  group_by(variable, Genotype, Treatment) %>%
  summarise(
    n      = sum(!is.na(value)),
    min    = round(min(value, na.rm = TRUE), 1),
    median = round(median(value, na.rm = TRUE), 1),
    mean   = round(mean(value, na.rm = TRUE), 1),
    max    = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  )
```

```{r unit-check-concentrations}
conc_cols <- c(
  paste0(reliable, "_seed"),
  paste0(reliable, "_stover")
)

ionome %>%
  select(all_of(conc_cols)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    n      = sum(!is.na(value)),
    min    = round(min(value, na.rm = TRUE), 1),
    median = round(median(value, na.rm = TRUE), 1),
    mean   = round(mean(value, na.rm = TRUE), 1),
    max    = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(variable)
```

## Content summary per mineral × genotype × treatment

```{r content-summary}
content %>%
  select(Genotype, Treatment, all_of(content_vars)) %>%
  pivot_longer(all_of(content_vars), names_to = "variable", values_to = "value") %>%
  separate(variable, into = c("mineral", "pool"), sep = "_", extra = "merge") %>%
  mutate(
    pool = recode(pool,
                  grain_mg = "grain (mg)",
                  stover_mg = "stover (mg)",
                  HI = "mineral HI")
  ) %>%
  group_by(mineral, pool, Genotype, Treatment) %>%
  summarise(
    n      = sum(!is.na(value)),
    median = signif(median(value, na.rm = TRUE), 3),
    mean   = signif(mean(value, na.rm = TRUE), 3),
    sd     = signif(sd(value, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(mineral, pool, Genotype, Treatment) %>%
  print(n = Inf)
```

## Biomass HI vs mineral HI per mineral

Rapid visual sanity check: does mineral HI track biomass HI, or does
it depart (true partitioning signal beyond biomass reallocation)?

```{r biomass-vs-mineral-hi, fig.width=10, fig.height=6}
# Order minerals by median (grain + stover) content per plant, descending:
# highest-flux first so the eye lands on the dominant pools. Sum of two
# SpATS-corrected pools, not the SpATS-corrected total (we don't test total
# anywhere in the paper, so it's not a corrected response).
mineral_order <- content %>%
  select(all_of(paste0(reliable, "_grain_mg")),
         all_of(paste0(reliable, "_stover_mg"))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "mg") %>%
  mutate(mineral = sub("_(grain|stover)_mg$", "", var)) %>%
  group_by(mineral) %>%
  summarise(median_total = median(mg, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(median_total)) %>%
  pull(mineral)

cat("Facet order (median total mg/plant, descending):\n")
print(mineral_order)

hi_long <- content %>%
  select(Genotype, Treatment, biomass_HI_frac,
         all_of(paste0(reliable, "_HI"))) %>%
  pivot_longer(ends_with("_HI"), names_to = "mineral", values_to = "mineral_HI") %>%
  mutate(
    mineral = sub("_HI$", "", mineral),
    mineral = factor(mineral, levels = mineral_order)
  ) %>%
  filter(!is.na(mineral_HI), !is.na(biomass_HI_frac))

ggplot(hi_long, aes(biomass_HI_frac, mineral_HI,
                    color = Treatment, shape = Genotype)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7, size = 1.5) +
  facet_wrap(~ mineral, scales = "free_y") +
  scale_color_manual(values = c("+P" = "gold", "-P" = "purple4")) +
  labs(
    x = "Biomass harvest index  [TKW / (TKW + STWHV)]",
    y = "Mineral HI  [grain content / total content]"
  ) +
  theme_classic(base_size = 12)
```

## Fit linear models across content + HI variables

```{r fit-models}
effects <- lapply(content_vars, function(x) {
  n_obs <- sum(!is.na(content[[x]]))
  if (n_obs < 20) return(NULL)

  model <- tryCatch(
    lm(as.formula(paste(x, "~ Genotype * Treatment")), data = content),
    error = function(e) NULL
  )
  if (is.null(model)) return(NULL)

  out <- coef(summary(model)) %>%
    as.data.frame() %>%
    rownames_to_column("predictor")
  colnames(out)[3] <- "Std.Error"
  colnames(out)[5] <- "p.value"
  out$response <- x
  out %>% select(response, everything())
}) %>%
  bind_rows() %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr")) %>%
  arrange(p.adjust)

cat("Total models fitted:", length(unique(effects$response)), "\n")
cat("Total effects tested:", nrow(effects), "\n")

cat("\n=== Significant effects (FDR < 0.05, no Intercept) ===\n")
effects %>%
  filter(p.adjust < 0.05, !grepl("Intercept", predictor)) %>%
  select(response, predictor, Estimate, p.value, p.adjust) %>%
  as_tibble() %>%
  print(n = Inf)
```

```{r interaction-effects}
interaction_effects <- effects %>%
  filter(grepl(":", predictor)) %>%
  separate(response, into = c("mineral", "pool"),
           sep = "_", extra = "merge", remove = FALSE) %>%
  mutate(
    signif = case_when(
      p.adjust < 0.001 ~ "***",
      p.adjust < 0.01  ~ "**",
      p.adjust < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  arrange(p.adjust)

cat("Genotype x Treatment interactions tested:", nrow(interaction_effects), "\n")
cat("Significant (FDR < 0.05):", sum(interaction_effects$p.adjust < 0.05), "\n")

if (any(interaction_effects$p.adjust < 0.05)) {
  interaction_effects %>%
    filter(p.adjust < 0.05) %>%
    select(mineral, pool, Estimate, p.value, p.adjust, signif) %>%
    print()
}
```

## lm-based pre-selection for figure brackets

Hierarchical convention: the **lm with global FDR** controls the
family-wise error rate across all coefficients (Treatment + Genotype +
interaction over all 48 responses; conservative). Brackets in the
figures illustrate **only the lm-significant effects** as pairwise Welch
*t*-tests with **nominal** p-values. No secondary FDR is applied to the
brackets — the lm's correction is the test, the brackets are the
illustration. Methods describes the hierarchy.

```{r lm-preselection}
# Concentration lm effects come from PSU2022_ionome.Rmd's separate fit
conc_eff_path <- file.path(paths$intermediate,
                           "ionome_linear_model_effects.csv")
if (file.exists(conc_eff_path)) {
  conc_lm <- read.csv(conc_eff_path)
} else {
  warning("Concentration lm effects not found; concentration brackets ",
          "will be empty. Render PSU2022_ionome.Rmd first.")
  conc_lm <- tibble::tibble(predictor = character(),
                            response = character(),
                            p.adjust = numeric())
}

# Pre-selected response variables per panel family
treatment_lm_conc <- conc_lm %>%
  filter(predictor == "Treatment-P", p.adjust < 0.05) %>%
  pull(response)

treatment_lm_content <- effects %>%
  filter(predictor == "Treatment-P", p.adjust < 0.05,
         grepl("_grain_mg$|_stover_mg$", response)) %>%
  pull(response)

treatment_lm_hi <- effects %>%
  filter(predictor == "Treatment-P", p.adjust < 0.05,
         grepl("_HI$", response)) %>%
  pull(response)

genotype_lm_content <- effects %>%
  filter(predictor == "GenotypeInv4m", p.adjust < 0.05,
         grepl("_grain_mg$|_stover_mg$", response)) %>%
  pull(response)

genotype_lm_hi <- effects %>%
  filter(predictor == "GenotypeInv4m", p.adjust < 0.05,
         grepl("_HI$", response)) %>%
  pull(response)

cat("=== lm pre-selection summary ===\n")
cat("Concentration Treatment effects (Panel A):       ",
    length(treatment_lm_conc), "\n")
cat("Content      Treatment effects (Panel B):       ",
    length(treatment_lm_content), "\n")
cat("Mineral HI   Treatment effects (Panel C):       ",
    length(treatment_lm_hi), "\n")
cat("Content      Genotype  effects (Inv4m supp):    ",
    length(genotype_lm_content), "\n")
cat("Mineral HI   Genotype  effects (Inv4m supp):    ",
    length(genotype_lm_hi), "\n")
```

## Prepare long-format content for plotting + t-tests

```{r to-plot-content}
to_plot_content <- content %>%
  select(Genotype, Treatment,
         all_of(paste0(reliable, "_grain_mg")),
         all_of(paste0(reliable, "_stover_mg"))) %>%
  pivot_longer(
    cols = -c(Genotype, Treatment),
    names_to = "mineral_pool",
    values_to = "mg"
  ) %>%
  mutate(
    mineral = sub("_(grain|stover)_mg$", "", mineral_pool),
    pool    = sub("^.+_(grain|stover)_mg$", "\\1", mineral_pool)
  ) %>%
  select(-mineral_pool) %>%
  filter(!is.na(mg)) %>%
  mutate(
    pool     = factor(pool, levels = c("stover", "grain")),
    Genotype = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  )
```

```{r treatment-ttests}
treatment.t.test <- to_plot_content %>%
  group_by(pool, Genotype, mineral) %>%
  rstatix::t_test(mg ~ Treatment) %>%
  rstatix::add_significance("p") %>%
  rstatix::add_y_position(scales = "free_y", step.increase = 0.12) %>%
  mutate(lm_response = paste0(mineral, "_", pool, "_mg")) %>%
  filter(lm_response %in% treatment_lm_content)

cat("\n=== Treatment content t-tests (lm-preselected, nominal p) ===\n")
treatment.t.test %>%
  select(mineral, pool, Genotype, statistic, p, p.signif) %>%
  arrange(p) %>%
  as_tibble() %>%
  print(n = Inf)
```

## Concentration long-format + treatment t-tests (Panel A)

```{r to-plot-conc}
to_plot_conc <- content %>%
  select(Genotype, Treatment,
         all_of(paste0(reliable, "_seed")),
         all_of(paste0(reliable, "_stover"))) %>%
  pivot_longer(
    cols = -c(Genotype, Treatment),
    names_to = "mineral_pool",
    values_to = "ppm"
  ) %>%
  mutate(
    mineral = sub("_(seed|stover)$", "", mineral_pool),
    pool    = sub("^.+_(seed|stover)$", "\\1", mineral_pool)
  ) %>%
  select(-mineral_pool) %>%
  filter(!is.na(ppm)) %>%
  mutate(
    pool      = factor(pool, levels = c("stover", "seed")),
    Genotype  = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype  = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  )

treatment.conc.ttest <- to_plot_conc %>%
  group_by(pool, Genotype, mineral) %>%
  rstatix::t_test(ppm ~ Treatment) %>%
  rstatix::add_significance("p") %>%
  rstatix::add_y_position(scales = "free_y", step.increase = 0.12) %>%
  mutate(lm_response = paste0(mineral, "_", pool)) %>%
  filter(lm_response %in% treatment_lm_conc)
```

## Mineral HI long-format + treatment t-tests (Panel C)

```{r to-plot-hi}
to_plot_hi <- content %>%
  select(Genotype, Treatment, all_of(paste0(reliable, "_HI"))) %>%
  pivot_longer(
    cols = -c(Genotype, Treatment),
    names_to = "mineral",
    values_to = "hi"
  ) %>%
  mutate(
    mineral   = sub("_HI$", "", mineral),
    Genotype  = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype  = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  ) %>%
  filter(!is.na(hi), is.finite(hi))

treatment.hi.ttest <- to_plot_hi %>%
  group_by(Genotype, mineral) %>%
  filter(n() >= 6) %>%
  rstatix::t_test(hi ~ Treatment) %>%
  rstatix::add_significance("p")

if (nrow(treatment.hi.ttest) > 0) {
  treatment.hi.ttest <- treatment.hi.ttest %>%
    group_by(Genotype, mineral) %>%
    group_modify(~ {
      sub_data <- to_plot_hi %>%
        filter(Genotype == .y$Genotype, mineral == .y$mineral)
      .x %>%
        rstatix::add_y_position(
          data = sub_data,
          formula = hi ~ Treatment,
          step.increase = 0.12
        )
    }) %>%
    ungroup() %>%
    filter(!is.na(y.position), is.finite(y.position)) %>%
    mutate(lm_response = paste0(mineral, "_HI")) %>%
    filter(lm_response %in% treatment_lm_hi)
}
```

## Per-mineral content plot function

```{r plot-mineral-content}
pal <- c("+P" = "gold", "-P" = "purple4")

plot_mineral_content <- function(mineral_name) {
  plot_data <- to_plot_content %>%
    filter(mineral == mineral_name)

  ttest_subset <- treatment.t.test %>%
    filter(mineral == mineral_name)

  ggplot(plot_data, aes(x = Treatment, y = mg, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    stat_pvalue_manual(
      ttest_subset,
      label = "p.signif",
      size = 5,
      bracket.size = 0.5,
      hide.ns = TRUE
    ) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    ggh4x::facet_nested(~ pool + Genotype, scales = "free_y") +
    labs(title = mineral_name, x = NULL, y = "Content [mg / plant]") +
    theme_classic(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = 15),
      axis.title.y = element_text(face = "bold", size = 13),
      plot.title = element_text(face = "bold", size = 28, hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = 13),
      strip.background = element_blank(),
      panel.spacing = unit(0.1, "lines"),
      ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.5)
    )
}
```

## Per-mineral concentration plot function (Panel A)

```{r plot-mineral-conc}
plot_mineral_conc <- function(mineral_name, log_scale = FALSE) {
  plot_data <- to_plot_conc %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.conc.ttest %>% filter(mineral == mineral_name)

  p <- ggplot(plot_data, aes(x = Treatment, y = ppm, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_color_manual(values = pal) +
    labs(title = mineral_name, x = NULL, y = "Concentration [mg / kg]")

  if (log_scale) {
    ttest_log <- ttest_subset
    ttest_log$y.position <- log2(ttest_log$y.position)
    p <- p +
      scale_y_continuous(
        trans = "log2",
        breaks = c(5, 10, 25, 50, 100, 200, 400, 800, 1600, 3200, 6400),
        expand = expansion(mult = c(0.05, 0.15))
      ) +
      stat_pvalue_manual(
        ttest_log, label = "p.signif",
        size = 5, bracket.size = 0.5, hide.ns = TRUE
      )
  } else {
    p <- p +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
      stat_pvalue_manual(
        ttest_subset, label = "p.signif",
        size = 5, bracket.size = 0.5, hide.ns = TRUE
      )
  }

  p +
    ggh4x::facet_nested(~ pool + Genotype, scales = "free_y") +
    theme_classic(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = 15),
      axis.title.y = element_text(face = "bold", size = 13),
      plot.title = element_text(face = "bold", size = 28, hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = 13),
      strip.background = element_blank(),
      panel.spacing = unit(0.1, "lines"),
      ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.5)
    )
}
```

## Per-mineral HI plot function (Panel C)

```{r plot-mineral-hi}
plot_mineral_HI <- function(mineral_name) {
  plot_data <- to_plot_hi %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.hi.ttest %>% filter(mineral == mineral_name)

  p <- ggplot(plot_data, aes(x = Treatment, y = hi, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    labs(title = mineral_name, x = NULL, y = "Mineral HI [grain / total]") +
    facet_wrap(~ Genotype) +
    theme_classic(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = 15),
      axis.title.y = element_text(face = "bold", size = 13),
      plot.title = element_text(face = "bold", size = 28, hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = 13),
      strip.background = element_blank()
    )

  if (nrow(ttest_subset) > 0) {
    p <- p + stat_pvalue_manual(
      ttest_subset, label = "p.signif",
      size = 5, bracket.size = 0.5, hide.ns = TRUE
    )
  }
  p
}
```

## Main 3-panel figure (P, Zn, Ca, S): A concentration / B content / C HI

Panels A-row includes log y-axis for Ca (stover/seed concentration range) to
match the concentration workbook.

```{r main-3panel-figure, fig.width=14, fig.height=12}
# Row A: concentrations
rowA_main <- ggarrange(
  plot_mineral_conc("P"),
  plot_mineral_conc("Zn") + ggpubr::rremove("ylab"),
  plot_mineral_conc("Ca", log_scale = TRUE) + ggpubr::rremove("ylab"),
  plot_mineral_conc("S")  + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

# Row B: contents (drop titles since Row A already labels each column)
rowB_main <- ggarrange(
  plot_mineral_content("P")  + labs(title = NULL),
  plot_mineral_content("Zn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("Ca") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("S")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

# Row C: mineral HI (drop titles, same reason)
rowC_main <- ggarrange(
  plot_mineral_HI("P")  + labs(title = NULL),
  plot_mineral_HI("Zn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("Ca") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("S")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

p_main_3panel <- ggarrange(
  rowA_main, rowB_main, rowC_main,
  ncol = 1, nrow = 3,
  heights = c(1, 1, 1),
  labels = c("A", "B", "C"),
  font.label = list(size = 30)
)

print(p_main_3panel)
```

## Supplementary 3-panel figure (Mg, Mn, K, Fe)

```{r supp-3panel-figure, fig.width=14, fig.height=12}
rowA_supp <- ggarrange(
  plot_mineral_conc("Mg"),
  plot_mineral_conc("Mn", log_scale = TRUE) + ggpubr::rremove("ylab"),
  plot_mineral_conc("K")  + ggpubr::rremove("ylab"),
  plot_mineral_conc("Fe") + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowB_supp <- ggarrange(
  plot_mineral_content("Mg") + labs(title = NULL),
  plot_mineral_content("Mn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("K")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("Fe") + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowC_supp <- ggarrange(
  plot_mineral_HI("Mg") + labs(title = NULL),
  plot_mineral_HI("Mn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("K")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("Fe") + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

p_supp_3panel <- ggarrange(
  rowA_supp, rowB_supp, rowC_supp,
  ncol = 1, nrow = 3,
  heights = c(1, 1, 1),
  labels = c("A", "B", "C"),
  font.label = list(size = 30)
)

print(p_supp_3panel)
```


## Inv4m-effects supplementary figure

The lm Genotype(Inv4m) effect is estimated pooled across P treatments
(interaction lm with non-significant GxE). Visualization pools too: one
Inv4m vs CTRL contrast per (mineral, pool) for content, and per mineral
for HI. Brackets come from Welch *t*-tests on the pooled data at
nominal *p*, illustrating the lm-certified effects.

```{r genotype-content-ttests}
genotype.content.ttest <- to_plot_content %>%
  group_by(pool, mineral) %>%
  filter(n() >= 4, var(mg, na.rm = TRUE) > 0) %>%
  rstatix::t_test(mg ~ Genotype) %>%
  rstatix::add_significance("p") %>%
  rstatix::add_y_position(scales = "free_y", step.increase = 0.12) %>%
  mutate(lm_response = paste0(mineral, "_", pool, "_mg")) %>%
  filter(lm_response %in% genotype_lm_content)

genotype.hi.ttest <- to_plot_hi %>%
  group_by(mineral) %>%
  filter(n() >= 4, var(hi, na.rm = TRUE) > 0) %>%
  rstatix::t_test(hi ~ Genotype) %>%
  rstatix::add_significance("p")

if (nrow(genotype.hi.ttest) > 0) {
  genotype.hi.ttest <- genotype.hi.ttest %>%
    group_by(mineral) %>%
    group_modify(~ {
      sub_data <- to_plot_hi %>% filter(mineral == .y$mineral)
      .x %>%
        rstatix::add_y_position(
          data = sub_data, formula = hi ~ Genotype,
          step.increase = 0.12
        )
    }) %>%
    ungroup() %>%
    filter(!is.na(y.position), is.finite(y.position)) %>%
    mutate(lm_response = paste0(mineral, "_HI")) %>%
    filter(lm_response %in% genotype_lm_hi)
}

cat("\n=== Inv4m content t-tests (pooled across treatment, lm-preselected) ===\n")
genotype.content.ttest %>%
  select(mineral, pool, statistic, p, p.signif) %>%
  arrange(p) %>% as_tibble() %>% print(n = Inf)

cat("\n=== Inv4m HI t-tests (pooled across treatment, lm-preselected) ===\n")
genotype.hi.ttest %>%
  select(mineral, statistic, p, p.signif) %>%
  arrange(p) %>% as_tibble() %>% print(n = Inf)
```

```{r plot-genotype-panels}
pal_geno <- c("CTRL" = "gold", "Inv4m" = "purple4")

make_content_panel <- function(mineral_name) {
  d <- to_plot_content %>%
    filter(mineral == mineral_name) %>%
    mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
           Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))
  s <- genotype.content.ttest %>%
    filter(mineral == mineral_name) %>%
    mutate(group1 = gsub("<i>|</i>", "", group1),
           group2 = gsub("<i>|</i>", "", group2))

  p <- ggplot(d, aes(x = Genotype, y = mg, colour = Genotype)) +
    ggtitle(mineral_name) +
    ylab("Content [mg / plant]") +
    geom_boxplot(width = 0.25, linewidth = 1, alpha = 0) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 2) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_colour_manual(values = pal_geno) +
    scale_x_discrete(labels = c("CTRL" = "CTRL", "Inv4m" = "<i>Inv4m</i>")) +
    facet_wrap(~ pool, scales = "free_y") +
    theme_classic(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 22, hjust = 0.5),
      axis.title.y = element_text(face = "bold", size = 13),
      axis.title.x = element_blank(),
      axis.text.x = element_markdown(size = 14, face = "bold", color = "black"),
      strip.background = element_blank(),
      strip.text = element_markdown(face = "bold", size = 14),
      legend.position = "none"
    )

  if (nrow(s) > 0) {
    p <- p + stat_pvalue_manual(
      s, label = "p.signif",
      size = 7, bracket.size = 0.6, hide.ns = FALSE
    )
  }
  p
}

make_hi_panel <- function(mineral_name) {
  d <- to_plot_hi %>%
    filter(mineral == mineral_name) %>%
    mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
           Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))
  s <- genotype.hi.ttest %>%
    filter(mineral == mineral_name) %>%
    mutate(group1 = gsub("<i>|</i>", "", group1),
           group2 = gsub("<i>|</i>", "", group2))

  p <- ggplot(d, aes(x = Genotype, y = hi, colour = Genotype)) +
    ggtitle(mineral_name) +
    ylab("Mineral HI [grain / total]") +
    geom_boxplot(width = 0.25, linewidth = 1, alpha = 0) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 2) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_colour_manual(values = pal_geno) +
    scale_x_discrete(labels = c("CTRL" = "CTRL", "Inv4m" = "<i>Inv4m</i>")) +
    theme_classic(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 22, hjust = 0.5),
      axis.title.y = element_text(face = "bold", size = 13),
      axis.title.x = element_blank(),
      axis.text.x = element_markdown(size = 14, face = "bold", color = "black"),
      strip.background = element_blank(),
      strip.text = element_markdown(face = "bold", size = 14),
      legend.position = "none"
    )

  if (nrow(s) > 0) {
    p <- p + stat_pvalue_manual(
      s, label = "p.signif",
      size = 7, bracket.size = 0.6, hide.ns = FALSE
    )
  }
  p
}
```

```{r inv4m-figure, fig.width=16, fig.height=10}
# Minerals with any lm-significant Genotype effect (content or HI).
inv4m_minerals <- c("Ca", "Zn", "Fe", "K")

row_content <- ggarrange(
  plotlist = lapply(inv4m_minerals, function(m) {
    p <- make_content_panel(m)
    if (m != inv4m_minerals[1]) p <- p + ggpubr::rremove("ylab")
    p
  }),
  ncol = 4, nrow = 1, align = "h"
)

row_hi <- ggarrange(
  plotlist = lapply(inv4m_minerals, function(m) {
    p <- make_hi_panel(m)
    if (m != inv4m_minerals[1]) p <- p + ggpubr::rremove("ylab")
    p
  }),
  ncol = 4, nrow = 1, align = "h"
)

p_inv4m_effects <- ggarrange(
  row_content, row_hi,
  ncol = 1, nrow = 2,
  heights = c(1, 1),
  labels = c("A", "B"),
  font.label = list(size = 30)
)

print(p_inv4m_effects)
```

## Save results

Per-plant corrected content is already in the upstream
`PSU2022_spatially_corrected_both_treatments_with_ionome.csv`, so this
chunk saves only the model outputs, t-tests, and figures unique to this
analysis.

```{r save-derived}
write.csv(
  effects,
  file.path(paths$intermediate, "ionome_content_linear_model_effects.csv"),
  row.names = FALSE
)

write.csv(
  interaction_effects,
  file.path(paths$intermediate, "ionome_content_interaction_effects.csv"),
  row.names = FALSE
)

# Drop list columns so add_y_position() output serializes cleanly.
write_ttest <- function(x, name) {
  if (nrow(x) == 0) return(invisible(NULL))
  write.csv(
    x %>% select(where(~ !is.list(.))),
    file.path(paths$intermediate, name),
    row.names = FALSE
  )
}

write_ttest(treatment.t.test,       "ionome_content_treatment_ttests.csv")
write_ttest(treatment.conc.ttest,   "ionome_content_conc_treatment_ttests.csv")
write_ttest(treatment.hi.ttest,     "ionome_content_hi_treatment_ttests.csv")
write_ttest(genotype.content.ttest, "ionome_content_genotype_ttests.csv")
write_ttest(genotype.hi.ttest,      "ionome_content_hi_genotype_ttests.csv")

ggsave(
  file.path(paths$figures, "ionome_3panel_main_figure.pdf"),
  plot = p_main_3panel, width = 14, height = 12, units = "in"
)
ggsave(
  file.path(paths$figures, "ionome_3panel_main_figure.png"),
  plot = p_main_3panel, width = 14, height = 12, units = "in", dpi = 300
)
ggsave(
  file.path(paths$figures, "ionome_3panel_supplementary_figure.pdf"),
  plot = p_supp_3panel, width = 14, height = 12, units = "in"
)
ggsave(
  file.path(paths$figures, "ionome_3panel_supplementary_figure.png"),
  plot = p_supp_3panel, width = 14, height = 12, units = "in", dpi = 300
)
ggsave(
  file.path(paths$figures, "ionome_inv4m_effects_figure.pdf"),
  plot = p_inv4m_effects, width = 16, height = 10, units = "in"
)
ggsave(
  file.path(paths$figures, "ionome_inv4m_effects_figure.png"),
  plot = p_inv4m_effects, width = 16, height = 10, units = "in", dpi = 300
)
```

## Session info

```{r session-info}
sessionInfo()
```