---
title: "Inv4m Per-Plant Mineral Content (lm contrasts via emmeans)"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "phosphorus_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

source(here::here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("phosphorus_paper")
```

## Overview

Variant of `PSU2022_ionome_content.Rmd` where **figure brackets come from
linear-model contrasts** (`emmeans::emmeans` + `pairs`) instead of marginal
Welch *t*-tests. The brackets visualise the same kind of t-statistic that
underlies each lm coefficient — same data, same model — so the asterisks
in the figures match what the lm effect tables report, with full power
from the joint fit (Treatment absorbed as covariate, full residual df).

Three convention notes:

- For Treatment +P vs −P brackets in the main figure: the contrast is
  `pairs(emmeans(m, ~ Treatment | Genotype))`. CTRL and Inv4m facets get
  facet-specific p-values (handles the S_stover concentration × Genotype
  interaction correctly, for example).
- For Inv4m vs CTRL brackets in the supp figure: pooled-across-treatment
  marginal contrast `pairs(emmeans(m, ~ Genotype))`.
- FDR is applied within each metric (concentration / content / HI).

## Load libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(ggpubr)
library(ggbeeswarm)
library(ggfx)
library(rstatix)
library(ggtext)
library(ggh4x)
library(emmeans)
```

## Load spatially corrected data

```{r load-data}
ionome <- read.csv(
  file.path(
    paths$intermediate,
    "PSU2022_spatially_corrected_both_treatments_with_ionome.csv"
  )
) %>%
  rename(Genotype = genotype) %>%
  filter(Genotype %in% c("CTRL", "Inv4m")) %>%
  droplevels()

ionome$Treatment <- factor(ionome$Treatment, levels = c("+P", "-P"))
ionome$Genotype  <- factor(ionome$Genotype,  levels = c("CTRL", "Inv4m"))

reliable <- c("Ca", "Fe", "K", "Mg", "Mn", "P", "S", "Zn")

content_vars <- c(
  paste0(reliable, "_grain_mg"),
  paste0(reliable, "_stover_mg"),
  paste0(reliable, "_HI")
)

conc_vars <- c(
  paste0(reliable, "_grain"),
  paste0(reliable, "_stover")
)

ratio_vars <- paste0(reliable, "_ratio")

content <- ionome

cat("Sample sizes:\n")
ionome %>% count(Genotype, Treatment) %>% print()
```

## Long-format inputs (used by both contrasts and plots)

```{r long-format}
to_plot_content <- content %>%
  select(Genotype, Treatment,
         all_of(paste0(reliable, "_grain_mg")),
         all_of(paste0(reliable, "_stover_mg"))) %>%
  pivot_longer(cols = -c(Genotype, Treatment),
               names_to = "mineral_pool", values_to = "mg") %>%
  mutate(
    mineral = sub("_(grain|stover)_mg$", "", mineral_pool),
    pool    = sub("^.+_(grain|stover)_mg$", "\\1", mineral_pool)
  ) %>%
  select(-mineral_pool) %>%
  filter(!is.na(mg)) %>%
  mutate(
    pool      = factor(pool, levels = c("stover", "grain")),
    Genotype  = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype  = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  )

to_plot_conc <- content %>%
  select(Genotype, Treatment,
         all_of(paste0(reliable, "_grain")),
         all_of(paste0(reliable, "_stover"))) %>%
  pivot_longer(cols = -c(Genotype, Treatment),
               names_to = "mineral_pool", values_to = "ppm") %>%
  mutate(
    mineral = sub("_(grain|stover)$", "", mineral_pool),
    pool    = sub("^.+_(grain|stover)$", "\\1", mineral_pool)
  ) %>%
  select(-mineral_pool) %>%
  filter(!is.na(ppm)) %>%
  mutate(
    pool      = factor(pool, levels = c("stover", "grain")),
    Genotype  = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype  = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  )

to_plot_hi <- content %>%
  select(Genotype, Treatment, all_of(paste0(reliable, "_HI"))) %>%
  pivot_longer(cols = -c(Genotype, Treatment),
               names_to = "mineral", values_to = "hi") %>%
  mutate(
    mineral   = sub("_HI$", "", mineral),
    Genotype  = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype  = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  ) %>%
  filter(!is.na(hi), is.finite(hi))

to_plot_ratio <- content %>%
  select(Genotype, Treatment, all_of(paste0(reliable, "_ratio"))) %>%
  pivot_longer(cols = -c(Genotype, Treatment),
               names_to = "mineral", values_to = "ratio") %>%
  mutate(
    mineral   = sub("_ratio$", "", mineral),
    Genotype  = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype  = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  ) %>%
  filter(!is.na(ratio), is.finite(ratio))
```

## Fit lms per response (kept for emmeans)

```{r fit-lms}
# Fit lm per response and keep model objects. Same model spec everywhere.
# Results are cached to RDS; delete the RDS to force a refit.
fit_per_response <- function(responses, data) {
  models <- lapply(responses, function(resp) {
    if (!resp %in% colnames(data)) return(NULL)
    if (sum(!is.na(data[[resp]])) < 20) return(NULL)
    tryCatch(
      lm(as.formula(paste(resp, "~ Genotype * Treatment")), data = data),
      error = function(e) NULL
    )
  })
  names(models) <- responses
  models[!sapply(models, is.null)]
}

models_rds <- file.path(paths$intermediate, "ionome_content_models.rds")
if (file.exists(models_rds)) {
  cat("Loading cached lm models from", models_rds, "\n")
  .ml <- readRDS(models_rds)
  models_conc    <- .ml$conc
  models_content <- .ml$content
  models_hi      <- .ml$hi
  models_ratio   <- .ml$ratio
  rm(.ml)
} else {
  models_conc    <- fit_per_response(conc_vars, content)
  models_content <- fit_per_response(paste0(reliable, c(rep("_grain_mg", 8),
                                                       rep("_stover_mg", 8))),
                                     content)
  models_hi      <- fit_per_response(paste0(reliable, "_HI"), content)
  models_ratio   <- fit_per_response(ratio_vars, content)
  saveRDS(list(conc = models_conc, content = models_content,
               hi = models_hi, ratio = models_ratio), models_rds)
  cat("Saved lm models to", models_rds, "\n")
}

cat("Fitted lms — conc:", length(models_conc),
    "content:", length(models_content),
    "HI:", length(models_hi),
    "ratio:", length(models_ratio), "\n")
```

## Effect table (lm coefficients, FDR across all)

```{r effect-table}
extract_coefs <- function(models, family) {
  lapply(names(models), function(resp) {
    m <- models[[resp]]
    out <- coef(summary(m)) %>%
      as.data.frame() %>%
      rownames_to_column("predictor")
    colnames(out)[3] <- "Std.Error"
    colnames(out)[5] <- "p.value"
    out$response <- resp
    out$family   <- family
    out %>% select(family, response, everything())
  }) %>% bind_rows()
}

effects <- bind_rows(
  extract_coefs(models_conc,    "conc"),
  extract_coefs(models_content, "content"),
  extract_coefs(models_hi,      "hi"),
  extract_coefs(models_ratio,   "ratio")
) %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr")) %>%
  arrange(p.adjust)

cat("Total models fitted:", length(unique(effects$response)), "\n")
cat("Total effects tested:", nrow(effects), "\n")

cat("\n=== Significant effects (FDR < 0.05, no Intercept) ===\n")
effects %>%
  filter(p.adjust < 0.05, !grepl("Intercept", predictor)) %>%
  select(family, response, predictor, Estimate, p.value, p.adjust) %>%
  as_tibble() %>%
  print(n = Inf)
```

## emmeans contrast helpers

```{r emmeans-helpers}
# Treatment-within-Genotype contrast (CTRL facet bracket = -P vs +P at CTRL;
# Inv4m facet bracket = -P vs +P at Inv4m). Returns one row per (response,
# Genotype).
contrasts_treatment_per_geno <- function(models) {
  lapply(names(models), function(resp) {
    m <- models[[resp]]
    emm <- emmeans(m, ~ Treatment | Genotype)
    pairs(emm, reverse = TRUE) %>%   # (-P) - (+P)
      as_tibble() %>%
      mutate(response = resp)
  }) %>% bind_rows()
}

# Pooled Genotype contrast (Inv4m vs CTRL averaged over Treatment). Returns
# one row per response.
contrasts_genotype_pooled <- function(models) {
  lapply(names(models), function(resp) {
    m <- models[[resp]]
    emm <- emmeans(m, ~ Genotype)
    pairs(emm, reverse = TRUE) %>%   # Inv4m - CTRL
      as_tibble() %>%
      mutate(response = resp)
  }) %>% bind_rows()
}

# Marginal Treatment contrast (pooled across Genotype). Returns one row per
# response. Use only when Genotype x Treatment interaction is ns.
contrasts_treatment_marginal <- function(models) {
  lapply(names(models), function(resp) {
    m <- models[[resp]]
    emm <- emmeans(m, ~ Treatment)
    pairs(emm, reverse = TRUE) %>%   # (-P) - (+P)
      as_tibble() %>%
      mutate(response = resp)
  }) %>% bind_rows()
}

# Format an emmeans contrast tibble to look like rstatix t_test output for
# stat_pvalue_manual. The function strips the contrast string and assigns
# group1/group2 explicitly. Adds FDR within the tibble (one metric).
format_treatment_contrast <- function(ct_tbl, response_to_mineral_pool, value_var) {
  ct_tbl %>%
    response_to_mineral_pool() %>%
    rename(p = p.value, statistic = t.ratio) %>%
    mutate(
      .y. = value_var,
      group1 = "+P",
      group2 = "-P",
      Genotype = gsub("Inv4m", "<i>Inv4m</i>", as.character(Genotype)),
      Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))
    ) %>%
    rstatix::adjust_pvalue(method = "fdr") %>%
    rstatix::add_significance("p.adj")
}

format_genotype_contrast <- function(ct_tbl, response_to_mineral_pool, value_var) {
  ct_tbl %>%
    response_to_mineral_pool() %>%
    rename(p = p.value, statistic = t.ratio) %>%
    mutate(
      .y. = value_var,
      group1 = "CTRL",
      group2 = "Inv4m"
    ) %>%
    rstatix::adjust_pvalue(method = "fdr") %>%
    rstatix::add_significance("p.adj")
}
```

## Treatment contrasts → t-test-style tibbles for the main figure

```{r treatment-contrasts}
# Helpers to split response name into mineral + pool
split_content <- function(df) {
  df %>% mutate(
    mineral = sub("_(grain|stover)_mg$", "", response),
    pool    = sub("^.+_(grain|stover)_mg$", "\\1", response),
    pool    = factor(pool, levels = c("stover", "grain"))
  )
}
split_conc <- function(df) {
  df %>% mutate(
    mineral = sub("_(grain|stover)$", "", response),
    pool    = sub("^.+_(grain|stover)$", "\\1", response),
    pool    = factor(pool, levels = c("stover", "grain"))
  )
}
split_hi <- function(df) {
  df %>% mutate(
    mineral = sub("_HI$", "", response)
  )
}
split_ratio <- function(df) {
  df %>% mutate(
    mineral = sub("_ratio$", "", response)
  )
}

# Cache raw emmeans outputs (before formatting) to avoid refitting on every
# render. Delete the RDS to force a refresh when models change.
emmeans_rds <- file.path(paths$intermediate, "ionome_content_emmeans.rds")
if (file.exists(emmeans_rds)) {
  cat("Loading cached emmeans contrasts from", emmeans_rds, "\n")
  .em <- readRDS(emmeans_rds)
  ct_conc    <- .em$ct_conc
  ct_content <- .em$ct_content
  ct_hi      <- .em$ct_hi
  ct_ratio   <- .em$ct_ratio
  rm(.em)
} else {
  ct_conc    <- contrasts_treatment_per_geno(models_conc)
  ct_content <- contrasts_treatment_per_geno(models_content)
  ct_hi      <- contrasts_treatment_per_geno(models_hi)
  ct_ratio   <- contrasts_treatment_per_geno(models_ratio)
  saveRDS(list(ct_conc = ct_conc, ct_content = ct_content,
               ct_hi = ct_hi, ct_ratio = ct_ratio), emmeans_rds)
  cat("Saved emmeans contrasts to", emmeans_rds, "\n")
}

# Unified FDR pool across every bracket drawn on the 4-panel figure:
#   - Panel A conc (per-Genotype)
#   - Panel B ratio (per-Genotype)
#   - Panel C content (per-Genotype)
#   - Panel D HI (per-Genotype)
raw_conc <- ct_conc %>%
  split_conc() %>%
  rename(p = p.value, statistic = t.ratio) %>%
  mutate(
    .y. = "ppm", group1 = "+P", group2 = "-P", metric = "conc",
    Genotype = gsub("Inv4m", "<i>Inv4m</i>", as.character(Genotype)),
    Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))
  )

raw_ratio <- ct_ratio %>%
  split_ratio() %>%
  rename(p = p.value, statistic = t.ratio) %>%
  mutate(
    .y. = "ratio", group1 = "+P", group2 = "-P", metric = "ratio",
    Genotype = gsub("Inv4m", "<i>Inv4m</i>", as.character(Genotype)),
    Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))
  )

raw_content <- ct_content %>%
  split_content() %>%
  rename(p = p.value, statistic = t.ratio) %>%
  mutate(
    .y. = "mg", group1 = "+P", group2 = "-P", metric = "content",
    Genotype = gsub("Inv4m", "<i>Inv4m</i>", as.character(Genotype)),
    Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))
  )

raw_hi <- ct_hi %>%
  split_hi() %>%
  rename(p = p.value, statistic = t.ratio) %>%
  mutate(
    .y. = "hi", group1 = "+P", group2 = "-P", metric = "hi",
    Genotype = gsub("Inv4m", "<i>Inv4m</i>", as.character(Genotype)),
    Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))
  )

unified_bracket_pool <- bind_rows(raw_conc, raw_ratio, raw_content,
                                  raw_hi) %>%
  mutate(p.adj = p.adjust(p, method = "fdr")) %>%
  rstatix::add_significance("p.adj")

cat("\n=== Unified FDR pool (n =", nrow(unified_bracket_pool),
    " | conc =", sum(unified_bracket_pool$metric == "conc"),
    "| ratio =", sum(unified_bracket_pool$metric == "ratio"),
    "| content =", sum(unified_bracket_pool$metric == "content"),
    "| hi =", sum(unified_bracket_pool$metric == "hi"),
    ") ===\n")

treatment.conc.ttest <- unified_bracket_pool %>%
  filter(metric == "conc") %>%
  mutate(Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
         pool = factor(pool, levels = c("stover", "grain"))) %>%
  group_by(pool, mineral, Genotype) %>%
  group_modify(~ {
    sub <- to_plot_conc %>%
      filter(pool == .y$pool, mineral == .y$mineral, Genotype == .y$Genotype)
    .x %>%
      rstatix::add_y_position(
        data = sub, formula = ppm ~ Treatment,
        step.increase = 0.12
      )
  }) %>%
  ungroup() %>%
  filter(!is.na(y.position), is.finite(y.position))

treatment.ratio.ttest <- unified_bracket_pool %>%
  filter(metric == "ratio") %>%
  mutate(Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))) %>%
  group_by(mineral, Genotype) %>%
  group_modify(~ {
    sub <- to_plot_ratio %>%
      filter(mineral == .y$mineral, Genotype == .y$Genotype)
    .x %>%
      rstatix::add_y_position(
        data = sub, formula = ratio ~ Treatment,
        step.increase = 0.12
      )
  }) %>%
  ungroup() %>%
  filter(!is.na(y.position), is.finite(y.position))

treatment.t.test <- unified_bracket_pool %>%
  filter(metric == "content") %>%
  mutate(Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
         pool = factor(pool, levels = c("stover", "grain"))) %>%
  group_by(pool, mineral, Genotype) %>%
  group_modify(~ {
    sub <- to_plot_content %>%
      filter(pool == .y$pool, mineral == .y$mineral, Genotype == .y$Genotype)
    .x %>%
      rstatix::add_y_position(
        data = sub, formula = mg ~ Treatment,
        step.increase = 0.20
      )
  }) %>%
  ungroup() %>%
  filter(!is.na(y.position), is.finite(y.position)) %>%
  mutate(y.position = case_when(
    mineral == "Ca" & pool == "grain" ~ y.position + 35,
    mineral == "Mn"                   ~ y.position + 0.2,
    mineral == "K"                    ~ y.position + 100,
    TRUE                              ~ y.position
  ))

treatment.hi.ttest <- unified_bracket_pool %>%
  filter(metric == "hi") %>%
  mutate(Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>"))) %>%
  group_by(mineral, Genotype) %>%
  group_modify(~ {
    sub <- to_plot_hi %>%
      filter(mineral == .y$mineral, Genotype == .y$Genotype)
    .x %>%
      rstatix::add_y_position(
        data = sub, formula = hi ~ Treatment,
        step.increase = 0.12
      )
  }) %>%
  ungroup() %>%
  filter(!is.na(y.position), is.finite(y.position))

cat("\n=== Treatment concentration contrasts (unified FDR, p.adj < 0.05) ===\n")
treatment.conc.ttest %>% filter(p.adj < 0.05) %>%
  select(mineral, pool, Genotype, statistic, p, p.adj, p.adj.signif) %>%
  arrange(p.adj) %>% print(n = Inf)

cat("\n=== Treatment ratio contrasts (unified FDR, p.adj < 0.05) ===\n")
treatment.ratio.ttest %>% filter(p.adj < 0.05) %>%
  select(mineral, Genotype, statistic, p, p.adj, p.adj.signif) %>%
  arrange(p.adj) %>% print(n = Inf)

cat("\n=== Treatment content contrasts (unified FDR, p.adj < 0.05) ===\n")
treatment.t.test %>% filter(p.adj < 0.05) %>%
  select(mineral, pool, Genotype, statistic, p, p.adj, p.adj.signif) %>%
  arrange(p.adj) %>% print(n = Inf)

cat("\n=== Treatment HI contrasts (unified FDR, p.adj < 0.05) ===\n")
treatment.hi.ttest %>% filter(p.adj < 0.05) %>%
  select(mineral, Genotype, statistic, p, p.adj, p.adj.signif) %>%
  arrange(p.adj) %>% print(n = Inf)
```

## Genotype contrasts (pooled) → unified-FDR pool across conc / content / HI / ratio

```{r genotype-contrasts}
# Cache raw pooled-Genotype emmeans outputs (including ratio).
geno_emmeans_rds <- file.path(paths$intermediate,
                              "ionome_content_geno_emmeans.rds")
needs_recompute <- TRUE
if (file.exists(geno_emmeans_rds)) {
  .gm <- readRDS(geno_emmeans_rds)
  if (all(c("gp_conc", "gp_content", "gp_hi", "gp_ratio") %in% names(.gm))) {
    gp_conc    <- .gm$gp_conc
    gp_content <- .gm$gp_content
    gp_hi      <- .gm$gp_hi
    gp_ratio   <- .gm$gp_ratio
    needs_recompute <- FALSE
    cat("Loaded cached pooled-Genotype contrasts from", geno_emmeans_rds, "\n")
  }
  rm(.gm)
}
if (needs_recompute) {
  gp_conc    <- contrasts_genotype_pooled(models_conc)
  gp_content <- contrasts_genotype_pooled(models_content)
  gp_hi      <- contrasts_genotype_pooled(models_hi)
  gp_ratio   <- contrasts_genotype_pooled(models_ratio)
  saveRDS(list(gp_conc = gp_conc, gp_content = gp_content,
               gp_hi = gp_hi, gp_ratio = gp_ratio), geno_emmeans_rds)
  cat("Saved pooled-Genotype contrasts to", geno_emmeans_rds, "\n")
}

# Unified pooled-Genotype FDR across all four metrics. Mirrors the
# unified-bracket convention used in the main 4-panel figure.
format_raw_geno <- function(ct_tbl, response_split, value_var, metric_lbl) {
  ct_tbl %>%
    response_split() %>%
    rename(p = p.value, statistic = t.ratio) %>%
    mutate(.y. = value_var, group1 = "CTRL", group2 = "Inv4m",
           metric = metric_lbl)
}

unified_geno_pool <- bind_rows(
  format_raw_geno(gp_conc,    split_conc,    "ppm",   "conc"),
  format_raw_geno(gp_content, split_content, "mg",    "content"),
  format_raw_geno(gp_hi,      split_hi,      "hi",    "hi"),
  format_raw_geno(gp_ratio,   split_ratio,   "ratio", "ratio")
) %>%
  rstatix::adjust_pvalue(method = "fdr", p.col = "p", output.col = "p.adj") %>%
  rstatix::add_significance("p.adj")

cat(sprintf("Unified pooled-Genotype FDR: %d contrasts, %d sig at p.adj < 0.05\n",
            nrow(unified_geno_pool), sum(unified_geno_pool$p.adj < 0.05)))

# Clean copies of the long-format data with non-markdown Genotype labels
# (needed for rstatix::add_y_position which matches on Genotype values).
to_plot_conc_clean <- to_plot_conc %>%
  mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
         Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))
to_plot_content_clean <- to_plot_content %>%
  mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
         Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))
to_plot_hi_clean <- to_plot_hi %>%
  mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
         Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))
to_plot_ratio_clean <- to_plot_ratio %>%
  mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
         Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))

add_y_pos_pool <- function(tbl, data, formula_y) {
  tbl %>%
    group_by(pool, mineral) %>%
    group_modify(~ {
      sub <- data %>% filter(pool == .y$pool, mineral == .y$mineral)
      .x %>% rstatix::add_y_position(
        data = sub, formula = formula_y, step.increase = 0.12
      )
    }) %>%
    ungroup() %>%
    filter(!is.na(y.position), is.finite(y.position))
}

add_y_pos_nopool <- function(tbl, data, formula_y) {
  tbl %>%
    group_by(mineral) %>%
    group_modify(~ {
      sub <- data %>% filter(mineral == .y$mineral)
      .x %>% rstatix::add_y_position(
        data = sub, formula = formula_y, step.increase = 0.12
      )
    }) %>%
    ungroup() %>%
    filter(!is.na(y.position), is.finite(y.position))
}

genotype.conc.ttest <- unified_geno_pool %>%
  filter(metric == "conc") %>%
  add_y_pos_pool(to_plot_conc_clean, ppm ~ Genotype)

genotype.content.ttest <- unified_geno_pool %>%
  filter(metric == "content") %>%
  add_y_pos_pool(to_plot_content_clean, mg ~ Genotype)

genotype.hi.ttest <- unified_geno_pool %>%
  filter(metric == "hi") %>%
  add_y_pos_nopool(to_plot_hi_clean, hi ~ Genotype)

genotype.ratio.ttest <- unified_geno_pool %>%
  filter(metric == "ratio") %>%
  add_y_pos_nopool(to_plot_ratio_clean, ratio ~ Genotype)

cat("\n=== Inv4m contrasts (unified pooled-Genotype FDR, p.adj < 0.05) ===\n")
unified_geno_pool %>%
  filter(p.adj < 0.05) %>%
  select(metric, mineral, any_of("pool"), statistic, p, p.adj, p.adj.signif) %>%
  arrange(p.adj) %>%
  print(n = Inf)
```

## Per-mineral plot functions (unchanged from base notebook)

```{r plot-functions}
pal <- c("+P" = "gold", "-P" = "purple4")

plot_mineral_content <- function(mineral_name, base_size = 18) {
  plot_data <- to_plot_content %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.t.test %>% filter(mineral == mineral_name)

  ggplot(plot_data, aes(x = Treatment, y = mg, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    stat_pvalue_manual(ttest_subset, label = "p.adj.signif",
                       size = 8, bracket.size = 1, hide.ns = TRUE) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
    ggh4x::facet_nested(~ pool + Genotype, scales = "free_y") +
    labs(title = mineral_name, x = NULL, y = "Content [mg / plant]") +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = rel(1.0)),
      axis.title.y = element_text(face = "bold", size = rel(0.9)),
      plot.title = element_text(face = "bold", size = rel(1.55), hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = rel(0.9)),
      strip.background = element_blank(),
      panel.spacing = unit(0.1, "lines"),
      ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.5)
    )
}

plot_mineral_conc <- function(mineral_name, log_scale = FALSE, base_size = 18) {
  plot_data <- to_plot_conc %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.conc.ttest %>% filter(mineral == mineral_name)

  p <- ggplot(plot_data, aes(x = Treatment, y = ppm, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_color_manual(values = pal) +
    labs(title = mineral_name, x = NULL, y = "Concentration [mg / kg]")

  if (log_scale) {
    ttest_log <- ttest_subset
    ttest_log$y.position <- log2(ttest_log$y.position)
    p <- p +
      scale_y_continuous(trans = "log2",
        breaks = c(5, 10, 25, 50, 100, 200, 400, 800, 1600, 3200, 6400),
        expand = expansion(mult = c(0.05, 0.15))) +
      stat_pvalue_manual(ttest_log, label = "p.adj.signif",
                         size = 8, bracket.size = 1, hide.ns = TRUE)
  } else {
    p <- p + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
      stat_pvalue_manual(ttest_subset, label = "p.adj.signif",
                         size = 8, bracket.size = 1, hide.ns = TRUE)
  }

  p + ggh4x::facet_nested(~ pool + Genotype, scales = "free_y") +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = rel(1.0)),
      axis.title.y = element_text(face = "bold", size = rel(0.9)),
      plot.title = element_text(face = "bold", size = rel(1.55), hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = rel(0.9)),
      strip.background = element_blank(),
      panel.spacing = unit(0.1, "lines"),
      ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.5)
    )
}

plot_mineral_ratio <- function(mineral_name, base_size = 18) {
  plot_data <- to_plot_ratio %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.ratio.ttest %>% filter(mineral == mineral_name)

  ggplot(plot_data, aes(x = Treatment, y = ratio, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    stat_pvalue_manual(ttest_subset, label = "p.adj.signif",
                       size = 8, bracket.size = 1, hide.ns = TRUE) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    facet_wrap(~ Genotype, nrow = 1) +
    labs(title = mineral_name, x = NULL, y = "Grain / stover ratio") +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = rel(1.0)),
      axis.title.y = element_text(face = "bold", size = rel(0.9)),
      plot.title = element_text(face = "bold", size = rel(1.55), hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = rel(0.9)),
      strip.background = element_blank(),
      panel.spacing = unit(0.1, "lines")
    )
}

plot_mineral_HI_marginal <- function(mineral_name, base_size = 18) {
  plot_data <- to_plot_hi %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.hi.marginal.ttest %>% filter(mineral == mineral_name)

  p <- ggplot(plot_data, aes(x = Treatment, y = hi, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    labs(title = mineral_name, x = NULL, y = "Mineral HI [grain / total]") +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = rel(1.0)),
      axis.title.y = element_text(face = "bold", size = rel(0.9)),
      plot.title = element_text(face = "bold", size = rel(1.55), hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = rel(0.9)),
      strip.background = element_blank()
    )

  if (nrow(ttest_subset) > 0) {
    p <- p + stat_pvalue_manual(ttest_subset, label = "p.adj.signif",
                                size = 8, bracket.size = 1, hide.ns = TRUE)
  }
  p
}

plot_mineral_HI <- function(mineral_name, base_size = 18) {
  plot_data <- to_plot_hi %>% filter(mineral == mineral_name)
  ttest_subset <- treatment.hi.ttest %>% filter(mineral == mineral_name)

  p <- ggplot(plot_data, aes(x = Treatment, y = hi, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    labs(title = mineral_name, x = NULL, y = "Mineral HI [grain / total]") +
    facet_wrap(~ Genotype) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = rel(1.0)),
      axis.title.y = element_text(face = "bold", size = rel(0.9)),
      plot.title = element_text(face = "bold", size = rel(1.55), hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = rel(0.9)),
      strip.background = element_blank()
    )

  if (nrow(ttest_subset) > 0) {
    p <- p + stat_pvalue_manual(ttest_subset, label = "p.adj.signif",
                                size = 8, bracket.size = 1, hide.ns = TRUE)
  }
  p
}
```

## Main 4-panel figure

```{r main-4panel-figure, fig.width=14, fig.height=16}
rowA_main <- ggarrange(
  plot_mineral_conc("P"),
  plot_mineral_conc("Zn") + ggpubr::rremove("ylab"),
  plot_mineral_conc("Ca", log_scale = TRUE) + ggpubr::rremove("ylab"),
  plot_mineral_conc("S")  + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowB_main <- ggarrange(
  plot_mineral_ratio("P") + labs(title = NULL) +
    scale_y_continuous(breaks = 1:5,
                       expand = expansion(mult = c(0.05, 0.15))),
  plot_mineral_ratio("Zn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_ratio("Ca") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_ratio("S")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowC_main <- ggarrange(
  plot_mineral_content("P")  + labs(title = NULL),
  plot_mineral_content("Zn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("Ca") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("S")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowD_main <- ggarrange(
  plot_mineral_HI("P")  + labs(title = NULL),
  plot_mineral_HI("Zn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("Ca") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("S")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

p_main_4panel <- ggarrange(
  rowA_main, rowB_main, rowC_main, rowD_main,
  ncol = 1, nrow = 4,
  heights = c(1, 0.8, 0.9, 0.8),
  labels = c("A", "B", "C", "D"),
  font.label = list(size = 30),
  align = "hv"
)
print(p_main_4panel)
```

## Supplementary 4-panel figure

```{r supp-4panel-figure, fig.width=14, fig.height=16}
rowA_supp <- ggarrange(
  plot_mineral_conc("Mg"),
  plot_mineral_conc("Mn", log_scale = TRUE) + ggpubr::rremove("ylab"),
  plot_mineral_conc("K")  + ggpubr::rremove("ylab"),
  plot_mineral_conc("Fe") + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowB_supp <- ggarrange(
  plot_mineral_ratio("Mg") + labs(title = NULL),
  plot_mineral_ratio("Mn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_ratio("K")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_ratio("Fe") + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowC_supp <- ggarrange(
  plot_mineral_content("Mg") + labs(title = NULL),
  plot_mineral_content("Mn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("K")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_content("Fe") + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

rowD_supp <- ggarrange(
  plot_mineral_HI("Mg") + labs(title = NULL),
  plot_mineral_HI("Mn") + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("K")  + labs(title = NULL) + ggpubr::rremove("ylab"),
  plot_mineral_HI("Fe") + labs(title = NULL) + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

p_supp_4panel <- ggarrange(
  rowA_supp, rowB_supp, rowC_supp, rowD_supp,
  ncol = 1, nrow = 4,
  heights = c(1, 0.8, 0.9, 0.8),
  labels = c("A", "B", "C", "D"),
  font.label = list(size = 30),
  align = "hv"
)
print(p_supp_4panel)
```

## Inv4m supp figure (4 metrics x 5 ions; Figure-2-style organization)

Rows are metrics (conc, content, HI, ratio) and columns are ions (Mg, Ca,
Zn, Fe, K). For conc and content, a single pool is shown per cell — the
pool where the Inv4m effect is significant under the unified FDR, or
`stover` as a fallback. Ion symbols are shown as plot titles on the top
row (A, conc); pool (grain / stover) appears as a facet strip on A and B.
Brackets are drawn only when the unified pooled-Genotype FDR is < 0.05.

```{r plot-inv4m-cell-helper}
pal_geno <- c("CTRL" = "gold", "Inv4m" = "purple4")

plot_inv4m_cell <- function(mineral_name, metric, pool_name = NULL,
                            show_title = FALSE, show_ylab = TRUE,
                            base_size = 18, log_scale = FALSE) {
  if (metric == "conc") {
    d <- to_plot_conc %>% filter(mineral == mineral_name, pool == pool_name)
    s <- genotype.conc.ttest %>% filter(mineral == mineral_name, pool == pool_name)
    y_var <- "ppm"
    ylab_text <- "Concentration [mg / kg]"
  } else if (metric == "content") {
    d <- to_plot_content %>% filter(mineral == mineral_name, pool == pool_name)
    s <- genotype.content.ttest %>% filter(mineral == mineral_name, pool == pool_name)
    y_var <- "mg"
    ylab_text <- "Content [mg / plant]"
  } else if (metric == "hi") {
    d <- to_plot_hi %>% filter(mineral == mineral_name)
    s <- genotype.hi.ttest %>% filter(mineral == mineral_name)
    y_var <- "hi"
    ylab_text <- "Mineral HI [grain / total]"
  } else if (metric == "ratio") {
    d <- to_plot_ratio %>% filter(mineral == mineral_name)
    s <- genotype.ratio.ttest %>% filter(mineral == mineral_name)
    y_var <- "ratio"
    ylab_text <- "Grain / stover ratio"
  }

  d <- d %>%
    mutate(Genotype = gsub("<i>|</i>", "", as.character(Genotype)),
           Genotype = factor(Genotype, levels = c("CTRL", "Inv4m")))

  p <- ggplot(d, aes(x = Genotype, y = .data[[y_var]], colour = Genotype)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1.2) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_colour_manual(values = pal_geno) +
    scale_x_discrete(labels = c("CTRL" = "CTRL", "Inv4m" = "<i>Inv4m</i>")) +
    labs(
      title = if (show_title) mineral_name
              else if (metric %in% c("hi", "ratio")) " "
              else NULL,
      x = NULL,
      y = if (show_ylab) ylab_text else NULL
    )

  if (log_scale) {
    p <- p + scale_y_continuous(
      trans = "log10", expand = expansion(mult = c(0.05, 0.15))
    )
  } else {
    p <- p + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
  }

  if (metric %in% c("conc", "content")) {
    p <- p + facet_wrap(~ pool, nrow = 1, strip.position = "top")
  }

  if (nrow(s) > 0) {
    p <- p + stat_pvalue_manual(
      s, label = "p.adj.signif",
      size = 8, bracket.size = 1, hide.ns = TRUE
    )
  }

  p + theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x = element_markdown(face = "bold", size = rel(1.0), color = "black"),
      axis.title.y = element_text(face = "bold", size = rel(0.9)),
      plot.title = element_text(face = "bold", size = rel(1.55), hjust = 0.5),
      strip.text = element_text(face = "bold", size = rel(0.9)),
      strip.background = element_blank()
    )
}
```

```{r inv4m-figure, fig.width=18, fig.height=14}
ions <- c("Mg", "Ca", "Zn", "Fe", "K")

# Selected pool per cell for conc / content. Sig pool if any, else stover.
pool_conc    <- c(Mg = "grain",  Ca = "stover", Zn = "stover",
                  Fe = "stover", K  = "stover")
pool_content <- c(Mg = "stover", Ca = "stover", Zn = "stover",
                  Fe = "grain",  K  = "stover")

# Row order: A = conc (with ion titles), B = ratio, C = content, D = HI
row_A <- lapply(ions, function(m) {
  plot_inv4m_cell(m, "conc", pool_name = pool_conc[[m]],
                  show_title = TRUE, show_ylab = (m == "Mg"))
})
row_B <- lapply(ions, function(m) {
  plot_inv4m_cell(m, "ratio",
                  show_title = FALSE, show_ylab = (m == "Mg"))
})
row_C <- lapply(ions, function(m) {
  plot_inv4m_cell(m, "content", pool_name = pool_content[[m]],
                  show_title = FALSE, show_ylab = (m == "Mg"))
})
row_D <- lapply(ions, function(m) {
  plot_inv4m_cell(m, "hi",
                  show_title = FALSE, show_ylab = (m == "Mg"))
})

p_inv4m_effects <- ggarrange(
  ggarrange(plotlist = row_A, ncol = 5, align = "h"),
  ggarrange(plotlist = row_B, ncol = 5, align = "h"),
  ggarrange(plotlist = row_C, ncol = 5, align = "h"),
  ggarrange(plotlist = row_D, ncol = 5, align = "h"),
  ncol = 1, nrow = 4,
  heights = c(1, 1, 1, 1),
  labels = c("A", "B", "C", "D"),
  font.label = list(size = 30)
)

print(p_inv4m_effects)
```

## Save outputs

```{r save-derived}
write.csv(
  effects,
  file.path(paths$intermediate, "ionome_content_linear_model_effects.csv"),
  row.names = FALSE
)

write_ttest <- function(x, name) {
  if (nrow(x) == 0) return(invisible(NULL))
  write.csv(
    x %>% select(where(~ !is.list(.))),
    file.path(paths$intermediate, name),
    row.names = FALSE
  )
}

write_ttest(treatment.t.test,       "ionome_content_treatment_ttests.csv")
write_ttest(treatment.conc.ttest,   "ionome_content_conc_treatment_ttests.csv")
write_ttest(treatment.hi.ttest,     "ionome_content_hi_treatment_ttests.csv")
write_ttest(treatment.ratio.ttest,  "ionome_content_ratio_treatment_ttests.csv")
write_ttest(genotype.conc.ttest,    "ionome_content_conc_genotype_ttests.csv")
write_ttest(genotype.content.ttest, "ionome_content_genotype_ttests.csv")
write_ttest(genotype.hi.ttest,      "ionome_content_hi_genotype_ttests.csv")
write_ttest(genotype.ratio.ttest,   "ionome_content_ratio_genotype_ttests.csv")
write_ttest(unified_geno_pool %>% select(where(~ !is.list(.))),
            "ionome_content_unified_genotype_ttests.csv")

ggsave(file.path(paths$figures, "ionome_4panel_main_figure.png"),
       plot = p_main_4panel, width = 15, height = 16, units = "in", dpi = 300)
ggsave(file.path(paths$figures, "ionome_4panel_supplementary_figure.png"),
       plot = p_supp_4panel, width = 15, height = 16, units = "in", dpi = 300)
ggsave(file.path(paths$figures, "ionome_inv4m_effects_figure.png"),
       plot = p_inv4m_effects, width = 14, height = 16, units = "in", dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
