---
title: "Inv4m Per-Plant Mineral Content (Biomass × Concentration)"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "phosphorus_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

source(here::here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("phosphorus_paper")
```

## Overview

Companion to `PSU2022_ionome.Rmd`. That workbook analyzes mineral
**concentration** (ppm in grain and stover). This one analyzes per-plant
mineral **content** (mg per plant):

- `grain_content_mg  = TKW_g   × conc_seed_ppm   / 1000`
- `stover_content_mg = STWHV_g × conc_stover_ppm / 1000`
- `total_content_mg  = grain_content + stover_content`
- `mineral_HI        = grain_content / total_content`

These quantities are derived at the raw plot level inside
`spatial_correction_for_INV4MXP.Rmd` and fit by SpATS as first-class
traits (same way `HI = TKW / STWHV` is handled). This workbook reads the
already-corrected content and mineral-HI columns and analyzes them;
nothing is re-derived here.

Content separates two processes the concentration-only analysis confounds:
net uptake (total content) and partitioning (mineral HI). This directly
addresses Miguel Piñeros' MP3 ask (processes behind mineral-concentration
changes) and Sawers' L176 question on per-plant flux rather than per-unit
grain.

## Load libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(ggpubr)
library(ggbeeswarm)
library(ggfx)
library(rstatix)
library(ggtext)
library(ggh4x)
```

## Load spatially corrected data

```{r load-data}
ionome <- read.csv(
  file.path(
    paths$intermediate,
    "PSU2022_spatially_corrected_both_treatments_with_ionome.csv"
  )
) %>%
  rename(Genotype = genotype) %>%
  filter(Genotype %in% c("CTRL", "Inv4m")) %>%
  droplevels()

ionome$Treatment <- factor(ionome$Treatment, levels = c("+P", "-P"))
ionome$Genotype  <- factor(ionome$Genotype,  levels = c("CTRL", "Inv4m"))

reliable <- c("Ca", "Fe", "K", "Mg", "Mn", "P", "S", "Zn")

content_vars <- c(
  paste0(reliable, "_grain_mg"),
  paste0(reliable, "_stover_mg"),
  paste0(reliable, "_total_mg"),
  paste0(reliable, "_HI")
)

# Alias: downstream chunks work against `content`. Nothing is derived here;
# all content + HI columns come from the upstream SpATS-corrected CSV.
content <- ionome

cat("Sample sizes:\n")
ionome %>% count(Genotype, Treatment) %>% print()
```

## Unit check: biomass and concentration ranges

Sanity-check the inputs before computing content. Expected:

- `TKW`: tens to hundreds of grams per plant (range roughly 30–300 g).
- `STWHV`: tens to hundreds of grams per plant.
- Concentrations: P/K in the thousands (ppm), Zn/Fe/Mn in the tens,
  Ca in hundreds to low thousands depending on tissue.

```{r unit-check-biomass}
ionome %>%
  select(Genotype, Treatment, TKW, STWHV) %>%
  pivot_longer(c(TKW, STWHV), names_to = "variable", values_to = "value") %>%
  group_by(variable, Genotype, Treatment) %>%
  summarise(
    n      = sum(!is.na(value)),
    min    = round(min(value, na.rm = TRUE), 1),
    median = round(median(value, na.rm = TRUE), 1),
    mean   = round(mean(value, na.rm = TRUE), 1),
    max    = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  )
```

```{r unit-check-concentrations}
conc_cols <- c(
  paste0(reliable, "_seed"),
  paste0(reliable, "_stover")
)

ionome %>%
  select(all_of(conc_cols)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    n      = sum(!is.na(value)),
    min    = round(min(value, na.rm = TRUE), 1),
    median = round(median(value, na.rm = TRUE), 1),
    mean   = round(mean(value, na.rm = TRUE), 1),
    max    = round(max(value, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(variable)
```

## Content summary per mineral × genotype × treatment

```{r content-summary}
content %>%
  select(Genotype, Treatment, all_of(content_vars)) %>%
  pivot_longer(all_of(content_vars), names_to = "variable", values_to = "value") %>%
  separate(variable, into = c("mineral", "pool"), sep = "_", extra = "merge") %>%
  mutate(
    pool = recode(pool,
                  grain_mg = "grain (mg)",
                  stover_mg = "stover (mg)",
                  total_mg = "total (mg)",
                  HI = "mineral HI")
  ) %>%
  group_by(mineral, pool, Genotype, Treatment) %>%
  summarise(
    n      = sum(!is.na(value)),
    median = signif(median(value, na.rm = TRUE), 3),
    mean   = signif(mean(value, na.rm = TRUE), 3),
    sd     = signif(sd(value, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(mineral, pool, Genotype, Treatment) %>%
  print(n = Inf)
```

## Biomass HI vs mineral HI per mineral

Rapid visual sanity check: does mineral HI track biomass HI, or does
it depart (true partitioning signal beyond biomass reallocation)?

```{r biomass-vs-mineral-hi, fig.width=10, fig.height=6}
# Order minerals by median total content per plant (descending): highest-flux
# first, so the eye lands on the dominant pools.
mineral_order <- content %>%
  select(all_of(paste0(reliable, "_total_mg"))) %>%
  pivot_longer(everything(), names_to = "mineral", values_to = "total_mg") %>%
  mutate(mineral = sub("_total_mg$", "", mineral)) %>%
  group_by(mineral) %>%
  summarise(median_total = median(total_mg, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(median_total)) %>%
  pull(mineral)

cat("Facet order (median total mg/plant, descending):\n")
print(mineral_order)

hi_long <- content %>%
  select(Genotype, Treatment, biomass_HI_frac,
         all_of(paste0(reliable, "_HI"))) %>%
  pivot_longer(ends_with("_HI"), names_to = "mineral", values_to = "mineral_HI") %>%
  mutate(
    mineral = sub("_HI$", "", mineral),
    mineral = factor(mineral, levels = mineral_order)
  ) %>%
  filter(!is.na(mineral_HI), !is.na(biomass_HI_frac))

ggplot(hi_long, aes(biomass_HI_frac, mineral_HI,
                    color = Treatment, shape = Genotype)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7, size = 1.5) +
  facet_wrap(~ mineral, scales = "free_y") +
  scale_color_manual(values = c("+P" = "gold", "-P" = "purple4")) +
  labs(
    x = "Biomass harvest index  [TKW / (TKW + STWHV)]",
    y = "Mineral HI  [grain content / total content]"
  ) +
  theme_classic(base_size = 12)
```

## Fit linear models across content + HI variables

```{r fit-models}
effects <- lapply(content_vars, function(x) {
  n_obs <- sum(!is.na(content[[x]]))
  if (n_obs < 20) return(NULL)

  model <- tryCatch(
    lm(as.formula(paste(x, "~ Genotype * Treatment")), data = content),
    error = function(e) NULL
  )
  if (is.null(model)) return(NULL)

  out <- coef(summary(model)) %>%
    as.data.frame() %>%
    rownames_to_column("predictor")
  colnames(out)[3] <- "Std.Error"
  colnames(out)[5] <- "p.value"
  out$response <- x
  out %>% select(response, everything())
}) %>%
  bind_rows() %>%
  mutate(p.adjust = p.adjust(p.value, method = "fdr")) %>%
  arrange(p.adjust)

cat("Total models fitted:", length(unique(effects$response)), "\n")
cat("Total effects tested:", nrow(effects), "\n")

cat("\n=== Significant effects (FDR < 0.05, no Intercept) ===\n")
effects %>%
  filter(p.adjust < 0.05, !grepl("Intercept", predictor)) %>%
  select(response, predictor, Estimate, p.value, p.adjust) %>%
  as_tibble() %>%
  print(n = Inf)
```

```{r interaction-effects}
interaction_effects <- effects %>%
  filter(grepl(":", predictor)) %>%
  separate(response, into = c("mineral", "pool"),
           sep = "_", extra = "merge", remove = FALSE) %>%
  mutate(
    signif = case_when(
      p.adjust < 0.001 ~ "***",
      p.adjust < 0.01  ~ "**",
      p.adjust < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  arrange(p.adjust)

cat("Genotype x Treatment interactions tested:", nrow(interaction_effects), "\n")
cat("Significant (FDR < 0.05):", sum(interaction_effects$p.adjust < 0.05), "\n")

if (any(interaction_effects$p.adjust < 0.05)) {
  interaction_effects %>%
    filter(p.adjust < 0.05) %>%
    select(mineral, pool, Estimate, p.value, p.adjust, signif) %>%
    print()
}
```

## Prepare long-format content for plotting + t-tests

```{r to-plot-content}
to_plot_content <- content %>%
  select(Genotype, Treatment,
         all_of(paste0(reliable, "_grain_mg")),
         all_of(paste0(reliable, "_stover_mg"))) %>%
  pivot_longer(
    cols = -c(Genotype, Treatment),
    names_to = "mineral_pool",
    values_to = "mg"
  ) %>%
  mutate(
    mineral = sub("_(grain|stover)_mg$", "", mineral_pool),
    pool    = sub("^.+_(grain|stover)_mg$", "\\1", mineral_pool)
  ) %>%
  select(-mineral_pool) %>%
  filter(!is.na(mg)) %>%
  mutate(
    pool     = factor(pool, levels = c("stover", "grain")),
    Genotype = gsub("Inv4m", "<i>Inv4m</i>", Genotype),
    Genotype = factor(Genotype, levels = c("CTRL", "<i>Inv4m</i>")),
    Treatment = factor(Treatment, levels = c("+P", "-P"))
  )
```

```{r treatment-ttests}
treatment.t.test <- to_plot_content %>%
  group_by(pool, Genotype, mineral) %>%
  rstatix::t_test(mg ~ Treatment) %>%
  rstatix::adjust_pvalue(method = "fdr") %>%
  rstatix::add_significance() %>%
  rstatix::add_y_position(scales = "free_y", step.increase = 0.12)

cat("\n=== Treatment t-tests (FDR < 0.05) ===\n")
treatment.t.test %>%
  filter(p.adj < 0.05) %>%
  select(mineral, pool, Genotype, statistic, p, p.adj, p.adj.signif) %>%
  arrange(p.adj) %>%
  as_tibble() %>%
  print(n = Inf)
```

## Per-mineral content plot function

```{r plot-mineral-content}
pal <- c("+P" = "gold", "-P" = "purple4")

plot_mineral_content <- function(mineral_name) {
  plot_data <- to_plot_content %>%
    filter(mineral == mineral_name)

  ttest_subset <- treatment.t.test %>%
    filter(mineral == mineral_name)

  ggplot(plot_data, aes(x = Treatment, y = mg, color = Treatment)) +
    geom_boxplot(width = 0.5, linewidth = 1, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 1) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    stat_pvalue_manual(
      ttest_subset,
      label = "p.adj.signif",
      size = 5,
      bracket.size = 0.5,
      hide.ns = TRUE
    ) +
    scale_color_manual(values = pal) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    ggh4x::facet_nested(~ pool + Genotype, scales = "free_y") +
    labs(title = mineral_name, x = NULL, y = "Content [mg / plant]") +
    theme_classic(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(color = "black", face = "bold", size = 15),
      axis.title.y = element_text(face = "bold", size = 13),
      plot.title = element_text(face = "bold", size = 28, hjust = 0.5),
      strip.text = element_markdown(face = "bold", size = 13),
      strip.background = element_blank(),
      panel.spacing = unit(0.1, "lines"),
      ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.5)
    )
}
```

## Main content figure (P, Zn, Ca, S)

```{r main-content-figure, fig.width=12, fig.height=4}
p_P  <- plot_mineral_content("P")
p_Zn <- plot_mineral_content("Zn")
p_Ca <- plot_mineral_content("Ca")
p_S  <- plot_mineral_content("S")

p_main_content <- ggarrange(
  p_P,
  p_Zn + ggpubr::rremove("ylab"),
  p_Ca + ggpubr::rremove("ylab"),
  p_S  + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

print(p_main_content)
```

## Supplementary content figure (Mg, Mn, K, Fe)

```{r supp-content-figure, fig.width=12, fig.height=4}
p_Mg <- plot_mineral_content("Mg")
p_Mn <- plot_mineral_content("Mn")
p_K  <- plot_mineral_content("K")
p_Fe <- plot_mineral_content("Fe")

p_supp_content <- ggarrange(
  p_Mg,
  p_Mn + ggpubr::rremove("ylab"),
  p_K  + ggpubr::rremove("ylab"),
  p_Fe + ggpubr::rremove("ylab"),
  ncol = 4, nrow = 1, align = "h"
)

print(p_supp_content)
```

## Save results

Per-plant corrected content is already in the upstream
`PSU2022_spatially_corrected_both_treatments_with_ionome.csv`, so this
chunk saves only the model outputs, t-tests, and figures unique to this
analysis.

```{r save-derived}
write.csv(
  effects,
  file.path(paths$intermediate, "ionome_content_linear_model_effects.csv"),
  row.names = FALSE
)

write.csv(
  interaction_effects,
  file.path(paths$intermediate, "ionome_content_interaction_effects.csv"),
  row.names = FALSE
)

# Drop list columns so add_y_position() output serializes cleanly.
treatment_export <- treatment.t.test %>% select(where(~ !is.list(.)))
write.csv(
  treatment_export,
  file.path(paths$intermediate, "ionome_content_treatment_ttests.csv"),
  row.names = FALSE
)

ggsave(
  file.path(paths$figures, "ionome_content_main_figure.pdf"),
  plot = p_main_content, width = 12, height = 4, units = "in"
)
ggsave(
  file.path(paths$figures, "ionome_content_main_figure.png"),
  plot = p_main_content, width = 12, height = 4, units = "in", dpi = 300
)
ggsave(
  file.path(paths$figures, "ionome_content_supplementary_figure.pdf"),
  plot = p_supp_content, width = 12, height = 4, units = "in"
)
ggsave(
  file.path(paths$figures, "ionome_content_supplementary_figure.png"),
  plot = p_supp_content, width = 12, height = 4, units = "in", dpi = 300
)
```

## Session info

```{r session-info}
sessionInfo()
```