---
title: "Trans-regulatory Network Analysis: Reference vs Novel Edges"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
 fig.height = 8
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Identify and visualize trans-regulatory networks controlled by genes
within the Inv4m inversion region. We focus on the Inv4m-specific network (genes
whose strongest regulator is within Inv4m, not flanking) and split it into
Reference (previously reported in MaizeNetome) and Novel (Inv4m-specific) edges.
**Approach**:

1. Build bipartite cis→trans co-expression network from expression data
2. Apply **strongest-regulator criterion**: keep targets where Inv4m regulator
   has the strongest correlation (not flanking)
3. Classify edges as **Reference** (in MaizeNetome) or **Novel** (not reported)
4. Visualize networks as minimum spanning trees on mutual information

**Key visualizations**:

- **Whole Inv4m trans network**: All edges from Inv4m regulators to targets
  where Inv4m is the strongest regulator
- **Reference network MST**: Edges previously reported in MaizeNetome
  (includes jmj2, pcna2, mrlk, engd2)
- **Novel network MST**: New Inv4m-specific edges not in MaizeNetome
  (includes jmj4 neighborhood)

**Output files** (for GO_Enrichment_Trans_Network.Rmd):

- `trans_network_edges_annotated.csv`
- `trans_network_regulators.csv`
- `trans_network_targets.csv`
- `trans_network_all_genes.csv`

---

# Setup and Dependencies

```{r libraries}
library(dplyr)
library(tidyr)
library(rtracklayer)
library(GenomicRanges)
library(igraph)
library(visNetwork)
```

---

# Helper Functions

```{r helper_functions}
#' Calculate mutual information from correlation coefficient
mutual_info <- function(r) {
 0.5 * log2(1 / (1 - r^2))
}

#' Calculate standardized euclidean distance for MST construction
std_dist <- function(r_scaled, n) {
  sqrt(2 * n * (1 - abs(r_scaled)))
}

#' Format nodes for network visualization
format_network_nodes <- function(node_ids, effects_df,
                                 inv4m_regulators, flanking_regulators,
                                 custom_labels = NULL) {
  nodes <- data.frame(id = node_ids) %>%
    left_join(
      effects_df %>%
        group_by(gene) %>%
        dplyr::slice(1) %>%
        ungroup() %>%
        dplyr::select(gene, locus_label, logFC),
      by = c("id" = "gene")
    ) %>%
    mutate(
      region = case_when(
        id %in% inv4m_regulators ~ "inv4m",
        id %in% flanking_regulators ~ "flank",
        TRUE ~ "out"
      ),
      upregulated = logFC > 0,
      color.border = case_when(
        region == "inv4m" ~ "#551A8B",
        region == "flank" ~ "#FF1493",
        region == "out" ~ "#FFD700"
      ),
      color.background = case_when(
        upregulated & region == "inv4m" ~ "#551A8B",
        upregulated & region == "flank" ~ "#FF1493",
        upregulated & region == "out" ~ "#FFD700",
        !upregulated ~ "white"
      ),
      borderWidth = 4,
      label = coalesce(locus_label, id)
    )

  # Handle duplicate labels
 dup_labels <- nodes %>%
    filter(!is.na(locus_label), locus_label != "") %>%
    group_by(locus_label) %>%
    filter(n() > 1) %>%
    pull(id)

  if (length(dup_labels) > 0) {
    nodes <- nodes %>%
      mutate(label = ifelse(
        id %in% dup_labels,
        paste0(locus_label, " (", id, ")"),
        label
      ))
  }

  # Apply custom labels
  if (!is.null(custom_labels) && any(names(custom_labels) %in% nodes$id)) {
    for (gene_id in names(custom_labels)) {
      if (gene_id %in% nodes$id) {
        nodes$label[nodes$id == gene_id] <- custom_labels[gene_id]
      }
    }
  }

  nodes$label <- paste0("<i>", nodes$label, "</i>")
  nodes
}

#' Create MST visualization from edge list
visualize_mst <- function(edges, nodes, effects_df, font_size = 20) {

  if (any(duplicated(nodes$label))) {
    stop("Cannot proceed with duplicate labels")
 }

  g <- graph_from_data_frame(
    d = edges %>% dplyr::select(from, to, r, corr_std, std_dist, mi),
    directed = TRUE,
    vertices = nodes
  )

  g_mst <- mst(g, weights = E(g)$std_dist)

  cat("MST - Edges:", length(E(g_mst)), "| Nodes:", length(V(g_mst)), "\n")

  mst_edges <- igraph::as_data_frame(g_mst, what = "edges") %>%
    left_join(
      effects_df %>%
        group_by(gene) %>%
        dplyr::slice(1) %>%
        ungroup() %>%
        dplyr::select(gene, logFC_from = logFC),
      by = c("from" = "gene")
    ) %>%
    left_join(
      effects_df %>%
        group_by(gene) %>%
        dplyr::slice(1) %>%
        ungroup() %>%
        dplyr::select(gene, logFC_to = logFC),
      by = c("to" = "gene")
    ) %>%
    mutate(
      is_concordant = sign(logFC_from) == sign(logFC_to),
      width = mi
    )

  # Filter nodes to only those in MST
  mst_node_ids <- unique(c(mst_edges$from, mst_edges$to))
  mst_nodes <- nodes %>% filter(id %in% mst_node_ids)

  mst_edges_vis <- mst_edges %>%
    mutate(
      arrows.to.type = ifelse(is_concordant, "arrow", "bar"),
      color = "DarkGray"
    ) %>%
    dplyr::select(from, to, width, arrows.to.type, color)

  visNetwork(mst_nodes, mst_edges_vis) %>%
    visNodes(font = list(size = font_size, multi = "html")) %>%
    visEdges(arrows = "to") %>%
    visOptions(
      highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
      nodesIdSelection = TRUE
    )
}

#' Visualize full network (non-MST) for smaller networks
visualize_full_network <- function(edges, nodes, effects_df, font_size = 20) {

  edge_df <- edges %>%
    left_join(
      effects_df %>%
        group_by(gene) %>%
        dplyr::slice(1) %>%
        ungroup() %>%
        dplyr::select(gene, logFC_from = logFC),
      by = c("from" = "gene")
    ) %>%
    left_join(
      effects_df %>%
        group_by(gene) %>%
        dplyr::slice(1) %>%
        ungroup() %>%
        dplyr::select(gene, logFC_to = logFC),
      by = c("to" = "gene")
    ) %>%
    mutate(
      is_concordant = sign(logFC_from) == sign(logFC_to),
      arrows.to.type = ifelse(is_concordant, "arrow", "bar"),
      color = "DarkGray",
      width = mi
    ) %>%
    dplyr::select(from, to, width, arrows.to.type, color)

  cat("Full Network - Edges:", nrow(edge_df), "| Nodes:", nrow(nodes), "\n")

  visNetwork(nodes, edge_df) %>%
    visNodes(font = list(size = font_size, multi = "html")) %>%
    visEdges(arrows = "to") %>%
    visOptions(
      highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
      nodesIdSelection = TRUE
    )
}
```

---

# Define Genomic Regions

```{r load_v5_annotations}
v5_gff <- file.path(paths$data, "Zea_mays.Zm-B73-REFERENCE-NAM-5.0.59.chr.gff3")
v5 <- rtracklayer::import(v5_gff) %>%
  subset(type == "gene" & seqnames %in% 1:10)

genes <- as.data.frame(v5)
genes$ID <- gsub("gene:", "", genes$ID)

cat("Total genes loaded:", nrow(genes), "\n")
```

```{r define_inv4m}
inv4m_start <- genes[genes$ID == "Zm00001eb190470", "start"]
inv4m_end <- genes[genes$ID == "Zm00001eb194800", "end"]

cat("Inv4m region:", inv4m_start, "-", inv4m_end, "\n")

inv4m_gene_ids <- genes %>%
  filter(
    seqnames == 4,
    start >= inv4m_start,
    end <= inv4m_end
  ) %>%
  pull(ID)

cat("Genes in Inv4m:", length(inv4m_gene_ids), "\n")
```

```{r define_flanking}
introgression_start <- 157012149
introgression_end <- 195900523

cis_introgression_gene_ids <- genes %>%
 filter(
    seqnames == 4,
    start >= introgression_start,
    end <= introgression_end
  ) %>%
  pull(ID)

flanking_introgression_gene_ids <- cis_introgression_gene_ids[
  !cis_introgression_gene_ids %in% inv4m_gene_ids
]

cat("Genes in cis introgression:", length(cis_introgression_gene_ids), "\n")
cat("Genes in flanking regions:", length(flanking_introgression_gene_ids), "\n")
```

---

# Load Differential Expression Results

```{r load_degs}
effects_df <- read.csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  na.strings = c("NA", NA, "")
) %>%
 filter(predictor == "Inv4m")

# Load consolidated labels from master file (includes curated + symbol + LLM-proposed)
master_labels <- read.csv(
  file.path(paths$data, "locus_labels_master.csv"),
  na.strings = c("NA", NA, "")
) %>%
  dplyr::select(gene, locus_label)

effects_df <- effects_df %>%
  dplyr::select(-any_of("locus_label")) %>%
  left_join(master_labels, by = "gene")

field_degs <- effects_df %>%
  filter(is_strong_DEG) %>%
  pull(gene) %>%
  unique()

cat("Total strong DEGs:", length(field_degs), "\n")
```

---

# Load MaizeNetome Reference Data

```{r load_maizenetome}
left_flanking_maizenetome <- read.csv(
  file.path(paths$data, "left_MaizeNetome_coexpression_edges.csv")
)
Inv4m_maizenetome <- read.csv(
  file.path(paths$data, "inv4m_MaizeNetome_coexpression_edges.csv")
)
right_flanking_maizenetome <- read.csv(
  file.path(paths$data, "right_MaizeNetome_coexpression_edges.csv")
)

MaizeNetome <- rbind(
  left_flanking_maizenetome %>%
    separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n") %>%
    mutate(network = "coexpression"),
  Inv4m_maizenetome %>%
    separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n") %>%
    mutate(network = "coexpression"),
  right_flanking_maizenetome %>%
    separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n") %>%
    mutate(network = "coexpression")
) %>%
  dplyr::rename(Direct.Connection = Direct.Connetation.Node)

cat("Total MaizeNetome edges:", nrow(MaizeNetome), "\n")
```

```{r convert_v4_v5}
v4_v5 <- read.table(
  file = file.path(paths$data, "B73v4_to_B73v5.tsv"),
  sep = "\t",
  header = FALSE
) %>%
  dplyr::rename(v4 = V1, v5 = V2) %>%
  separate_longer_delim(cols = v5, delim = ",")

MaizeNetome_v5 <- MaizeNetome %>%
  left_join(v4_v5, by = c("Element.ID" = "v4")) %>%
  dplyr::rename(Element.ID.v5 = v5) %>%
  filter(!is.na(Element.ID.v5)) %>%
  left_join(v4_v5, by = c("Direct.Connection" = "v4")) %>%
  dplyr::rename(Direct.Connection.v5 = v5) %>%
  filter(!is.na(Direct.Connection.v5))

cat("Edges after v4->v5 conversion:", nrow(MaizeNetome_v5), "\n")
```

---

# Build Bipartite Trans Co-expression Network

```{r define_cis_trans_genes}
cis_degs <- effects_df %>%
  filter(in_cis == TRUE, is_strong_DEG == TRUE) %>%
  inner_join(v4_v5, by = c(gene = "v5")) %>%
  group_by(gene) %>%
  arrange(adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup()

trans_degs <- effects_df %>%
  filter(in_trans == TRUE, is_strong_DEG == TRUE) %>%
  inner_join(v4_v5, by = c(gene = "v5")) %>%
  group_by(gene) %>%
  arrange(adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup()

cis_genes_all <- cis_degs$gene %>% sort() %>% unique()
trans_genes_all <- trans_degs$gene %>% sort() %>% unique()

inv4m_regulators <- cis_genes_all[cis_genes_all %in% inv4m_gene_ids]
flanking_regulators <- cis_genes_all[cis_genes_all %in% flanking_introgression_gene_ids]

cat("Cis DEGs:", length(cis_genes_all), "\n")
cat("  Inv4m regulators:", length(inv4m_regulators), "\n")
cat("  Flanking regulators:", length(flanking_regulators), "\n")
cat("Trans DEGs:", length(trans_genes_all), "\n")
```

```{r filter_reference_edges}
reference_edges_maizenetome <- MaizeNetome_v5 %>%
  filter(
    Element.ID.v5 %in% cis_genes_all,
    Direct.Connection.v5 %in% trans_genes_all
  ) %>%
  dplyr::rename(
    cis_gene = Element.ID.v5,
    trans_gene = Direct.Connection.v5
  ) %>%
  dplyr::select(cis_gene, trans_gene, network)

cat("Reference edges from MaizeNetome:", nrow(reference_edges_maizenetome), "\n")
```

```{r load_expression}
voomR <- readRDS(
  file.path(paths$intermediate, "normalized_expression_voom_object_leaf_trt.rds")
)
n_samples <- ncol(voomR$E)

cat("Expression matrix:", nrow(voomR$E), "genes x", n_samples, "samples\n")
```

```{r filter_ld_linked}
linked <- read.table(
  file.path(paths$intermediate, "inv4m_linked_genes_to_filter.txt"),
  header = TRUE
)

trans_genes_filtered <- trans_genes_all[!trans_genes_all %in% linked$gene_id]

cat("Trans genes after LD filtering:",
    length(trans_genes_all), "->", length(trans_genes_filtered), "\n")

reference_edges_maizenetome <- reference_edges_maizenetome %>%
  filter(trans_gene %in% trans_genes_filtered)
```

```{r build_bipartite}
cis_genes <- cis_genes_all[cis_genes_all %in% rownames(voomR$E)]
trans_genes <- trans_genes_filtered[trans_genes_filtered %in% rownames(voomR$E)]

cat("Genes in expression data:\n")
cat("  Cis:", length(cis_genes), "\n")
cat("  Trans:", length(trans_genes), "\n")

# Calculate correlations
expr_cis <- t(voomR$E)[, cis_genes]
expr_trans <- t(voomR$E)[, trans_genes]
corr_bipartite <- cor(expr_cis, expr_trans, use = "pairwise.complete.obs")

# Scaled correlations for MST
expr_cis_scaled <- scale(expr_cis)
expr_trans_scaled <- scale(expr_trans)
corr_bipartite_scaled <- cor(expr_cis_scaled, expr_trans_scaled,
                              use = "pairwise.complete.obs")

# Calculate p-values
p_values <- matrix(NA, nrow = nrow(corr_bipartite), ncol = ncol(corr_bipartite))
for (i in 1:nrow(corr_bipartite)) {
  for (j in 1:ncol(corr_bipartite)) {
    r <- corr_bipartite[i, j]
    t_stat <- r * sqrt((n_samples - 2) / (1 - r^2))
    p_values[i, j] <- 2 * pt(-abs(t_stat), df = n_samples - 2)
  }
}
p_adj <- matrix(p.adjust(c(p_values), method = "bonferroni"), nrow = nrow(p_values))

# Filter by Bonferroni-corrected p < 0.001. The cis x trans test universe is
# small (m ~ 1095), so BH at 0.05 was nearly equivalent to raw p < 0.05 and
# admitted edges down to |r| = 0.30. Bonferroni at alpha = 0.001 enforces a
# credibility floor of |r| ~ 0.67 on the surviving edges.
corr_bipartite_filtered <- corr_bipartite
corr_bipartite_filtered[p_adj > 0.001] <- 0

cat("Significant pairs (Bonferroni < 0.001):",
    sum(corr_bipartite_filtered != 0), "\n")

# Create edge list
edges_idx <- which(corr_bipartite_filtered != 0, arr.ind = TRUE)
bipartite_edges <- data.frame(
  from = rownames(corr_bipartite_filtered)[edges_idx[, 1]],
  to = colnames(corr_bipartite_filtered)[edges_idx[, 2]],
  r = corr_bipartite_filtered[edges_idx],
  corr_std = corr_bipartite_scaled[edges_idx]
)

cat("Bipartite network edges:", nrow(bipartite_edges), "\n")
```

---

# Define Inv4m Network (Strongest Regulator Criterion)

**Purpose**: Filter to targets where the Inv4m regulator has the strongest
correlation, ensuring we capture Inv4m-specific effects rather than shared
flanking regulation.

```{r strongest_regulator}
# For each target, find the strongest correlated regulator
strongest_edge_per_target <- bipartite_edges %>%
  mutate(abs_r = abs(r)) %>%
  group_by(to) %>%
  slice_max(abs_r, n = 1, with_ties = FALSE) %>%
  ungroup()

# Targets whose strongest regulator is NOT in flanking region
targets_strongest_nonflanking <- strongest_edge_per_target %>%
  filter(!from %in% flanking_regulators) %>%
  pull(to)

# Inv4m edges to these targets
inv4m_trans_edges <- bipartite_edges %>%
  filter(
    to %in% targets_strongest_nonflanking,
    from %in% inv4m_regulators
  )

cat("\n=== Inv4m Network (Strongest Regulator Criterion) ===\n")
cat("Targets with non-flanking as strongest:", length(targets_strongest_nonflanking), "\n")
cat("Inv4m trans edges:", nrow(inv4m_trans_edges), "\n")
cat("Unique genes:", length(unique(c(inv4m_trans_edges$from, inv4m_trans_edges$to))), "\n")
```

---

# Classify Edges: Reference vs Novel

```{r classify_edges}
inv4m_trans_edges$in_maizenetome <- mapply(
  function(f, t) {
    any(reference_edges_maizenetome$cis_gene == f &
        reference_edges_maizenetome$trans_gene == t)
  },
  inv4m_trans_edges$from,
  inv4m_trans_edges$to
)

inv4m_trans_edges$network_module <- ifelse(
  inv4m_trans_edges$in_maizenetome,
  "Reference",
  "Novel"
)

cat("\n=== Edge Classification ===\n")
print(table(inv4m_trans_edges$network_module))

reference_edges <- inv4m_trans_edges %>% filter(network_module == "Reference")
novel_edges <- inv4m_trans_edges %>% filter(network_module == "Novel")

cat("\nReference edges:", nrow(reference_edges), "\n")
cat("Novel edges:", nrow(novel_edges), "\n")
```

---

# Visualize Whole Inv4m Trans Network

**Purpose**: Overview of all Inv4m trans-regulatory connections before splitting
into Reference and Novel components.

```{r whole_inv4m_network}
# Prepare edges with distance metrics
whole_edges <- inv4m_trans_edges %>%
  mutate(
    std_dist = std_dist(corr_std, n_samples),
    mi = mutual_info(r)
  )

# Format nodes
whole_node_ids <- unique(c(whole_edges$from, whole_edges$to))
whole_nodes <- format_network_nodes(
  whole_node_ids, effects_df,
  inv4m_regulators, flanking_regulators
)

cat("\n=== Whole Inv4m Trans Network ===\n")
cat("Edges:", nrow(whole_edges), "\n")
cat("Nodes:", nrow(whole_nodes), "\n")

visualize_mst(whole_edges, whole_nodes, effects_df, font_size = 20)
```

---

# Visualize Reference Network (MaizeNetome Edges)
**Purpose**: Edges previously reported in MaizeNetome database. These represent
conserved co-expression relationships. Key genes: jmj2, pcna2, mrlk, engd2.

```{r reference_network}
if (nrow(reference_edges) > 0) {
  ref_edges <- reference_edges %>%
    mutate(
      std_dist = std_dist(corr_std, n_samples),
      mi = mutual_info(r)
    )

  ref_node_ids <- unique(c(ref_edges$from, ref_edges$to))
  ref_nodes <- format_network_nodes(
    ref_node_ids, effects_df,
    inv4m_regulators, flanking_regulators
  )

  cat("\n=== Reference Network (MaizeNetome) ===\n")
  cat("Edges:", nrow(ref_edges), "\n")
  cat("Nodes:", nrow(ref_nodes), "\n")

  visualize_mst(ref_edges, ref_nodes, effects_df, font_size = 25)
} else {
  cat("No Reference edges found\n")
}
```

---

# Visualize Novel Network (Inv4m-Specific Edges)

**Purpose**: New co-expression relationships not previously reported in
MaizeNetome. These represent Inv4m-specific regulatory effects.
Key genes: jmj4 neighborhood including pcna2, yhcf.

```{r novel_network}
if (nrow(novel_edges) > 0) {
  novel_edges_prep <- novel_edges %>%
    mutate(
      std_dist = std_dist(corr_std, n_samples),
      mi = mutual_info(r)
    )

  novel_node_ids <- unique(c(novel_edges_prep$from, novel_edges_prep$to))
  novel_nodes <- format_network_nodes(
    novel_node_ids, effects_df,
    inv4m_regulators, flanking_regulators
  )

  cat("\n=== Novel Network (Inv4m-Specific) ===\n")
  cat("Edges:", nrow(novel_edges_prep), "\n")
  cat("Nodes:", nrow(novel_nodes), "\n")

  visualize_mst(novel_edges_prep, novel_nodes, effects_df, font_size = 22)
} else {
  cat("No Novel edges found\n")
}
```

---

# Full Novel MST — Static Figure (Supplementary)

```{r novel_mst_static}
if (nrow(novel_edges) > 0) {
  library(htmlwidgets)
  library(webshot)

  vn_novel_full <- visualize_mst(novel_edges_prep, novel_nodes,
                                  effects_df, font_size = 22)

  # Set widget canvas to square before screenshotting
  vn_novel_full$width  <- 1400
  vn_novel_full$height <- 1400

  out_png  <- file.path(paths$figures, "supp_novel_trans_network_full.png")
  tmp_html <- tempfile(fileext = ".html")

  saveWidget(vn_novel_full, tmp_html, selfcontained = TRUE)
  webshot(tmp_html, file = out_png,
          selector = ".vis-network",
          vwidth = 1400, vheight = 1400,
          zoom = 2, delay = 2)
  unlink(tmp_html)

  cat("Saved:", out_png, "\n")
  vn_novel_full
}
```

---

# Visualize jmj4 Neighborhood (Novel MST Subgraph)

**Purpose**: Extract the jmj4-centered subgraph from the novel network MST.
This preserves the MST structure while focusing on jmj4, engd2, and their
targets (kan1, sec6, nii2, roc4, etc.).

```{r jmj4_neighborhood}
# Define the jmj4 neighborhood genes (from manual selection of novel MST)
jmj4_neighborhood_labels <- c(
  "jmj4", "engd2", "roc4", "kan1", "sec6", "nii2", "tet2", "flz22", "plat1"
)

# Genes with only assembly IDs (no locus labels)
jmj4_neighborhood_ids <- c(
  "Zm00001eb132370",
  "Zm00001eb193130",

  "Zm00001eb192970"
)

# Get gene IDs matching labels
jmj4_genes_by_label <- effects_df %>%
  filter(locus_label %in% jmj4_neighborhood_labels) %>%
  pull(gene) %>%
  unique()

# Combine both sources
jmj4_all_genes <- unique(c(jmj4_genes_by_label, jmj4_neighborhood_ids))

cat("jmj4 neighborhood genes:", length(jmj4_all_genes), "\n")

if (nrow(novel_edges_prep) > 0 && length(jmj4_all_genes) > 0) {
  # Build the full novel MST first
  g_novel <- graph_from_data_frame(
    d = novel_edges_prep %>% dplyr::select(from, to, r, corr_std, std_dist, mi),
    directed = TRUE,
    vertices = novel_nodes
  )
  g_novel_mst <- mst(g_novel, weights = E(g_novel)$std_dist)

  # Extract subgraph containing jmj4 neighborhood nodes
  jmj4_nodes_in_mst <- V(g_novel_mst)$name[V(g_novel_mst)$name %in% jmj4_all_genes]

  if (length(jmj4_nodes_in_mst) > 0) {
    g_jmj4_subgraph <- induced_subgraph(g_novel_mst, jmj4_nodes_in_mst)

    # Get edges and nodes from subgraph
    jmj4_mst_edges <- igraph::as_data_frame(g_jmj4_subgraph, what = "edges")
    jmj4_node_ids <- V(g_jmj4_subgraph)$name

    jmj4_nodes <- format_network_nodes(
      jmj4_node_ids, effects_df,
      inv4m_regulators, flanking_regulators
    )

    cat("\n=== jmj4 Neighborhood (Novel MST Subgraph) ===\n")
    cat("MST Edges:", nrow(jmj4_mst_edges), "\n")
    cat("Nodes:", length(jmj4_node_ids), "\n")

    # Prepare edges for visualization
    jmj4_vis_edges <- jmj4_mst_edges %>%
      left_join(
        effects_df %>% group_by(gene) %>% dplyr::slice(1) %>% ungroup() %>%
          dplyr::select(gene, logFC_from = logFC),
        by = c("from" = "gene")
      ) %>%
      left_join(
        effects_df %>% group_by(gene) %>% dplyr::slice(1) %>% ungroup() %>%
          dplyr::select(gene, logFC_to = logFC),
        by = c("to" = "gene")
      ) %>%
      mutate(
        is_concordant = sign(logFC_from) == sign(logFC_to),
        arrows.to.type = ifelse(is_concordant, "arrow", "bar"),
        color = "DarkGray",
        width = mi
      ) %>%
      dplyr::select(from, to, width, arrows.to.type, color)

    vn_jmj4 <- visNetwork(jmj4_nodes, jmj4_vis_edges) %>%
      visNodes(font = list(size = 25, multi = "html")) %>%
      visEdges(arrows = "to") %>%
      visOptions(
        highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
        nodesIdSelection = TRUE
      )

    # Save Figure 4 panel C (jmj4 neighborhood subgraph) as PNG via webshot
    library(htmlwidgets); library(webshot)
    vn_jmj4$width  <- 1400
    vn_jmj4$height <- 1400
    out_png_jmj4 <- file.path(paths$figures, "Figure_4_C_trans_network.png")
    tmp_html_jmj4 <- tempfile(fileext = ".html")
    saveWidget(vn_jmj4, tmp_html_jmj4, selfcontained = TRUE)
    webshot(tmp_html_jmj4, file = out_png_jmj4, selector = ".vis-network",
            vwidth = 1400, vheight = 1400, zoom = 2, delay = 2)
    unlink(tmp_html_jmj4)
    cat("Saved:", out_png_jmj4, "\n")

    vn_jmj4
  } else {
    cat("No jmj4 neighborhood genes found in novel MST\n")
  }
} else {
  cat("No novel edges or jmj4 genes to process\n")
}
```

---

# Export Files for GO Enrichment Analysis

```{r export_go_files}
# Add regulator_region to full bipartite edges for GO enrichment
trans_network_edges_annotated <- bipartite_edges %>%
  mutate(
    regulator_region = case_when(
      from %in% inv4m_regulators ~ "inv4m",
      from %in% flanking_regulators ~ "flanking",
      TRUE ~ "other"
    )
  ) %>%
  dplyr::select(from, to, r, regulator_region)

write.csv(
  trans_network_edges_annotated,
  file.path(paths$intermediate, "trans_network_edges_annotated.csv"),
  row.names = FALSE
)

cat("=== trans_network_edges_annotated.csv ===\n")
cat("Total edges:", nrow(trans_network_edges_annotated), "\n")
cat("By regulator_region:\n")
print(table(trans_network_edges_annotated$regulator_region))

# Regulators
trans_network_regulators <- effects_df %>%
  filter(gene %in% unique(bipartite_edges$from)) %>%
  group_by(gene) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene, locus_label, locus_symbol, desc_merged, logFC, adj.P.Val) %>%
  mutate(
    node_type = "regulator",
    region = case_when(
      gene %in% inv4m_regulators ~ "inv4m",
      gene %in% flanking_regulators ~ "flanking",
      TRUE ~ "other"
    )
  )

write.csv(
  trans_network_regulators,
  file.path(paths$intermediate, "trans_network_regulators.csv"),
  row.names = FALSE
)

cat("\n=== trans_network_regulators.csv ===\n")
cat("Total:", nrow(trans_network_regulators), "\n")
print(table(trans_network_regulators$region))

# Targets
trans_network_targets <- effects_df %>%
  filter(gene %in% unique(bipartite_edges$to)) %>%
  group_by(gene) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene, locus_label, locus_symbol, desc_merged, logFC, adj.P.Val) %>%
  mutate(node_type = "target", region = "trans")

write.csv(
  trans_network_targets,
  file.path(paths$intermediate, "trans_network_targets.csv"),
  row.names = FALSE
)

cat("\n=== trans_network_targets.csv ===\n")
cat("Total:", nrow(trans_network_targets), "\n")

# All genes
trans_network_all_genes <- bind_rows(
  trans_network_regulators,
  trans_network_targets
) %>%
  distinct(gene, .keep_all = TRUE)

write.csv(
  trans_network_all_genes,
  file.path(paths$intermediate, "trans_network_all_genes.csv"),
  row.names = FALSE
)

cat("\n=== trans_network_all_genes.csv ===\n")
cat("Total unique genes:", nrow(trans_network_all_genes), "\n")
```

---

# Export Reference and Novel Network Files

```{r export_modules}
# Reference network genes and edges
reference_genes <- unique(c(reference_edges$from, reference_edges$to))

reference_gene_info <- effects_df %>%
  filter(gene %in% reference_genes) %>%
  group_by(gene) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene, locus_label, desc_merged, logFC, adj.P.Val) %>%
  mutate(gene_location = ifelse(gene %in% cis_genes, "cis", "trans"))

write.csv(
  reference_gene_info,
  file.path(paths$intermediate, "reference_trans_network_genes.csv"),
  row.names = FALSE
)

write.csv(
  reference_edges,
  file.path(paths$intermediate, "reference_trans_network_edges.csv"),
  row.names = FALSE
)

cat("\n=== Reference Network ===\n")
cat("Genes:", nrow(reference_gene_info), "\n")
cat("Edges:", nrow(reference_edges), "\n")

# Novel network genes and edges
novel_genes <- unique(c(novel_edges$from, novel_edges$to))

novel_gene_info <- effects_df %>%
  filter(gene %in% novel_genes) %>%
  group_by(gene) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene, locus_label, desc_merged, logFC, adj.P.Val) %>%
  mutate(gene_location = ifelse(gene %in% cis_genes, "cis", "trans"))

write.csv(
  novel_gene_info,
  file.path(paths$intermediate, "novel_trans_network_genes.csv"),
  row.names = FALSE
)

write.csv(
  novel_edges,
  file.path(paths$intermediate, "novel_trans_network_edges.csv"),
  row.names = FALSE
)

cat("\n=== Novel Network ===\n")
cat("Genes:", nrow(novel_gene_info), "\n")
cat("Edges:", nrow(novel_edges), "\n")
```

---

# Supplementary Table — Novel Trans-coexpression Network Genes

```{r novel_network_table}
master_desc <- read.csv(
  file.path(paths$data, "locus_labels_master.csv"),
  na.strings = c("NA", "")
) %>%
  dplyr::select(gene, desc_pannzer, desc_merged_master = desc_merged)

# Join novel_nodes (region) with gene annotations and DEG stats
novel_table_df <- novel_nodes %>%
  dplyr::select(id, region) %>%
  left_join(
    effects_df %>%
      group_by(gene) %>%
      dplyr::slice(1) %>%
      ungroup() %>%
      dplyr::select(gene, locus_label, desc_merged, logFC, adj.P.Val),
    by = c("id" = "gene")
  ) %>%
  left_join(master_desc, by = c("id" = "gene")) %>%
  mutate(
    description = dplyr::coalesce(desc_merged_master, desc_merged, desc_pannzer),
    tex_gene  = paste0("\\textit{", id, "}"),
    tex_label = ifelse(!is.na(locus_label),
                       paste0("\\textit{", locus_label, "}"), "---"),
    tex_desc  = ifelse(!is.na(description),
                       gsub("%", "\\\\%", description), "---"),
    tex_lfc   = sprintf("%.2f", logFC),
    tex_fdr   = formatC(adj.P.Val, format = "e", digits = 2)
  ) %>%
  arrange(region, adj.P.Val)

inv4m_rows   <- novel_table_df %>% filter(region == "inv4m")
outside_rows <- novel_table_df %>% filter(region != "inv4m")

make_rows <- function(df) {
  apply(df, 1, function(r) {
    paste0("  ", r["tex_gene"], " & ", r["tex_label"], " & ",
           r["tex_desc"], " & ", r["tex_lfc"], " & ", r["tex_fdr"], " \\\\")
  })
}

header <- c(
  "\\toprule",
  "\\textbf{Gene ID} & \\textbf{Label} & \\textbf{Description} &",
  "  \\textbf{$\\log_2$FC} & \\textbf{FDR} \\\\",
  "\\midrule"
)

out_lines <- c(
  "\\begin{table*}[!ht]",
  "\\caption{\\textbf{Novel \\invfour trans-coexpression network genes.}",
  "All genes in the dataset-specific co-expression network (Bonferroni $< 0.001$),",
  "split by genomic location relative to the \\invfour introgression.",
  "$\\log_2$FC and FDR are from the genotype DEG analysis.}",
  "\\label{tab::novel_trans_network}",
  "\\resizebox{\\textwidth}{!}{%",
  "\\begin{tabular}{lllrr}",
  header,
  paste0("\\multicolumn{5}{l}{\\textbf{\\invfour region}} \\\\"),
  "\\midrule",
  make_rows(inv4m_rows),
  "\\midrule",
  paste0("\\multicolumn{5}{l}{\\textbf{Outside introgression}} \\\\"),
  "\\midrule",
  make_rows(outside_rows),
  "\\bottomrule",
  "\\end{tabular}}",
  "\\end{table*}"
)

out_tex <- file.path(paths$tables, "novel_trans_network_genes.tex")
writeLines(out_lines, out_tex)

cat("Saved:", out_tex, "\n")
cat("Inv4m region genes:", nrow(inv4m_rows), "\n")
cat("Outside introgression genes:", nrow(outside_rows), "\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```
