---
title: "Figure 3 Assembly: Global and Local Transcriptomic Effects of Inv4m"
author: "Fausto Rodriguez Zapata"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 10,
  dpi = 150
)

# Setup project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This notebook assembles individual panels into the final Figure 3 composite.
Panels B (MDS) and C (Volcano) are loaded from RDS files for proper ggplot assembly.
Other panels are loaded from PNG files.

**Panel layout (8 panels, A-H):**

```
  +---------+---------+
  |    A    |    D    |
  +---------+---------+
  |    B    |    E    |
  +---------+---------+
  |    C    |    F    |
  +---------+---------+
  |         |    G    |
  |         +---------+
  |         |    H    |
  +---------+---------+
```

- **Column 1:** A (Leaf sampling), B (MDS dim3 vs dim4), C (Volcano Inv4m)
- **Column 2:** D (Manhattan Leaf), E (Manhattan Inv4m), F (Haplotype), G (Local Manhattan), H (Local logFC)

**Output:** `results/inversion_paper/figures/Figure3_RNAseq.png`

---

# Setup

## Load Libraries

```{r libraries}
library(cowplot)
library(ggpubr)
library(ggtext)
library(magick)
library(grid)
library(ggplot2)
```

## Configuration

```{r configuration}
# Panel directory for PNGs
panel_dir <- file.path(paths$figures, "panels")

# Final figure dimensions (2-column width for journal)
fig_width_mm <- 180
fig_height_mm <- 200
dpi <- 300

# Convert mm to inches for ggsave
fig_width_in <- fig_width_mm / 25.4
fig_height_in <- fig_height_mm / 25.4
```

---

# Load Panels

## Panels B and C from RDS (ggplot objects)

```{r load_rds_panels}
# Load MDS plot (Panel B)
mds_file <- file.path(paths$intermediate, "transcript_MDS_34.rds")
if (file.exists(mds_file)) {
  panel_B <- readRDS(mds_file)
  # Remove axis text and ticks for figure assembly
  panel_B <- panel_B +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
  cat("Loaded MDS plot from RDS\n")
} else {
  warning("MDS RDS not found. Run differential_expression_leaf_treatment_model.Rmd first.")
  panel_B <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Panel B\n(missing)", size = 8, color = "gray50") +
    theme_void()
}

# Load Volcano plot (Panel C)
volcano_file <- file.path(paths$intermediate, "volcano_inv4m.rds")
if (file.exists(volcano_file)) {
  panel_C <- readRDS(volcano_file)
  cat("Loaded Volcano plot from RDS\n")
} else {
  warning("Volcano RDS not found. Run volcano_plot_analysis.Rmd first.")
  panel_C <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Panel C\n(missing)", size = 8, color = "gray50") +
    theme_void()
}
```

## Other Panels from PNG

```{r load_png_panels}
# Helper function to load PNG as ggplot
load_png_panel <- function(filepath, label) {
  if (file.exists(filepath)) {
    img <- image_read(filepath)
    grob <- rasterGrob(img, interpolate = TRUE)
    p <- ggdraw() + draw_grob(grob)
  } else {
    p <- ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = paste("Panel", label, "\n(missing)"),
               size = 8, color = "gray50") +
      theme_void()
  }
  return(p)
}

# Load PNG panels
panel_A <- load_png_panel(file.path(panel_dir, "panel_A_leaf_sampling.png"), "A")
panel_D <- load_png_panel(file.path(panel_dir, "panel_D_manhattan_leaf.png"), "D")
panel_E <- load_png_panel(file.path(panel_dir, "panel_E_manhattan_inv4m.png"), "E")
panel_F <- load_png_panel(file.path(panel_dir, "panel_F_haplotype.png"), "F")
panel_G <- load_png_panel(file.path(panel_dir, "panel_G_local_manhattan.png"), "G")
panel_H <- load_png_panel(file.path(panel_dir, "panel_H_local_logfc.png"), "H")

# Check status
png_files <- c(
  A = file.path(panel_dir, "panel_A_leaf_sampling.png"),
  D = file.path(panel_dir, "panel_D_manhattan_leaf.png"),
  E = file.path(panel_dir, "panel_E_manhattan_inv4m.png"),
  F = file.path(panel_dir, "panel_F_haplotype.png"),
  G = file.path(panel_dir, "panel_G_local_manhattan.png"),
  H = file.path(panel_dir, "panel_H_local_logfc.png")
)

status_df <- data.frame(
  Panel = c("A", "B", "C", "D", "E", "F", "G", "H"),
  Source = c("PNG", "RDS", "RDS", "PNG", "PNG", "PNG", "PNG", "PNG"),
  Status = c(
    ifelse(file.exists(png_files["A"]), "OK", "MISSING"),
    ifelse(file.exists(mds_file), "OK", "MISSING"),
    ifelse(file.exists(volcano_file), "OK", "MISSING"),
    ifelse(file.exists(png_files["D"]), "OK", "MISSING"),
    ifelse(file.exists(png_files["E"]), "OK", "MISSING"),
    ifelse(file.exists(png_files["F"]), "OK", "MISSING"),
    ifelse(file.exists(png_files["G"]), "OK", "MISSING"),
    ifelse(file.exists(png_files["H"]), "OK", "MISSING")
  )
)

knitr::kable(status_df, caption = "Panel Availability")
```

---

# Add Panel Labels

```{r add_labels}
add_label <- function(plot, label, fontface = "bold", size = 14) {
  ggdraw(plot) +
    draw_label(label, x = 0.02, y = 0.98, hjust = 0, vjust = 1,
               fontface = fontface, size = size)
}

panel_A_labeled <- add_label(panel_A, "A")
panel_B_labeled <- add_label(panel_B, "B")
panel_C_labeled <- add_label(panel_C, "C")
panel_D_labeled <- add_label(panel_D, "D")
panel_E_labeled <- add_label(panel_E, "E")
panel_F_labeled <- add_label(panel_F, "F")
panel_G_labeled <- add_label(panel_G, "G")
panel_H_labeled <- add_label(panel_H, "H")
```

---

# Assemble Figure

```{r assemble_figure, fig.width=12, fig.height=14}
# Column 1: A, B, C
col1 <- plot_grid(
  panel_A_labeled,
  panel_B_labeled,
  panel_C_labeled,
  ncol = 1,
  rel_heights = c(1, 1, 1),
  align = "v",
  axis = "lr"
)

# Column 2: D, E, F, G, H
col2 <- plot_grid(
  panel_D_labeled,
  panel_E_labeled,
  panel_F_labeled,
  panel_G_labeled,
  panel_H_labeled,
  ncol = 1,
  rel_heights = c(0.8, 0.8, 1.2, 1.2, 1.2),
  align = "v",
  axis = "lr"
)

# Combine columns
figure3 <- plot_grid(
  col1,
  col2,
  ncol = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tb"
)

figure3
```

---

# Save Output

```{r save_figure}
output_file <- file.path(paths$figures, "Figure3_RNAseq.png")

ggsave(
  output_file,
  plot = figure3,
  width = fig_width_in,
  height = fig_height_in,
  dpi = dpi,
  bg = "white"
)

cat(sprintf("Figure saved to: %s\n", output_file))
cat(sprintf("Dimensions: %.0f x %.0f mm at %d DPI\n", fig_width_mm, fig_height_mm, dpi))
```

---

# Summary

```{r summary}
cat("=== Assembly Complete ===\n\n")
cat("Panel sources:\n")
cat("  B (MDS): RDS -> axis removed for assembly\n")
cat("  C (Volcano): RDS\n")
cat("  A, D, E, F, G, H: PNG\n")
cat("\nTo update panels, run:\n")
cat("  - differential_expression_leaf_treatment_model.Rmd -> Panel B (MDS)\n")
cat("  - volcano_plot_analysis.Rmd -> Panel C (Volcano)\n")
cat("  - make_manhattan_plots.Rmd -> Panels D, E, G, H\n")
cat("  - plot_genotype_get_correlated_loci.Rmd -> Panel F\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```
