---
title: "Figure 3 Assembly: Global and Local Transcriptomic Effects of Inv4m"
author: "Fausto Rodriguez Zapata"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 10,
  dpi = 150
)

# Setup project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This notebook assembles individual panel PNGs into the final Figure 3 composite.
Each panel is generated by upstream Rmd scripts and saved to the `panels/` directory.

**Panel layout (8 panels, A-H):**

```
  +---------+---------+
  |    A    |    D    |
  +---------+---------+
  |    B    |    E    |
  +---------+---------+
  |    C    |    F    |
  +---------+---------+
  |         |    G    |
  |         +---------+
  |         |    H    |
  +---------+---------+
```

- **Column 1:** A (Leaf sampling), B (MDS dim3 vs dim4), C (Volcano Inv4m)
- **Column 2:** D (Manhattan Leaf), E (Manhattan Inv4m), F (Haplotype), G (Local Manhattan), H (Local logFC)

**Output:** `results/inversion_paper/figures/Figure3_RNAseq.png`

---

# Setup

## Load Libraries

```{r libraries}
library(cowplot)
library(magick)
library(grid)
library(png)
library(ggplot2)
```

## Configuration

```{r configuration}
# Panel directory
panel_dir <- file.path(paths$figures, "panels")

# Final figure dimensions (2-column width for journal)
# Nature/Science: ~180mm width, Cell: ~174mm
fig_width_mm <- 180
fig_height_mm <- 200
dpi <- 300

# Convert mm to inches for ggsave
fig_width_in <- fig_width_mm / 25.4
fig_height_in <- fig_height_mm / 25.4
```

---

# Panel Status Check

```{r panel_files}
# Panel file paths
panel_files <- list(
  A = file.path(panel_dir, "panel_A_leaf_sampling.png"),
  B = file.path(panel_dir, "panel_B_MDS.png"),
  C = file.path(panel_dir, "panel_C_volcano.png"),
  D = file.path(panel_dir, "panel_D_manhattan_leaf.png"),
  E = file.path(panel_dir, "panel_E_manhattan_inv4m.png"),
  F = file.path(panel_dir, "panel_F_haplotype.png"),
  G = file.path(panel_dir, "panel_G_local_manhattan.png"),
  H = file.path(panel_dir, "panel_H_local_logfc.png")
)

# Check which panels exist
panel_status <- sapply(panel_files, file.exists)

status_df <- data.frame(
  Panel = names(panel_status),
  Status = ifelse(panel_status, "OK", "MISSING"),
  File = basename(unlist(panel_files))
)

knitr::kable(status_df, caption = "Panel Availability")

if (!all(panel_status)) {
  missing <- names(panel_status)[!panel_status]
  warning(sprintf("Missing panels: %s. Using placeholders.", paste(missing, collapse = ", ")))
}
```

---

# Helper Functions

```{r helper_functions}
#' Load panel PNG or create placeholder
load_panel <- function(filepath, label) {
  if (file.exists(filepath)) {
    # Load PNG and convert to ggplot grob
    img <- image_read(filepath)
    grob <- rasterGrob(img, interpolate = TRUE)
    p <- ggdraw() + draw_grob(grob)
  } else {
    # Create placeholder with label
    p <- ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = paste("Panel", label, "\n(missing)"),
               size = 8, color = "gray50") +
      theme_void() +
      theme(
        panel.background = element_rect(fill = "gray95", color = "gray80"),
        plot.margin = margin(5, 5, 5, 5)
      )
  }
  return(p)
}

#' Add panel label (A, B, C, etc.)
add_label <- function(plot, label, fontface = "bold", size = 14) {
  ggdraw(plot) +
    draw_label(label, x = 0.02, y = 0.98, hjust = 0, vjust = 1,
               fontface = fontface, size = size)
}
```

---

# Load Panels

```{r load_panels}
# Load all panels
panels <- lapply(names(panel_files), function(label) {
  load_panel(panel_files[[label]], label)
})
names(panels) <- names(panel_files)

# Add panel labels
panels_labeled <- mapply(add_label, panels, names(panels), SIMPLIFY = FALSE)
```

---

# Assemble Figure

```{r assemble_figure, fig.width=12, fig.height=14}
# Column 1: A, B, C (equally spaced, 1/3 each)
col1 <- plot_grid(
  panels_labeled$A,
  panels_labeled$B,
  panels_labeled$C,
  ncol = 1,
  rel_heights = c(1, 1, 1),
  align = "v",
  axis = "lr"
)

# Column 2: D, E, F, G, H (stacked vertically)
# D, E are shorter (Manhattan plots); F, G, H are taller
col2 <- plot_grid(
  panels_labeled$D,
  panels_labeled$E,
  panels_labeled$F,
  panels_labeled$G,
  panels_labeled$H,
  ncol = 1,
  rel_heights = c(0.8, 0.8, 1.2, 1.2, 1.2),
  align = "v",
  axis = "lr"
)

# Combine columns side by side
figure3 <- plot_grid(
  col1,
  col2,
  ncol = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tb"
)

figure3
```

---

# Save Output

```{r save_figure}
# Save PNG
output_file <- file.path(paths$figures, "Figure3_RNAseq.png")

ggsave(
  output_file,
  plot = figure3,
  width = fig_width_in,
  height = fig_height_in,
  dpi = dpi,
  bg = "white"
)

cat(sprintf("Figure saved to: %s\n", output_file))
cat(sprintf("Dimensions: %.0f x %.0f mm at %d DPI\n", fig_width_mm, fig_height_mm, dpi))
```

---

# Summary

```{r summary}
cat("=== Assembly Complete ===\n\n")
cat("Panels used:\n")
for (p in names(panel_status)) {
  src <- if (panel_status[p]) basename(panel_files[[p]]) else "(placeholder)"
  cat(sprintf("  %s: %s\n", p, src))
}

cat("\nTo update panels, run the source scripts:\n")
cat("  - differential_expression_leaf_treatment_model.Rmd -> Panel B (MDS)\n")
cat("  - volcano_plot_analysis.Rmd -> Panel C (Volcano)\n
")
cat("  - make_manhattan_plots.Rmd -> Panels D, E, G, H (Manhattan plots)\n")
cat("  - plot_genotype_get_correlated_loci.Rmd -> Panel F (Haplotype)\n")
cat("  - Panel A: Create manually (leaf sampling schematic)\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```
