---
title: "WGCNA Module Perturbation Figure Assembly"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 12
)

knitr::opts_knit$set(root.dir = here::here())
```

# Overview

This notebook assembles the WGCNA module perturbation figure (Figure 6) from
pre-generated panel RDS files.

**Panels:**
- **A**: CTRL consensus network dendrogram with module colors
- **B**: Delta kWithin boxplot (connectivity change per module) with GO annotations
- **C**: Hub gene connectivity scatter plot (CTRL vs Inv4m)

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(cowplot)
  library(patchwork)
})

# Load project paths
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

```{r define_paths}
# Pipeline run directory
run_dir <- here("results/inversion_paper/field_perturbation/run_20251231_201332")

# Panel paths
panel_paths <- list(
  A = file.path(run_dir, "04_consensus_networks/ctrl/dendrogram_plot.rds"),
  B = file.path(run_dir, "06_preservation/delta_kWithin_boxplot_annotated.rds"),
  C = file.path(run_dir, "07_module_annotation/hub_connectivity_plot.rds")
)

# Verify all panels exist
for (panel in names(panel_paths)) {
  if (!file.exists(panel_paths[[panel]])) {
    stop(paste("Panel", panel, "not found:", panel_paths[[panel]]))
  }
}
cat("All panel files found.\n")
```

---

# Load Panels

```{r load_panels}
p_A <- readRDS(panel_paths$A)
p_B <- readRDS(panel_paths$B)
p_C <- readRDS(panel_paths$C)

cat("Panels loaded successfully.\n")
```

---

# Preview Individual Panels

## Panel A: Consensus Network Dendrogram

```{r preview_A, fig.width=10, fig.height=6}
print(p_A)
```

## Panel B: Delta kWithin Boxplot

```{r preview_B, fig.width=12, fig.height=10}
print(p_B)
```

## Panel C: Hub Gene Connectivity

```{r preview_C, fig.width=12, fig.height=16
print(p_C)
```

---

# Alternative Layouts

## Vertical Stack (A, B, C)

```{r vertical_layout, fig.width=12, fig.height=20}
fig_vertical <- plot_grid(
  p_A,
  p_B,
  p_C,
  ncol = 1,
  labels = c("A", "B", "C"),
  rel_heights = c(0.1, 0.5, 0.4)
)

ggsave(
  file.path(paths$figures, "WGCNA_module_perturbation.png"),
  fig_vertical,
  width = 12, height = 20
)

ggsave(
  file.path(paths$figures, "WGCNA_module_perturbation.svg"),
  fig_vertical,
  width = 12, height = 20
)

print(fig_vertical)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
