---
title: "WGCNA Module Perturbation Figure Assembly"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 12
)

knitr::opts_knit$set(root.dir = here::here())
```

# Overview

This notebook assembles the WGCNA module perturbation figure (Figure 6) from
pre-generated panel RDS files.

**Panels:**
- **A**: CTRL consensus network dendrogram with module colors
- **B**: Z-score ridge plot (kWithin permutation test, genotype-response + leaf-gradient)
- **C**: Hub gene connectivity scatter plot (CTRL vs Inv4m)

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(cowplot)
  library(patchwork)
})

# Load project paths
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

```{r define_paths}
# Pipeline run directory
run_dir <- here("results/inversion_paper/field_perturbation/run_20251231_201332")

# Panel paths.
# Panel C was previously the WGCNA hub-gene connectivity scatter from the
# field_perturbation pipeline. Per Ruben's restructure (2026-05-18) it is now
# the hand-assembled trans-network MST composite produced from the Bonferroni
# iteration in agent/inversion_paper/novel_edge_filter_iterate.html.
panel_paths <- list(
  A = file.path(run_dir, "04_consensus_networks/ctrl/dendrogram_plot.rds"),
  B = file.path(run_dir, "06_preservation/zscore_density_kWithin.rds"),
  C = here("results/inversion_paper/figures/Figure_4_C_trans_network.png")
)

# Verify all panels exist
for (panel in names(panel_paths)) {
  if (!file.exists(panel_paths[[panel]])) {
    stop(paste("Panel", panel, "not found:", panel_paths[[panel]]))
  }
}
cat("All panel files found.\n")
```

---

# Load Panels

```{r load_panels}
p_A <- readRDS(panel_paths$A)
p_B <- readRDS(panel_paths$B)
p_C <- cowplot::ggdraw() + cowplot::draw_image(panel_paths$C)

cat("Panels loaded successfully.\n")
```

---

# Preview Individual Panels

## Panel A: Consensus Network Dendrogram

```{r preview_A, fig.width=10, fig.height=6}
print(p_A)
```

## Panel B: kWithin Permutation Ridge Plot

```{r preview_B, fig.width=12, fig.height=10}
print(p_B)
```

## Panel C: Hub Gene Connectivity

```{r preview_C, fig.width=12, fig.height=16}
print(p_C)
```

---

# Alternative Layouts

## Layout: A on top, B and C side by side below

```{r combined_layout, fig.width=14, fig.height=12}
# Bottom row: B and C side by side
bottom_row <- plot_grid(
  p_B,
  p_C,
  ncol = 2,
  labels = c("B", "C"),
  label_size = 35,
  rel_widths = c(1, 1)
)

# Combined: A on top, B+C below
fig_combined <- plot_grid(
  p_A,
  bottom_row,
  ncol = 1,
  labels = c("A", ""),
  label_size = 35,
  rel_heights = c(0.15, 0.85)
)

ggsave(
  file.path(paths$figures, "WGCNA_module_perturbation.png"),
  fig_combined,
  width = 14, height = 12,
  dpi = 300
)

ggsave(
  file.path(paths$figures, "WGCNA_module_perturbation.pdf"),
  fig_combined,
  width = 14, height = 12
)

print(fig_combined)
```

---

# Supplementary Figure — Hub Gene Connectivity

```{r supp_hub_connectivity, fig.width=10, fig.height=8}
p_hub <- readRDS(file.path(run_dir, "07_module_annotation/hub_connectivity_plot.rds")) +
  theme(
    legend.position      = c(0.05, 0.95),
    legend.justification = c(0, 1)
  )

ggsave(
  file.path(paths$figures, "supp_hub_connectivity.png"),
  p_hub,
  width = 10, height = 8, dpi = 300
)

print(p_hub)
```

---

# Supplementary Figure — sec6 Leaf Trajectory

```{r supp_sec6_trajectory, fig.width=10, fig.height=5}
library(ggtext)

psu_voom <- readRDS(file.path(paths$intermediate,
                               "normalized_expression_voom_object_leaf_trt.rds"))
expr_mat  <- psu_voom$E
full_meta <- psu_voom$targets %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  mutate(Genotype = recode(Genotype, "B73" = "CTRL"))

sec6_gene <- "Zm00001eb194380"

sec6_df <- tibble(
  sample_id = colnames(expr_mat),
  log2CPM   = expr_mat[sec6_gene, ]
) %>%
  left_join(full_meta %>% select(sample_id, Genotype, leaf_tissue, Treatment),
            by = "sample_id") %>%
  mutate(leaf = factor(leaf_tissue))

sec6_centered <- sec6_df %>%
  group_by(Genotype) %>%
  mutate(centered_log2CPM = log2CPM - mean(log2CPM)) %>%
  ungroup()

summary_raw <- sec6_df %>%
  group_by(Genotype, leaf) %>%
  summarise(mean_expr = mean(log2CPM),
            se_expr   = sd(log2CPM) / sqrt(n()),
            .groups   = "drop")

summary_centered <- sec6_centered %>%
  group_by(Genotype, leaf) %>%
  summarise(mean_expr = mean(centered_log2CPM),
            se_expr   = sd(centered_log2CPM) / sqrt(n()),
            .groups   = "drop")

geno_colors <- c("CTRL" = "gold", "Inv4m" = "purple4")

p_raw <- ggplot(summary_raw,
                aes(x = leaf, y = mean_expr, color = Genotype, group = Genotype)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_expr - se_expr, ymax = mean_expr + se_expr),
                width = 0.2) +
  geom_jitter(data = sec6_df,
              aes(x = leaf, y = log2CPM, color = Genotype),
              alpha = 0.4, size = 1.5, width = 0.1, inherit.aes = FALSE) +
  scale_color_manual(values = geno_colors) +
  theme_minimal(base_size = 14) +
  labs(x     = "Leaf",
       y     = expression(log[2] * "(CPM)"),
       title = "*sec6* (*Zm00001eb194380*)",
       subtitle = "Raw expression",
       color = NULL) +
  theme(plot.title = element_markdown())

p_centered <- ggplot(summary_centered,
                     aes(x = leaf, y = mean_expr, color = Genotype, group = Genotype)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_expr - se_expr, ymax = mean_expr + se_expr),
                width = 0.2) +
  geom_jitter(data = sec6_centered,
              aes(x = leaf, y = centered_log2CPM, color = Genotype),
              alpha = 0.4, size = 1.5, width = 0.1, inherit.aes = FALSE) +
  scale_color_manual(values = geno_colors) +
  theme_minimal(base_size = 14) +
  labs(x        = "Leaf",
       y        = expression("Centered " * log[2] * "(CPM)"),
       title    = "*sec6* (*Zm00001eb194380*)",
       subtitle = "Centered per genotype",
       color    = NULL) +
  theme(plot.title = element_markdown())

p_sec6 <- p_raw + p_centered +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "top")

ggsave(
  file.path(paths$figures, "sec6_leaf_trajectory.png"),
  p_sec6, width = 10, height = 5, dpi = 300
)

print(p_sec6)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
