---
title: "Figure 3 Assembly: Global and Local Transcriptomic Effects of Inv4m"
author: "Fausto Rodriguez Zapata"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 10,
  dpi = 150
)

# Setup project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This notebook assembles individual panels into the final Figure 3 composite.
All panels (except A which is a user-provided PNG) are loaded from RDS files
for font size control and proper ggplot assembly.

**Panel layout (8 panels, A-H):**

```
  +---------+---------+
  |    A    |    D    |
  +---------+---------+
  |    B    |    E    |
  +---------+---------+
  |    C    |    F    |
  +---------+---------+
  |         |    G    |
  |         +---------+
  |         |    H    |
  +---------+---------+
```

- **Column 1:** A (Leaf sampling), B (MDS dim3 vs dim4), C (Volcano Inv4m)
- **Column 2:** D (Manhattan Leaf), E (Manhattan Inv4m), F (Haplotype), G (Local Manhattan), H (Local logFC)

**Output:** `results/inversion_paper/figures/Figure3_RNAseq.png`

---

# Setup

## Load Libraries

```{r libraries}
library(cowplot)
library(ggpubr)
library(ggtext)
library(magick)
library(grid)
library(ggplot2)
```

## Configuration

```{r configuration}
# Panel directory for PNGs
panel_dir <- file.path(paths$figures, "panels")

# Final figure dimensions (2-column width for journal)
# Note: height increased to accommodate 8 panels with ggrepel labels
fig_width_mm <- 280
fig_height_mm <- 370
dpi <- 300

# Convert mm to inches for ggsave
fig_width_in <- fig_width_mm / 25.4
fig_height_in <- fig_height_mm / 25.4
```

---

# Load Panels

## Load RDS Panels (B, C, D, E, F, G, H)

```{r load_rds_panels}
# Helper to load RDS panel with fallback
load_rds_panel <- function(filepath, label, script_hint = "") {
  if (file.exists(filepath)) {
    p <- readRDS(filepath)
    cat(sprintf("Loaded Panel %s from RDS\n", label))
    return(p)
  } else {
    warning(sprintf("Panel %s RDS not found. %s", label, script_hint))
    return(ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = paste("Panel", label, "\n(missing)"),
               size = 8, color = "gray50") +
      theme_void())
  }
}

# Panel B: MDS plot
mds_file <- file.path(paths$intermediate, "transcript_MDS_34.rds")
panel_B <- load_rds_panel(mds_file, "B",
  "Run differential_expression_leaf_treatment_model.Rmd first.")
# Remove axis text/ticks for figure assembly
if (file.exists(mds_file)) {
  panel_B <- panel_B +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}

# Panel C: Volcano plot
volcano_file <- file.path(paths$intermediate, "volcano_inv4m.rds")
panel_C <- load_rds_panel(volcano_file, "C",
  "Run volcano_plot_analysis.Rmd first.")

# Panel D: Manhattan Leaf
panel_D_file <- file.path(paths$intermediate, "panel_D_manhattan_leaf.rds")
panel_D <- load_rds_panel(panel_D_file, "D",
  "Run make_manhattan_plots.Rmd first.")

# Panel E: Manhattan Inv4m
panel_E_file <- file.path(paths$intermediate, "panel_E_manhattan_inv4m.rds")
panel_E <- load_rds_panel(panel_E_file, "E",
  "Run make_manhattan_plots.Rmd first.")

# Panel F: Haplotype
panel_F_file <- file.path(paths$intermediate, "panel_F_haplotype.rds")
panel_F <- load_rds_panel(panel_F_file, "F",
  "Run plot_genotype_get_correlated_loci.Rmd first.")

# Panel G: Local Manhattan
panel_G_file <- file.path(paths$intermediate, "panel_G_local_manhattan.rds")
panel_G <- load_rds_panel(panel_G_file, "G",
  "Run make_manhattan_plots.Rmd first.")

# Panel H: Local logFC
panel_H_file <- file.path(paths$intermediate, "panel_H_local_logfc.rds")
panel_H <- load_rds_panel(panel_H_file, "H",
  "Run make_manhattan_plots.Rmd first.")
```

## Panel A from PNG (user-provided schematic)

```{r load_png_panels}
# Helper function to load PNG as ggplot
load_png_panel <- function(filepath, label) {
  if (file.exists(filepath)) {
    img <- image_read(filepath)
    grob <- rasterGrob(img, interpolate = TRUE)
    p <- ggdraw() + draw_grob(grob)
  } else {
    p <- ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = paste("Panel", label, "\n(missing)"),
               size = 8, color = "gray50") +
      theme_void()
  }
  return(p)
}

# Load Panel A (leaf sampling schematic - user-provided PNG)
panel_A_file <- file.path(panel_dir, "panel_A_leaf_sampling.png")
panel_A <- load_png_panel(panel_A_file, "A")
```

## Panel Status

```{r panel_status}
# Check status of all panels
status_df <- data.frame(
  Panel = c("A", "B", "C", "D", "E", "F", "G", "H"),
  Source = c("PNG", "RDS", "RDS", "RDS", "RDS", "RDS", "RDS", "RDS"),
  Status = c(
    ifelse(file.exists(panel_A_file), "OK", "MISSING"),
    ifelse(file.exists(mds_file), "OK", "MISSING"),
    ifelse(file.exists(volcano_file), "OK", "MISSING"),
    ifelse(file.exists(panel_D_file), "OK", "MISSING"),
    ifelse(file.exists(panel_E_file), "OK", "MISSING"),
    ifelse(file.exists(panel_F_file), "OK", "MISSING"),
    ifelse(file.exists(panel_G_file), "OK", "MISSING"),
    ifelse(file.exists(panel_H_file), "OK", "MISSING")
  )
)

knitr::kable(status_df, caption = "Panel Availability")
```

---

# Adjust Panel Fonts

Balance font sizes across all panels for consistent appearance.

```{r adjust_fonts}
# Target base size for readable axes
base_size <- 15

# Panel B (MDS): Keep axes readable, shrink legend (legend.text uses element_markdown)
panel_B <- panel_B +
  theme(
    axis.title = element_text(size = base_size),
    legend.title = element_markdown(size = 15),
    legend.background = element_rect(fill = "transparent"),
    legend.key = element_rect(fill = "transparent", colour = NA),
    legend.position = "top",
    plot.margin = unit(c(5.5, 5.5, 5.5, 40), "pt")
  )

# Panel C (Volcano): Readable axes, smaller legend
panel_C <- panel_C +
  theme(
    axis.title = element_text(size = base_size),
    axis.text = element_text(size = base_size - 2),
    legend.key.size = unit(0.12, "cm"),
    legend.position = c(0.81, 0.94),
    legend.background = element_rect(fill = "transparent"),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )

# Panels D, E (Manhattan): Readable axes, smaller title and legend
manhattan_theme <- theme(
  plot.title = element_markdown(size = base_size, face = "bold", hjust = 1),
  axis.title = element_text(size = base_size),
  axis.text = element_text(size = base_size - 2),
  axis.text.x = element_text(size = base_size - 3)
)
panel_D <- panel_D + manhattan_theme
panel_E <- panel_E + manhattan_theme

# Panels G, H (Local Manhattan/logFC): Readable axes
local_theme <- theme(
  axis.title = element_text(size = base_size),
  axis.text.x = element_text(size = base_size - 3),
  plot.margin = unit(c(5.5, 5.5, 5.5, 20), "pt")
)

# Panel F: Haplotype
panel_F <- panel_F + local_theme +
  theme(
    legend.title = element_markdown(size = 15)
  )

panel_G <- panel_G + local_theme
panel_H <- panel_H + local_theme
```

---

# Add Panel Labels

```{r add_labels}
add_label <- function(plot, label, fontface = "bold", size = 14) {
  ggdraw(plot) +
    draw_label(label, x = 0.02, y = 0.98, hjust = 0, vjust = 1,
               fontface = fontface, size = size)
}

panel_A_labeled <- add_label(panel_A, "A")
panel_B_labeled <- add_label(panel_B, "B")
panel_C_labeled <- add_label(panel_C, "C")
panel_D_labeled <- add_label(panel_D, "D")
panel_E_labeled <- add_label(panel_E, "E")
panel_F_labeled <- add_label(panel_F, "F")
panel_G_labeled <- add_label(panel_G, "G")
panel_H_labeled <- add_label(panel_H, "H")
```

---

# Assemble Figure

```{r assemble_figure, fig.width=12, fig.height=14}
# Column 1: A, B, C
col1 <- plot_grid(
  panel_A_labeled,
  panel_B_labeled,
  panel_C_labeled,
  ncol = 1,
  rel_heights = c(1, 1, 1),
  align = "v",
  axis = "lr"
)

# Column 2: D, E, F, G, H
# Panel F needs much more height for the 3 haplotype lines to be clearly visible
col2 <- plot_grid(
  panel_D_labeled,
  panel_E_labeled,
  panel_F_labeled,
  panel_G_labeled,
  panel_H_labeled,
  ncol = 1,
  rel_heights = c(0.6, 0.6, 0.6, 0.9, 0.9),
  align = "v",
  axis = "lr"
)

# Combine columns
figure3 <- plot_grid(
  col1,
  col2,
  ncol = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tb"
)

figure3
```

---

# Save Output

```{r save_figure}
output_file <- file.path(paths$figures, "Figure3_RNAseq.png")

ggsave(
  output_file,
  plot = figure3,
  width = fig_width_in,
  height = fig_height_in,
  dpi = dpi,
  bg = "white"
)

cat(sprintf("Figure saved to: %s\n", output_file))
cat(sprintf("Dimensions: %.0f x %.0f mm at %d DPI\n", fig_width_mm, fig_height_mm, dpi))
```

---

# Summary

```{r summary}
cat("=== Assembly Complete ===\n\n")
cat("Panel sources:\n")
cat("  A (Leaf sampling): PNG (user-provided schematic)\n")
cat("  B (MDS): RDS -> axis removed for assembly\n")
cat("  C (Volcano): RDS\n")
cat("  D (Manhattan Leaf): RDS\n")
cat("  E (Manhattan Inv4m): RDS\n")
cat("  F (Haplotype): RDS\n")
cat("  G (Local Manhattan): RDS\n")
cat("  H (Local logFC): RDS\n")
cat("\nTo regenerate panels, run:\n")
cat("  - differential_expression_leaf_treatment_model.Rmd -> Panel B (MDS)\n")
cat("  - volcano_plot_analysis.Rmd -> Panel C (Volcano)\n")
cat("  - make_manhattan_plots.Rmd -> Panels D, E, G, H\n")
cat("  - plot_genotype_get_correlated_loci.Rmd -> Panel F\n")
cat("\nAll RDS panels can have fonts adjusted in this assembly script.\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```
