---
title: "WGCNA Module Perturbation Figure Assembly"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 12
)

knitr::opts_knit$set(root.dir = here::here())
```

# Overview

This notebook assembles the WGCNA module perturbation figure (Figure 6) from
pre-generated panel RDS files.

**Panels:**
- **A**: CTRL consensus network dendrogram with module colors
- **B**: Z-score ridge plot (kWithin permutation test, genotype-response + leaf-gradient)
- **C**: Hub gene connectivity scatter plot (CTRL vs Inv4m)

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(cowplot)
  library(patchwork)
})

# Load project paths
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

```{r define_paths}
# Pipeline run directory
run_dir <- here("results/inversion_paper/field_perturbation/run_20251231_201332")

# Panel paths.
# Panel C was previously the WGCNA hub-gene connectivity scatter from the
# field_perturbation pipeline. Per Ruben's restructure (2026-05-18) it is now
# the hand-assembled trans-network MST composite produced from the Bonferroni
# iteration in agent/inversion_paper/novel_edge_filter_iterate.html.
panel_paths <- list(
  A = file.path(run_dir, "04_consensus_networks/ctrl/dendrogram_plot.rds"),
  B = file.path(run_dir, "06_preservation/zscore_density_kWithin.rds"),
  C = here("results/inversion_paper/figures/Figure_4_C_trans_network.png")
)

# Verify all panels exist
for (panel in names(panel_paths)) {
  if (!file.exists(panel_paths[[panel]])) {
    stop(paste("Panel", panel, "not found:", panel_paths[[panel]]))
  }
}
cat("All panel files found.\n")
```

---

# Load Panels

```{r load_panels}
p_A <- readRDS(panel_paths$A)
p_B <- readRDS(panel_paths$B)
p_C <- cowplot::ggdraw() + cowplot::draw_image(panel_paths$C)

cat("Panels loaded successfully.\n")
```

---

# Preview Individual Panels

## Panel A: Consensus Network Dendrogram

```{r preview_A, fig.width=10, fig.height=6}
print(p_A)
```

## Panel B: kWithin Permutation Ridge Plot

```{r preview_B, fig.width=12, fig.height=10}
print(p_B)
```

## Panel C: Hub Gene Connectivity

```{r preview_C, fig.width=12, fig.height=16}
print(p_C)
```

---

# Alternative Layouts

## Layout: A on top, B and C side by side below

```{r combined_layout, fig.width=14, fig.height=12}
# Bottom row: B and C side by side
bottom_row <- plot_grid(
  p_B,
  p_C,
  ncol = 2,
  labels = c("B", "C"),
  label_size = 35,
  rel_widths = c(1, 1)
)

# Combined: A on top, B+C below
fig_combined <- plot_grid(
  p_A,
  bottom_row,
  ncol = 1,
  labels = c("A", ""),
  label_size = 35,
  rel_heights = c(0.15, 0.85)
)

ggsave(
  file.path(paths$figures, "WGCNA_module_perturbation.png"),
  fig_combined,
  width = 14, height = 12,
  dpi = 300
)

ggsave(
  file.path(paths$figures, "WGCNA_module_perturbation.pdf"),
  fig_combined,
  width = 14, height = 12
)

print(fig_combined)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
