---
title: "Regional correlation structure and expression magnitude along the Inv4m introgression"
author: "FV Rodriguez"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4
)
```

# Overview

This notebook examines how expression correlation and fold change magnitude
vary along the Inv4m introgression on chromosome 4. Three questions:

1. **Inter-regional correlation:** Do DEGs inside Inv4m correlate more or less
   strongly with DEGs outside the introgression than flanking DEGs do?
2. **Intra-regional correlation:** How tightly are DEGs co-regulated within
   each region (inv4m, flanking, outside)?
3. **Sliding window patterns:** How do correlation with outside DEGs and
   fold change magnitude vary along the physical map?

# Setup

```{r libraries}
library(here)
library(dplyr)
library(ggplot2)
library(zoo)
library(tidyr)

source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

```{r load-data}
effects_df <- read.csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv")
) %>%
  filter(predictor == "Inv4m")

nrow(effects_df)

effects_df$P <- effects_df$P.Value
effects_df$neglogP <- -log10(effects_df$P)

xp <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_leaf_trt.rds"))
```

```{r define-regions}
# Remap column names to match the original analysis
# DE script uses: in_Inv4m, in_cis (shared introgression), in_flank, in_trans
# Add convenience columns
effects_df <- effects_df %>%
  mutate(
    in_inv4m = in_Inv4m,
    in_flanking = in_flank,
    in_shared = in_cis,
    region = case_when(
      in_Inv4m ~ "inv4m",
      in_flank ~ "flanking",
      TRUE ~ "outside"
    ),
    is_significant = adj.P.Val < 0.05
  )

bonferroni_threshold <- 0.05 / nrow(effects_df)
```

# DEG lists by region

Identify Bonferroni-significant DEGs in each region, removing duplicated gene
names (e.g., JMJ paralogs mapped to the same locus label).

```{r deg-lists}
inv4_inv4 <- effects_df$gene[effects_df$P < bonferroni_threshold & effects_df$in_inv4m] %>%
  sort() %>% unique()

inv4_flanking <- effects_df$gene[effects_df$P < bonferroni_threshold & effects_df$in_flanking] %>%
  sort() %>% unique()

inv4_outside <- effects_df$gene[effects_df$P < bonferroni_threshold & !effects_df$in_shared] %>%
  sort() %>% unique()

inv4_introgression <- effects_df$gene[effects_df$P < bonferroni_threshold & effects_df$in_shared] %>%
  sort() %>% unique()

cat("DEGs by region (Bonferroni):\n")
cat("  Inv4m:", length(inv4_inv4), "\n")
cat("  Flanking:", length(inv4_flanking), "\n")
cat("  Outside:", length(inv4_outside), "\n")
cat("  Total introgression:", length(inv4_introgression), "\n")
```

# Inter-regional correlations (DEGs vs outside DEGs)

Compute pairwise expression correlations between outside DEGs and DEGs
inside Inv4m or in the flanking region. This tests whether genes in one
region of the introgression are more tightly co-regulated with the rest
of the genome than genes in the other region.

```{r inter-regional-correlations}
out_inv4m <- cor(t(xp[inv4_outside, ]), t(xp[inv4_inv4, ]))
out_flanking <- cor(t(xp[inv4_outside, ]), t(xp[inv4_flanking, ]))
out_introgression <- cor(t(xp[inv4_outside, ]), t(xp[inv4_introgression, ]))

cat("Correlation matrix dimensions:\n")
cat("  Outside x Inv4m:", paste(dim(out_inv4m), collapse = " x "), "\n")
cat("  Outside x Flanking:", paste(dim(out_flanking), collapse = " x "), "\n")
cat("  Outside x Introgression:", paste(dim(out_introgression), collapse = " x "), "\n")
```

```{r inter-regional-boxplot, fig.width=5, fig.height=5}
to_plot <- rbind(
  data.frame(region = "inv4m", cor = as.numeric(out_inv4m)),
  data.frame(region = "flanking", cor = as.numeric(out_flanking))
)
to_plot$R2 <- to_plot$cor^2

to_plot %>%
  ggplot(aes(x = region, y = R2)) +
  geom_boxplot() +
  ylab(expression(R^2 ~ "(squared correlation with outside DEGs)")) +
  ggpubr::theme_classic2(base_size = 16)
```

```{r inter-regional-tests}
cat("=== R2 ~ region (all correlations) ===\n")
lm(R2 ~ region, data = to_plot) %>% summary() %>% print()

cat("\n=== R2 ~ region (negative correlations only) ===\n")
lm(R2 ~ region, data = to_plot[to_plot$cor < 0, ]) %>% summary() %>% print()

cat("\n=== R2 ~ region (positive correlations only) ===\n")
lm(R2 ~ region, data = to_plot[to_plot$cor > 0, ]) %>% summary() %>% print()
```

# Intra-regional correlations

How tightly co-regulated are DEGs within each region? Compute pairwise
correlations within inv4m, flanking, and outside DEGs separately.

```{r intra-regional-correlations}
inner_inv4m <- cor(t(xp[inv4_inv4, ]))
inner_flanking <- cor(t(xp[inv4_flanking, ]))
inner_outside <- cor(t(xp[inv4_outside, ]))

to_plot <- rbind(
  data.frame(
    comparison = "inv4m\noutside",
    cor = out_inv4m[upper.tri(out_inv4m, diag = FALSE)]
  ),
  data.frame(
    comparison = "flanking\noutside",
    cor = out_flanking[upper.tri(out_flanking, diag = FALSE)]
  ),
  data.frame(
    comparison = "inv4m\ninv4m",
    cor = inner_inv4m[upper.tri(inner_inv4m, diag = FALSE)]
  ),
  data.frame(
    comparison = "flanking\nflanking",
    cor = inner_flanking[upper.tri(inner_flanking, diag = FALSE)]
  ),
  data.frame(
    comparison = "outside\noutside",
    cor = inner_outside[upper.tri(inner_outside, diag = FALSE)]
  )
)

to_plot$R2 <- to_plot$cor^2
```

```{r intra-regional-boxplot, fig.width=8, fig.height=5}
to_plot %>%
  mutate(comparison = factor(comparison)) %>%
  mutate(comparison = forcats::fct_reorder(comparison, -R2)) %>%
  ggplot(aes(x = comparison, y = R2)) +
  ggtitle("Correlation between Inv4m DEGs by region") +
  ylab(expression(R^2 ~ "(squared correlation)")) +
  geom_boxplot() +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none")
```

```{r intra-regional-test}
lm(R2 ~ comparison, data = to_plot) %>% summary()
```

# Sliding window: R² with outside DEGs along chromosome 4

For each 25-gene window along chromosome 4, compute the mean R² between
genes in that window and all outside DEGs. This shows how the strength
of trans-correlation varies along the introgression.

```{r sliding-window-r2}
in_out_cor <- t(out_introgression)
in_out_R2 <- in_out_cor^2

rollme <- cbind(
  data.frame(gene = unique(colnames(out_introgression))),
  as.data.frame(in_out_cor)
)

get_midelement <- function(vec) vec[ceiling(length(vec) / 2)]

r2 <- rollapply(
  rollme$gene,
  width = 25,
  function(x) mean(as.numeric(in_out_R2[x, ])),
  by.column = FALSE
)

mid_gene <- rollapply(
  rollme$gene,
  width = 25,
  function(x) get_midelement(x),
  by.column = FALSE
)

by_window <- data.frame(r2_win = r2, gene = mid_gene)
```

```{r define-inv4m-boundaries}
# Derive boundaries from gene positions in the effects data
introgression_limit <- c(
  min(effects_df$BP[effects_df$in_shared]),
  max(effects_df$BP[effects_df$in_shared])
) / 1e6

inversion_limit <- c(
  min(effects_df$BP[effects_df$in_inv4m]),
  max(effects_df$BP[effects_df$in_inv4m])
) / 1e6
```

```{r sliding-window-r2-plot, fig.width=8, fig.height=5}
to_plot <- effects_df %>%
  mutate(BP = BP / 1e6) %>%
  inner_join(by_window, by = "gene")

to_plot %>%
  ggplot(aes(x = BP, y = r2_win, col = region)) +
  geom_vline(
    xintercept = c(introgression_limit, inversion_limit),
    linetype = "dashed"
  ) +
  ylab(expression(R^2 ~ "with outside DEGs (25-gene window mean)")) +
  xlab("Chromosome 4 Position [Mb]") +
  geom_point() +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "top")
```

# Sliding window: fold change along chromosome 4

Separate sliding windows for upregulated and downregulated genes show how
the magnitude and direction of the Inv4m effect vary along the physical map.
This addresses whether expression changes are directionally biased (as in
Guerrero et al. 2016, Solanum introgression lines) or symmetrically dampened.

```{r sliding-window-fc-up}
up_genes <- effects_df %>%
  filter(logFC > 0, CHR == 4) %>%
  arrange(BP) %>%
  distinct(gene, .keep_all = TRUE)

fc_up <- rollapply(up_genes$logFC, FUN = mean, width = 10)
mid_up <- rollapply(up_genes$gene, width = 10,
                    function(x) get_midelement(x), by.column = FALSE)
by_window_up <- data.frame(foldchange = fc_up, gene = mid_up)

to_plot_up <- effects_df %>%
  filter(CHR == 4) %>%
  arrange(BP) %>%
  inner_join(by_window_up, by = "gene")
```

```{r sliding-window-fc-down}
down_genes <- effects_df %>%
  filter(logFC < 0, CHR == 4) %>%
  arrange(BP) %>%
  distinct(gene, .keep_all = TRUE)

fc_down <- rollapply(down_genes$logFC, FUN = mean, width = 10)
mid_down <- rollapply(down_genes$gene, width = 10,
                      function(x) get_midelement(x), by.column = FALSE)
by_window_down <- data.frame(foldchange = fc_down, gene = mid_down)

to_plot_down <- effects_df %>%
  filter(CHR == 4) %>%
  arrange(BP) %>%
  inner_join(by_window_down, by = "gene")
```

```{r sliding-window-fc-plot, fig.width=7, fig.height=4}
pal <- c("purple4", "plum", "gold")
names(pal) <- c("inv4m", "flanking", "outside")

logFC_label <- expression(log[2]("Fold change"))

effects_df %>%
  filter(CHR == 4, BP > 150e6, BP < 200e6) %>%
  ggplot(aes(x = BP / 1e6, y = logFC, col = region)) +
  annotate("rect", xmin = 150, xmax = 200, ymin = -2, ymax = 2,
           alpha = 0.5, fill = "grey75") +
  geom_point() +
  scale_color_manual(values = pal) +
  coord_cartesian(xlim = c(150, 200)) +
  geom_line(
    data = to_plot_up,
    aes(x = BP / 1e6, y = foldchange), inherit.aes = FALSE,
    col = "darkgreen", linewidth = 1
  ) +
  geom_line(
    data = to_plot_down,
    aes(x = BP / 1e6, y = foldchange), inherit.aes = FALSE,
    col = "darkred", linewidth = 1
  ) +
  ylab(logFC_label) +
  xlab("Chromosome 4 Position [Mb]") +
  annotate("text", x = c(151, 151), y = c(6.5, -5),
           label = c("Upregulated", "Downregulated"),
           color = c("darkgreen", "darkred"),
           fontface = "bold", size = 5, hjust = 0) +
  geom_hline(yintercept = 0) +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none")
```

# Boxplots: fold change and significance by region

## Setup

```{r shared-setup}
regulation_order <- c("Downregulated", "Upregulated")
names(regulation_order) <- c("-1", "1")

to_plot <- effects_df %>%
  filter(CHR == 4) %>%
  mutate(
    regulation = regulation_order[as.character(sign(logFC))],
    zone = case_when(
      in_flanking & BP < 180e6 ~ "left",
      in_flanking & BP > 180e6 ~ "right",
      in_inv4m ~ "inv4m",
      .default = "outside"
    ),
    region = factor(region, levels = c("inv4m", "flanking", "outside"))
  )

to_plot$zone <- factor(to_plot$zone,
                       levels = c("left", "inv4m", "right", "outside"))
```

## Without splitting by direction

All expressed genes on chromosome 4, no up/down split.

### logFC by region (inv4m, flanking, outside)

```{r boxplot-logfc-by-region-nosplit, fig.width=6, fig.height=5}
wilcox_region <- to_plot %>%
  rstatix::wilcox_test(logFC ~ region) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance()

wilcox_region

# Manual bracket positions:
# inv4m=1, flanking=2, outside=3
# Level 1: inv4m-flanking and flanking-outside at same height
# Level 2: inv4m-outside spanning above both
ymax <- max(to_plot$logFC, na.rm = TRUE)
level1 <- ymax * 1.05
level2 <- ymax * 1.20

wilcox_region$xmin <- c(1, 1, 2.1)
wilcox_region$xmax <- c(1.9, 3, 3)
wilcox_region$y.position <- c(level1, level2, level1)

p_logfc_region <- to_plot %>%
  ggplot(aes(x = region, y = logFC, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_region, hide.ns = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
  ggpubr::theme_classic2(base_size = 25) +
  theme(legend.position = "none")

p_logfc_region

saveRDS(p_logfc_region, file.path(paths$intermediate, "panel_logfc_by_region.rds"))
```

### logFC by zone (left, inv4m, right, outside)

```{r boxplot-logfc-by-zone-nosplit, fig.width=7, fig.height=5}
wilcox_zone <- to_plot %>%
  rstatix::wilcox_test(logFC ~ zone, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "zone", step.increase = 0.05)

wilcox_zone

to_plot %>%
  ggplot(aes(x = zone, y = logFC, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_zone, hide.ns = TRUE) +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none")
```

### -log10(P) by region (inv4m, flanking, outside)

```{r boxplot-neglogP-by-region-nosplit, fig.width=6, fig.height=5}
wilcox_nlp_region <- to_plot %>%
  rstatix::wilcox_test(neglogP ~ region, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "region", step.increase = 0.03)

wilcox_nlp_region

to_plot %>%
  ggplot(aes(x = region, y = neglogP, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_nlp_region, hide.ns = TRUE) +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none")
```

### -log10(P) by zone (left, inv4m, right, outside)

```{r boxplot-neglogP-by-zone-nosplit, fig.width=7, fig.height=5}
wilcox_nlp_zone <- to_plot %>%
  rstatix::wilcox_test(neglogP ~ zone, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "zone", step.increase = 0.03)

wilcox_nlp_zone

to_plot %>%
  ggplot(aes(x = zone, y = neglogP, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_nlp_zone, hide.ns = TRUE) +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none")
```

## Split by direction of regulation

Same panels but faceted by up/down.

### logFC by region

```{r boxplot-logfc-by-region, fig.width=8, fig.height=5}
wilcox_region_dir <- to_plot %>%
  group_by(regulation) %>%
  rstatix::wilcox_test(logFC ~ region, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "region", step.increase = 0.05)

wilcox_region_dir

to_plot %>%
  ggplot(aes(x = region, y = logFC, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_region_dir, hide.ns = TRUE) +
  facet_wrap(. ~ regulation, scales = "free_y") +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none", strip.background = element_blank())
```

### logFC by zone

```{r boxplot-logfc-by-zone, fig.width=8, fig.height=5}
wilcox_zone_dir <- to_plot %>%
  group_by(regulation) %>%
  rstatix::wilcox_test(logFC ~ zone, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "zone", step.increase = 0.05)

wilcox_zone_dir

to_plot %>%
  ggplot(aes(x = zone, y = logFC, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_zone_dir, hide.ns = TRUE) +
  facet_wrap(. ~ regulation, scales = "free_y") +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none", strip.background = element_blank())
```

### -log10(P) by region

```{r boxplot-neglogP-by-region, fig.width=8, fig.height=5}
wilcox_nlp_region_dir <- to_plot %>%
  group_by(regulation) %>%
  rstatix::wilcox_test(neglogP ~ region, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "region", step.increase = 0.03)

wilcox_nlp_region_dir

to_plot %>%
  ggplot(aes(x = region, y = neglogP, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_nlp_region_dir, hide.ns = TRUE) +
  facet_wrap(. ~ regulation) +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none", strip.background = element_blank())
```

### -log10(P) by zone

```{r boxplot-neglogP-by-zone, fig.width=8, fig.height=5}
wilcox_nlp_zone_dir <- to_plot %>%
  group_by(regulation) %>%
  rstatix::wilcox_test(neglogP ~ zone, ref.group = "inv4m") %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_xy_position(x = "zone", step.increase = 0.03)

to_plot %>%
  ggplot(aes(x = zone, y = neglogP, color = region)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(width = 0.25, fill = NA, color = "black", outliers = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  ggpubr::stat_pvalue_manual(wilcox_nlp_zone_dir, hide.ns = TRUE) +
  facet_wrap(. ~ regulation) +
  ggpubr::theme_classic2(base_size = 16) +
  theme(legend.position = "none", strip.background = element_blank())
```

# Summary

```{r region-counts}
with(effects_df %>% filter(is_significant), table(region))
```

# Session info

```{r session-info}
sessionInfo()
```