---
title: "PA2024 Spatial Analysis: SpATS Spatial Correction (preliminary)"
author: "Fausto Rodriguez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    fig_width: 12
    fig_height: 8
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

Preliminary spatial correction of PA2024 (PSU2024) field phenotypes, mirroring
the PA2022 pipeline (`Corrected_phenotype_analysis_PSU2022.Rmd`):

1. Load PSU2024 raw data and assemble physical X/Y from the 2×2 P_SQUARE layout.
2. Recode genotype (NIL_xxxx → CTRL / Inv4m / Unknown / Check) via a stub key.
3. Fit a single SpATS PSANOVA surface **per phenotype** with `Rep` and `P_LEVEL`
   as fixed factors.
4. Extract corrected plot-level values.
5. Second-stage CTRL vs Inv4m t-test (FDR) + Figure 2-style boxplot panel.

The field is a 2×2 of P_SQUAREs: P_SQ 1 = bottom-left (LowP), 2 = bottom-right
(HighP), 3 = top-left (LowP), 4 = top-right (HighP). Each P_SQUARE is 10 rows ×
24 cols, so the assembled field is 20 rows × 48 cols. RS 2001 sits at (X=1, Y=1)
and RS 2960 at (X=48, Y=20).

## 1. Libraries

```{r load-libraries}
library(tidyverse)
library(ggtext)
library(ggfx)
library(rstatix)
library(ggpubr)
library(SpATS)
library(statgenHTP)
library(ggbeeswarm)
library(cowplot)
```

## 2. Load and prepare PSU2024 data

```{r load-data}
DATA_DIR <- paths$data
plant_csv <- file.path(DATA_DIR, "PSU2024.csv")

field_raw <- read_csv(plant_csv, show_col_types = FALSE)

## Assembled X/Y physical coords (2x2 layout)
n_col_psq <- max(field_raw$COLUMN)
n_row_psq <- max(field_raw$ROW)
psq_x_offset <- c(`1` = 0L, `3` = 0L, `2` = n_col_psq, `4` = n_col_psq)
psq_y_offset <- c(`1` = 0L, `2` = 0L, `3` = n_row_psq, `4` = n_row_psq)

field_data <- field_raw %>%
  rename(rowid = RS) %>%
  mutate(
    X = psq_x_offset[as.character(P_SQUARE)] + COLUMN,
    Y = psq_y_offset[as.character(P_SQUARE)] + ROW,
    Treatment = factor(P_LEVEL, levels = c("HIGH_P", "LOW_P")),
    Rep      = as.factor(REP),
    Genotype = as.character(GENOTYPE)
  ) %>%
  mutate(Treatment = fct_recode(Treatment, "+P" = "HIGH_P", "-P" = "LOW_P"))

cat("Field data dims:", dim(field_data), "\n")
cat("X range:", range(field_data$X), "  Y range:", range(field_data$Y), "\n")
```

## 3. Genotype classification (Check / NIL population)

The PSU2024 trial has two biological tiers:

- **Checks (fixed effects in SpATS via `addCheck`):** every entry that is *not*
  part of the segregating NIL mapping population. That is `B73`, `Inv4_B73`,
  `Inv4_MEX`, `Inv4_Mi21`, `Inv4_PT`, `Oh43`, `Z031E0591`, `Z031E0594` — eight
  entries total.
- **NIL population (random effect):** the 41 `NIL_xxxx` segregating lines.
  Residual / BLUP normality applies *within* this cohort.
- **Nuisance fixed effects:** `Rep`, `P_LEVEL`, spatial smooth on assembled X/Y.

The **focal contrast** is `Inv4_Mi21` vs `Inv4_B73`. Both are checks (fixed
effects), so their BLUEs and the difference SE come straight out of the check
coefficient block.

For this paper, the contrast of interest is **CTRL = `Inv4_B73`** vs
**Inv4m = `Inv4_Mi21`**. Each entry has a `donor` (the source of the Inv4m
introgression at chromosome 4 — `B73`/`Mi21`/`MEX`/`PT`) that is parsed from
the GENOTYPE label where possible.

```{r genotype-classify}
check_set <- c("B73", "Inv4_B73", "Inv4_MEX", "Inv4_Mi21", "Inv4_PT",
               "Oh43", "Z031E0591", "Z031E0594")
focal_pair <- c("Inv4_B73", "Inv4_Mi21")

donor_lookup <- c(
  "B73"        = "B73",
  "Inv4_B73"   = "B73",
  "Inv4_MEX"   = "MEX",
  "Inv4_Mi21"  = "Mi21",
  "Inv4_PT"    = "PT",
  "Oh43"       = "Oh43",
  "Z031E0591"  = "hybrid_check",
  "Z031E0594"  = "hybrid_check"
)
## NIL_xxxx donor not encoded in the genotype code — flagged as "NIL_pop" for now
## (user: replace if/when an external donor key becomes available)

field_data <- field_data %>%
  mutate(
    geno_class = if_else(Genotype %in% check_set, "Check", "NILpop"),
    donor      = unname(donor_lookup[Genotype]),
    donor      = if_else(is.na(donor), "NIL_pop", donor),
    geno_label = case_when(
      Genotype == "Inv4_B73"  ~ "CTRL",
      Genotype == "Inv4_Mi21" ~ "Inv4m",
      TRUE                    ~ NA_character_
    ),
    Genotype   = factor(Genotype)
  )

cat("\n--- genotype tier summary ---\n")
print(field_data %>% count(geno_class))
print(field_data %>% filter(geno_class == "Check") %>%
        count(Genotype, donor))
cat("\nFocal contrast: ",
    "CTRL = ", focal_pair[1], ", Inv4m = ", focal_pair[2], "\n", sep = "")
```

## 4. Identify usable phenotypes (<50% missing)

```{r usable-traits}
base_phenotypes <- c("PH", "DTA", "DTS", "SDW_11wk", "SDW_H")
ear_phenotypes  <- c("TEW", "TEN", "TGW", "GWE", "HGW", "TGN", "AMF")
ndvi_phenotypes <- c("NDVI_T1", "NDVI_T2")
all_phenotypes  <- intersect(
  unique(c(base_phenotypes, ear_phenotypes, ndvi_phenotypes)),
  names(field_data)
)

## Strip stray spaces (e.g. "SDW_11wk ")
fix_names <- setNames(names(field_data), trimws(names(field_data)))
if (!"SDW_11wk" %in% names(field_data) &&
    "SDW_11wk " %in% names(field_data)) {
  field_data <- field_data %>% rename(SDW_11wk = `SDW_11wk `)
}

missing_summary <- field_data %>%
  select(all_of(all_phenotypes)) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "phenotype",
               values_to = "missing_count") %>%
  mutate(total_obs = nrow(field_data),
         missing_pct = round(missing_count / total_obs * 100, 1),
         available_n = total_obs - missing_count) %>%
  arrange(missing_pct)

print(missing_summary)

usable_phenotypes <- missing_summary %>%
  filter(missing_pct < 50) %>% pull(phenotype)

cat("\nUsable phenotypes (<50% missing):",
    paste(usable_phenotypes, collapse = ", "), "\n")
```

## 5. Build TimePoints object for statgenHTP

```{r build-timepoints}
environment_name <- "PSU2024"
single_time <- field_data %>%
  filter(!is.na(X), !is.na(Y)) %>%
  mutate(time = as.POSIXct("2024-05-22 10:00:00"))

tp_data <- createTimePoints(
  dat = single_time,
  experimentName = environment_name,
  genotype = "Genotype",
  timePoint = "time",
  plotId = "rowid",
  repId = "Rep",
  rowNum = "Y",
  colNum = "X",
  checkGenotypes = check_set,
  addCheck = TRUE
)

summary(tp_data)
```

## 6. Fit SpATS per phenotype

```{r fit-spats, results='hide'}
corrected_list <- list()
fit_info <- list()
all_models <- list()

for (phen in usable_phenotypes) {
  message("Fitting SpATS model for: ", phen)
  no_nas <- single_time %>% filter(!is.na(.data[[phen]]))
  if (nrow(no_nas) < 10) {
    message("Insufficient data for ", phen, " (n=", nrow(no_nas), ")")
    next
  }
  fit_try <- try(
    fitModels(
      TP = tp_data,
      trait = phen,
      extraFixedFactors = c("repId", "Treatment"),
      timePoints = 1,
      what = "random"   # NIL_xxxx random; checks fixed via addCheck
    ),
    silent = TRUE
  )
  if (inherits(fit_try, "try-error")) {
    fit_info[[phen]] <- list(n_obs = nrow(no_nas), status = "failed")
    next
  }
  all_models[[phen]] <- fit_try
  corr <- getCorrected(fit_try) %>%
    rename(!!phen := !!phen)
  corrected_list[[phen]] <- corr %>%
    select(plotId, genotype, !!phen)
  fit_info[[phen]] <- list(n_obs = nrow(no_nas), status = "ok")
}

sp_corrected <- Reduce(function(a, b)
                       full_join(a, b, by = c("plotId", "genotype")),
                       corrected_list)

## Attach Treatment, geno_class, donor, geno_label back on
sp_corrected <- sp_corrected %>%
  mutate(plotId = as.character(plotId)) %>%
  left_join(field_data %>%
              mutate(rowid = as.character(rowid)) %>%
              select(rowid, Treatment, geno_class, donor, geno_label),
            by = c("plotId" = "rowid"))

cat("Plots with corrected data:", nrow(sp_corrected), "\n")
cat("Traits with corrected values:",
    paste(intersect(usable_phenotypes, names(sp_corrected)), collapse = ", "),
    "\n")
```

## 7. SpATS diagnostics (PH, DTA, DTS)

```{r spats-PH}
if ("PH" %in% names(all_models)) {
  plot(all_models$PH, timePoints = 1)
  plot(all_models$PH, timePoints = 1, plotType = "spatial", spaTrend = "raw")
}
```

```{r spats-DTA}
if ("DTA" %in% names(all_models)) {
  plot(all_models$DTA, timePoints = 1)
  plot(all_models$DTA, timePoints = 1, plotType = "spatial", spaTrend = "raw")
}
```

```{r spats-DTS}
if ("DTS" %in% names(all_models)) {
  plot(all_models$DTS, timePoints = 1)
  plot(all_models$DTS, timePoints = 1, plotType = "spatial", spaTrend = "raw")
}
```

## 8. Genotype BLUEs from check fixed effects

For each fitted model we pull the check-genotype BLUEs (their fixed-effect
coefficients) plus the NIL_xxxx BLUPs via `getGenoPred()`. The focal contrast
`Inv4_Mi21 − Inv4_B73` and its SE come straight from the check block of the
variance-covariance matrix.

```{r genotype-blues}
get_focal_contrast <- function(model_list, focal_pair) {
  out <- list()
  for (trait in names(model_list)) {
    pred <- try(getGenoPred(model_list[[trait]]), silent = TRUE)
    if (inherits(pred, "try-error")) next
    pred_df <- pred$genoPred
    if (!all(focal_pair %in% pred_df$genotype)) next

    sub <- pred_df %>%
      filter(genotype %in% focal_pair) %>%
      select(genotype, predicted.values, standard.errors)
    if (nrow(sub) != 2) next

    ctrl  <- sub %>% filter(genotype == focal_pair[1])
    inv4m <- sub %>% filter(genotype == focal_pair[2])
    diff  <- inv4m$predicted.values - ctrl$predicted.values
    se_diff <- sqrt(inv4m$standard.errors^2 + ctrl$standard.errors^2)
    z_stat <- diff / se_diff
    p_z    <- 2 * pnorm(-abs(z_stat))

    out[[trait]] <- tibble(
      trait = trait,
      BLUE_CTRL      = ctrl$predicted.values,
      SE_CTRL        = ctrl$standard.errors,
      BLUE_Inv4m     = inv4m$predicted.values,
      SE_Inv4m       = inv4m$standard.errors,
      effect         = diff,
      se_diff        = se_diff,
      z              = z_stat,
      p              = p_z
    )
  }
  bind_rows(out)
}

contrast_df <- get_focal_contrast(all_models, focal_pair) %>%
  mutate(CTRL  = focal_pair[1],
         Inv4m = focal_pair[2],
         donor_CTRL  = donor_lookup[focal_pair[1]],
         donor_Inv4m = donor_lookup[focal_pair[2]],
         p_adj = p.adjust(p, method = "fdr"),
         signif = case_when(
           p_adj < 0.001 ~ "***",
           p_adj < 0.01  ~ "**",
           p_adj < 0.05  ~ "*",
           TRUE          ~ "ns"
         )) %>%
  arrange(p_adj)

print(contrast_df)
```

## 9. Figure 2-style panel: per-plot boxplots for the focal pair

Show the spatially-corrected plot-level values for `Inv4_Mi21` vs `Inv4_B73`
(20 plots each) per trait, with the BLUE difference overlaid as the bracket
label.

```{r main-panel, fig.width=10, fig.height=7}
focal_corrected <- sp_corrected %>%
  filter(genotype %in% focal_pair) %>%
  mutate(genotype = factor(geno_label, levels = c("CTRL", "Inv4m")))

pheno_theme_panel <- theme_classic2(base_size = 20) +
  theme(
    plot.title = element_blank(),
    axis.title.y = element_text(size = 18),
    axis.title.x = element_blank(),
    axis.text.x = element_markdown(size = 18, face = "bold", color = "black"),
    axis.text.y = element_text(size = 16),
    strip.background = element_blank(),
    strip.text = element_text(size = 20, face = "bold"),
    legend.position = "none"
  )

make_trait_plot <- function(trait_name, data, contrast_tbl, units) {
  d <- filter(data, trait == trait_name)
  cr <- contrast_tbl %>% filter(trait == trait_name)
  y_label <- units[trait_name]
  star <- if (nrow(cr)) cr$signif else "ns"

  ymax <- max(d$value, na.rm = TRUE)
  yrng <- diff(range(d$value, na.rm = TRUE))

  p <- ggplot(d, aes(x = genotype, y = value, color = genotype)) +
    geom_boxplot(width = 0.25, linewidth = 1.5, alpha = 0) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    geom_quasirandom(size = 2) %>%
      with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
    scale_color_manual(values = c("CTRL" = "gold", "Inv4m" = "purple4")) +
    scale_x_discrete(labels = c("CTRL" = "CTRL", "Inv4m" = "*Inv4m*")) +
    annotate("text", x = 1.5, y = ymax + 0.10 * yrng, label = star,
             size = 8, fontface = "bold") +
    labs(title = trait_name, y = y_label, x = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.20))) +
    pheno_theme_panel +
    theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5))
  if (trait_name %in% c("DTA", "DTS")) {
    y_range <- range(d$value, na.rm = TRUE)
    p <- p + scale_y_continuous(
      breaks = seq(floor(y_range[1]), ceiling(y_range[2]), by = 1),
      expand = expansion(mult = c(0.1, 0.20)))
  }
  p
}

main_traits <- intersect(c("PH", "DTA", "DTS", "SDW_H", "GWE", "HGW"),
                         usable_phenotypes)
trait_units <- c(PH = "cm", DTA = "Days", DTS = "Days",
                 SDW_H = "g", GWE = "g", HGW = "g", TGW = "g",
                 SDW_11wk = "g", TEW = "g")

plot_data_main <- focal_corrected %>%
  pivot_longer(cols = any_of(usable_phenotypes),
               names_to = "trait", values_to = "value") %>%
  filter(trait %in% main_traits, !is.na(value)) %>%
  mutate(trait = factor(trait, levels = main_traits))

trait_plots <- lapply(main_traits, make_trait_plot,
                      data = plot_data_main, contrast_tbl = contrast_df,
                      units = trait_units)

main_panel <- cowplot::plot_grid(plotlist = trait_plots,
                                 nrow = 2, ncol = 3,
                                 align = "hv", axis = "tblr")

print(main_panel)

ggsave(file.path(paths$figures, "PSU2024_field_phenotypes_panel.png"),
       main_panel, width = 10, height = 7, dpi = 300)
cat("Saved panel to:",
    file.path(paths$figures, "PSU2024_field_phenotypes_panel.png"), "\n")
```

## 10. Cohen's d forest plot + export

To compare effects across traits on a common scale, we standardize each BLUE
difference by the **pooled within-genotype SD** of the spatially corrected
plot-level values for Inv4_B73 and Inv4_Mi21 (20 plots each). With n=20 per
group, we also report Hedges' g (small-sample correction).

```{r cohens-d}
focal_long <- sp_corrected %>%
  filter(genotype %in% focal_pair) %>%
  pivot_longer(cols = any_of(usable_phenotypes),
               names_to = "trait", values_to = "value") %>%
  filter(!is.na(value))

pooled_sd_df <- focal_long %>%
  group_by(trait, genotype) %>%
  summarise(sd_g = sd(value), n_g = n(), .groups = "drop") %>%
  group_by(trait) %>%
  summarise(pooled_sd = sqrt(sum((n_g - 1) * sd_g^2) / sum(n_g - 1)),
            n_total   = sum(n_g),
            .groups = "drop")

contrast_df <- contrast_df %>%
  left_join(pooled_sd_df, by = "trait") %>%
  mutate(cohens_d  = effect / pooled_sd,
         se_d      = se_diff / pooled_sd,
         hedges_g  = cohens_d * (1 - 3 / (4 * n_total - 9)))

print(contrast_df %>%
        select(trait, effect, se_diff, pooled_sd,
               cohens_d, se_d, hedges_g, p_adj, signif))
```

```{r forest-plot, fig.width=7, fig.height=5}
forest_df <- contrast_df %>%
  arrange(cohens_d) %>%
  mutate(trait = factor(trait, levels = trait))

p_forest <- ggplot(forest_df,
                   aes(x = cohens_d, y = trait,
                       xmin = cohens_d - 1.96 * se_d,
                       xmax = cohens_d + 1.96 * se_d)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-0.8, -0.5, -0.2, 0.2, 0.5, 0.8),
             linetype = "dotted", color = "grey80") +
  geom_pointrange(size = 0.8, color = "purple4") +
  geom_text(aes(label = signif), nudge_y = 0.25, size = 5,
            fontface = "bold") +
  labs(x = "Cohen's d (Inv4m − CTRL)  ±95% CI",
       y = NULL,
       title = "PA2024 — focal contrast (standardized effect)") +
  theme_classic2(base_size = 14)

print(p_forest)

ggsave(file.path(paths$figures, "PSU2024_focal_contrast_forest.png"),
       p_forest, width = 7, height = 5, dpi = 300)

write_csv(contrast_df,
          file.path(paths$intermediate,
                    paste0(environment_name, "_focal_contrast_BLUEs.csv")))
write_csv(sp_corrected,
          file.path(paths$intermediate,
                    paste0(environment_name,
                           "_spatially_corrected_phenotypes.csv")))

cat("Outputs:\n  ",
    file.path(paths$intermediate,
              paste0(environment_name, "_focal_contrast_BLUEs.csv")), "\n  ",
    file.path(paths$intermediate,
              paste0(environment_name, "_spatially_corrected_phenotypes.csv")),
    "\n  ",
    file.path(paths$figures, "PSU2024_focal_contrast_forest.png"), "\n  ",
    file.path(paths$figures, "PSU2024_field_phenotypes_panel.png"), "\n")
```

## 11. Notes / next steps

- Model design:
  - `Genotype` random over the 41 `NIL_xxxx` lines (residual normality applies here)
  - 8 non-NIL entries (`B73`, `Inv4_B73`, `Inv4_MEX`, `Inv4_Mi21`, `Inv4_PT`,
    `Oh43`, `Z031E0591/4`) added as checks (fixed via `addCheck = TRUE`)
  - `Rep` and `Treatment` (P_LEVEL) as fixed nuisance terms
  - Spatial smooth on assembled X (1–48) and Y (1–20)
- The focal contrast `Inv4_Mi21 − Inv4_B73` is extracted from the check BLUE
  block. SE-of-difference is computed as `sqrt(SE_A^2 + SE_B^2)` (independent
  approximation; replace with full Var-Cov if check-pair covariance becomes a
  concern).
- Compare these BLUE-based corrected values against Sandra's `_SPAT` columns
  as a methods-parity check before finalizing the Figure 2 hybrid panel.
