---
title: "Floral Candidate Gene Expression: mads69, ub2, ub3, rap2.7"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.width = 12,
 fig.height = 8
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This notebook checks for Inv4m effects on floral transition candidate genes
highlighted by Dong et al. 2026 (Fang Yang's group, Genome Biology):

- **mads69** (Zm00001eb143080) — floral promoter; mutants have smaller meristems
- **ub2** (Zm00001eb035030) — promotes floral transition; double mutants have shorter meristems
- **ub3** (Zm00001eb199880) — promotes floral transition (with ub2)
- **rap2.7** (Zm00001eb355240) — floral repressor; mutants have larger meristems

We examine expression across:

1. **Crow et al. 2020** growth chamber data — all tissues including SAM, split by donor (Mi21/PT) and temperature (warm/cold)
2. **PA2022** field experiment — leaf tissue

## Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggfx)
library(ggbeeswarm)
library(limma)
library(here)
```

## Define Candidate Genes

```{r candidate_genes}
candidate_genes <- tibble(
  gene = c("Zm00001eb143080", "Zm00001eb035030", "Zm00001eb199880", "Zm00001eb355240"),
  symbol = c("mads69", "ub2", "ub3", "rap2.7"),
  description = c(
    "MADS-transcription factor 69 (floral promoter)",
    "Unbranched2 (floral transition / SPL)",
    "Unbranched3 (floral transition / SPL)",
    "RAP2.7 orthologue (floral repressor)"
  )
)

cat("Candidate genes:\n")
print(candidate_genes)
```

# Load PA2022 Data

```{r load_psu2022}
expr_psu <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_leaf_trt.rds"))
meta_psu <- read.csv(file.path(paths$data, "PSU2022_metadata.csv"))

{
 cat("=== PA2022 DATA ===\n")
 cat("Expression matrix:", nrow(expr_psu), "genes x", ncol(expr_psu), "samples\n")
 cat("Metadata samples:", nrow(meta_psu), "\n")
}
```

```{r extract_psu2022}
genes_in_psu <- candidate_genes$gene[candidate_genes$gene %in% rownames(expr_psu)]
cat("Candidate genes found in PA2022:", length(genes_in_psu), "of", nrow(candidate_genes), "\n")
cat("Missing:", setdiff(candidate_genes$gene, genes_in_psu), "\n")

expr_psu_long <- expr_psu[genes_in_psu, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(cols = -gene, names_to = "sample", values_to = "logCPM") %>%
  left_join(
    meta_psu %>%
      dplyr::select(library, Genotype, leaf_tissue, Treatment) %>%
      mutate(sample = library),
    by = "sample"
  ) %>%
  filter(!is.na(Genotype)) %>%
  left_join(candidate_genes, by = "gene") %>%
  mutate(
    experiment = "PA2022 (Field)",
    tissue = paste0("Leaf ", leaf_tissue),
    genotype = ifelse(Genotype == "CTRL", "B73", "Inv4m"),
    donor = NA_character_,
    temp = NA_character_
  )

cat("PA2022 expression records:", nrow(expr_psu_long), "\n")
```

# Load Crow2020 Data

```{r load_crow2020}
expr_crow <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_crow2020.rds"))

# Load and join metadata to get Donor information
sra_meta <- read.csv(file.path(paths$data, "SraRunTable.csv"), check.names = FALSE)
sub_meta <- read.csv(file.path(paths$data, "gc7_sample_submission.csv"), stringsAsFactors = FALSE)

meta_crow <- sub_meta %>%
  filter(old_line %in% c("PT_NIL", "Mi21_NIL")) %>%
  dplyr::select(samp, old_line) %>%
  inner_join(
    sra_meta %>% dplyr::rename(samp = `Library Name`),
    by = "samp"
  ) %>%
  mutate(
    sample = Run,
    donor = case_when(
      grepl("^PT_", old_line) ~ "PT",
      grepl("^Mi21_", old_line) ~ "Mi21"
    ),
    genotype = ifelse(genotype == "B73", "B73", "Inv4m"),
    tissue_clean = case_when(
      tissue == "V1_Leaf_tissue"    ~ "V1 Leaf",
      tissue == "V1_Primary_root"   ~ "V1 Root",
      tissue == "V3_Leaf_tissue"    ~ "V3 Leaf",
      tissue == "V3_leaf_base"      ~ "V3 Leaf Base",
      tissue == "V3_leaf_sheath"    ~ "V3 Leaf Sheath",
      tissue == "V3_S2_leaf_base"   ~ "V3 S2 Leaf Base",
      tissue == "V3_S2_leaf_tip"    ~ "V3 S2 Leaf Tip",
      tissue == "V3_Primary_root"   ~ "V3 Root",
      tissue == "V3_Stem_SAM"       ~ "V3 SAM",
      TRUE ~ tissue
    )
  )

{
 cat("=== CROW2020 DATA ===\n")
 cat("Expression matrix:", nrow(expr_crow), "genes x", ncol(expr_crow), "samples\n")
 cat("\nSamples with donor info:", nrow(meta_crow), "\n")
 cat("\nDonor x Genotype:\n")
 print(table(meta_crow$donor, meta_crow$genotype))
 cat("\nTissue x Temp:\n")
 print(table(meta_crow$tissue_clean, meta_crow$temp))
}
```

```{r extract_crow2020}
genes_in_crow <- candidate_genes$gene[candidate_genes$gene %in% rownames(expr_crow)]
cat("Candidate genes found in Crow2020:", length(genes_in_crow), "of", nrow(candidate_genes), "\n")
cat("Missing:", setdiff(candidate_genes$gene, genes_in_crow), "\n")

expr_crow_long <- expr_crow[genes_in_crow, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(cols = -gene, names_to = "sample", values_to = "logCPM") %>%
  inner_join(
    meta_crow %>% dplyr::select(sample, genotype, tissue_clean, temp, donor),
    by = "sample"
  ) %>%
  left_join(candidate_genes, by = "gene") %>%
  mutate(
    experiment = "Crow et al., 2020 (Growth Chamber)",
    tissue = tissue_clean
  )

cat("Crow2020 expression records:", nrow(expr_crow_long), "\n")
```

# Crow2020 DEG Results: Inv4m x Tissue Interactions

Extract the Inv4m effect coefficients from the Crow2020 limma model for these genes.

```{r crow2020_deg_check}
# Load the voom object and refit the model
voom_obj <- readRDS(file.path(paths$intermediate, "normalized_expression_voom_object_crow2020.rds"))

design <- voom_obj$design
fit <- lmFit(voom_obj, design)
ebfit <- eBayes(fit)

# Coefficients of interest for Inv4m effects
inv4m_coefs <- c(
  "GenotypeInv4m"                   = "Inv4m (main)",
  "TissueV3_SAM:GenotypeInv4m"     = "V3_SAM x Inv4m",
  "GenotypeInv4m:tempcold"          = "Inv4m x cold"
)

# Extract results for our candidate genes
results_list <- list()
for (i in seq_along(inv4m_coefs)) {
  coef_name <- names(inv4m_coefs)[i]
  coef_idx <- which(colnames(coef(ebfit)) == coef_name)
  if (length(coef_idx) == 0) {
    cat("WARNING: coefficient", coef_name, "not found\n")
    next
  }

  tt <- topTable(ebfit, coef = coef_idx, sort.by = "none", n = Inf)
  tt <- tt[genes_in_crow, , drop = FALSE]
  tt$coefficient <- inv4m_coefs[i]
  tt$gene <- rownames(tt)
  results_list[[i]] <- tt
}

crow_effects <- bind_rows(results_list) %>%
  left_join(candidate_genes, by = "gene") %>%
  dplyr::select(symbol, coefficient, logFC, P.Value, adj.P.Val) %>%
  as_tibble() %>%
  mutate(
    sig = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.05  ~ "*",
      P.Value < 0.05    ~ "(nominal)",
      TRUE ~ "ns"
    )
  )

cat("=== Crow2020 Inv4m Effects on Candidate Genes ===\n\n")
crow_effects %>% arrange(symbol, coefficient) %>% print(n = 20)
```

# Expression Plots

## Crow2020: SAM Samples by Donor and Temperature

```{r crow_sam_plot, fig.width=12, fig.height=8}
genotype_colors <- c("B73" = "gold", "Inv4m" = "#4a0f82")

# Filter to SAM samples only
sam_data <- expr_crow_long %>%
  filter(tissue == "V3 SAM")

cat("SAM samples:", nrow(sam_data) / length(genes_in_crow), "\n")
cat("By donor x genotype x temp:\n")
print(table(sam_data$donor, sam_data$genotype, sam_data$temp) / length(genes_in_crow))

p_sam <- ggplot(
  sam_data,
  aes(x = interaction(temp, donor, sep = "\n"), y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.6,
    linewidth = 0.8,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) +
  geom_point(
    size = 2,
    alpha = 0.8,
    position = position_dodge(width = 0.7)
  ) +
  facet_wrap(~ symbol, scales = "free_y", nrow = 1) +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(color = "black", face = "bold", size = 12),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold.italic", size = 14),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "top",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    title = "Crow et al. 2020 — V3 SAM samples",
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_sam)
```

## All Tissues: Crow2020 + PA2022

Faceted by gene (rows) and experiment (columns), matching JMJ plot style.
Crow2020 data includes donor as point shape.

```{r combined_plot, fig.width=16, fig.height=10}
# Combine both experiments
combined <- bind_rows(
  expr_psu_long %>%
    dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment, donor),
  expr_crow_long %>%
    dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment, donor)
) %>%
  mutate(
    symbol = factor(symbol, levels = c("mads69", "ub2", "ub3", "rap2.7")),
    genotype = factor(genotype, levels = c("B73", "Inv4m")),
    experiment = factor(experiment, levels = c(
      "Crow et al., 2020 (Growth Chamber)", "PA2022 (Field)"
    ))
  )

# Drop ub3 if not in data (filtered out due to low expression)
genes_present <- unique(combined$symbol[!is.na(combined$symbol)])
cat("Genes present in combined data:", paste(genes_present, collapse = ", "), "\n")

p_combined <- ggplot(
  combined,
  aes(x = tissue, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    aes(shape = donor),
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_grid(symbol ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  scale_shape_manual(
    values = c("Mi21" = 16, "PT" = 17),
    name = "Donor",
    na.value = 4
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 30, hjust = 1,
      color = "black", face = "bold", size = 12
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text.x = element_text(face = "bold", size = 14),
    strip.text.y = element_text(face = "bold.italic", size = 14),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "top",
    legend.text = element_text(face = "bold", size = 11),
    legend.title = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_combined)
```

## Crow2020: SAM Only, Split by Donor

```{r sam_by_donor, fig.width=14, fig.height=8}
sam_donor <- expr_crow_long %>%
  filter(tissue == "V3 SAM") %>%
  mutate(
    symbol = factor(symbol, levels = c("mads69", "ub2", "ub3", "rap2.7")),
    genotype = factor(genotype, levels = c("B73", "Inv4m")),
    donor = factor(donor, levels = c("Mi21", "PT")),
    temp = factor(temp, levels = c("warm", "cold"))
  )

p_sam_donor <- ggplot(
  sam_donor,
  aes(x = temp, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 0.8,
    outlier.shape = NA,
    position = position_dodge(width = 0.6)
  ) +
  geom_point(
    size = 2.5,
    alpha = 0.8,
    position = position_dodge(width = 0.6)
  ) +
  facet_grid(symbol ~ donor, scales = "free_y") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(color = "black", face = "bold", size = 13),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text.x = element_text(face = "bold", size = 14),
    strip.text.y = element_text(face = "bold.italic", size = 14),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "top",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    title = "Crow et al. 2020 — V3 SAM, by Donor",
    x = "Temperature",
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_sam_donor)
```

# Summary Statistics

```{r summary_stats}
# SAM-specific summary
sam_summary <- expr_crow_long %>%
  filter(tissue == "V3 SAM") %>%
  group_by(symbol, donor, temp, genotype) %>%
  summarise(
    mean_logCPM = mean(logCPM, na.rm = TRUE),
    sd_logCPM = sd(logCPM, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

cat("=== SAM Expression Summary (Crow2020) ===\n\n")
print(sam_summary %>% arrange(symbol, donor, temp, genotype), n = 40)

# Overall summary across all tissues
overall_summary <- combined %>%
  group_by(experiment, tissue, symbol, genotype) %>%
  summarise(
    mean_logCPM = mean(logCPM, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = genotype,
    values_from = c(mean_logCPM, n),
    names_sep = "_"
  ) %>%
  mutate(log2FC = mean_logCPM_Inv4m - mean_logCPM_B73)

cat("\n=== Log2FC Summary (Inv4m vs B73) ===\n\n")
print(overall_summary %>%
  dplyr::select(experiment, tissue, symbol, log2FC) %>%
  arrange(symbol, experiment, tissue),
  n = 60
)
```

# Save Outputs

```{r save_outputs}
# Save summary tables
write_csv(
  sam_summary,
  file.path(paths$intermediate, "floral_candidates_SAM_expression_summary.csv")
)

write_csv(
  crow_effects,
  file.path(paths$intermediate, "floral_candidates_crow2020_inv4m_effects.csv")
)

# Save plots
ggsave(
  file.path(paths$figures, "floral_candidates_SAM_by_donor.pdf"),
  p_sam_donor,
  width = 14, height = 8
)

ggsave(
  file.path(paths$figures, "floral_candidates_all_tissues.pdf"),
  p_combined,
  width = 16, height = 10
)

{
  cat("=== OUTPUTS SAVED ===\n")
  cat("  floral_candidates_SAM_expression_summary.csv\n")
  cat("  floral_candidates_crow2020_inv4m_effects.csv\n")
  cat("  floral_candidates_SAM_by_donor.pdf\n")
  cat("  floral_candidates_all_tissues.pdf\n")
}
```

# Session Info

```{r session_info}
sessionInfo()
```