---
title: "GO Enrichment Analysis: Inv4m Trans Co-expression Network"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Perform Gene Ontology enrichment analysis on the validated 
trans-regulatory network to identify biological processes associated with genes 
in trans-regulatory networks controlled by different genomic regions.

**Approach**:
1. Load trans network node lists from MaizeNetome analysis
2. Define gene sets by network sub-structure (flanking-regulated vs Inv4m-regulated)
3. Perform GO enrichment separately for each network subset
4. Compare enrichment across network regions
5. Validate with Top DEGs subset (|log2FC| > 2)

**Expected outcomes**:
- Identification of enriched biological processes in trans-regulatory networks
- Comparison of functional categories between flanking and Inv4m networks
- Visualization of GO terms with representative gene examples

---

# Setup and Dependencies

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(clusterProfiler)
library(forcats)
library(DOSE)
```

---

# Load Data

## Trans Co-expression Network Node Lists

**Purpose**: Load gene lists defining trans-regulatory network substructures.

```{r load_network_nodes}
trans_regulators <- read.csv(
  file.path(paths$intermediate, "trans_network_regulators.csv")
)
trans_targets <- read.csv(
  file.path(paths$intermediate, "trans_network_targets.csv")
)
trans_network_all <- read.csv(
  file.path(paths$intermediate, "trans_network_all_genes.csv")
)
trans_edges <- read.csv(
  file.path(paths$intermediate, "trans_network_edges_annotated.csv")
)

cat("Network composition:\n")
cat("  Regulators:", nrow(trans_regulators), "\n")
cat("  Targets:", nrow(trans_targets), "\n")
cat("  Total genes:", nrow(trans_network_all), "\n")
cat("  Edges:", nrow(trans_edges), "\n")
```

## Load Effects Table for Background

**Purpose**: Define universe of analyzed genes as background for enrichment.

```{r load_effects}
effects_df <- read.csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv")
) %>%
  filter(predictor == "Inv4m") %>%
  group_by(gene) %>%
  arrange(adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup()

background <- effects_df %>%
  pull(gene) %>%
  sort() %>%
  unique()

cat("Background genes:", length(background), "\n")
cat("DEGs:", sum(effects_df$is_DEG), "\n")
cat("High confidence DEGs:", sum(effects_df$is_hiconf_DEG), "\n")
```

## Load GO Annotation

**Purpose**: Load maize B73 NAM5 GO annotations from GOMAP.

```{r load_go_annotation}
TERM2NAME <- readRDS(file.path(paths$data, "TERM2NAME.rds"))
TERM2GENE <- readRDS(file.path(paths$data, "TERM2GENE_PANNZER_full.rds"))

cat("GO annotation loaded:\n")
cat("  Biological Process terms:", 
    length(unique(TERM2NAME$BP$go_id)), "\n")
cat("  Gene-term associations:", nrow(TERM2GENE), "\n")
```

---

# Define Gene Sets for Enrichment

## Extract Network Substructure Node Lists

**Purpose**: Identify nodes in flanking-regulated vs Inv4m-regulated 
sub-networks from edge structure.

```{r define_network_subsets}
flanking_network_genes <- trans_edges %>%
  filter(regulator_region == "flanking") %>%
  {union(.$from, .$to)} %>%
  sort() %>%
  unique()

inv4m_network_genes <- trans_edges %>%
  filter(regulator_region == "inv4m") %>%
  {union(.$from, .$to)} %>%
  setdiff(flanking_network_genes) %>%
  sort() %>%
  unique()

cat("Network subsets (Method 1 - Location setdiff):\n")
cat("  Flanking-regulated network:", length(flanking_network_genes), "genes\n")
cat("  Inv4m-regulated network:", length(inv4m_network_genes), "genes\n")
cat("  Overlap:",
    length(intersect(flanking_network_genes, inv4m_network_genes)), "\n")
```

## Method 2: Strongest Non-Flanking Regulator

**Purpose**: Alternative gene set definition that keeps trans targets where Inv4m
is the strongest regulator, even if they also connect to flanking regulators.

This recovers trans targets that Method 1 excludes due to any connection to
flanking network.

```{r define_method2_subsets}
# Define regulators by region
inv4m_regulators <- trans_edges %>%
  filter(regulator_region == "inv4m") %>%
  pull(from) %>%
  unique()

flanking_regulators <- trans_edges %>%
  filter(regulator_region == "flanking") %>%
  pull(from) %>%
  unique()

# For each target, find strongest correlated regulator
strongest_edge_per_target <- trans_edges %>%
  mutate(abs_r = abs(r)) %>%
  group_by(to) %>%
  slice_max(abs_r, n = 1, with_ties = FALSE) %>%
  ungroup()

# Targets whose strongest regulator is NOT in flanking
targets_strongest_nonflanking <- strongest_edge_per_target %>%
  filter(regulator_region != "flanking") %>%
  pull(to)

# Inv4m edges to these targets
edges_strongest_nonflanking_inv4m <- trans_edges %>%
  filter(
    to %in% targets_strongest_nonflanking,
    regulator_region == "inv4m"
  )

# Method 2 gene set
inv4m_network_genes_method2 <- union(
  edges_strongest_nonflanking_inv4m$from,
  edges_strongest_nonflanking_inv4m$to
) %>%
  sort() %>%
  unique()

cat("\nNetwork subsets (Method 2 - Strongest non-flanking regulator):\n")
cat("  Inv4m regulators:", length(inv4m_regulators), "\n")
cat("  Flanking regulators:", length(flanking_regulators), "\n")
cat("  Targets with Inv4m as strongest:", length(targets_strongest_nonflanking), "\n")
cat("  Inv4m network (Method 2):", length(inv4m_network_genes_method2), "genes\n")

# Compare methods
cat("\nMethod comparison:\n")
cat("  Method 1 (setdiff):", length(inv4m_network_genes), "genes\n")
cat("  Method 2 (strongest):", length(inv4m_network_genes_method2), "genes\n")
cat("  Overlap:", length(intersect(inv4m_network_genes, inv4m_network_genes_method2)), "\n")
cat("  Only in Method 2:", length(setdiff(inv4m_network_genes_method2, inv4m_network_genes)),
    "(trans targets recovered)\n")
```

## Create Directional Gene Sets

**Purpose**: Split network genes by direction of regulation for enrichment.

```{r define_gene_sets}
# Method 1 gene sets (location-based setdiff)
network_gene_sets_m1 <- list(
  `flank.Up` = trans_network_all %>%
    filter(gene %in% flanking_network_genes, logFC > 0) %>%
    pull(gene),

  `flank.Down` = trans_network_all %>%
    filter(gene %in% flanking_network_genes, logFC < 0) %>%
    pull(gene),

  `Inv4m.Up` = trans_network_all %>%
    filter(gene %in% inv4m_network_genes, logFC > 0) %>%
    pull(gene),

  `Inv4m.Down` = trans_network_all %>%
    filter(gene %in% inv4m_network_genes, logFC < 0) %>%
    pull(gene)
)

# Method 2 gene sets (strongest non-flanking regulator)
# For Method 2, we need logFC info - join with effects_df
method2_genes_with_logfc <- effects_df %>%
  filter(gene %in% inv4m_network_genes_method2) %>%
  group_by(gene) %>%
  slice(1) %>%
  ungroup()

network_gene_sets_m2 <- list(
  `flank.Up` = trans_network_all %>%
    filter(gene %in% flanking_network_genes, logFC > 0) %>%
    pull(gene),

  `flank.Down` = trans_network_all %>%
    filter(gene %in% flanking_network_genes, logFC < 0) %>%
    pull(gene),

  `Inv4m.Up.M2` = method2_genes_with_logfc %>%
    filter(logFC > 0) %>%
    pull(gene),

  `Inv4m.Down.M2` = method2_genes_with_logfc %>%
    filter(logFC < 0) %>%
    pull(gene)
)

# Use Method 1 for main analysis (backward compatible)
network_gene_sets <- network_gene_sets_m1

n_genes_m1 <- sapply(network_gene_sets_m1, length)
n_genes_m2 <- sapply(network_gene_sets_m2, length)

cat("=== Method 1 (Location setdiff) ===\n")
cat("Gene set sizes:\n")
print(n_genes_m1)
cat("  Inv4m total:", n_genes_m1["Inv4m.Up"] + n_genes_m1["Inv4m.Down"], "\n")

cat("\n=== Method 2 (Strongest non-flanking) ===\n")
cat("Gene set sizes:\n")
print(n_genes_m2)
cat("  Inv4m total:", n_genes_m2["Inv4m.Up.M2"] + n_genes_m2["Inv4m.Down.M2"], "\n")

cat("\nDifference: Method 2 adds",
    (n_genes_m2["Inv4m.Up.M2"] + n_genes_m2["Inv4m.Down.M2"]) -
    (n_genes_m1["Inv4m.Up"] + n_genes_m1["Inv4m.Down"]),
    "trans targets\n")
```

---

# GO Enrichment Analysis

## Define Enrichment Function

**Purpose**: Wrapper for clusterProfiler enrichment with consistent parameters.

```{r enrichment_function}
#' Perform GO enrichment analysis
#'
#' Wrapper around clusterProfiler::enricher with standardized parameters
#' for maize GO annotation.
#'
#' @param x Character vector of gene IDs
#' @param ontology GO ontology: "BP", "MF", or "CC"
#' @param universe Character vector of background genes
#'
#' @return enrichResult object from clusterProfiler
ego <- function(x, ontology = "BP", universe = NULL) {
  enricher(
    gene = x,
    pvalueCutoff = 0.05,
    pAdjustMethod = "none",
    universe = universe,
    minGSSize = 10,
    maxGSSize = 500,
    qvalueCutoff = 0.2,
    TERM2GENE = TERM2GENE[TERM2GENE$GO %in% TERM2NAME[[ontology]]$go_id, ],
    TERM2NAME = TERM2NAME[[ontology]]
  )
}
```

## Compare Networks: Biological Process

**Purpose**: Identify enriched GO Biological Process terms across network 
subsets.

```{r compare_networks_bp}
compGO_BP <- compareCluster(
  geneCluster = network_gene_sets,
  fun = function(x) {
    ego(x, ontology = "BP", universe = background)
  }
)

filtered_result <- compGO_BP@compareClusterResult %>%
  filter(Count > 1) %>%
  group_by(Cluster, geneID) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  ungroup()

{
cat("\nEnrichment results by network:\n")
print(table(compGO_BP@compareClusterResult$Cluster))
print(table(filtered_result$Cluster))
}
```

---

# Filter and Process Results

## Remove Redundant and Low-Count Terms

**Purpose**: Filter GO results to reduce semantic redundancy and focus on 
well-supported terms.

```{r filter_results}
library(AnnotationHub)

ah <- AnnotationHub()
org.Zm.eg.db <- ah[["AH119718"]]

library(rrvgo)


# Calculate similarity matrix
go_terms <- unique(compGO_BP@compareClusterResult$ID)
simMatrix <- calculateSimMatrix(go_terms,
  orgdb = org.Zm.eg.db,  # Maize annotation
  ont = "BP",
  method = "Rel"
)

# Reduce to representative terms
scores <- setNames(-log10(compGO_BP@compareClusterResult$p.adjust),
                   compGO_BP@compareClusterResult$ID)
reducedTerms <- reduceSimMatrix(simMatrix,
  scores,
  threshold = 0.7,
  orgdb = org.Zm.eg.db
)

# Keep only parent (representative) terms
compGO_BP@compareClusterResult  <- 
  filtered_result  %>%
  filter(
    ID %in% c(reducedTerms$parent,
              "GO:0040007")
  )

{
cat("\nEnrichment results by network:\n")
table(compGO_BP@compareClusterResult$Cluster)
}
```

## Identify Terms with Known Genes

**Purpose**: Focus on GO terms that include genes with characterized symbols.

```{r identify_known_genes}
effects_symbol <- effects_df %>%
  filter(is_DEG) %>%
  dplyr::select(gene, SYMBOL = locus_symbol, locus_label, locus_name, 
         logFC, adj.P.Val) %>%
  filter(!is.na(SYMBOL) & !is.na(locus_name)) %>%
  mutate(neglogP = -log10(adj.P.Val))

with_known_gene <- compGO_BP@compareClusterResult %>%
  separate_longer_delim(geneID, delim = "/") %>%
  dplyr::rename(gene = geneID) %>%
  inner_join(
    effects_symbol %>% 
      dplyr::select(gene, SYMBOL, locus_name),
    relationship = "many-to-many"
  ) %>%
  pull(ID) %>%
  sort() %>%
  unique()

cat("GO terms with known genes:", length(with_known_gene), "\n")
```

---

# Visualize Enrichment Results

## Dotplot: All Networks

**Purpose**: Overview of enriched GO terms across network subsets.

```{r dotplot_all, fig.height=10, fig.width=12}
cluster_order <- names(network_gene_sets)
levels(compGO_BP@compareClusterResult$Cluster) <- cluster_order

dotplot(compGO_BP, size = "Count", showCategory = 10) +
  ggtitle("GO Biological Process Annotation\nInv4m Trans Co-expression Network") +
  scale_size(range = c(4, 8), breaks = 1:4) +
  scale_x_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Prepare Data for Annotated Plot

**Purpose**: Select top terms per cluster and identify representative genes.

```{r prepare_plot_data}
sorted_by_p <- compGO_BP@compareClusterResult %>%
  filter(ID %in% with_known_gene) %>%
  mutate(
    neglogP = -log10(p.adjust),
    Cluster = factor(Cluster, levels = cluster_order)
  ) %>%
  group_by(Cluster) %>%
  arrange(Cluster, p.adjust) %>%
  dplyr::slice(1:5) %>%
  mutate(label_order = 1:n()) %>%
  mutate(Description = fct_reorder(Description, label_order))

top_terms <- levels(sorted_by_p$Description)

to_plot <- compGO_BP@compareClusterResult %>%
  filter(ID %in% with_known_gene) %>%
  mutate(Cluster = factor(Cluster, levels = cluster_order)) %>%
  filter(Description %in% top_terms) %>%
  mutate(
    neglogP = -log10(p.adjust),
    Description = factor(Description, levels = top_terms)
  ) %>%
  group_by(Cluster) %>%
  arrange(Cluster, p.adjust) %>%
  mutate(
    label_order = 1:n(),
    Description = fct_reorder(Description, label_order)
  )

term2symbol <- to_plot %>%
  separate_longer_delim(geneID, delim = "/") %>%
  dplyr::rename(gene = geneID) %>%
  inner_join(effects_symbol, relationship = "many-to-many", by="gene") %>%
  dplyr::select(Cluster, gene, SYMBOL, locus_label, ID, Description, 
         p.adjust, Count, logFC, adj.P.Val) %>%
  distinct() %>%
  group_by(Cluster, ID) %>%
  arrange(Cluster, ID, adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  mutate(gene_label = ifelse(
    !is.na(locus_label) & locus_label != "", 
    locus_label, 
    SYMBOL
  ))

cat("\nTerms with gene labels:", nrow(term2symbol), "\n")
```

## Bubble Plot with Gene Annotations

**Purpose**: Detailed visualization showing enriched GO terms with 
representative gene symbols.

```{r bubble_plot_annotated, fig.height=5, fig.width=14}
labels <- paste0(
  gsub(".", " ", cluster_order, fixed = TRUE),
  "\n(", n_genes_m1, ")"
)
names(labels) <- names(n_genes_m1)

bg <- to_plot %>%
  ggplot(aes(x = Cluster, y = Description, size = Count, fill = neglogP)) +
  ggtitle("GO Biological Process Annotation\nInv4m Trans Co-expression Network") +
  scale_fill_gradient(
    low = "dodgerblue", 
    high = "tomato", 
    name = "-log10(FDR)",
    limits = c(1, max(to_plot$neglogP, na.rm = TRUE)),
    guide = guide_colorbar()
  ) +
  scale_y_discrete(
    labels = scales::wrap_format(60),
    limits = rev(levels(to_plot$Description))
  ) +
  scale_x_discrete(
    labels = labels, 
    drop = FALSE,
    expand = expansion(add = c(0.5, 1))
  ) +
  geom_point(shape = 21) +
  scale_size(range =c(6,8),breaks = 2:4) +
  ylab(NULL) +
  xlab(NULL) +
  DOSE::theme_dose(font.size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    legend.position = "right",
    axis.text.x = element_text(vjust = 0.5, angle = 0, hjust = 0.5)
  )

annotation_plot <- bg +
  geom_text(
    data = term2symbol,
    position = position_nudge(0.15),
    hjust = 0,
    vjust = 0.5,
    fontface = "italic",
    size = 5,
    mapping = aes(x = Cluster, y = Description, label = gene_label),
    inherit.aes = FALSE
  )

print(annotation_plot)
```

---

# Method 2: GO Enrichment Comparison

**Purpose**: Compare GO enrichment between Method 1 (location setdiff) and
Method 2 (strongest non-flanking regulator) to assess impact of including
trans targets.

## Run Method 2 Enrichment

```{r method2_enrichment}
# Run GO enrichment on Method 2 gene sets (Inv4m only, since flanking is same)
compGO_BP_M2 <- compareCluster(
  geneCluster = network_gene_sets_m2,
  fun = function(x) {
    ego(x, ontology = "BP", universe = background)
  }
)

cat("Method 2 RAW enrichment results by cluster:\n")
print(table(compGO_BP_M2@compareClusterResult$Cluster))

# Apply same filtering as Method 1: Count > 1, then rrvgo semantic reduction
filtered_result_m2 <- compGO_BP_M2@compareClusterResult %>%
  filter(Count > 1) %>%
  group_by(Cluster, geneID) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  ungroup()

# Calculate similarity matrix for Method 2 terms
go_terms_m2 <- unique(compGO_BP_M2@compareClusterResult$ID)

# Only run rrvgo if there are terms to process
if (length(go_terms_m2) > 1) {
  simMatrix_m2 <- calculateSimMatrix(go_terms_m2,
    orgdb = org.Zm.eg.db,
    ont = "BP",
    method = "Rel"
  )

  # Reduce to representative terms
  scores_m2 <- setNames(-log10(compGO_BP_M2@compareClusterResult$p.adjust),
                        compGO_BP_M2@compareClusterResult$ID)
  reducedTerms_m2 <- reduceSimMatrix(simMatrix_m2,
    scores_m2,
    threshold = 0.7,
    orgdb = org.Zm.eg.db
  )

  # Keep only parent (representative) terms + manually add growth (GO:0040007)
  # Growth is significant (p.adjust < 0.05) and biologically relevant:
  # - Plant height differs significantly in field
  # - Annotated genes include jmj cluster
  compGO_BP_M2@compareClusterResult <- filtered_result_m2 %>%
    filter(ID %in% c(reducedTerms_m2$parent, "GO:0040007"))
}

cat("\nMethod 2 FILTERED enrichment results by cluster:\n")
print(table(compGO_BP_M2@compareClusterResult$Cluster))
```

## Compare Methods

```{r compare_methods}
# Extract Inv4m results from both methods
m1_inv4m <- compGO_BP@compareClusterResult %>%
  filter(grepl("Inv4m", Cluster))

m2_inv4m <- compGO_BP_M2@compareClusterResult %>%
  filter(grepl("Inv4m", Cluster))

cat("\n=== GO ENRICHMENT COMPARISON ===\n")
cat("\nMethod 1 (Location setdiff):\n")
cat("  Inv4m gene sets:", n_genes_m1["Inv4m.Up"] + n_genes_m1["Inv4m.Down"], "genes\n")
cat("  Significant GO terms:", nrow(m1_inv4m %>% filter(p.adjust < 0.05)), "\n")

cat("\nMethod 2 (Strongest non-flanking):\n")
cat("  Inv4m gene sets:", n_genes_m2["Inv4m.Up.M2"] + n_genes_m2["Inv4m.Down.M2"], "genes\n")
cat("  Significant GO terms:", nrow(m2_inv4m %>% filter(p.adjust < 0.05)), "\n")

# Search for growth-related terms
growth_keywords <- c("growth", "cell cycle", "division", "meristem",
                     "proliferat", "development", "morphogenesis")

m1_growth <- m1_inv4m %>%
  filter(p.adjust < 0.05) %>%
  filter(str_detect(tolower(Description), paste(growth_keywords, collapse = "|")))

m2_growth <- m2_inv4m %>%
  filter(p.adjust < 0.05) %>%
  filter(str_detect(tolower(Description), paste(growth_keywords, collapse = "|")))

cat("\n=== GROWTH-RELATED TERMS ===\n")
cat("\nMethod 1 growth terms:", nrow(m1_growth), "\n")
if (nrow(m1_growth) > 0) {
  print(m1_growth %>% dplyr::select(Cluster, Description, Count, p.adjust))
}

cat("\nMethod 2 growth terms:", nrow(m2_growth), "\n")
if (nrow(m2_growth) > 0) {
  print(m2_growth %>% dplyr::select(Cluster, Description, Count, p.adjust))
}
```

## Method 2 Top Terms

```{r method2_top_terms}
cat("\n=== Method 2: Top GO Terms by Cluster ===\n")

m2_inv4m %>%
  filter(p.adjust < 0.05) %>%
  group_by(Cluster) %>%
  arrange(p.adjust) %>%
  slice_head(n = 10) %>%
  dplyr::select(Cluster, Description, Count, p.adjust) %>%
  print(n = 30)
```

## Method 2 Visualization

```{r method2_bubble_plot, fig.height=5, fig.width=14}
# Prepare data: top 5 terms per group by p.adjust
cluster_order_m2 <- c("flank.Up", "flank.Down", "Inv4m.Up.M2", "Inv4m.Down.M2")

to_plot_m2 <- compGO_BP_M2@compareClusterResult %>%
  filter(p.adjust < 0.05) %>%
  mutate(
    neglogP = -log10(p.adjust),
    Cluster = factor(Cluster, levels = cluster_order_m2)
  ) %>%
  group_by(Cluster) %>%
  arrange(p.adjust) %>%
  slice_head(n = 5) %>%
  mutate(
    label_order = 1:n(),
    Description = fct_reorder(Description, label_order)
  ) %>%
  ungroup()

# Get gene labels for each term (most significant gene per term)
term2symbol_m2 <- to_plot_m2 %>%
  separate_longer_delim(geneID, delim = "/") %>%
  dplyr::rename(gene = geneID) %>%
  inner_join(effects_symbol, relationship = "many-to-many", by = "gene") %>%
  dplyr::select(Cluster, gene, SYMBOL, locus_label, ID, Description,
         p.adjust, Count, logFC, adj.P.Val) %>%
  distinct() %>%
  group_by(Cluster, ID) %>%
  arrange(Cluster, ID, adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  mutate(gene_label = ifelse(
    !is.na(locus_label) & locus_label != "",
    locus_label,
    SYMBOL
  ))

cat("\nMethod 2 bubble plot:\n")
cat("  Terms to plot:", nrow(to_plot_m2), "\n")
cat("  Terms with gene labels:", nrow(term2symbol_m2), "\n")

# Create facet labels with gene counts
labels_m2 <- paste0(
  gsub(".", " ", cluster_order_m2, fixed = TRUE),
  "\n(", n_genes_m2, ")"
)
names(labels_m2) <- names(n_genes_m2)

# Build bubble plot
bg_m2 <- to_plot_m2 %>%
  ggplot(aes(x = Cluster, y = Description, size = Count, fill = neglogP)) +
  ggtitle("GO Biological Process Annotation\nMethod 2: Strongest Non-Flanking Regulator") +
  scale_fill_gradient(
    low = "dodgerblue",
    high = "tomato",
    name = "-log10(FDR)",
    limits = c(1, max(to_plot_m2$neglogP, na.rm = TRUE)),
    guide = guide_colorbar()
  ) +
  scale_y_discrete(
    labels = scales::wrap_format(60),
    limits = rev(levels(to_plot_m2$Description))
  ) +
  scale_x_discrete(
    labels = labels_m2,
    drop = FALSE,
    expand = expansion(add = c(0.5, 1))
  ) +
  geom_point(shape = 21) +
  scale_size(range = c(6, 8), breaks = 2:4) +
  ylab(NULL) +
  xlab(NULL) +
  DOSE::theme_dose(font.size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    legend.position = "right",
    axis.text.x = element_text(vjust = 0.5, angle = 0, hjust = 0.5)
  )

# Add gene annotations
annotation_plot_m2 <- bg_m2 +
  geom_text(
    data = term2symbol_m2,
    position = position_nudge(0.15),
    hjust = 0,
    vjust = 0.5,
    fontface = "italic",
    size = 5,
    mapping = aes(x = Cluster, y = Description, label = gene_label),
    inherit.aes = FALSE
  )

print(annotation_plot_m2)
```

---

# Export Results

**Purpose**: Save enrichment results and visualizations.

```{r export_results}
write.csv(
  compGO_BP@compareClusterResult,
  file = file.path(paths$intermediate, "GO_enrichment_trans_network_all.csv"),
  row.names = FALSE
)

gene_term_associations <- to_plot %>%
  separate_longer_delim(geneID, delim = "/") %>%
  dplyr::rename(gene = geneID) %>%
  left_join(effects_symbol, relationship = "many-to-many") %>%
  dplyr::select(Cluster, gene, SYMBOL, locus_name, ID, Description, 
         p.adjust, Count, logFC, adj.P.Val)

write.csv(
  gene_term_associations,
  file = file.path(paths$intermediate, "GO_gene_term_associations.csv"),
  row.names = FALSE
)

ggsave(
  annotation_plot,
  file = file.path(paths$figures, "GO_trans_network_annotated.svg"),
  height = 10,
  width = 14
)

ggsave(
  annotation_plot,
  file = file.path(paths$figures, "GO_trans_network_annotated.pdf"),
  height = 10,
  width = 14
)

# Export Method 2 results
write.csv(
  compGO_BP_M2@compareClusterResult,
  file = file.path(paths$intermediate, "GO_enrichment_trans_network_method2.csv"),
  row.names = FALSE
)

cat("\nFiles exported to paths$intermediate and paths$figures\n")
cat("  - GO_enrichment_trans_network_all.csv (Method 1)\n")
cat("  - GO_enrichment_trans_network_method2.csv (Method 2)\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```