---
title: "Greenyellow Module Characterization"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)

knitr::opts_knit$set(root.dir = here::here())
```

# Overview

**Purpose**: Characterize the greenyellow WGCNA module, which contains the plant
height GWAS candidate *sec6* and the cell proliferation marker *pcna2*. This
module has the highest bootstrap support (LFP = 0.80) among all WGCNA modules.

**Key findings**:

- The greenyellow module contains 16 DEGs including sec6 and pcna2
- GO enrichment reveals "developmental cell growth" (p = 8.5e-4) driven by ropgef14/pla2
- Module includes DNA replication (pcna2), exocytosis (sec6), and cell polarity genes
- Supports the hypothesis that Inv4m perturbs a growth/cell division network

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(knitr)
})

# Load project paths
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

```{r define_paths}
# Field perturbation run directory
run_dir <- here("results/inversion_paper/field_perturbation/run_20251231_201332")

# Input files
CONSENSUS_FILE <- file.path(run_dir, "04_consensus_networks/ctrl/consensus_modules.csv")
GO_BP_FILE <- file.path(run_dir, "07_module_annotation/all_modules_GO_BP.csv")
LABELS_FILE <- here("data", "locus_labels_master.csv")
SUPPORT_FILE <- file.path(run_dir, "05_bootstrap_support/ctrl/module_support.csv")

# Verify files exist
stopifnot(file.exists(CONSENSUS_FILE))
stopifnot(file.exists(GO_BP_FILE))
stopifnot(file.exists(LABELS_FILE))
stopifnot(file.exists(SUPPORT_FILE))

cat("All input files found.\n")
```

---

# Load Data

```{r load_data}
# Load CTRL consensus module assignments
ctrl_modules <- read_csv(CONSENSUS_FILE, show_col_types = FALSE)

# Load GO BP enrichment results
go_bp <- read_csv(GO_BP_FILE, show_col_types = FALSE)

# Load gene annotations with expression data
labels <- read_csv(LABELS_FILE, show_col_types = FALSE)

# Load module support statistics
support <- read_csv(SUPPORT_FILE, show_col_types = FALSE)

cat("Loaded", nrow(ctrl_modules), "genes with module assignments\n")
cat("Loaded", nrow(go_bp), "GO BP enrichment terms\n")
cat("Loaded", nrow(labels), "gene annotations\n")
```

---

# Module Bootstrap Support

The greenyellow module has the highest bootstrap support among all modules.

```{r module_support}
# Extract greenyellow support stats
gy_support <- support %>%
  filter(color == "greenyellow")

cat("Greenyellow module bootstrap support:\n")
cat("  Mean LFP:", round(gy_support$mean_lfp, 3), "\n")
cat("  Stability:", gy_support$stability, "\n")
cat("  N genes:", gy_support$n_genes, "\n")

# Compare to other modules
cat("\nTop 5 modules by LFP:\n")
support %>%
  arrange(desc(mean_lfp)) %>%
  head(5) %>%
  select(color, mean_lfp, n_genes, stability) %>%
  kable()
```

---

# Greenyellow Module Genes

The greenyellow module contains sec6 (plant height GWAS candidate) and pcna2
(proliferating cell nuclear antigen), both upregulated in Inv4m plants.

```{r extract_greenyellow_module}
# Extract greenyellow module genes
gy_genes <- ctrl_modules %>%
  filter(color == "greenyellow") %>%
  pull(gene)

cat("Greenyellow module contains", length(gy_genes), "genes\n")

# Verify key genes are present
key_genes <- c(
  "Zm00001eb194380",  # sec6
  "Zm00001eb192050",  # pcna2
  "Zm00001eb188240"   # ropgef14
)
cat("\nsec6 in greenyellow module:", key_genes[1] %in% gy_genes, "\n")
cat("pcna2 in greenyellow module:", key_genes[2] %in% gy_genes, "\n")
cat("ropgef14 in greenyellow module:", key_genes[3] %in% gy_genes, "\n")
```

---

# Annotated DEG Table

Table sorted by statistical significance (FDR, ascending = most significant first).

```{r annotate_greenyellow_genes}
# Join greenyellow genes with annotations
gy_annotated <- ctrl_modules %>%
  filter(color == "greenyellow") %>%
  left_join(labels, by = "gene") %>%
  mutate(
    # Use locus_label if available, else locus_symbol
    symbol = case_when(
      !is.na(locus_label) & locus_label != "" ~ locus_label,
      !is.na(locus_symbol) & locus_symbol != "" ~ locus_symbol,
      TRUE ~ "---"
    ),
    # Clean up description - use desc_merged or desc_pannzer
    description = case_when(
      !is.na(desc_merged) & desc_merged != "" ~ desc_merged,
      !is.na(desc_pannzer) & desc_pannzer != "" ~ desc_pannzer,
      TRUE ~ "---"
    ),
    # Format log2FC and FDR
    log2FC = round(logFC, 2),
    FDR_val = adj.P.Val
  ) %>%
  # Sort by FDR (most significant first)
  arrange(FDR_val) %>%
  select(gene, symbol, description, log2FC, FDR_val)

# Display count
cat("Annotated", nrow(gy_annotated), "greenyellow module genes\n")
```

```{r display_table}
# Create formatted table
gy_table <- gy_annotated %>%
  mutate(
    FDR = formatC(FDR_val, format = "e", digits = 1),
    # Truncate long descriptions
    description = str_trunc(description, 50)
  ) %>%
  select(gene, symbol, description, log2FC, FDR)

# Display with kable
kable(
  gy_table,
  col.names = c("Gene ID", "Label", "Description", "log2FC", "FDR"),
  caption = "Greenyellow module DEGs sorted by statistical significance (FDR)"
)
```

---

# GO Enrichment for Greenyellow Module

```{r go_enrichment}
# Filter GO BP results for greenyellow module
gy_go <- go_bp %>%
  filter(module_color == "greenyellow") %>%
  arrange(pvalue) %>%
  select(ID, Description, GeneRatio, pvalue, Count, geneID) %>%
  mutate(
    pvalue_fmt = formatC(pvalue, format = "e", digits = 2),
    Description = str_trunc(Description, 60)
  )

cat("Greenyellow module has", nrow(gy_go), "enriched GO BP terms\n")
```

```{r display_go_table}
# Display top GO terms
kable(
  gy_go %>%
    head(20) %>%
    select(ID, Description, GeneRatio, pvalue_fmt, Count),
  col.names = c("GO ID", "Description", "Gene Ratio", "p-value", "Count"),
  caption = "Top 20 GO Biological Process terms for greenyellow module"
)
```

---

# Growth-Related GO Terms

Highlight GO terms related to growth, cell cycle, and cell proliferation.

```{r growth_go_terms}
# Filter for growth-related terms
growth_go <- gy_go %>%
  filter(grepl("growth|proliferation|cell cycle|DNA replication|cell division|cytokinesis|cell tip",
               Description, ignore.case = TRUE))

cat("Growth-related GO terms in greenyellow module:\n\n")

kable(
  growth_go %>%
    select(ID, Description, GeneRatio, pvalue_fmt, geneID),
  col.names = c("GO ID", "Description", "Gene Ratio", "p-value", "Genes"),
  caption = "Growth-related GO terms in greenyellow module"
)
```

---

# Functional Categories

```{r functional_categories}
# Manually categorize key genes
categories <- tribble(
  ~category, ~genes, ~description,
  "Cell proliferation", "pcna2", "DNA replication licensing - cell cycle S phase",
  "Exocytosis/Cytokinesis", "sec6", "Exocyst complex - vesicle trafficking, cell plate formation",
  "Cell polarity/Growth", "ropgef14, pla2", "Rho GTPase signaling - polarized cell growth",
  "DNA repair", "parg, pcna2", "Poly(ADP-ribose) metabolism, DNA damage response",
  "Transcription factor", "mybr58", "MYB-related TF - potentially growth-related targets",
  "Signaling kinases", "pbl19, pk3, pk8", "Protein kinases - signal transduction"
)

kable(
  categories,
  col.names = c("Category", "Genes", "Biological Role"),
  caption = "Functional categories in the greenyellow module"
)
```

---

# Key Genes: sec6 and pcna2

## sec6 (Exocyst complex component)

- **Gene ID**: Zm00001eb194380
- **log2FC**: +2.87 (upregulated in Inv4m)
- **FDR**: 2.9e-17
- **GWAS support**: Plant height candidate (Peiffer et al. 2014)
- **Function**: Exocyst complex component involved in vesicle trafficking and
  cell plate formation during cytokinesis

## pcna2 (Proliferating cell nuclear antigen 2)

- **Gene ID**: Zm00001eb192050
- **log2FC**: +2.99 (upregulated in Inv4m)
- **FDR**: 9.3e-03
- **Function**: DNA replication processivity factor, marker of cell proliferation,
  essential for S phase progression

## ropgef14 (Rop guanine nucleotide exchange factor 14)

- **Gene ID**: Zm00001eb188240
- **log2FC**: -2.98 (downregulated in Inv4m)
- **FDR**: 2.6e-14
- **Function**: Activator of Rho-type GTPases, required for polarized cell growth
  including pollen tube elongation

---

# Summary

The greenyellow module represents a coherent **cell growth/division network**
with the highest bootstrap support (LFP = 0.80) among all WGCNA modules:

1. **Cell proliferation**: pcna2 (DNA replication), parg (DNA repair)
2. **Cytokinesis**: sec6 (exocyst complex, cell plate formation)
3. **Polarized growth**: ropgef14, pla2 (Rho GTPase signaling, tip growth)
4. **Signaling**: Multiple protein kinases (pbl19, pk3, pk8)

The GO enrichment confirms:

- **"Developmental cell growth"** (p = 8.5e-4) - polarized growth genes
- **"Cell tip growth"** (p = 6.0e-4) - pollen tube elongation genes
- **"Regulation of DNA replication"** (p = 0.019) - pcna2

The presence of both sec6 (GWAS plant height candidate) and pcna2 (cell
proliferation marker) in this highly stable module supports the hypothesis
that Inv4m introgression perturbs a coordinated cell growth/division program,
consistent with the altered SAM morphology and plant height phenotypes.

---

# Save Outputs

```{r save_outputs}
# Save annotated table (CSV)
write_csv(
  gy_annotated %>%
    mutate(FDR = formatC(FDR_val, format = "e", digits = 2)) %>%
    select(gene, symbol, description, log2FC, FDR),
  file.path(paths$intermediate, "greenyellow_module_annotated_DEGs.csv")
)

# Save GO enrichment
write_csv(
  gy_go %>% select(-pvalue_fmt),
  file.path(paths$intermediate, "greenyellow_module_GO_BP.csv")
)

cat("\nSaved outputs:\n")
cat("  - greenyellow_module_annotated_DEGs.csv\n")
cat("  - greenyellow_module_GO_BP.csv\n")
```

```{r generate_latex_table}
# Generate LaTeX table for main.tex
latex_header <- '\\begin{table}[!ht]
\\caption{\\textbf{Greenyellow module DEGs containing growth and cell division genes.} Genes assigned to the greenyellow WGCNA consensus module, which has the highest bootstrap support (LFP = 0.80) among all modules. Contains the plant height GWAS candidate \\textit{sec6} and the cell proliferation marker \\textit{pcna2}. Sorted by statistical significance (FDR). GO enrichment includes ``developmental cell growth\'\' ($p = 8.5 \\times 10^{-4}$) and ``regulation of DNA replication\'\' ($p = 0.019$).}
\\label{tab::greenyellow_module}
\\centering
\\small
\\begin{tabular}{@{}lllrr@{}}
\\toprule
\\textbf{Gene ID} & \\textbf{Label} & \\textbf{Description} & \\textbf{log$_2$FC} & \\textbf{FDR} \\\\
\\midrule'

latex_footer <- '\\bottomrule
\\end{tabular}
\\end{table}'

# Format each row
format_latex_row <- function(gene, symbol, desc, log2fc, fdr) {
  # Format label
  label <- if (is.na(symbol) || symbol == "" || symbol == "---") {
    "---"
  } else {
    paste0("\\textit{", symbol, "}")
  }

  # Format description
  desc_clean <- if (is.na(desc) || desc == "" || desc == "---") {
    "---"
  } else {
    str_trunc(desc, 40) %>%
      str_replace_all("_", "\\\\_") %>%
      str_replace_all("&", "\\\\&") %>%
      str_replace_all("%", "\\\\%")
  }

  # Format log2FC with sign
  log2fc_fmt <- if (log2fc >= 0) {
    paste0("$+", sprintf("%.2f", log2fc), "$")
  } else {
    paste0("$", sprintf("%.2f", log2fc), "$")
  }

  # Format FDR in scientific notation
  fdr_exp <- floor(log10(fdr))
  fdr_mant <- fdr / 10^fdr_exp
  fdr_fmt <- paste0("$", sprintf("%.1f", fdr_mant), " \\times 10^{", fdr_exp, "}$")

  paste(gene, label, desc_clean, log2fc_fmt, fdr_fmt, sep = " & ", collapse = "") %>%
    paste0(" \\\\")
}

# Generate all rows
latex_rows <- gy_annotated %>%
  rowwise() %>%
  mutate(
    latex_row = format_latex_row(gene, symbol, description, log2FC, FDR_val)
  ) %>%
  pull(latex_row)

# Combine into full table
latex_table <- paste(
  latex_header,
  paste(latex_rows, collapse = "\n"),
  latex_footer,
  sep = "\n"
)

# Save to tables folder
writeLines(latex_table, file.path(paths$tables, "greenyellow_module_DEGs.tex"))

cat("\nLaTeX table saved:\n")
cat("  - tables/greenyellow_module_DEGs.tex\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
