---
title: "Internode Length Analysis: NC2025 Field Experiment"
author: "Fausto Rodriguez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    fig_width: 12
    fig_height: 8
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)

# Setup project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

## Overview

This analysis examines **internode length measurements** from the NC2025 field experiment to characterize the effect of the Inv4m introgression on plant architecture.

**Research Questions:**
1. Does Inv4m affect internode length patterns along the stem?
2. Does Inv4m affect the total number of nodes?
3. How do internode profiles relate to overall plant height?

## 1. Load Libraries and Data

```{r load-libraries}
library(tidyverse)
library(rstatix)
library(ggpubr)
library(ggtext)
library(ggbeeswarm)
library(broom)
library(patchwork)
```
```{r load-data}
# Load field metadata
field_data <- read_csv(file.path(paths$data, "CLY25_Inv4m.csv"),
                       na = c("", "NA", NA))

# Extract metadata columns
metadata <- field_data %>%
  select(plant:group, X_pos, Y_pos)

# Load internode measurements
phytomere <- read_csv(file.path(paths$data, "CLY25_Inv4m_Internode.csv"),
                      na = c("", "na", "NA", NA))

# Load dissection validation data
dissection <- read_csv(file.path(paths$data, "CLY25_Inv4m_dissection.csv"))

cat("Field data dimensions:", dim(field_data), "\n")
cat("Internode measurements:", nrow(phytomere), "records\n")
cat("Dissection records:", nrow(dissection), "records\n")
```

## 2. Data Processing

```{r data-processing}
# Join metadata with internode measurements
phytomere_processed <- metadata %>%
  inner_join(phytomere, by = "plant") %>%
  rename(genotype = "inv4m_gt") %>%
  filter(!is.na(length)) %>%
  filter(!(grepl("no tassel", notes))) %>%
  group_by(plant) %>%
  arrange(-internode) %>%
  mutate(from_top = row_number()) %>%
  ungroup()

# Filter by donor (focus on MI21 and TMEX with ≤10 internodes from top)
phytomere_processed <- phytomere_processed %>%
  filter(donor == "MI21" | (donor == "TMEX" & from_top <= 10))

# Create genotype labels for plotting
phytomere_processed <- phytomere_processed %>%
  mutate(
    Genotype_label = case_when(
      genotype == "CTRL" ~ "CTRL",
      genotype == "INV4M" ~ "<i>Inv4m</i>",
      TRUE ~ genotype
    )
  )

cat("\nProcessed data dimensions:", dim(phytomere_processed), "\n")
cat("Plants:", n_distinct(phytomere_processed$plant), "\n")
cat("Genotypes:", paste(unique(phytomere_processed$genotype), collapse = ", "), "\n")
cat("Donors:", paste(unique(phytomere_processed$donor), collapse = ", "), "\n")
```

## 3. Summary Statistics

```{r summary-stats}
# Summary by genotype and donor
summary_by_genotype <- phytomere_processed %>%
  group_by(donor, genotype) %>%
  summarise(
    n_plants = n_distinct(plant),
    n_internodes = n(),
    mean_length = mean(length, na.rm = TRUE),
    sd_length = sd(length, na.rm = TRUE),
    .groups = "drop"
  )

cat("Summary by genotype and donor:\n")
print(summary_by_genotype)

# Node count per plant
node_count <- phytomere_processed %>%
  group_by(donor, genotype, plant) %>%
  summarise(n_nodes = n(), .groups = "drop")

cat("\nNode count summary:\n")
node_count %>%
  group_by(donor, genotype) %>%
  summarise(
    n_plants = n(),
    mean_nodes = mean(n_nodes),
    sd_nodes = sd(n_nodes),
    .groups = "drop"
  ) %>%
  print()
```

## 4. Internode Length Profiles

### 4.1 Theme and Color Palette

```{r plot-theme}
# Color palette (gold for CTRL, purple for Inv4m)
pal <- c("CTRL" = "gold", "<i>Inv4m</i>" = "purple4", "INV4M" = "purple4")

# Phenotype theme (matching ionome style)
pheno_theme <- theme_classic(base_size = 16) +
  theme(
    plot.title = element_markdown(hjust = 0.5, face = "bold"),
    axis.title.y = element_markdown(),
    axis.title.x = element_markdown(),
    axis.text.x = element_text(face = "bold", color = "black"),
    strip.background = element_blank(),
    strip.text = element_markdown(size = 14),
    legend.position = "top",
    legend.title = element_blank()
  )
```

### 4.2 Line Plot by Position from Top

```{r internode-lines, fig.width=12, fig.height=10}
plot_from_top_lines <- phytomere_processed %>%
  mutate(Genotype_label = factor(Genotype_label, levels = c("CTRL", "<i>Inv4m</i>"))) %>%
  arrange(Genotype_label, length) %>%
  ggplot(aes(x = factor(from_top), y = length, color = Genotype_label, group = plant)) +
  geom_line(alpha = 0.6) +
  geom_point(data = phytomere_processed %>%
               filter(has_ear == 1) %>%
               mutate(Genotype_label = factor(Genotype_label, levels = c("CTRL", "<i>Inv4m</i>"))),
             aes(x = factor(from_top), y = length, fill = Genotype_label),
             size = 3,
             shape = 21,
             color = "white",
             stroke = 1) +
  xlab("Internode position from top") +
  ylab("Internode length (cm)") +
  scale_x_discrete(limits = rev) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  facet_wrap(~ donor, nrow = 2) +
  labs(title = "**Internode Length Profiles by Position from Top**",
       subtitle = "Filled points indicate ear node position") +
  pheno_theme +
  theme(strip.background = element_blank())

print(plot_from_top_lines)

# Save figure
ggsave(file.path(paths$figures, "internode_profiles_from_top.pdf"),
       plot_from_top_lines, width = 12, height = 10)
ggsave(file.path(paths$figures, "internode_profiles_from_top.png"),
       plot_from_top_lines, width = 12, height = 10, dpi = 300)
```

### 4.3 Boxplot by Position from Top

```{r internode-boxplot, fig.width=12, fig.height=10}
plot_from_top_boxplot <- phytomere_processed %>%
  mutate(Genotype_label = factor(Genotype_label, levels = c("CTRL", "<i>Inv4m</i>"))) %>%
  ggplot(aes(x = factor(from_top), y = length, color = Genotype_label)) +
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.7)) +
  xlab("Internode position from top") +
  ylab("Internode length (cm)") +
  scale_color_manual(values = pal) +
  scale_x_discrete(limits = rev) +
  facet_wrap(~ donor, nrow = 2) +
  labs(title = "**Internode Length Distribution by Position**") +
  pheno_theme +
  theme(strip.background = element_blank())

print(plot_from_top_boxplot)

ggsave(file.path(paths$figures, "internode_boxplot_by_position.pdf"),
       plot_from_top_boxplot, width = 12, height = 10)
ggsave(file.path(paths$figures, "internode_boxplot_by_position.png"),
       plot_from_top_boxplot, width = 12, height = 10, dpi = 300)
```

## 5. Node Count Analysis

```{r node-count-analysis, fig.width=10, fig.height=8}
# Statistical tests: comparing genotypes WITHIN each donor
stat_test <- node_count %>%
  group_by(donor) %>%
  t_test(n_nodes ~ genotype) %>%
  add_significance("p") %>%
  add_xy_position(x = "genotype")

cat("Statistical test results by donor:\n")
print(stat_test)

# Summary statistics for reporting
node_summary_stats <- node_count %>%
  group_by(donor, genotype) %>%
  get_summary_stats(n_nodes, type = "mean_sd")
cat("\nSummary statistics:\n")
print(node_summary_stats)

# Create plot with significance brackets
node_count_plot <- node_count %>%
  mutate(
    Genotype_label = case_when(
      genotype == "CTRL" ~ "CTRL",
      genotype == "INV4M" ~ "<i>Inv4m</i>",
      TRUE ~ genotype
    ),
    Genotype_label = factor(Genotype_label, levels = c("CTRL", "<i>Inv4m</i>"))
  ) %>%
  ggplot(aes(x = Genotype_label, y = n_nodes, color = Genotype_label)) +
  geom_boxplot(width = 0.4, linewidth = 1, outlier.shape = NA) +
  geom_beeswarm(size = 2.5, alpha = 0.7) +
  scale_color_manual(values = pal) +
  facet_wrap(~ donor, nrow = 1) +
  labs(
    title = "**Node Count by Genotype**",
    x = NULL,
    y = "Number of nodes"
  ) +
  pheno_theme +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_markdown()
  )

print(node_count_plot)

ggsave(file.path(paths$figures, "node_count_boxplot.pdf"),
       node_count_plot, width = 10, height = 8)
ggsave(file.path(paths$figures, "node_count_boxplot.png"),
       node_count_plot, width = 10, height = 8, dpi = 300)
```

## 6. Height Validation: Sum of Internodes vs Direct Measurement

```{r height-comparison, fig.width=10, fig.height=8}
# Calculate sum of internodes per plant
sum_internodes <- phytomere_processed %>%
  group_by(plant, donor, genotype) %>%
  summarise(
    sum_nodes = sum(length, na.rm = TRUE),
    n_nodes = n(),
    .groups = "drop"
  )

# Join with field data for direct height measurement
compare_height <- sum_internodes %>%
  inner_join(field_data %>% select(plant, PH, X_pos, Y_pos), by = "plant") %>%
  filter(!is.na(PH), !is.na(sum_nodes))

# Linear regression
height_model <- lm(sum_nodes ~ PH, data = compare_height)
r_squared <- summary(height_model)$r.squared
slope <- coef(height_model)[2]
intercept <- coef(height_model)[1]

cat("\nHeight comparison regression:\n")
cat(sprintf("y = %.2fx + %.2f\n", slope, intercept))
cat(sprintf("R² = %.3f\n", r_squared))

# Create comparison plot
equation_text <- sprintf("y = %.2fx + %.2f\nR² = %.3f", slope, intercept, r_squared)

height_comparison_plot <- compare_height %>%
  mutate(
    Genotype_label = case_when(
      genotype == "CTRL" ~ "CTRL",
      genotype == "INV4M" ~ "<i>Inv4m</i>",
      TRUE ~ genotype
    )
  ) %>%
  ggplot(aes(x = PH, y = sum_nodes, color = Genotype_label)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "solid", linewidth = 1) +
  annotate("text",
           x = min(compare_height$PH, na.rm = TRUE) +
             0.05 * diff(range(compare_height$PH, na.rm = TRUE)),
           y = max(compare_height$sum_nodes, na.rm = TRUE) -
             0.1 * diff(range(compare_height$sum_nodes, na.rm = TRUE)),
           label = equation_text,
           hjust = 0, vjust = 1, size = 5) +
  labs(
    title = "**Plant Height Comparison**",
    subtitle = "Red line = 1:1 reference; Dashed = regression fit",
    x = "Direct height measurement (cm)",
    y = "Sum of internode lengths (cm)"
  ) +
  scale_color_manual(values = pal) +
  pheno_theme

print(height_comparison_plot)

ggsave(file.path(paths$figures, "internode_height_validation.pdf"),
       height_comparison_plot, width = 10, height = 8)
ggsave(file.path(paths$figures, "internode_height_validation.png"),
       height_comparison_plot, width = 10, height = 8, dpi = 300)
```

## 7. Dissection Validation

```{r dissection-validation, fig.width=10, fig.height=8}
# Join dissection data with processed internode data
# Note: dissection uses 'from_top' column, not 'internode'
compare_dissection <- dissection %>%
  left_join(phytomere_processed, by = c("plant", "from_top")) %>%
  filter(!is.na(length), !is.na(length_dissection))

cat("Dissection comparison records:", nrow(compare_dissection), "\n")

# Linear regression
dissection_model <- lm(length_dissection ~ length, data = compare_dissection)
slope_d <- round(coef(dissection_model)[2], 3)
intercept_d <- round(coef(dissection_model)[1], 3)
r_squared_d <- round(summary(dissection_model)$r.squared, 3)

# Create equation text
if (intercept_d >= 0) {
  equation_d <- paste0("y = ", slope_d, "x + ", intercept_d)
} else {
  equation_d <- paste0("y = ", slope_d, "x - ", abs(intercept_d))
}

dissection_plot <- compare_dissection %>%
  ggplot(aes(x = length, y = length_dissection)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  annotate("text",
           x = 15, y = 30,
           label = paste(equation_d, "\nR² =", r_squared_d),
           hjust = 1.1, vjust = 1.1,
           size = 6,
           fontface = "bold") +
  labs(
    title = "**Dissection Validation**",
    subtitle = "Comparing field measurements vs dissection measurements",
    x = "Field measurement (cm)",
    y = "Dissection measurement (cm)"
  ) +
  pheno_theme

print(dissection_plot)

ggsave(file.path(paths$figures, "internode_dissection_validation.pdf"),
       dissection_plot, width = 10, height = 8)
ggsave(file.path(paths$figures, "internode_dissection_validation.png"),
       dissection_plot, width = 10, height = 8, dpi = 300)
```

## 8. Statistical Summary

```{r statistical-summary}
# Genotype effects on internode length by position
internode_effects <- phytomere_processed %>%
  group_by(donor, from_top) %>%
  filter(n_distinct(genotype) == 2) %>%
  summarise(
    ctrl_mean = mean(length[genotype == "CTRL"], na.rm = TRUE),
    inv4m_mean = mean(length[genotype == "INV4M"], na.rm = TRUE),
    diff = inv4m_mean - ctrl_mean,
    p_value = tryCatch(
      t.test(length ~ genotype)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cat("\nInternode length effects by position (Inv4m - CTRL):\n")
print(internode_effects %>% filter(!is.na(p_value)))

# Export summary statistics
write_csv(summary_by_genotype, file.path(paths$intermediate, "internode_summary_by_genotype.csv"))
write_csv(node_count, file.path(paths$intermediate, "internode_node_count_by_plant.csv"))
write_csv(compare_height, file.path(paths$intermediate, "internode_height_comparison.csv"))
write_csv(internode_effects, file.path(paths$intermediate, "internode_position_effects.csv"))

cat("\nResults exported to:", paths$intermediate, "\n")
```

## 9. Internode-by-Position T-Tests with FDR Correction

```{r internode-ttest-fdr}
# Hypothesis: The difference in plant height is mainly due to internode 2 (just below tassel)
# Test: t-test for each internode position, corrected for multiple testing

# Focus on MI21 donor for the main analysis
internode_ttest <- phytomere_processed %>%
  filter(donor == "MI21") %>%
  group_by(from_top) %>%
  filter(n_distinct(genotype) == 2, n() >= 6) %>%  # Ensure both genotypes present with sufficient n
  t_test(length ~ genotype, detailed = TRUE) %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance("p.adj") %>%
  arrange(from_top)

cat("=== T-TESTS BY INTERNODE POSITION (MI21 donor) ===\n")
cat("H₀: No difference in internode length between CTRL and Inv4m\n")
cat("FDR correction applied across all positions\n\n")

print(internode_ttest %>%
        select(from_top, n1, n2, estimate, estimate1, estimate2, statistic, p, p.adj, p.adj.signif))

# Calculate effect sizes (difference and percent change)
internode_effects_detailed <- phytomere_processed %>%
  filter(donor == "MI21") %>%
  group_by(from_top) %>%
  filter(n_distinct(genotype) == 2) %>%
  summarise(
    n_ctrl = sum(genotype == "CTRL"),
    n_inv4m = sum(genotype == "INV4M"),
    ctrl_mean = mean(length[genotype == "CTRL"], na.rm = TRUE),
    inv4m_mean = mean(length[genotype == "INV4M"], na.rm = TRUE),
    diff_cm = inv4m_mean - ctrl_mean,
    pct_change = 100 * (inv4m_mean - ctrl_mean) / ctrl_mean,
    .groups = "drop"
  ) %>%
  left_join(internode_ttest %>% select(from_top, p, p.adj, p.adj.signif), by = "from_top") %>%
  arrange(from_top)

cat("\n=== EFFECT SIZES BY INTERNODE POSITION ===\n")
print(internode_effects_detailed)

# Summarize contribution to total height difference
total_diff <- sum(internode_effects_detailed$diff_cm, na.rm = TRUE)
internode_contribution <- internode_effects_detailed %>%
  mutate(
    contribution_cm = diff_cm,
    contribution_pct = 100 * diff_cm / total_diff
  ) %>%
  select(from_top, diff_cm, contribution_pct, p.adj, p.adj.signif) %>%
  arrange(desc(abs(contribution_pct)))

cat("\n=== CONTRIBUTION TO TOTAL HEIGHT DIFFERENCE ===\n")
cat(sprintf("Total height difference (sum of internode diffs): %.2f cm\n\n", total_diff))
print(internode_contribution)

# Test hypothesis: Is internode 2 the main contributor?
internode2_contrib <- internode_contribution %>% filter(from_top == 2)
if (nrow(internode2_contrib) > 0) {
  cat("\n=== HYPOTHESIS TEST: Internode 2 contribution ===\n")
  cat(sprintf("Internode 2 difference: %.2f cm (%.1f%% of total)\n",
              internode2_contrib$diff_cm, internode2_contrib$contribution_pct))
  cat(sprintf("FDR-adjusted p-value: %.4f (%s)\n",
              internode2_contrib$p.adj, internode2_contrib$p.adj.signif))
}

# Export results
write_csv(internode_ttest, file.path(paths$intermediate, "internode_ttest_by_position.csv"))
write_csv(internode_effects_detailed, file.path(paths$intermediate, "internode_effects_detailed.csv"))
write_csv(internode_contribution, file.path(paths$intermediate, "internode_contribution_to_height.csv"))

cat("\nResults exported to:", paths$intermediate, "\n")
```

### 9.1 Visualization of Significant Differences

```{r internode-significance-plot, fig.width=10, fig.height=8}
# Create a bar plot showing internode contribution with significance markers
contribution_plot <- internode_contribution %>%
  mutate(
    from_top = factor(from_top),
    significant = p.adj < 0.05
  ) %>%
  ggplot(aes(x = from_top, y = diff_cm, fill = significant)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_text(aes(label = p.adj.signif,
                y = diff_cm + sign(diff_cm) * 0.3),
            size = 6, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "purple4", "FALSE" = "gray70"),
                    labels = c("TRUE" = "FDR < 0.05", "FALSE" = "ns"),
                    name = "Significance") +
  scale_x_discrete(limits = rev) +
  coord_flip() +
  labs(
    title = "**Internode Length Difference (*Inv4m* − CTRL)**",
    subtitle = "MI21 donor; significance after FDR correction",
    x = "Internode position (from top)",
    y = "Difference (cm)"
  ) +
  theme_classic(base_size = 18) +
  theme(
    plot.title = element_markdown(),
    plot.subtitle = element_text(size = 14),
    legend.position = "top"
  )

print(contribution_plot)

ggsave(file.path(paths$figures, "internode_contribution_barplot.pdf"),
       contribution_plot, width = 10, height = 8)
ggsave(file.path(paths$figures, "internode_contribution_barplot.png"),
       contribution_plot, width = 10, height = 8, dpi = 300)
```

## 11. Key Findings

```{r key-findings}
cat("=== KEY FINDINGS ===\n\n")

# Node count differences
node_count_summary <- node_count %>%
  group_by(donor, genotype) %>%
  summarise(mean_nodes = mean(n_nodes), .groups = "drop") %>%
  pivot_wider(names_from = genotype, values_from = mean_nodes) %>%
  mutate(diff = INV4M - CTRL)

cat("1. Node Count Differences:\n")
print(node_count_summary)

# Overall internode length effect
overall_effect <- phytomere_processed %>%
  group_by(genotype) %>%
  summarise(mean_length = mean(length, na.rm = TRUE), .groups = "drop")

cat("\n2. Overall Mean Internode Length:\n")
print(overall_effect)

# Height validation
cat("\n3. Height Validation:\n")
cat(sprintf("   Sum of internodes explains %.1f%% of plant height variance\n", r_squared * 100))
```

## 12. Session Information

```{r session-info}
sessionInfo()
```
