---
title: "JMJ Pink Module Characterization"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(

  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)

knitr::opts_knit$set(root.dir = here::here())
```

# Overview

**Purpose**: Characterize the pink WGCNA module, which contains the JMJ2/JMJ4
histone demethylase cluster. This post-hoc analysis explores the functional
coherence of genes co-expressed with JMJ2/JMJ4 in CTRL samples.

**Key findings**:

- The pink module contains 30 DEGs including jmj2 and jmj4
- GO enrichment reveals "positive regulation of growth" driven by jmj2/jmj4
- Module includes DNA replication (rfc3), ribosome biogenesis, and chromatin genes
- Supports the hypothesis that JMJ2/JMJ4 are epigenetic regulators of growth

---

# Setup

```{r load_packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(knitr)
})

# Load project paths
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

```{r define_paths}
# Field perturbation run directory
run_dir <- here("results/inversion_paper/field_perturbation/run_20251231_201332")

# Input files
CONSENSUS_FILE <- file.path(run_dir, "04_consensus_networks/ctrl/consensus_modules.csv")
GO_BP_FILE <- file.path(run_dir, "07_module_annotation/all_modules_GO_BP.csv")
LABELS_FILE <- here("data", "locus_labels_master.csv")

# Verify files exist
stopifnot(file.exists(CONSENSUS_FILE))
stopifnot(file.exists(GO_BP_FILE))
stopifnot(file.exists(LABELS_FILE))

cat("All input files found.\n")
```

---

# Load Data

```{r load_data}
# Load CTRL consensus module assignments
ctrl_modules <- read_csv(CONSENSUS_FILE, show_col_types = FALSE)

# Load GO BP enrichment results
go_bp <- read_csv(GO_BP_FILE, show_col_types = FALSE)

# Load gene annotations with expression data
labels <- read_csv(LABELS_FILE, show_col_types = FALSE)

cat("Loaded", nrow(ctrl_modules), "genes with module assignments\n")
cat("Loaded", nrow(go_bp), "GO BP enrichment terms\n")
cat("Loaded", nrow(labels), "gene annotations\n")
```

---

# Pink Module Genes

The pink module contains jmj2 and jmj4, the JUMONJI histone demethylases that
are consistently downregulated in Inv4m plants.

```{r extract_pink_module}
# Extract pink module genes
pink_genes <- ctrl_modules %>%
  filter(color == "pink") %>%
  pull(gene)

cat("Pink module contains", length(pink_genes), "genes\n")

# Verify jmj2/jmj4 are present
jmj_genes <- c("Zm00001eb191790", "Zm00001eb191820")
cat("jmj2 in pink module:", jmj_genes[1] %in% pink_genes, "\n")
cat("jmj4 in pink module:", jmj_genes[2] %in% pink_genes, "\n")
```

---

# Annotated DEG Table

Table sorted by statistical significance (FDR, ascending = most significant first).

```{r annotate_pink_genes}
# Join pink genes with annotations
pink_annotated <- ctrl_modules %>%
  filter(color == "pink") %>%
  left_join(labels, by = "gene") %>%
  mutate(
    # Use locus_label if available; fall back to locus_symbol only when it is
    # not a marker (is_marker=TRUE means the symbol is a clone/marker ID
    # like cl59140_1, which is uninformative as a display label).
    symbol = case_when(
      !is.na(locus_label) & locus_label != "" ~ locus_label,
      !is.na(locus_symbol) & locus_symbol != "" & (is.na(is_marker) | !is_marker) ~ locus_symbol,
      TRUE ~ "—"
    ),
    # Clean up description - use desc_merged or desc_pannzer
    description = case_when(
      !is.na(desc_merged) & desc_merged != "" ~ desc_merged,
      !is.na(desc_pannzer) & desc_pannzer != "" ~ desc_pannzer,
      TRUE ~ "—"
    ),
    # Format log2FC and FDR
    log2FC = round(logFC, 2),
    FDR_val = adj.P.Val
  ) %>%
  # Sort by FDR (most significant first)
  arrange(FDR_val) %>%
  select(gene, symbol, description, log2FC, FDR_val)

# Display count
cat("Annotated", nrow(pink_annotated), "pink module genes\n")
```

```{r display_table}
# Create formatted table
pink_table <- pink_annotated %>%
  mutate(
    FDR = formatC(FDR_val, format = "e", digits = 1),
    # Truncate long descriptions
    description = str_trunc(description, 50)
  ) %>%
  select(gene, symbol, description, log2FC, FDR)

# Display with kable
kable(
  pink_table,
  col.names = c("Gene ID", "Label", "Description", "log2FC", "FDR"),
  caption = "Pink module DEGs sorted by statistical significance (FDR)"
)
```

---

# GO Enrichment for Pink Module

```{r go_enrichment}
# Filter GO BP results for pink module
pink_go <- go_bp %>%
  filter(module_color == "pink") %>%
  arrange(pvalue) %>%
  select(ID, Description, GeneRatio, pvalue, Count, geneID) %>%
  mutate(
    pvalue_fmt = formatC(pvalue, format = "e", digits = 2),
    Description = str_trunc(Description, 60)
  )

cat("Pink module has", nrow(pink_go), "enriched GO BP terms\n")
```

```{r display_go_table}
# Display top GO terms
kable(
  pink_go %>%
    head(15) %>%
    select(ID, Description, GeneRatio, pvalue_fmt, Count),
  col.names = c("GO ID", "Description", "Gene Ratio", "p-value", "Count"),
  caption = "Top 15 GO Biological Process terms for pink module"
)
```

---

# Growth-Related Genes

Highlight genes related to growth, cell cycle, and cell proliferation.

```{r growth_genes}
# Define growth-related GO terms
growth_go_ids <- pink_go %>%
  filter(grepl("growth|proliferation|cell cycle|chromatin|epigenetic|ribosom",
               Description, ignore.case = TRUE)) %>%
  pull(ID)

# Extract genes from these terms
growth_gene_string <- pink_go %>%
  filter(ID %in% growth_go_ids) %>%
  pull(geneID) %>%
  paste(collapse = "/")

growth_genes <- unique(unlist(str_split(growth_gene_string, "/")))

cat("Growth-related genes in pink module:", length(growth_genes), "\n")

# Annotate growth genes
growth_annotated <- pink_annotated %>%
  filter(gene %in% growth_genes) %>%
  mutate(
    FDR = formatC(FDR_val, format = "e", digits = 1)
  ) %>%
  select(gene, symbol, description, log2FC, FDR)

kable(
  growth_annotated,
  col.names = c("Gene ID", "Symbol", "Description", "log2FC", "FDR"),
  caption = "Pink module genes in growth-related GO terms"
)
```

---

# Functional Categories

```{r functional_categories}
# Manually categorize key genes
categories <- tribble(
  ~category, ~genes, ~description,
  "Epigenetic regulators", "jmj2, jmj4, swc6", "Histone demethylases and chromatin remodeling",
  "DNA replication", "rfc3", "Replication factor C - cell cycle progression",
  "Ribosome biogenesis", "rpl29, rps8, rps18, ddx14, noc2, bcd135b", "Protein synthesis capacity for growth",
  "Translation initiation", "eif2b, eif2b2, lurp8", "Translation machinery",
  "Growth signaling", "bpg2", "Brassinosteroid pathway - growth regulation"
)

kable(
  categories,
  col.names = c("Category", "Genes", "Biological Role"),
  caption = "Functional categories in the pink module"
)
```

---

# Summary

The pink module represents a coherent **growth/chromatin regulatory network**
centered on the JMJ2/JMJ4 histone demethylases:

1. **Epigenetic control**: jmj2/jmj4 (histone demethylation) and swc6 (chromatin remodeling)
2. **Cell proliferation machinery**: rfc3 (DNA replication), ribosomal proteins
3. **Translation capacity**: Multiple translation initiation factors
4. **Growth signaling**: bpg2 (brassinosteroid pathway)

The GO enrichment confirms:

- **"Positive regulation of growth"** (p = 0.0017) - driven by jmj2/jmj4
- **"Ribosome biogenesis"** (p = 0.0003) - protein synthesis capacity
- **"Chromatin remodeling"** (p = 0.013) - epigenetic regulation

This supports the hypothesis that JMJ2/JMJ4 downregulation in Inv4m disrupts
a coordinated program of epigenetic regulation and growth-related gene expression.

---

# Save Outputs

```{r save_outputs}
# Save annotated table (CSV)
write_csv(
  pink_annotated %>%
    mutate(FDR = formatC(FDR_val, format = "e", digits = 2)) %>%
    select(gene, symbol, description, log2FC, FDR),
  file.path(paths$intermediate, "pink_module_annotated_DEGs.csv")
)

# Save GO enrichment
write_csv(
  pink_go %>% select(-pvalue_fmt),
  file.path(paths$intermediate, "pink_module_GO_BP.csv")
)

cat("\nSaved outputs:\n")
cat("  - pink_module_annotated_DEGs.csv\n")
cat("  - pink_module_GO_BP.csv\n")
```

```{r generate_latex_table}
# Generate LaTeX table for main.tex
latex_header <- '\\begin{table}[!ht]
\\caption{\\textbf{Pink module DEGs co-expressed with \\jmjii/\\jmjiv.} Genes assigned to the pink WGCNA consensus module, which contains the JMJ2/JMJ4 histone demethylase cluster. Sorted by statistical significance (FDR). GO enrichment for this module includes ``positive regulation of growth\'\' ($p = 0.0017$) and ``ribosome biogenesis\'\' ($p = 0.0003$).}
\\label{tab::pink_module}
\\centering
\\small
\\begin{tabular}{@{}lllrr@{}}
\\toprule
\\textbf{Gene ID} & \\textbf{Label} & \\textbf{Description} & \\textbf{log$_2$FC} & \\textbf{FDR} \\\\
\\midrule'

latex_footer <- '\\bottomrule
\\end{tabular}
\\end{table}'

# Format each row
format_latex_row <- function(gene, symbol, desc, log2fc, fdr) {
  # Format label
  label <- if (is.na(symbol) || symbol == "" || symbol == "—") {
    "---"
  } else {
    paste0("\\textit{", symbol, "}")
  }

  # Format description
  desc_clean <- if (is.na(desc) || desc == "" || desc == "—") {
    "---"
  } else {
    str_trunc(desc, 40) %>%
      str_replace_all("_", "\\\\_") %>%
      str_replace_all("&", "\\\\&") %>%
      str_replace_all("%", "\\\\%")
  }

  # Format log2FC with sign
  log2fc_fmt <- if (log2fc >= 0) {
    paste0("$+", sprintf("%.2f", log2fc), "$")
  } else {
    paste0("$", sprintf("%.2f", log2fc), "$")
  }

  # Format FDR in scientific notation
  fdr_exp <- floor(log10(fdr))
  fdr_mant <- fdr / 10^fdr_exp
  fdr_fmt <- paste0("$", sprintf("%.1f", fdr_mant), " \\times 10^{", fdr_exp, "}$")

  paste(gene, label, desc_clean, log2fc_fmt, fdr_fmt, sep = " & ", collapse = "") %>%
    paste0(" \\\\")
}

# Generate all rows
latex_rows <- pink_annotated %>%
  rowwise() %>%
  mutate(
    latex_row = format_latex_row(gene, symbol, description, log2FC, FDR_val)
  ) %>%
  pull(latex_row)

# Combine into full table
latex_table <- paste(
  latex_header,
  paste(latex_rows, collapse = "\n"),
  latex_footer,
  sep = "\n"
)

# Save to tables folder
writeLines(latex_table, file.path(paths$tables, "pink_module_DEGs.tex"))

cat("\nLaTeX table saved:\n")
cat("  - tables/pink_module_DEGs.tex\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
