---
title: "JMJ Paralog Expression Boxplot (PSU2022)"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 5
)

# Load project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This script creates an expression boxplot for all 5 JMJ cluster paralogs using
transcript-level kallisto quantification from the JMJ-corrected cDNA reference.

**JMJ Cluster Paralogs (genomic order):**

| Transcript | Symbol | Description |
|------------|--------|-------------|
| Zm00001eb191790_T001 | jmj9 | First tandem copy |
| Zm00001eb191790_T017 | psi | Pseudogene (from V4 annotation) |
| Zm00001eb191790_T006 | jmj6 | Third tandem copy |
| Zm00001eb191790_T013 | jmj2 | Fourth tandem copy |
| Zm00001eb191820_T001 | jmj4 | Separate gene model |

**Background:** The V5 annotation incorrectly merges jmj9, psi, jmj6, and jmj2
into a single gene model (Zm00001eb191790) with 16 chimeric transcripts. We
created a corrected cDNA reference that treats each paralog as a separate entry,
enabling accurate transcript-level quantification.

## Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(ggfx)
library(ggbeeswarm)
```

## Define JMJ Paralogs

```{r jmj_paralogs}
# JMJ paralog transcripts (in genomic order)
jmj_paralogs <- tibble(
  transcript = c(
    "Zm00001eb191790_T001",
    "Zm00001eb191790_T017",
    "Zm00001eb191790_T006",
    "Zm00001eb191790_T013",
    "Zm00001eb191820_T001"
  ),
  symbol = c("jmj9", "psi", "jmj6", "jmj2", "jmj4"),
  position = c(1, 2, 3, 4, 5)
)

cat("JMJ cluster paralogs:\n")
print(jmj_paralogs)
```

# Load Kallisto Data

```{r config}
# Kallisto output directory (JMJ-corrected run)
quant_dir <- "/Volumes/rsstu/users/r/rrellan/DOE_CAREER/inv4m/PSU2022/kallisto/quant_out_jmj_corrected/kallisto_quant"

# Check server mount
if (!dir.exists(quant_dir)) {
  stop("Server not mounted or kallisto output not found.\n  ", quant_dir)
}
```

```{r load_kallisto}
cat("Loading kallisto transcript-level data...\n")

samples <- dir(quant_dir)
cat("Found", length(samples), "samples\n")

# Extract TPM for JMJ paralogs AND compute library sizes
kallisto_data <- lapply(samples, function(x) {
  abundance_file <- file.path(quant_dir, x, x, "abundance.tsv")

  if (!file.exists(abundance_file)) {
    warning("File not found: ", abundance_file)
    return(NULL)
  }

  abundance <- read.table(abundance_file, sep = "\t", header = TRUE)

  # Compute library size (total estimated counts)
  lib_size <- sum(abundance$est_counts)

  # Filter for JMJ paralogs
  jmj_data <- abundance %>%
    filter(target_id %in% jmj_paralogs$transcript) %>%
    select(target_id, tpm, est_counts) %>%
    mutate(sample = x, lib_size = lib_size)

  jmj_data
}) %>%
  bind_rows()

cat("Extracted", nrow(kallisto_data), "records\n")
```

## Filter Low Quality Libraries

Samples with library size < 20 million reads are considered low quality
(same threshold as DEG analysis).

```{r filter_quality}
# Get library sizes per sample
lib_sizes <- kallisto_data %>%
  distinct(sample, lib_size)

# Mark low quality samples
lib_sizes <- lib_sizes %>%
  mutate(low_quality = lib_size < 2e7)

{
  cat("=== LIBRARY SIZE QC ===\n")
  cat("Low quality libraries (< 20M reads):\n")
  print(table(lib_sizes$low_quality))

  if (any(lib_sizes$low_quality)) {
    cat("\nExcluded samples:\n")
    print(lib_sizes %>% filter(low_quality) %>% select(sample, lib_size))
  }
}

# Filter out low quality samples
good_samples <- lib_sizes %>%
  filter(!low_quality) %>%
  pull(sample)

jmj_tpm <- kallisto_data %>%
  filter(sample %in% good_samples) %>%
  select(-lib_size)

cat("\nRetained", n_distinct(jmj_tpm$sample), "samples\n")
```

# Add Metadata

```{r add_metadata}
# Load PSU2022 metadata
meta <- read.csv(file.path(paths$data, "PSU2022_metadata.csv"))

# Join with paralog info and metadata
jmj_expr <- jmj_tpm %>%
  left_join(jmj_paralogs, by = c("target_id" = "transcript")) %>%
  left_join(
    meta %>% select(library, Genotype, leaf_tissue),
    by = c("sample" = "library")
  ) %>%
  filter(!is.na(Genotype)) %>%
  mutate(
    genotype = ifelse(Genotype == "CTRL", "B73", "Inv4m"),
    tissue = paste0("Leaf ", leaf_tissue),
    # Order symbols by genomic position
    symbol = factor(symbol, levels = c("jmj9", "psi", "jmj6", "jmj2", "jmj4")),
    genotype = factor(genotype, levels = c("B73", "Inv4m"))
  )

{
  cat("=== DATA SUMMARY ===\n")
  cat("Samples:", n_distinct(jmj_expr$sample), "\n")
  cat("Paralogs:", n_distinct(jmj_expr$symbol), "\n")
  cat("\nBy genotype:\n")
  print(table(jmj_expr$genotype) / 5)
  cat("\nBy tissue:\n")
  print(table(jmj_expr$tissue) / 5)
}
```

# Expression Boxplot

```{r boxplot, fig.width=14, fig.height=5}
genotype_colors <- c("B73" = "gold", "Inv4m" = "#4a0f82")

p_jmj_paralogs <- ggplot(
  jmj_expr,
  aes(x = tissue, y = log2(tpm + 1), color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_wrap(~ symbol, nrow = 1) +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 11
    ),
    axis.text.y = element_text(color = "black", face = "bold"),
    axis.title = element_text(face = "bold", size = 13),
    strip.text = element_text(face = "bold.italic", size = 14),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(TPM + 1)"))
  )

print(p_jmj_paralogs)
```

# Summary Statistics

```{r summary_stats}
jmj_summary <- jmj_expr %>%
  group_by(tissue, symbol, genotype) %>%
  summarise(
    mean_tpm = mean(tpm, na.rm = TRUE),
    sd_tpm = sd(tpm, na.rm = TRUE),
    mean_log2 = mean(log2(tpm + 1), na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = genotype,
    values_from = c(mean_tpm, sd_tpm, mean_log2, n),
    names_sep = "_"
  ) %>%
  mutate(
    log2FC = mean_log2_Inv4m - mean_log2_B73,
    direction = ifelse(log2FC < 0, "Down", "Up")
  )

cat("=== EXPRESSION SUMMARY ===\n")
jmj_summary %>%
  select(tissue, symbol, mean_tpm_B73, mean_tpm_Inv4m, log2FC, direction) %>%
  arrange(symbol, tissue) %>%
  print(n = 25)
```

## Mean log2FC by Paralog

```{r mean_fc}
jmj_summary %>%
  group_by(symbol) %>%
  summarise(
    mean_log2FC = mean(log2FC),
    .groups = "drop"
  ) %>%
  arrange(symbol)
```

# Save Outputs

```{r save_outputs}
# Save plot
saveRDS(p_jmj_paralogs, file.path(paths$figures, "jmj_paralogs_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "jmj_paralogs_expression_boxplot.pdf"),
  p_jmj_paralogs,
  width = 14, height = 5
)
ggsave(
  file.path(paths$figures, "jmj_paralogs_expression_boxplot.png"),
  p_jmj_paralogs,
  width = 14, height = 5, dpi = 300
)

# Save data
write_csv(jmj_expr, file.path(paths$intermediate, "jmj_paralogs_expression_data.csv"))
write_csv(jmj_summary, file.path(paths$intermediate, "jmj_paralogs_expression_summary.csv"))

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("Figures:\n")
  cat("  jmj_paralogs_expression_boxplot.pdf\n")
  cat("  jmj_paralogs_expression_boxplot.png\n")
  cat("  jmj_paralogs_expression_boxplot.rds\n")
  cat("\nData:\n")
  cat("  jmj_paralogs_expression_data.csv\n")
  cat("  jmj_paralogs_expression_summary.csv\n")
}
```

# Session Info

```{r session_info}
sessionInfo()
```
