---
title: "JMJ Paralog Expression Boxplot (PSU2022)"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 5
)

# Load project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This script analyzes JMJ cluster expression from transcript-level kallisto
quantification using the JMJ-corrected cDNA reference.

**Two analyses are performed:**

1. **Total Cluster Expression** (main analysis): Sum of CPM across all 5 paralogs
   per sample. This is the biologically meaningful comparison because:
   - **B73/CTRL**: Has 5 tandem JMJ gene copies (>97% identical)
   - **Inv4m**: Has only 1 gene copy in the cluster

   Comparing total cluster output shows whether the single Inv4m copy produces
   as much transcript as the 5 B73 copies combined.

2. **Per-Paralog Expression**: Individual expression for each paralog. Note that
   because the paralogs are >97% identical, kallisto distributes Inv4m reads
   across all 5 loci artificially. This view is useful for understanding read
   distribution but not for biological interpretation of Inv4m expression.

**JMJ Cluster Paralogs (genomic order):**

| Transcript | Symbol | Description |
|------------|--------|-------------|
| Zm00001eb191790_T001 | jmj9 | First tandem copy |
| Zm00001eb191790_T017 | psi | Pseudogene (from V4 annotation) |
| Zm00001eb191790_T006 | jmj6 | Third tandem copy |
| Zm00001eb191790_T013 | jmj2 | Fourth tandem copy |
| Zm00001eb191820_T001 | jmj4 | Separate gene model |

**Background:** The V5 annotation incorrectly merges jmj9, psi, jmj6, and jmj2
into a single gene model (Zm00001eb191790) with 16 chimeric transcripts. We
created a corrected cDNA reference that treats each paralog as a separate entry.

## Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(ggfx)
library(ggbeeswarm)
```

## Define JMJ Paralogs

```{r jmj_paralogs}
# JMJ paralog transcripts (in genomic order)
jmj_paralogs <- tibble(
  transcript = c(
    "Zm00001eb191790_T001",
    "Zm00001eb191790_T017",
    "Zm00001eb191790_T006",
    "Zm00001eb191790_T013",
    "Zm00001eb191820_T001"
  ),
  symbol = c("jmj9", "psi", "jmj6", "jmj2", "jmj4"),
  position = c(1, 2, 3, 4, 5)
)

cat("JMJ cluster paralogs:\n")
print(jmj_paralogs)
```

# Load Counts Matrix

The counts matrix is pre-extracted from kallisto outputs using the standalone
script `extract_kallisto_counts_matrix.R`. This analysis reads from the cached
intermediate file only.

```{r load_counts}
# Counts matrix path (extracted from kallisto by separate script)
counts_matrix_file <- file.path(
  paths$intermediate,
  "PSU2022_kallisto_jmj_corrected_counts.csv"
)

# Check if counts matrix exists
if (!file.exists(counts_matrix_file)) {
  stop(
    "Counts matrix not found: ", counts_matrix_file, "\n\n",
    "Run the extraction script first:\n",
    "  Rscript scripts/inversion_paper/extract_kallisto_counts_matrix.R\n",
    "(Requires server mounted)"
  )
}

cat("Loading counts matrix...\n")
full_counts_matrix <- read_csv(counts_matrix_file, show_col_types = FALSE)
cat("Loaded:", nrow(full_counts_matrix), "transcripts x",
    ncol(full_counts_matrix) - 1, "samples\n")
```

## Extract JMJ Paralogs and Calculate CPM

```{r extract_jmj}
# Calculate library sizes from full counts matrix
lib_sizes_df <- full_counts_matrix %>%
  pivot_longer(cols = -target_id, names_to = "sample", values_to = "counts") %>%
  group_by(sample) %>%
  summarise(lib_size = sum(counts, na.rm = TRUE), .groups = "drop")

cat("Library sizes calculated for", nrow(lib_sizes_df), "samples\n")

# Extract JMJ counts (long format)
kallisto_data <- full_counts_matrix %>%
  filter(target_id %in% jmj_paralogs$transcript) %>%
  pivot_longer(
    cols = -target_id,
    names_to = "sample",
    values_to = "est_counts"
  ) %>%
  left_join(lib_sizes_df, by = "sample") %>%
  mutate(
    # Calculate CPM (counts per million)
    cpm = (est_counts / lib_size) * 1e6
  )

cat("Extracted JMJ paralogs:", nrow(kallisto_data), "records\n")
cat("(", n_distinct(kallisto_data$target_id), "transcripts x",
    n_distinct(kallisto_data$sample), "samples )\n")
```

## Filter Low Quality Libraries

Samples with library size < 20 million reads are considered low quality
(same threshold as DEG analysis).

```{r filter_quality}
# Get library sizes per sample
lib_sizes <- kallisto_data %>%
  distinct(sample, lib_size)

# Mark low quality samples
lib_sizes <- lib_sizes %>%
  mutate(low_quality = lib_size < 2e7)

{
  cat("=== LIBRARY SIZE QC ===\n")
  cat("Low quality libraries (< 20M reads):\n")
  print(table(lib_sizes$low_quality))

  if (any(lib_sizes$low_quality)) {
    cat("\nExcluded samples:\n")
    print(lib_sizes %>% filter(low_quality) %>% select(sample, lib_size))
  }
}

# Filter out low quality samples
good_samples <- lib_sizes %>%
  filter(!low_quality) %>%
  pull(sample)

jmj_filtered <- kallisto_data %>%
  filter(sample %in% good_samples)

cat("\nRetained", n_distinct(jmj_filtered$sample), "samples\n")
```

# Add Metadata

```{r add_metadata}
# Load PSU2022 metadata
meta <- read.csv(file.path(paths$data, "PSU2022_metadata.csv"))

# Join with paralog info and metadata
jmj_expr <- jmj_filtered %>%
  left_join(jmj_paralogs, by = c("target_id" = "transcript")) %>%
  left_join(
    meta %>% select(library, Genotype, leaf_tissue),
    by = c("sample" = "library")
  ) %>%
  filter(!is.na(Genotype)) %>%
  mutate(
    genotype = ifelse(Genotype == "CTRL", "B73", "Inv4m"),
    tissue = paste0("Leaf ", leaf_tissue),
    # Order symbols by genomic position
    symbol = factor(symbol, levels = c("jmj9", "psi", "jmj6", "jmj2", "jmj4")),
    genotype = factor(genotype, levels = c("B73", "Inv4m"))
  )

{
  cat("=== DATA SUMMARY ===\n")
  cat("Samples:", n_distinct(jmj_expr$sample), "\n")
  cat("Paralogs:", n_distinct(jmj_expr$symbol), "\n")
  cat("\nBy genotype:\n")
  print(table(jmj_expr$genotype) / 5)
  cat("\nBy tissue:\n")
  print(table(jmj_expr$tissue) / 5)
}
```

# Per-Paralog Expression Boxplot

This shows expression distribution per paralog. **Note:** Because JMJ paralogs
are >97% identical, kallisto artificially distributes Inv4m reads across all 5
loci. See "Total Cluster Expression" section below for the main analysis.

```{r boxplot, fig.width=14, fig.height=5}
genotype_colors <- c("B73" = "gold", "Inv4m" = "#4a0f82")

p_jmj_paralogs <- ggplot(
  jmj_expr,
  aes(x = tissue, y = log2(cpm + 1), color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_wrap(~ symbol, nrow = 1) +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 11
    ),
    axis.text.y = element_text(color = "black", face = "bold"),
    axis.title = element_text(face = "bold", size = 13),
    strip.text = element_text(face = "bold.italic", size = 14),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM + 1)"))
  )

print(p_jmj_paralogs)
```

# Total Cluster Expression Analysis

Since B73/CTRL has 5 tandem JMJ paralogs (>97% identical) while Inv4m has only
1 gene copy, the biologically meaningful comparison is **total cluster expression**:
the sum of CPM across all 5 paralogs per sample.

```{r total_cluster_expression}
# Sum CPM across all 5 paralogs for each sample
jmj_total <- jmj_expr %>%
  group_by(sample, genotype, tissue) %>%
  summarise(
    total_cpm = sum(cpm, na.rm = TRUE),
    total_counts = sum(est_counts, na.rm = TRUE),
    n_paralogs = n(),
    .groups = "drop"
  )

cat("=== TOTAL CLUSTER EXPRESSION ===\n")
cat("Samples:", n_distinct(jmj_total$sample), "\n")
cat("Mean total CPM by genotype:\n")
jmj_total %>%
  group_by(genotype) %>%
  summarise(
    mean_total_cpm = mean(total_cpm),
    sd_total_cpm = sd(total_cpm),
    .groups = "drop"
  ) %>%
  print()
```

## Total Cluster Expression Boxplot

This plot shows the **sum of expression across all 5 JMJ paralogs** for each sample,
comparing B73 (5 gene copies) vs Inv4m (1 gene copy).

```{r total_boxplot, fig.width=8, fig.height=6}
p_jmj_total <- ggplot(
  jmj_total,
  aes(x = tissue, y = log2(total_cpm + 1), color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 2,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 13
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40")
  ) +
  labs(
    title = "Total JMJ Cluster Expression",
    subtitle = "Sum of 5 paralogs: B73 (5 copies) vs Inv4m (1 copy)",
    x = NULL,
    y = expression(bold(log[2]*"(Total CPM + 1)"))
  )

print(p_jmj_total)
```

## Total Cluster Expression Boxplot (Linear Scale)

Same data as above but on linear scale to emphasize the magnitude of difference.
The dashed line shows the expected Inv4m expression if it scaled linearly with
copy number (B73 mean / 5).

```{r calculate_expected, include=FALSE}
# Calculate expected Inv4m expression (B73 / 5) for each tissue
expected_df <- jmj_total %>%
  filter(genotype == "B73") %>%
  group_by(tissue) %>%
  summarise(
    B73_mean = mean(total_cpm),
    expected_inv4m = B73_mean / 5,
    .groups = "drop"
  )
```

```{r total_boxplot_linear, fig.width=8, fig.height=6}
# Add expected values to data for plotting
expected_df <- expected_df %>%
  mutate(tissue = factor(tissue, levels = levels(factor(jmj_total$tissue))))

p_jmj_total_linear <- ggplot(
  jmj_total,
  aes(x = tissue, y = total_cpm, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 2,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  # Add expected line (B73/5) - dashed red crossbars
  geom_errorbar(
    data = expected_df,
    aes(x = tissue, ymin = expected_inv4m, ymax = expected_inv4m),
    color = "red", linetype = "dashed", linewidth = 1,
    width = 0.6, inherit.aes = FALSE
  ) +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 13
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40")
  ) +
  labs(
    title = "Total JMJ Cluster Expression (Linear Scale)",
    subtitle = "Dashed red line = expected if Inv4m = B73/5 (copy number scaling)",
    x = NULL,
    y = "Total CPM"
  )

print(p_jmj_total_linear)
```

## One-Sample t-Test: Inv4m vs Expected (B73/5)

**Null hypothesis:** Inv4m mean expression = B73 mean / 5 (linear copy number scaling)

```{r ttest_expected}
# Calculate summary stats by tissue and genotype
tissue_stats <- jmj_total %>%
  group_by(tissue, genotype) %>%
  summarise(
    mean_cpm = mean(total_cpm),
    sd_cpm = sd(total_cpm),
    n = n(),
    values = list(total_cpm),
    .groups = "drop"
  )

# Get B73 stats and calculate expected
b73_stats <- tissue_stats %>%
  filter(genotype == "B73") %>%
  select(tissue, B73_mean = mean_cpm, B73_sd = sd_cpm, B73_n = n) %>%
  mutate(expected = B73_mean / 5)

# Get Inv4m stats
inv4m_stats <- tissue_stats %>%
  filter(genotype == "Inv4m") %>%
  select(tissue, Inv4m_mean = mean_cpm, Inv4m_sd = sd_cpm, Inv4m_n = n,
         inv4m_values = values)

# Merge and run t-tests
ttest_results <- b73_stats %>%
  left_join(inv4m_stats, by = "tissue") %>%
  rowwise() %>%
  mutate(
    ttest = list(t.test(inv4m_values, mu = expected)),
    t_stat = ttest$statistic,
    p_value = ttest$p.value,
    ci_lower = ttest$conf.int[1],
    ci_upper = ttest$conf.int[2],
    pct_of_expected = (Inv4m_mean / expected) * 100
  ) %>%
  ungroup() %>%
  select(-inv4m_values, -ttest)

# Display results table
cat("=== ONE-SAMPLE T-TEST: Inv4m vs Expected (B73/5) ===\n\n")

ttest_table <- ttest_results %>%
  select(tissue, B73_mean, expected, Inv4m_mean, pct_of_expected, t_stat, p_value) %>%
  mutate(
    B73_mean = round(B73_mean, 1),
    expected = round(expected, 1),
    Inv4m_mean = round(Inv4m_mean, 1),
    pct_of_expected = paste0(round(pct_of_expected, 1), "%"),
    t_stat = round(t_stat, 2),
    p_value = format(p_value, digits = 2, scientific = TRUE)
  ) %>%
  rename(
    Tissue = tissue,
    `B73 Mean` = B73_mean,
    `Expected (B73/5)` = expected,
    `Inv4m Mean` = Inv4m_mean,
    `% of Expected` = pct_of_expected,
    `t` = t_stat,
    `p-value` = p_value
  )

knitr::kable(ttest_table, align = "lrrrcrr")
```

```{r ttest_overall}
# Overall test (pooling across tissues)
overall_B73_mean <- mean(jmj_total$total_cpm[jmj_total$genotype == "B73"])
overall_expected <- overall_B73_mean / 5
overall_Inv4m_values <- jmj_total$total_cpm[jmj_total$genotype == "Inv4m"]

overall_ttest <- t.test(overall_Inv4m_values, mu = overall_expected)

cat("\n=== OVERALL TEST (pooled across tissues) ===\n\n")
cat("B73 mean:           ", round(overall_B73_mean, 2), "CPM\n")
cat("Expected (B73/5):   ", round(overall_expected, 2), "CPM\n")
cat("Inv4m mean:         ", round(mean(overall_Inv4m_values), 2), "CPM\n")
cat("% of expected:      ", round(mean(overall_Inv4m_values) / overall_expected * 100, 1), "%\n\n")
cat("One-sample t-test:\n")
cat("  t =", round(overall_ttest$statistic, 2), "\n")
cat("  df =", round(overall_ttest$parameter, 1), "\n")
cat("  p =", format(overall_ttest$p.value, digits = 3, scientific = TRUE), "\n")
cat("  95% CI: [", round(overall_ttest$conf.int[1], 2), ",",
    round(overall_ttest$conf.int[2], 2), "]\n")
```

## Total Cluster Summary Statistics

```{r total_summary}
jmj_total_summary <- jmj_total %>%
  group_by(tissue, genotype) %>%
  summarise(
    mean_total_cpm = mean(total_cpm, na.rm = TRUE),
    sd_total_cpm = sd(total_cpm, na.rm = TRUE),
    mean_log2 = mean(log2(total_cpm + 1), na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = genotype,
    values_from = c(mean_total_cpm, sd_total_cpm, mean_log2, n),
    names_sep = "_"
  ) %>%
  mutate(
    log2FC = mean_log2_Inv4m - mean_log2_B73,
    fold_change = 2^log2FC,
    direction = ifelse(log2FC < 0, "Down", "Up")
  )

cat("=== TOTAL CLUSTER EXPRESSION BY TISSUE ===\n")
jmj_total_summary %>%
  select(tissue, mean_total_cpm_B73, mean_total_cpm_Inv4m, log2FC, fold_change, direction) %>%
  print()

cat("\n=== OVERALL MEAN ===\n")
cat("Mean log2FC across tissues:", round(mean(jmj_total_summary$log2FC), 3), "\n")
cat("Mean fold change:", round(mean(jmj_total_summary$fold_change), 3), "\n")
```

## Statistical Test (Total Cluster Expression)

```{r total_stats}
# Test for genotype effect on total cluster expression
cat("=== STATISTICAL ANALYSIS ===\n\n")

# Overall t-test (ignoring tissue)
t_overall <- t.test(
  log2(total_cpm + 1) ~ genotype,
  data = jmj_total
)
cat("Overall t-test (B73 vs Inv4m):\n")
cat("  t =", round(t_overall$statistic, 3), "\n")
cat("  p =", format(t_overall$p.value, digits = 3, scientific = TRUE), "\n")
cat("  Mean B73:", round(t_overall$estimate[1], 3), "\n")
cat("  Mean Inv4m:", round(t_overall$estimate[2], 3), "\n")

# ANOVA with tissue interaction
model <- aov(log2(total_cpm + 1) ~ genotype * tissue, data = jmj_total)
cat("\n\nTwo-way ANOVA (genotype × tissue):\n")
print(summary(model))
```

## Chi-Square Test: Does Inv4m Expression Match Copy Number Expectation?

**Null hypothesis:** Expression scales linearly with copy number.
If B73 has 5 copies and Inv4m has 1 copy, then under H0:
- Expected ratio = 5:1 (B73 : Inv4m)
- Inv4m should contribute 1/6 = 16.7% of combined expression

```{r chisq_test}
# Get total counts (summed across all samples) for each genotype
total_by_genotype <- jmj_total %>%
  group_by(genotype) %>%
  summarise(
    total_counts = sum(total_counts),
    n_samples = n(),
    .groups = "drop"
  )

cat("=== CHI-SQUARE GOODNESS OF FIT TEST ===\n\n")
cat("H0: Expression ratio matches copy number ratio (5:1)\n")
cat("H1: Inv4m expression is less than expected from copy number\n\n")

# Observed counts
obs_B73 <- total_by_genotype$total_counts[total_by_genotype$genotype == "B73"]
obs_Inv4m <- total_by_genotype$total_counts[total_by_genotype$genotype == "Inv4m"]
total_obs <- obs_B73 + obs_Inv4m

cat("Observed total counts:\n")
cat("  B73:   ", format(round(obs_B73), big.mark = ","), "\n")
cat("  Inv4m: ", format(round(obs_Inv4m), big.mark = ","), "\n")
cat("  Total: ", format(round(total_obs), big.mark = ","), "\n\n")

# Observed proportion
obs_prop_inv4m <- obs_Inv4m / total_obs
cat("Observed Inv4m proportion:", round(obs_prop_inv4m * 100, 2), "%\n")
cat("Expected under H0 (1/6):  ", round(100/6, 2), "%\n\n")

# Expected counts under H0 (5:1 ratio = 5/6 : 1/6)
exp_B73 <- total_obs * (5/6)
exp_Inv4m <- total_obs * (1/6)

cat("Expected counts under H0 (5:1 ratio):\n")
cat("  B73:   ", format(round(exp_B73), big.mark = ","), "\n")
cat("  Inv4m: ", format(round(exp_Inv4m), big.mark = ","), "\n\n")

# Chi-square goodness of fit test
observed <- c(obs_B73, obs_Inv4m)
expected_prop <- c(5/6, 1/6)

chisq_result <- chisq.test(observed, p = expected_prop)

cat("Chi-square test result:\n")
cat("  X² =", round(chisq_result$statistic, 2), "\n")
cat("  df =", chisq_result$parameter, "\n")
cat("  p  =", format(chisq_result$p.value, digits = 3, scientific = TRUE), "\n\n")

# Effect size: how much less than expected?
deficit <- (exp_Inv4m - obs_Inv4m) / exp_Inv4m * 100
cat("Inv4m expression deficit:\n")
cat("  Observed:", format(round(obs_Inv4m), big.mark = ","), "counts\n")
cat("  Expected:", format(round(exp_Inv4m), big.mark = ","), "counts\n")
cat("  Deficit: ", round(deficit, 1), "% below expectation\n")
```

## Alternative: Per-Tissue Chi-Square Tests

```{r chisq_by_tissue}
cat("=== PER-TISSUE CHI-SQUARE TESTS ===\n\n")

tissue_chisq <- jmj_total %>%
  group_by(tissue, genotype) %>%
  summarise(total_counts = sum(total_counts), .groups = "drop") %>%
  pivot_wider(names_from = genotype, values_from = total_counts) %>%
  mutate(
    total = B73 + Inv4m,
    obs_prop_inv4m = Inv4m / total,
    exp_Inv4m = total * (1/6),
    exp_B73 = total * (5/6)
  ) %>%
  rowwise() %>%
  mutate(
    chisq = chisq.test(c(B73, Inv4m), p = c(5/6, 1/6))$statistic,
    p_value = chisq.test(c(B73, Inv4m), p = c(5/6, 1/6))$p.value,
    deficit_pct = (exp_Inv4m - Inv4m) / exp_Inv4m * 100
  ) %>%
  ungroup()

tissue_chisq %>%
  select(tissue, obs_prop_inv4m, deficit_pct, chisq, p_value) %>%
  mutate(
    obs_prop_inv4m = paste0(round(obs_prop_inv4m * 100, 1), "%"),
    deficit_pct = paste0(round(deficit_pct, 1), "%"),
    chisq = round(chisq, 1),
    p_value = format(p_value, digits = 2, scientific = TRUE)
  ) %>%
  rename(
    `Inv4m %` = obs_prop_inv4m,
    `Deficit` = deficit_pct,
    `X²` = chisq,
    `p-value` = p_value
  ) %>%
  print()

cat("\nExpected Inv4m % under H0: 16.7%\n")
```

# Per-Paralog Summary Statistics

```{r summary_stats}
jmj_summary <- jmj_expr %>%
  group_by(tissue, symbol, genotype) %>%
  summarise(
    mean_cpm = mean(cpm, na.rm = TRUE),
    sd_cpm = sd(cpm, na.rm = TRUE),
    mean_log2 = mean(log2(cpm + 1), na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = genotype,
    values_from = c(mean_cpm, sd_cpm, mean_log2, n),
    names_sep = "_"
  ) %>%
  mutate(
    log2FC = mean_log2_Inv4m - mean_log2_B73,
    direction = ifelse(log2FC < 0, "Down", "Up")
  )

cat("=== EXPRESSION SUMMARY ===\n")
jmj_summary %>%
  select(tissue, symbol, mean_cpm_B73, mean_cpm_Inv4m, log2FC, direction) %>%
  arrange(symbol, tissue) %>%
  print(n = 25)
```

## Mean log2FC by Paralog

```{r mean_fc}
jmj_summary %>%
  group_by(symbol) %>%
  summarise(
    mean_log2FC = mean(log2FC),
    .groups = "drop"
  ) %>%
  arrange(symbol)
```

# Save Outputs

```{r save_outputs}
# Save per-paralog plot
saveRDS(p_jmj_paralogs, file.path(paths$figures, "jmj_paralogs_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "jmj_paralogs_expression_boxplot.pdf"),
  p_jmj_paralogs,
  width = 14, height = 5
)
ggsave(
  file.path(paths$figures, "jmj_paralogs_expression_boxplot.png"),
  p_jmj_paralogs,
  width = 14, height = 5, dpi = 300
)

# Save total cluster expression plot (log scale)
saveRDS(p_jmj_total, file.path(paths$figures, "jmj_total_cluster_expression.rds"))
ggsave(
  file.path(paths$figures, "jmj_total_cluster_expression.pdf"),
  p_jmj_total,
  width = 8, height = 6
)
ggsave(
  file.path(paths$figures, "jmj_total_cluster_expression.png"),
  p_jmj_total,
  width = 8, height = 6, dpi = 300
)

# Save total cluster expression plot (linear scale)
saveRDS(p_jmj_total_linear, file.path(paths$figures, "jmj_total_cluster_expression_linear.rds"))
ggsave(
  file.path(paths$figures, "jmj_total_cluster_expression_linear.pdf"),
  p_jmj_total_linear,
  width = 8, height = 6
)
ggsave(
  file.path(paths$figures, "jmj_total_cluster_expression_linear.png"),
  p_jmj_total_linear,
  width = 8, height = 6, dpi = 300
)

# Save per-paralog data
write_csv(jmj_expr, file.path(paths$intermediate, "jmj_paralogs_expression_data.csv"))
write_csv(jmj_summary, file.path(paths$intermediate, "jmj_paralogs_expression_summary.csv"))

# Save total cluster data
write_csv(jmj_total, file.path(paths$intermediate, "jmj_total_cluster_expression_data.csv"))
write_csv(jmj_total_summary, file.path(paths$intermediate, "jmj_total_cluster_expression_summary.csv"))

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("\nPer-Paralog Analysis:\n")
  cat("  Figures: jmj_paralogs_expression_boxplot.pdf/png/rds\n")
  cat("  Data: jmj_paralogs_expression_data.csv\n")
  cat("  Data: jmj_paralogs_expression_summary.csv\n")
  cat("\nTotal Cluster Analysis:\n")
  cat("  Figures: jmj_total_cluster_expression.pdf/png/rds (log scale)\n")
  cat("  Figures: jmj_total_cluster_expression_linear.pdf/png/rds (linear scale)\n")
  cat("  Data: jmj_total_cluster_expression_data.csv\n")
  cat("  Data: jmj_total_cluster_expression_summary.csv\n")
}
```

# Session Info

```{r session_info}
sessionInfo()
```
