---
title: "Joint Consensus WGCNA: Genotype-Response and Leaf-Gradient Modules"
author: "F. V. Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = TRUE
)
knitr::opts_knit$set(root.dir = here::here())
```

# Overview

The permutation test notebook builds genotype-response modules (Inv4m DEGs) via
1000 consensus iterations (parameter sweeps + gene subsampling), but the topology
control modules (photosynthesis + senescence genes from Ojeda 2026) are built with
a single `cutreeDynamic` call. This asymmetry makes the LFP comparison unfair and
may cause bimodality in the leaf-gradient kWithin null distributions, because
photosynthesis and senescence genes are anti-correlated along the leaf developmental
gradient and collapse into mixed modules under single-pass clustering.

This notebook fixes the asymmetry by running the same consensus pipeline on the
**combined gene set**: 465 genotype-response genes (Inv4m DEGs) + ~218 leaf-gradient
genes (photosynthesis + senescence), ~683 total. Modules are classified post-hoc
by gene-set composition.

**Gene set terminology:**
- **Genotype-response genes** — differentially expressed between CTRL and Inv4m
  plants; coexpression driven by the Inv4m introgression.
- **Leaf-gradient genes** — photosynthesis and senescence hub genes (Ojeda 2026);
  coexpression driven by the leaf developmental gradient (young base = photosynthetic,
  old tip = senescing), independent of Inv4m genotype.

**Network:** built on CTRL plants only (reference), then tested for connectivity
loss in Inv4m plants.

Predictions:
- Leaf-gradient modules should show **no** systematic kWithin loss between
  CTRL and Inv4m genotypes.
- Genotype-response modules should recapitulate the connectivity loss seen in
  the main permutation test.
- Joint clustering should better separate photosynthesis and senescence genes
  into distinct modules, resolving the bimodality.

**Fixed parameters (justified in permutation test notebook):**
- Soft power = 18 (estimated for both CTRL and Inv4m DEG networks)
- 1K consensus iterations (slow step)
- 10K genotype label permutations

# Setup

```{r load_packages}
library(here)
library(tidyverse)
library(WGCNA)
library(ggridges)
library(furrr)
library(future)

options(stringsAsFactors = FALSE)
disableWGCNAThreads()
set.seed(42)
```

```{r config}
SOFT_POWER            <- 18
N_CONSENSUS           <- 1000
N_BOOT                <- 1000
N_PERM                <- 10000
GENE_SUBSAMPLE_RATE   <- 0.80
MIN_MODULE_SIZE_RANGE <- c(10L, 15L, 20L, 25L, 30L, 35L)
CONSENSUS_MIN_SIZE    <- min(MIN_MODULE_SIZE_RANGE)
DEEP_SPLIT            <- 4L
NETWORK_TYPE          <- "signed"
LOG_FREQUENCY         <- 100L
DEG_FRACTION_THRESH   <- 0.70   # fraction to call a module response- or gradient-dominated
LFP_STABLE            <- 0.70   # LFP >= this: stable
LFP_MODERATE          <- 0.50   # LFP >= this: moderate; < this: unstable

RUN_DIR    <- here("results/inversion_paper/field_perturbation/run_20251231_201332")
OUTPUT_DIR <- file.path(RUN_DIR, "joint_consensus")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

FILTER_DIR    <- file.path(RUN_DIR, "02_gene_filter")
DATA_PREP_DIR <- file.path(RUN_DIR, "01_data_prep")

source(here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")

cat("=== CONFIGURATION ===\n")
cat("Soft power (fixed):", SOFT_POWER, "\n")
cat("Consensus iterations:", N_CONSENSUS, "\n")
cat("Bootstrap iterations:", N_BOOT, "\n")
cat("Permutation iterations:", N_PERM, "\n")
cat("Gene subsample rate:", GENE_SUBSAMPLE_RATE, "\n")
cat("minModuleSize range:", paste(MIN_MODULE_SIZE_RANGE, collapse = ", "), "\n")
cat("DEG fraction threshold:", DEG_FRACTION_THRESH, "\n")
cat("Output directory:", OUTPUT_DIR, "\n")
```

# Load Data

```{r load_data}
# DEG-filtered expression (465 genes x 43 samples)
expr_filtered <- readRDS(file.path(FILTER_DIR, "filtered_expression.rds"))

# Full expression matrix (24k genes x 43 samples, PSU2022)
full_expr_raw <- readRDS(file.path(DATA_PREP_DIR, "combined_expression.rds"))
full_expr_mat <- if (inherits(full_expr_raw, "EList")) full_expr_raw$E else as.matrix(full_expr_raw)

# Sample metadata
sample_meta <- read_csv(file.path(DATA_PREP_DIR, "sample_metadata.csv"),
                         show_col_types = FALSE)

cat("DEG expression:", nrow(expr_filtered), "genes x", ncol(expr_filtered), "samples\n")
cat("Full expression:", nrow(full_expr_mat), "genes x", ncol(full_expr_mat), "samples\n")
```

```{r load_control_genes}
photo_genes <- read.table(
  file.path(paths$data, "photosynthesis_expression_index_genes_ojeda2026.txt"),
  header = FALSE, stringsAsFactors = FALSE
)$V1

sag_genes <- read.table(
  file.path(paths$data, "senesence_hubs_ojeda2026.tab"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)$gene

deg_genes <- rownames(expr_filtered)

photo_control <- setdiff(intersect(photo_genes, rownames(full_expr_mat)), deg_genes)
sag_control   <- setdiff(intersect(sag_genes,  rownames(full_expr_mat)), deg_genes)
control_genes <- unique(c(photo_control, sag_control))

cat("Photosynthesis (non-DEG, in matrix):", length(photo_control), "\n")
cat("Senescence hubs (non-DEG, in matrix):", length(sag_control), "\n")
cat("Combined control genes:", length(control_genes), "\n")
```

# Combine Expression Matrix

```{r combine_expression}
joint_genes <- c(deg_genes, control_genes)
gene_type   <- c(
  setNames(rep("genotype", length(deg_genes)),     deg_genes),
  setNames(rep("leaf",     length(control_genes)), control_genes)
)

# Expression matrix for joint gene set (genes x samples)
ctrl_samples  <- sample_meta |> filter(Genotype == "CTRL")  |> pull(sample_id)
inv4m_samples <- sample_meta |> filter(Genotype == "Inv4m") |> pull(sample_id)
all_samples   <- sample_meta$sample_id

deg_expr <- expr_filtered[deg_genes, all_samples]
ctrl_expr <- full_expr_mat[control_genes, all_samples]

joint_expr_mat <- rbind(deg_expr, ctrl_expr)   # genes x samples

# WGCNA format: samples x genes
datExpr_joint <- t(joint_expr_mat[joint_genes, all_samples])
datExpr_CTRL  <- datExpr_joint[ctrl_samples,  ]
datExpr_Inv4m <- datExpr_joint[inv4m_samples, ]

true_labels <- setNames(sample_meta$Genotype, sample_meta$sample_id)
n_ctrl  <- length(ctrl_samples)
n_inv4m <- length(inv4m_samples)
n_total <- n_ctrl + n_inv4m

cat("Joint expression:", nrow(joint_expr_mat), "genes x", ncol(joint_expr_mat), "samples\n")
cat("  DEG genes:", length(deg_genes), "\n")
cat("  Control genes:", length(control_genes), "\n")
cat("WGCNA matrix: CTRL", n_ctrl, "samples | Inv4m", n_inv4m, "samples\n")
```

# Joint Consensus Network

Build the consensus network on **CTRL samples only** (n = `r n_ctrl`), mirroring
the approach in step 04 of the field perturbation pipeline.

## Consensus Functions

```{r define_consensus_functions}
#' Run single WGCNA iteration on gene subsample
run_subsample_wgcna <- function(datExpr, gene_indices, power,
                                 min_module_size, deep_split = 4L,
                                 network_type = "signed") {
  expr_subset <- datExpr[, gene_indices]
  net <- blockwiseModules(
    expr_subset,
    power          = power,
    networkType    = network_type,
    TOMType        = network_type,
    minModuleSize  = min_module_size,
    mergeCutHeight = 0,
    reassignThreshold = 0,
    maxBlockSize   = ncol(expr_subset),
    numericLabels  = TRUE,
    deepSplit      = deep_split,
    verbose        = 0
  )
  data.frame(gene = colnames(expr_subset), module = net$colors,
             stringsAsFactors = FALSE)
}

build_indicator_matrix <- function(subsample_configs, gene_names) {
  n <- length(gene_names)
  I <- matrix(0L, n, n, dimnames = list(gene_names, gene_names))
  for (cfg in subsample_configs) {
    g <- gene_names[cfg$gene_indices]
    I[g, g] <- I[g, g] + 1L
  }
  I
}

build_coclustering_matrix <- function(module_results, gene_names) {
  n <- length(gene_names)
  C <- matrix(0L, n, n, dimnames = list(gene_names, gene_names))
  for (res in module_results) {
    assigned <- res[res$module != 0, ]
    for (mod in unique(assigned$module)) {
      g <- assigned$gene[assigned$module == mod]
      if (length(g) > 1) C[g, g] <- C[g, g] + 1L
    }
  }
  C
}

compute_consensus_adjacency <- function(C, I) {
  A <- C / I
  A[is.nan(A) | is.infinite(A)] <- 0
  A
}

cut_consensus_tree <- function(A, min_module_size = 10L, deep_split = 4L) {
  D    <- 1 - A
  tree <- hclust(as.dist(D), method = "average")
  invisible(capture.output(
    mods <- cutreeDynamic(dendro = tree, distM = D,
                           minClusterSize = min_module_size,
                           deepSplit = deep_split, verbose = 0)
  ))
  names(mods) <- rownames(A)
  list(tree = tree, modules = mods)
}

# File-based atomic counter for furrr workers
increment_and_get_count <- function(counter_file, lock_file) {
  tryCatch({
    for (attempt in seq_len(100)) {
      if (!file.exists(lock_file)) {
        tryCatch(writeLines(as.character(Sys.getpid()), lock_file),
                 error = function(e) NULL)
        break
      }
      Sys.sleep(0.01 * sample(1:10, 1))
    }
    lines <- readLines(counter_file, warn = FALSE)
    cnt   <- if (length(lines) > 0 && !is.na(suppressWarnings(as.integer(lines[1])))) {
      as.integer(lines[1])
    } else 0L
    writeLines(as.character(cnt + 1L), counter_file)
    tryCatch(unlink(lock_file), error = function(e) NULL)
    cnt + 1L
  }, error = function(e) NA_integer_)
}

cat("Consensus functions defined\n")
```

## Run 1K Consensus Iterations (CTRL samples)

```{r consensus_iterations, cache=TRUE}
n_joint <- ncol(datExpr_CTRL)
subsample_size <- floor(GENE_SUBSAMPLE_RATE * n_joint)

joint_configs <- lapply(seq_len(N_CONSENSUS), function(i) {
  list(
    iteration    = i,
    gene_indices = sample(n_joint, subsample_size, replace = FALSE),
    min_module_size = sample(MIN_MODULE_SIZE_RANGE, 1L)
  )
})

n_cores <- max(1L, parallel::detectCores() - 1L)
plan(multisession, workers = n_cores)

counter_file <- file.path(OUTPUT_DIR, ".progress_counter")
lock_file    <- file.path(OUTPUT_DIR, ".progress_lock")
writeLines("0", counter_file)

start_time <- Sys.time()

joint_iter_results <- future_map(joint_configs, function(cfg) {
  res <- run_subsample_wgcna(
    datExpr_CTRL, cfg$gene_indices, SOFT_POWER,
    cfg$min_module_size, DEEP_SPLIT, NETWORK_TYPE
  )
  cnt <- increment_and_get_count(counter_file, lock_file)
  if (!is.na(cnt) && (cnt %% LOG_FREQUENCY == 0 || cnt == N_CONSENSUS)) {
    elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
    eta     <- (N_CONSENSUS - cnt) * (elapsed / cnt)
    cat(sprintf("[%s] Joint consensus: %d/%d (%.0f%%) | ETA: %.1f min\n",
                format(Sys.time(), "%H:%M:%S"), cnt, N_CONSENSUS,
                100 * cnt / N_CONSENSUS, eta / 60))
  }
  res
}, .options = furrr_options(seed = TRUE), .progress = FALSE)

plan(sequential)
unlink(counter_file); unlink(lock_file)

elapsed_min <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
cat(sprintf("Consensus complete: %.1f minutes\n", elapsed_min))
saveRDS(joint_iter_results, file.path(OUTPUT_DIR, "consensus_iter_results.rds"))
```

## Build Consensus Adjacency

```{r consensus_adjacency, cache=TRUE}
cat("Building indicator matrix...\n")
I_joint <- build_indicator_matrix(joint_configs, colnames(datExpr_CTRL))
saveRDS(I_joint, file.path(OUTPUT_DIR, "indicator_matrix.rds"))

cat("Building co-clustering matrix...\n")
C_joint <- build_coclustering_matrix(joint_iter_results, colnames(datExpr_CTRL))
saveRDS(C_joint, file.path(OUTPUT_DIR, "coclustering_matrix.rds"))

cat("Computing consensus adjacency...\n")
A_joint <- compute_consensus_adjacency(C_joint, I_joint)
saveRDS(A_joint, file.path(OUTPUT_DIR, "consensus_adjacency.rds"))

cat("Indicator range:    [", min(I_joint), ",", max(I_joint), "]\n")
cat("Co-clustering range:[", min(C_joint), ",", max(C_joint), "]\n")
cat("Adjacency range:    [", round(min(A_joint), 3), ",", round(max(A_joint), 3), "]\n")
```

## Cut Consensus Tree

```{r consensus_cut, cache=TRUE}
cat("Cutting joint consensus tree...\n")
joint_tree <- cut_consensus_tree(A_joint, CONSENSUS_MIN_SIZE, DEEP_SPLIT)

raw_modules <- joint_tree$modules

# Assign size-sorted consensus labels: consensus01 (largest) ... consensusN
# This replaces stochastic WGCNA color names with stable index labels.
mod_sizes <- sort(
  table(raw_modules[raw_modules != 0]),
  decreasing = TRUE
)
consensus_labels <- setNames(
  sprintf("consensus%02d", seq_along(mod_sizes)),
  names(mod_sizes)   # keys are character representations of integer module IDs
)

# Named vector: gene -> "consensus01" ... "consensusN" | "unassigned"
joint_colors <- setNames(
  ifelse(raw_modules == 0, "unassigned",
         consensus_labels[as.character(raw_modules)]),
  names(raw_modules)
)

joint_module_df <- data.frame(
  gene      = names(raw_modules),
  wgcna_int = raw_modules,         # original integer (for reference)
  color     = joint_colors,        # "consensus01", ..., "unassigned"
  stringsAsFactors = FALSE
)

saveRDS(joint_tree,      file.path(OUTPUT_DIR, "consensus_tree.rds"))
saveRDS(consensus_labels, file.path(OUTPUT_DIR, "consensus_labels.rds"))
write_csv(joint_module_df, file.path(OUTPUT_DIR, "joint_consensus_modules_raw.csv"))

n_mods   <- length(unique(raw_modules[raw_modules != 0]))
n_grey   <- sum(raw_modules == 0)
n_assign <- sum(raw_modules != 0)

cat("Modules found:", n_mods, "\n")
cat("Assigned genes:", n_assign, "| Unassigned:", n_grey, "\n\n")
print(table(joint_module_df$color))
```

# Module Classification

Classify each module by the fraction of DEG vs control genes it contains.
Threshold = `r DEG_FRACTION_THRESH`.

```{r module_classification}
joint_module_df <- joint_module_df |>
  mutate(gene_type = gene_type[gene])

module_composition <- joint_module_df |>
  filter(color != "unassigned") |>
  group_by(color) |>
  summarise(
    n_genes   = n(),
    n_genotype = sum(gene_type == "genotype"),
    n_leaf     = sum(gene_type == "leaf"),
    frac_DEG   = n_genotype / n_genes,
    .groups    = "drop"
  ) |>
  mutate(
    module_type = case_when(
      frac_DEG >= DEG_FRACTION_THRESH        ~ "genotype-dominated",
      (1 - frac_DEG) >= DEG_FRACTION_THRESH  ~ "leaf-dominated",
      TRUE                                    ~ "mixed"
    )
  ) |>
  arrange(desc(frac_DEG))

cat("=== MODULE COMPOSITION ===\n")
print(module_composition)

cat("\nModule counts by type:\n")
print(table(module_composition$module_type))

# Attach module_type to full gene-level table
joint_module_df <- joint_module_df |>
  left_join(module_composition |> select(color, module_type, frac_DEG),
            by = "color")   # module_type: "genotype-dominated" | "leaf-dominated" | "mixed"

write_csv(joint_module_df,
          file.path(OUTPUT_DIR, "joint_consensus_modules.csv"))
```

```{r composition_barplot, fig.width=9, fig.height=4}
module_composition |>
  pivot_longer(c(n_genotype, n_leaf), names_to = "gene_set", values_to = "count") |>
  mutate(
    gene_set = recode(gene_set, n_genotype = "Genotype response", n_leaf = "Leaf gradient"),
    color_ordered = fct_reorder(color, frac_DEG, .desc = TRUE)
  ) |>
  ggplot(aes(x = color_ordered, y = count, fill = gene_set)) +
  geom_col(position = "stack") +
  geom_text(
    data = module_composition,
    aes(x = fct_reorder(color, frac_DEG, .desc = TRUE),
        y = n_genes + 2, label = module_type, fill = NULL),
    size = 3, angle = 45, hjust = 0
  ) +
  scale_fill_manual(values = c("Genotype response" = "#3B528B", "Leaf gradient" = "#21908C")) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Module", y = "Gene count", fill = NULL,
       title = "Joint CTRL reference network: module composition")
```

# Palette Reassignment

Map the temporary `consensus01`...`consensusN` index labels to the color palette
used in the main analysis, so that downstream plots (bootstrap LFP, ridgeline
distributions) share colors with Figure 6 and the permutation test notebook.

- **Genotype-dominated modules** → matched to existing DEG consensus module colors
  (`04_consensus_networks/ctrl/consensus_modules.csv`) by maximum Jaccard gene-set
  overlap. These are the standard WGCNA color names (turquoise, blue, brown, ...).
- **Leaf-dominated modules** → assigned the fixed leaf gradient palette used in
  `permutation_test_delta_kWithin.Rmd` (`ctrl_color_remap`: sky, peach, jade, ...).
  Chosen to not clash with WGCNA standard colors.
- **Mixed modules** → grey70.

After this chunk, `joint_colors` and `joint_module_df$color` use the reassigned
palette for all downstream analysis.

```{r palette_reassignment}
# Existing DEG module assignments from main consensus pipeline
CONSENSUS_DIR <- file.path(RUN_DIR, "04_consensus_networks")
deg_ref_modules <- read_csv(
  file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)

# Leaf gradient palette (matches ctrl_color_remap in permutation_test notebook)
LEAF_COLOR_HEX <- c(
  sky      = "#5DA5DA",
  peach    = "#FAA43A",
  jade     = "#60BD68",
  coral    = "#F15854",
  lavender = "#B276B2",
  umber    = "#8D4B08",
  mustard  = "#DECF3F",
  rose     = "#F17CB0"
)

jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

# --- Genotype-dominated: match to DEG palette by Jaccard overlap ---
geno_mods     <- module_composition |>
  filter(module_type == "genotype-dominated") |> pull(color)
deg_mod_colors <- unique(deg_ref_modules$color[deg_ref_modules$color != "grey"])

geno_color_map <- setNames(
  map_chr(geno_mods, function(cons_label) {
    genes_joint <- names(joint_colors[joint_colors == cons_label])
    j_scores <- sapply(deg_mod_colors, function(deg_col) {
      jaccard(genes_joint, deg_ref_modules$gene[deg_ref_modules$color == deg_col])
    })
    best <- names(which.max(j_scores))
    if (max(j_scores) < 0.05)
      warning(cons_label, ": low best Jaccard = ", round(max(j_scores), 3))
    best
  }),
  geno_mods
)

cat("=== GENOTYPE MODULE PALETTE ===\n")
for (cm in geno_mods) {
  g_j <- names(joint_colors[joint_colors == cm])
  g_d <- deg_ref_modules$gene[deg_ref_modules$color == geno_color_map[cm]]
  cat(sprintf("  %-14s -> %-12s  Jaccard=%.3f  n=%d\n",
              cm, geno_color_map[cm], jaccard(g_j, g_d), length(g_j)))
}

# --- Leaf-dominated: assign fixed palette by descending module size ---
leaf_mods <- module_composition |>
  filter(module_type == "leaf-dominated") |>
  arrange(desc(n_genes)) |> pull(color)
if (length(leaf_mods) > length(LEAF_COLOR_HEX))
  warning("More leaf modules (", length(leaf_mods),
          ") than palette entries (", length(LEAF_COLOR_HEX), ")")
leaf_color_map <- setNames(
  names(LEAF_COLOR_HEX)[seq_along(leaf_mods)],
  leaf_mods
)

cat("\n=== LEAF GRADIENT MODULE PALETTE ===\n")
for (cm in leaf_mods)
  cat(sprintf("  %-14s -> %-12s  n=%d\n",
              cm, leaf_color_map[cm], sum(joint_colors == cm)))

# --- Mixed: grey ---
mixed_mods <- module_composition |>
  filter(module_type == "mixed") |> pull(color)
mixed_color_map <- setNames(rep("grey70", length(mixed_mods)), mixed_mods)

# --- Build full palette map and hex lookup ---
palette_map <- c(geno_color_map, leaf_color_map, mixed_color_map,
                 unassigned = "unassigned")

# Hex lookup for scale_fill_manual / scale_color_manual in all downstream plots.
# Standard R color names (WGCNA DEG palette) are resolved via col2rgb.
deg_hex <- setNames(
  apply(col2rgb(geno_color_map), 2,
        function(x) rgb(x[1], x[2], x[3], maxColorValue = 255)),
  geno_color_map
)
color_hex <- c(deg_hex, LEAF_COLOR_HEX, grey70 = "#B3B3B3",
               unassigned = "#FFFFFF")
color_hex <- color_hex[!duplicated(names(color_hex))]

cat("\n=== FULL PALETTE MAP ===\n")
print(palette_map)

# --- Update joint_colors and joint_module_df ---
joint_colors_idx <- joint_colors   # keep consensusXX for reference
joint_colors     <- unname(palette_map[joint_colors_idx])
names(joint_colors) <- names(joint_colors_idx)

joint_module_df <- joint_module_df |>
  mutate(
    consensus_id = color,           # original index label
    color        = palette_map[color]
  )
module_composition <- module_composition |>
  mutate(
    consensus_id = color,
    color        = palette_map[color]
  )

cat("\nPalette reassignment complete.",
    "joint_colors and joint_module_df updated.\n")
```

# Bootstrap LFP

Quantify module stability by resampling CTRL samples with replacement 1000 times,
rebuilding the full adjacency + hierarchical clustering + `cutreeDynamic`, and
computing the Largest Fragment Proportion (LFP) per module.

```{r bootstrap_lfp, cache=TRUE}
# Reference module assignments (numeric, non-grey)
ref_modules <- setNames(joint_tree$modules, names(joint_tree$modules))

boot_results_lfp <- vector("list", N_BOOT)

for (b in seq_len(N_BOOT)) {
  n_samp     <- nrow(datExpr_CTRL)
  boot_idx   <- sample(n_samp, n_samp, replace = TRUE)
  expr_boot  <- datExpr_CTRL[boot_idx, ]

  adj_boot <- adjacency(expr_boot, power = SOFT_POWER, type = NETWORK_TYPE)
  tree_boot <- hclust(as.dist(1 - adj_boot), method = "average")
  invisible(capture.output(
    mods_boot <- cutreeDynamic(
      dendro        = tree_boot,
      distM         = 1 - adj_boot,
      deepSplit     = DEEP_SPLIT,
      minClusterSize = CONSENSUS_MIN_SIZE
    )
  ))

  boot_results_lfp[[b]] <- data.frame(
    gene   = colnames(datExpr_CTRL),
    module = mods_boot,
    stringsAsFactors = FALSE
  )

  if (b %% 100 == 0) cat("Bootstrap:", b, "/", N_BOOT, "\n")
}

saveRDS(boot_results_lfp, file.path(OUTPUT_DIR, "bootstrap_results.rds"))
cat("Bootstrap complete\n")
```

```{r compute_lfp, cache=TRUE}
cons_ids <- sort(unique(joint_colors[joint_colors != "unassigned"]))

lfp_results <- map_dfr(cons_ids, function(cons_label) {
  mod_genes <- names(joint_colors[joint_colors == cons_label])

  lfp_vals <- sapply(boot_results_lfp, function(boot) {
    boot_sub <- boot[boot$gene %in% mod_genes, ]
    counts   <- table(boot_sub$module)
    counts_nz <- counts[names(counts) != "0"]
    if (length(counts_nz) == 0) return(0)
    max(counts_nz) / length(mod_genes)
  })

  tibble(
    color   = cons_label,
    n_genes = length(mod_genes),
    mean_lfp   = mean(lfp_vals),
    sd_lfp     = sd(lfp_vals),
    se_lfp     = sd(lfp_vals) / sqrt(N_BOOT),
    stability  = case_when(
      mean(lfp_vals) >= LFP_STABLE   ~ "stable",
      mean(lfp_vals) >= LFP_MODERATE ~ "moderate",
      TRUE                            ~ "unstable"
    )
  )
}) |>
  left_join(module_composition |> select(color, module_type, frac_DEG), by = "color") |>
  arrange(desc(mean_lfp))

cat("=== BOOTSTRAP LFP ===\n")
print(lfp_results)

write_csv(lfp_results, file.path(OUTPUT_DIR, "bootstrap_lfp.csv"))
```

```{r lfp_plot, fig.width=7, fig.height=4}
lfp_results |>
  filter(!is.na(module_type)) |>
  mutate(color_ordered = fct_reorder(color, mean_lfp)) |>
  ggplot(aes(x = color_ordered, y = mean_lfp,
             ymin = mean_lfp - se_lfp, ymax = mean_lfp + se_lfp,
             color = module_type)) +
  geom_pointrange() +
  geom_hline(yintercept = c(LFP_MODERATE, LFP_STABLE), linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(
    "genotype-dominated" = "#3B528B",   # viridis D mid (temporary; no single color for a type)
    "leaf-dominated"     = "#21908C",
    "mixed"              = "grey70"
  )) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(x = NULL, y = "Mean LFP ± SE", color = "Module type",
       title = "Bootstrap module stability (LFP) by module type")
```

# Permutation Test: Genotype Label Shuffling

Shuffle genotype labels 10K times (keeping group sizes fixed: `r n_ctrl` CTRL,
`r n_inv4m` Inv4m). Recompute mean $\Delta k_{Within}$ and $\Delta k_{Out}$
per joint consensus module. Module definitions remain fixed throughout.

The prediction: DEG-dominated modules show kWithin loss; control-dominated
modules do not.

## Helper Function

```{r delta_k_helper}
compute_module_delta_k <- function(datExpr_a, datExpr_b, power, mod_colors) {
  adj_a <- adjacency(datExpr_a, power = power, type = "signed")
  adj_b <- adjacency(datExpr_b, power = power, type = "signed")

  k_a <- intramodularConnectivity(adj_a, mod_colors)
  k_b <- intramodularConnectivity(adj_b, mod_colors)

  delta_kWithin <- k_b$kWithin - k_a$kWithin
  delta_kOut    <- k_b$kOut    - k_a$kOut

  module_ids <- unique(mod_colors[mod_colors != "unassigned" & !is.na(mod_colors)])
  list(
    kWithin = sapply(module_ids, function(m) {
      mean(delta_kWithin[mod_colors == m], na.rm = TRUE)
    }),
    kOut = sapply(module_ids, function(m) {
      mean(delta_kOut[mod_colors == m], na.rm = TRUE)
    })
  )
}
```

## Observed Delta kWithin

```{r observed_delta}
obs <- compute_module_delta_k(
  datExpr_joint[ctrl_samples,  ],
  datExpr_joint[inv4m_samples, ],
  SOFT_POWER, joint_colors
)

cat("Observed mean delta kWithin per module (sorted):\n")
print(round(sort(obs$kWithin), 4))
cat("\nObserved mean delta kOut per module (sorted):\n")
print(round(sort(obs$kOut), 4))
```

## Timing Probe

Run 20 iterations first to measure actual per-iteration cost on this machine,
then project the full run time before committing to 10K.

```{r permutation_timing_probe}
N_PROBE <- 20L

probe_start <- proc.time()
for (i in seq_len(N_PROBE)) {
  perm_idx     <- sample(n_total)
  pseudo_ctrl  <- perm_idx[seq_len(n_ctrl)]
  pseudo_inv4m <- perm_idx[(n_ctrl + 1):n_total]
  res <- compute_module_delta_k(
    datExpr_joint[pseudo_ctrl,  ],
    datExpr_joint[pseudo_inv4m, ],
    SOFT_POWER, joint_colors
  )
}
probe_elapsed <- (proc.time() - probe_start)[["elapsed"]]

sec_per_iter <- probe_elapsed / N_PROBE
projected_min <- sec_per_iter * N_PERM / 60

cat(sprintf(
  "Timing probe: %d iterations in %.1f s\n  -> %.3f s / iteration\n  -> %.1f min projected for N_PERM = %d\n",
  N_PROBE, probe_elapsed, sec_per_iter, projected_min, N_PERM
))
```

## Permutation Loop

```{r permutation_loop, cache=TRUE}
null_kWithin <- matrix(NA_real_, nrow = N_PERM, ncol = length(obs$kWithin),
                        dimnames = list(NULL, names(obs$kWithin)))
null_kOut    <- matrix(NA_real_, nrow = N_PERM, ncol = length(obs$kOut),
                        dimnames = list(NULL, names(obs$kOut)))

perm_start <- Sys.time()

for (i in seq_len(N_PERM)) {
  perm_idx      <- sample(n_total)
  pseudo_ctrl   <- perm_idx[seq_len(n_ctrl)]
  pseudo_inv4m  <- perm_idx[(n_ctrl + 1):n_total]

  res <- compute_module_delta_k(
    datExpr_joint[pseudo_ctrl,  ],
    datExpr_joint[pseudo_inv4m, ],
    SOFT_POWER, joint_colors
  )
  null_kWithin[i, ] <- res$kWithin
  null_kOut[i, ]    <- res$kOut

  if (i %% 500 == 0) {
    elapsed <- as.numeric(difftime(Sys.time(), perm_start, units = "mins"))
    eta     <- (N_PERM - i) * (elapsed / i)
    cat(sprintf("[%s] %d/%d | %.1f min elapsed | ETA %.1f min\n",
                format(Sys.time(), "%H:%M:%S"), i, N_PERM, elapsed, eta))
  }
}

elapsed_total <- as.numeric(difftime(Sys.time(), perm_start, units = "mins"))
cat(sprintf("Permutation test complete: %.1f min total\n", elapsed_total))
```

## ECDF P-values

```{r ecdf_pvalues}
# kWithin: one-tailed, test for loss (observed <= null)
ecdf_pvals_kWithin <- sapply(names(obs$kWithin), function(m) {
  mean(null_kWithin[, m] <= obs$kWithin[m])
})

# kOut: one-tailed, test for gain (observed >= null)
ecdf_pvals_kOut <- sapply(names(obs$kOut), function(m) {
  mean(null_kOut[, m] >= obs$kOut[m])
})
```

## FDR Correction

BH correction applied jointly across all module tests.

```{r fdr_correction}
n_genes_per_mod <- sapply(names(obs$kWithin), function(m) {
  sum(joint_colors == m, na.rm = TRUE)
})

results_joint <- tibble(
  color                  = names(obs$kWithin),
  n_genes                = n_genes_per_mod,
  observed_delta_kWithin = round(obs$kWithin, 4),
  null_mean_kWithin      = round(colMeans(null_kWithin), 4),
  null_sd_kWithin        = round(apply(null_kWithin, 2, sd), 4),
  ecdf_p_kWithin         = round(ecdf_pvals_kWithin, 4),
  observed_delta_kOut    = round(obs$kOut, 4),
  null_mean_kOut         = round(colMeans(null_kOut), 4),
  null_sd_kOut           = round(apply(null_kOut, 2, sd), 4),
  ecdf_p_kOut            = round(ecdf_pvals_kOut, 4)
) |>
  left_join(module_composition |> select(color, module_type, n_genotype, n_leaf, frac_DEG),
            by = "color") |>
  mutate(
    fdr_kWithin = round(p.adjust(ecdf_p_kWithin, method = "BH"), 4),
    fdr_kOut    = round(p.adjust(ecdf_p_kOut,    method = "BH"), 4),
    sig_kWithin = ifelse(fdr_kWithin < 0.05, "*", "")
  ) |>
  arrange(fdr_kWithin)

cat("=== PERMUTATION TEST RESULTS ===\n\n")

cat("Genotype-dominated modules:\n")
results_joint |>
  filter(module_type == "genotype-dominated") |>
  select(color, n_genes, n_genotype, observed_delta_kWithin, ecdf_p_kWithin, fdr_kWithin, sig_kWithin) |>
  print()

cat("\nLeaf-dominated modules:\n")
results_joint |>
  filter(module_type == "leaf-dominated") |>
  select(color, n_genes, n_leaf, observed_delta_kWithin, ecdf_p_kWithin, fdr_kWithin, sig_kWithin) |>
  print()

cat("\nMixed modules:\n")
results_joint |>
  filter(module_type == "mixed") |>
  select(color, n_genes, n_genotype, n_leaf, observed_delta_kWithin, ecdf_p_kWithin, fdr_kWithin, sig_kWithin) |>
  print()

cat("\nSignificant at FDR < 0.05:", sum(results_joint$fdr_kWithin < 0.05, na.rm = TRUE),
    "of", nrow(results_joint), "\n")
```

## Save Results

```{r save_results}
write_csv(results_joint, file.path(OUTPUT_DIR, "joint_permutation_results.csv"))

saveRDS(
  list(
    observed_kWithin = obs$kWithin,
    observed_kOut    = obs$kOut,
    null_kWithin     = null_kWithin,
    null_kOut        = null_kOut,
    pvals_kWithin    = ecdf_pvals_kWithin,
    pvals_kOut       = ecdf_pvals_kOut,
    joint_colors     = joint_colors,
    module_composition = module_composition,
    control_genes    = control_genes,
    deg_genes        = deg_genes
  ),
  file.path(OUTPUT_DIR, "joint_permutation_results.rds")
)

cat("Results saved to:", OUTPUT_DIR, "\n")
```

# Visualization

## Ridgeline Density Plots

Null distributions of $\Delta k_{Within}$ from 10K genotype label permutations.
Red line = observed value. Modules ordered by observed delta within each type.

```{r ridgeline_deg, fig.width=10, fig.height=7}
deg_mods <- results_joint |>
  filter(module_type == "genotype-dominated") |>
  arrange(observed_delta_kWithin) |>
  pull(color)

if (length(deg_mods) > 0) {
  deg_color_hex <- setNames(deg_mods, deg_mods)

  null_deg_df <- map_dfr(deg_mods, function(m) {
    tibble(module = m, z_value = null_kWithin[, m])
  }) |>
    mutate(module = factor(module, levels = deg_mods))

  pts_deg <- results_joint |>
    filter(color %in% deg_mods) |>
    transmute(
      module  = factor(color, levels = deg_mods),
      z_observed = observed_delta_kWithin,
      fdr     = fdr_kWithin
    )

  p_deg <- ggplot(null_deg_df, aes(x = z_value, y = module, fill = module)) +
    geom_density_ridges(color = "grey40", alpha = 0.7, scale = 0.9,
                         rel_min_height = 0.01) +
    scale_fill_manual(values = color_hex) +
    guides(fill = "none") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_segment(
      data = pts_deg,
      aes(x = z_observed, xend = z_observed,
          y = as.numeric(module), yend = as.numeric(module) + 0.8),
      color = "red", linewidth = 0.8, inherit.aes = FALSE
    ) +
    geom_text(
      data = pts_deg |> filter(fdr < 0.05),
      aes(x = z_observed - 0.6, y = as.numeric(module) + 0.35,
          label = sprintf("FDR=%.3f", fdr)),
      color = "red", size = 3, hjust = 0.5, inherit.aes = FALSE
    ) +
    theme_minimal(base_size = 14) +
    labs(y = NULL, x = expression(Delta * k[Within]),
         title = "Genotype-response modules: genotype label permutation null")
  print(p_deg)
} else {
  cat("No genotype-dominated modules to plot\n")
}
```

```{r ridgeline_control, fig.width=10, fig.height=5}
ctrl_mods <- results_joint |>
  filter(module_type == "leaf-dominated") |>
  arrange(observed_delta_kWithin) |>
  pull(color)

if (length(ctrl_mods) > 0) {
  null_ctrl_df <- map_dfr(ctrl_mods, function(m) {
    tibble(module = m, z_value = null_kWithin[, m])
  }) |>
    mutate(module = factor(module, levels = ctrl_mods))

  pts_ctrl <- results_joint |>
    filter(color %in% ctrl_mods) |>
    transmute(
      module  = factor(color, levels = ctrl_mods),
      z_observed = observed_delta_kWithin,
      fdr     = fdr_kWithin
    )

  p_ctrl <- ggplot(null_ctrl_df, aes(x = z_value, y = module, fill = module)) +
    geom_density_ridges(color = "grey40", alpha = 0.7,
                         scale = 0.9, rel_min_height = 0.01) +
    scale_fill_manual(values = color_hex) +
    guides(fill = "none") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_segment(
      data = pts_ctrl,
      aes(x = z_observed, xend = z_observed,
          y = as.numeric(module), yend = as.numeric(module) + 0.8),
      color = "red", linewidth = 0.8, inherit.aes = FALSE
    ) +
    geom_text(
      data = pts_ctrl |> filter(fdr < 0.05),
      aes(x = z_observed - 0.6, y = as.numeric(module) + 0.35,
          label = sprintf("FDR=%.3f", fdr)),
      color = "red", size = 3, hjust = 0.5, inherit.aes = FALSE
    ) +
    theme_minimal(base_size = 14) +
    labs(y = NULL, x = expression(Delta * k[Within]),
         title = "Leaf gradient modules: genotype label permutation null")
  print(p_ctrl)
} else {
  cat("No leaf-dominated modules to plot\n")
}
```

## Bimodality Check: Control Gene kWithin Distributions

Compare the kWithin distributions of photosynthesis vs senescence genes
within each control-dominated module. This checks whether the joint
clustering resolved the bimodality seen in single-pass clustering.

```{r bimodality_check, fig.width=9, fig.height=4}
ctrl_gene_info <- joint_module_df |>
  filter(gene_type == "leaf", color != "unassigned") |>
  mutate(
    leaf_program = case_when(
      gene %in% photo_control ~ "photosynthesis",
      gene %in% sag_control   ~ "senescence",
      TRUE                     ~ "other"
    )
  )

adj_all_ctrl <- adjacency(datExpr_CTRL, power = SOFT_POWER, type = NETWORK_TYPE)
k_ctrl_full <- intramodularConnectivity(
  adj_all_ctrl,
  joint_colors[rownames(adj_all_ctrl)]
)

ctrl_kw_df <- tibble(
  gene        = names(joint_colors),
  color       = joint_colors,
  kWithin     = k_ctrl_full$kWithin
) |>
  filter(gene %in% control_genes, color != "unassigned") |>  # control_genes = leaf-gradient gene set
  mutate(
    leaf_program = case_when(
      gene %in% photo_control ~ "photosynthesis",
      gene %in% sag_control   ~ "senescence",
      TRUE                     ~ "other"
    )
  )

ggplot(ctrl_kw_df, aes(x = kWithin, fill = leaf_program)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  facet_wrap(~color, scales = "free") +
  scale_fill_manual(values = c(photosynthesis = "forestgreen",
                                senescence    = "goldenrod3",
                                other         = "grey60")) +
  theme_minimal(base_size = 12) +
  labs(x = expression(k[Within]), y = "Count", fill = "Leaf program",
       title = "kWithin by leaf program within leaf gradient modules",
       subtitle = "Bimodality = photosynthesis/senescence separation unresolved")
```

# Summary

```{r summary}
cat("=== JOINT CONSENSUS ANALYSIS SUMMARY ===\n\n")

cat("Gene set sizes:\n")
cat("  Genotype response genes:", length(deg_genes), "\n")
cat("  Leaf gradient genes:    ", length(control_genes), "  (",
    length(photo_control), "photosynthesis +",
    length(sag_control), "senescence)\n")
cat("  Joint total:            ", length(joint_genes), "\n\n")

cat("Consensus network (CTRL samples, power =", SOFT_POWER, "):\n")
cat("  Modules:", n_mods, " | Assigned:", n_assign, " | Grey:", n_grey, "\n\n")

cat("Module classification (threshold =", DEG_FRACTION_THRESH, "):\n")
print(module_composition |> select(color, n_genes, n_genotype, n_leaf, frac_DEG, module_type))

cat("\nBootstrap LFP (stability):\n")
print(lfp_results |> select(color, n_genes, mean_lfp, stability, module_type))

cat("\nPermutation test (10K iterations):\n")
cat("  Significant genotype-dominated (FDR < 0.05):",
    results_joint |> filter(module_type == "genotype-dominated", fdr_kWithin < 0.05) |> nrow(), "\n")
cat("  Significant leaf-dominated (FDR < 0.05):",
    results_joint |> filter(module_type == "leaf-dominated", fdr_kWithin < 0.05) |> nrow(), "\n")
cat("  Significant mixed (FDR < 0.05):",
    results_joint |> filter(module_type == "mixed", fdr_kWithin < 0.05) |> nrow(), "\n")

cat("\nConclusion: ",
    if (results_joint |> filter(module_type == "leaf-dominated", fdr_kWithin < 0.05) |> nrow() == 0)
      "Leaf gradient modules show no significant kWithin loss. Connectivity loss is specific to genotype response modules."
    else
      "Some leaf gradient modules show significant kWithin loss. Investigate further.",
    "\n")
```