---
title: "JMJ Cluster Expression Boxplot"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.width = 10,
 fig.height = 6
)

# Load project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This script creates an expression boxplot for the JMJ cluster genes (jmj2, jmj4, jmj21)
comparing Inv4m vs CTRL across all tissues in:

1. **PSU2022**: Field experiment with 4 leaf positions
2. **Crow2020**: Growth chamber experiment with 5 tissues (V1_Leaf, V1_Root, V3_Leaf, V3_Root, V3_SAM)

The output is saved as an RDS file for composition with the JMJ microsynteny figure.

## Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggfx)
library(ggbeeswarm)
library(here)
```

## Define JMJ Genes

```{r jmj_genes}
jmj_genes <- tibble(
 gene = c("Zm00001eb191790", "Zm00001eb191820", "Zm00001eb190750"),
 symbol = c("jmj2", "jmj4", "jmj21"),
 description = c(
   "JUMONJI-transcription factor 2",
   "JUMONJI-transcription factor 4",
   "JUMONJI-transcription factor 21"
 )
)

cat("JMJ cluster genes:\n")
print(jmj_genes)
```



# Load PSU2022 Data

```{r load_psu2022}
# Load expression matrix (logCPM)
expr_psu <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_leaf_trt.rds"))

# Load metadata
meta_psu <- read.csv(file.path(paths$data, "PSU2022_metadata.csv"))

{
 cat("=== PSU2022 DATA ===\n")
 cat("Expression matrix:", nrow(expr_psu), "genes x", ncol(expr_psu), "samples\n")
 cat("Metadata samples:", nrow(meta_psu), "\n")
 cat("\nGenotypes:\n")
 print(table(meta_psu$Genotype))
 cat("\nLeaf positions:\n")
 print(table(meta_psu$leaf_tissue))
}
```

```{r extract_psu2022}
# Extract JMJ expression
jmj_in_psu <- jmj_genes$gene[jmj_genes$gene %in% rownames(expr_psu)]
cat("JMJ genes found in PSU2022:", length(jmj_in_psu), "\n")

if (length(jmj_in_psu) > 0) {
 # Extract expression for JMJ genes
 jmj_expr_psu <- expr_psu[jmj_in_psu, , drop = FALSE] %>%
   as.data.frame() %>%
   rownames_to_column("gene") %>%
   pivot_longer(
     cols = -gene,
     names_to = "sample",
     values_to = "logCPM"
   )

 # Match samples to metadata
 # Sample IDs in expression matrix should match library column in metadata
 jmj_expr_psu <- jmj_expr_psu %>%
   left_join(
     meta_psu %>%
       dplyr::select(library, Genotype, leaf_tissue, Treatment) %>%
       mutate(sample = library),
     by = "sample"
   ) %>%
   filter(!is.na(Genotype)) %>%
   left_join(jmj_genes, by = "gene") %>%
   mutate(
     experiment = "PSU2022 (Field)",
     tissue = paste0("Leaf ", leaf_tissue),
     genotype = ifelse(Genotype == "CTRL", "B73", "Inv4m")
   )

 cat("PSU2022 expression records:", nrow(jmj_expr_psu), "\n")
}
```



# Load Crow2020 Data

```{r load_crow2020}
# Load expression matrix (logCPM)
expr_crow <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_crow2020.rds"))

# Load metadata
meta_crow <- read.csv(file.path(paths$data, "SraRunTable.csv"), check.names = FALSE)

# Standardize column names - use Run (SRR ID) as sample ID
meta_crow <- meta_crow %>%
 dplyr::rename(
   sample = Run,
   genotype_raw = genotype,
   tissue_raw = tissue
 ) %>%
 mutate(
   genotype = ifelse(genotype_raw == "B73", "B73", "Inv4m"),
   tissue = tissue_raw
 )

{
 cat("=== CROW2020 DATA ===\n")
 cat("Expression matrix:", nrow(expr_crow), "genes x", ncol(expr_crow), "samples\n")
 cat("Metadata samples:", nrow(meta_crow), "\n")
 cat("\nGenotypes:\n")
 print(table(meta_crow$genotype))
 cat("\nTissues:\n")
 print(table(meta_crow$tissue))
 cat("\nTemperatures:\n")
 print(table(meta_crow$temp))
}
```

```{r extract_crow2020}
# Extract JMJ expression
jmj_in_crow <- jmj_genes$gene[jmj_genes$gene %in% rownames(expr_crow)]
cat("JMJ genes found in Crow2020:", length(jmj_in_crow), "\n")

if (length(jmj_in_crow) > 0) {
 # Extract expression for JMJ genes
 jmj_expr_crow <- expr_crow[jmj_in_crow, , drop = FALSE] %>%
   as.data.frame() %>%
   rownames_to_column("gene") %>%
   pivot_longer(
     cols = -gene,
     names_to = "sample",
     values_to = "logCPM"
   )

 # Match samples to metadata
 jmj_expr_crow <- jmj_expr_crow %>%
   left_join(
     meta_crow %>%
       dplyr::select(sample, genotype, tissue, temp),
     by = "sample"
   ) %>%
   filter(!is.na(genotype)) %>%
   left_join(jmj_genes, by = "gene") %>%
   mutate(
     experiment = "Crow2020 (Growth Chamber)",
     # Clean up tissue names
     tissue = str_replace(tissue, "_tissue$", "") %>%
       str_replace("_", " ") %>%
       str_replace("Primary root", "Root")
   )

 cat("Crow2020 expression records:", nrow(jmj_expr_crow), "\n")
}
```


# Combine Data

```{r combine_data}
# Combine both experiments
jmj_combined <- bind_rows(
 jmj_expr_psu %>%
   dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment),
 jmj_expr_crow %>%
   dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment)
)

# Order factors
jmj_combined <- jmj_combined %>%
 mutate(
   symbol = factor(symbol, levels = c("jmj2", "jmj4", "jmj21")),
   genotype = factor(genotype, levels = c("B73", "Inv4m")),
   experiment = factor(experiment, levels = c("Crow2020 (Growth Chamber)", "PSU2022 (Field)"))
 )

{
 cat("=== COMBINED DATA ===\n")
 cat("Total records:", nrow(jmj_combined), "\n")
 cat("\nBy experiment:\n")
 print(table(jmj_combined$experiment))
 cat("\nBy gene:\n")
 print(table(jmj_combined$symbol))
}
```

---

# Total JMJ Cluster Expression Analysis

The JMJ cluster in B73 contains 5 paralogs (jmj9, psi, jmj6, jmj2, jmj4) that are >97% identical.
In the V5 annotation, jmj9, psi, jmj6, and jmj2 are merged into a single gene model (Zm00001eb191790),
while jmj4 is separate (Zm00001eb191820). The Inv4m inversion removes 4 copies, leaving only jmj4.

To compare total cluster expression between genotypes, we sum the CPM values for jmj2 + jmj4
(which captures all 5 paralogs in the V5 annotation). The expected Inv4m expression under
copy-number scaling is B73/5 (one copy out of five).

## Calculate Total Cluster Expression

```{r total_cluster_calc}
# Convert logCPM to CPM and sum jmj2 + jmj4 per sample
# First, get unique sample info from PSU2022
psu_samples <- jmj_expr_psu %>%
  filter(symbol %in% c("jmj2", "jmj4")) %>%
  mutate(CPM = 2^logCPM) %>%
  group_by(sample, genotype, tissue, experiment) %>%
  summarise(total_CPM = sum(CPM), .groups = "drop")

# Then from Crow2020
crow_samples <- jmj_expr_crow %>%
  filter(symbol %in% c("jmj2", "jmj4")) %>%
  mutate(CPM = 2^logCPM) %>%
  group_by(sample, genotype, tissue, experiment) %>%
  summarise(total_CPM = sum(CPM), .groups = "drop")

# Combine and rename B73 to CTRL
total_cluster_samples <- bind_rows(psu_samples, crow_samples) %>%
  mutate(
    genotype = recode(genotype, "B73" = "CTRL"),
    genotype = factor(genotype, levels = c("CTRL", "Inv4m")),
    experiment = factor(experiment, levels = c("Crow2020 (Growth Chamber)", "PSU2022 (Field)"))
  )

{
  cat("=== TOTAL JMJ CLUSTER EXPRESSION ===\n")
  cat("Samples per experiment:\n")
  print(table(total_cluster_samples$experiment, total_cluster_samples$genotype))
  cat("\nMean total CPM by genotype and experiment:\n")
  print(
    total_cluster_samples %>%
      group_by(experiment, genotype) %>%
      summarise(mean_CPM = mean(total_CPM), sd_CPM = sd(total_CPM), n = n(), .groups = "drop")
  )
}
```

## Calculate Expected Values

```{r expected_values}
# Calculate CTRL mean per experiment and tissue for expected lines
expected_by_tissue <- total_cluster_samples %>%
  filter(genotype == "CTRL") %>%
  group_by(experiment, tissue) %>%
  summarise(
    CTRL_mean = mean(total_CPM),
    expected_Inv4m = mean(total_CPM) / 5,
    .groups = "drop"
  )

# Also calculate overall expected per experiment (for stats)
expected_overall <- total_cluster_samples %>%
  filter(genotype == "CTRL") %>%
  group_by(experiment) %>%
  summarise(
    CTRL_mean = mean(total_CPM),
    expected_Inv4m = mean(total_CPM) / 5,
    .groups = "drop"
  )

{
  cat("=== EXPECTED VALUES (CTRL/5) ===\n")
  cat("\nBy tissue:\n")
  print(expected_by_tissue)
  cat("\nOverall by experiment:\n")
  print(expected_overall)
}
```

## Statistical Test (Per Tissue)

### Linear Contrast t-Test

Test whether expression scales with copy number using a linear contrast:

**H₀: μ_CTRL - 5×μ_Inv4m = 0** (expression scales with 5:1 copy number)

This Welch-type t-test properly accounts for uncertainty in both group means and is appropriate
for small sample sizes (uses t-distribution with Welch-Satterthwaite degrees of freedom).

```{r linear_contrast_test}
# Linear contrast t-test for each experiment AND tissue
# H0: mean_CTRL - 5*mean_Inv4m = 0 (expression scales with copy number)
contrast_by_tissue <- total_cluster_samples %>%
  group_by(experiment, tissue) %>%
  summarise(
    mean_CTRL = mean(total_CPM[genotype == "CTRL"]),
    mean_Inv4m = mean(total_CPM[genotype == "Inv4m"]),
    var_CTRL = var(total_CPM[genotype == "CTRL"]),
    var_Inv4m = var(total_CPM[genotype == "Inv4m"]),
    n_CTRL = sum(genotype == "CTRL"),
    n_Inv4m = sum(genotype == "Inv4m"),
    .groups = "drop"
  ) %>%
  mutate(
    # Observed ratio
    observed_ratio = mean_CTRL / mean_Inv4m,
    # Expected Inv4m under copy number scaling
    expected_Inv4m = mean_CTRL / 5,
    # Percent of expected
    percent_of_expected = (mean_Inv4m / expected_Inv4m) * 100,
    # Linear contrast: mean_CTRL - 5*mean_Inv4m
    contrast = mean_CTRL - 5 * mean_Inv4m,
    # Standard error of contrast (note: coefficient 5 gets squared for variance)
    se_contrast = sqrt(var_CTRL / n_CTRL + 25 * var_Inv4m / n_Inv4m),
    # t statistic
    t_stat = contrast / se_contrast,
    # Welch-Satterthwaite degrees of freedom
    df = (var_CTRL / n_CTRL + 25 * var_Inv4m / n_Inv4m)^2 /
         ((var_CTRL / n_CTRL)^2 / (n_CTRL - 1) + (25 * var_Inv4m / n_Inv4m)^2 / (n_Inv4m - 1)),
    # Two-tailed p-value
    p_value = 2 * pt(abs(t_stat), df = df, lower.tail = FALSE)
  ) %>%
  # Add FDR adjustment across all tissues and significance stars
  mutate(
    fdr = p.adjust(p_value, method = "fdr"),
    sig_stars = case_when(
      fdr < 0.001 ~ "***",
      fdr < 0.01 ~ "**",
      fdr < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

# Display as table
contrast_table <- contrast_by_tissue %>%
  dplyr::select(experiment, tissue, n_CTRL, n_Inv4m, observed_ratio, t_stat, df, p_value, fdr, percent_of_expected, sig_stars) %>%
  mutate(
    observed_ratio = sprintf("%.2f:1", observed_ratio),
    t_stat = sprintf("%.2f", t_stat),
    df = sprintf("%.1f", df),
    p_value = sprintf("%.3f", p_value),
    fdr = sprintf("%.3f", fdr),
    percent_of_expected = sprintf("%.1f%%", percent_of_expected)
  )

knitr::kable(
  contrast_table,
  col.names = c("Experiment", "Tissue", "n CTRL", "n Inv4m", "Observed Ratio", "t", "df", "P-value", "FDR", "% of Expected", "Sig"),
  caption = "Linear contrast t-test: H₀ = μ_CTRL - 5×μ_Inv4m = 0 (expression scales with copy number)"
)
```

## Total Cluster Expression Boxplot

```{r total_cluster_boxplot, fig.width=14, fig.height=7}
# Define colors (matching ionome style: gold for CTRL, purple for Inv4m)
genotype_colors <- c("CTRL" = "gold", "Inv4m" = "#4a0f82")

# Prepare expected line data - add genotype = "Inv4m" so it dodges to Inv4m position
expected_lines <- expected_by_tissue %>%
  mutate(genotype = factor("Inv4m", levels = c("CTRL", "Inv4m")))

# Prepare significance annotation data
# Position stars just above the Inv4m boxplots (below CTRL)
sig_annotations <- contrast_by_tissue %>%
  dplyr::select(experiment, tissue, sig_stars, mean_Inv4m) %>%
  # Calculate y position: slightly above the Inv4m data
  left_join(
    total_cluster_samples %>%
      filter(genotype == "Inv4m") %>%
      group_by(experiment, tissue) %>%
      summarise(y_max = max(total_CPM), .groups = "drop"),
    by = c("experiment", "tissue")
  ) %>%
  mutate(
    y_pos = y_max * 1.15,  # Position 15% above max Inv4m value
    genotype = factor("Inv4m", levels = c("CTRL", "Inv4m"))
  )

# Create boxplot with tissue on x-axis (matching jmj2/jmj4 plot layout)
p_total_cluster <- ggplot(
  total_cluster_samples,
  aes(x = tissue, y = total_CPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  # Add expected line (CTRL/5) for each tissue at the Inv4m position
  geom_errorbar(
    data = expected_lines,
    aes(x = tissue, ymin = expected_Inv4m, ymax = expected_Inv4m, group = genotype),
    inherit.aes = FALSE,
    position = position_dodge(width = 0.7),
    width = 0.4,
    linewidth = 1,
    linetype = "dashed",
    color = "red"
  ) +
  # Add significance stars above Inv4m boxplots
  geom_text(
    data = sig_annotations,
    aes(x = tissue, y = y_pos, label = sig_stars, group = genotype),
    inherit.aes = FALSE,
    position = position_dodge(width = 0.7),
    size = 5,
    fontface = "bold",
    color = "black"
  ) +
  facet_grid(. ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 12
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 18),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = "Total JMJ Cluster Expression (CPM)",
    caption = "Dashed red line = expected Inv4m (CTRL/5). Significance: * FDR<0.05, ** FDR<0.01, *** FDR<0.001"
  )

print(p_total_cluster)
```

## Save Total Cluster Analysis Outputs

```{r save_total_cluster}
# Save total cluster expression plot
saveRDS(p_total_cluster, file.path(paths$figures, "jmj_total_cluster_expression.rds"))
ggsave(
  file.path(paths$figures, "jmj_total_cluster_expression.pdf"),
  p_total_cluster,
  width = 14, height = 7
)
ggsave(
  file.path(paths$figures, "jmj_total_cluster_expression.png"),
  p_total_cluster,
  width = 14, height = 7, dpi = 300
)

# Save per-tissue linear contrast t-test results
write_csv(
  contrast_by_tissue %>%
    dplyr::select(experiment, tissue, n_CTRL, n_Inv4m, mean_CTRL, mean_Inv4m,
                  observed_ratio, expected_Inv4m, percent_of_expected,
                  contrast, se_contrast, t_stat, df, p_value, fdr, sig_stars),
  file.path(paths$intermediate, "jmj_total_cluster_contrast_test.csv")
)

# Save sample-level data
write_csv(
  total_cluster_samples,
  file.path(paths$intermediate, "jmj_total_cluster_samples.csv")
)

{
  cat("=== TOTAL CLUSTER ANALYSIS SAVED ===\n")
  cat("  jmj_total_cluster_expression.rds/pdf/png\n")
  cat("  jmj_total_cluster_contrast_test.csv\n")
  cat("  jmj_total_cluster_samples.csv\n")
}
```

---

# Create Expression Boxplot (All 3 genes)

```{r boxplot_all, fig.width=14, fig.height=10}
# Define colors (matching ionome style: gold for B73, purple for Inv4m)
genotype_colors <- c("B73" = "gold", "Inv4m" = "#4a0f82")

p_jmj_expr <- ggplot(
  jmj_combined,
  aes(x = tissue, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_grid(symbol ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 11
    ),
    axis.text.y = element_text(color = "black", face = "bold"),
    axis.title = element_text(face = "bold", size = 13),
    strip.text.x = element_text(face = "bold", size = 16),
    strip.text.y = element_text(face = "bold.italic", size = 16),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_jmj_expr)
```

# Create Expression Boxplot (jmj2 and jmj4 only)

```{r boxplot_jmj2_jmj4, fig.width=14, fig.height=7}
# Filter for jmj2 and jmj4 only
jmj_cluster <- jmj_combined %>%
  filter(symbol %in% c("jmj2", "jmj4")) %>%
  mutate(symbol = factor(symbol, levels = c("jmj2", "jmj4")))

p_jmj_cluster <- ggplot(
  jmj_cluster,
  aes(x = tissue, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_grid(symbol ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 12
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text.x = element_text(face = "bold", size = 18),
    strip.text.y = element_text(face = "bold.italic", size = 18),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_jmj_cluster)
```



# Statistical Summary

```{r stats_summary}
# Calculate means and fold changes
jmj_summary <- jmj_combined %>%
 group_by(experiment, tissue, symbol, genotype) %>%
 summarise(
   mean_logCPM = mean(logCPM, na.rm = TRUE),
   sd_logCPM = sd(logCPM, na.rm = TRUE),
   n = n(),
   .groups = "drop"
 ) %>%
 pivot_wider(
   names_from = genotype,
   values_from = c(mean_logCPM, sd_logCPM, n),
   names_sep = "_"
 ) %>%
 mutate(
   log2FC = mean_logCPM_Inv4m - mean_logCPM_B73,
   direction = ifelse(log2FC < 0, "Down", "Up")
 )

{
 cat("=== EXPRESSION SUMMARY ===\n")
 print(
   jmj_summary %>%
     dplyr::select(experiment, tissue, symbol, log2FC, direction) %>%
     arrange(symbol, experiment, tissue)
 )
}
```



# Save Outputs

```{r save_outputs}
# Save all 3 genes plot
saveRDS(p_jmj_expr, file.path(paths$figures, "jmj_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "jmj_expression_boxplot.pdf"),
  p_jmj_expr,
  width = 14, height = 10
)
ggsave(
  file.path(paths$figures, "jmj_expression_boxplot.png"),
  p_jmj_expr,
  width = 14, height = 10, dpi = 300
)

# Save jmj2/jmj4 cluster plot
saveRDS(p_jmj_cluster, file.path(paths$figures, "jmj2_jmj4_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "jmj2_jmj4_expression_boxplot.pdf"),
  p_jmj_cluster,
  width = 14, height = 7
)
ggsave(
  file.path(paths$figures, "jmj2_jmj4_expression_boxplot.png"),
  p_jmj_cluster,
  width = 14, height = 7, dpi = 300
)

# Save summary table
write_csv(
  jmj_summary,
  file.path(paths$intermediate, "jmj_expression_summary.csv")
)

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("All 3 genes:\n")
  cat("  jmj_expression_boxplot.rds/pdf/png\n")
  cat("\njmj2/jmj4 cluster only:\n")
  cat("  jmj2_jmj4_expression_boxplot.rds/pdf/png\n")
  cat("\nSummary table:\n")
  cat("  jmj_expression_summary.csv\n")
}
```

---

# Cell Proliferation Genes Expression (Supplementary Figure)

This section plots expression of two cell proliferation-related genes across tissues:

- **pcna2** (Zm00001eb192050): Proliferating cell nuclear antigen 2, a DNA replication
  processivity factor. Upregulated in Inv4m (log2FC = 2.96, p < 0.001).
- **smo4** (Zm00001eb192850): Ribosome biogenesis protein NOP53 (SMO4 ortholog).
  The Arabidopsis ortholog regulates cell proliferation and organ size (Zhang et al.
  2015 JIPB). Downregulated in Inv4m (log2FC = -1.98).

## Define Cell Proliferation Genes

```{r cellprolif_genes}
cellprolif_genes <- tibble(
  gene = c("Zm00001eb192050", "Zm00001eb192850"),
  symbol = c("pcna2", "smo4"),
  description = c(
    "Proliferating cell nuclear antigen 2",
    "Ribosome biogenesis protein NOP53 (SMO4 ortholog)"
  )
)

cat("Cell proliferation genes:\n")
print(cellprolif_genes)
```

## Extract Cell Proliferation Gene Expression

```{r extract_cellprolif_psu2022}
# Extract from PSU2022
cellprolif_in_psu <- cellprolif_genes$gene[cellprolif_genes$gene %in% rownames(expr_psu)]
cat("Cell proliferation genes found in PSU2022:", length(cellprolif_in_psu), "\n")

if (length(cellprolif_in_psu) > 0) {
  cellprolif_expr_psu <- expr_psu[cellprolif_in_psu, , drop = FALSE] %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    pivot_longer(
      cols = -gene,
      names_to = "sample",
      values_to = "logCPM"
    ) %>%
    left_join(
      meta_psu %>%
        dplyr::select(library, Genotype, leaf_tissue, Treatment) %>%
        mutate(sample = library),
      by = "sample"
    ) %>%
    filter(!is.na(Genotype)) %>%
    left_join(cellprolif_genes, by = "gene") %>%
    mutate(
      experiment = "PSU2022 (Field)",
      tissue = paste0("Leaf ", leaf_tissue),
      genotype = ifelse(Genotype == "CTRL", "B73", "Inv4m")
    )

  cat("PSU2022 expression records:", nrow(cellprolif_expr_psu), "\n")
}
```

```{r extract_cellprolif_crow2020}
# Extract from Crow2020
cellprolif_in_crow <- cellprolif_genes$gene[cellprolif_genes$gene %in% rownames(expr_crow)]
cat("Cell proliferation genes found in Crow2020:", length(cellprolif_in_crow), "\n")

if (length(cellprolif_in_crow) > 0) {
  cellprolif_expr_crow <- expr_crow[cellprolif_in_crow, , drop = FALSE] %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    pivot_longer(
      cols = -gene,
      names_to = "sample",
      values_to = "logCPM"
    ) %>%
    left_join(
      meta_crow %>%
        dplyr::select(sample, genotype, tissue, temp),
      by = "sample"
    ) %>%
    filter(!is.na(genotype)) %>%
    left_join(cellprolif_genes, by = "gene") %>%
    mutate(
      experiment = "Crow2020 (Growth Chamber)",
      tissue = str_replace(tissue, "_tissue$", "") %>%
        str_replace("_", " ") %>%
        str_replace("Primary root", "Root")
    )

  cat("Crow2020 expression records:", nrow(cellprolif_expr_crow), "\n")
}
```

```{r combine_cellprolif}
# Combine both experiments
cellprolif_combined <- bind_rows(
  cellprolif_expr_psu %>%
    dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment),
  cellprolif_expr_crow %>%
    dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment)
)

# Order factors (pcna2 first, then smo4)
cellprolif_combined <- cellprolif_combined %>%
  mutate(
    symbol = factor(symbol, levels = c("pcna2", "smo4")),
    genotype = factor(genotype, levels = c("B73", "Inv4m")),
    experiment = factor(experiment, levels = c("Crow2020 (Growth Chamber)", "PSU2022 (Field)"))
  )

{
  cat("=== CELL PROLIFERATION GENES COMBINED DATA ===\n")
  cat("Total records:", nrow(cellprolif_combined), "\n")
  cat("\nBy gene:\n")
  print(table(cellprolif_combined$symbol))
  cat("\nBy experiment:\n")
  print(table(cellprolif_combined$experiment))
}
```

## Cell Proliferation Genes Boxplot

```{r cellprolif_boxplot, fig.width=14, fig.height=7}
p_cellprolif_expr <- ggplot(
  cellprolif_combined,
  aes(x = tissue, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_grid(symbol ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 12
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text.x = element_text(face = "bold", size = 18),
    strip.text.y = element_text(face = "bold.italic", size = 18),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_cellprolif_expr)
```

## Cell Proliferation Genes Statistics

```{r cellprolif_stats}
cellprolif_summary <- cellprolif_combined %>%
  group_by(experiment, tissue, symbol, genotype) %>%
  summarise(
    mean_logCPM = mean(logCPM, na.rm = TRUE),
    sd_logCPM = sd(logCPM, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = genotype,
    values_from = c(mean_logCPM, sd_logCPM, n),
    names_sep = "_"
  ) %>%
  mutate(
    log2FC = mean_logCPM_Inv4m - mean_logCPM_B73,
    direction = ifelse(log2FC < 0, "Down", "Up")
  )

{
  cat("=== CELL PROLIFERATION GENES EXPRESSION SUMMARY ===\n")
  print(
    cellprolif_summary %>%
      dplyr::select(experiment, tissue, symbol, log2FC, direction) %>%
      arrange(symbol, experiment, tissue)
  )
}
```

## Save Cell Proliferation Outputs

```{r save_cellprolif}
# Save cell proliferation plot (pcna2 + smo4)
saveRDS(p_cellprolif_expr, file.path(paths$figures, "cellprolif_pcna2_smo4_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "cellprolif_pcna2_smo4_expression_boxplot.pdf"),
  p_cellprolif_expr,
  width = 14, height = 7
)
ggsave(
  file.path(paths$figures, "cellprolif_pcna2_smo4_expression_boxplot.png"),
  p_cellprolif_expr,
  width = 14, height = 7, dpi = 300
)

# Save summary table
write_csv(
  cellprolif_summary,
  file.path(paths$intermediate, "cellprolif_pcna2_smo4_expression_summary.csv")
)

{
  cat("=== CELL PROLIFERATION SAVED OUTPUTS ===\n")
  cat("  cellprolif_pcna2_smo4_expression_boxplot.rds/pdf/png\n")
  cat("  cellprolif_pcna2_smo4_expression_summary.csv\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```
