---
title: "JMJ Cluster Expression Boxplot"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 warning = FALSE,
 message = FALSE,
 fig.width = 10,
 fig.height = 6
)

# Load project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

This script creates an expression boxplot for the JMJ cluster genes (jmj2, jmj4, jmj21)
comparing Inv4m vs CTRL across all tissues in:

1. **PSU2022**: Field experiment with 4 leaf positions
2. **Crow2020**: Growth chamber experiment with 5 tissues (V1_Leaf, V1_Root, V3_Leaf, V3_Root, V3_SAM)

The output is saved as an RDS file for composition with the JMJ microsynteny figure.

## Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggfx)
library(ggbeeswarm)
library(here)
```

## Define JMJ Genes

```{r jmj_genes}
jmj_genes <- tibble(
 gene = c("Zm00001eb191790", "Zm00001eb191820", "Zm00001eb190750"),
 symbol = c("jmj2", "jmj4", "jmj21"),
 description = c(
   "JUMONJI-transcription factor 2",
   "JUMONJI-transcription factor 4",
   "JUMONJI-transcription factor 21"
 )
)

cat("JMJ cluster genes:\n")
print(jmj_genes)
```



# Load PSU2022 Data

```{r load_psu2022}
# Load expression matrix (logCPM)
expr_psu <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_leaf_trt.rds"))

# Load metadata
meta_psu <- read.csv(file.path(paths$data, "PSU2022_metadata.csv"))

{
 cat("=== PSU2022 DATA ===\n")
 cat("Expression matrix:", nrow(expr_psu), "genes x", ncol(expr_psu), "samples\n")
 cat("Metadata samples:", nrow(meta_psu), "\n")
 cat("\nGenotypes:\n")
 print(table(meta_psu$Genotype))
 cat("\nLeaf positions:\n")
 print(table(meta_psu$leaf_tissue))
}
```

```{r extract_psu2022}
# Extract JMJ expression
jmj_in_psu <- jmj_genes$gene[jmj_genes$gene %in% rownames(expr_psu)]
cat("JMJ genes found in PSU2022:", length(jmj_in_psu), "\n")

if (length(jmj_in_psu) > 0) {
 # Extract expression for JMJ genes
 jmj_expr_psu <- expr_psu[jmj_in_psu, , drop = FALSE] %>%
   as.data.frame() %>%
   rownames_to_column("gene") %>%
   pivot_longer(
     cols = -gene,
     names_to = "sample",
     values_to = "logCPM"
   )

 # Match samples to metadata
 # Sample IDs in expression matrix should match library column in metadata
 jmj_expr_psu <- jmj_expr_psu %>%
   left_join(
     meta_psu %>%
       dplyr::select(library, Genotype, leaf_tissue, Treatment) %>%
       mutate(sample = library),
     by = "sample"
   ) %>%
   filter(!is.na(Genotype)) %>%
   left_join(jmj_genes, by = "gene") %>%
   mutate(
     experiment = "PSU2022 (Field)",
     tissue = paste0("Leaf ", leaf_tissue),
     genotype = ifelse(Genotype == "CTRL", "B73", "Inv4m")
   )

 cat("PSU2022 expression records:", nrow(jmj_expr_psu), "\n")
}
```



# Load Crow2020 Data

```{r load_crow2020}
# Load expression matrix (logCPM)
expr_crow <- readRDS(file.path(paths$intermediate, "normalized_expression_logCPM_crow2020.rds"))

# Load metadata
meta_crow <- read.csv(file.path(paths$data, "SraRunTable.csv"), check.names = FALSE)

# Standardize column names - use Run (SRR ID) as sample ID
meta_crow <- meta_crow %>%
 dplyr::rename(
   sample = Run,
   genotype_raw = genotype,
   tissue_raw = tissue
 ) %>%
 mutate(
   genotype = ifelse(genotype_raw == "B73", "B73", "Inv4m"),
   tissue = tissue_raw
 )

{
 cat("=== CROW2020 DATA ===\n")
 cat("Expression matrix:", nrow(expr_crow), "genes x", ncol(expr_crow), "samples\n")
 cat("Metadata samples:", nrow(meta_crow), "\n")
 cat("\nGenotypes:\n")
 print(table(meta_crow$genotype))
 cat("\nTissues:\n")
 print(table(meta_crow$tissue))
 cat("\nTemperatures:\n")
 print(table(meta_crow$temp))
}
```

```{r extract_crow2020}
# Extract JMJ expression
jmj_in_crow <- jmj_genes$gene[jmj_genes$gene %in% rownames(expr_crow)]
cat("JMJ genes found in Crow2020:", length(jmj_in_crow), "\n")

if (length(jmj_in_crow) > 0) {
 # Extract expression for JMJ genes
 jmj_expr_crow <- expr_crow[jmj_in_crow, , drop = FALSE] %>%
   as.data.frame() %>%
   rownames_to_column("gene") %>%
   pivot_longer(
     cols = -gene,
     names_to = "sample",
     values_to = "logCPM"
   )

 # Match samples to metadata
 jmj_expr_crow <- jmj_expr_crow %>%
   left_join(
     meta_crow %>%
       dplyr::select(sample, genotype, tissue, temp),
     by = "sample"
   ) %>%
   filter(!is.na(genotype)) %>%
   left_join(jmj_genes, by = "gene") %>%
   mutate(
     experiment = "Crow2020 (Growth Chamber)",
     # Clean up tissue names
     tissue = str_replace(tissue, "_tissue$", "") %>%
       str_replace("_", " ") %>%
       str_replace("Primary root", "Root")
   )

 cat("Crow2020 expression records:", nrow(jmj_expr_crow), "\n")
}
```


# Combine Data

```{r combine_data}
# Combine both experiments
jmj_combined <- bind_rows(
 jmj_expr_psu %>%
   dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment),
 jmj_expr_crow %>%
   dplyr::select(gene, symbol, logCPM, genotype, tissue, experiment)
)

# Order factors
jmj_combined <- jmj_combined %>%
 mutate(
   symbol = factor(symbol, levels = c("jmj2", "jmj4", "jmj21")),
   genotype = factor(genotype, levels = c("B73", "Inv4m")),
   experiment = factor(experiment, levels = c("Crow2020 (Growth Chamber)", "PSU2022 (Field)"))
 )

{
 cat("=== COMBINED DATA ===\n")
 cat("Total records:", nrow(jmj_combined), "\n")
 cat("\nBy experiment:\n")
 print(table(jmj_combined$experiment))
 cat("\nBy gene:\n")
 print(table(jmj_combined$symbol))
}
```



# Create Expression Boxplot (All 3 genes)

```{r boxplot_all, fig.width=14, fig.height=10}
# Define colors (matching ionome style: gold for B73, purple for Inv4m)
genotype_colors <- c("B73" = "gold", "Inv4m" = "#4a0f82")

p_jmj_expr <- ggplot(
  jmj_combined,
  aes(x = tissue, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_grid(symbol ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 11
    ),
    axis.text.y = element_text(color = "black", face = "bold"),
    axis.title = element_text(face = "bold", size = 13),
    strip.text.x = element_text(face = "bold", size = 16),
    strip.text.y = element_text(face = "bold.italic", size = 16),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_jmj_expr)
```

# Create Expression Boxplot (jmj2 and jmj4 only)

```{r boxplot_jmj2_jmj4, fig.width=14, fig.height=7}
# Filter for jmj2 and jmj4 only
jmj_cluster <- jmj_combined %>%
  filter(symbol %in% c("jmj2", "jmj4")) %>%
  mutate(symbol = factor(symbol, levels = c("jmj2", "jmj4")))

p_jmj_cluster <- ggplot(
  jmj_cluster,
  aes(x = tissue, y = logCPM, color = genotype)
) +
  geom_boxplot(
    width = 0.5,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.7)
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(
    size = 1.5,
    dodge.width = 0.7,
    alpha = 0.7
  ) %>%
  with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  facet_grid(symbol ~ experiment, scales = "free_x", space = "free_x") +
  scale_color_manual(values = genotype_colors, name = "Genotype") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      color = "black", face = "bold", size = 12
    ),
    axis.text.y = element_text(color = "black", face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 14),
    strip.text.x = element_text(face = "bold", size = 18),
    strip.text.y = element_text(face = "bold.italic", size = 18),
    strip.background = element_rect(fill = "grey95", color = NA),
    legend.position = "bottom",
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3)
  ) +
  labs(
    x = NULL,
    y = expression(bold(log[2]*"(CPM)"))
  )

print(p_jmj_cluster)
```



# Statistical Summary

```{r stats_summary}
# Calculate means and fold changes
jmj_summary <- jmj_combined %>%
 group_by(experiment, tissue, symbol, genotype) %>%
 summarise(
   mean_logCPM = mean(logCPM, na.rm = TRUE),
   sd_logCPM = sd(logCPM, na.rm = TRUE),
   n = n(),
   .groups = "drop"
 ) %>%
 pivot_wider(
   names_from = genotype,
   values_from = c(mean_logCPM, sd_logCPM, n),
   names_sep = "_"
 ) %>%
 mutate(
   log2FC = mean_logCPM_Inv4m - mean_logCPM_B73,
   direction = ifelse(log2FC < 0, "Down", "Up")
 )

{
 cat("=== EXPRESSION SUMMARY ===\n")
 print(
   jmj_summary %>%
     dplyr::select(experiment, tissue, symbol, log2FC, direction) %>%
     arrange(symbol, experiment, tissue)
 )
}
```



# Save Outputs

```{r save_outputs}
# Save all 3 genes plot
saveRDS(p_jmj_expr, file.path(paths$figures, "jmj_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "jmj_expression_boxplot.pdf"),
  p_jmj_expr,
  width = 14, height = 10
)
ggsave(
  file.path(paths$figures, "jmj_expression_boxplot.png"),
  p_jmj_expr,
  width = 14, height = 10, dpi = 300
)

# Save jmj2/jmj4 cluster plot
saveRDS(p_jmj_cluster, file.path(paths$figures, "jmj2_jmj4_expression_boxplot.rds"))
ggsave(
  file.path(paths$figures, "jmj2_jmj4_expression_boxplot.pdf"),
  p_jmj_cluster,
  width = 14, height = 7
)
ggsave(
  file.path(paths$figures, "jmj2_jmj4_expression_boxplot.png"),
  p_jmj_cluster,
  width = 14, height = 7, dpi = 300
)

# Save summary table
write_csv(
  jmj_summary,
  file.path(paths$intermediate, "jmj_expression_summary.csv")
)

{
  cat("=== SAVED OUTPUTS ===\n")
  cat("All 3 genes:\n")
  cat("  jmj_expression_boxplot.rds/pdf/png\n")
  cat("\njmj2/jmj4 cluster only:\n")
  cat("  jmj2_jmj4_expression_boxplot.rds/pdf/png\n")
  cat("\nSummary table:\n")
  cat("  jmj_expression_summary.csv\n")
}
```



# Session Info

```{r session_info}
sessionInfo()
```
