---
title: "Inv4m Breakpoint Tables S1 and S2"
author: "inv4m"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")

library(dplyr)
library(tidyr)
library(readr)
library(knitr)

# Display labels for genomes (internal key -> table label)
genome_labels <- c(B73 = "B73", PT = "PT", TIL18 = "TIL18", Mi21 = "Mi21 NIL")
```

# Overview

This notebook generates the data for Supplementary Tables S1 (Inv4m
delimitation and breakpoints) and S2 (breakpoint knob repeats) from:

1. **Manually curated coordinates** — breakpoint boundaries and bounding genes
   determined by inspecting AnchorWave CDS-anchored whole-genome alignments
   and identifying the genes flanking the strand-transition zones.
2. **BLAST megablast results** — knob180 (M32524.1) and TR-1 (AY083942.1)
   repeat hits against chromosome 4, generated by
   `scripts/02_genomics_foundation/run_repeat_annotation_4genomes.sh`.

## Genomes

| Label    | Assembly                              | Chr4 name            |
|----------|---------------------------------------|----------------------|
| B73      | Zm-B73-REFERENCE-NAM-5.0              | chr4                 |
| PT       | Zm-PT-REFERENCE-HiLo-1.0             | PT04                 |
| TIL18    | Zx-TIL18-REFERENCE-PanAnd-1.0        | chr4                 |
| Mi21 NIL | Mi21 NIL HiFi (hifiasm + RagTag)      | scf00000010_RagTag   |

## Curation method

Breakpoint coordinates were defined by the AnchorWave alignment endpoints
where reference-orientation synteny was lost. Bounding genes were identified
by inspecting gene annotations (B73 NAM5 GFF3 for B73, PT HiLo-1.0 GFF3 for
PT, TIL18 PanAnd-1.0 GFF3 for TIL18, and Liftoff B73-to-Mi21 GFF3 for
Mi21 NIL) flanking the breakpoint segments.

# Table S1: Inv4m delimitation and breakpoints

## Inv4m limits

Curated inversion boundaries — the first and last genes within the inverted
region, as identified by AnchorWave strand analysis.

```{r inv4m-limits}
inv4m_limits <- tribble(
  ~genome,     ~start,      ~end,        ~start_gene,           ~end_gene,
  "B73",       172883881,   188131462,   "Zm00001eb190470",     "Zm00001eb194800",
  "PT",        186925484,   173486186,   "Zm00109aa017629",     "Zm00109aa018009",
  "TIL18",     180365316,   193570652,   "Zx00002aa015554",     "Zx00002aa015905",
  "Mi21",      156752807,   167620500,   "Zm00001eb194800",     "Zm00001eb190470"
) %>%
  mutate(length = abs(end - start))

kable(inv4m_limits, format.args = list(big.mark = ","))
```

## Breakpoint segments

Gene-free segments flanking the inversion, defined as the gap between the
last gene on the collinear side and the first gene on the inverted side.

"Upstream" and "Downstream" are defined in B73-sense orientation: upstream is
the breakpoint at lower B73 coordinates (~172.9 Mb), downstream at higher
(~188.1 Mb). For highland genomes carrying the inversion, the physical
positions are swapped but the labels follow the B73 convention.

```{r breakpoint-segments}
breakpoint_segments <- tribble(
  ~genome,     ~up_start,   ~up_end,     ~down_start, ~down_end,
  # Upstream/downstream follow B73-sense orientation (~172.9 Mb / ~188.1 Mb).
  # For highland genomes the inversion swaps the physical positions, but
  # the labels are defined by which B73 breakpoint each segment corresponds to.
  "B73",       172561330,   172883881,   188131461,   188220418,
  "PT",        173369064,   173486186,   186925483,   187092654,
  "TIL18",     180269950,   180365316,   193570651,   193734606,
  # Mi21: B73-sense upstream breakpoint maps to Mi21 ~167.6 Mb (higher coord),
  #        B73-sense downstream maps to Mi21 ~156.6 Mb (lower coord).
  "Mi21",      167621241,   167838764,   156560061,   156750619
) %>%
  mutate(
    up_span   = up_end - up_start,
    down_span = down_end - down_start
  )

kable(breakpoint_segments, format.args = list(big.mark = ","))
```

## Bounding genes

Genes flanking each breakpoint segment (outside → inside the inversion).

```{r bounding-genes}
bounding_genes <- tribble(
  ~genome,     ~up_outer,             ~up_inner,             ~down_inner,           ~down_outer,
  "B73",       "Zm00001eb190450",     "Zm00001eb190470",     "Zm00109aa017627",     "Zm00001eb194820",
  "PT",        "Zm00109aa017627",     "Zm00109aa017629",     "Zx00002aa015905",     "Zm00109aa017629",
  "TIL18",     "Zx00002aa015905",     "Zx00002aa015906",     "Zx00002aa015906",     "Zx00002aa015554",
  "Mi21",      "Zm00001eb190450",     "Zm00001eb194800",     "Zm00001eb190470",     "Zm00001eb194820"
)

kable(bounding_genes)
```

## Export Table S1

```{r export-s1}
apply_label <- function(df) {
  df %>% mutate(genome = genome_labels[genome])
}

s1_limits <- inv4m_limits %>%
  select(genome, start, end, length, start_gene, end_gene) %>%
  apply_label()

s1_segments <- breakpoint_segments %>%
  select(genome, up_start, up_end, up_span, down_start, down_end, down_span) %>%
  apply_label()

s1_genes <- bounding_genes %>% apply_label()

write_csv(s1_limits,   file.path(paths$intermediate, "table_S1_inv4m_limits.csv"))
write_csv(s1_segments, file.path(paths$intermediate, "table_S1_breakpoint_segments.csv"))
write_csv(s1_genes,    file.path(paths$intermediate, "table_S1_bounding_genes.csv"))
```

# Table S2: Breakpoint knob repeats

## Load BLAST results

BLAST+ megablast (outfmt 6) results from
`run_repeat_annotation_4genomes.sh`. Columns: qseqid, sseqid, pident,
length, mismatch, gapopen, qstart, qend, sstart, send, evalue, bitscore.

```{r load-blast}
blast_cols <- c("qseqid", "sseqid", "pident", "length", "mismatch",
                "gapopen", "qstart", "qend", "sstart", "send",
                "evalue", "bitscore")

genomes <- c("B73", "PT", "TIL18", "Mi21")

read_blast <- function(repeat_type, genome) {
  f <- file.path(paths$intermediate, paste0(repeat_type, "_", genome, ".blast"))
  read_tsv(f, col_names = blast_cols, col_types = cols(.default = "c")) %>%
    mutate(
      across(c(length, sstart, send, bitscore), as.numeric),
      genome = genome,
      repeat_type = repeat_type,
      # Normalize subject coordinates so smin < smax
      smin = pmin(sstart, send),
      smax = pmax(sstart, send)
    )
}

all_blast <- bind_rows(
  lapply(genomes, function(g) read_blast("knob180", g)),
  lapply(genomes, function(g) read_blast("TR-1", g))
)

cat(sprintf("Loaded %s BLAST hits across %d genomes\n",
            format(nrow(all_blast), big.mark = ","), length(genomes)))
```

## Count repeats in breakpoint segments

```{r count-repeats}
# Breakpoint segment coordinates per genome (B73-sense upstream/downstream)
segments <- breakpoint_segments %>%
  select(genome, up_start, up_end, down_start, down_end)

# Internal knob region: between inv4m boundaries
internal_regions <- inv4m_limits %>%
  mutate(
    int_start = pmin(start, end),
    int_end   = pmax(start, end)
  ) %>%
  select(genome, int_start, int_end)

count_in_region <- function(blast_df, genome_id, region_start, region_end) {
  blast_df %>%
    filter(
      genome == genome_id,
      repeat_type == "knob180",
      (smin >= region_start & smin <= region_end) |
      (smax >= region_start & smax <= region_end)
    ) %>%
    summarise(
      count       = n(),
      match_len   = sum(length),
      hit_start   = ifelse(n() > 0, min(smin), NA_real_),
      hit_end     = ifelse(n() > 0, max(smax), NA_real_),
      span        = ifelse(n() > 0, max(smax) - min(smin), NA_real_),
      .groups     = "drop"
    )
}

# Upstream breakpoint repeats
up_repeats <- lapply(seq_len(nrow(segments)), function(i) {
  s <- segments[i, ]
  count_in_region(all_blast, s$genome, s$up_start, s$up_end) %>%
    mutate(genome = s$genome, section = "upstream",
           seg_span = s$up_end - s$up_start)
}) %>% bind_rows()

# Downstream breakpoint repeats
down_repeats <- lapply(seq_len(nrow(segments)), function(i) {
  s <- segments[i, ]
  count_in_region(all_blast, s$genome, s$down_start, s$down_end) %>%
    mutate(genome = s$genome, section = "downstream",
           seg_span = s$down_end - s$down_start)
}) %>% bind_rows()

# Internal repeats
int_repeats <- lapply(seq_len(nrow(internal_regions)), function(i) {
  r <- internal_regions[i, ]
  count_in_region(all_blast, r$genome, r$int_start, r$int_end) %>%
    mutate(genome = r$genome, section = "internal",
           seg_span = r$int_end - r$int_start)
}) %>% bind_rows()

knob_repeats <- bind_rows(up_repeats, down_repeats, int_repeats) %>%
  mutate(
    pct_knob_span = ifelse(!is.na(span), round(span / seg_span * 100), NA),
    pct_knob      = ifelse(!is.na(span) & span > 0,
                           round(match_len / span * 100), NA),
    genome_label  = genome_labels[genome]
  ) %>%
  select(genome = genome_label, section, count, match_len,
         hit_start, hit_end, span, pct_knob_span, pct_knob)
```

## Results

### Upstream breakpoint segment

```{r show-upstream}
knob_repeats %>% filter(section == "upstream") %>%
  kable(format.args = list(big.mark = ","))
```

### Internal knob repeat

```{r show-internal}
knob_repeats %>% filter(section == "internal") %>%
  select(-pct_knob_span, -pct_knob) %>%
  kable(format.args = list(big.mark = ","))
```

### Downstream breakpoint segment

```{r show-downstream}
knob_repeats %>% filter(section == "downstream") %>%
  kable(format.args = list(big.mark = ","))
```

## Export Table S2

```{r export-s2}
write_csv(knob_repeats, file.path(paths$intermediate, "table_S2_knob_repeats.csv"))
```

# Validation

Compare computed values against the hardcoded values in `main.tex`.

```{r validate}
# Spot-check B73 upstream: expect count=352, match_len=59580
b73_up <- knob_repeats %>%
  filter(genome == "B73", section == "upstream")
cat(sprintf("B73 upstream: count=%d (expect 352), match_len=%d (expect 59580)\n",
            b73_up$count, b73_up$match_len))

# Spot-check Mi21 downstream: expect count=57, match_len=9865
mi21_down <- knob_repeats %>%
  filter(genome == "Mi21 NIL", section == "downstream")
cat(sprintf("Mi21 downstream: count=%d (expect 57), match_len=%d (expect 9865)\n",
            mi21_down$count, mi21_down$match_len))
```

# Session info

```{r session}
sessionInfo()
```
