---
title: "Permutation Test for WGCNA Connectivity Loss"
author: "F. V. Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = TRUE
)
knitr::opts_knit$set(root.dir = here::here())
```

# Overview

The WGCNA analysis (step 06) found that 12 of 13 DEG coexpression modules
lose intramodular connectivity ($\Delta k_{Within}$) when comparing Inv4m
to CTRL networks. A potential concern is that this connectivity loss could
be an artifact of splitting samples by genotype: genes selected as DEGs
will, by definition, have different mean expression between groups, and
this could inflate correlation differences when networks are built
separately.

This notebook addresses that concern with two complementary approaches:

1. **Genotype label permutation test**: Shuffle genotype labels 1000 times,
   recompute $\Delta k_{Within}$ per module, and compare the observed
   connectivity loss to the empirical null distribution. This directly tests
   whether the observed loss exceeds what random sample splitting produces.

2. **Topology control gene set**: Apply the same analysis to 218
   photosynthesis and senescence genes (Ojeda 2026) that are NOT
   differentially expressed between genotypes but DO have real coexpression
   structure driven by the leaf developmental gradient. If the connectivity
   loss is an artifact of sample splitting, these control genes should show
   the same pattern. If it is specific to DEGs, they should not.

# Setup

```{r load_packages}
library(here)
library(tidyverse)
library(WGCNA)
library(ggridges)

allowWGCNAThreads()
set.seed(42)
```

```{r config}
N_PERM <- 10000
RUN_DIR <- here("results/inversion_paper/field_perturbation/run_20251231_201332")
OUTPUT_DIR <- file.path(RUN_DIR, "06_preservation")

FILTER_DIR    <- file.path(RUN_DIR, "02_gene_filter")
DATA_PREP_DIR <- file.path(RUN_DIR, "01_data_prep")
CONSENSUS_DIR <- file.path(RUN_DIR, "04_consensus_networks")
REFERENCE_DIR <- file.path(RUN_DIR, "03_reference_network")

source(here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Load Data

```{r load_data}
# DEG-filtered expression (465 genes x 43 samples)
expr_filtered <- readRDS(file.path(FILTER_DIR, "filtered_expression.rds"))
sample_meta   <- read_csv(file.path(DATA_PREP_DIR, "sample_metadata.csv"),
                           show_col_types = FALSE)
ctrl_modules  <- read_csv(file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
                           show_col_types = FALSE)
SOFT_POWER    <- readRDS(file.path(REFERENCE_DIR, "ctrl", "selected_power.rds"))

# Full expression matrix (24k genes x 43 samples, PSU2022 only)
full_expr <- readRDS(file.path(DATA_PREP_DIR, "combined_expression.rds"))
if (inherits(full_expr, "EList")) {
  full_expr_mat <- full_expr$E
} else {
  full_expr_mat <- as.matrix(full_expr)
}

cat("DEG expression:", nrow(expr_filtered), "genes x", ncol(expr_filtered), "samples\n")
cat("Full expression:", nrow(full_expr_mat), "genes x", ncol(full_expr_mat), "samples\n")
cat("Soft power:", SOFT_POWER, "\n")
```

# Control Gene Sets

Photosynthesis and senescence hub genes from Ojeda 2026. In the companion
phosphorus paper, we showed that the expression indices for these gene sets
do not differ between Inv4m and CTRL genotypes. They are driven by the leaf
developmental gradient (young leaves = photosynthesis, old leaves =
senescence), providing biologically structured coexpression that is
independent of the Inv4m introgression.

```{r load_control_genes}
photo_genes <- read.table(
  file.path(paths$data, "photosynthesis_expression_index_genes_ojeda2026.txt"),
  header = FALSE, stringsAsFactors = FALSE
)$V1

sag_genes <- read.table(
  file.path(paths$data, "senesence_hubs_ojeda2026.tab"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)$gene

# Exclude any overlap with DEGs
deg_genes <- ctrl_modules$gene
photo_control <- setdiff(intersect(photo_genes, rownames(full_expr_mat)), deg_genes)
sag_control   <- setdiff(intersect(sag_genes, rownames(full_expr_mat)), deg_genes)
control_genes <- unique(c(photo_control, sag_control))

cat("Photosynthesis (non-DEG, in matrix):", length(photo_control), "\n")
cat("Senescence hubs (non-DEG, in matrix):", length(sag_control), "\n")
cat("Combined control genes:", length(control_genes), "\n")
```

# Prepare Expression Matrices

```{r prepare_expression}
# DEG module colors aligned to gene order
module_colors <- setNames(ctrl_modules$color, ctrl_modules$gene)
module_colors <- module_colors[rownames(expr_filtered)]

# Genotype labels
all_samples <- sample_meta$sample_id
true_labels <- setNames(sample_meta$Genotype, sample_meta$sample_id)
n_ctrl  <- sum(true_labels == "CTRL")
n_inv4m <- sum(true_labels == "Inv4m")
n_total <- length(true_labels)

# WGCNA format: samples x genes
datExpr_all     <- t(expr_filtered[, all_samples])
datExpr_control <- t(full_expr_mat[control_genes, all_samples])

cat("DEG matrix:", nrow(datExpr_all), "samples x", ncol(datExpr_all), "genes\n")
cat("Control matrix:", nrow(datExpr_control), "samples x", ncol(datExpr_control), "genes\n")
cat("CTRL:", n_ctrl, "| Inv4m:", n_inv4m, "| Total:", n_total, "\n")
```

# WGCNA Modules for Control Genes

Build modules from control genes using CTRL samples only, mirroring
how the DEG modules were constructed.

```{r control_modules}
datExpr_ctrl_only <- datExpr_control[which(true_labels == "CTRL"), ]
adj_ctrl_control <- adjacency(datExpr_ctrl_only, power = SOFT_POWER, type = "signed")
ctrl_tree <- hclust(as.dist(1 - adj_ctrl_control), method = "average")
invisible(capture.output(
  ctrl_control_modules <- cutreeDynamic(
    dendro = ctrl_tree,
    distM = 1 - adj_ctrl_control,
    deepSplit = 4,
    minClusterSize = 10
  )
))
control_module_colors_raw <- labels2colors(ctrl_control_modules)

# Remap WGCNA standard colors to non-clashing names with matching hex
ctrl_color_remap <- c(
  turquoise = "sky",
  blue      = "peach",
  brown     = "jade",
  yellow    = "coral",
  green     = "lavender",
  black     = "umber",
  red       = "mustard",
  pink      = "rose",
  magenta   = "aqua",
  purple    = "spring",
  grey      = "grey"
)
# Hex codes for plotting (keyed by new name)
ctrl_color_hex <- c(
  sky      = "#5DA5DA",
  peach    = "#FAA43A",
  jade     = "#60BD68",
  coral    = "#F15854",
  lavender = "#B276B2",
  umber    = "#8D4B08",
  mustard  = "#DECF3F",
  rose     = "#F17CB0",
  aqua     = "#66E3D9",
  spring   = "#00FF7F",
  grey     = "grey"
)

control_module_colors <- ctrl_color_remap[control_module_colors_raw]
names(control_module_colors) <- control_genes

cat("Control modules found:", length(unique(ctrl_control_modules[ctrl_control_modules != 0])), "\n")
cat("Unassigned (grey):", sum(ctrl_control_modules == 0), "of", length(ctrl_control_modules), "\n\n")

tibble(module = control_module_colors) |>
  filter(module != "grey") |>
  count(module, sort = TRUE) |>
  print()
```

## Bootstrap Support (LFP) for Control Modules

Quantify module stability using the same bootstrap resampling approach
as the DEG modules (step 05). Resample CTRL samples with replacement
1000 times, rebuild the network each time, and compute the Largest
Fragment Proportion (LFP) per module.

```{r control_bootstrap, cache=TRUE}
N_BOOT <- 1000

# Reference module assignments (numeric, excluding grey)
ctrl_ref_modules <- data.frame(
  gene = control_genes,
  module = ctrl_control_modules,
  color = control_module_colors,
  stringsAsFactors = FALSE
)

boot_results <- vector("list", N_BOOT)

for (b in seq_len(N_BOOT)) {
  n_samp <- nrow(datExpr_ctrl_only)
  boot_idx <- sample(n_samp, n_samp, replace = TRUE)
  expr_boot <- datExpr_ctrl_only[boot_idx, ]

  # Match original control module construction:
  # adjacency + hclust + cutreeDynamic
  adj_boot <- adjacency(expr_boot, power = SOFT_POWER,
                         type = "signed")
  tree_boot <- hclust(as.dist(1 - adj_boot),
                       method = "average")
  invisible(capture.output(
    mods_boot <- cutreeDynamic(
      dendro = tree_boot,
      distM = 1 - adj_boot,
      deepSplit = 4,
      minClusterSize = 10
    )
  ))

  boot_results[[b]] <- data.frame(
    gene = colnames(expr_boot),
    module = mods_boot,
    stringsAsFactors = FALSE
  )
}

# Compute LFP per module
ctrl_modules_list <- unique(
  ctrl_ref_modules$module[ctrl_ref_modules$module != 0]
)

ctrl_lfp <- map_dfr(ctrl_modules_list, function(mod) {
  mod_genes <- ctrl_ref_modules$gene[ctrl_ref_modules$module == mod]
  mod_color <- ctrl_ref_modules$color[ctrl_ref_modules$module == mod][1]

  lfp_vals <- sapply(boot_results, function(boot) {
    boot_sub <- boot[boot$gene %in% mod_genes, ]
    counts <- table(boot_sub$module)
    counts_nz <- counts[names(counts) != "0"]
    if (length(counts_nz) == 0) return(0)
    max(counts_nz) / length(mod_genes)
  })

  tibble(
    module = mod,
    color = mod_color,
    n_genes = length(mod_genes),
    mean_lfp = mean(lfp_vals),
    sd_lfp = sd(lfp_vals),
    se_lfp = sd(lfp_vals) / sqrt(N_BOOT)
  )
})

cat("=== CONTROL MODULE BOOTSTRAP SUPPORT ===\n")
print(ctrl_lfp |> arrange(desc(mean_lfp)))
```

## GO Enrichment for Control Modules

Run GO biological process enrichment on the control modules using the same
approach as the WGCNA perturbation pipeline (step 07), then reduce redundant
terms with rrvgo.

```{r control_go_enrichment, cache=TRUE}
library(clusterProfiler)
library(rrvgo)
library(AnnotationHub)

TERM2GENE <- readRDS(here("data", "TERM2GENE.rds"))
TERM2NAME <- readRDS(here("data", "TERM2NAME.rds"))

# Universe = all genes in the full expression matrix
ctrl_universe <- rownames(full_expr_mat)

# Gene lists per control module (excluding grey)
ctrl_gene_lists <- split(
  names(control_module_colors[control_module_colors != "grey"]),
  control_module_colors[control_module_colors != "grey"]
)

cat("Control modules for GO:", length(ctrl_gene_lists), "\n")
cat("Module sizes:", paste(names(ctrl_gene_lists),
    sapply(ctrl_gene_lists, length), sep = "=", collapse = ", "), "\n")

# Enrichment function
ctrl_ego <- function(x) {
  enricher(
    gene = x,
    pvalueCutoff = 0.05,
    pAdjustMethod = "none",
    universe = ctrl_universe,
    minGSSize = 10,
    maxGSSize = 500,
    qvalueCutoff = 0.2,
    TERM2GENE = TERM2GENE[TERM2GENE$GO %in% TERM2NAME[["BP"]]$go_id, ],
    TERM2NAME = TERM2NAME[["BP"]]
  )
}

# Run compareCluster
ctrl_compGO <- compareCluster(
  geneCluster = ctrl_gene_lists,
  fun = ctrl_ego
)

ctrl_go_results <- as.data.frame(ctrl_compGO@compareClusterResult) |>
  dplyr::rename(module_color = Cluster) |>
  arrange(module_color, pvalue)

cat("Total enriched BP terms:", nrow(ctrl_go_results), "\n")
cat("Terms per module:\n")
print(table(ctrl_go_results$module_color))

# rrvgo semantic reduction
ah <- AnnotationHub()
org.Zm.eg.db <- ah[["AH119718"]]

ctrl_go_filtered <- ctrl_go_results |>
  filter(Count > 1) |>
  group_by(module_color) |>
  arrange(pvalue, .by_group = TRUE) |>
  ungroup()

ctrl_go_ids <- unique(ctrl_go_results$ID)
if (length(ctrl_go_ids) > 1) {
  simMatrix <- calculateSimMatrix(
    ctrl_go_ids,
    orgdb = org.Zm.eg.db,
    ont = "BP",
    method = "Rel"
  )
  scores <- setNames(-log10(ctrl_go_results$pvalue), ctrl_go_results$ID)
  reducedTerms <- reduceSimMatrix(
    simMatrix, scores, threshold = 0.7, orgdb = org.Zm.eg.db
  )
  ctrl_go_rrvgo <- ctrl_go_filtered |>
    filter(ID %in% reducedTerms$parent)
} else {
  ctrl_go_rrvgo <- ctrl_go_filtered
}

# Top 1 BP term per control module
ctrl_top_bp <- ctrl_go_rrvgo |>
  group_by(module_color) |>
  slice_min(pvalue, n = 1, with_ties = FALSE) |>
  ungroup() |>
  dplyr::select(module_color, ID, Description, pvalue, Count)

cat("\nTop BP term per control module:\n")
print(ctrl_top_bp)
```

## Module Size Comparison

```{r module_size_comparison, fig.width=8, fig.height=4}
deg_sizes <- tibble(
  module = module_colors,
  type = "DEG"
) |>
  filter(module != "grey", !is.na(module)) |>
  count(module, type, name = "n_genes")

ctrl_sizes <- tibble(
  module = control_module_colors,
  type = "Control"
) |>
  filter(module != "grey") |>
  count(module, type, name = "n_genes")

bind_rows(deg_sizes, ctrl_sizes) |>
  ggplot(aes(x = n_genes, fill = type)) +
  geom_histogram(bins = 15, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c(DEG = "grey60", Control = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(x = "Module size (genes)", y = "Count", fill = NULL,
       title = "Module size distributions overlap between DEG and control")
```

# Helper Function

```{r helper_function}
compute_module_delta_k <- function(datExpr_a, datExpr_b, power, mod_colors) {
  adj_a <- adjacency(datExpr_a, power = power, type = "signed")
  adj_b <- adjacency(datExpr_b, power = power, type = "signed")

  k_a <- intramodularConnectivity(adj_a, mod_colors)
  k_b <- intramodularConnectivity(adj_b, mod_colors)

  delta_kWithin <- k_b$kWithin - k_a$kWithin
  delta_kOut    <- k_b$kOut - k_a$kOut

  module_ids <- unique(mod_colors[mod_colors != "grey" & !is.na(mod_colors)])
  list(
    kWithin = sapply(module_ids, function(m) {
      idx <- which(mod_colors == m)
      mean(delta_kWithin[idx], na.rm = TRUE)
    }),
    kOut = sapply(module_ids, function(m) {
      idx <- which(mod_colors == m)
      mean(delta_kOut[idx], na.rm = TRUE)
    })
  )
}

# Density ridgeline panel constructor (used by kWithin sections)
# points_df must have columns: module, z_observed, fdr
# fill_color: single color string OR named vector (module -> hex)
# go_terms: optional named vector (module -> GO description) for y-axis labels
make_density_panel <- function(density_df, points_df, panel_title,
                               fill_color = "identity",
                               go_terms = NULL,
                               module_sizes = NULL) {
  # Build y-axis labels with module size: "green (n=30)"
  mod_levels <- levels(density_df$module)
  if (!is.null(module_sizes)) {
    y_labels <- sapply(mod_levels, function(m) {
      n <- module_sizes[m]
      if (!is.na(n)) sprintf("%s (n=%d)", m, n) else m
    })
  } else {
    y_labels <- setNames(mod_levels, mod_levels)
  }

  if (fill_color[1] == "identity") {
    p <- ggplot(density_df, aes(x = z_value, y = module, fill = module)) +
      geom_density_ridges(color = "grey40", alpha = 0.7,
                          scale = 0.9, rel_min_height = 0.01) +
      scale_fill_identity() +
      guides(fill = "none")
  } else if (length(fill_color) == 1) {
    p <- ggplot(density_df, aes(x = z_value, y = module)) +
      geom_density_ridges(fill = fill_color, color = "grey40", alpha = 0.7,
                          scale = 0.9, rel_min_height = 0.01)
  } else {
    p <- ggplot(density_df, aes(x = z_value, y = module, fill = module)) +
      geom_density_ridges(color = "grey40", alpha = 0.7,
                          scale = 0.9, rel_min_height = 0.01) +
      scale_fill_manual(values = fill_color) +
      guides(fill = "none")
  }
  p <- p +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_segment(
      data = points_df,
      aes(x = z_observed, xend = z_observed,
          y = as.numeric(module),
          yend = as.numeric(module) + 0.8),
      color = "red", linewidth = 0.8, inherit.aes = FALSE
    ) +
    geom_text(
      data = points_df |> filter(fdr < 0.05),
      aes(x = z_observed - 0.6, y = as.numeric(module) + 0.35,
          label = sprintf("FDR=%.3f", fdr)),
      color = "red", size = 3, hjust = 0.5, inherit.aes = FALSE
    ) +
    scale_y_discrete(labels = y_labels) +
    scale_x_continuous(limits = c(NA, 5),
                       breaks = seq(-10, 10, 2),
                       minor_breaks = seq(-10, 10, 2)) +
    theme_minimal(base_size = 14) +
    labs(y = NULL, x = expression(Z[Delta * k]), title = panel_title)

  # GO annotations inside plot area, right-aligned at x = 2.5
  if (!is.null(go_terms)) {
    go_df <- tibble(
      module = names(go_terms),
      go_desc = unname(go_terms)
    ) |>
      filter(module %in% levels(density_df$module)) |>
      mutate(
        module = factor(module, levels = levels(density_df$module)),
        y_pos = as.numeric(module) + 0.35
      )
    p <- p +
      geom_text(
        data = go_df,
        aes(x = 2.5, y = y_pos, label = go_desc),
        hjust = 0, size = 14 * 0.8 / .pt, fontface = "italic",
        inherit.aes = FALSE
      )
  }

  p
}
```

# Observed Delta kWithin

```{r observed}
ctrl_idx  <- which(true_labels == "CTRL")
inv4m_idx <- which(true_labels == "Inv4m")

obs_deg <- compute_module_delta_k(
  datExpr_all[ctrl_idx, ], datExpr_all[inv4m_idx, ],
  SOFT_POWER, module_colors
)
observed_delta     <- obs_deg$kWithin
observed_delta_out <- obs_deg$kOut

obs_ctrl <- compute_module_delta_k(
  datExpr_control[ctrl_idx, ], datExpr_control[inv4m_idx, ],
  SOFT_POWER, control_module_colors
)
observed_control     <- obs_ctrl$kWithin
observed_control_out <- obs_ctrl$kOut

cat("Observed mean delta kWithin per DEG module:\n")
print(round(sort(observed_delta), 4))
cat("\nObserved mean delta kOut per DEG module:\n")
print(round(sort(observed_delta_out), 4))

cat("\nObserved mean delta kWithin per control module:\n")
print(round(sort(observed_control), 4))
cat("\nObserved mean delta kOut per control module:\n")
print(round(sort(observed_control_out), 4))
```

# Module Label Permutation: Is the Fine Structure Real?

Before testing connectivity *change*, we first verify that the modules
capture real coexpression structure in the CTRL network. We keep the
adjacency matrix fixed and shuffle module labels 1000 times. If observed
mean $k_{Within}$ exceeds the shuffled null, the modules are more
internally connected than random gene groupings of the same sizes.

```{r module_label_permutation, cache=TRUE}
N_LABEL_PERM <- 1000

# --- DEG modules in CTRL network ---
adj_ctrl_deg_full <- adjacency(
  datExpr_all[ctrl_idx, ],
  power = SOFT_POWER, type = "signed"
)

# Observed kWithin per DEG module
k_obs_deg <- intramodularConnectivity(
  adj_ctrl_deg_full, module_colors
)
obs_kw_deg <- sapply(
  unique(module_colors[module_colors != "grey" &
                        !is.na(module_colors)]),
  function(m) {
    mean(k_obs_deg$kWithin[module_colors == m],
         na.rm = TRUE)
  }
)

# Null: shuffle module labels, keep sizes
null_kw_deg <- matrix(NA, nrow = N_LABEL_PERM,
                       ncol = length(obs_kw_deg))
colnames(null_kw_deg) <- names(obs_kw_deg)

for (i in seq_len(N_LABEL_PERM)) {
  shuffled <- sample(module_colors)
  names(shuffled) <- names(module_colors)
  k_shuf <- intramodularConnectivity(
    adj_ctrl_deg_full, shuffled
  )
  null_kw_deg[i, ] <- sapply(names(obs_kw_deg), function(m) {
    mean(k_shuf$kWithin[shuffled == m], na.rm = TRUE)
  })
}

# P-values: fraction of null >= observed
pvals_label_deg <- sapply(names(obs_kw_deg), function(m) {
  mean(null_kw_deg[, m] >= obs_kw_deg[m])
})

cat("=== DEG MODULE STRUCTURE TEST ===\n")
cat("Observed kWithin vs label-shuffled null (CTRL network)\n\n")
deg_structure <- tibble(
  module = names(obs_kw_deg),
  n_genes = sapply(names(obs_kw_deg),
    function(m) sum(module_colors == m, na.rm = TRUE)),
  observed_kWithin = round(obs_kw_deg, 3),
  null_mean = round(colMeans(null_kw_deg), 3),
  null_sd = round(apply(null_kw_deg, 2, sd), 3),
  p_value = pvals_label_deg
) |> arrange(p_value)
print(deg_structure)

# --- Control modules in CTRL network ---
adj_ctrl_ctrl_full <- adjacency(
  datExpr_control[ctrl_idx, ],
  power = SOFT_POWER, type = "signed"
)

k_obs_ctrl <- intramodularConnectivity(
  adj_ctrl_ctrl_full, control_module_colors
)
obs_kw_ctrl <- sapply(
  unique(control_module_colors[
    control_module_colors != "grey"]),
  function(m) {
    mean(k_obs_ctrl$kWithin[
      control_module_colors == m], na.rm = TRUE)
  }
)

null_kw_ctrl <- matrix(NA, nrow = N_LABEL_PERM,
                        ncol = length(obs_kw_ctrl))
colnames(null_kw_ctrl) <- names(obs_kw_ctrl)

for (i in seq_len(N_LABEL_PERM)) {
  shuffled <- sample(control_module_colors)
  names(shuffled) <- names(control_module_colors)
  k_shuf <- intramodularConnectivity(
    adj_ctrl_ctrl_full, shuffled
  )
  null_kw_ctrl[i, ] <- sapply(
    names(obs_kw_ctrl), function(m) {
      mean(k_shuf$kWithin[shuffled == m], na.rm = TRUE)
    }
  )
}

pvals_label_ctrl <- sapply(names(obs_kw_ctrl), function(m) {
  mean(null_kw_ctrl[, m] >= obs_kw_ctrl[m])
})

cat("\n=== CONTROL MODULE STRUCTURE TEST ===\n")
cat("Observed kWithin vs label-shuffled null (CTRL network)\n\n")
ctrl_structure <- tibble(
  module = names(obs_kw_ctrl),
  n_genes = sapply(names(obs_kw_ctrl),
    function(m) sum(control_module_colors == m)),
  observed_kWithin = round(obs_kw_ctrl, 3),
  null_mean = round(colMeans(null_kw_ctrl), 3),
  null_sd = round(apply(null_kw_ctrl, 2, sd), 3),
  p_value = pvals_label_ctrl
) |> arrange(p_value)
print(ctrl_structure)

cat("\n=== SUMMARY ===\n")
cat("DEG modules with p < 0.05:",
    sum(pvals_label_deg < 0.05), "of",
    length(pvals_label_deg), "\n")
cat("Control modules with p < 0.05:",
    sum(pvals_label_ctrl < 0.05), "of",
    length(pvals_label_ctrl), "\n")
```

# Genotype Label Permutation Test

Shuffle genotype labels 1000 times, keeping group sizes fixed (20 CTRL,
23 Inv4m). For each permutation, recompute $\Delta k_{Within}$ for both
DEG and control modules. The module definitions remain fixed throughout.

```{r permutation_loop, cache=TRUE}
null_deltas      <- matrix(NA, nrow = N_PERM, ncol = length(observed_delta))
null_deltas_out  <- matrix(NA, nrow = N_PERM, ncol = length(observed_delta_out))
colnames(null_deltas)     <- names(observed_delta)
colnames(null_deltas_out) <- names(observed_delta_out)

null_control     <- matrix(NA, nrow = N_PERM, ncol = length(observed_control))
null_control_out <- matrix(NA, nrow = N_PERM, ncol = length(observed_control_out))
colnames(null_control)     <- names(observed_control)
colnames(null_control_out) <- names(observed_control_out)

for (i in seq_len(N_PERM)) {
  perm_idx <- sample(n_total)
  pseudo_ctrl  <- perm_idx[1:n_ctrl]
  pseudo_inv4m <- perm_idx[(n_ctrl + 1):n_total]

  perm_deg <- compute_module_delta_k(
    datExpr_all[pseudo_ctrl, ], datExpr_all[pseudo_inv4m, ],
    SOFT_POWER, module_colors
  )
  null_deltas[i, ]     <- perm_deg$kWithin
  null_deltas_out[i, ] <- perm_deg$kOut

  perm_ctrl <- compute_module_delta_k(
    datExpr_control[pseudo_ctrl, ], datExpr_control[pseudo_inv4m, ],
    SOFT_POWER, control_module_colors
  )
  null_control[i, ]     <- perm_ctrl$kWithin
  null_control_out[i, ] <- perm_ctrl$kOut

}
```

# Results

## ECDF P-values

One-tailed test: fraction of permuted values $\leq$ observed (testing for
connectivity loss more extreme than the null).

```{r ecdf_pvalues}
# kWithin
ecdf_pvals <- sapply(names(observed_delta), function(m) {
  mean(null_deltas[, m] <= observed_delta[m])
})
ecdf_pvals_control <- sapply(names(observed_control), function(m) {
  mean(null_control[, m] <= observed_control[m])
})

# kOut (one-tailed: test for gain in inter-modular connectivity)
ecdf_pvals_out <- sapply(names(observed_delta_out), function(m) {
  mean(null_deltas_out[, m] >= observed_delta_out[m])
})
ecdf_pvals_control_out <- sapply(
  names(observed_control_out), function(m) {
    mean(null_control_out[, m] >= observed_control_out[m])
  }
)
```

### DEG Modules

```{r results_deg}
results_deg <- tibble(
  module = names(observed_delta),
  type = "DEG",
  n_genes = sapply(names(observed_delta), function(m) sum(module_colors == m, na.rm = TRUE)),
  observed_mean_delta_k = round(observed_delta, 4),
  null_mean = round(colMeans(null_deltas), 4),
  null_sd = round(apply(null_deltas, 2, sd), 4),
  ecdf_pvalue = round(ecdf_pvals, 4)
) |>
  arrange(ecdf_pvalue)

results_deg
```

### Control Modules

```{r results_control}
results_ctrl <- tibble(
  module = names(observed_control),
  type = "control",
  n_genes = sapply(names(observed_control), function(m) sum(control_module_colors == m)),
  observed_mean_delta_k = round(observed_control, 4),
  null_mean = round(colMeans(null_control), 4),
  null_sd = round(apply(null_control, 2, sd), 4),
  ecdf_pvalue = round(ecdf_pvals_control, 4)
) |>
  arrange(ecdf_pvalue)

results_ctrl
```

### BH FDR Correction Across All Tests

```{r fdr_correction}
all_results <- bind_rows(results_deg, results_ctrl)

all_results <- all_results |>
  mutate(
    fdr = p.adjust(ecdf_pvalue, method = "BH"),
    sig = ifelse(fdr < 0.05, "*", "")
  ) |>
  arrange(ecdf_pvalue)

cat("Significant at FDR < 0.05:\n")
all_results |>
  filter(fdr < 0.05) |>
  dplyr::select(module, type, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()

cat("\nAll control modules:\n")
all_results |>
  filter(type == "control") |>
  dplyr::select(module, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()
```

## Save Results

```{r save_results}
write_csv(all_results,
          file.path(OUTPUT_DIR,
                    "permutation_delta_kWithin_results.csv"))
saveRDS(
  list(
    observed = observed_delta,
    observed_control = observed_control,
    observed_out = observed_delta_out,
    observed_control_out = observed_control_out,
    null = null_deltas,
    null_control = null_control,
    null_out = null_deltas_out,
    null_control_out = null_control_out,
    pvals = ecdf_pvals,
    pvals_control = ecdf_pvals_control,
    pvals_out = ecdf_pvals_out,
    pvals_control_out = ecdf_pvals_control_out,
    control_genes = control_genes,
    control_module_colors = control_module_colors,
    photo_genes = photo_control,
    sag_genes = sag_control
  ),
  file.path(OUTPUT_DIR, "permutation_delta_kWithin.rds")
)
```

# Visualization

## Permutation Null Distributions

Each panel shows the null distribution of mean $\Delta k_{Within}$ from
1000 genotype label permutations. The red line marks the observed value.
Grey panels = DEG modules; blue panels = control modules.

```{r null_distributions, fig.width=16, fig.height=14}
# DEG modules
plot_deg <- map_dfr(names(observed_delta), function(m) {
  tibble(module = m, type = "DEG", null_value = null_deltas[, m])
})
# Control modules (prefix to avoid color name collisions)
plot_ctrl <- map_dfr(names(observed_control), function(m) {
  tibble(module = paste0("ctrl_", m), type = "control", null_value = null_control[, m])
})
plot_data <- bind_rows(plot_deg, plot_ctrl)

obs_data <- tibble(
  module = c(names(observed_delta), paste0("ctrl_", names(observed_control))),
  type = c(rep("DEG", length(observed_delta)), rep("control", length(observed_control))),
  observed = c(observed_delta, observed_control),
  ecdf_pvalue = c(ecdf_pvals, ecdf_pvals_control)
)

module_levels <- c(sort(names(observed_delta)), sort(paste0("ctrl_", names(observed_control))))
plot_data$module <- factor(plot_data$module, levels = module_levels)
obs_data$module  <- factor(obs_data$module, levels = module_levels)

p_null <- ggplot(plot_data, aes(x = null_value)) +
  geom_histogram(
    aes(y = after_stat(density), fill = type),
    bins = 40, color = "grey50", alpha = 0.7
  ) +
  geom_vline(data = obs_data, aes(xintercept = observed),
             color = "red", linewidth = 1) +
  geom_text(
    data = obs_data,
    aes(x = observed, y = Inf, label = sprintf("p = %.3f", ecdf_pvalue)),
    vjust = 1.5, hjust = -0.1, color = "red", size = 3
  ) +
  scale_fill_manual(values = c(DEG = "grey70", control = "steelblue")) +
  facet_wrap(~module, scales = "free_x", ncol = 4) +
  theme_minimal(base_size = 12) +
  labs(
    x = expression(Mean ~ Delta ~ k[Within] ~ "(permuted)"),
    y = "Density",
    title = "Genotype label permutation null for connectivity loss",
    subtitle = sprintf(
      "Red = observed; %d permutations. Grey = DEG modules, blue = control (non-DEG, Ojeda 2026)",
      N_PERM
    ),
    fill = "Gene set"
  )

print(p_null)

ggsave(file.path(OUTPUT_DIR, "permutation_null_delta_kWithin.pdf"),
       p_null, width = 16, height = 14)
ggsave(file.path(OUTPUT_DIR, "permutation_null_delta_kWithin.png"),
       p_null, width = 16, height = 14, dpi = 300)
```

## |logFC| vs Delta kWithin

Scatterplot of Inv4m expression effect size against connectivity change,
for DEG and control genes separately.

```{r logfc_vs_delta_k, fig.width=14, fig.height=6}
library(patchwork)

# Load DEG effects
effects <- read_csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  show_col_types = FALSE
)
inv4m_effects <- effects |>
  filter(grepl("Inv4m", predictor), !grepl(":", predictor)) |>
  dplyr::select(gene, logFC, adj.P.Val)

# Gene-level delta kWithin for DEGs
adj_ctrl_deg  <- adjacency(datExpr_all[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_deg <- adjacency(datExpr_all[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_deg  <- intramodularConnectivity(adj_ctrl_deg, module_colors)
k_inv4m_deg <- intramodularConnectivity(adj_inv4m_deg, module_colors)

deg_dk <- tibble(
  gene = rownames(k_ctrl_deg),
  delta_kWithin = k_inv4m_deg$kWithin - k_ctrl_deg$kWithin,
  group = "DEG"
)

# Gene-level delta kWithin for control genes
adj_ctrl_c  <- adjacency(datExpr_control[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_c <- adjacency(datExpr_control[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_c  <- intramodularConnectivity(adj_ctrl_c, control_module_colors)
k_inv4m_c <- intramodularConnectivity(adj_inv4m_c, control_module_colors)

ctrl_dk <- tibble(
  gene = names(control_module_colors),
  delta_kWithin = k_inv4m_c$kWithin - k_ctrl_c$kWithin,
  group = "Topology control"
)

# Merge with logFC and p-value
scatter_df <- bind_rows(deg_dk, ctrl_dk) |>
  left_join(inv4m_effects |> dplyr::select(gene, logFC, adj.P.Val), by = "gene") |>
  mutate(
    abs_logFC = abs(logFC),
    neg_log10p = -log10(adj.P.Val)
  ) |>
  filter(!is.na(logFC))

# Correlations per group
cor_logfc <- scatter_df |>
  group_by(group) |>
  summarise(
    rho = cor(abs_logFC, delta_kWithin, method = "spearman"),
    .groups = "drop"
  )
cor_pval <- scatter_df |>
  group_by(group) |>
  summarise(
    rho = cor(neg_log10p, delta_kWithin, method = "spearman"),
    .groups = "drop"
  )
cat("Spearman correlations (|logFC| vs delta_k):\n")
print(cor_logfc)
cat("\nSpearman correlations (-log10P vs delta_k):\n")
print(cor_pval)

# Panel A: |logFC| vs delta kWithin
p_logfc <- ggplot(scatter_df, aes(x = abs_logFC, y = delta_kWithin, color = group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("DEG" = "grey40", "Topology control" = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("|" * log[2] * "FC" * "| (Inv4m vs CTRL)"),
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)"),
    color = NULL
  ) +
  theme(legend.position = c(0.2, 0.15))

# Panel B: -log10P vs delta kWithin
p_pval <- ggplot(scatter_df, aes(x = neg_log10p, y = delta_kWithin, color = group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("DEG" = "grey40", "Topology control" = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("-" * log[10] * "(FDR)"),
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)"),
    color = NULL
  ) +
  theme(legend.position = "none")

p_combined <- p_logfc + p_pval + plot_annotation(tag_levels = "A")
print(p_combined)

ggsave(file.path(OUTPUT_DIR, "logFC_pval_vs_delta_kWithin.pdf"),
       p_combined, width = 14, height = 6)
ggsave(file.path(OUTPUT_DIR, "logFC_pval_vs_delta_kWithin.png"),
       p_combined, width = 14, height = 6, dpi = 300)
```

### Linear model: does expression effect size predict connectivity loss?

If connectivity loss were an artifact of differential expression, |logFC| should
be a strong predictor of $\Delta k_{Within}$. We fit gene-level linear models
for DEG and control genes separately, and a combined model with an interaction
term to test whether the relationship differs between groups.

```{r lm_logfc_delta_k}
# DEG-only model
deg_only <- scatter_df |> filter(group == "DEG")
fit_deg <- lm(delta_kWithin ~ abs_logFC, data = deg_only)

cat("=== DEG genes: lm(delta_kWithin ~ |logFC|) ===\n")
summary(fit_deg)
cat(sprintf("\nR² = %.4f  (|logFC| explains %.1f%% of variance in delta_k)\n",
            summary(fit_deg)$r.squared,
            summary(fit_deg)$r.squared * 100))

# Control-only model
ctrl_only <- scatter_df |> filter(group == "Topology control")
fit_ctrl <- lm(delta_kWithin ~ abs_logFC, data = ctrl_only)

cat("\n=== Topology control genes: lm(delta_kWithin ~ |logFC|) ===\n")
summary(fit_ctrl)
cat(sprintf("\nR² = %.4f  (|logFC| explains %.1f%% of variance in delta_k)\n",
            summary(fit_ctrl)$r.squared,
            summary(fit_ctrl)$r.squared * 100))

# Combined model with interaction
fit_interaction <- lm(delta_kWithin ~ abs_logFC * group, data = scatter_df)

cat("\n=== Combined: lm(delta_kWithin ~ |logFC| * group) ===\n")
summary(fit_interaction)
cat(sprintf("\nR² = %.4f\n", summary(fit_interaction)$r.squared))
```

## Effect Size Distribution: DEG vs Control

```{r effect_size_boxplots, fig.width=10, fig.height=5}
# Panel A: |logFC|
p_box_logfc <- ggplot(scatter_df, aes(x = group, y = abs_logFC, fill = group)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.3) +
  scale_fill_manual(values = c("Topology control" = "steelblue", "DEG" = "grey60")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(x = NULL, y = expression("|" * log[2] * "FC" * "|"))

# Panel B: -log10(FDR)
p_box_pval <- ggplot(scatter_df, aes(x = group, y = neg_log10p, fill = group)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.3) +
  scale_fill_manual(values = c("Topology control" = "steelblue", "DEG" = "grey60")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(x = NULL, y = expression("-" * log[10] * "(FDR)"))

p_box_combined <- p_box_logfc + p_box_pval + plot_annotation(tag_levels = "A")
print(p_box_combined)

wt_logfc <- wilcox.test(abs_logFC ~ group, data = scatter_df)
wt_pval  <- wilcox.test(neg_log10p ~ group, data = scatter_df)
cat("|logFC| Wilcoxon p:", format.pval(wt_logfc$p.value), "\n")
cat("-log10(FDR) Wilcoxon p:", format.pval(wt_pval$p.value), "\n")

ggsave(file.path(OUTPUT_DIR, "effect_size_DEG_vs_control.pdf"),
       p_box_combined, width = 10, height = 5)
ggsave(file.path(OUTPUT_DIR, "effect_size_DEG_vs_control.png"),
       p_box_combined, width = 10, height = 5, dpi = 300)

ggsave(file.path(OUTPUT_DIR, "logFC_DEG_vs_control.pdf"),
       p_logfc, width = 5, height = 5)
ggsave(file.path(OUTPUT_DIR, "logFC_DEG_vs_control.png"),
       p_logfc, width = 5, height = 5, dpi = 300)
```

## Z-score Density Plot (Main Figure Candidate)

Two-panel figure comparing connectivity disruption in DEG modules versus
topology control modules. Z-scores are computed per module as
$(observed - mean_{null}) / sd_{null}$. Red lines mark observed values.
GO biological process annotations (from module annotation step 07) are
shown on the left axis for DEG modules.

```{r zscore_violins, fig.width=10, fig.height=10}
# Load GO annotations for DEG modules
go_bp_file <- file.path(RUN_DIR, "07_module_annotation", "top_BP_per_module_rrvgo.csv")
if (file.exists(go_bp_file)) {
  go_bp <- read_csv(go_bp_file, show_col_types = FALSE)
  deg_go_terms <- setNames(
    str_trunc(go_bp$Description, 35),
    go_bp$module_color
  )
} else {
  deg_go_terms <- NULL
}

# Z-score the null distributions
deg_null_z <- perm_null_data <- null_deltas
deg_null_mean <- colMeans(null_deltas)
deg_null_sd <- apply(null_deltas, 2, sd)
for (m in colnames(deg_null_z)) {
  deg_null_z[, m] <- (deg_null_z[, m] - deg_null_mean[m]) / deg_null_sd[m]
}
deg_obs_z <- (observed_delta - deg_null_mean) / deg_null_sd

ctrl_null_z <- null_control
ctrl_null_mean <- colMeans(null_control)
ctrl_null_sd <- apply(null_control, 2, sd)
for (m in colnames(ctrl_null_z)) {
  ctrl_null_z[, m] <- (ctrl_null_z[, m] - ctrl_null_mean[m]) / ctrl_null_sd[m]
}
ctrl_obs_z <- (observed_control - ctrl_null_mean) / ctrl_null_sd

# DEG violin data
deg_violin <- map_dfr(colnames(deg_null_z), function(m) {
  tibble(module = m, z_value = deg_null_z[, m])
})
# Compute FDR across all kWithin modules (DEG + control)
# Use positional indexing to avoid duplicate name collisions
n_deg <- length(ecdf_pvals)
n_ctrl <- length(ecdf_pvals_control)
all_kw_pvals <- c(ecdf_pvals, ecdf_pvals_control)
all_kw_fdr <- p.adjust(all_kw_pvals, method = "BH")
kw_fdr_deg  <- setNames(all_kw_fdr[1:n_deg], names(ecdf_pvals))
kw_fdr_ctrl <- setNames(all_kw_fdr[(n_deg + 1):(n_deg + n_ctrl)], names(ecdf_pvals_control))

deg_points <- tibble(
  module = names(deg_obs_z),
  z_observed = deg_obs_z,
  fdr = kw_fdr_deg[names(deg_obs_z)]
)
deg_order <- deg_points |> arrange(z_observed) |> pull(module)
deg_violin$module <- factor(deg_violin$module, levels = deg_order)
deg_points$module <- factor(deg_points$module, levels = deg_order)

# Control violin data
ctrl_violin <- map_dfr(colnames(ctrl_null_z), function(m) {
  tibble(module = m, z_value = ctrl_null_z[, m])
})
ctrl_points <- tibble(
  module = names(ctrl_obs_z),
  z_observed = ctrl_obs_z,
  fdr = kw_fdr_ctrl[names(ctrl_obs_z)]
)
ctrl_order <- ctrl_points |> arrange(z_observed) |> pull(module)
ctrl_violin$module <- factor(ctrl_violin$module, levels = ctrl_order)
ctrl_points$module <- factor(ctrl_points$module, levels = ctrl_order)

# Module sizes for y-axis labels
deg_mod_sizes <- sapply(levels(deg_violin$module), function(m) {
  sum(module_colors == m, na.rm = TRUE)
})
ctrl_mod_sizes <- sapply(levels(ctrl_violin$module), function(m) {
  sum(control_module_colors == m)
})

p_deg_density <- make_density_panel(deg_violin, deg_points,
                                    "DEG coexpression modules (kWithin)",
                                    fill_color = "identity",
                                    go_terms = deg_go_terms,
                                    module_sizes = deg_mod_sizes)
# GO terms for control modules from enrichment
ctrl_go_annot <- setNames(
  str_trunc(ctrl_top_bp$Description, 35),
  ctrl_top_bp$module_color
)

p_ctrl_density <- make_density_panel(ctrl_violin, ctrl_points,
                                     "Topology control (kWithin)",
                                     fill_color = ctrl_color_hex,
                                     go_terms = ctrl_go_annot,
                                     module_sizes = ctrl_mod_sizes)

p_density_combined <- p_deg_density / p_ctrl_density +
  plot_annotation(tag_levels = "A")
print(p_density_combined)

ggsave(file.path(OUTPUT_DIR, "zscore_density_DEG_vs_control.pdf"),
       p_density_combined, width = 10, height = 10)
ggsave(file.path(OUTPUT_DIR, "zscore_density_DEG_vs_control.png"),
       p_density_combined, width = 10, height = 10, dpi = 300)
```

## Z-score Density Plot: Inter-modular Connectivity (kOut)

Same approach as the kWithin plot but for extra-modular connectivity
($\Delta k_{Out}$). Positive values indicate genes gaining connections
to genes outside their module in Inv4m.

```{r zscore_kout, fig.width=10, fig.height=10}
# Z-score kOut null distributions
deg_null_z_out <- null_deltas_out
deg_null_mean_out <- colMeans(null_deltas_out)
deg_null_sd_out <- apply(null_deltas_out, 2, sd)
for (m in colnames(deg_null_z_out)) {
  deg_null_z_out[, m] <- (deg_null_z_out[, m] -
    deg_null_mean_out[m]) / deg_null_sd_out[m]
}
deg_obs_z_out <- (observed_delta_out -
  deg_null_mean_out) / deg_null_sd_out

ctrl_null_z_out <- null_control_out
ctrl_null_mean_out <- colMeans(null_control_out)
ctrl_null_sd_out <- apply(null_control_out, 2, sd)
for (m in colnames(ctrl_null_z_out)) {
  ctrl_null_z_out[, m] <- (ctrl_null_z_out[, m] -
    ctrl_null_mean_out[m]) / ctrl_null_sd_out[m]
}
ctrl_obs_z_out <- (observed_control_out -
  ctrl_null_mean_out) / ctrl_null_sd_out

# FDR across all kOut tests
n_deg_out <- length(ecdf_pvals_out)
n_ctrl_out <- length(ecdf_pvals_control_out)
all_kout_pvals <- c(ecdf_pvals_out, ecdf_pvals_control_out)
all_kout_fdr <- p.adjust(all_kout_pvals, method = "BH")
kout_fdr_deg <- setNames(
  all_kout_fdr[1:n_deg_out],
  names(ecdf_pvals_out)
)
kout_fdr_ctrl <- setNames(
  all_kout_fdr[(n_deg_out + 1):(n_deg_out + n_ctrl_out)],
  names(ecdf_pvals_control_out)
)

# DEG kOut density data
deg_violin_out <- map_dfr(colnames(deg_null_z_out), function(m) {
  tibble(module = m, z_value = deg_null_z_out[, m])
})
deg_points_out <- tibble(
  module = names(deg_obs_z_out),
  z_observed = deg_obs_z_out,
  fdr = kout_fdr_deg[names(deg_obs_z_out)]
)
deg_order_out <- deg_points_out |>
  arrange(z_observed) |>
  pull(module)
deg_violin_out$module <- factor(
  deg_violin_out$module, levels = deg_order_out
)
deg_points_out$module <- factor(
  deg_points_out$module, levels = deg_order_out
)

# Control kOut density data
ctrl_violin_out <- map_dfr(
  colnames(ctrl_null_z_out), function(m) {
    tibble(module = m, z_value = ctrl_null_z_out[, m])
  }
)
ctrl_points_out <- tibble(
  module = names(ctrl_obs_z_out),
  z_observed = ctrl_obs_z_out,
  fdr = kout_fdr_ctrl[names(ctrl_obs_z_out)]
)
ctrl_order_out <- ctrl_points_out |>
  arrange(z_observed) |>
  pull(module)
ctrl_violin_out$module <- factor(
  ctrl_violin_out$module, levels = ctrl_order_out
)
ctrl_points_out$module <- factor(
  ctrl_points_out$module, levels = ctrl_order_out
)

p_deg_kout <- make_density_panel(
  deg_violin_out, deg_points_out,
  "DEG coexpression modules (kOut)",
  fill_color = "identity",
  go_terms = deg_go_terms,
  module_sizes = deg_mod_sizes
)
p_ctrl_kout <- make_density_panel(
  ctrl_violin_out, ctrl_points_out,
  "Topology control (kOut)",
  fill_color = ctrl_color_hex,
  go_terms = ctrl_go_annot,
  module_sizes = ctrl_mod_sizes
)

p_kout_combined <- p_deg_kout / p_ctrl_kout +
  plot_annotation(tag_levels = "A")
print(p_kout_combined)

ggsave(file.path(OUTPUT_DIR,
                 "zscore_density_kOut_DEG_vs_control.pdf"),
       p_kout_combined, width = 10, height = 10)
ggsave(file.path(OUTPUT_DIR,
                 "zscore_density_kOut_DEG_vs_control.png"),
       p_kout_combined, width = 10, height = 10, dpi = 300)
```

## kOut: CTRL vs Inv4m Gene-level Scatter

Analogous to Figure 4C (kWithin hub scatter) but for inter-modular
connectivity. Each point is a gene; axes are kOut in CTRL vs kOut in
Inv4m. Points above the diagonal gained inter-modular connectivity.

```{r kout_scatter, fig.width=8, fig.height=8}
# Gene-level kOut (already computed in logfc_vs_delta_k chunk)
kout_scatter <- tibble(
  gene = rownames(k_ctrl_deg),
  module = module_colors[rownames(k_ctrl_deg)],
  kOut_CTRL = k_ctrl_deg$kOut,
  kOut_Inv4m = k_inv4m_deg$kOut
) |>
  filter(module != "grey", !is.na(module))

p_kout_scatter <- ggplot(
  kout_scatter,
  aes(x = kOut_CTRL, y = kOut_Inv4m, color = module)
) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50") +
  scale_color_identity() +
  theme_minimal(base_size = 14) +
  labs(
    x = expression(k[Out] ~ "(CTRL)"),
    y = expression(k[Out] ~ "(Inv4m)"),
    title = "Inter-modular connectivity: CTRL vs Inv4m"
  ) +
  coord_equal()

print(p_kout_scatter)

# Summary stats
cat("Genes above diagonal (kOut gain):",
    sum(kout_scatter$kOut_Inv4m > kout_scatter$kOut_CTRL),
    "of", nrow(kout_scatter), "\n")
cat("Paired Wilcoxon (kOut Inv4m vs CTRL):\n")
wt_kout <- wilcox.test(kout_scatter$kOut_Inv4m,
                        kout_scatter$kOut_CTRL,
                        paired = TRUE)
print(wt_kout)

ggsave(file.path(OUTPUT_DIR, "kOut_CTRL_vs_Inv4m_scatter.pdf"),
       p_kout_scatter, width = 8, height = 8)
ggsave(file.path(OUTPUT_DIR, "kOut_CTRL_vs_Inv4m_scatter.png"),
       p_kout_scatter, width = 8, height = 8, dpi = 300)
```

# Module Preservation (Langfelder 2011) and Combined Stats Table

Run `modulePreservation()` for both gene sets (CTRL → Inv4m) to get
Z_summary, Z_density, and Z_connectivity. Then compile a single table
with all statistics for the supplementary material.

```{r module_preservation_both, cache=TRUE}
# --- DEG modules: load existing results from 06_preservation ---
# These were computed with power = 18 (the correct value)
mp_deg_file <- file.path(OUTPUT_DIR,
                          "modulePreservation_results.rds")
mp_deg <- readRDS(mp_deg_file)

z_deg <- mp_deg$preservation$Z$ref.CTRL$inColumnsAlsoPresentIn.Inv4m
deg_pres <- tibble(
  module = rownames(z_deg),
  Zsummary = round(z_deg[, "Zsummary.pres"], 2),
  Zdensity = round(z_deg[, "Zdensity.pres"], 2),
  Zconnectivity = round(z_deg[, "Zconnectivity.pres"], 2)
) |> filter(module != "gold", module != "grey")

cat("=== DEG MODULE PRESERVATION (from 06_preservation) ===\n")
print(deg_pres |> arrange(Zsummary))

# --- Control modules: run modulePreservation ---
# NOTE: modulePreservation uses Pearson correlation internally.
# The joint analysis (TODO) will use power = 18 for both sets.
multiExpr_ctrl <- list(
  CTRL = list(data = datExpr_control[ctrl_idx, ]),
  Inv4m = list(data = datExpr_control[inv4m_idx, ])
)
multiColor_ctrl <- list(CTRL = control_module_colors)

mp_ctrl <- modulePreservation(
  multiData = multiExpr_ctrl,
  multiColor = multiColor_ctrl,
  referenceNetworks = 1,
  testNetworks = 2,
  nPermutations = 200,
  networkType = "signed",
  verbose = 0
)

z_ctrl <- mp_ctrl$preservation$Z$ref.CTRL$inColumnsAlsoPresentIn.Inv4m
ctrl_pres <- tibble(
  module = rownames(z_ctrl),
  Zsummary = round(z_ctrl[, "Zsummary.pres"], 2),
  Zdensity = round(z_ctrl[, "Zdensity.pres"], 2),
  Zconnectivity = round(z_ctrl[, "Zconnectivity.pres"], 2)
) |> filter(module != "gold", module != "grey")

cat("\n=== CONTROL MODULE PRESERVATION ===\n")
print(ctrl_pres |> arrange(Zsummary))
```

## Combined Statistics Table

```{r combined_table}
# Load DEG preservation summary for LFP values
preservation_summary <- read_csv(
  file.path(OUTPUT_DIR, "preservation_summary.csv"),
  show_col_types = FALSE
)

# DEG table
deg_table <- deg_pres |>
  left_join(
    tibble(
      module = names(observed_delta),
      delta_kWithin = round(observed_delta, 3),
      kWithin_FDR = round(kw_fdr_deg[names(observed_delta)], 4),
      delta_kOut = round(observed_delta_out, 3),
      kOut_FDR = round(kout_fdr_deg[names(observed_delta_out)], 4)
    ),
    by = "module"
  ) |>
  mutate(
    gene_set = "DEG",
    n_genes = sapply(module, function(m)
      sum(module_colors == m, na.rm = TRUE)),
    LFP = round(sapply(module, function(m) {
      pres_row <- preservation_summary[
        preservation_summary$color == m, ]
      if (nrow(pres_row) > 0) pres_row$ctrl_lfp[1]
      else NA_real_
    }), 3)
  )

# Control table
ctrl_table <- ctrl_pres |>
  left_join(
    tibble(
      module = names(observed_control),
      delta_kWithin = round(observed_control, 3),
      kWithin_FDR = round(
        kw_fdr_ctrl[names(observed_control)], 4),
      delta_kOut = round(observed_control_out, 3),
      kOut_FDR = round(
        kout_fdr_ctrl[names(observed_control_out)], 4)
    ),
    by = "module"
  ) |>
  mutate(
    gene_set = "Control",
    n_genes = sapply(module, function(m)
      sum(control_module_colors == m)),
    LFP = round(ctrl_lfp$mean_lfp[
      match(module, ctrl_lfp$color)], 3)
  )

# Combine
combined_table <- bind_rows(deg_table, ctrl_table) |>
  dplyr::select(
    gene_set, module, n_genes, LFP,
    Zsummary, Zdensity, Zconnectivity,
    delta_kWithin, kWithin_FDR,
    delta_kOut, kOut_FDR
  ) |>
  arrange(gene_set, kWithin_FDR)

cat("=== COMBINED MODULE STATISTICS ===\n\n")
print(combined_table, n = 30, width = Inf)

# Save
write_csv(
  combined_table,
  file.path(OUTPUT_DIR,
            "combined_module_stats_table.csv")
)
cat("\nSaved: combined_module_stats_table.csv\n")
```

# Summary

```{r summary}
n_deg_sig <- sum(all_results$type == "DEG" & all_results$fdr < 0.05)
n_ctrl_sig <- sum(all_results$type == "control" & all_results$fdr < 0.05)

cat("=== PERMUTATION TEST SUMMARY ===\n\n")
cat("DEG modules significant (FDR < 0.05):", n_deg_sig, "of",
    sum(all_results$type == "DEG"), "\n")
cat("Control modules significant (FDR < 0.05):", n_ctrl_sig, "of",
    sum(all_results$type == "control"), "\n\n")

cat("Significant DEG modules:\n")
all_results |>
  filter(type == "DEG", fdr < 0.05) |>
  dplyr::select(module, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()

n_deg_sig_out <- sum(kout_fdr_deg < 0.05)
n_ctrl_sig_out <- sum(kout_fdr_ctrl < 0.05)

cat("
=== kOut (inter-modular) ===
DEG modules with significant kOut gain (FDR < 0.05):",
    n_deg_sig_out, "of", length(kout_fdr_deg), "
Control modules with significant kOut gain (FDR < 0.05):",
    n_ctrl_sig_out, "of", length(kout_fdr_ctrl), "

=== CONCLUSIONS ===

1. kWithin: DEG modules show significant intramodular
   connectivity loss beyond the permutation null.
   Control modules do not.

2. kOut: No significant gain in inter-modular connectivity
   for DEG or control modules. The perturbation degrades
   within-module coexpression structure without causing
   genes to gain spurious cross-module connections.

3. This asymmetry (kWithin loss without kOut gain) argues
   against a global artifact of mean expression shifts:
   if the connectivity change were driven by correlated
   mean shifts inflating all pairwise differences, both
   kWithin and kOut would be affected. Instead, only
   within-module relationships are disrupted, consistent
   with degradation of specific coexpression programs.

4. The linear model confirms this: |logFC| does not
   predict delta kWithin (p = 0.65, R2 = 0.0004),
   ruling out differential expression magnitude as a
   driver of connectivity loss.
")
```

# Session Info

```{r session_info}
sessionInfo()
```
