---
title: "Permutation Test for WGCNA Connectivity Loss"
author: "F. V. Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = TRUE
)
knitr::opts_knit$set(root.dir = here::here())
```

# Overview

The WGCNA analysis (step 06) found that 12 of 13 DEG coexpression modules
lose intramodular connectivity ($\Delta k_{Within}$) when comparing Inv4m
to CTRL networks. A potential concern is that this connectivity loss could
be an artifact of splitting samples by genotype: genes selected as DEGs
will, by definition, have different mean expression between groups, and
this could inflate correlation differences when networks are built
separately.

This notebook addresses that concern with two complementary approaches:

1. **Genotype label permutation test**: Shuffle genotype labels 1000 times,
   recompute $\Delta k_{Within}$ per module, and compare the observed
   connectivity loss to the empirical null distribution. This directly tests
   whether the observed loss exceeds what random sample splitting produces.

2. **Topology control gene set**: Apply the same analysis to 218
   photosynthesis and senescence genes (Ojeda 2026) that are NOT
   differentially expressed between genotypes but DO have real coexpression
   structure driven by the leaf developmental gradient. If the connectivity
   loss is an artifact of sample splitting, these control genes should show
   the same pattern. If it is specific to DEGs, they should not.

# Setup

```{r load_packages}
library(here)
library(tidyverse)
library(WGCNA)

allowWGCNAThreads()
set.seed(42)
```

```{r config}
N_PERM <- 10000
RUN_DIR <- here("results/inversion_paper/field_perturbation/run_20251231_201332")
OUTPUT_DIR <- file.path(RUN_DIR, "06_preservation")

FILTER_DIR    <- file.path(RUN_DIR, "02_gene_filter")
DATA_PREP_DIR <- file.path(RUN_DIR, "01_data_prep")
CONSENSUS_DIR <- file.path(RUN_DIR, "04_consensus_networks")
REFERENCE_DIR <- file.path(RUN_DIR, "03_reference_network")

source(here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Load Data

```{r load_data}
# DEG-filtered expression (465 genes x 43 samples)
expr_filtered <- readRDS(file.path(FILTER_DIR, "filtered_expression.rds"))
sample_meta   <- read_csv(file.path(DATA_PREP_DIR, "sample_metadata.csv"),
                           show_col_types = FALSE)
ctrl_modules  <- read_csv(file.path(CONSENSUS_DIR, "ctrl", "consensus_modules.csv"),
                           show_col_types = FALSE)
SOFT_POWER    <- readRDS(file.path(REFERENCE_DIR, "ctrl", "selected_power.rds"))

# Full expression matrix (24k genes x 43 samples, PSU2022 only)
full_expr <- readRDS(file.path(DATA_PREP_DIR, "combined_expression.rds"))
if (inherits(full_expr, "EList")) {
  full_expr_mat <- full_expr$E
} else {
  full_expr_mat <- as.matrix(full_expr)
}

cat("DEG expression:", nrow(expr_filtered), "genes x", ncol(expr_filtered), "samples\n")
cat("Full expression:", nrow(full_expr_mat), "genes x", ncol(full_expr_mat), "samples\n")
cat("Soft power:", SOFT_POWER, "\n")
```

# Control Gene Sets

Photosynthesis and senescence hub genes from Ojeda 2026. In the companion
phosphorus paper, we showed that the expression indices for these gene sets
do not differ between Inv4m and CTRL genotypes. They are driven by the leaf
developmental gradient (young leaves = photosynthesis, old leaves =
senescence), providing biologically structured coexpression that is
independent of the Inv4m introgression.

```{r load_control_genes}
photo_genes <- read.table(
  file.path(paths$data, "photosynthesis_expression_index_genes_ojeda2026.txt"),
  header = FALSE, stringsAsFactors = FALSE
)$V1

sag_genes <- read.table(
  file.path(paths$data, "senesence_hubs_ojeda2026.tab"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)$gene

# Exclude any overlap with DEGs
deg_genes <- ctrl_modules$gene
photo_control <- setdiff(intersect(photo_genes, rownames(full_expr_mat)), deg_genes)
sag_control   <- setdiff(intersect(sag_genes, rownames(full_expr_mat)), deg_genes)
control_genes <- unique(c(photo_control, sag_control))

cat("Photosynthesis (non-DEG, in matrix):", length(photo_control), "\n")
cat("Senescence hubs (non-DEG, in matrix):", length(sag_control), "\n")
cat("Combined control genes:", length(control_genes), "\n")
```

# Prepare Expression Matrices

```{r prepare_expression}
# DEG module colors aligned to gene order
module_colors <- setNames(ctrl_modules$color, ctrl_modules$gene)
module_colors <- module_colors[rownames(expr_filtered)]

# Genotype labels
all_samples <- sample_meta$sample_id
true_labels <- setNames(sample_meta$Genotype, sample_meta$sample_id)
n_ctrl  <- sum(true_labels == "CTRL")
n_inv4m <- sum(true_labels == "Inv4m")
n_total <- length(true_labels)

# WGCNA format: samples x genes
datExpr_all     <- t(expr_filtered[, all_samples])
datExpr_control <- t(full_expr_mat[control_genes, all_samples])

cat("DEG matrix:", nrow(datExpr_all), "samples x", ncol(datExpr_all), "genes\n")
cat("Control matrix:", nrow(datExpr_control), "samples x", ncol(datExpr_control), "genes\n")
cat("CTRL:", n_ctrl, "| Inv4m:", n_inv4m, "| Total:", n_total, "\n")
```

# WGCNA Modules for Control Genes

Build modules from control genes using CTRL samples only, mirroring
how the DEG modules were constructed.

```{r control_modules}
datExpr_ctrl_only <- datExpr_control[which(true_labels == "CTRL"), ]
adj_ctrl_control <- adjacency(datExpr_ctrl_only, power = SOFT_POWER, type = "signed")
ctrl_tree <- hclust(as.dist(1 - adj_ctrl_control), method = "average")
ctrl_control_modules <- cutreeDynamic(
  dendro = ctrl_tree,
  distM = 1 - adj_ctrl_control,
  deepSplit = 4,
  minClusterSize = 10
)
control_module_colors_raw <- labels2colors(ctrl_control_modules)

# Remap WGCNA standard colors to non-clashing names with matching hex
ctrl_color_remap <- c(
  turquoise = "sky",
  blue      = "peach",
  brown     = "jade",
  yellow    = "coral",
  green     = "lavender",
  black     = "umber",
  red       = "mustard",
  pink      = "rose",
  magenta   = "aqua",
  purple    = "spring",
  grey      = "grey"
)
# Hex codes for plotting (keyed by new name)
ctrl_color_hex <- c(
  sky      = "#5DA5DA",
  peach    = "#FAA43A",
  jade     = "#60BD68",
  coral    = "#F15854",
  lavender = "#B276B2",
  umber    = "#8D4B08",
  mustard  = "#DECF3F",
  rose     = "#F17CB0",
  aqua     = "#66E3D9",
  spring   = "#00FF7F",
  grey     = "grey"
)

control_module_colors <- ctrl_color_remap[control_module_colors_raw]
names(control_module_colors) <- control_genes

cat("Control modules found:", length(unique(ctrl_control_modules[ctrl_control_modules != 0])), "\n")
cat("Unassigned (grey):", sum(ctrl_control_modules == 0), "of", length(ctrl_control_modules), "\n\n")

tibble(module = control_module_colors) |>
  filter(module != "grey") |>
  count(module, sort = TRUE) |>
  print()
```

## Module Size Comparison

```{r module_size_comparison, fig.width=8, fig.height=4}
deg_sizes <- tibble(
  module = module_colors,
  type = "DEG"
) |>
  filter(module != "grey", !is.na(module)) |>
  count(module, type, name = "n_genes")

ctrl_sizes <- tibble(
  module = control_module_colors,
  type = "Control"
) |>
  filter(module != "grey") |>
  count(module, type, name = "n_genes")

bind_rows(deg_sizes, ctrl_sizes) |>
  ggplot(aes(x = n_genes, fill = type)) +
  geom_histogram(bins = 15, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c(DEG = "grey60", Control = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(x = "Module size (genes)", y = "Count", fill = NULL,
       title = "Module size distributions overlap between DEG and control")
```

# Helper Function

```{r helper_function}
compute_module_delta_k <- function(datExpr_a, datExpr_b, power, mod_colors) {
  adj_a <- adjacency(datExpr_a, power = power, type = "signed")
  adj_b <- adjacency(datExpr_b, power = power, type = "signed")

  k_a <- intramodularConnectivity(adj_a, mod_colors)
  k_b <- intramodularConnectivity(adj_b, mod_colors)

  delta_k <- k_b$kWithin - k_a$kWithin

  module_ids <- unique(mod_colors[mod_colors != "grey" & !is.na(mod_colors)])
  sapply(module_ids, function(m) {
    idx <- which(mod_colors == m)
    mean(delta_k[idx], na.rm = TRUE)
  })
}

# Violin plot panel constructor (used by kWithin and kME violin sections)
# points_df must have columns: module, z_observed, fdr
# fill_color: single color string OR named vector (module -> hex)
make_violin_panel <- function(violin_df, points_df, panel_title, fill_color) {
  if (length(fill_color) == 1) {
    p <- ggplot(violin_df, aes(x = module, y = z_value)) +
      geom_violin(fill = fill_color, color = "grey40", alpha = 0.6, scale = "width")
  } else {
    p <- ggplot(violin_df, aes(x = module, y = z_value, fill = module)) +
      geom_violin(color = "grey40", alpha = 0.6, scale = "width") +
      scale_fill_manual(values = fill_color) +
      guides(fill = "none")
  }
  p +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(
      data = points_df,
      aes(x = module, y = z_observed),
      color = "red", size = 3
    ) +
    geom_text(
      data = points_df |> filter(fdr < 0.05),
      aes(x = module, y = z_observed, label = sprintf("FDR=%.3f", fdr)),
      color = "red", size = 3, vjust = -1, hjust = 0.5
    ) +
    coord_flip() +
    theme_minimal(base_size = 14) +
    labs(x = NULL, y = expression(Z[Delta * k]), title = panel_title)
}
```

# Observed Delta kWithin

```{r observed}
ctrl_idx  <- which(true_labels == "CTRL")
inv4m_idx <- which(true_labels == "Inv4m")

observed_delta <- compute_module_delta_k(
  datExpr_all[ctrl_idx, ], datExpr_all[inv4m_idx, ],
  SOFT_POWER, module_colors
)

observed_control <- compute_module_delta_k(
  datExpr_control[ctrl_idx, ], datExpr_control[inv4m_idx, ],
  SOFT_POWER, control_module_colors
)

cat("Observed mean delta kWithin per DEG module:\n")
print(round(sort(observed_delta), 4))

cat("\nObserved mean delta kWithin per control module:\n")
print(round(sort(observed_control), 4))
```

# Permutation Test

Shuffle genotype labels 1000 times, keeping group sizes fixed (20 CTRL,
23 Inv4m). For each permutation, recompute $\Delta k_{Within}$ for both
DEG and control modules. The module definitions remain fixed throughout.

```{r permutation_loop, cache=TRUE}
null_deltas  <- matrix(NA, nrow = N_PERM, ncol = length(observed_delta))
colnames(null_deltas) <- names(observed_delta)
null_control <- matrix(NA, nrow = N_PERM, ncol = length(observed_control))
colnames(null_control) <- names(observed_control)

pb <- txtProgressBar(min = 0, max = N_PERM, style = 3)
for (i in seq_len(N_PERM)) {
  perm_idx <- sample(n_total)
  pseudo_ctrl  <- perm_idx[1:n_ctrl]
  pseudo_inv4m <- perm_idx[(n_ctrl + 1):n_total]

  null_deltas[i, ] <- compute_module_delta_k(
    datExpr_all[pseudo_ctrl, ], datExpr_all[pseudo_inv4m, ],
    SOFT_POWER, module_colors
  )

  null_control[i, ] <- compute_module_delta_k(
    datExpr_control[pseudo_ctrl, ], datExpr_control[pseudo_inv4m, ],
    SOFT_POWER, control_module_colors
  )

  setTxtProgressBar(pb, i)
}
close(pb)
```

# Results

## ECDF P-values

One-tailed test: fraction of permuted values $\leq$ observed (testing for
connectivity loss more extreme than the null).

```{r ecdf_pvalues}
ecdf_pvals <- sapply(names(observed_delta), function(m) {
  mean(null_deltas[, m] <= observed_delta[m])
})
ecdf_pvals_control <- sapply(names(observed_control), function(m) {
  mean(null_control[, m] <= observed_control[m])
})
```

### DEG Modules

```{r results_deg}
results_deg <- tibble(
  module = names(observed_delta),
  type = "DEG",
  n_genes = sapply(names(observed_delta), function(m) sum(module_colors == m, na.rm = TRUE)),
  observed_mean_delta_k = round(observed_delta, 4),
  null_mean = round(colMeans(null_deltas), 4),
  null_sd = round(apply(null_deltas, 2, sd), 4),
  ecdf_pvalue = round(ecdf_pvals, 4)
) |>
  arrange(ecdf_pvalue)

results_deg
```

### Control Modules

```{r results_control}
results_ctrl <- tibble(
  module = names(observed_control),
  type = "control",
  n_genes = sapply(names(observed_control), function(m) sum(control_module_colors == m)),
  observed_mean_delta_k = round(observed_control, 4),
  null_mean = round(colMeans(null_control), 4),
  null_sd = round(apply(null_control, 2, sd), 4),
  ecdf_pvalue = round(ecdf_pvals_control, 4)
) |>
  arrange(ecdf_pvalue)

results_ctrl
```

### BH FDR Correction Across All Tests

```{r fdr_correction}
all_results <- bind_rows(results_deg, results_ctrl)

all_results <- all_results |>
  mutate(
    fdr = p.adjust(ecdf_pvalue, method = "BH"),
    sig = ifelse(fdr < 0.05, "*", "")
  ) |>
  arrange(ecdf_pvalue)

cat("Significant at FDR < 0.05:\n")
all_results |>
  filter(fdr < 0.05) |>
  select(module, type, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()

cat("\nAll control modules:\n")
all_results |>
  filter(type == "control") |>
  select(module, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()
```

## Save Results

```{r save_results}
write_csv(all_results, file.path(OUTPUT_DIR, "permutation_delta_kWithin_results.csv"))
saveRDS(
  list(
    observed = observed_delta,
    observed_control = observed_control,
    null = null_deltas,
    null_control = null_control,
    pvals = ecdf_pvals,
    pvals_control = ecdf_pvals_control,
    control_genes = control_genes,
    control_module_colors = control_module_colors,
    photo_genes = photo_control,
    sag_genes = sag_control
  ),
  file.path(OUTPUT_DIR, "permutation_delta_kWithin.rds")
)
```

# Visualization

## Permutation Null Distributions

Each panel shows the null distribution of mean $\Delta k_{Within}$ from
1000 genotype label permutations. The red line marks the observed value.
Grey panels = DEG modules; blue panels = control modules.

```{r null_distributions, fig.width=16, fig.height=14}
# DEG modules
plot_deg <- map_dfr(names(observed_delta), function(m) {
  tibble(module = m, type = "DEG", null_value = null_deltas[, m])
})
# Control modules (prefix to avoid color name collisions)
plot_ctrl <- map_dfr(names(observed_control), function(m) {
  tibble(module = paste0("ctrl_", m), type = "control", null_value = null_control[, m])
})
plot_data <- bind_rows(plot_deg, plot_ctrl)

obs_data <- tibble(
  module = c(names(observed_delta), paste0("ctrl_", names(observed_control))),
  type = c(rep("DEG", length(observed_delta)), rep("control", length(observed_control))),
  observed = c(observed_delta, observed_control),
  ecdf_pvalue = c(ecdf_pvals, ecdf_pvals_control)
)

module_levels <- c(sort(names(observed_delta)), sort(paste0("ctrl_", names(observed_control))))
plot_data$module <- factor(plot_data$module, levels = module_levels)
obs_data$module  <- factor(obs_data$module, levels = module_levels)

p_null <- ggplot(plot_data, aes(x = null_value)) +
  geom_histogram(
    aes(y = after_stat(density), fill = type),
    bins = 40, color = "grey50", alpha = 0.7
  ) +
  geom_vline(data = obs_data, aes(xintercept = observed),
             color = "red", linewidth = 1) +
  geom_text(
    data = obs_data,
    aes(x = observed, y = Inf, label = sprintf("p = %.3f", ecdf_pvalue)),
    vjust = 1.5, hjust = -0.1, color = "red", size = 3
  ) +
  scale_fill_manual(values = c(DEG = "grey70", control = "steelblue")) +
  facet_wrap(~module, scales = "free_x", ncol = 4) +
  theme_minimal(base_size = 12) +
  labs(
    x = expression(Mean ~ Delta ~ k[Within] ~ "(permuted)"),
    y = "Density",
    title = "Genotype label permutation null for connectivity loss",
    subtitle = sprintf(
      "Red = observed; %d permutations. Grey = DEG modules, blue = control (non-DEG, Ojeda 2026)",
      N_PERM
    ),
    fill = "Gene set"
  )

print(p_null)

ggsave(file.path(OUTPUT_DIR, "permutation_null_delta_kWithin.pdf"),
       p_null, width = 16, height = 14)
ggsave(file.path(OUTPUT_DIR, "permutation_null_delta_kWithin.png"),
       p_null, width = 16, height = 14, dpi = 300)
```

## |logFC| vs Delta kWithin

Scatterplot of Inv4m expression effect size against connectivity change,
for DEG and control genes separately.

```{r logfc_vs_delta_k, fig.width=14, fig.height=6}
library(patchwork)

# Load DEG effects
effects <- read_csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  show_col_types = FALSE
)
inv4m_effects <- effects |>
  filter(grepl("Inv4m", predictor), !grepl(":", predictor)) |>
  select(gene, logFC, adj.P.Val)

# Gene-level delta kWithin for DEGs
adj_ctrl_deg  <- adjacency(datExpr_all[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_deg <- adjacency(datExpr_all[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_deg  <- intramodularConnectivity(adj_ctrl_deg, module_colors)
k_inv4m_deg <- intramodularConnectivity(adj_inv4m_deg, module_colors)

deg_dk <- tibble(
  gene = rownames(k_ctrl_deg),
  delta_kWithin = k_inv4m_deg$kWithin - k_ctrl_deg$kWithin,
  group = "DEG"
)

# Gene-level delta kWithin for control genes
adj_ctrl_c  <- adjacency(datExpr_control[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_c <- adjacency(datExpr_control[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_c  <- intramodularConnectivity(adj_ctrl_c, control_module_colors)
k_inv4m_c <- intramodularConnectivity(adj_inv4m_c, control_module_colors)

ctrl_dk <- tibble(
  gene = names(control_module_colors),
  delta_kWithin = k_inv4m_c$kWithin - k_ctrl_c$kWithin,
  group = "Topology control"
)

# Merge with logFC and p-value
scatter_df <- bind_rows(deg_dk, ctrl_dk) |>
  left_join(inv4m_effects |> select(gene, logFC, adj.P.Val), by = "gene") |>
  mutate(
    abs_logFC = abs(logFC),
    neg_log10p = -log10(adj.P.Val)
  ) |>
  filter(!is.na(logFC))

# Correlations per group
cor_logfc <- scatter_df |>
  group_by(group) |>
  summarise(
    rho = cor(abs_logFC, delta_kWithin, method = "spearman"),
    .groups = "drop"
  )
cor_pval <- scatter_df |>
  group_by(group) |>
  summarise(
    rho = cor(neg_log10p, delta_kWithin, method = "spearman"),
    .groups = "drop"
  )
cat("Spearman correlations (|logFC| vs delta_k):\n")
print(cor_logfc)
cat("\nSpearman correlations (-log10P vs delta_k):\n")
print(cor_pval)

# Panel A: |logFC| vs delta kWithin
p_logfc <- ggplot(scatter_df, aes(x = abs_logFC, y = delta_kWithin, color = group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("DEG" = "grey40", "Topology control" = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("|" * log[2] * "FC" * "| (Inv4m vs CTRL)"),
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)"),
    color = NULL
  ) +
  theme(legend.position = c(0.2, 0.15))

# Panel B: -log10P vs delta kWithin
p_pval <- ggplot(scatter_df, aes(x = neg_log10p, y = delta_kWithin, color = group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("DEG" = "grey40", "Topology control" = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("-" * log[10] * "(FDR)"),
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)"),
    color = NULL
  ) +
  theme(legend.position = "none")

p_combined <- p_logfc + p_pval + plot_annotation(tag_levels = "A")
print(p_combined)

ggsave(file.path(OUTPUT_DIR, "logFC_pval_vs_delta_kWithin.pdf"),
       p_combined, width = 14, height = 6)
ggsave(file.path(OUTPUT_DIR, "logFC_pval_vs_delta_kWithin.png"),
       p_combined, width = 14, height = 6, dpi = 300)
```

## Effect Size Distribution: DEG vs Control

```{r effect_size_boxplots, fig.width=10, fig.height=5}
# Panel A: |logFC|
p_box_logfc <- ggplot(scatter_df, aes(x = group, y = abs_logFC, fill = group)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.3) +
  scale_fill_manual(values = c("Topology control" = "steelblue", "DEG" = "grey60")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(x = NULL, y = expression("|" * log[2] * "FC" * "|"))

# Panel B: -log10(FDR)
p_box_pval <- ggplot(scatter_df, aes(x = group, y = neg_log10p, fill = group)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.3) +
  scale_fill_manual(values = c("Topology control" = "steelblue", "DEG" = "grey60")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(x = NULL, y = expression("-" * log[10] * "(FDR)"))

p_box_combined <- p_box_logfc + p_box_pval + plot_annotation(tag_levels = "A")
print(p_box_combined)

wt_logfc <- wilcox.test(abs_logFC ~ group, data = scatter_df)
wt_pval  <- wilcox.test(neg_log10p ~ group, data = scatter_df)
cat("|logFC| Wilcoxon p:", format.pval(wt_logfc$p.value), "\n")
cat("-log10(FDR) Wilcoxon p:", format.pval(wt_pval$p.value), "\n")

ggsave(file.path(OUTPUT_DIR, "effect_size_DEG_vs_control.pdf"),
       p_box_combined, width = 10, height = 5)
ggsave(file.path(OUTPUT_DIR, "effect_size_DEG_vs_control.png"),
       p_box_combined, width = 10, height = 5, dpi = 300)

ggsave(file.path(OUTPUT_DIR, "logFC_DEG_vs_control.pdf"),
       p_logfc, width = 5, height = 5)
ggsave(file.path(OUTPUT_DIR, "logFC_DEG_vs_control.png"),
       p_logfc, width = 5, height = 5, dpi = 300)
```

## Z-score Violin Plot (Main Figure Candidate)

Two-panel figure comparing connectivity disruption in DEG modules versus
topology control modules. Z-scores are computed per module as
$(observed - mean_{null}) / sd_{null}$. Red dots mark observed values.

```{r zscore_violins, fig.width=10, fig.height=10}
# Z-score the null distributions
deg_null_z <- perm_null_data <- null_deltas
deg_null_mean <- colMeans(null_deltas)
deg_null_sd <- apply(null_deltas, 2, sd)
for (m in colnames(deg_null_z)) {
  deg_null_z[, m] <- (deg_null_z[, m] - deg_null_mean[m]) / deg_null_sd[m]
}
deg_obs_z <- (observed_delta - deg_null_mean) / deg_null_sd

ctrl_null_z <- null_control
ctrl_null_mean <- colMeans(null_control)
ctrl_null_sd <- apply(null_control, 2, sd)
for (m in colnames(ctrl_null_z)) {
  ctrl_null_z[, m] <- (ctrl_null_z[, m] - ctrl_null_mean[m]) / ctrl_null_sd[m]
}
ctrl_obs_z <- (observed_control - ctrl_null_mean) / ctrl_null_sd

# DEG violin data
deg_violin <- map_dfr(colnames(deg_null_z), function(m) {
  tibble(module = m, z_value = deg_null_z[, m])
})
# Compute FDR across all kWithin modules (DEG + control)
# Use positional indexing to avoid duplicate name collisions
n_deg <- length(ecdf_pvals)
n_ctrl <- length(ecdf_pvals_control)
all_kw_pvals <- c(ecdf_pvals, ecdf_pvals_control)
all_kw_fdr <- p.adjust(all_kw_pvals, method = "BH")
kw_fdr_deg  <- setNames(all_kw_fdr[1:n_deg], names(ecdf_pvals))
kw_fdr_ctrl <- setNames(all_kw_fdr[(n_deg + 1):(n_deg + n_ctrl)], names(ecdf_pvals_control))

deg_points <- tibble(
  module = names(deg_obs_z),
  z_observed = deg_obs_z,
  fdr = kw_fdr_deg[names(deg_obs_z)]
)
deg_order <- deg_points |> arrange(z_observed) |> pull(module)
deg_violin$module <- factor(deg_violin$module, levels = deg_order)
deg_points$module <- factor(deg_points$module, levels = deg_order)

# Control violin data
ctrl_violin <- map_dfr(colnames(ctrl_null_z), function(m) {
  tibble(module = m, z_value = ctrl_null_z[, m])
})
ctrl_points <- tibble(
  module = names(ctrl_obs_z),
  z_observed = ctrl_obs_z,
  fdr = kw_fdr_ctrl[names(ctrl_obs_z)]
)
ctrl_order <- ctrl_points |> arrange(z_observed) |> pull(module)
ctrl_violin$module <- factor(ctrl_violin$module, levels = ctrl_order)
ctrl_points$module <- factor(ctrl_points$module, levels = ctrl_order)

p_deg_violin <- make_violin_panel(deg_violin, deg_points,
                                   "DEG coexpression modules (kWithin)", "grey70")
p_ctrl_violin <- make_violin_panel(ctrl_violin, ctrl_points,
                                    "Topology control (kWithin)", ctrl_color_hex)

p_violin_combined <- p_deg_violin / p_ctrl_violin +
  plot_annotation(tag_levels = "A")
print(p_violin_combined)

ggsave(file.path(OUTPUT_DIR, "zscore_violin_DEG_vs_control.pdf"),
       p_violin_combined, width = 10, height = 10)
ggsave(file.path(OUTPUT_DIR, "zscore_violin_DEG_vs_control.png"),
       p_violin_combined, width = 10, height = 10, dpi = 300)
```

# Summary

```{r summary}
n_deg_sig <- sum(all_results$type == "DEG" & all_results$fdr < 0.05)
n_ctrl_sig <- sum(all_results$type == "control" & all_results$fdr < 0.05)

cat("=== PERMUTATION TEST SUMMARY ===\n\n")
cat("DEG modules significant (FDR < 0.05):", n_deg_sig, "of",
    sum(all_results$type == "DEG"), "\n")
cat("Control modules significant (FDR < 0.05):", n_ctrl_sig, "of",
    sum(all_results$type == "control"), "\n\n")

cat("Significant DEG modules:\n")
all_results |>
  filter(type == "DEG", fdr < 0.05) |>
  select(module, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()

cat("\nConclusion: Connectivity loss in DEG modules exceeds the genotype\n")
cat("label permutation null, while non-DEG control genes (matched for\n")
cat("coexpression structure and sample composition) show no significant\n")
cat("connectivity loss. This rules out the possibility that the observed\n")
cat("disruption is an artifact of building separate genotype networks\n")
cat("from differentially expressed genes.\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
