---
title: "Phenotype Association Filter — Inv4m DEGs × FT/PH Candidates"
author: "Fausto Rodriguez Zapata"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Identify Inv4m DEGs that overlap published flowering time (FT) and
plant height (PH) candidate gene lists. Produces Table 2 (`tab:FT_PH_candidates`)
in the manuscript.

**Methods reference**: Lines 1406–1432 of `main.tex`

**Sources for FT candidates**:

1. Wang et al. 2021 — 991 highland-adapted FT genes (v3 IDs)
2. Li et al. 2022 — 62 FT genes (v4 IDs)
3. GWAS Atlas — "days to flowering" PPTO:0000155 (v4 coords)
4. Tibbs-Cortes et al. 2024 — plasticity SNPs for DTS/DTA (v5 coords)

**Sources for PH candidates**:

1. Li et al. 2022 — 23 PH genes (v4 IDs, hardcoded)
2. GWAS Atlas — "plant height" PPTO:0000126 (v4 coords)
3. Tibbs-Cortes et al. 2024 — plasticity SNPs for PH (v5 coords)

# Setup

```{r libraries}
library(tidyverse)
library(GenomicRanges)
library(rtracklayer)
library(xtable)
```

```{r constants}
# Thresholds
FDR_THRESHOLD <- 0.05
LOGFC_THRESHOLD <- 1.5
FLANK_BP <- 5000 # ±5 kb extension for GWAS SNP–gene overlap

# Inv4m region boundaries (v5 coordinates)
INV4M_START <- 172883675 # Zm00001eb190470 gene start
INV4M_END   <- 188132113 # Zm00001eb194800 gene end

# Shared introgression region (broader)
SHARED_START <- 157012149
SHARED_END   <- 195900523
```

# Load Gene Annotations

```{r load-annotations}
# Gene symbols (for multi-alias expansion, e.g. jmj2/jmj6/jmj9)
gene_symbols <- read.delim(
  file.path(paths$data, "gene_symbol.tab"),
  stringsAsFactors = FALSE
)

# Master labels (consolidated: curated > symbol > llm_proposed)
master_labels <- read.csv(
  file.path(paths$data, "locus_labels_master.csv"),
  stringsAsFactors = FALSE
) %>%
  dplyr::select(gene_model = 1, master_label = 4) %>%
  dplyr::filter(!is.na(master_label) & master_label != "")

# PANNZER functional descriptions — keep first per gene
gene_pannzer <- read.delim(
  file.path(paths$data, "PANNZER_DESC.tab"),
  stringsAsFactors = FALSE
) %>%
  group_by(gene_model) %>%
  dplyr::slice(1) %>%
  ungroup()

# v3 → v5 cross-reference (no header; one-to-many comma-separated)
v3_v5_raw <- read.delim(
  file.path(paths$data, "B73v3_to_B73v5.tsv"),
  header = FALSE,
  col.names = c("v3", "v5_list"),
  stringsAsFactors = FALSE
)
v3_v5 <- v3_v5_raw %>%
  separate_rows(v5_list, sep = ",") %>%
  dplyr::rename(v5 = v5_list) %>%
  dplyr::filter(v5 != "")

# v4 → v5 cross-reference
v4_v5_raw <- read.delim(
  file.path(paths$data, "B73v4_to_B73v5.tsv"),
  header = FALSE,
  col.names = c("v4", "v5_list"),
  stringsAsFactors = FALSE
)
v4_v5 <- v4_v5_raw %>%
  separate_rows(v5_list, sep = ",") %>%
  dplyr::rename(v5 = v5_list) %>%
  dplyr::filter(v5 != "")

cat("v3→v5 mappings:", nrow(v3_v5), "\n")
cat("v4→v5 mappings:", nrow(v4_v5), "\n")
```

# Load Gene Coordinates

```{r load-gff-v5}
# v5 GFF — genes on chr 1-10 only
v5_gff <- import(file.path(paths$data,
  "Zea_mays.Zm-B73-REFERENCE-NAM-5.0.59.chr.gff3"))
v5_genes <- v5_gff[v5_gff$type == "gene" &
                    seqnames(v5_gff) %in% as.character(1:10)]
v5_genes$gene_id <- sub("^gene:", "", v5_genes$ID)
names(v5_genes) <- v5_genes$gene_id

# Extended ±5 kb version for SNP overlap
v5_genes_ext <- v5_genes
start(v5_genes_ext) <- pmax(1, start(v5_genes_ext) - FLANK_BP)
end(v5_genes_ext) <- end(v5_genes_ext) + FLANK_BP

cat("v5 genes loaded:", length(v5_genes), "\n")
```

```{r load-gff-v4}
# v4 GFF — genes on chr 1-10 only
v4_gff <- import(file.path(paths$data,
  "Zm-B73-REFERENCE-GRAMENE-4.0_Zm00001d.2.gff3"))
v4_genes <- v4_gff[v4_gff$type == "gene" &
                    seqnames(v4_gff) %in% as.character(1:10)]
v4_genes$gene_id <- sub("^gene:", "", v4_genes$ID)
names(v4_genes) <- v4_genes$gene_id

# Extended ±5 kb version for GWAS Atlas SNP overlap
v4_genes_ext <- v4_genes
start(v4_genes_ext) <- pmax(1, start(v4_genes_ext) - FLANK_BP)
end(v4_genes_ext) <- end(v4_genes_ext) + FLANK_BP

cat("v4 genes loaded:", length(v4_genes), "\n")
```

# Load Differential Expression Results

```{r load-effects}
effects_all <- read.csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  stringsAsFactors = FALSE
)

# Genes whose own SNPs are in LD with the Inv4m tagging marker. Differential
# expression at these loci is confounded with cis effects of the linked SNPs,
# so they are excluded from candidate selection. The same filter is applied
# upstream in the trans-coexpression network notebook.
ld_linked_path <- file.path(paths$intermediate, "inv4m_linked_genes_to_filter.txt")
ld_linked_genes <- if (file.exists(ld_linked_path)) {
  read.table(ld_linked_path, header = TRUE, stringsAsFactors = FALSE)$gene_id
} else {
  character(0)
}

# Filter to Inv4m predictor, keep best p-value per gene
effects <- effects_all %>%
  dplyr::filter(predictor == "Inv4m") %>%
  dplyr::filter(!gene %in% ld_linked_genes) %>%
  mutate(
    is_significant = adj.P.Val < FDR_THRESHOLD,
    is_topDEG = is_significant & abs(logFC) > LOGFC_THRESHOLD
  ) %>%
  group_by(gene) %>%
  arrange(adj.P.Val) %>%
  dplyr::slice(1) %>%
  ungroup()

cat("LD-linked genes excluded:", length(ld_linked_genes), "\n")
cat("Total Inv4m-tested genes (after LD filter):", nrow(effects), "\n")
cat("Significant DEGs (FDR < 0.05):", sum(effects$is_significant), "\n")
cat("Top DEGs (|logFC| >", LOGFC_THRESHOLD, "):",
    sum(effects$is_topDEG), "\n")
```

# Build Flowering Time Candidate List

## Wang et al. 2021 (991 highland-adapted FT genes)

```{r ft-wang2021}
liwang <- read.csv(
  file.path(paths$data, "floweringTime_genes_liwang.csv"),
  stringsAsFactors = FALSE
)

# Convert v3 IDs to v5
wang_v5 <- liwang %>%
  dplyr::select(v3 = gene) %>%
  inner_join(v3_v5, by = "v3", relationship = "many-to-many") %>%
  pull(v5) %>%
  unique()

cat("Wang 2021: ", nrow(liwang), "v3 genes →", length(wang_v5), "v5 genes\n")
```

## Li et al. 2022 (62 FT genes)

```{r ft-li2022}
li2022_ft_v4 <- readLines(file.path(paths$data, "li2022_FT.list"))
li2022_ft_v4 <- li2022_ft_v4[li2022_ft_v4 != ""]

li2022_ft_v5 <- v4_v5 %>%
  dplyr::filter(v4 %in% li2022_ft_v4) %>%
  pull(v5) %>%
  unique()

cat("Li 2022 FT:", length(li2022_ft_v4), "v4 genes →",
    length(li2022_ft_v5), "v5 genes\n")
```

## GWAS Atlas — days to flowering (PPTO:0000155)

```{r ft-gwas-atlas}
atlas <- read.delim(
  file.path(paths$data, "gwas_association_result_for_maize.txt"),
  stringsAsFactors = FALSE,
  comment.char = "",
  check.names = FALSE
)

# Standardise column names
colnames(atlas)[colnames(atlas) == "#GaP_id"] <- "idx"
colnames(atlas)[colnames(atlas) == "Trait accession"] <- "TOID"
colnames(atlas)[colnames(atlas) == "P-value"] <- "p.value"
colnames(atlas)[colnames(atlas) == "PMID"] <- "PUBMID"
colnames(atlas)[colnames(atlas) == "First author"] <- "first_author"

# FT associations
atlas_ft <- atlas %>% dplyr::filter(TOID == "PPTO:0000155")

# Build GRanges for FT SNPs (v4 coordinates)
atlas_ft_gr <- with(atlas_ft,
  GRanges(
    seqnames = Chr,
    ranges = IRanges(start = Pos, end = Pos),
    idx = idx,
    p.value = as.numeric(p.value),
    PUBMID = PUBMID,
    first_author = first_author
  )
)

# Overlap with v4 genes (±5 kb)
olap <- findOverlaps(atlas_ft_gr, v4_genes_ext)

atlas_ft_hits <- data.frame(
  v4 = v4_genes_ext$gene_id[subjectHits(olap)],
  p.value = atlas_ft_gr$p.value[queryHits(olap)],
  stringsAsFactors = FALSE
) %>%
  group_by(v4) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%  # best p-value per gene
  ungroup()

# Convert to v5
atlas_ft_v5 <- atlas_ft_hits %>%
  inner_join(v4_v5, by = "v4", relationship = "many-to-many") %>%
  group_by(v5) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup()

cat("GWAS Atlas FT: ", nrow(atlas_ft), "SNPs →",
    nrow(atlas_ft_hits), "v4 genes →",
    length(unique(atlas_ft_v5$v5)), "v5 genes\n")
```

## Tibbs-Cortes et al. 2024 — FT plasticity SNPs

```{r ft-tibbs-cortes}
plasticity <- read.csv(
  file.path(paths$data, "Supplemental_Table_S3.csv"),
  skip = 1,
  stringsAsFactors = FALSE
)

# FT traits: DTS_slope, DTS_intcp, DTA_slope, DTA_intcp
tibbs_ft_long <- plasticity %>%
  dplyr::select(seqnames = Chr, start = bp, Marker,
                DTS_slope, DTS_intcp, DTA_slope, DTA_intcp) %>%
  tidyr::pivot_longer(
    cols = c(DTS_slope, DTS_intcp, DTA_slope, DTA_intcp),
    values_to = "P", names_to = "trait"
  ) %>%
  dplyr::filter(!is.na(P))

# Build GRanges (v5 coordinates)
tibbs_ft_gr <- with(tibbs_ft_long,
  GRanges(
    seqnames = seqnames,
    ranges = IRanges(start = start, end = start),
    P = P,
    Marker = Marker
  )
)

# Overlap with v5 genes (±5 kb)
olap <- findOverlaps(tibbs_ft_gr, v5_genes_ext)

tibbs_ft_hits <- data.frame(
  gene = v5_genes_ext$gene_id[subjectHits(olap)],
  p.value = tibbs_ft_gr$P[queryHits(olap)],
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(!is.na(gene)) %>%
  dplyr::distinct() %>%
  group_by(gene) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup()

cat("Tibbs-Cortes FT:", nrow(tibbs_ft_long), "SNP-trait pairs →",
    length(unique(tibbs_ft_hits$gene)), "v5 genes\n")
```

## Combine FT Candidate List

```{r ft-combine}
FT_candidates <- sort(unique(c(
  wang_v5,
  li2022_ft_v5,
  atlas_ft_v5$v5,
  tibbs_ft_hits$gene
)))

cat("Nonredundant FT candidates:", length(FT_candidates),
    "(expected ~2224)\n")
```

# Build Plant Height Candidate List

## Li et al. 2022 (23 PH genes)

```{r ph-li2022}
# Hardcoded v4 IDs from original analysis
li2022_ph_v4 <- c(
  "Zm00001d010396", "Zm00001d011069", "Zm00001d029657", "Zm00001d029656",
  "Zm00001d029658", "Zm00001d029659", "Zm00001d029758", "Zm00001d029884",
  "Zm00001d032747", "Zm00001d032776", "Zm00001d032788", "Zm00001d032790",
  "Zm00001d032791", "Zm00001d033204", "Zm00001d033206", "Zm00001d033205",
  "Zm00001d033223", "Zm00001d033224", "Zm00001d033225", "Zm00001d033226",
  "Zm00001d033228", "Zm00001d033227"
)

li2022_ph_v5 <- v4_v5 %>%
  dplyr::filter(v4 %in% li2022_ph_v4) %>%
  pull(v5) %>%
  unique()

cat("Li 2022 PH:", length(li2022_ph_v4), "v4 genes →",
    length(li2022_ph_v5), "v5 genes\n")
```

## GWAS Atlas — plant height (PPTO:0000126)

```{r ph-gwas-atlas}
atlas_ph <- atlas %>% dplyr::filter(TOID == "PPTO:0000126")

atlas_ph_gr <- with(atlas_ph,
  GRanges(
    seqnames = Chr,
    ranges = IRanges(start = Pos, end = Pos),
    idx = idx,
    p.value = as.numeric(p.value),
    PUBMID = PUBMID,
    first_author = first_author
  )
)

olap <- findOverlaps(atlas_ph_gr, v4_genes_ext)

atlas_ph_hits <- data.frame(
  v4 = v4_genes_ext$gene_id[subjectHits(olap)],
  p.value = atlas_ph_gr$p.value[queryHits(olap)],
  stringsAsFactors = FALSE
) %>%
  group_by(v4) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup()

atlas_ph_v5 <- atlas_ph_hits %>%
  inner_join(v4_v5, by = "v4", relationship = "many-to-many") %>%
  group_by(v5) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup()

cat("GWAS Atlas PH:", nrow(atlas_ph), "SNPs →",
    nrow(atlas_ph_hits), "v4 genes →",
    length(unique(atlas_ph_v5$v5)), "v5 genes\n")
```

## Tibbs-Cortes et al. 2024 — PH plasticity SNPs

```{r ph-tibbs-cortes}
tibbs_ph_long <- plasticity %>%
  dplyr::select(seqnames = Chr, start = bp, Marker,
                PH_slope, PH_intcp) %>%
  tidyr::pivot_longer(
    cols = c(PH_slope, PH_intcp),
    values_to = "P", names_to = "trait"
  ) %>%
  dplyr::filter(!is.na(P))

tibbs_ph_gr <- with(tibbs_ph_long,
  GRanges(
    seqnames = seqnames,
    ranges = IRanges(start = start, end = start),
    P = P,
    Marker = Marker
  )
)

olap <- findOverlaps(tibbs_ph_gr, v5_genes_ext)

tibbs_ph_hits <- data.frame(
  gene = v5_genes_ext$gene_id[subjectHits(olap)],
  p.value = tibbs_ph_gr$P[queryHits(olap)],
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(!is.na(gene)) %>%
  dplyr::distinct() %>%
  group_by(gene) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup()

cat("Tibbs-Cortes PH:", nrow(tibbs_ph_long), "SNP-trait pairs →",
    length(unique(tibbs_ph_hits$gene)), "v5 genes\n")
```

## Combine PH Candidate List

```{r ph-combine}
PH_candidates <- sort(unique(c(
  li2022_ph_v5,
  atlas_ph_v5$v5,
  tibbs_ph_hits$gene
)))

cat("Nonredundant PH candidates:", length(PH_candidates),
    "(expected ~1272)\n")
```

# Join DEGs × Candidates

```{r annotate-effects}
# Flag candidate status in effects table
effects <- effects %>%
  mutate(
    is_FT_candidate = gene %in% FT_candidates,
    is_PH_candidate = gene %in% PH_candidates
  )

cat("Inv4m DEGs that are FT candidates:",
    sum(effects$is_significant & effects$is_FT_candidate), "\n")
cat("Inv4m top DEGs that are FT candidates:",
    sum(effects$is_topDEG & effects$is_FT_candidate), "\n")
cat("Inv4m DEGs that are PH candidates:",
    sum(effects$is_significant & effects$is_PH_candidate), "\n")
cat("Inv4m top DEGs that are PH candidates:",
    sum(effects$is_topDEG & effects$is_PH_candidate), "\n")
```

```{r build-gwas-pvalues}
# Collect best GWAS p-value per gene across all sources (FT)
gwas_pval_ft <- bind_rows(
  atlas_ft_v5 %>% dplyr::select(gene = v5, p.value),
  tibbs_ft_hits %>% dplyr::select(gene, p.value)
) %>%
  group_by(gene) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::rename(gwas_pvalue_ft = p.value)

# Collect best GWAS p-value per gene across all sources (PH)
gwas_pval_ph <- bind_rows(
  atlas_ph_v5 %>% dplyr::select(gene = v5, p.value),
  tibbs_ph_hits %>% dplyr::select(gene, p.value)
) %>%
  group_by(gene) %>%
  arrange(p.value) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::rename(gwas_pvalue_ph = p.value)
```

```{r pmid-mapping}
# Map GWAS Atlas PMIDs to BibTeX citation keys
pmid_to_cite <- c(
  "24514905" = "peiffer2014",
  "27012534" = "li2016",
  "22660547" = "jiao2012",
  "23840585" = "peiffer2013"
)

# Helper: resolve PMIDs to cite keys, adding liu2023 (the GWAS Atlas database)
resolve_atlas_cites <- function(pmids) {
  keys <- unique(na.omit(pmid_to_cite[as.character(pmids)]))
  # Always include liu2023 as the database reference
  sort(unique(c(keys, "liu2023")))
}
```

```{r build-reference-tracking}
# --- GWAS Atlas: track actual PMIDs per gene ---

# FT atlas refs with PMIDs
atlas_ft_with_pmid <- data.frame(
  v4 = v4_genes_ext$gene_id[subjectHits(findOverlaps(atlas_ft_gr, v4_genes_ext))],
  PUBMID = atlas_ft_gr$PUBMID[queryHits(findOverlaps(atlas_ft_gr, v4_genes_ext))],
  stringsAsFactors = FALSE
) %>%
  inner_join(v4_v5, by = "v4", relationship = "many-to-many") %>%
  dplyr::select(gene = v5, PUBMID) %>%
  distinct()

# PH atlas refs with PMIDs
atlas_ph_with_pmid <- data.frame(
  v4 = v4_genes_ext$gene_id[subjectHits(findOverlaps(atlas_ph_gr, v4_genes_ext))],
  PUBMID = atlas_ph_gr$PUBMID[queryHits(findOverlaps(atlas_ph_gr, v4_genes_ext))],
  stringsAsFactors = FALSE
) %>%
  inner_join(v4_v5, by = "v4", relationship = "many-to-many") %>%
  dplyr::select(gene = v5, PUBMID) %>%
  distinct()

# Build per-gene citation lists for FT
ft_cite_list <- list()

# wang2021
for (g in wang_v5) ft_cite_list[[g]] <- c(ft_cite_list[[g]], "wang2021")
# li2022
for (g in li2022_ft_v5) ft_cite_list[[g]] <- c(ft_cite_list[[g]], "li2022")
# GWAS Atlas (with PMID resolution)
for (i in seq_len(nrow(atlas_ft_with_pmid))) {
  g <- atlas_ft_with_pmid$gene[i]
  cites <- resolve_atlas_cites(atlas_ft_with_pmid$PUBMID[i])
  ft_cite_list[[g]] <- c(ft_cite_list[[g]], cites)
}
# tibbs-cortes
for (g in tibbs_ft_hits$gene) ft_cite_list[[g]] <- c(ft_cite_list[[g]], "tibbs-cortes2024")

ft_refs <- data.frame(
  gene = names(ft_cite_list),
  ft_refs = sapply(ft_cite_list, function(x) paste(sort(unique(x)), collapse = ", ")),
  stringsAsFactors = FALSE,
  row.names = NULL
)

# Build per-gene citation lists for PH
ph_cite_list <- list()

# li2022
for (g in li2022_ph_v5) ph_cite_list[[g]] <- c(ph_cite_list[[g]], "li2022")
# GWAS Atlas (with PMID resolution)
for (i in seq_len(nrow(atlas_ph_with_pmid))) {
  g <- atlas_ph_with_pmid$gene[i]
  cites <- resolve_atlas_cites(atlas_ph_with_pmid$PUBMID[i])
  ph_cite_list[[g]] <- c(ph_cite_list[[g]], cites)
}
# tibbs-cortes
for (g in tibbs_ph_hits$gene) ph_cite_list[[g]] <- c(ph_cite_list[[g]], "tibbs-cortes2024")

ph_refs <- data.frame(
  gene = names(ph_cite_list),
  ph_refs = sapply(ph_cite_list, function(x) paste(sort(unique(x)), collapse = ", ")),
  stringsAsFactors = FALSE,
  row.names = NULL
)
```

```{r build-tables}
# Location label helper
location_label <- function(in_inv4m, in_flank) {
  case_when(
    in_inv4m ~ "Inv4m",
    in_flank ~ "flanking",
    TRUE ~ "outside"
  )
}

# Genes with multiple aliases in gene_symbol.tab (for row expansion, e.g. jmj2/jmj6/jmj9)
gene_symbol_aliases <- gene_symbols %>%
  dplyr::filter(!is.na(locus_symbol) & locus_symbol != "")

# Identify multi-alias genes (genes with >1 alias in gene_symbol.tab)
multi_alias_genes <- gene_symbol_aliases %>%
  group_by(gene_model) %>%
  dplyr::filter(n() > 1) %>%
  ungroup()

# Annotate and expand: use desc (PANNZER) as primary description,
# use locus_labels_master.csv for shorthand labels,
# expand only multi-alias genes (e.g. jmj2/jmj6/jmj9) from gene_symbol.tab
annotate_and_expand <- function(df) {
  # Supplement description: prefer desc (PANNZER functional) over desc_merged
  df <- df %>%
    mutate(
      description = ifelse(!is.na(desc) & desc != "", desc, desc_merged),
      description = ifelse(is.na(description) | description == "",
                           "Uncharacterized protein", description)
    )

  # Expand multi-alias genes (creates multiple rows)
  genes_with_multi_aliases <- df %>%
    dplyr::filter(gene %in% multi_alias_genes$gene_model) %>%
    inner_join(
      multi_alias_genes %>% dplyr::select(gene_model, alias = locus_symbol),
      by = c("gene" = "gene_model"),
      relationship = "many-to-many"
    ) %>%
    mutate(locus_symbol = alias) %>%
    dplyr::select(-alias)

  # All other genes: use master label
  other_genes <- df %>%
    dplyr::filter(!(gene %in% multi_alias_genes$gene_model)) %>%
    left_join(master_labels, by = c("gene" = "gene_model")) %>%
    mutate(locus_symbol = ifelse(
      !is.na(master_label) & master_label != "",
      master_label,
      ifelse(is.na(locus_symbol) | locus_symbol == "", "", locus_symbol)
    )) %>%
    dplyr::select(-master_label)

  bind_rows(genes_with_multi_aliases, other_genes)
}

# FT top DEGs table
FT_table <- effects %>%
  dplyr::filter(is_topDEG & is_FT_candidate) %>%
  left_join(gwas_pval_ft, by = "gene") %>%
  left_join(ft_refs, by = "gene") %>%
  mutate(location = location_label(in_Inv4m, in_flank)) %>%
  dplyr::select(gene, locus_symbol, desc_merged, desc, logFC, neglogP,
                location, gwas_pvalue_ft, ft_refs) %>%
  annotate_and_expand() %>%
  arrange(-neglogP, gene, locus_symbol)

# PH top DEGs table
PH_table <- effects %>%
  dplyr::filter(is_topDEG & is_PH_candidate) %>%
  left_join(gwas_pval_ph, by = "gene") %>%
  left_join(ph_refs, by = "gene") %>%
  mutate(location = location_label(in_Inv4m, in_flank)) %>%
  dplyr::select(gene, locus_symbol, desc_merged, desc, logFC, neglogP,
                location, gwas_pvalue_ph, ph_refs) %>%
  annotate_and_expand() %>%
  arrange(-neglogP, gene, locus_symbol)

cat("FT top DEGs for table:", nrow(FT_table), "\n")
cat("PH top DEGs for table:", nrow(PH_table), "\n")
```

```{r display-tables}
knitr::kable(
  FT_table %>% dplyr::select(gene, locus_symbol, description, logFC, neglogP,
                              location, gwas_pvalue_ft, ft_refs),
  digits = 2, caption = "Flowering Time candidate DEGs (|logFC| > 1.5)")
knitr::kable(
  PH_table %>% dplyr::select(gene, locus_symbol, description, logFC, neglogP,
                              location, gwas_pvalue_ph, ph_refs),
  digits = 2, caption = "Plant Height candidate DEGs (|logFC| > 1.5)")
```

# Fisher Enrichment Tests

```{r fisher-ft}
cat("=== Flowering Time Enrichment ===\n\n")

# Top DEGs vs FT candidates
ft_top <- table(effects$is_topDEG, effects$is_FT_candidate)
dimnames(ft_top) <- list(TopDEG = c("No", "Yes"),
                         FT_candidate = c("No", "Yes"))
cat("Top DEGs (|logFC| >", LOGFC_THRESHOLD, ") × FT candidates:\n")
print(ft_top)
cat("\nFisher exact test (two-sided):\n")
print(fisher.test(ft_top))
cat("\nFisher exact test (greater):\n")
print(fisher.test(ft_top, alternative = "greater"))

# All significant DEGs vs FT candidates
ft_sig <- table(effects$is_significant, effects$is_FT_candidate)
dimnames(ft_sig) <- list(DEG = c("No", "Yes"),
                         FT_candidate = c("No", "Yes"))
cat("\nAll significant DEGs × FT candidates:\n")
print(ft_sig)
cat("\nFisher exact test (two-sided):\n")
print(fisher.test(ft_sig))
cat("\nFisher exact test (greater):\n")
print(fisher.test(ft_sig, alternative = "greater"))
```

```{r fisher-ph}
cat("=== Plant Height Enrichment ===\n\n")

ph_top <- table(effects$is_topDEG, effects$is_PH_candidate)
dimnames(ph_top) <- list(TopDEG = c("No", "Yes"),
                         PH_candidate = c("No", "Yes"))
cat("Top DEGs (|logFC| >", LOGFC_THRESHOLD, ") × PH candidates:\n")
print(ph_top)
cat("\nFisher exact test (two-sided):\n")
print(fisher.test(ph_top))

ph_sig <- table(effects$is_significant, effects$is_PH_candidate)
dimnames(ph_sig) <- list(DEG = c("No", "Yes"),
                         PH_candidate = c("No", "Yes"))
cat("\nAll significant DEGs × PH candidates:\n")
print(ph_sig)
cat("\nFisher exact test (two-sided):\n")
print(fisher.test(ph_sig))
```

# Output

## CSV

```{r write-csv}
out_csv <- bind_rows(
  FT_table %>% mutate(
    pheno = "Flowering Time",
    gwas_pvalue = gwas_pvalue_ft,
    refs = ft_refs
  ) %>% dplyr::select(-gwas_pvalue_ft, -ft_refs),
  PH_table %>% mutate(
    pheno = "Plant Height",
    gwas_pvalue = gwas_pvalue_ph,
    refs = ph_refs
  ) %>% dplyr::select(-gwas_pvalue_ph, -ph_refs)
) %>%
  dplyr::select(pheno, gene, locus_symbol, description, logFC, neglogP,
                location, gwas_pvalue, refs) %>%
  arrange(pheno, -neglogP)

write.csv(out_csv,
  file.path(paths$intermediate, "inv4m_DEGs_by_phenotype_association.csv"),
  row.names = FALSE
)
cat("Wrote:", file.path(paths$intermediate,
    "inv4m_DEGs_by_phenotype_association.csv"), "\n")
```

## LaTeX Table

```{r write-latex}
# Format for LaTeX
format_pval <- function(x) {
  ifelse(is.na(x), "", formatC(x, format = "e", digits = 2))
}

tex_lines <- c(
  "\\begin{table*}[!ht]",
  "\\caption{\\textbf{Effect of \\invfour on Flowering Time and Plant Height Gene",
  "Candidates}}",
  "\\label{tab:FT_PH_candidates}",
  "\\resizebox{\\textwidth}{!}{%",
  "\\begin{tabular}{llllllll}",
  "\\toprule",
  "  \\textbf{Gene} &",
  "  \\textbf{Label} &",
  "  \\textbf{Description} &",
  "  \\textbf{\\begin{tabular}[c]{@{}l@{}}$\\bf{log_2}$\\\\(FC)\\end{tabular}} &",
  "  \\textbf{\\begin{tabular}[c]{@{}l@{}}$\\bf{-log_{10}}$\\\\",
  "(\\textit{FDR})\\end{tabular}} &",
  "  \\textbf{Location} &",
  "  \\textbf{\\begin{tabular}[c]{@{}l@{}} GWAS \\\\",
  "\\textit{p-value}\\end{tabular}} &",
  "  \\textbf{ref} \\\\ \\midrule"
)

# Helper: format a single gene row
format_row <- function(gene, symbol, desc, logfc, nlogp, loc, pval, refs_str) {
  # Escape LaTeX special characters in description
  desc_clean <- gsub("&", "\\\\&", desc)
  desc_clean <- gsub("%", "\\\\%", desc_clean)
  desc_clean <- gsub("_", "\\\\_", desc_clean)

  # Format location: bold Inv4m
  loc_fmt <- if (loc == "Inv4m") "\\textbf{\\invfour}" else loc

  # Format GWAS p-value
  pval_fmt <- format_pval(pval)

  # Build citation string — refs_str already contains proper cite keys
  cite_fmt <- ""
  if (!is.na(refs_str) && refs_str != "") {
    cite_keys <- trimws(unlist(strsplit(refs_str, ",")))
    cite_fmt <- paste0("\\cite{", paste(unique(cite_keys), collapse = ", "), "}")
  }

  # Format symbol
  sym_fmt <- if (is.na(symbol) || symbol == "") "" else paste0("\\textit{", symbol, "}")

  paste0(
    "  \\textit{", gene, "} &\n",
    "  ", sym_fmt, " &\n",
    "  ", desc_clean, " &\n",
    "  ", sprintf("%.2f", logfc), " &\n",
    "  ", sprintf("%.2f", nlogp), " &\n",
    "  ", loc_fmt, " &\n",
    "  ", pval_fmt, " &\n",
    "  ", cite_fmt, " \\\\"
  )
}

# FT section
tex_lines <- c(tex_lines,
  "\\multicolumn{8}{l}{\\textbf{Flowering Time}} \\\\ \\midrule")

for (i in seq_len(nrow(FT_table))) {
  r <- FT_table[i, ]
  tex_lines <- c(tex_lines, format_row(
    r$gene, r$locus_symbol, r$description,
    r$logFC, r$neglogP, r$location,
    r$gwas_pvalue_ft, r$ft_refs
  ))
}

# PH section
tex_lines <- c(tex_lines,
  " \\midrule",
  "\\multicolumn{8}{l}{\\textbf{Plant Height}} \\\\ \\midrule")

for (i in seq_len(nrow(PH_table))) {
  r <- PH_table[i, ]
  tex_lines <- c(tex_lines, format_row(
    r$gene, r$locus_symbol, r$description,
    r$logFC, r$neglogP, r$location,
    r$gwas_pvalue_ph, r$ph_refs
  ))
}

tex_lines <- c(tex_lines,
  " \\bottomrule",
  "\\end{tabular}%",
  "}",
  "\\end{table*}"
)

writeLines(tex_lines,
  file.path(paths$tables, "FT_PH_candidates.tex"))
cat("Wrote:", file.path(paths$tables, "FT_PH_candidates.tex"), "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
