---
title: "Figure 1 Assembly"
author: "inv4m"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")

library(ggplot2)
library(cowplot)
library(ggtext)
```

# Load pre-built panels

Panels B–D come from `plot_synteny_and_repeats.Rmd`.
Panel E (genotype map) with Panel F (chr4 inset) come from
`plot_genotype_get_correlated_loci.Rmd` as `plot_with_inset`.

```{r load-panels}
p_bcd <- readRDS(file.path(paths$intermediate, "fig1_panel_BCD.rds"))
p_ef  <- readRDS(file.path(paths$intermediate, "fig1_panel_E.rds"))
```

# Panel A placeholder

Panel A (breeding scheme) is a hand-drawn/Inkscape illustration added in
post-production. We reserve space with the correct aspect ratio.

```{r panel-a-placeholder}
p_a <- ggdraw() +
  draw_label("Panel A\n(breeding scheme — added in Inkscape)",
             size = 18, color = "gray60")
```

# Panels E + F (pre-built)

`p_ef` is the full bottom row from `plot_genotype_get_correlated_loci.Rmd`:
E = whole-genome genotype map, F = chr4 zoom inset, with spike-calibrated
brackets and panel labels already included.

```{r panel-ef-preview, fig.width=18, fig.height=8}
p_ef
```

# Full Figure 1

```{r figure1-full, fig.width=18, fig.height=22}
fig1 <- plot_grid(
  p_a,
  p_bcd,
  p_ef,
  ncol = 1,
  rel_heights = c(0.20, 0.45, 0.35),
  labels = c("A", "", ""),
  label_size = 32, label_fontface = "bold",
  label_x = 0, label_y = 1)

fig1
```

```{r figure1-save}
ggsave(file.path(paths$figures, "Figure_1.png"),
       plot = fig1, width = 18, height = 22, dpi = 300, bg = "white")
ggsave(file.path(paths$figures, "Figure_1.svg"),
       plot = fig1, width = 18, height = 22, device = "svg",
       fix_text_size = TRUE)
```

# Session info

```{r session-info}
sessionInfo()
```
