---
title: "Chromosomal Knob Repeat Analysis — Figure 1 Panels A & B"
author: "inv4m"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

BLAST+ megablast hits of chromosomal knob repeats (knob180 M32524.1 and TR-1
AY083942.1) plotted along chromosome 4 of four *Zea* genomes. Vertical lines
mark inversion breakpoint boundaries from AnchorWave genome alignments.

This generates **Figure 1 Panel A** of the inversion paper in two variants:

- **Plot 2** (PT): TIL18 / PT / B73 — three-genome cowplot with crossing lines
- **Plot 3** (Mi21): TIL18 / Mi21 / B73 — same architecture, Mi21 NIL assembly

BLAST searches were run with `scripts/02_genomics_foundation/run_repeat_annotation_4genomes.sh`.

# Load BLAST results

```{r load-data}
library(dplyr)
library(ggplot2)
library(scales)
library(cowplot)

blastcols <- c("qseqid", "sseqid", "pident", "length", "mismatch",
               "gapopen", "qstart", "qend", "sstart", "send",
               "evalue", "bitscore")

genomes <- c("TIL18", "PT", "Mi21", "B73")

read_blast <- function(rep_label, prefix) {
  files <- setNames(
    file.path(paths$intermediate, paste0(prefix, "_", genomes, ".blast")),
    genomes)
  lapply(genomes, function(g) {
    blast <- read.table(files[g], sep = "\t", header = FALSE,
      col.names = blastcols,
      colClasses = c(qseqid = "character", sseqid = "character"))
    blast$genome <- g
    blast
  }) %>% bind_rows() %>%
    mutate(bitscaled = rescale(bitscore, to = c(0, 100)), rep = rep_label)
}

to_plot <- bind_rows(
  read_blast("knob180", "knob180"),
  read_blast("TR-1", "TR-1")
) %>% mutate(genome = factor(genome, levels = genomes), rep = factor(rep))
```

```{r hit-summary}
to_plot %>%
  group_by(genome, rep) %>%
  summarise(n_hits = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = rep, values_from = n_hits)
```

# Breakpoint alignment

All genomes are aligned so that the upstream breakpoint maps to the B73
upstream coordinate (172,882,309). This allows visual comparison of inversion
extent across genomes.

```{r breakpoints}
bp_all <- data.frame(
  genome = c("TIL18","TIL18","PT","PT","B73","B73","Mi21","Mi21"),
  side   = rep(c("upstream","downstream"), 4),
  xpos   = c(180365316, 193570651, 173486186, 186925483,
             172882309, 188131461, 156752504, 167621035),
  stringsAsFactors = FALSE)

b73_upstream <- 172882309
offsets <- bp_all %>% filter(side == "upstream") %>%
  mutate(offset = b73_upstream - xpos) %>% select(genome, offset)
bp_aligned <- bp_all %>% left_join(offsets, by = "genome") %>%
  mutate(xpos_aligned = xpos + offset)
to_plot <- to_plot %>% left_join(offsets, by = "genome") %>%
  mutate(sstart_aligned = sstart + offset)

# Midlines: mean repeat position between breakpoints per genome
midlines <- lapply(unique(bp_all$genome), function(g) {
  bps <- bp_all$xpos[bp_all$genome == g]
  hits <- to_plot %>% filter(genome == g,
    sstart >= min(bps) + 1e6, sstart <= max(bps) - 1e6)
  if (nrow(hits) == 0) return(NULL)
  data.frame(genome = g,
    xmid_aligned = mean(hits$sstart) + offsets$offset[offsets$genome == g])
}) %>% bind_rows()
```

```{r aligned-breakpoints}
bp_aligned %>% select(genome, side, xpos, xpos_aligned) %>%
  tidyr::pivot_wider(names_from = side, values_from = c(xpos, xpos_aligned))
```

# Plot 1: Four-genome faceted overview

Simple faceted plot of all four genomes at native coordinates (no alignment).

```{r plot1-faceted, fig.width=10, fig.height=10}
annotations <- bp_all %>%
  mutate(genome = factor(genome, levels = genomes))

p_faceted <- to_plot %>%
  ggplot(aes(x = sstart, y = bitscaled, color = rep)) +
  ylab("Normalized Repeat Match Score") +
  geom_vline(data = annotations, aes(xintercept = xpos),
             color = "#3232a0", linewidth = 1.2) +
  geom_point(size = 0.8) +
  facet_wrap(~ genome, nrow = 4, ncol = 1, scales = "free_x") +
  scale_color_manual(values = c("knob180" = "#1d7f7a", "TR-1" = "gold")) +
  scale_x_continuous(labels = unit_format(unit = "", scale = 1e-6),
                     breaks = seq(150e6, 275e6, by = 25e6)) +
  annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
  annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
  theme_minimal(base_size = 16) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(hjust = 0.8),
        strip.background = element_blank(),
        legend.position = "none")

p_faceted
```

# Cowplot assembly helpers

Shared infrastructure for the three-genome cowplot figures (Plots 2 and 3).

```{r cowplot-shared}
col_inv4m <- "purple4"
col_knob  <- "#1d7f7a"
col_gold  <- "gold"
col_gray  <- "gray50"
lw        <- 1.2

cfg <- list(
  xlim    = c(163e6, 250e6),
  xbreaks = seq(175e6, 250e6, by = 25e6))

# SVG canvas dimensions (9x6 inch at 72 dpi)
svg <- list(W = 648, H = 432, lw = 2.56)

# Shared x-axis tick positions (same for all variants)
xtick_x <- c("175" = 183.17, "200" = 314.23, "225" = 445.29, "250" = 576.35)

# Arrow geometry (svg units)
a_h <- 4; a_gap <- 1; a_pt <- 8

# --- Panel builder ---
make_panel <- function(genome_id, show_points = FALSE, bot_margin = 15) {
  gbp  <- bp_aligned %>% filter(genome == genome_id)
  gmid <- midlines %>% filter(genome == genome_id)
  gdat <- to_plot %>% filter(genome == genome_id)

  p <- ggplot() +
    geom_hline(yintercept = 0, color = "gray70", linewidth = 0.4) +
    { if (nrow(gmid) > 0)
        geom_vline(xintercept = gmid$xmid_aligned,
                   color = "gray55", linewidth = lw) } +
    geom_vline(data = gbp, aes(xintercept = xpos_aligned),
               color = col_inv4m, linewidth = lw) +
    { if (show_points)
        geom_point(data = gdat, aes(x = sstart_aligned, y = bitscaled, color = rep),
                   size = 1.2, alpha = 0.7) } +
    { if (show_points)
        scale_color_manual(values = c("knob180" = col_knob, "TR-1" = col_gold),
                           guide = "none") } +
    scale_x_continuous(labels = unit_format(unit = "", scale = 1e-6),
                       breaks = cfg$xbreaks) +
    scale_y_continuous(breaks = c(0, 25, 50, 75, 100), limits = c(-5, 105)) +
    coord_cartesian(xlim = cfg$xlim) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.margin = margin(t = 15, r = 10, b = bot_margin, l = 10),
          axis.title = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_text(color = "white"))

  if (genome_id != "B73")
    p <- p + theme(axis.text.x = element_blank())
  p
}

# --- Arrow polygon builder ---
make_arrow_grob <- function(vlines_cfg, panel, direction) {
  v <- vlines_cfg[[panel]]
  if (direction == "left") {
    cy <- v$top_y - a_gap - a_h - svg$lw
    xs <- c(v$up_x, v$up_x + a_pt, v$dn_x, v$dn_x, v$up_x + a_pt)
    ys <- c(cy,     cy + a_h, cy + a_h, cy - a_h, cy - a_h)
  } else {
    cy <- v$bot_y + a_gap + a_h + svg$lw
    xs <- c(v$up_x, v$dn_x - a_pt, v$dn_x, v$dn_x - a_pt, v$up_x)
    ys <- c(cy + a_h, cy + a_h, cy, cy - a_h, cy - a_h)
  }
  draw_grob(grid::polygonGrob(
    x = unit(xs / svg$W, "npc"), y = unit(1 - ys / svg$H, "npc"),
    gp = grid::gpar(fill = col_inv4m, col = col_inv4m)))
}

# --- Sigmoid + connection line helpers ---
sigmoid_path <- function(x1, y1, x2, y2, n = 50) {
  t <- seq(-6, 6, length.out = n)
  s <- 1 / (1 + exp(-t))
  data.frame(x = x1 + (x2 - x1) * s,
             y = y1 + (y2 - y1) * seq(0, 1, length.out = n))
}

draw_sigmoid <- function(x1, y1, x2, y2, color, lwd = lw) {
  pts <- sigmoid_path(x1, y1, x2, y2)
  draw_plot(
    ggplot(pts, aes(x, y)) +
      geom_path(color = color, linewidth = lwd) +
      scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
      scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
      theme_void(),
    x = 0, y = 0, width = 1, height = 1)
}

# --- Connection line builder ---
build_crossing_lines <- function(vlines_cfg, mid_x_cfg, pairs) {
  unlist(lapply(pairs, function(pr) {
    tv <- vlines_cfg[[pr$top]]; bv <- vlines_cfg[[pr$bot]]
    tm <- mid_x_cfg[[pr$top]]; bm <- mid_x_cfg[[pr$bot]]
    y1 <- 1 - tv$bot_y / svg$H
    y2 <- 1 - bv$top_y / svg$H
    t_up <- tv$up_x / svg$W; t_dn <- tv$dn_x / svg$W
    b_up <- bv$up_x / svg$W; b_dn <- bv$dn_x / svg$W

    if (pr$inverted) {
      # Sigmoid crossing: upstream↔downstream swap
      list(
        draw_sigmoid(tm / svg$W, y1, bm / svg$W, y2, color = "gray55"),
        draw_sigmoid(t_up, y1, b_dn, y2, color = col_inv4m),
        draw_sigmoid(t_dn, y1, b_up, y2, color = col_inv4m))
    } else {
      # Sigmoid parallel: same-side connections
      list(
        draw_sigmoid(tm / svg$W, y1, bm / svg$W, y2, color = "gray55"),
        draw_sigmoid(t_up, y1, b_up, y2, color = col_inv4m),
        draw_sigmoid(t_dn, y1, b_dn, y2, color = col_inv4m))
    }
  }), recursive = FALSE)
}

# --- Overlay annotation builder ---
legend_dot <- function(col, x, y) {
  draw_plot(
    ggplot() + annotate("point", x = 0, y = 0, color = col, size = 4) + theme_void(),
    x = x, y = y, width = 0.02, height = 0.02)
}

build_overlay <- function(genome_labels) {
  sz_name    <- 20
  sz_species <- 16
  y_header   <- 0.955
  y_bump     <- 0.02
  x_gname    <- 0.54
  x_species  <- 0.90
  y_xtick    <- 1 - 375 / svg$H

  genome_annots <- unlist(lapply(genome_labels, function(g) {
    out <- list(draw_label(g$name, x = x_gname, y = g$y + y_bump,
                           size = sz_name, fontface = "bold"))
    if (!is.null(g$desc_html))
      out <- c(out, list(draw_grob(gridtext::richtext_grob(
        g$desc_html,
        x = unit(x_species, "npc"), y = unit(g$y + y_bump, "npc"),
        hjust = 1, gp = grid::gpar(fontsize = sz_species, col = col_gray),
        box_gp = grid::gpar(col = NA, fill = NA)))))
    if (!is.null(g$desc))
      out <- c(out, list(draw_label(g$desc, x = x_species, y = g$y + y_bump,
        size = sz_species, fontface = "plain", color = col_gray, hjust = 1)))
    out
  }), recursive = FALSE)

  xtick_annots <- lapply(names(xtick_x), function(lab) {
    draw_label(lab, x = xtick_x[lab] / svg$W, y = y_xtick,
               size = 16, color = "#4D4D4D")
  })

  c(list(
    draw_label("Normalized Repeat Match Score", x = 0.02, y = 0.5,
               angle = 90, size = 20),
    draw_label("Chromosome 4 Position [Mb]", x = 0.50, y = 0.055, size = 20),
    draw_label("A", x = 0.02, y = 0.98, size = 24, fontface = "bold"),
    draw_label("Inv4m", x = 0.32, y = 0.917, size = sz_name,
               fontface = "bold.italic", color = col_inv4m),
    legend_dot(col_knob, 0.55, y_header - 0.012),
    draw_label("knob 180", x = 0.58, y = y_header, size = sz_name,
               fontface = "bold.italic", color = col_knob, hjust = 0),
    legend_dot(col_gold, 0.73, y_header - 0.012),
    draw_label("TR-1", x = 0.761, y = y_header - 0.002, size = sz_name,
               fontface = "bold.italic", color = "gray30", hjust = 0),
    draw_label("TR-1", x = 0.76, y = y_header, size = sz_name,
               fontface = "bold.italic", color = col_gold, hjust = 0)),
    genome_annots,
    xtick_annots)
}

# --- Full figure assembler ---
spacer <- ggplot() + theme_void() + theme(plot.margin = margin(0, 10, 0, 10))

build_panel_a <- function(variant) {
  vl    <- variant$vlines
  mx    <- variant$mid_x
  pairs <- variant$pairs
  gl    <- variant$genome_labels
  gnames <- variant$genomes

  crossing <- build_crossing_lines(vl, mx, pairs)
  overlay  <- build_overlay(gl)
  arrow_top <- make_arrow_grob(vl, gnames[1], "left")
  arrow_bot <- make_arrow_grob(vl, gnames[3], "right")

  function(show_points) {
    stk <- plot_grid(
      spacer,
      make_panel(gnames[1], show_points = show_points),
      make_panel(gnames[2], show_points = show_points),
      make_panel(gnames[3], show_points = show_points, bot_margin = 2),
      spacer,
      ncol = 1, align = "v", axis = "lr",
      rel_heights = c(0.15, 1, 1, 1, 0.15))
    p <- ggdraw() + draw_plot(stk, x = 0.08, y = 0.08, width = 0.86, height = 0.86)
    for (a in overlay)  p <- p + a
    for (a in crossing) p <- p + a
    p + arrow_top + arrow_bot
  }
}
```

# Plot 2: TIL18 / PT / B73

```{r plot2-pt-config}
variant_pt <- list(
  genomes = c("TIL18", "PT", "B73"),
  vlines = list(
    TIL18 = list(up_x = 172.07, dn_x = 241.30, top_y = 57.75,  bot_y = 140.45),
    PT    = list(up_x = 172.07, dn_x = 242.52, top_y = 170.33, bot_y = 253.03),
    B73   = list(up_x = 172.07, dn_x = 252.01, top_y = 282.91, bot_y = 354.59)),
  mid_x = list(TIL18 = 190.90, PT = 192.58, B73 = 231.98),
  pairs = list(
    list(top = "TIL18", bot = "PT",  inverted = FALSE),
    list(top = "PT",    bot = "B73", inverted = TRUE)),
  genome_labels = list(
    list(name = "TIL18", y = 0.875,  desc_html = "teosinte *mexicana*"),
    list(name = "PT",    y = 0.6144, desc = "highland maize"),
    list(name = "B73",   y = 0.3539, desc = NULL)))
```

```{r plot2-pt-render, fig.width=9, fig.height=6}
make_pt <- build_panel_a(variant_pt)
make_pt(TRUE)
```

```{r plot2-pt-save}
ggsave(file.path(paths$figures, "fig1_panel_A_skeleton.svg"),
       plot = make_pt(FALSE), width = 9, height = 6, device = "svg",
       fix_text_size = TRUE)
ggsave(file.path(paths$figures, "fig1_panel_A_skeleton.png"),
       plot = make_pt(TRUE), width = 9, height = 6, dpi = 300, bg = "white")
```

# Plot 3: TIL18 / Mi21 / B73

```{r plot3-mi21-config}
variant_mi21 <- list(
  genomes = c("TIL18", "Mi21", "B73"),
  vlines = list(
    TIL18 = list(up_x = 172.07, dn_x = 241.30, top_y = 57.75,  bot_y = 140.45),
    Mi21  = list(up_x = 172.07, dn_x = 229.05, top_y = 170.33, bot_y = 253.03),
    B73   = list(up_x = 172.07, dn_x = 252.01, top_y = 282.91, bot_y = 354.59)),
  mid_x = list(TIL18 = 190.90, Mi21 = 192.47, B73 = 231.98),
  pairs = list(
    list(top = "TIL18", bot = "Mi21", inverted = FALSE),
    list(top = "Mi21",  bot = "B73",  inverted = TRUE)),
  genome_labels = list(
    list(name = "TIL18", y = 0.875,  desc_html = "teosinte *mexicana*"),
    list(name = "Mi21 NIL", y = 0.6144, desc = "highland maize"),
    list(name = "B73",   y = 0.3539, desc = NULL)))
```

```{r plot3-mi21-render, fig.width=9, fig.height=6}
make_mi21 <- build_panel_a(variant_mi21)
make_mi21(TRUE)
```

```{r plot3-mi21-save}
ggsave(file.path(paths$figures, "fig1_panel_A_mi21_skeleton.svg"),
       plot = make_mi21(FALSE), width = 9, height = 6, device = "svg",
       fix_text_size = TRUE)
ggsave(file.path(paths$figures, "fig1_panel_A_mi21_skeleton.png"),
       plot = make_mi21(TRUE), width = 9, height = 6, dpi = 300, bg = "white")
```

# Panel B: Breakpoint self-similarity dotplots

LAST self-vs-self dotplots at upstream and downstream Inv4m breakpoint segments.
Dotplots were generated by `scripts/02_genomics_foundation/run_breakpoint_dotplots_4genomes.sh`.

The figure uses three genomes matching the Panel A Mi21 variant: TIL18, Mi21 NIL, B73.

```{r panel-b-config}
library(png)

panel_b_genomes <- c("TIL18", "Mi21", "B73")

panel_b_coords <- list(
  TIL18 = list(up = "180269950-180365316", down = "193570651-193734606"),
  Mi21  = list(up = "156519809-156752504", down = "167621035-167839092"),
  B73   = list(up = "172561959-172882309", down = "188131461-188220418"))

panel_b_labels <- c(TIL18 = "TIL18", Mi21 = "Mi21 NIL", B73 = "B73")

seg_size_kb <- function(range_str) {
  parts <- as.numeric(strsplit(range_str, "-")[[1]])
  round((parts[2] - parts[1]) / 1e3)
}

# Rotate PNG array 90 degrees counter-clockwise
rotate_ccw <- function(img) {
  aperm(img[, rev(seq_len(ncol(img))), , drop = FALSE], c(2, 1, 3))
}

read_dotplot <- function(genome, direction) {
  fname <- sprintf("%s_brkpt_%s.png", genome, direction)
  fpath <- file.path(paths$intermediate, fname)
  if (!file.exists(fpath)) stop("Missing: ", fpath)
  rotate_ccw(readPNG(fpath))
}

make_dotplot_cell <- function(img, size_kb) {
  ggdraw() +
    draw_image(img, x = 0, y = 0, width = 1, height = 1) +
    draw_label(paste0(size_kb, " kb"), x = 0.95, y = 0.02, hjust = 1, vjust = 0,
               size = 10, color = "gray30")
}
```

```{r panel-b-assemble, fig.width=5.5, fig.height=8}
# Mi21 PNGs are named by B73-sense orientation but the inversion flips
# upstream/downstream, so swap which file goes in which column
png_dir <- list(
  TIL18 = c(up = "up", down = "down"),
  Mi21  = c(up = "down", down = "up"),
  B73   = c(up = "up", down = "down"))

cells <- list()
for (g in panel_b_genomes) {
  for (dir in c("up", "down")) {
    img <- read_dotplot(g, png_dir[[g]][dir])
    coord_key <- dir
    size <- seg_size_kb(panel_b_coords[[g]][[coord_key]])
    cells[[paste0(g, "_", dir)]] <- make_dotplot_cell(img, size)
  }
}

# Assemble 3x2 grid with genome labels
rows <- lapply(panel_b_genomes, function(g) {
  label <- ggdraw() + draw_label(panel_b_labels[g], size = 12, fontface = "bold")
  plot_grid(label, cells[[paste0(g, "_up")]], cells[[paste0(g, "_down")]],
            nrow = 1, rel_widths = c(0.18, 0.41, 0.41))
})

col_headers <- plot_grid(
  ggdraw(),
  ggdraw() + draw_label("Upstream", size = 12),
  ggdraw() + draw_label("Downstream", size = 12),
  nrow = 1, rel_widths = c(0.18, 0.41, 0.41))

title_row <- ggdraw() +
  draw_label("B", x = 0.06, y = 0.5, size = 16, fontface = "bold") +
  draw_label(expression(italic("Inv4m") ~ "Breakpoints"), x = 0.45, y = 0.5, size = 14)

p_panelb <- plot_grid(
  title_row,
  rows[[1]], rows[[2]], rows[[3]],
  col_headers,
  ncol = 1, rel_heights = c(0.06, 0.26, 0.26, 0.26, 0.06))

p_panelb
```

```{r panel-b-save}
ggsave(file.path(paths$figures, "Fig1_panel_B.png"),
       plot = p_panelb, width = 5.5, height = 8, dpi = 300, bg = "white")
ggsave(file.path(paths$figures, "Fig1_panel_B.svg"),
       plot = p_panelb, width = 5.5, height = 8, device = "svg",
       fix_text_size = TRUE)
```

# Session info

```{r session-info}
sessionInfo()
```
