---
title: "Figure 1 Assembly"
author: "inv4m"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")

library(ggplot2)
library(cowplot)
library(ggtext)
library(magick)
```

# Load pre-built panels

Panels B–D come from `plot_synteny_and_repeats.Rmd`.
Panel E (genotype map) with Panel F (chr4 inset) come from
`plot_genotype_get_correlated_loci.Rmd` as `plot_with_inset`.

```{r load-panels}
p_bcd <- readRDS(file.path(paths$intermediate, "fig1_panel_BCD.rds"))
p_ef  <- readRDS(file.path(paths$intermediate, "fig1_panel_E.rds"))
```

# Panel A (breeding scheme SVG)

Panel A is the breeding scheme diagram created in Inkscape.

```{r panel-a}
panel_a_svg <- file.path(paths$figures, "panelA.svg")
img_a <- magick::image_read_svg(panel_a_svg, width = 18 * 300)
p_a <- ggdraw() + draw_image(img_a)
```

# Panels E + F (pre-built)

`p_ef` is the full bottom row from `plot_genotype_get_correlated_loci.Rmd`:
E = whole-genome genotype map, F = chr4 zoom inset, with spike-calibrated
brackets and panel labels already included.

```{r panel-ef-preview, fig.width=18, fig.height=8}
p_ef
```

# Full Figure 1

```{r figure1-full, fig.width=18, fig.height=19}
fig1_grid <- plot_grid(
  p_a,
  p_bcd,
  p_ef,
  ncol = 1,
  rel_heights = c(0.23, 0.37, 0.40))

# Add A and B labels on the Figure 1 canvas so they share NPC space with E
fig1 <- ggdraw(fig1_grid) +
  draw_label("A", x = 0.005, y = 0.995, size = 45, fontface = "bold",
             hjust = 0, vjust = 1) +
  draw_label("B", x = 0.005, y = 0.76, size = 45, fontface = "bold",
             hjust = 0, vjust = 1)

fig1
```

```{r figure1-save}
ggsave(file.path(paths$figures, "Figure_1.png"),
       plot = fig1, width = 18, height = 19, dpi = 300, bg = "white")
ggsave(file.path(paths$figures, "Figure_1.svg"),
       plot = fig1, width = 18, height = 19, device = "svg",
       fix_text_size = TRUE, bg = "white")
```

# Session info

```{r session-info}
sessionInfo()
```
