---
title: "Chromosomal Knob Repeat Analysis — Figure 1 Panel A"
author: "inv4m"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

BLAST+ megablast hits of chromosomal knob repeats (knob180 M32524.1 and TR-1
AY083942.1) plotted along chromosome 4 of four *Zea* genomes. Vertical lines
mark inversion breakpoint boundaries from AnchorWave genome alignments.

This generates **Figure 1 Panel A** of the inversion paper.

BLAST searches were run with `scripts/02_genomics_foundation/run_repeat_annotation_4genomes.sh`.

# Load BLAST results

```{r load-data}
library(dplyr)
library(ggplot2)
library(scales)

blastcols <- c("qseqid", "sseqid", "pident", "length", "mismatch",
               "gapopen", "qstart", "qend", "sstart", "send",
               "evalue", "bitscore")

# Genome order for facets (top to bottom)
genomes <- c("TIL18", "PT", "Mi21", "B73")

# BLAST result files in intermediate/
knob_files <- setNames(
  file.path(paths$intermediate, paste0("knob180_", genomes, ".blast")),
  genomes
)

tr1_files <- setNames(
  file.path(paths$intermediate, paste0("TR-1_", genomes, ".blast")),
  genomes
)
```

```{r read-blast}
read_blast <- function(files, genomes, rep_label) {
  lapply(genomes, function(g) {
    blast <- read.table(
      files[g], sep = "\t", header = FALSE,
      col.names = blastcols,
      colClasses = c(qseqid = "character", sseqid = "character")
    )
    blast$genome <- g
    blast
  }) %>%
    bind_rows() %>%
    mutate(
      bitscaled = rescale(bitscore, to = c(0, 100)),
      rep = rep_label
    )
}

knob <- read_blast(knob_files, genomes, "knob180")
tr1  <- read_blast(tr1_files, genomes, "TR-1")

to_plot <- bind_rows(knob, tr1) %>%
  mutate(
    genome = factor(genome, levels = genomes),
    rep = factor(rep)
  )
```

```{r hit-summary}
to_plot %>%
  group_by(genome, rep) %>%
  summarise(n_hits = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = rep, values_from = n_hits)
```

# Breakpoint annotations

```{r breakpoints}
# Breakpoint coordinates per genome (from AnchorWave alignments / Table 1)
# Each genome has two vertical lines: upstream and downstream of Inv4m
annotations <- data.frame(
  genome = factor(
    c("TIL18", "TIL18", "PT", "PT", "B73", "B73", "Mi21", "Mi21"),
    levels = genomes
  ),
  xpos = c(
    # TIL18 (inverted karyotype)
    180365316, 193570651,
    # PT (inverted karyotype)
    173486186, 186925483,
    # B73 (standard karyotype)
    172882309, 188131461,
    # Mi21 NIL (inverted karyotype, from AnchorWave anchors 2026-03-07)
    # Upstream boundary gene Zm00001eb190470 → Mi21:167621035
    # Downstream boundary gene Zm00001eb194800 → Mi21:156752504
    156752504, 167621035
  )
)
```

# Figure 1 Panel A

```{r panel-a, fig.width=10, fig.height=10}
p <- to_plot %>%
  ggplot(aes(x = sstart, y = bitscaled, color = rep)) +
  ylab("Normalized Repeat Match Score") +
  geom_vline(
    data = annotations,
    mapping = aes(xintercept = xpos),
    color = "#3232a0",
    linewidth = 1.2
  ) +
  geom_point(size = 0.8) +
  facet_wrap(~ genome, nrow = 4, ncol = 1, scales = "free_x") +
  scale_color_manual(values = c("knob180" = "#1d7f7a", "TR-1" = "gold")) +
  scale_x_continuous(
    labels = unit_format(unit = "", scale = 1e-6),
    breaks = seq(150e6, 275e6, by = 25e6)
  ) +
  annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
  annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
  theme_minimal(base_size = 16) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(hjust = 0.8),
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    legend.position = "none"
  )

p
```

# Save figure

```{r save}
ggsave(
  file.path(paths$figures, "fig1_panel_A_repeat_annotation.pdf"),
  plot = p,
  width = 10, height = 10
)

ggsave(
  file.path(paths$figures, "fig1_panel_A_repeat_annotation.png"),
  plot = p,
  width = 10, height = 10, dpi = 300
)
```

# Session info

```{r session-info}
sessionInfo()
```
