---
title: "Figure 2 draft assembly (2x2 macro layout)"
author: "Fausto Rodriguez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    fig_width: 12
    fig_height: 8.05
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Macro layout (850 x 570 px draft)

```
+-------------------------------+-------------------------------+
|  A  PSU2022 field phenotypes  |  B  PSU2024 significant       |
|     (6-trait, 2 x 3 grid)     |     (4-trait 2 x 2 grid)      |
+-------------------------------+-------------------------------+
|  C  SAM quantification        |  D  SAM micrograph (DIC)      |
|     (4 morphometric panels)   |     (placeholder — Overleaf)  |
+-------------------------------+-------------------------------+
```

Panel A and C are imported as PNGs already saved by their producer notebooks.
Panel B is rebuilt fresh from the PSU2024 corrected-phenotypes CSV restricted
to the four traits significant at FDR < 0.05 in the marginal contrast (GWE,
TGW, TGN, PH). Panel D currently shows a placeholder; the DIC micrograph PNG
lives on Overleaf and would need to be added under `data/` (or pointed at via
an absolute path) to be embedded here.

# 1. Libraries

```{r load-libraries}
library(tidyverse)
library(ggtext)
library(ggfx)
library(ggpubr)
library(ggbeeswarm)
library(cowplot)
library(magick)
```

# 2. Harmonized theme + boxplot helper

All boxplot panels (A, B, C) share this theme. `base_size = 14` matches the
effective rendered size of the legacy 6-trait Panel A image at the 12-inch
canvas width, so panel typography is internally consistent.

```{r theme-helper}
pheno_theme_panel <- theme_classic2(base_size = 14) +
  theme(
    plot.title       = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title.y     = element_text(size = 12),
    axis.title.x     = element_blank(),
    axis.text.x      = element_markdown(size = 12, face = "bold",
                                        color = "black"),
    axis.text.y      = element_text(size = 10),
    legend.position  = "none"
  )

# Single helper: trait_data has columns (genotype, trait, value);
# label_lookup maps trait -> y-axis label; star_lookup maps trait -> sig star
make_boxplot <- function(tr, trait_data, label_lookup, star_lookup) {
  d <- trait_data %>% filter(trait == tr, !is.na(value))
  if (!nrow(d)) return(NULL)
  ymax <- max(d$value, na.rm = TRUE)
  yrng <- diff(range(d$value, na.rm = TRUE))
  star <- if (tr %in% names(star_lookup)) star_lookup[[tr]] else ""
  bracket_df <- tibble(group1 = "CTRL", group2 = "Inv4m",
                       y.position = ymax + 0.08 * yrng,
                       label = star)
  y_lab <- if (tr %in% names(label_lookup)) label_lookup[[tr]] else tr
  ggplot(d, aes(x = genotype, y = value, color = genotype)) +
    geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0) +
    geom_quasirandom(size = 1.4) +
    scale_color_manual(values = c("CTRL" = "gold", "Inv4m" = "purple4")) +
    scale_x_discrete(labels = c("CTRL" = "CTRL", "Inv4m" = "*Inv4m*")) +
    stat_pvalue_manual(bracket_df, tip.length = 0.01, size = 5,
                       bracket.size = 0.6, inherit.aes = FALSE) +
    labs(title = tr, y = y_lab, x = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0.08, 0.20))) +
    pheno_theme_panel
}
```

# 3. Panel A — PSU2022 6-trait phenotypes (rebuilt native)

```{r panel-A, fig.width=7.2, fig.height=5.5}
psu2022 <- read_csv(
  file.path(paths$intermediate,
            "PSU2022_spatially_corrected_phenotypes.csv"),
  show_col_types = FALSE) %>%
  filter(genotype %in% c("CTRL", "Inv4m")) %>%
  mutate(genotype = factor(genotype, levels = c("CTRL", "Inv4m")))

psu2022_eff <- read_csv(
  file.path(paths$intermediate, "PSU2022_inv4m_effect_sizes.csv"),
  show_col_types = FALSE)

A_traits <- c("PH", "DTA", "DTS", "HI", "STWHV", "CD")
A_labels <- c(PH = "Plant height (cm)", DTA = "Days to anthesis",
              DTS = "Days to silking", HI = "Harvest index",
              STWHV = "Stover dry wt (g)", CD = "Cob diameter (mm)")
A_stars  <- setNames(psu2022_eff$p.adj.signif, psu2022_eff$trait)

psu2022_long <- psu2022 %>%
  pivot_longer(cols = any_of(A_traits), names_to = "trait",
               values_to = "value") %>%
  mutate(trait = factor(trait, levels = A_traits))

A_plots <- lapply(A_traits, make_boxplot,
                  trait_data = psu2022_long,
                  label_lookup = A_labels,
                  star_lookup = A_stars)
panel_A <- cowplot::plot_grid(plotlist = A_plots, nrow = 2, ncol = 3,
                              align = "hv", axis = "tblr")
panel_A
```

# 4. Panel B — PSU2024 4 significant traits (rebuilt native)

```{r panel-B, fig.width=4.8, fig.height=5.5}
sp_corrected <- read_csv(
  file.path(paths$intermediate,
            "PSU2024_spatially_corrected_phenotypes.csv"),
  show_col_types = FALSE)
all_contrasts <- read_csv(
  file.path(paths$intermediate, "PSU2024_focal_contrast_all_27.csv"),
  show_col_types = FALSE)
marginal <- all_contrasts %>% filter(contrast_type == "marginal") %>%
  arrange(p_adj)
sig_traits <- marginal %>% filter(p_adj < 0.05) %>% pull(trait)
cat("Significant traits (FDR < 0.05):",
    paste(sig_traits, collapse = ", "), "\n")

B_labels <- c(PH = "Plant height (cm)",
              TGW = "Total grain weight (g)",
              GWE = "Grain weight / ear (g)",
              TGN = "Total grain number")
B_stars <- setNames(marginal$signif, marginal$trait)

focal_long <- sp_corrected %>%
  filter(geno_label %in% c("CTRL", "Inv4m")) %>%
  mutate(genotype = factor(geno_label, levels = c("CTRL", "Inv4m"))) %>%
  pivot_longer(cols = any_of(sig_traits),
               names_to = "trait", values_to = "value") %>%
  filter(trait %in% sig_traits) %>%
  mutate(trait = factor(trait, levels = sig_traits))

B_plots <- lapply(sig_traits, make_boxplot,
                  trait_data = focal_long,
                  label_lookup = B_labels,
                  star_lookup = B_stars)
panel_B <- cowplot::plot_grid(plotlist = B_plots, nrow = 2, ncol = 2,
                              align = "hv", axis = "tblr")
panel_B
```

# 5. Panel C — SAM 4-trait morphometry (rebuilt native)

```{r panel-C, fig.width=7.2, fig.height=2.5}
sam_raw <- read.csv(
  here("data", "MeristemMesurements_B73-INV_December2024.csv")) %>%
  dplyr::select(-starts_with("X")) %>%
  mutate(
    radius       = Width / 2,
    hr_ratio     = Height / radius,
    shape_factor = 1000 * Height / (radius^2),
    genotype = factor(case_when(Genotype == "INV" ~ "Inv4m",
                                Genotype == "B73" ~ "CTRL",
                                TRUE              ~ as.character(Genotype)),
                      levels = c("CTRL", "Inv4m"))
  )

C_traits <- c("Height", "radius", "hr_ratio", "shape_factor")
C_labels <- c(Height       = "Height (μm)",
              radius       = "Radius (μm)",
              hr_ratio     = "h / r",
              shape_factor = expression(h / r^2 ~ "(nm"^{-1}*")"))
C_titles <- c(Height       = "Height",
              radius       = "Radius",
              hr_ratio     = "HR ratio",
              shape_factor = "Shape Factor")

## One-tailed t-tests (Inv4m > CTRL), matching SAM_morphology_analysis.Rmd
sam_one_tailed <- sam_raw %>%
  pivot_longer(cols = all_of(C_traits), names_to = "trait",
               values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(trait) %>%
  summarise(
    p = t.test(value ~ genotype, alternative = "less")$p.value,
    .groups = "drop"
  )

star_from_p <- function(p) {
  if (is.na(p)) return("ns")
  if (p < 0.001) return("***")
  if (p < 0.01)  return("**")
  if (p < 0.05)  return("*")
  "ns"
}
C_stars <- setNames(vapply(sam_one_tailed$p, star_from_p, character(1)),
                    sam_one_tailed$trait)
print(sam_one_tailed)

sam_long <- sam_raw %>%
  pivot_longer(cols = all_of(C_traits),
               names_to = "trait", values_to = "value") %>%
  mutate(trait = factor(trait, levels = C_traits))

make_sam_panel <- function(tr) {
  d <- sam_long %>% filter(trait == tr, !is.na(value))
  ymax <- max(d$value); yrng <- diff(range(d$value))
  bracket_df <- tibble(group1 = "CTRL", group2 = "Inv4m",
                       y.position = ymax + 0.08 * yrng,
                       label = C_stars[[tr]])
  y_lab <- C_labels[[tr]]
  p <- ggplot(d, aes(x = genotype, y = value, color = genotype)) +
    geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0) +
    geom_quasirandom(size = 1.4) +
    scale_color_manual(values = c("CTRL" = "gold", "Inv4m" = "purple4")) +
    scale_x_discrete(labels = c("CTRL" = "CTRL", "Inv4m" = "*Inv4m*")) +
    stat_pvalue_manual(bracket_df, tip.length = 0.01, size = 5,
                       bracket.size = 0.6, inherit.aes = FALSE) +
    labs(title = C_titles[[tr]], x = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0.08, 0.20))) +
    pheno_theme_panel
  if (is.expression(y_lab)) p + ylab(y_lab) else p + ylab(y_lab)
}

C_plots <- lapply(C_traits, make_sam_panel)
panel_C <- cowplot::plot_grid(plotlist = C_plots, nrow = 1, ncol = 4,
                              align = "hv", axis = "tblr")
panel_C
```

# 6. Panel D — SAM DIC micrograph

```{r panel-D, fig.width=4.8, fig.height=2.5}
panel_D <- ggdraw() +
  draw_image(image_read(here("data", "SAM_DIC_micrograph.png")), scale = 1)
panel_D
```

# 6. Assemble 2x2 macro layout at 850 x 570 px

```{r figure-2-draft, fig.width=12, fig.height=8.05}
## Asymmetric grid:
##   - columns: 60/40 (left wider than right; right = 2/3 of left)
##   - rows:    top row ~2.2x taller than bottom row (D is short, height of
##              one panel y-axis in C)
top_row <- cowplot::plot_grid(
  panel_A, panel_B,
  ncol = 2, rel_widths = c(0.6, 0.4),
  labels = c("A", "B"),
  label_size = 18, label_fontface = "bold"
)
bottom_row <- cowplot::plot_grid(
  panel_C, panel_D,
  ncol = 2, rel_widths = c(0.6, 0.4),
  labels = c("C", "D"),
  label_size = 18, label_fontface = "bold"
)
fig2 <- cowplot::plot_grid(
  top_row, bottom_row,
  ncol = 1, rel_heights = c(5.55, 2.5)
)

print(fig2)

ggsave(file.path(paths$figures, "Figure_2_draft.png"),
       fig2, width = 12, height = 8.05, dpi = 300, bg = "white")

cat("Saved:",
    file.path(paths$figures, "Figure_2_draft.png"), "\n")
```

# 7. Notes

- All boxplot panels (A, B, C) rendered native with a shared
  `pheno_theme_panel` (base_size = 14, linewidth = 1.0, dot size = 1.4) so
  typography and stroke weight match.
- Significance stars: Panel A from `PSU2022_inv4m_effect_sizes.csv` (Welch
  t-test, FDR), Panel B from the 27-row FDR pool in
  `PSU2024_focal_contrast_all_27.csv` (marginal contrast row), Panel C
  computed inline as **one-tailed Welch t-tests** (Inv4m > CTRL), matching
  the convention used in `SAM_morphology_analysis.Rmd`.
- Output dimensions: 12 x 8.05 in @ 300 dpi (3600 x 2415 px), aspect ≈ 1.49.
