---
title: "SAM morphometry: estimation statistics with dabestr"
author: "Fausto Rodriguez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    fig_width: 9
    fig_height: 6
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Purpose

Apply estimation statistics (DABEST, Ho et al. 2019) to the same SAM
morphometry data analysed parametrically in `SAM_morphology_analysis.Rmd`.
For each of the four SAM traits (height, radius, $h/r$, $h/r^2$), report the
mean difference Inv4m $-$ CTRL with bootstrap 95% confidence intervals via
the multi two-group estimation idiom (`dabestr`). Effect sizes are not
$p$-thresholded; significance is read from whether the bootstrap CI overlaps
zero.

# 1. Libraries and data

```{r libs}
library(tidyverse)
library(dabestr)
library(cowplot)
library(knitr)

sam_raw <- read.csv(
  here("data", "MeristemMesurements_B73-INV_December2024.csv")) %>%
  dplyr::select(-starts_with("X")) %>%
  mutate(
    radius       = Width / 2,
    hr_ratio     = Height / radius,
    shape_factor = 1000 * Height / (radius^2),
    genotype = factor(case_when(Genotype == "INV" ~ "Inv4m",
                                Genotype == "B73" ~ "CTRL",
                                TRUE              ~ as.character(Genotype)),
                      levels = c("CTRL", "Inv4m"))
  )

cat("rows:", nrow(sam_raw), "\n")
sam_raw %>% count(genotype)
```

# 2. Per-trait dabest analyses

`dabestr::load()` builds a dabest object; `mean_diff()` computes the
unpaired mean difference with 5000 bootstrap resamples (defaults). The
returned object carries the bootstrap distribution, point estimate, and
95% CI.

```{r per-trait-dabest, results = "hide"}
sam_traits <- c("Height", "radius", "hr_ratio", "shape_factor")
trait_units <- c(Height       = "Height (μm)",
                 radius       = "Radius (μm)",
                 hr_ratio     = "h / r",
                 shape_factor = expression(h / r^2 ~ "(nm"^{-1}*")"))

run_dabest <- function(tr) {
  d <- sam_raw %>% dplyr::select(genotype, !!sym(tr)) %>%
    rename(value = !!sym(tr)) %>%
    dplyr::filter(!is.na(value))
  dabest_obj <- dabestr::load(d, x = genotype, y = value,
                              idx = c("CTRL", "Inv4m"))
  mean_diff(dabest_obj)
}

dabest_results <- setNames(lapply(sam_traits, run_dabest), sam_traits)
```

# 3. Effect size summary

```{r summary-table}
summary_tbl <- lapply(sam_traits, function(tr) {
  bs <- dabest_results[[tr]]$boot_result
  tibble(trait        = tr,
         mean_diff    = bs$difference[1],
         ci_lo_95     = bs$bca_ci_low[1],
         ci_hi_95     = bs$bca_ci_high[1],
         ci_overlap_0 = (bs$bca_ci_low[1] <= 0) & (bs$bca_ci_high[1] >= 0))
}) %>% bind_rows()

knitr::kable(summary_tbl, digits = 3,
             caption = paste("Mean differences (Inv4m - CTRL) with",
                             "bootstrap 95% BCa intervals."))

write_csv(summary_tbl,
          file.path(paths$intermediate,
                    "SAM_dabestr_mean_diff_BCa.csv"))
```

# 4. Estimation plots

```{r plot-each, fig.width=8, fig.height=5}
plot_list <- setNames(
  lapply(sam_traits, function(tr) {
    dabest_plot(dabest_results[[tr]],
                raw_marker_alpha = 0.8,
                contrast_label   = paste("Inv4m −\nCTRL"))
  }),
  sam_traits
)

for (tr in sam_traits) {
  cat("\n\n### ", tr, " — ", as.character(trait_units[[tr]]), "\n\n", sep = "")
  print(plot_list[[tr]])
}
```

# 5. Composite grid (all four traits)

```{r grid, fig.width=14, fig.height=10}
grid_plot <- cowplot::plot_grid(
  plot_list$Height,
  plot_list$radius,
  plot_list$hr_ratio,
  plot_list$shape_factor,
  nrow = 2, ncol = 2,
  labels = sam_traits,
  label_size = 14, label_fontface = "bold"
)

print(grid_plot)

ggsave(file.path(paths$figures, "SAM_dabestr_estimation_plots.png"),
       grid_plot, width = 14, height = 10, dpi = 200, bg = "white")
cat("\nSaved:",
    file.path(paths$figures, "SAM_dabestr_estimation_plots.png"), "\n")
```

# 6. Interpretation notes

- Whisker bars in each contrast panel are bias-corrected and accelerated
  (BCa) bootstrap 95% confidence intervals on the mean difference.
- A trait is conventionally "significant at $\alpha = 0.05$" under this
  framework if its 95% CI does not cross zero. This is conceptually
  similar to a two-sided test at $\alpha = 0.05$ but does not require an
  FDR correction across traits since the inference target is the
  per-trait effect size and CI rather than a global null-hypothesis test.
- The CI half-widths report inferential precision and are interpretable
  on the natural scale of each trait.
