---
title: "Sequence Divergence vs Differential Expression in the Inv4m Introgression"
author: "Fausto Rodriguez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

Are the expression differences between Inv4m and B73 haplotypes explained by
per-gene CDS sequence divergence, or do some genes show excess DE beyond what
divergence predicts?

We use LAST alignments (B73 vs Mi21 CDS) already computed by the jcvi ortholog
pipeline. Only genes within the shared introgression on chromosome 4 are
analyzed.

```{r load_libraries}
library(here)
library(tidyverse)
library(ggrepel)
library(GenomicRanges)
library(MASS)      # rlm (robust regression)
library(quantreg)  # rq  (quantile regression)
library(evd)       # GPD fitting for extreme value analysis
library(ggtext)    # richtext labels
library(ggbeeswarm) # beeswarm plots
library(ggfx)      # shadow effects

source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Load data

## LAST alignments (CDS percent identity)

```{r load_last}
last_file <- "/Users/fvrodriguez/mnt/rrellan/DOE_CAREER/inv4m/nilhifimi21/microsynteny/B73.Mi21.last"

last_cols <- c("query", "subject", "pct_identity", "aln_len", "mismatches",
               "gaps", "qstart", "qend", "sstart", "send", "evalue", "score")

last_raw <- read.table(last_file, header = FALSE, sep = "\t",
                       col.names = last_cols, comment.char = "#")

# Strip transcript suffix to get gene IDs
last_raw <- last_raw %>%
  mutate(
    b73_gene = sub("_T\\d+$", "", query),
    mi21_gene = sub("^Mi21_", "", sub("_T\\d+$", "", subject))
  )

cat("Raw LAST alignments:", nrow(last_raw), "\n")
```

## DEG results

```{r load_degs}
degs <- read.csv(file.path(paths$intermediate, "deg_both_models_genotype.csv"))
cat("DEG results:", nrow(degs), "genes\n")
```

## Gene coordinates (B73 v5 GFF)

```{r load_coordinates}
v5_gff_file <- file.path(paths$data, "Zea_mays.Zm-B73-REFERENCE-NAM-5.0.59.chr.gff3")
genes_gr <- rtracklayer::import(v5_gff_file)
genes_gr <- genes_gr[genes_gr$type == "gene" &
                     as.character(seqnames(genes_gr)) %in% as.character(1:10)]
genes <- as.data.frame(genes_gr)
genes$ID <- gsub("gene:", "", genes$ID)
```

# Define regions and filter

```{r define_regions}
# Boundaries
introgression_start <- 157012149
introgression_end   <- 195900523
inv4m_start <- genes$start[genes$ID == "Zm00001eb190470"]
inv4m_end   <- genes$end[genes$ID == "Zm00001eb194800"]

# Chr4 introgression genes
introgression_genes <- genes %>%
  filter(seqnames == 4, start >= introgression_start, end <= introgression_end) %>%
  mutate(location = ifelse(start >= inv4m_start & end <= inv4m_end,
                           "Inv4m", "flanking"))

cat("Introgression genes:", nrow(introgression_genes), "\n")
cat("  Inv4m:", sum(introgression_genes$location == "Inv4m"), "\n")
cat("  CTRL:", sum(introgression_genes$location == "flanking"), "\n")
```

# Best LAST hit per gene in introgression

Keep the highest-scoring ortholog alignment for each B73 gene within the
introgression.

```{r best_hits}
best_hits <- last_raw %>%
  filter(b73_gene %in% introgression_genes$ID) %>%
  # Keep only self-ortholog hits (same gene ID in B73 and Mi21)
  filter(b73_gene == mi21_gene) %>%
  group_by(b73_gene) %>%
  slice_max(score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(gene = b73_gene, pct_identity, aln_len, score)

cat("Genes with ortholog hits:", nrow(best_hits), "\n")
cat("Identity range:", round(min(best_hits$pct_identity), 2), "-",
    round(max(best_hits$pct_identity), 2), "%\n")
```

# Merge datasets

```{r merge}
# Load master locus labels
locus_labels <- read.csv(file.path(paths$data, "locus_labels_master.csv")) %>%
  dplyr::select(gene, locus_label)

merged <- best_hits %>%
  inner_join(degs, by = "gene") %>%
  inner_join(
    introgression_genes %>% dplyr::select(ID, location, start),
    by = c("gene" = "ID")
  ) %>%
  left_join(locus_labels, by = "gene") %>%
  mutate(
    cds_divergence = 100 - pct_identity,
    abs_logFC = abs(logFC),
    direction = ifelse(logFC > 0, "Up in Inv4m", "Down in Inv4m"),
    display_label = coalesce(locus_label, gene)
  )

cat("Merged dataset:", nrow(merged), "genes\n")
cat("  DEGs (FDR < 0.05):", sum(merged$adj.P.Val < 0.05), "\n")
cat("  Up:", sum(merged$direction == "Up in Inv4m"),
    " Down:", sum(merged$direction == "Down in Inv4m"), "\n")
```

# Analysis

## Model selection

```{r models}
fit_null <- lm(abs_logFC ~ 1, data = merged)
fit_div  <- lm(abs_logFC ~ cds_divergence, data = merged)
fit_loc  <- lm(abs_logFC ~ location, data = merged)
fit_add  <- lm(abs_logFC ~ cds_divergence + location, data = merged)
fit_int  <- lm(abs_logFC ~ cds_divergence * location, data = merged)

models <- list(null = fit_null, divergence = fit_div, location = fit_loc,
               additive = fit_add, interaction = fit_int)

data.frame(
  model = names(models),
  df = sapply(models, function(m) length(coef(m))),
  AIC = round(sapply(models, AIC), 1),
  BIC = round(sapply(models, BIC), 1),
  adj_R2 = round(sapply(models, function(m) summary(m)$adj.r.squared), 4)
) %>% arrange(BIC)
```

```{r model_summaries}
cat("=== Additive model ===\n")
summary(fit_add)

cat("\n=== Interaction model ===\n")
summary(fit_int)

anova(fit_add, fit_int)

cat("\n=== ANOVA: location-only vs additive ===\n")
anova(fit_loc, fit_add)
```

## Scatter: |log2FC| vs CDS divergence

```{r scatter_plot, fig.width=8, fig.height=6}
# Use additive model for ANCOVA lines
fit_ancova <- fit_add

# Predicted ANCOVA lines
ancova_pred <- expand.grid(
  cds_divergence = seq(0, max(merged$cds_divergence), length.out = 200),
  location = c("Inv4m", "flanking")
)
ancova_pred$abs_logFC <- predict(fit_ancova, newdata = ancova_pred)

# Residuals for outlier labeling
merged$residual <- residuals(fit_ancova)

# Top outliers by residual magnitude
top_outliers <- merged %>%
  filter(abs(residual) > quantile(abs(residual), 0.95) | abs_logFC > 2) %>%
  mutate(
    fontface = ifelse(grepl("^jmj3$", display_label, ignore.case = TRUE),
                      "bold.italic", "italic")
  )

# Extract stats for annotation
ancova_sum <- summary(fit_ancova)
ancova_coefs <- coef(ancova_sum)
r2 <- round(ancova_sum$adj.r.squared, 3)
p_div <- ancova_coefs["cds_divergence", "Pr(>|t|)"]
p_loc <- ancova_coefs["locationInv4m", "Pr(>|t|)"]

x_max <- ceiling(max(merged$cds_divergence))
y_max <- ceiling(max(merged$abs_logFC))

# Stats annotation (ggtext richtext for italic p)
stat_label <- paste0(
  "Region *p* = ", format(p_loc, digits = 2), "<br>",
  "Divergence *p* = ", format(p_div, digits = 2), "<br>",
  "adj. R\u00B2 = ", r2
)

p_scatter <- ggplot(merged, aes(x = cds_divergence, y = abs_logFC,
                                 color = location)) +
  geom_point(aes(shape = adj.P.Val < 0.05), size = 2, alpha = 0.5) +
  geom_line(data = ancova_pred, linewidth = 1) +
  geom_text_repel(
    data = top_outliers,
    aes(label = display_label, fontface = fontface),
    size = 2.5, max.overlaps = 15, segment.alpha = 0.4,
    show.legend = FALSE
  ) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +
  scale_color_manual(
    values = c("Inv4m" = "purple4", "flanking" = "plum")
  ) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1), guide = "none") +
  geom_richtext(
    data = data.frame(x = 5.1, y = 11.2),
    aes(x = x, y = y), label = stat_label,
    inherit.aes = FALSE, hjust = 0, vjust = 1,
    size = 4.5, label.colour = NA, fill = NA
  ) +
  labs(
    x = "CDS divergence (100 - % identity)",
    y = "|log2 fold change|",
    title = "Expression change vs sequence divergence in the Inv4m introgression"
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = c(0.18, 0.98),
    legend.justification = c(0, 1),
    legend.title = element_blank(),
    legend.text = element_text(size = 13),
    legend.key.size = unit(1.2, "lines"),
    legend.background = element_rect(fill = alpha("white", 0.8), color = NA)
  )

print(p_scatter)
```


```{r models_by_direction}
for (dir in c("Up in Inv4m", "Down in Inv4m")) {
  cat("\n===", dir, "===\n")
  d <- merged %>% filter(direction == dir)
  cat("n =", nrow(d), "\n")
  fit_dir <- lm(logFC ~ cds_divergence + location, data = d)
  print(summary(fit_dir))
}
```

## Robust regression

OLS is sensitive to the few high-divergence, high-|logFC| genes that act as
leverage points. Robust regression (Huber M-estimator) downweights these
automatically.

```{r robust_regression}
fit_rlm_add <- rlm(abs_logFC ~ cds_divergence + location, data = merged)
cat("=== Robust regression (rlm, additive) ===\n")
summary(fit_rlm_add)

# Huber weights: values near 1 = well-behaved, <<1 = downweighted
cat("\nDownweighted genes (Huber weight < 0.5):\n")
merged$huber_w <- fit_rlm_add$w
merged %>%
  filter(huber_w < 0.5) %>%
  dplyr::select(gene, location, cds_divergence, abs_logFC, huber_w) %>%
  arrange(huber_w)
```

```{r compare_coefficients}
coef_table <- data.frame(
  term = names(coef(fit_add)),
  OLS = round(coef(fit_add), 4),
  Robust = round(coef(fit_rlm_add), 4)
)
cat("=== Coefficient comparison ===\n")
print(coef_table, row.names = FALSE)
```

## Binned regression (visual summary)

Bin genes by CDS divergence and regress on bin means. This upweights sparse
high-divergence bins that would otherwise be overwhelmed by the mass of
low-divergence genes. Shown for visual clarity; inferential results come from
the robust regression above.

```{r binned_regression, fig.width=8, fig.height=6}
merged$div_bin <- cut(merged$cds_divergence,
                      breaks = seq(0, ceiling(max(merged$cds_divergence)), by = 1),
                      include.lowest = TRUE)

bin_means <- merged %>%
  group_by(div_bin) %>%
  summarise(
    n = n(),
    mean_div = mean(cds_divergence),
    mean_absFC = mean(abs_logFC),
    se_absFC = sd(abs_logFC) / sqrt(n()),
    .groups = "drop"
  )

bin_means

fit_bin_uw <- lm(mean_absFC ~ mean_div, data = bin_means)
fit_bin_w  <- lm(mean_absFC ~ mean_div, data = bin_means, weights = n)

cat("=== Unweighted bin-mean regression ===\n")
summary(fit_bin_uw)

cat("\n=== Weighted bin-mean regression (weights = n) ===\n")
summary(fit_bin_w)
```

```{r binned_by_location, fig.width=8, fig.height=5.5}
bin_means_loc <- merged %>%
  group_by(div_bin, location) %>%
  summarise(
    n = n(),
    mean_div = mean(cds_divergence),
    mean_absFC = mean(abs_logFC),
    se_absFC = sd(abs_logFC) / sqrt(n()),
    .groups = "drop"
  )

# ANCOVA on bin means (abs_logFC)
fit_bin_loc <- lm(mean_absFC ~ mean_div + location, data = bin_means_loc, weights = n)
cat("=== Binned ANCOVA model (divergence + location) ===\n")
summary(fit_bin_loc)

# Predicted lines from additive model
pred_grid <- expand.grid(
  mean_div = seq(0, max(bin_means_loc$mean_div, na.rm = TRUE), length.out = 100),
  location = c("flanking", "Inv4m")
)
pred_grid$mean_absFC <- predict(fit_bin_loc, newdata = pred_grid)

# Stats for annotation (from binned ANCOVA)
bin_sum <- summary(fit_bin_loc)
bin_coefs <- coef(bin_sum)
bin_r2 <- round(bin_sum$adj.r.squared, 3)
bin_p_div <- bin_coefs["mean_div", "Pr(>|t|)"]
bin_p_loc <- bin_coefs["locationInv4m", "Pr(>|t|)"]

bin_stat_label <- paste0(
  "Region *p* = ", format(bin_p_loc, digits = 2), "<br>",
  "Divergence *p* = ", format(bin_p_div, digits = 2), "<br>",
  "adj. R\u00B2 = ", bin_r2
)

p_ancova <- ggplot(bin_means_loc, aes(x = mean_div, y = mean_absFC,
                                       color = location)) +
  geom_point(aes(size = n), alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_absFC - se_absFC,
                    ymax = mean_absFC + se_absFC),
                width = 0.15, alpha = 0.4) +
  geom_line(data = pred_grid, linewidth = 0.9) +
  geom_richtext(
    data = data.frame(x = 0, y = 8.7),
    aes(x = x, y = y), label = bin_stat_label,
    inherit.aes = FALSE, hjust = 0, vjust = 1,
    size = 4.5, label.colour = NA, fill = NA
  ) +
  scale_color_manual(values = c("Inv4m" = "purple4", "flanking" = "plum"),
                     name = "Region") +
  scale_size_continuous(name = "Genes per bin") +
  labs(
    x = "Mean CDS divergence (%)",
    y = "Mean |log2 fold change|",
    title = "Binned ANCOVA: shared slope (divergence), offset = location"
  ) +
  theme_bw(base_size = 13)

print(p_ancova)
```

## Outlier genes (excess DE beyond divergence)

Genes with large positive residuals have more expression change than expected
from their sequence divergence alone — candidates for trans-regulation or CNV
effects.

```{r outliers}
outliers <- merged %>%
  arrange(desc(abs(residual))) %>%
  dplyr::select(gene, location, pct_identity, cds_divergence,
                logFC, abs_logFC, adj.P.Val, residual) %>%
  head(20)

outliers
```

## Distribution of divergence by region

```{r divergence_boxplot, fig.width=6, fig.height=4}
# Pseudocount for zeros: one log10 unit below the minimum nonzero value
min_nz <- min(merged$cds_divergence[merged$cds_divergence > 0])
pseudo_zero <- min_nz / 10

merged_div_plot <- merged %>%
  mutate(div_plot = ifelse(cds_divergence == 0, pseudo_zero, cds_divergence))

# Wilcoxon test for annotation
wt <- wilcox.test(cds_divergence ~ location, data = merged)

p_beeswarm <- ggplot(merged_div_plot, aes(x = location, y = div_plot, color = location)) +
  geom_boxplot(width = 0.25, linewidth = 1.5, alpha = 0, outlier.shape = NA) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  geom_quasirandom(size = 2) %>%
    with_shadow(colour = "black", x_offset = 0, y_offset = 0, sigma = 1) +
  scale_color_manual(values = c("Inv4m" = "purple4", "flanking" = "plum")) +
  scale_x_discrete(labels = c("Inv4m" = "*Inv4m*", "flanking" = "flanking")) +
  annotate("segment", x = 1, xend = 2, y = max(merged_div_plot$div_plot) * 1.3,
           yend = max(merged_div_plot$div_plot) * 1.3) +
  geom_richtext(
    data = data.frame(x = 1.5, y = max(merged_div_plot$div_plot) * 1.6),
    aes(x = x, y = y),
    label = paste0("Wilcoxon *p* = ", format(wt$p.value, digits = 2)),
    inherit.aes = FALSE, size = 4, label.colour = NA, fill = NA
  ) +
  scale_y_continuous(trans = "log10", expand = expansion(mult = c(0.1, 0.15))) +
  labs(x = NULL, y = "CDS divergence (%)") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "none",
    axis.text.x = element_markdown(size = 13)
  )

print(p_beeswarm)
```

```{r divergence_test}
cat("=== Wilcoxon rank-sum test ===\n")
print(wt)

cat("\n=== Welch t-test on log-transformed divergence ===\n")
merged_div_plot$log_div <- log(merged_div_plot$div_plot)
print(t.test(log_div ~ location, data = merged_div_plot))
```

# Extreme value analysis

## Rationale

The ANCOVA and quantile regressions model the *average* or *median* relationship
between divergence and expression change. But our interest is in the DEGs — the
genes in the *tail* of the distribution that show extreme expression responses.
Extreme value theory (EVT) models this tail behavior directly.

The approach:

1. Regress `logFC ~ cds_divergence` to remove the divergence effect
2. Take `|residual|` to pool both tails (comparable to the `abs_logFC` additive model)
3. Fit a Generalized Pareto Distribution (GPD) to the extreme |residuals|
4. Compare GPD parameters between Inv4m and flanking regions
5. If the scale or shape parameter differs, the inversion produces heavier tails
   (more extreme DE) than divergence alone predicts

## Residuals from divergence model

```{r evt_residuals}
# Robust residuals from divergence-only model, then take absolute value
# to pool both tails — directly comparable to the abs_logFC additive model
merged$robust_resid <- merged$logFC - predict(
  rlm(logFC ~ cds_divergence, data = merged),
  newdata = merged
)
merged$abs_resid <- abs(merged$robust_resid)

cat("Absolute residual summary:\n")
summary(merged$abs_resid)
```

## GPD threshold selection

```{r gpd_threshold}
# Use 75th percentile of |residual| as threshold
thresh <- quantile(merged$abs_resid, 0.75)

cat("Threshold (75th pct of |residual|):", round(thresh, 3), "\n")
cat("Total exceedances:", sum(merged$abs_resid > thresh), "\n")
cat("  Inv4m:", sum(merged$abs_resid[merged$location == "Inv4m"] > thresh), "\n")
cat("  flanking:", sum(merged$abs_resid[merged$location == "flanking"] > thresh), "\n")
```

## GPD fits by location (pooled tails)

```{r gpd_by_location}
gpd_results <- list()

for (loc in c("Inv4m", "flanking")) {
  d <- merged %>% filter(location == loc)
  exc <- d$abs_resid[d$abs_resid > thresh]

  if (length(exc) >= 5) {
    fit <- fpot(d$abs_resid, threshold = thresh, model = "gpd")
    gpd_results[[loc]] <- list(
      location = loc,
      n_exceed = length(exc),
      scale = fit$estimate["scale"],
      shape = fit$estimate["shape"],
      se_scale = fit$std.err["scale"],
      se_shape = fit$std.err["shape"],
      nllh = fit$deviance  # negative log-likelihood
    )
    cat("===", loc, "===\n")
    cat("  Exceedances:", length(exc), "\n")
    cat("  Scale:", round(fit$estimate["scale"], 4),
        "±", round(fit$std.err["scale"], 4), "\n")
    cat("  Shape:", round(fit$estimate["shape"], 4),
        "±", round(fit$std.err["shape"], 4), "\n\n")
  }
}

gpd_df <- bind_rows(lapply(gpd_results, as.data.frame))
gpd_df
```

## Compare tail behavior

```{r gpd_comparison, fig.width=7, fig.height=5}
gpd_long <- gpd_df %>%
  pivot_longer(cols = c(scale, shape), names_to = "parameter", values_to = "estimate") %>%
  mutate(se = ifelse(parameter == "scale", se_scale, se_shape))

ggplot(gpd_long, aes(x = location, y = estimate, color = location)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
                width = 0.15) +
  facet_wrap(~ parameter, scales = "free_y") +
  scale_color_manual(values = c("Inv4m" = "purple4", "flanking" = "plum")) +
  labs(
    x = NULL, y = "GPD parameter estimate",
    title = "Generalized Pareto fits (pooled tails): |residual| by region",
    caption = "Error bars: 95% CI from MLE standard errors"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

## Likelihood ratio test

Compare a model where both regions share GPD parameters vs separate fits.

```{r gpd_lrt}
# Pooled fit (all locations combined)
fit_pooled <- fpot(merged$abs_resid, threshold = thresh, model = "gpd")

# Sum of separate negative log-likelihoods
nllh_separate <- sum(sapply(gpd_results, function(x) x$nllh))
nllh_pooled <- fit_pooled$deviance

# LRT: 2 * (loglik_separate - loglik_pooled), df = 2 (extra scale + shape)
lrt_stat <- 2 * (nllh_pooled - nllh_separate)
lrt_pval <- pchisq(lrt_stat, df = 2, lower.tail = FALSE)

cat("=== Likelihood ratio test: shared vs separate GPD ===\n")
cat("  Pooled nllh:", round(nllh_pooled, 2), "\n")
cat("  Separate nllh:", round(nllh_separate, 2), "\n")
cat("  LRT statistic:", round(lrt_stat, 3), "\n")
cat("  p-value:", format(lrt_pval, digits = 4), "\n")
```

## Exceedance probability by location

```{r exceedance_rate}
exc_table <- merged %>%
  group_by(location) %>%
  summarise(
    n = n(),
    n_extreme = sum(abs_resid > thresh),
    pct_extreme = round(100 * n_extreme / n, 1),
    .groups = "drop"
  )

exc_table

cat("\n=== Fisher's test: exceedance rate by location ===\n")
mat <- matrix(c(
  exc_table$n_extreme[exc_table$location == "Inv4m"],
  exc_table$n[exc_table$location == "Inv4m"] - exc_table$n_extreme[exc_table$location == "Inv4m"],
  exc_table$n_extreme[exc_table$location == "flanking"],
  exc_table$n[exc_table$location == "flanking"] - exc_table$n_extreme[exc_table$location == "flanking"]
), nrow = 2, dimnames = list(c("extreme", "not"), c("Inv4m", "flanking")))
print(fisher.test(mat))
```

## QQ plots

```{r gpd_qq, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

for (loc in c("Inv4m", "flanking")) {
  d <- merged %>% filter(location == loc)
  if (sum(d$abs_resid > thresh) >= 5) {
    fit <- fpot(d$abs_resid, threshold = thresh, model = "gpd")
    plot(fit, which = 4, main = paste(loc, "- GPD QQ (pooled tails)"))
  }
}
```

# Save outputs

```{r save}
write.csv(merged, file.path(paths$intermediate, "sequence_divergence_vs_DE.csv"),
          row.names = FALSE)

ggsave(file.path(paths$figures, "sequence_divergence_vs_DE.pdf"),
       p_scatter, width = 8, height = 6)
ggsave(file.path(paths$figures, "sequence_divergence_vs_DE.svg"),
       p_scatter, width = 8, height = 6, fix_text_size = TRUE)

cat("Outputs saved to:", paths$intermediate, "and", paths$figures, "\n")
```

## Multipanel figure

```{r multipanel, fig.width=14, fig.height=6}
library(cowplot)
library(gridExtra)

# Remove title from plots for the assembled figure
p_scatter_notitle <- p_scatter + labs(title = NULL)
p_beeswarm_notitle <- p_beeswarm + labs(title = NULL)

# Assemble: beeswarm (A) | scatter (B)
p_combined <- plot_grid(
  p_beeswarm_notitle,
  p_scatter_notitle,
  nrow = 1,
  rel_widths = c(0.3, 0.7),
  labels = c("A", "B"),
  label_size = 18
)

print(p_combined)

ggsave(file.path(paths$figures, "sequence_divergence_multipanel.pdf"),
       p_combined, width = 14, height = 6)
ggsave(file.path(paths$figures, "sequence_divergence_multipanel.svg"),
       p_combined, width = 14, height = 6, fix_text_size = TRUE)
```

# Zero-divergence DEGs and flanking enrichment

```{r zero_div_fisher}
# Genes with 0% CDS divergence
n_zero_inv4m <- sum(merged$cds_divergence == 0 & merged$location == "Inv4m")
n_zero_flank <- sum(merged$cds_divergence == 0 & merged$location == "flanking")
n_nonzero_inv4m <- sum(merged$cds_divergence > 0 & merged$location == "Inv4m")
n_nonzero_flank <- sum(merged$cds_divergence > 0 & merged$location == "flanking")

mat_zero <- matrix(
  c(n_zero_inv4m, n_nonzero_inv4m,
    n_zero_flank, n_nonzero_flank),
  nrow = 2,
  dimnames = list(c("zero_div", "nonzero_div"), c("Inv4m", "flanking"))
)

cat("Contingency table:\n")
print(mat_zero)

ft_zero <- fisher.test(mat_zero)
cat("\nFisher's exact test: zero-divergence genes by location\n")
print(ft_zero)

cat("\nAll", n_zero_inv4m + n_zero_flank, "zero-divergence genes are DEGs (FDR < 0.05)\n")
cat("  Inv4m:", n_zero_inv4m, "\n")
cat("  flanking:", n_zero_flank, "\n")
```

# Results paragraph and figure caption

```{r results_values, include=FALSE}
# Precompute values for LaTeX
med_inv4m <- round(median(merged$cds_divergence[merged$location == "Inv4m"]), 2)
med_flank <- round(median(merged$cds_divergence[merged$location == "flanking"]), 2)
wt_p <- format(wt$p.value, digits = 2)
add_r2 <- round(summary(fit_add)$adj.r.squared, 3)
add_r2_pct <- round(summary(fit_add)$adj.r.squared * 100, 1)
add_p_div <- format(coef(summary(fit_add))[2,4], digits = 2)
add_p_loc <- format(coef(summary(fit_add))[3,4], digits = 2)
n_zero <- n_zero_inv4m + n_zero_flank
ft_p <- format(ft_zero$p.value, digits = 2)
ft_or <- round(ft_zero$estimate, 1)
n_inv4m <- sum(merged$location == "Inv4m")
n_flank <- sum(merged$location == "flanking")
```

```{r results_paragraph, echo=FALSE, results='asis'}
results_latex <- paste0(
'To test whether differential expression in the introgression is explained by
coding sequence divergence between the B73 and Mi21 haplotypes, we computed
CDS percent identity from LAST pairwise alignments for ', nrow(merged), ' expressed
genes with ortholog hits in the shared introgression.
Genes inside the inversion had significantly higher CDS divergence than flanking
genes (median ', med_inv4m, '\\% vs ', med_flank, '\\%;
Wilcoxon $p$ = ', wt_p, '; Fig.~\\ref{fig::divergence} A), consistent with
suppressed recombination preserving ancestral \\textit{mexicana} divergence within the
inversion.
An additive linear model ($|\\text{log}_2\\text{FC}| \\sim \\text{CDS divergence} + \\text{region}$)
explained ', add_r2_pct, '\\% of the variance
in expression change (adj.\\ $R^2$ = ', add_r2, ';
divergence $p$ = ', add_p_div, ';
region $p$ = ', add_p_loc, '; Fig.~\\ref{fig::divergence} B).
After accounting for divergence, genes inside the inversion showed smaller
absolute expression changes than flanking genes, indicating that the inversion
dampens rather than amplifies expression divergence.
Notably, ', n_zero, ' genes had identical CDS between B73 and
Mi21, and all were significantly differentially expressed (FDR $<$ 0.05).
Of these, ', n_zero_flank, ' of ', n_zero, ' were located in the
flanking region, a significant overrepresentation
(Fisher\'s exact test, $p$ = ', ft_p, ',
OR = ', ft_or, ').
Because these genes lack coding sequence divergence, their differential expression
is most parsimoniously attributed to \\textit{trans}-regulatory effects originating from
divergent loci elsewhere in the introgression.')

caption_latex <- paste0(
'\\begin{figure}[!ht]
\\includegraphics[width=\\linewidth]{figs/sequence_divergence_multipanel.pdf}
\\caption{\\textbf{CDS sequence divergence and expression change in the \\invfour introgression.}
\\textbf{(A)} Distribution of CDS divergence (100 $-$ \\% identity) for expressed genes inside the inversion (\\invfour, $n$ = ', n_inv4m, ') and in the flanking introgression (flanking, $n$ = ', n_flank, ').
Points show individual genes (beeswarm); boxes show median and interquartile range.
$y$-axis is log$_{10}$-scaled.
Wilcoxon rank-sum $p$ = ', wt_p, '.
\\textbf{(B)} Absolute log$_2$ fold change vs CDS divergence.
Parallel lines show the additive ANCOVA model (shared slope for divergence effect, vertical offset for region effect).
Region $p$ = ', add_p_loc, ';
divergence $p$ = ', add_p_div, ';
adj.\\ $R^2$ = ', add_r2, '.}
\\label{fig::divergence}
\\end{figure}')

cat("\n\n````\nRESULTS PARAGRAPH\n\n")
cat(results_latex)
cat("\n````\n\n")
# Build model comparison table LaTeX
bic_df <- data.frame(
  model = names(models),
  df = sapply(models, function(m) length(coef(m))),
  AIC = round(sapply(models, AIC), 1),
  BIC = round(sapply(models, BIC), 1),
  adj_R2 = round(sapply(models, function(m) summary(m)$adj.r.squared), 4),
  stringsAsFactors = FALSE
) %>% arrange(BIC)

# Format model names for LaTeX
model_labels <- c(
  null = "Null (intercept only)",
  divergence = "CDS divergence",
  location = "Region",
  additive = "CDS divergence + region",
  interaction = "CDS divergence $\\times$ region"
)

table_rows <- paste0(
  sapply(seq_len(nrow(bic_df)), function(i) {
    paste0(model_labels[bic_df$model[i]], " & ",
           bic_df$df[i], " & ",
           bic_df$AIC[i], " & ",
           bic_df$BIC[i], " & ",
           bic_df$adj_R2[i])
  }),
  collapse = " \\\\\n"
)

table_latex <- paste0(
'\\begin{table}[!ht]
\\caption{\\textbf{Model comparison for the relationship between CDS divergence and expression change in the \\invfour introgression.}
Five linear models predicting $|\\text{log}_2\\text{FC}|$ were compared by BIC and AIC.
The additive model (CDS divergence + region) provided the best fit, with the interaction term adding no improvement.
Region distinguishes genes inside \\invfour from the flanking introgression.}
\\label{tab::divergence_models}
\\centering
\\begin{tabular}{@{}lcccc@{}}
\\toprule
\\textbf{Model} & \\textbf{df} & \\textbf{AIC} & \\textbf{BIC} & \\textbf{adj.\\ $R^2$} \\\\ \\midrule
', table_rows, ' \\\\ \\bottomrule
\\end{tabular}
\\end{table}')

cat("````\nFIGURE FLOAT\n\n")
cat(caption_latex)
cat("\n````\n\n")
cat("````\nMODEL COMPARISON TABLE\n\n")
cat(table_latex)
cat("\n````\n\n")
```

# Session info

```{r session_info}
sessionInfo()
```
