---
title: "Sequence Divergence vs Differential Expression in the Inv4m Introgression"
author: "Fausto Rodriguez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

Are the expression differences between Inv4m and B73 haplotypes explained by
per-gene CDS sequence divergence, or do some genes show excess DE beyond what
divergence predicts?

We use LAST alignments (B73 vs Mi21 CDS) already computed by the jcvi ortholog
pipeline. Only genes within the shared introgression on chromosome 4 are
analyzed.

```{r load_libraries}
library(here)
library(tidyverse)
library(ggrepel)
library(GenomicRanges)
library(MASS)      # rlm (robust regression)
library(quantreg)  # rq  (quantile regression)
library(evd)       # GPD fitting for extreme value analysis
library(ggtext)    # richtext labels

source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Load data

## LAST alignments (CDS percent identity)

```{r load_last}
last_file <- "/Users/fvrodriguez/mnt/rrellan/DOE_CAREER/inv4m/nilhifimi21/microsynteny/B73.Mi21.last"

last_cols <- c("query", "subject", "pct_identity", "aln_len", "mismatches",
               "gaps", "qstart", "qend", "sstart", "send", "evalue", "score")

last_raw <- read.table(last_file, header = FALSE, sep = "\t",
                       col.names = last_cols, comment.char = "#")

# Strip transcript suffix to get gene IDs
last_raw <- last_raw %>%
  mutate(
    b73_gene = sub("_T\\d+$", "", query),
    mi21_gene = sub("^Mi21_", "", sub("_T\\d+$", "", subject))
  )

cat("Raw LAST alignments:", nrow(last_raw), "\n")
```

## DEG results

```{r load_degs}
degs <- read.csv(file.path(paths$intermediate, "deg_both_models_genotype.csv"))
cat("DEG results:", nrow(degs), "genes\n")
```

## Gene coordinates (B73 v5 GFF)

```{r load_coordinates}
v5_gff_file <- file.path(paths$data, "Zea_mays.Zm-B73-REFERENCE-NAM-5.0.59.chr.gff3")
genes_gr <- rtracklayer::import(v5_gff_file)
genes_gr <- genes_gr[genes_gr$type == "gene" &
                     as.character(seqnames(genes_gr)) %in% as.character(1:10)]
genes <- as.data.frame(genes_gr)
genes$ID <- gsub("gene:", "", genes$ID)
```

# Define regions and filter

```{r define_regions}
# Boundaries
introgression_start <- 157012149
introgression_end   <- 195900523
inv4m_start <- genes$start[genes$ID == "Zm00001eb190470"]
inv4m_end   <- genes$end[genes$ID == "Zm00001eb194800"]

# Chr4 introgression genes
introgression_genes <- genes %>%
  filter(seqnames == 4, start >= introgression_start, end <= introgression_end) %>%
  mutate(location = ifelse(start >= inv4m_start & end <= inv4m_end,
                           "Inv4m", "flanking"))

cat("Introgression genes:", nrow(introgression_genes), "\n")
cat("  Inv4m:", sum(introgression_genes$location == "Inv4m"), "\n")
cat("  CTRL:", sum(introgression_genes$location == "flanking"), "\n")
```

# Best LAST hit per gene in introgression

Keep the highest-scoring ortholog alignment for each B73 gene within the
introgression.

```{r best_hits}
best_hits <- last_raw %>%
  filter(b73_gene %in% introgression_genes$ID) %>%
  # Keep only self-ortholog hits (same gene ID in B73 and Mi21)
  filter(b73_gene == mi21_gene) %>%
  group_by(b73_gene) %>%
  slice_max(score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(gene = b73_gene, pct_identity, aln_len, score)

cat("Genes with ortholog hits:", nrow(best_hits), "\n")
cat("Identity range:", round(min(best_hits$pct_identity), 2), "-",
    round(max(best_hits$pct_identity), 2), "%\n")
```

# Merge datasets

```{r merge}
# Load master locus labels
locus_labels <- read.csv(file.path(paths$data, "locus_labels_master.csv")) %>%
  dplyr::select(gene, locus_label)

merged <- best_hits %>%
  inner_join(degs, by = "gene") %>%
  inner_join(
    introgression_genes %>% dplyr::select(ID, location, start),
    by = c("gene" = "ID")
  ) %>%
  left_join(locus_labels, by = "gene") %>%
  mutate(
    cds_divergence = 100 - pct_identity,
    abs_logFC = abs(logFC),
    direction = ifelse(logFC > 0, "Up in Inv4m", "Down in Inv4m"),
    display_label = coalesce(locus_label, gene)
  )

cat("Merged dataset:", nrow(merged), "genes\n")
cat("  DEGs (FDR < 0.05):", sum(merged$adj.P.Val < 0.05), "\n")
cat("  Up:", sum(merged$direction == "Up in Inv4m"),
    " Down:", sum(merged$direction == "Down in Inv4m"), "\n")
```

# Analysis

## Scatter: |log2FC| vs CDS divergence

```{r scatter_plot, fig.width=8, fig.height=6}
# ANCOVA fit: parallel slopes by location
fit_ancova <- lm(abs_logFC ~ cds_divergence + location, data = merged)

# Predicted ANCOVA lines
ancova_pred <- expand.grid(
  cds_divergence = seq(0, max(merged$cds_divergence), length.out = 100),
  location = c("Inv4m", "flanking")
)
ancova_pred$abs_logFC <- predict(fit_ancova, newdata = ancova_pred)

# Residuals for outlier labeling
merged$residual <- residuals(fit_ancova)

# Top outliers by residual magnitude
top_outliers <- merged %>%
  filter(abs(residual) > quantile(abs(residual), 0.95) | abs_logFC > 2) %>%
  mutate(
    fontface = ifelse(grepl("^jmj3$", display_label, ignore.case = TRUE),
                      "bold.italic", "italic")
  )

# Extract stats for legend labels
ancova_sum <- summary(fit_ancova)
ancova_coefs <- coef(ancova_sum)
r2 <- round(ancova_sum$adj.r.squared, 3)
p_div <- ancova_coefs["cds_divergence", "Pr(>|t|)"]
p_loc <- ancova_coefs["locationInv4m", "Pr(>|t|)"]

# Fixed coordinate range
x_max <- ceiling(max(merged$cds_divergence))
y_max <- ceiling(max(merged$abs_logFC))

# Stats annotation (ggtext richtext for italic p)
stat_label <- paste0(
  "Region *p* = ", format(p_loc, digits = 2), "<br>",
  "Divergence *p* = ", format(p_div, digits = 2), "<br>",
  "adj. R\u00B2 = ", r2
)

p_scatter <- ggplot(merged, aes(x = cds_divergence, y = abs_logFC,
                                 color = location)) +
  geom_point(aes(shape = adj.P.Val < 0.05), size = 2, alpha = 0.5) +
  geom_line(data = ancova_pred, linewidth = 1) +
  geom_text_repel(
    data = top_outliers,
    aes(label = display_label, fontface = fontface),
    size = 2.5, max.overlaps = 15, segment.alpha = 0.4,
    show.legend = FALSE
  ) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +
  scale_color_manual(
    values = c("Inv4m" = "purple4", "flanking" = "plum")
  ) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1), guide = "none") +
  geom_richtext(
    data = data.frame(x = 5.1, y = 11.2),
    aes(x = x, y = y), label = stat_label,
    inherit.aes = FALSE, hjust = 0, vjust = 1,
    size = 4.5, label.colour = NA, fill = NA
  ) +
  labs(
    x = "CDS divergence (100 - % identity)",
    y = "|log2 fold change|",
    title = "Expression change vs sequence divergence in the Inv4m introgression"
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = c(0.18, 0.98),
    legend.justification = c(0, 1),
    legend.title = element_blank(),
    legend.text = element_text(size = 13),
    legend.key.size = unit(1.2, "lines"),
    legend.background = element_rect(fill = alpha("white", 0.8), color = NA)
  )

print(p_scatter)
```

## Linear models

```{r models}
fit_null <- lm(abs_logFC ~ 1, data = merged)
fit_div  <- lm(abs_logFC ~ cds_divergence, data = merged)
fit_loc  <- lm(abs_logFC ~ location, data = merged)
fit_add  <- lm(abs_logFC ~ cds_divergence + location, data = merged)
fit_int  <- lm(abs_logFC ~ cds_divergence * location, data = merged)

models <- list(null = fit_null, divergence = fit_div, location = fit_loc,
               additive = fit_add, interaction = fit_int)

data.frame(
  model = names(models),
  df = sapply(models, function(m) length(coef(m))),
  AIC = round(sapply(models, AIC), 1),
  BIC = round(sapply(models, BIC), 1),
  adj_R2 = round(sapply(models, function(m) summary(m)$adj.r.squared), 4)
) %>% arrange(BIC)
```

```{r model_summaries}
cat("=== Additive model ===\n")
summary(fit_add)

cat("\n=== Interaction model ===\n")
summary(fit_int)

anova(fit_add, fit_int)
```

```{r models_by_direction}
for (dir in c("Up in Inv4m", "Down in Inv4m")) {
  cat("\n===", dir, "===\n")
  d <- merged %>% filter(direction == dir)
  cat("n =", nrow(d), "\n")
  fit_dir <- lm(logFC ~ cds_divergence + location, data = d)
  print(summary(fit_dir))
}
```

## Robust and quantile regression

OLS is sensitive to the few high-divergence, high-|logFC| genes that act as
leverage points. Robust regression (Huber M-estimator) downweights these
automatically. Quantile regression at the median provides a complementary
check that does not assume symmetric errors.

```{r robust_regression}
fit_rlm_add <- rlm(abs_logFC ~ cds_divergence + location, data = merged)
cat("=== Robust regression (rlm, additive) ===\n")
summary(fit_rlm_add)

# Huber weights: values near 1 = well-behaved, <<1 = downweighted
cat("\nDownweighted genes (Huber weight < 0.5):\n")
merged$huber_w <- fit_rlm_add$w
merged %>%
  filter(huber_w < 0.5) %>%
  dplyr::select(gene, location, cds_divergence, abs_logFC, huber_w) %>%
  arrange(huber_w)
```

```{r quantile_regression}
fit_rq_add <- rq(abs_logFC ~ cds_divergence + location, data = merged, tau = 0.5)
cat("=== Quantile regression (median, additive) ===\n")
summary(fit_rq_add, se = "boot")
```

```{r compare_coefficients}
coef_table <- data.frame(
  term = names(coef(fit_add)),
  OLS = round(coef(fit_add), 4),
  Robust = round(coef(fit_rlm_add), 4),
  Quantile = round(coef(fit_rq_add), 4)
)
cat("=== Coefficient comparison ===\n")
print(coef_table, row.names = FALSE)
```

## Binned regression (visual summary)

Bin genes by CDS divergence and regress on bin means. This upweights sparse
high-divergence bins that would otherwise be overwhelmed by the mass of
low-divergence genes. Shown for visual clarity; inferential results come from
the robust regression above.

```{r binned_regression, fig.width=8, fig.height=6}
merged$div_bin <- cut(merged$cds_divergence,
                      breaks = seq(0, ceiling(max(merged$cds_divergence)), by = 1),
                      include.lowest = TRUE)

bin_means <- merged %>%
  group_by(div_bin) %>%
  summarise(
    n = n(),
    mean_div = mean(cds_divergence),
    mean_absFC = mean(abs_logFC),
    se_absFC = sd(abs_logFC) / sqrt(n()),
    .groups = "drop"
  )

bin_means

fit_bin_uw <- lm(mean_absFC ~ mean_div, data = bin_means)
fit_bin_w  <- lm(mean_absFC ~ mean_div, data = bin_means, weights = n)

cat("=== Unweighted bin-mean regression ===\n")
summary(fit_bin_uw)

cat("\n=== Weighted bin-mean regression (weights = n) ===\n")
summary(fit_bin_w)
```

```{r binned_by_location_direction, fig.width=10, fig.height=5.5}
bin_means_loc <- merged %>%
  group_by(div_bin, location, direction) %>%
  summarise(
    n = n(),
    mean_div = mean(cds_divergence),
    mean_logFC = mean(logFC),
    se_logFC = sd(logFC) / sqrt(n()),
    .groups = "drop"
  )

# Fit separate ANCOVA per direction (on signed logFC)
ancova_fits <- bin_means_loc %>%
  group_by(direction) %>%
  group_map(~ {
    fit <- lm(mean_logFC ~ mean_div + location, data = .x, weights = n)
    cat("===", .y$direction, "===\n")
    print(summary(fit))
    cat("\n")

    pred <- expand.grid(
      mean_div = seq(0, max(.x$mean_div, na.rm = TRUE), length.out = 100),
      location = c("flanking", "Inv4m")
    )
    pred$mean_logFC <- predict(fit, newdata = pred)
    pred$direction <- .y$direction
    pred
  }, .keep = TRUE) %>%
  bind_rows()

p_ancova <- ggplot(bin_means_loc, aes(x = mean_div, y = mean_logFC,
                                       color = location)) +
  geom_point(aes(size = n), alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_logFC - se_logFC,
                    ymax = mean_logFC + se_logFC),
                width = 0.15, alpha = 0.4) +
  geom_line(data = ancova_fits, linewidth = 0.9) +
  facet_wrap(~ direction) +
  scale_color_manual(values = c("Inv4m" = "purple4", "flanking" = "plum"),
                     name = "Region") +
  scale_size_continuous(name = "Genes per bin") +
  labs(
    x = "Mean CDS divergence (%)",
    y = "Mean log2 fold change",
    title = "ANCOVA by direction: shared slope (divergence), offset = location"
  ) +
  theme_bw(base_size = 13)

print(p_ancova)
```

## Quantile regression on raw data (no binning)

Raw scatter with median regression lines per location, split by direction.
No binning or aggregation — all 254 points shown.

```{r quantile_by_direction, fig.width=10, fig.height=5.5}
# Fit quantile regression (median) per direction
rq_fits <- merged %>%
  group_by(direction) %>%
  group_map(~ {
    fit <- rq(logFC ~ cds_divergence + location, data = .x, tau = 0.5)
    cat("===", .y$direction, "===\n")
    print(summary(fit, se = "boot"))
    cat("\n")

    pred <- expand.grid(
      cds_divergence = seq(0, max(.x$cds_divergence), length.out = 100),
      location = c("flanking", "Inv4m")
    )
    pred$logFC <- predict(fit, newdata = pred)
    pred$direction <- .y$direction
    pred
  }, .keep = TRUE) %>%
  bind_rows()

p_qr <- ggplot(merged, aes(x = cds_divergence, y = logFC, color = location)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_line(data = rq_fits, linewidth = 1) +
  facet_wrap(~ direction) +
  scale_color_manual(values = c("Inv4m" = "purple4", "flanking" = "plum"),
                     name = "Region") +
  labs(
    x = "CDS divergence (100 - % identity)",
    y = "log2 fold change",
    title = "Quantile regression (median) on raw data, by direction"
  ) +
  theme_bw(base_size = 13)

print(p_qr)
```

## Outlier genes (excess DE beyond divergence)

Genes with large positive residuals have more expression change than expected
from their sequence divergence alone — candidates for trans-regulation or CNV
effects.

```{r outliers}
outliers <- merged %>%
  arrange(desc(abs(residual))) %>%
  dplyr::select(gene, location, pct_identity, cds_divergence,
                logFC, abs_logFC, adj.P.Val, residual) %>%
  head(20)

outliers
```

## Distribution of divergence by region

```{r divergence_boxplot, fig.width=6, fig.height=4}
ggplot(merged, aes(x = location, y = cds_divergence, fill = location)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_manual(values = c("Inv4m" = "purple4", "flanking" = "plum")) +
  labs(x = NULL, y = "CDS divergence (%)") +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

```{r divergence_test}
wilcox.test(cds_divergence ~ location, data = merged)
```

# Extreme value analysis

## Rationale

The ANCOVA and quantile regressions model the *average* or *median* relationship
between divergence and expression change. But our interest is in the DEGs — the
genes in the *tail* of the distribution that show extreme expression responses.
Extreme value theory (EVT) models this tail behavior directly.

The approach:

1. Regress `logFC ~ cds_divergence` to remove the divergence effect
2. Take `|residual|` to pool both tails (comparable to the `abs_logFC` additive model)
3. Fit a Generalized Pareto Distribution (GPD) to the extreme |residuals|
4. Compare GPD parameters between Inv4m and flanking regions
5. If the scale or shape parameter differs, the inversion produces heavier tails
   (more extreme DE) than divergence alone predicts

## Residuals from divergence model

```{r evt_residuals}
# Robust residuals from divergence-only model, then take absolute value
# to pool both tails — directly comparable to the abs_logFC additive model
merged$robust_resid <- merged$logFC - predict(
  rlm(logFC ~ cds_divergence, data = merged),
  newdata = merged
)
merged$abs_resid <- abs(merged$robust_resid)

cat("Absolute residual summary:\n")
summary(merged$abs_resid)
```

## GPD threshold selection

```{r gpd_threshold}
# Use 75th percentile of |residual| as threshold
thresh <- quantile(merged$abs_resid, 0.75)

cat("Threshold (75th pct of |residual|):", round(thresh, 3), "\n")
cat("Total exceedances:", sum(merged$abs_resid > thresh), "\n")
cat("  Inv4m:", sum(merged$abs_resid[merged$location == "Inv4m"] > thresh), "\n")
cat("  flanking:", sum(merged$abs_resid[merged$location == "flanking"] > thresh), "\n")
```

## GPD fits by location (pooled tails)

```{r gpd_by_location}
gpd_results <- list()

for (loc in c("Inv4m", "flanking")) {
  d <- merged %>% filter(location == loc)
  exc <- d$abs_resid[d$abs_resid > thresh]

  if (length(exc) >= 5) {
    fit <- fpot(d$abs_resid, threshold = thresh, model = "gpd")
    gpd_results[[loc]] <- list(
      location = loc,
      n_exceed = length(exc),
      scale = fit$estimate["scale"],
      shape = fit$estimate["shape"],
      se_scale = fit$std.err["scale"],
      se_shape = fit$std.err["shape"],
      nllh = fit$deviance  # negative log-likelihood
    )
    cat("===", loc, "===\n")
    cat("  Exceedances:", length(exc), "\n")
    cat("  Scale:", round(fit$estimate["scale"], 4),
        "±", round(fit$std.err["scale"], 4), "\n")
    cat("  Shape:", round(fit$estimate["shape"], 4),
        "±", round(fit$std.err["shape"], 4), "\n\n")
  }
}

gpd_df <- bind_rows(lapply(gpd_results, as.data.frame))
gpd_df
```

## Compare tail behavior

```{r gpd_comparison, fig.width=7, fig.height=5}
gpd_long <- gpd_df %>%
  pivot_longer(cols = c(scale, shape), names_to = "parameter", values_to = "estimate") %>%
  mutate(se = ifelse(parameter == "scale", se_scale, se_shape))

ggplot(gpd_long, aes(x = location, y = estimate, color = location)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
                width = 0.15) +
  facet_wrap(~ parameter, scales = "free_y") +
  scale_color_manual(values = c("Inv4m" = "purple4", "flanking" = "plum")) +
  labs(
    x = NULL, y = "GPD parameter estimate",
    title = "Generalized Pareto fits (pooled tails): |residual| by region",
    caption = "Error bars: 95% CI from MLE standard errors"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

## Likelihood ratio test

Compare a model where both regions share GPD parameters vs separate fits.

```{r gpd_lrt}
# Pooled fit (all locations combined)
fit_pooled <- fpot(merged$abs_resid, threshold = thresh, model = "gpd")

# Sum of separate negative log-likelihoods
nllh_separate <- sum(sapply(gpd_results, function(x) x$nllh))
nllh_pooled <- fit_pooled$deviance

# LRT: 2 * (loglik_separate - loglik_pooled), df = 2 (extra scale + shape)
lrt_stat <- 2 * (nllh_pooled - nllh_separate)
lrt_pval <- pchisq(lrt_stat, df = 2, lower.tail = FALSE)

cat("=== Likelihood ratio test: shared vs separate GPD ===\n")
cat("  Pooled nllh:", round(nllh_pooled, 2), "\n")
cat("  Separate nllh:", round(nllh_separate, 2), "\n")
cat("  LRT statistic:", round(lrt_stat, 3), "\n")
cat("  p-value:", format(lrt_pval, digits = 4), "\n")
```

## Exceedance probability by location

```{r exceedance_rate}
exc_table <- merged %>%
  group_by(location) %>%
  summarise(
    n = n(),
    n_extreme = sum(abs_resid > thresh),
    pct_extreme = round(100 * n_extreme / n, 1),
    .groups = "drop"
  )

exc_table

cat("\n=== Fisher's test: exceedance rate by location ===\n")
mat <- matrix(c(
  exc_table$n_extreme[exc_table$location == "Inv4m"],
  exc_table$n[exc_table$location == "Inv4m"] - exc_table$n_extreme[exc_table$location == "Inv4m"],
  exc_table$n_extreme[exc_table$location == "flanking"],
  exc_table$n[exc_table$location == "flanking"] - exc_table$n_extreme[exc_table$location == "flanking"]
), nrow = 2, dimnames = list(c("extreme", "not"), c("Inv4m", "flanking")))
print(fisher.test(mat))
```

## QQ plots

```{r gpd_qq, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

for (loc in c("Inv4m", "flanking")) {
  d <- merged %>% filter(location == loc)
  if (sum(d$abs_resid > thresh) >= 5) {
    fit <- fpot(d$abs_resid, threshold = thresh, model = "gpd")
    plot(fit, which = 4, main = paste(loc, "- GPD QQ (pooled tails)"))
  }
}
```

# Save outputs

```{r save}
write.csv(merged, file.path(paths$intermediate, "sequence_divergence_vs_DE.csv"),
          row.names = FALSE)

ggsave(file.path(paths$figures, "sequence_divergence_vs_DE.pdf"),
       p_scatter, width = 8, height = 6)
ggsave(file.path(paths$figures, "sequence_divergence_vs_DE.svg"),
       p_scatter, width = 8, height = 6, fix_text_size = TRUE)

cat("Outputs saved to:", paths$intermediate, "and", paths$figures, "\n")
```

# Session info

```{r session_info}
sessionInfo()
```
