---
title: "MaizeNetome Reference Network Disruption by Inv4m"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Question**: Does Inv4m introgression disrupt established coexpression networks?

**Approach**: We test whether MaizeNetome reference edges (known coexpression
relationships) are preserved in our WGCNA gene set when comparing CTRL vs Inv4m
genotypes. This is a MaizeNetome-centric analysis: we start with reference edges
and ask what fraction are disrupted.

**Key metric**: Of MaizeNetome edges detected in CTRL, what percentage are lost
in Inv4m?

***

# Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
```

***

# Load Data

```{r load_expression}
# Load normalized expression data
voomR <- readRDS(
  file.path(paths$intermediate, "normalized_expression_voom_object_leaf_trt.rds")
)

# Load sample metadata
wgcna_run_dir <- here::here(
  "results", "inversion_paper", "field_perturbation", "run_20251231_201332"
)
sample_meta <- read_csv(
  file.path(wgcna_run_dir, "01_data_prep", "sample_metadata.csv"),
  show_col_types = FALSE
)

# Split samples by genotype
ctrl_samples <- sample_meta %>% filter(Genotype == "CTRL") %>% pull(sample_id)
inv4m_samples <- sample_meta %>% filter(Genotype == "Inv4m") %>% pull(sample_id)
```

```{r load_wgcna_genes}
# Load WGCNA module assignments (genes assigned to modules, excluding grey)
ctrl_modules <- read_csv(
  file.path(wgcna_run_dir, "04_consensus_networks", "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)

wgcna_genes <- ctrl_modules %>%
  filter(module != 0) %>%
  pull(gene) %>%
  intersect(rownames(voomR$E))

cat("WGCNA genes:", length(wgcna_genes), "\n")
```

```{r load_maizenetome}
# Load and combine MaizeNetome coexpression edges
left_mn <- read.csv(file.path(paths$data, "left_MaizeNetome_coexpression_edges.csv"))
inv4m_mn <- read.csv(file.path(paths$data, "inv4m_MaizeNetome_coexpression_edges.csv"))
right_mn <- read.csv(file.path(paths$data, "right_MaizeNetome_coexpression_edges.csv"))

MaizeNetome <- rbind(
  left_mn %>% separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n"),
  inv4m_mn %>% separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n"),
  right_mn %>% separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n")
) %>%
  dplyr::rename(Direct.Connection = Direct.Connetation.Node)

# Convert v4 to v5 gene IDs
v4_v5 <- read.table(
  file.path(paths$data, "B73v4_to_B73v5.tsv"),
  sep = "\t", header = FALSE
) %>%
  dplyr::rename(v4 = V1, v5 = V2) %>%
  separate_longer_delim(cols = v5, delim = ",")

MaizeNetome_v5 <- MaizeNetome %>%
  left_join(v4_v5, by = c("Element.ID" = "v4"), relationship = "many-to-many") %>%
  dplyr::rename(Element.ID.v5 = v5) %>%
  filter(!is.na(Element.ID.v5)) %>%
  left_join(v4_v5, by = c("Direct.Connection" = "v4"), relationship = "many-to-many") %>%
  dplyr::rename(Direct.Connection.v5 = v5) %>%
  filter(!is.na(Direct.Connection.v5))
```

***
# Filter MaizeNetome to WGCNA Gene Set

```{r filter_maizenetome}
# Get MaizeNetome edges where BOTH genes are in our WGCNA set
mn_edges <- MaizeNetome_v5 %>%
  filter(Element.ID.v5 %in% wgcna_genes & Direct.Connection.v5 %in% wgcna_genes) %>%
  rowwise() %>%
  mutate(
    gene1 = min(Element.ID.v5, Direct.Connection.v5),
    gene2 = max(Element.ID.v5, Direct.Connection.v5),
    edge_id = paste(gene1, gene2, sep = "_")
  ) %>%
  ungroup() %>%
  distinct(edge_id, gene1, gene2)

cat("MaizeNetome reference edges among WGCNA genes:", nrow(mn_edges), "\n")
```

***

# Test Reference Edges in CTRL and Inv4m

For each MaizeNetome edge, we compute the Pearson correlation in CTRL and Inv4m
samples separately. An edge is "detected" if the correlation is significant
(FDR < 0.05).

```{r compute_correlations}
# Extract expression matrices
expr_ctrl <- voomR$E[wgcna_genes, ctrl_samples]
expr_inv4m <- voomR$E[wgcna_genes, inv4m_samples]
n_ctrl <- ncol(expr_ctrl)
n_inv4m <- ncol(expr_inv4m)

# Function to compute correlation p-value
cor_pvalue <- function(r, n) {
  t_stat <- r * sqrt((n - 2) / (1 - r^2))
  2 * pt(-abs(t_stat), df = n - 2)
}

# Compute correlations for each MaizeNetome edge
mn_results <- mn_edges %>%
  rowwise() %>%
  mutate(
    # CTRL correlation
    r_ctrl = cor(expr_ctrl[gene1, ], expr_ctrl[gene2, ], use = "complete.obs"),
    p_ctrl = cor_pvalue(r_ctrl, n_ctrl),
    # Inv4m correlation
    r_inv4m = cor(expr_inv4m[gene1, ], expr_inv4m[gene2, ], use = "complete.obs"),
    p_inv4m = cor_pvalue(r_inv4m, n_inv4m)
  ) %>%
  ungroup()

# FDR correction (across all tested edges)
mn_results <- mn_results %>%
  mutate(
    fdr_ctrl = p.adjust(p_ctrl, method = "fdr"),
    fdr_inv4m = p.adjust(p_inv4m, method = "fdr"),
    sig_ctrl = fdr_ctrl < 0.05,
    sig_inv4m = fdr_inv4m < 0.05
  )
```

***

# Edge-Level Results

```{r edge_results}
# Classify edge fate
mn_results <- mn_results %>%
  mutate(
    status = case_when(
      sig_ctrl & sig_inv4m ~ "preserved",
      sig_ctrl & !sig_inv4m ~ "lost",
      !sig_ctrl & sig_inv4m ~ "gained",
      TRUE ~ "not_detected"
    )
  )

edge_summary <- mn_results %>%
  count(status) %>%
  mutate(pct = round(100 * n / sum(n), 1))

cat("=== MAIZENETOME EDGE FATE ===\n")
knitr::kable(edge_summary, format = "simple")

# Key statistics
n_reference <- nrow(mn_results)
n_detected_ctrl <- sum(mn_results$sig_ctrl)
n_detected_inv4m <- sum(mn_results$sig_inv4m)
n_preserved <- sum(mn_results$status == "preserved")
n_lost <- sum(mn_results$status == "lost")
n_gained <- sum(mn_results$status == "gained")

cat("\n=== KEY STATISTICS ===\n")
cat("MaizeNetome reference edges:", n_reference, "\n")
cat("Detected in CTRL (FDR < 0.05):", n_detected_ctrl,
    "(", round(100 * n_detected_ctrl / n_reference, 1), "%)\n")
cat("Detected in Inv4m (FDR < 0.05):", n_detected_inv4m,
    "(", round(100 * n_detected_inv4m / n_reference, 1), "%)\n")
cat("Lost in Inv4m:", n_lost, "/", n_detected_ctrl,
    "(", round(100 * n_lost / n_detected_ctrl, 1), "%)\n")
cat("Preserved in Inv4m:", n_preserved, "/", n_detected_ctrl,
    "(", round(100 * n_preserved / n_detected_ctrl, 1), "%)\n")
cat("Gained in Inv4m (not in CTRL):", n_gained, "\n")
cat("Net change:", n_detected_inv4m - n_detected_ctrl, "edges\n")
```

***

# Vertex-Level Results

How many genes lose all their MaizeNetome connections in Inv4m?

```{r vertex_results}
# Count edges per gene by status
vertex_stats <- bind_rows(
  mn_results %>% filter(sig_ctrl) %>% dplyr::select(gene = gene1, status),
  mn_results %>% filter(sig_ctrl) %>% dplyr::select(gene = gene2, status)
) %>%
  count(gene, status) %>%
  pivot_wider(names_from = status, values_from = n, values_fill = 0)

# Ensure columns exist
for (col in c("preserved", "lost")) {
  if (!col %in% names(vertex_stats)) vertex_stats[[col]] <- 0
}

vertex_stats <- vertex_stats %>%
  mutate(
    total_ctrl = preserved + lost,
    pct_lost = round(100 * lost / total_ctrl, 1)
  )

# Genes with complete loss
total_loss_genes <- vertex_stats %>%
  filter(total_ctrl > 0, preserved == 0)

n_genes_with_edges <- sum(vertex_stats$total_ctrl > 0)
n_genes_total_loss <- nrow(total_loss_genes)

cat("=== VERTEX-LEVEL DISRUPTION ===\n")
cat("Genes with MaizeNetome edges in CTRL:", n_genes_with_edges, "\n")
cat("Genes losing ALL connections:", n_genes_total_loss,
    "(", round(100 * n_genes_total_loss / n_genes_with_edges, 1), "%)\n")
```

***

# Visualization

```{r figure_edge_fate, fig.width=10, fig.height=5}
# Edge fate summary - show detection in both genotypes and edge dynamics
edge_plot_data <- data.frame(
  Status = c("CTRL", "Inv4m", "Lost", "Preserved", "Gained"),
  Count = c(n_detected_ctrl, n_detected_inv4m, n_lost, n_preserved, n_gained),
  Type = c("Detection", "Detection", "Dynamics", "Dynamics", "Dynamics")
)
edge_plot_data$Status <- factor(edge_plot_data$Status, levels = edge_plot_data$Status)

p_edges <- ggplot(edge_plot_data, aes(x = Status, y = Count, fill = Status)) +
  geom_col(color = "black", width = 0.6) +
  geom_text(aes(label = Count), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c(
    "CTRL" = "#3498DB",
    "Inv4m" = "#9B59B6",
    "Lost" = "#E74C3C",
    "Preserved" = "#27AE60",
    "Gained" = "#F39C12"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    title = "MaizeNetome Reference Edge Detection and Dynamics",
    subtitle = sprintf(
      "%d reference edges | CTRL detects %d, Inv4m detects %d (net: %+d)",
      n_reference, n_detected_ctrl, n_detected_inv4m, n_detected_inv4m - n_detected_ctrl
    ),
    x = NULL,
    y = "Number of MaizeNetome Edges"
  )

print(p_edges)

ggsave(
  file.path(paths$figures, "maizenetome_edge_disruption.pdf"),
  p_edges, width = 10, height = 5
)
```

```{r figure_vertex_loss, fig.width=8, fig.height=5}
# Vertex loss distribution
vertex_plot_data <- vertex_stats %>%
  filter(total_ctrl > 0) %>%
  mutate(
    loss_category = case_when(
      pct_lost == 0 ~ "0% (fully preserved)",
      pct_lost == 100 ~ "100% (complete loss)",
      TRUE ~ "Partial loss"
    )
  ) %>%
  count(loss_category)

p_vertex <- ggplot(vertex_plot_data, aes(x = loss_category, y = n, fill = loss_category)) +
  geom_col(color = "black", width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c(
    "0% (fully preserved)" = "#27AE60",
    "Partial loss" = "#F39C12",
    "100% (complete loss)" = "#E74C3C"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    title = "MaizeNetome Connection Loss per Gene",
    subtitle = sprintf(
      "%d genes with reference edges | %d%% lose all connections",
      n_genes_with_edges, round(100 * n_genes_total_loss / n_genes_with_edges)
    ),
    x = "Edge Loss Category",
    y = "Number of Genes"
  )

print(p_vertex)

ggsave(
  file.path(paths$figures, "maizenetome_vertex_disruption.pdf"),
  p_vertex, width = 8, height = 5
)
```

***

# Export Results

```{r export}
# Summary statistics
summary_stats <- tibble(
  Metric = c(
    "WGCNA genes analyzed",
    "MaizeNetome reference edges (among WGCNA genes)",
    "Reference edges detected in CTRL (FDR < 0.05)",
    "Reference edges detected in Inv4m (FDR < 0.05)",
    "CTRL detection rate (%)",
    "Inv4m detection rate (%)",
    "Edges preserved (in both)",
    "Edges lost (CTRL only)",
    "Edges gained (Inv4m only)",
    "Edge loss rate (% of CTRL detected)",
    "Net edge change",
    "Genes with reference edges in CTRL",
    "Genes losing all reference connections",
    "Complete loss rate (%)"
  ),
  Value = c(
    length(wgcna_genes),
    n_reference,
    n_detected_ctrl,
    n_detected_inv4m,
    round(100 * n_detected_ctrl / n_reference, 1),
    round(100 * n_detected_inv4m / n_reference, 1),
    n_preserved,
    n_lost,
    n_gained,
    round(100 * n_lost / n_detected_ctrl, 1),
    n_detected_inv4m - n_detected_ctrl,
    n_genes_with_edges,
    n_genes_total_loss,
    round(100 * n_genes_total_loss / n_genes_with_edges, 1)
  )
)

knitr::kable(summary_stats, format = "simple")

write_csv(summary_stats, file.path(paths$intermediate, "maizenetome_disruption_summary.csv"))
write_csv(mn_results, file.path(paths$intermediate, "maizenetome_edge_results.csv"))
write_csv(vertex_stats, file.path(paths$intermediate, "maizenetome_vertex_results.csv"))
```

***

# Summary

```{r summary, comment=""}
cat("============================================================\n")
cat("MAIZENETOME REFERENCE NETWORK DISRUPTION\n")
cat("============================================================\n\n")

cat("QUESTION: Does Inv4m disrupt established coexpression networks?\n\n")

cat("EDGE-LEVEL ANALYSIS\n")
cat(sprintf("  Reference edges tested: %d\n", n_reference))
cat(sprintf("  Detected in CTRL: %d (%.1f%%)\n",
    n_detected_ctrl, 100 * n_detected_ctrl / n_reference))
cat(sprintf("  Detected in Inv4m: %d (%.1f%%)\n",
    n_detected_inv4m, 100 * n_detected_inv4m / n_reference))
cat(sprintf("  Lost (CTRL only): %d (%.1f%% of CTRL)\n",
    n_lost, 100 * n_lost / n_detected_ctrl))
cat(sprintf("  Preserved (both): %d\n", n_preserved))
cat(sprintf("  Gained (Inv4m only): %d\n", n_gained))
cat(sprintf("  Net change: %+d edges\n\n", n_detected_inv4m - n_detected_ctrl))

cat("VERTEX-LEVEL ANALYSIS\n")
cat(sprintf("  Genes with reference edges: %d\n", n_genes_with_edges))
cat(sprintf("  Genes losing ALL connections: %d (%.1f%%)\n\n",
    n_genes_total_loss, 100 * n_genes_total_loss / n_genes_with_edges))

cat("CONCLUSION: Inv4m causes net loss of MaizeNetome reference edges.\n")
cat("While some edges are gained, more are lost than gained.\n")
cat("============================================================\n")
```

***

# Session Information

```{r session_info}
sessionInfo()
```
