---
title: "WGCNA Network Validation Against MaizeNetome"
author: "Fausto Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,

  message = FALSE,
  fig.width = 10,
  fig.height = 8
)

library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Validate WGCNA module networks against the MaizeNetome reference
database and quantify network disruption caused by Inv4m introgression.

**Approach**:
1. Build binary correlation networks from CTRL and Inv4m expression data
2. Use same methodology as trans-network analysis (Pearson correlation, FDR < 0.05)
3. Compare edges to MaizeNetome reference
4. Quantify edge loss between CTRL and Inv4m conditions

**Key outputs**:
- Edge counts: CTRL vs Inv4m
- Reference vs Novel edge classification
- Edge loss statistics (network disruption)

---

# Setup and Dependencies

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(patchwork)
library(eulerr)
```

---

# Load Data

## Expression Data

```{r load_expression}
# Load normalized expression data
voomR <- readRDS(
 file.path(paths$intermediate, "normalized_expression_voom_object_leaf_trt.rds")
)

# Load sample metadata from WGCNA pipeline
wgcna_run_dir <- here::here(
  "results", "inversion_paper", "field_perturbation",
  "run_20251231_201332"
)
sample_meta <- read_csv(
  file.path(wgcna_run_dir, "01_data_prep", "sample_metadata.csv"),
  show_col_types = FALSE
)

# Split samples by genotype
ctrl_samples <- sample_meta %>%
  filter(Genotype == "CTRL") %>%
  pull(sample_id)

inv4m_samples <- sample_meta %>%
  filter(Genotype == "Inv4m") %>%
  pull(sample_id)

cat("=== EXPRESSION DATA ===\n")
cat("Total genes:", nrow(voomR$E), "\n")
cat("CTRL samples:", length(ctrl_samples), "\n")
cat("Inv4m samples:", length(inv4m_samples), "\n")
```

## WGCNA Module Genes

```{r load_wgcna_genes}
# Load WGCNA module assignments (CTRL modules as reference)
ctrl_modules <- read_csv(
  file.path(wgcna_run_dir, "04_consensus_networks", "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)

# Get genes assigned to modules (exclude grey/unassigned)
wgcna_genes <- ctrl_modules %>%
  filter(module != 0) %>%
  pull(gene)

# Filter to genes in expression matrix
wgcna_genes <- wgcna_genes[wgcna_genes %in% rownames(voomR$E)]

cat("\n=== WGCNA GENES ===\n")
cat("Genes in modules:", length(wgcna_genes), "\n")
cat("Genes in expression matrix:", sum(wgcna_genes %in% rownames(voomR$E)), "\n")
```

## MaizeNetome Reference

```{r load_maizenetome}
# Load MaizeNetome coexpression edges (same as trans-network analysis)
left_flanking_maizenetome <- read.csv(
  file.path(paths$data, "left_MaizeNetome_coexpression_edges.csv")
)
Inv4m_maizenetome <- read.csv(
  file.path(paths$data, "inv4m_MaizeNetome_coexpression_edges.csv")
)
right_flanking_maizenetome <- read.csv(
  file.path(paths$data, "right_MaizeNetome_coexpression_edges.csv")
)

MaizeNetome <- rbind(
  left_flanking_maizenetome %>%
    separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n") %>%
    mutate(network = "coexpression"),
  Inv4m_maizenetome %>%
    separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n") %>%
    mutate(network = "coexpression"),
  right_flanking_maizenetome %>%
    separate_longer_delim(cols = Direct.Connetation.Node, delim = "\n") %>%
    mutate(network = "coexpression")
) %>%
  dplyr::rename(Direct.Connection = Direct.Connetation.Node)

cat("\n=== MAIZENETOME ===\n")
cat("Total edges (v4 IDs):", nrow(MaizeNetome), "\n")
```

```{r convert_v4_v5}
# Convert v4 to v5 gene IDs
v4_v5 <- read.table(
  file = file.path(paths$data, "B73v4_to_B73v5.tsv"),
  sep = "\t",
  header = FALSE
) %>%
  dplyr::rename(v4 = V1, v5 = V2) %>%
  separate_longer_delim(cols = v5, delim = ",")

MaizeNetome_v5 <- MaizeNetome %>%
  left_join(v4_v5, by = c("Element.ID" = "v4")) %>%
  dplyr::rename(Element.ID.v5 = v5) %>%
  filter(!is.na(Element.ID.v5)) %>%
  left_join(v4_v5, by = c("Direct.Connection" = "v4")) %>%
  dplyr::rename(Direct.Connection.v5 = v5) %>%
  filter(!is.na(Direct.Connection.v5))

# Create edge set for fast lookup (undirected - sort gene pairs)
maizenetome_edges <- MaizeNetome_v5 %>%
  rowwise() %>%
  mutate(
    gene1 = min(Element.ID.v5, Direct.Connection.v5),
    gene2 = max(Element.ID.v5, Direct.Connection.v5),
    edge_id = paste(gene1, gene2, sep = "_")
  ) %>%
  ungroup() %>%
  distinct(edge_id, .keep_all = TRUE)

cat("Edges after v4->v5 conversion:", nrow(MaizeNetome_v5), "\n
")
cat("Unique undirected edges:", nrow(maizenetome_edges), "\n")
```

---

# Build Correlation Networks

## Helper Functions

```{r helper_functions}
#' Build binary edge list from expression matrix
#'
#' @param expr Expression matrix (genes x samples)
#' @param genes Genes to include
#' @param fdr_threshold FDR threshold for significance
#'
#' @return Data frame with significant edges
build_correlation_network <- function(expr, genes, fdr_threshold = 0.05) {

  # Filter to genes in expression matrix
  genes <- genes[genes %in% rownames(expr)]
  expr_subset <- expr[genes, ]

  n_genes <- length(genes)
  n_samples <- ncol(expr_subset)

  cat("Building network for", n_genes, "genes x", n_samples, "samples\n")
  cat("Total possible edges:", n_genes * (n_genes - 1) / 2, "\n")

 # Compute correlation matrix
  cor_matrix <- cor(t(expr_subset), use = "pairwise.complete.obs")

  # Compute p-values for correlations
  # t-statistic: t = r * sqrt((n-2)/(1-r^2))
  t_matrix <- cor_matrix * sqrt((n_samples - 2) / (1 - cor_matrix^2))
  p_matrix <- 2 * pt(-abs(t_matrix), df = n_samples - 2)

  # Extract lower triangle (undirected network)
  lower_idx <- which(lower.tri(cor_matrix), arr.ind = TRUE)

  edges_df <- data.frame(
    gene1 = genes[lower_idx[, 1]],
    gene2 = genes[lower_idx[, 2]],
    r = cor_matrix[lower_idx],
    p_value = p_matrix[lower_idx]
  )

  # FDR correction
  edges_df$p_adj <- p.adjust(edges_df$p_value, method = "fdr")

  # Filter significant edges
  sig_edges <- edges_df %>%
    filter(p_adj < fdr_threshold) %>%
    rowwise() %>%
    mutate(
      # Ensure consistent ordering for comparison
      g1 = min(gene1, gene2),
      g2 = max(gene1, gene2),
      edge_id = paste(g1, g2, sep = "_")
    ) %>%
    ungroup() %>%
    dplyr::select(gene1 = g1, gene2 = g2, edge_id, r, p_value, p_adj)

  cat("Significant edges (FDR <", fdr_threshold, "):", nrow(sig_edges), "\n")

  return(sig_edges)
}

#' Check edges against MaizeNetome
#'
#' @param edges Edge data frame with edge_id column
#' @param maizenetome_edge_ids Vector of MaizeNetome edge IDs
#'
#' @return Edge data frame with in_maizenetome column
check_maizenetome <- function(edges, maizenetome_edge_ids) {
  edges %>%
    mutate(in_maizenetome = edge_id %in% maizenetome_edge_ids)
}
```

## CTRL Network

```{r build_ctrl_network}
cat("=== BUILDING CTRL NETWORK ===\n\n")

# Extract CTRL expression
expr_ctrl <- voomR$E[, ctrl_samples]

# Build network
ctrl_edges <- build_correlation_network(
  expr = expr_ctrl,
  genes = wgcna_genes,
  fdr_threshold = 0.05
)

# Check against MaizeNetome
ctrl_edges <- check_maizenetome(ctrl_edges, maizenetome_edges$edge_id)

cat("\n=== CTRL NETWORK SUMMARY ===\n")
cat("Total edges:", nrow(ctrl_edges), "\n")
cat("Reference (in MaizeNetome):", sum(ctrl_edges$in_maizenetome), "\n")
cat("Novel (not in MaizeNetome):", sum(!ctrl_edges$in_maizenetome), "\n")
cat("% in MaizeNetome:", round(100 * mean(ctrl_edges$in_maizenetome), 2), "%\n")
```

## Inv4m Network

```{r build_inv4m_network}
cat("=== BUILDING INV4M NETWORK ===\n\n")

# Extract Inv4m expression
expr_inv4m <- voomR$E[, inv4m_samples]

# Build network
inv4m_edges <- build_correlation_network(
  expr = expr_inv4m,
  genes = wgcna_genes,
  fdr_threshold = 0.05
)

# Check against MaizeNetome
inv4m_edges <- check_maizenetome(inv4m_edges, maizenetome_edges$edge_id)

cat("\n=== INV4M NETWORK SUMMARY ===\n")
cat("Total edges:", nrow(inv4m_edges), "\n")
cat("Reference (in MaizeNetome):", sum(inv4m_edges$in_maizenetome), "\n")
cat("Novel (not in MaizeNetome):", sum(!inv4m_edges$in_maizenetome), "\n")
cat("% in MaizeNetome:", round(100 * mean(inv4m_edges$in_maizenetome), 2), "%\n")
```

---

# Compare CTRL vs Inv4m Networks

## Edge Overlap Analysis

```{r edge_comparison}
# Create edge sets
ctrl_edge_set <- ctrl_edges$edge_id
inv4m_edge_set <- inv4m_edges$edge_id

# Calculate overlaps
shared_edges <- intersect(ctrl_edge_set, inv4m_edge_set)
ctrl_only_edges <- setdiff(ctrl_edge_set, inv4m_edge_set)
inv4m_only_edges <- setdiff(inv4m_edge_set, ctrl_edge_set)

cat("=== EDGE COMPARISON ===\n")
cat("Shared edges (in both):", length(shared_edges), "\n")
cat("CTRL-only edges (lost in Inv4m):", length(ctrl_only_edges), "\n")
cat("Inv4m-only edges (gained):", length(inv4m_only_edges), "\n")
cat("\nEdge loss rate:", round(100 * length(ctrl_only_edges) / length(ctrl_edge_set), 2), "%\n")
cat("Edge gain rate:", round(100 * length(inv4m_only_edges) / length(inv4m_edge_set), 2), "%\n")
```

```{r edge_categories}
# Categorize all edges
all_edges <- unique(c(ctrl_edge_set, inv4m_edge_set))

edge_summary <- data.frame(edge_id = all_edges) %>%
  mutate(
    in_ctrl = edge_id %in% ctrl_edge_set,
    in_inv4m = edge_id %in% inv4m_edge_set,
    in_maizenetome = edge_id %in% maizenetome_edges$edge_id,
    category = case_when(
      in_ctrl & in_inv4m ~ "Shared",
      in_ctrl & !in_inv4m ~ "Lost",
      !in_ctrl & in_inv4m ~ "Gained"
    )
  )

cat("\n=== EDGE CATEGORIES BY MAIZENETOME STATUS ===\n")
edge_summary %>%
  group_by(category, in_maizenetome) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = in_maizenetome, values_from = n, names_prefix = "MaizeNetome_") %>%
  print()
```

## Sign Concordance Analysis

Since WGCNA uses signed networks, we verify that "shared" edges have concordant
correlation signs between CTRL and Inv4m. An edge with r = +0.8 in CTRL and
r = -0.7 in Inv4m would be significant in both but represents a biological
reversal, not preservation.

```{r sign_concordance}
# Join CTRL and Inv4m edges to compare correlation signs
shared_edge_correlations <- inner_join(
 ctrl_edges %>% dplyr::select(edge_id, r_ctrl = r),
  inv4m_edges %>% dplyr::select(edge_id, r_inv4m = r),
  by = "edge_id"
) %>%
  mutate(
    sign_ctrl = sign(r_ctrl),
    sign_inv4m = sign(r_inv4m),
    same_sign = sign_ctrl == sign_inv4m,
    sign_pattern = case_when(
      r_ctrl > 0 & r_inv4m > 0 ~ "pos → pos",
      r_ctrl > 0 & r_inv4m < 0 ~ "pos → neg (FLIP)",
      r_ctrl < 0 & r_inv4m > 0 ~ "neg → pos (FLIP)",
      r_ctrl < 0 & r_inv4m < 0 ~ "neg → neg"
    )
  )

cat("=== SIGN CONCORDANCE IN SHARED EDGES ===\n")
cat("Total shared edges:", nrow(shared_edge_correlations), "\n\n")

cat("Same sign (truly preserved):", sum(shared_edge_correlations$same_sign),
    "(", round(100 * mean(shared_edge_correlations$same_sign), 1), "%)\n")
cat("Opposite sign (sign flip):", sum(!shared_edge_correlations$same_sign),
    "(", round(100 * mean(!shared_edge_correlations$same_sign), 1), "%)\n\n")

cat("=== SIGN PATTERN BREAKDOWN ===\n")
shared_edge_correlations %>%
  count(sign_pattern) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n)) %>%
  print()

# Store for later use
n_truly_preserved <- sum(shared_edge_correlations$same_sign)
n_sign_flips <- sum(!shared_edge_correlations$same_sign)
```

```{r sign_concordance_maizenetome}
# Check sign concordance specifically for MaizeNetome edges
shared_mn_correlations <- shared_edge_correlations %>%
  filter(edge_id %in% maizenetome_edges$edge_id)

cat("\n=== SIGN CONCORDANCE IN MAIZENETOME EDGES ===\n")
cat("MaizeNetome edges in both networks:", nrow(shared_mn_correlations), "\n")
cat("Same sign:", sum(shared_mn_correlations$same_sign),
    "(", round(100 * mean(shared_mn_correlations$same_sign), 1), "%)\n")
cat("Sign flip:", sum(!shared_mn_correlations$same_sign),
    "(", round(100 * mean(!shared_mn_correlations$same_sign), 1), "%)\n")

# Store for summary
n_mn_truly_preserved <- sum(shared_mn_correlations$same_sign)
n_mn_sign_flips <- sum(!shared_mn_correlations$same_sign)
```

---

# Network Statistics Summary

```{r network_stats}
# Compile network statistics
network_stats <- tibble(
  Metric = c(
    "WGCNA genes analyzed",
    "Possible edges",
    "--- CTRL Network ---",
    "CTRL total edges",
    "CTRL Reference edges (in MaizeNetome)",
    "CTRL Novel edges (not in MaizeNetome)",
    "CTRL % in MaizeNetome",
    "--- Inv4m Network ---",
    "Inv4m total edges",
    "Inv4m Reference edges (in MaizeNetome)",
    "Inv4m Novel edges (not in MaizeNetome)",
    "Inv4m % in MaizeNetome",
    "--- Network Disruption ---",
    "Shared edges (significant in both)",
    "Lost edges (CTRL only)",
    "Gained edges (Inv4m only)",
    "Edge loss rate (%)",
    "Edge gain rate (%)",
    "Net edge change",
    "--- Sign Concordance ---",
    "Shared edges with same sign",
    "Shared edges with sign flip",
    "% same sign",
    "MaizeNetome edges same sign",
    "MaizeNetome edges sign flip"
  ),
  Value = c(
    length(wgcna_genes),
    length(wgcna_genes) * (length(wgcna_genes) - 1) / 2,
    "",
    nrow(ctrl_edges),
    sum(ctrl_edges$in_maizenetome),
    sum(!ctrl_edges$in_maizenetome),
    round(100 * mean(ctrl_edges$in_maizenetome), 2),
    "",
    nrow(inv4m_edges),
    sum(inv4m_edges$in_maizenetome),
    sum(!inv4m_edges$in_maizenetome),
    round(100 * mean(inv4m_edges$in_maizenetome), 2),
    "",
    length(shared_edges),
    length(ctrl_only_edges),
    length(inv4m_only_edges),
    round(100 * length(ctrl_only_edges) / length(ctrl_edge_set), 2),
    round(100 * length(inv4m_only_edges) / length(inv4m_edge_set), 2),
    nrow(inv4m_edges) - nrow(ctrl_edges),
    "",
    n_truly_preserved,
    n_sign_flips,
    round(100 * n_truly_preserved / length(shared_edges), 2),
    n_mn_truly_preserved,
    n_mn_sign_flips
  )
)

print(network_stats)

write_csv(network_stats, file.path(paths$intermediate, "wgcna_maizenetome_validation_stats.csv"))
```

---

# Visualization

## Euler Diagram: CTRL vs Inv4m vs MaizeNetome

```{r euler_diagram, fig.width=8, fig.height=8}
# Create sets for Euler diagram
# Note: MaizeNetome is much larger, so we focus on overlap with our edges
our_edges <- unique(c(ctrl_edge_set, inv4m_edge_set))
maizenetome_relevant <- maizenetome_edges$edge_id[maizenetome_edges$edge_id %in% our_edges]

euler_data <- list(
  CTRL = ctrl_edge_set,
  Inv4m = inv4m_edge_set
)

p_euler <- plot(
  euler(euler_data),
  quantities = TRUE,
  fills = list(fill = c("#DAA520", "#551A8B"), alpha = 0.6),
  labels = list(fontsize = 14),
  main = "Edge Overlap: CTRL vs Inv4m Networks"
)

print(p_euler)

# Save
pdf(file.path(paths$figures, "wgcna_edge_euler.pdf"), width = 8, height = 8)
print(p_euler)
dev.off()
```

## Bar Plot: Edge Counts by Category

```{r edge_barplot, fig.width=10, fig.height=6}
edge_counts <- data.frame(
  Category = c("CTRL total", "Inv4m total", "Shared", "Lost\n(CTRL only)", "Gained\n(Inv4m only)"),
  Count = c(
    nrow(ctrl_edges),
    nrow(inv4m_edges),
    length(shared_edges),
    length(ctrl_only_edges),
    length(inv4m_only_edges)
  ),
  Type = c("Network", "Network", "Comparison", "Comparison", "Comparison")
)

edge_counts$Category <- factor(edge_counts$Category, levels = edge_counts$Category)

p_bar <- ggplot(edge_counts, aes(x = Category, y = Count, fill = Type)) +
  geom_col(color = "black", width = 0.7) +
  geom_text(aes(label = scales::comma(Count)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Network" = "steelblue", "Comparison" = "coral")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12)
  ) +
  labs(
    title = "WGCNA Network Edge Comparison",
    subtitle = paste0("Genes: ", length(wgcna_genes), " | FDR threshold: 0.05"),
    x = NULL,
    y = "Number of Edges"
  )

print(p_bar)

ggsave(
  file.path(paths$figures, "wgcna_edge_comparison_barplot.pdf"),
  p_bar, width = 10, height = 6
)
ggsave(
  file.path(paths$figures, "wgcna_edge_comparison_barplot.png"),
  p_bar, width = 10, height = 6, dpi = 300
)
```

## MaizeNetome Validation: Reference vs Novel

```{r maizenetome_validation_plot, fig.width=10, fig.height=6}
validation_data <- data.frame(
  Network = rep(c("CTRL", "Inv4m"), each = 2),
  EdgeType = rep(c("Reference\n(in MaizeNetome)", "Novel"), 2),
  Count = c(
    sum(ctrl_edges$in_maizenetome),
    sum(!ctrl_edges$in_maizenetome),
    sum(inv4m_edges$in_maizenetome),
    sum(!inv4m_edges$in_maizenetome)
  )
)

p_validation <- ggplot(validation_data, aes(x = Network, y = Count, fill = EdgeType)) +
  geom_col(position = "dodge", color = "black", width = 0.7) +
  geom_text(
    aes(label = scales::comma(Count)),
    position = position_dodge(width = 0.7),
    vjust = -0.5, size = 4
  ) +
  scale_fill_manual(values = c("Reference\n(in MaizeNetome)" = "#2E8B57", "Novel" = "#CD853F")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top") +
  labs(
    title = "MaizeNetome Validation of WGCNA Networks",
    subtitle = "Reference edges confirm biology; Novel edges are experiment-specific",
    x = NULL,
    y = "Number of Edges",
    fill = NULL
  )

print(p_validation)

ggsave(
  file.path(paths$figures, "wgcna_maizenetome_validation.pdf"),
  p_validation, width = 10, height = 6
)
ggsave(
  file.path(paths$figures, "wgcna_maizenetome_validation.png"),
  p_validation, width = 10, height = 6, dpi = 300
)
```

## Edge Loss by MaizeNetome Status

```{r edge_loss_maizenetome, fig.width=10, fig.height=6}
# Are lost edges more likely to be Reference or Novel?
lost_edge_df <- ctrl_edges %>%
  filter(edge_id %in% ctrl_only_edges)

shared_edge_df <- ctrl_edges %>%
  filter(edge_id %in% shared_edges)

loss_by_type <- data.frame(
  Category = c("Preserved\n(Shared)", "Preserved\n(Shared)", "Lost\n(CTRL only)", "Lost\n(CTRL only)"),
  MaizeNetome = c("Reference", "Novel", "Reference", "Novel"),
  Count = c(
    sum(shared_edge_df$in_maizenetome),
    sum(!shared_edge_df$in_maizenetome),
    sum(lost_edge_df$in_maizenetome),
    sum(!lost_edge_df$in_maizenetome)
  )
)

# Add percentages within category
loss_by_type <- loss_by_type %>%
  group_by(Category) %>%
  mutate(
    Total = sum(Count),
    Pct = round(100 * Count / Total, 1)
  ) %>%
  ungroup()

p_loss <- ggplot(loss_by_type, aes(x = Category, y = Count, fill = MaizeNetome)) +
  geom_col(position = "fill", color = "black", width = 0.7) +
  geom_text(
    aes(label = paste0(Pct, "%")),
    position = position_fill(vjust = 0.5),
    size = 5, color = "white", fontface = "bold"
  ) +
  scale_fill_manual(values = c("Reference" = "#2E8B57", "Novel" = "#CD853F")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top") +
  labs(
    title = "MaizeNetome Status of Preserved vs Lost Edges",
    subtitle = "Are Reference edges more stable than Novel edges?",
    x = NULL,
    y = "Proportion of Edges",
    fill = NULL
  )

print(p_loss)

ggsave(
  file.path(paths$figures, "wgcna_edge_loss_by_maizenetome.pdf"),
  p_loss, width = 8, height = 6
)
```

```{r fisher_test}
# Fisher's exact test: Are Reference edges more likely to be preserved?
fisher_table <- matrix(
  c(
    sum(shared_edge_df$in_maizenetome),    # Preserved & Reference
    sum(!shared_edge_df$in_maizenetome),   # Preserved & Novel
    sum(lost_edge_df$in_maizenetome),      # Lost & Reference
    sum(!lost_edge_df$in_maizenetome)      # Lost & Novel
  ),
  nrow = 2,
  dimnames = list(
    c("Reference", "Novel"),
    c("Preserved", "Lost")
  )
)

cat("\n=== CONTINGENCY TABLE ===\n")
print(fisher_table)

fisher_result <- fisher.test(fisher_table)
cat("\n=== FISHER'S EXACT TEST ===\n")
cat("Are Reference edges more likely to be preserved?\n")
cat("Odds ratio:", round(fisher_result$estimate, 3), "\n")
cat("P-value:", signif(fisher_result$p.value, 3), "\n")
cat("95% CI:", round(fisher_result$conf.int[1], 3), "-", round(fisher_result$conf.int[2], 3), "\n")
```

---

# Module-Level Edge Analysis

```{r module_edge_analysis}
# Add module information to edges
ctrl_modules_lookup <- ctrl_modules %>%
  filter(module != 0) %>%
  dplyr::select(gene, module, color)

# For each edge, get modules of both genes
ctrl_edges_with_modules <- ctrl_edges %>%
  left_join(ctrl_modules_lookup, by = c("gene1" = "gene")) %>%
  dplyr::rename(module1 = module, color1 = color) %>%
  left_join(ctrl_modules_lookup, by = c("gene2" = "gene")) %>%
  dplyr::rename(module2 = module, color2 = color) %>%
  mutate(
    same_module = module1 == module2,
    edge_type = ifelse(same_module, "Within-module", "Between-module"),
    is_lost = edge_id %in% ctrl_only_edges
  )

cat("=== EDGE TYPES ===\n")
print(table(ctrl_edges_with_modules$edge_type))

cat("\n=== EDGE LOSS BY TYPE ===\n")
ctrl_edges_with_modules %>%
  group_by(edge_type) %>%
  summarise(
    Total = n(),
    Lost = sum(is_lost),
    Preserved = sum(!is_lost),
    Loss_Rate = round(100 * Lost / Total, 2)
  ) %>%
  print()
```

```{r module_edge_plot, fig.width=10, fig.height=6}
module_loss <- ctrl_edges_with_modules %>%
  filter(same_module) %>%
  group_by(color1) %>%
  summarise(
    Total_Edges = n(),
    Lost = sum(is_lost),
    Preserved = sum(!is_lost),
    Loss_Rate = 100 * Lost / Total_Edges,
    .groups = "drop"
  ) %>%
  arrange(desc(Loss_Rate)) %>%
  dplyr::rename(Module = color1)

# Only plot modules with at least 10 edges
module_loss_plot <- module_loss %>%
  filter(Total_Edges >= 10)

if (nrow(module_loss_plot) > 0) {
  p_module_loss <- ggplot(
    module_loss_plot,
    aes(x = reorder(Module, -Loss_Rate), y = Loss_Rate, fill = Module)
  ) +
    geom_col(color = "black") +
    geom_text(aes(label = paste0(round(Loss_Rate, 1), "%")), vjust = -0.5, size = 3) +
    scale_fill_identity() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = "Within-Module Edge Loss Rate by Module",
      subtitle = "Modules with >= 10 within-module edges",
      x = "Module",
      y = "Edge Loss Rate (%)"
    )

  print(p_module_loss)

  ggsave(
    file.path(paths$figures, "wgcna_module_edge_loss.pdf"),
    p_module_loss, width = 12, height = 6
  )
}

write_csv(module_loss, file.path(paths$intermediate, "wgcna_module_edge_loss.csv"))
```

---

# Export Edge Lists

```{r export_edges}
# Export CTRL edges with annotations
ctrl_edges_export <- ctrl_edges_with_modules %>%
  dplyr::select(
    gene1, gene2, edge_id, r, p_adj,
    in_maizenetome, module1, module2, same_module, is_lost
  )

write_csv(ctrl_edges_export, file.path(paths$intermediate, "wgcna_ctrl_edges.csv"))

# Export Inv4m edges
inv4m_edges_export <- inv4m_edges %>%
  dplyr::select(gene1, gene2, edge_id, r, p_adj, in_maizenetome)

write_csv(inv4m_edges_export, file.path(paths$intermediate, "wgcna_inv4m_edges.csv"))

# Export edge comparison summary with sign concordance
edge_comparison <- data.frame(
  edge_id = unique(c(ctrl_edge_set, inv4m_edge_set))
) %>%
  mutate(
    in_ctrl = edge_id %in% ctrl_edge_set,
    in_inv4m = edge_id %in% inv4m_edge_set,
    in_maizenetome = edge_id %in% maizenetome_edges$edge_id,
    status = case_when(
      in_ctrl & in_inv4m ~ "Shared",
      in_ctrl & !in_inv4m ~ "Lost",
      !in_ctrl & in_inv4m ~ "Gained"
    )
  ) %>%
  left_join(
    shared_edge_correlations %>%
      dplyr::select(edge_id, r_ctrl, r_inv4m, same_sign, sign_pattern),
    by = "edge_id"
  )

write_csv(edge_comparison, file.path(paths$intermediate, "wgcna_edge_comparison.csv"))

cat("=== EXPORTED FILES ===\n")
cat("wgcna_ctrl_edges.csv:", nrow(ctrl_edges_export), "edges\n")
cat("wgcna_inv4m_edges.csv:", nrow(inv4m_edges_export), "edges\n")
cat("wgcna_edge_comparison.csv:", nrow(edge_comparison), "unique edges\n")
cat("wgcna_maizenetome_validation_stats.csv: summary statistics\n")
cat("wgcna_module_edge_loss.csv: module-level edge loss\n")
```

---

# Summary

```{r summary}
cat("\n")
cat("=" %>% rep(60) %>% paste(collapse = ""), "\n")
cat("WGCNA NETWORK VALIDATION SUMMARY\n")
cat("=" %>% rep(60) %>% paste(collapse = ""), "\n\n")

cat("NETWORK CONSTRUCTION\n")
cat("  Genes:", length(wgcna_genes), "\n")
cat("  Method: Pearson correlation, FDR < 0.05\n")
cat("  (Same parameters as trans-network analysis)\n\n")

cat("CTRL NETWORK\n")
cat("  Total edges:", scales::comma(nrow(ctrl_edges)), "\n")
cat("  Reference (MaizeNetome):", scales::comma(sum(ctrl_edges$in_maizenetome)),
    "(", round(100 * mean(ctrl_edges$in_maizenetome), 1), "%)\n")
cat("  Novel:", scales::comma(sum(!ctrl_edges$in_maizenetome)), "\n\n")

cat("INV4M NETWORK\n")
cat("  Total edges:", scales::comma(nrow(inv4m_edges)), "\n")
cat("  Reference (MaizeNetome):", scales::comma(sum(inv4m_edges$in_maizenetome)),
    "(", round(100 * mean(inv4m_edges$in_maizenetome), 1), "%)\n")
cat("  Novel:", scales::comma(sum(!inv4m_edges$in_maizenetome)), "\n\n")

cat("NETWORK DISRUPTION\n")
cat("  Shared edges:", scales::comma(length(shared_edges)), "\n")
cat("  Lost edges:", scales::comma(length(ctrl_only_edges)),
    "(", round(100 * length(ctrl_only_edges) / nrow(ctrl_edges), 1), "% of CTRL)\n")
cat("  Gained edges:", scales::comma(length(inv4m_only_edges)), "\n\n")

cat("SIGN CONCORDANCE (signed network validation)\n")
cat("  Shared edges with same sign:", scales::comma(n_truly_preserved),
    "(", round(100 * n_truly_preserved / length(shared_edges), 1), "%)\n")
cat("  Shared edges with sign flip:", scales::comma(n_sign_flips),
    "(", round(100 * n_sign_flips / length(shared_edges), 1), "%)\n")
cat("  MaizeNetome edges - same sign:", n_mn_truly_preserved, "/", nrow(shared_mn_correlations),
    "(", round(100 * n_mn_truly_preserved / nrow(shared_mn_correlations), 1), "%)\n\n")

if (fisher_result$p.value < 0.05) {
  cat("MAIZENETOME VALIDATION\n")
  cat("  Reference edges are", ifelse(fisher_result$estimate > 1, "MORE", "LESS"),
      "likely to be preserved\n")
  cat("  Odds ratio:", round(fisher_result$estimate, 2),
      "(p =", signif(fisher_result$p.value, 3), ")\n")
}

cat("\n")
cat("=" %>% rep(60) %>% paste(collapse = ""), "\n")
```

---

# Session Information

```{r session_info}
sessionInfo()
```
