---
title: "Differential Expression Volcano Plot Analysis - Inv4m Effect"
author: "Fausto Rodriguez Zapata"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 300
)

# Setup project paths
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Overview

**Purpose**: Generate volcano plot for Inv4m genotype effect on differential
gene expression, highlighting genes selected by Mahalanobis distance outlier
detection.

**Input**: CSV file containing differential expression results with Mahalanobis
outlier identification (`is_selected_DEG` column) from the inversion paper
DE analysis.

**Output**:
- Inv4m volcano plot (Panel C for Figure 3)
- Standalone volcano figure for manuscript

---

# Setup

## Load Libraries
```{r libraries}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(ggpubr)
```

---

# Data Import and Filtering

**Purpose**: Load differential expression results and filter to Inv4m genotype
effect.

**Approach**: Import CSV and subset to Inv4m genotype effects. The input
contains pre-calculated Mahalanobis distances and outlier flags from upstream
analysis.

**Expected outcome**: Filtered dataset with Inv4m predictor, retaining all
columns including `is_selected_DEG` for gene labeling.
```{r import_filter}

predictor_order <- "Inv4m"

# Load DE results with Mahalanobis outlier detection
effects_df <- read.csv(file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv")) %>%
  filter(predictor %in% predictor_order) %>%
  droplevels()

# Set predictor factor

effects_df$predictor <- factor(effects_df$predictor, levels = predictor_order)

# Diagnostics
cat("Predictors in dataset (ordered):", levels(effects_df$predictor), "\n")
cat("Total genes:", nrow(effects_df), "\n")
cat("Genes per predictor:\n")
print(table(effects_df$predictor))
```

---

# Gene Selection for Labeling

**Purpose**: Identify top genes to label in volcano plots based on 
Mahalanobis distance ranking.

**Approach**: For each predictor and regulation direction, select the top N 
genes (default N=10) that are marked as `is_selected_DEG` and have 
locus_label annotation in selected_DEGs_curated_locus_label.csv. Load curated 
labels from the external file, coalesce with existing labels, then filter for 
genes with valid annotations to label in the plot.

**Expected outcome**: Data frame of genes to label, with custom label text 
from curated annotations.
```{r select_labels}
# Load curated labels from intermediate results
curated <- read.csv(file.path(paths$intermediate, "selected_DEGs_leaf_interaction_model.csv")) %>%
  select(
    regulation,
    predictor,
    gene,
    locus_symbol,
    locus_label,
    desc_merged,
    mahalanobis,
    logFC,
    neglogP
  )

# Set predictor order for curated data
curated$predictor <- factor(curated$predictor, levels = predictor_order)

# Define number of labels per predictor and regulation
label_counts <- data.frame(
  predictor = c("Inv4m", "Inv4m"),
  regulation = c("Downregulated", "Upregulated"),
  n_labels = c(15, 15),
  stringsAsFactors = FALSE
)

# Select genes for labeling based on specified counts
to_plot <- curated %>%
  ungroup() %>%
  filter(!is.na(locus_label) & locus_label != "") %>%
  arrange(regulation, predictor, -mahalanobis) %>%
  left_join(label_counts, by = c("predictor", "regulation")) %>%
  group_by(regulation, predictor) %>%
  mutate(row_num = row_number()) %>%
  filter(row_num <= n_labels) %>%
  select(-row_num, -n_labels) %>%
  arrange(-mahalanobis) %>%
  ungroup() %>%
  mutate(predictor = factor(predictor, levels = predictor_order)) %>%  
  data.frame()

# Diagnostics
cat("Total curated genes:", nrow(curated), "\n")
cat("Genes with locus_label:", sum(!is.na(curated$locus_label) & curated$locus_label != ""), "\n")
cat("\nSelected DEGs by regulation and predictor:\n")
with(curated, print(table(regulation, predictor)))
cat("\nGenes selected for labeling:", nrow(to_plot), "\n")
cat("Labels per predictor/regulation:\n")
with(to_plot, print(table(regulation, predictor)))

# Show configured label counts
cat("\nConfigured label counts:\n")
print(label_counts)
```

---

# Visualization

## Inv4m Volcano Plot

**Purpose**: Create volcano plot showing logFC vs. FDR for Inv4m genotype effect.

**Approach**: Use ggplot with:
- Points colored by regulation status (down/below threshold/up)
- Text labels for top Mahalanobis-selected genes
- Expression formatting for axis labels

**Expected outcome**: Single-panel figure saved as Panel C for Figure 3.
```{r volcano_inv4m, fig.width=8, fig.height=8}
# Custom formatting function for y-axis
scaleFUN <- function(x) sprintf("%.0f", x)

# Set regulation based on high-confidence DEG flag
effects_df$regulation[!effects_df$is_hiconf_DEG] <- "Unregulated"

# Generate Inv4m volcano plot
volcano_inv4m <- effects_df %>%
  ggplot(aes(x = logFC, y = neglogP, col = regulation)) +
  ylab(expression(-log[10](italic(FDR)))) +
  xlab(expression(log[2]("Fold Change"))) +
  scale_y_continuous(labels = scaleFUN, expand = expansion(mult = c(0, 0.1))) +
  geom_point(alpha = 0.3, size = 1) +
  geom_text_repel(
    data = to_plot,
    aes(label = locus_label),
    color = "black",
    fontface = "bold.italic",
    segment.color = "grey",
    force = 2,
    size = 5,
    min.segment.length = 0,
    max.overlaps = 50
  ) +
  scale_color_manual(
    name = "",
    values = c("#00AFBB", "grey", "#bb0c00"),
    labels = c("Downregulated", "Below threshold", "Upregulated")
  ) +
  guides(color = guide_legend(
    reverse = TRUE,
    override.aes = list(size = 5)
  )) +
  theme_classic2(base_size = 20) +
  theme(
    plot.title = element_blank(),
    legend.position = c(0.85, 0.90),
    legend.background = element_rect(fill = "transparent"),
    legend.spacing = unit(0, "line"),
    legend.box.spacing = unit(0, "line"),
    legend.text = element_text(size = 14),
    legend.direction = "vertical"
  )

# Save as Panel C for Figure 3
ggsave(
  file.path(paths$figures, "panels", "panel_C_volcano.png"),
  plot = volcano_inv4m,
  width = 6, height = 6, dpi = 300, bg = "white"
)

# Also save standalone figure
ggsave(
  file.path(paths$figures, "volcano_inv4m.png"),
  plot = volcano_inv4m,
  width = 8, height = 8, dpi = 150, bg = "white"
)

volcano_inv4m
```

# Summary

```{r summary}
cat("Inv4m Volcano Plot Analysis Complete\n")
cat("=====================================\n")
cat("Total genes analyzed:", nrow(effects_df), "\n")
cat("Genes labeled in plot:", nrow(to_plot), "\n")
cat("\nOutput files:\n")
cat("  - Panel C: panels/panel_C_volcano.png\n")
cat("  - Standalone: volcano_inv4m.png\n")
```

---

# Session Information
```{r session_info}
sessionInfo()
```