---
title: "WGCNA Module Perturbation Test"
author: "F. V. Rodriguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = TRUE
)
knitr::opts_knit$set(root.dir = here::here())
```

# Overview

This notebook tests whether the intramodular connectivity loss observed in
genotype-response modules (Inv4m DEGs) exceeds what random genotype label
shuffling produces. The topology control uses leaf-gradient genes
(photosynthesis + senescence hub genes from Ojeda 2026) that have real
coexpression structure but are not differentially expressed between genotypes.

Both module sets come from separate consensus WGCNA pipeline runs (1000
iterations each), with bootstrap-validated stability (LFP from step 05).

**Gene set terminology:**

- **Genotype-response genes** -- differentially expressed between CTRL and
  Inv4m plants; coexpression driven by the Inv4m introgression.
- **Leaf-gradient genes** -- photosynthesis and senescence hub genes
  (Ojeda 2026); coexpression driven by the leaf developmental gradient,
  independent of Inv4m genotype.

**Prediction:** Genotype-response modules show significant kWithin loss;
leaf-gradient modules do not.

# Setup

```{r load_packages}
library(here)
library(tidyverse)
library(WGCNA)
library(ggridges)
library(patchwork)

allowWGCNAThreads()
set.seed(42)
```

```{r config}
N_PERM <- 10000

# Genotype-response run
GR_RUN <- here("results/inversion_paper/field_perturbation/run_20251231_201332")
# Leaf-gradient run
LG_RUN <- here("results/inversion_paper/field_perturbation/run_leaf_gradient")

source(here("scripts/utils/setup_paths.R"))
paths <- setup_project_paths("inversion_paper")
```

# Load Pre-computed Data

```{r load_genotype_response}
# Genotype-response consensus modules (from CTRL reference network)
gr_modules <- read_csv(
  file.path(GR_RUN, "04_consensus_networks", "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)
gr_lfp <- read_csv(
  file.path(GR_RUN, "05_bootstrap_support", "ctrl", "module_support.csv"),
  show_col_types = FALSE
)
gr_preservation <- read_csv(
  file.path(GR_RUN, "06_preservation", "zsummary_table.csv"),
  show_col_types = FALSE
)

# Expression and metadata (shared across runs)
expr_gr <- readRDS(file.path(GR_RUN, "02_gene_filter", "filtered_expression.rds"))
sample_meta <- read_csv(
  file.path(GR_RUN, "01_data_prep", "sample_metadata.csv"),
  show_col_types = FALSE
)
SOFT_POWER <- readRDS(
  file.path(GR_RUN, "03_reference_network", "ctrl", "selected_power.rds")
)

# GO annotations
go_bp_file <- file.path(GR_RUN, "07_module_annotation", "top_BP_per_module_rrvgo.csv")
gr_go_terms <- if (file.exists(go_bp_file)) {
  go_bp <- read_csv(go_bp_file, show_col_types = FALSE)
  setNames(str_trunc(go_bp$Description, 35), go_bp$module_color)
} else {
  NULL
}

cat("Genotype-response:", nrow(gr_modules), "genes,",
    n_distinct(gr_modules$color[gr_modules$color != "grey"]), "modules\n")
cat("Soft power:", SOFT_POWER, "\n")
```

```{r load_leaf_gradient}
# Leaf-gradient consensus modules (from CTRL reference network)
lg_modules <- read_csv(
  file.path(LG_RUN, "04_consensus_networks", "ctrl", "consensus_modules.csv"),
  show_col_types = FALSE
)
lg_lfp <- read_csv(
  file.path(LG_RUN, "05_bootstrap_support", "ctrl", "module_support.csv"),
  show_col_types = FALSE
)
lg_preservation <- read_csv(
  file.path(LG_RUN, "06_preservation", "zsummary_table.csv"),
  show_col_types = FALSE
)

# Leaf-gradient expression matrix
expr_lg <- readRDS(file.path(LG_RUN, "02_gene_filter", "filtered_expression.rds"))

# Remap leaf-gradient colors to avoid collisions with genotype-response
lg_color_remap <- c(
  turquoise = "sky",
  blue      = "peach",
  brown     = "jade",
  yellow    = "coral",
  green     = "lavender",
  grey      = "grey"
)
lg_color_hex <- c(
  sky      = "#5DA5DA",
  peach    = "#FAA43A",
  jade     = "#60BD68",
  coral    = "#F15854",
  lavender = "#B276B2",
  grey     = "grey"
)

cat("Leaf-gradient:", nrow(lg_modules), "genes,",
    n_distinct(lg_modules$color[lg_modules$color != "grey"]), "modules\n")
```

## GO Enrichment for Leaf-gradient Modules

Step 07 was skipped for the leaf-gradient run (DEG-specific inputs).
Run GO BP enrichment here using clusterProfiler + rrvgo, matching the
approach used for genotype-response modules.

```{r lg_go_enrichment, cache=TRUE}
library(clusterProfiler)
library(rrvgo)
library(AnnotationHub)

TERM2GENE <- readRDS(here("data", "TERM2GENE.rds"))
TERM2NAME <- readRDS(here("data", "TERM2NAME.rds"))

# Universe = all genes in the full expression matrix
full_expr <- readRDS(file.path(GR_RUN, "01_data_prep", "combined_expression.rds"))
if (inherits(full_expr, "EList")) {
  go_universe <- rownames(full_expr$E)
} else {
  go_universe <- rownames(as.matrix(full_expr))
}
rm(full_expr)

# Gene lists per leaf-gradient module (use remapped colors, exclude grey)
lg_remapped <- lg_color_remap[lg_modules$color]
names(lg_remapped) <- lg_modules$gene
lg_gene_lists <- split(
  names(lg_remapped[lg_remapped != "grey"]),
  lg_remapped[lg_remapped != "grey"]
)

cat("Leaf-gradient modules for GO:", length(lg_gene_lists), "\n")
cat("Module sizes:", paste(names(lg_gene_lists),
    sapply(lg_gene_lists, length), sep = "=", collapse = ", "), "\n")

lg_ego <- function(x) {
  enricher(
    gene = x,
    pvalueCutoff = 0.05,
    pAdjustMethod = "none",
    universe = go_universe,
    minGSSize = 10,
    maxGSSize = 500,
    qvalueCutoff = 0.2,
    TERM2GENE = TERM2GENE[TERM2GENE$GO %in% TERM2NAME[["BP"]]$go_id, ],
    TERM2NAME = TERM2NAME[["BP"]]
  )
}

lg_compGO <- compareCluster(
  geneCluster = lg_gene_lists,
  fun = lg_ego
)

lg_go_results <- as.data.frame(lg_compGO@compareClusterResult) |>
  dplyr::rename(module_color = Cluster) |>
  arrange(module_color, pvalue)

cat("Total enriched BP terms:", nrow(lg_go_results), "\n")
cat("Terms per module:\n")
print(table(lg_go_results$module_color))

# rrvgo semantic reduction
ah <- AnnotationHub()
org.Zm.eg.db <- ah[["AH119718"]]

lg_go_filtered <- lg_go_results |>
  filter(Count > 1) |>
  group_by(module_color) |>
  arrange(pvalue, .by_group = TRUE) |>
  ungroup()

lg_go_ids <- unique(lg_go_results$ID)
if (length(lg_go_ids) > 1) {
  simMatrix <- calculateSimMatrix(
    lg_go_ids,
    orgdb = org.Zm.eg.db,
    ont = "BP",
    method = "Rel"
  )
  scores <- setNames(-log10(lg_go_results$pvalue), lg_go_results$ID)
  reducedTerms <- reduceSimMatrix(
    simMatrix, scores, threshold = 0.7, orgdb = org.Zm.eg.db
  )
  lg_go_rrvgo <- lg_go_filtered |>
    filter(ID %in% reducedTerms$parent)
} else {
  lg_go_rrvgo <- lg_go_filtered
}

# Top 1 BP term per leaf-gradient module
lg_top_bp <- lg_go_rrvgo |>
  group_by(module_color) |>
  slice_min(pvalue, n = 1, with_ties = FALSE) |>
  ungroup() |>
  dplyr::select(module_color, ID, Description, pvalue, Count)

lg_go_terms <- setNames(
  str_trunc(lg_top_bp$Description, 35),
  lg_top_bp$module_color
)

cat("\nTop BP term per leaf-gradient module:\n")
print(lg_top_bp)
```

# Prepare Expression Matrices

```{r prepare_expression}
# Module color vectors
gr_colors <- setNames(gr_modules$color, gr_modules$gene)
gr_colors <- gr_colors[rownames(expr_gr)]

lg_colors <- lg_color_remap[lg_modules$color]
names(lg_colors) <- lg_modules$gene
lg_colors <- lg_colors[rownames(expr_lg)]

# Genotype labels
true_labels <- setNames(sample_meta$Genotype, sample_meta$sample_id)
all_samples <- sample_meta$sample_id
n_ctrl  <- sum(true_labels == "CTRL")
n_inv4m <- sum(true_labels == "Inv4m")
n_total <- length(true_labels)

# WGCNA format: samples x genes
datExpr_gr <- t(expr_gr[, all_samples])
datExpr_lg <- t(expr_lg[, all_samples])

cat("Genotype-response matrix:", nrow(datExpr_gr), "samples x",
    ncol(datExpr_gr), "genes\n")
cat("Leaf-gradient matrix:", nrow(datExpr_lg), "samples x",
    ncol(datExpr_lg), "genes\n")
cat("CTRL:", n_ctrl, "| Inv4m:", n_inv4m, "\n")
```

# Supplementary Figure S5: Bootstrap Support

Two-panel figure showing CTRL consensus module bootstrap support (LFP)
for both gene sets.

```{r bootstrap_support_figure, fig.width=10, fig.height=5.5}
STABILITY_STABLE   <- 0.70
STABILITY_MODERATE <- 0.50

# Combine both gene sets into one data frame
gr_support_df <- gr_lfp |>
  filter(color != "grey") |>
  mutate(
    display_color = color,
    fill_hex = color,
    gene_set = "Genotype Response DEGs"
  )

lg_support_df <- lg_lfp |>
  filter(color != "grey") |>
  mutate(
    display_color = lg_color_remap[color],
    fill_hex = lg_color_hex[lg_color_remap[color]],
    gene_set = "Leaf Gradient Hubs"
  )

support_df <- bind_rows(gr_support_df, lg_support_df) |>
  mutate(gene_set = recode(gene_set,
    "Genotype Response DEGs" = "Genotype Response",
    "Leaf Gradient Hubs" = "Leaf Gradient"))

p_support <- ggplot(
  support_df,
  aes(x = reorder(display_color, mean_lfp), y = mean_lfp, fill = fill_hex)
) +
  geom_col(color = "black", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = pmax(0, mean_lfp - se_lfp),
        ymax = pmin(1, mean_lfp + se_lfp)),
    width = 0.3
  ) +
  geom_hline(yintercept = STABILITY_STABLE, linetype = "dashed",
             color = "darkgreen") +
  geom_hline(yintercept = STABILITY_MODERATE, linetype = "dashed",
             color = "orange") +
  scale_fill_identity() +
  coord_flip() +
  facet_grid(gene_set ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 14) +
  theme(panel.spacing.y = unit(1, "lines"),
        strip.text.y = element_text(size = 14)) +
  labs(x = "Module", y = "Mean Support (LFP)")

print(p_support)

ggsave(file.path(paths$figures, "module_support_combined.pdf"),
       p_support, width = 10, height = 5.5)
ggsave(file.path(paths$figures, "module_support_combined.png"),
       p_support, width = 10, height = 8, dpi = 300)
```

# Helper Function

```{r helper_function}
compute_module_delta_k <- function(datExpr_a, datExpr_b, power, mod_colors) {
  adj_a <- adjacency(datExpr_a, power = power, type = "signed")
  adj_b <- adjacency(datExpr_b, power = power, type = "signed")

  k_a <- intramodularConnectivity(adj_a, mod_colors)
  k_b <- intramodularConnectivity(adj_b, mod_colors)

  delta_kWithin <- k_b$kWithin - k_a$kWithin
  delta_kOut    <- k_b$kOut - k_a$kOut

  module_ids <- unique(mod_colors[mod_colors != "grey" & !is.na(mod_colors)])
  list(
    kWithin = sapply(module_ids, function(m) {
      idx <- which(mod_colors == m)
      mean(delta_kWithin[idx], na.rm = TRUE)
    }),
    kOut = sapply(module_ids, function(m) {
      idx <- which(mod_colors == m)
      mean(delta_kOut[idx], na.rm = TRUE)
    })
  )
}
```

# Observed Delta kWithin

```{r observed}
ctrl_idx  <- which(true_labels == "CTRL")
inv4m_idx <- which(true_labels == "Inv4m")

obs_gr <- compute_module_delta_k(
  datExpr_gr[ctrl_idx, ], datExpr_gr[inv4m_idx, ],
  SOFT_POWER, gr_colors
)
observed_gr_kw  <- obs_gr$kWithin
observed_gr_out <- obs_gr$kOut

obs_lg <- compute_module_delta_k(
  datExpr_lg[ctrl_idx, ], datExpr_lg[inv4m_idx, ],
  SOFT_POWER, lg_colors
)
observed_lg_kw  <- obs_lg$kWithin
observed_lg_out <- obs_lg$kOut

cat("Observed mean delta kWithin per genotype-response module:\n")
print(round(sort(observed_gr_kw), 4))
cat("\nObserved mean delta kWithin per leaf-gradient module:\n")
print(round(sort(observed_lg_kw), 4))

cat("\nObserved mean delta kOut per genotype-response module:\n")
print(round(sort(observed_gr_out), 4))
cat("\nObserved mean delta kOut per leaf-gradient module:\n")
print(round(sort(observed_lg_out), 4))
```

# Genotype Label Permutation Test

Shuffle genotype labels `r format(N_PERM, big.mark=",")` times, keeping
group sizes fixed (`r n_ctrl` CTRL, `r n_inv4m` Inv4m). For each
permutation, recompute $\Delta k_{Within}$ and $\Delta k_{Out}$ for both
gene sets. Module definitions remain fixed throughout.

```{r permutation_loop, cache=TRUE}
null_gr_kw  <- matrix(NA, nrow = N_PERM, ncol = length(observed_gr_kw))
null_gr_out <- matrix(NA, nrow = N_PERM, ncol = length(observed_gr_out))
colnames(null_gr_kw)  <- names(observed_gr_kw)
colnames(null_gr_out) <- names(observed_gr_out)

null_lg_kw  <- matrix(NA, nrow = N_PERM, ncol = length(observed_lg_kw))
null_lg_out <- matrix(NA, nrow = N_PERM, ncol = length(observed_lg_out))
colnames(null_lg_kw)  <- names(observed_lg_kw)
colnames(null_lg_out) <- names(observed_lg_out)

for (i in seq_len(N_PERM)) {
  perm_idx <- sample(n_total)
  pseudo_ctrl  <- perm_idx[1:n_ctrl]
  pseudo_inv4m <- perm_idx[(n_ctrl + 1):n_total]

  perm_gr <- compute_module_delta_k(
    datExpr_gr[pseudo_ctrl, ], datExpr_gr[pseudo_inv4m, ],
    SOFT_POWER, gr_colors
  )
  null_gr_kw[i, ]  <- perm_gr$kWithin
  null_gr_out[i, ] <- perm_gr$kOut

  perm_lg <- compute_module_delta_k(
    datExpr_lg[pseudo_ctrl, ], datExpr_lg[pseudo_inv4m, ],
    SOFT_POWER, lg_colors
  )
  null_lg_kw[i, ]  <- perm_lg$kWithin
  null_lg_out[i, ] <- perm_lg$kOut
}
```

# Results

## ECDF P-values

One-tailed test: fraction of permuted values $\leq$ observed (testing for
connectivity loss more extreme than the null).

```{r ecdf_pvalues}
# kWithin
pvals_gr_kw <- sapply(names(observed_gr_kw), function(m) {
  mean(null_gr_kw[, m] <= observed_gr_kw[m])
})
pvals_lg_kw <- sapply(names(observed_lg_kw), function(m) {
  mean(null_lg_kw[, m] <= observed_lg_kw[m])
})

# kOut (one-tailed: test for gain in inter-modular connectivity)
pvals_gr_out <- sapply(names(observed_gr_out), function(m) {
  mean(null_gr_out[, m] >= observed_gr_out[m])
})
pvals_lg_out <- sapply(names(observed_lg_out), function(m) {
  mean(null_lg_out[, m] >= observed_lg_out[m])
})
```

### Genotype-response Modules

```{r results_gr}
results_gr <- tibble(
  module = names(observed_gr_kw),
  gene_set = "Genotype Response DEGs",
  n_genes = sapply(names(observed_gr_kw), function(m)
    sum(gr_colors == m, na.rm = TRUE)),
  observed_mean_delta_k = round(observed_gr_kw, 4),
  null_mean = round(colMeans(null_gr_kw), 4),
  null_sd = round(apply(null_gr_kw, 2, sd), 4),
  ecdf_pvalue = round(pvals_gr_kw, 4)
) |>
  arrange(ecdf_pvalue)

results_gr
```

### Leaf-gradient Modules

```{r results_lg}
results_lg <- tibble(
  module = names(observed_lg_kw),
  gene_set = "Leaf Gradient Hubs",
  n_genes = sapply(names(observed_lg_kw), function(m)
    sum(lg_colors == m)),
  observed_mean_delta_k = round(observed_lg_kw, 4),
  null_mean = round(colMeans(null_lg_kw), 4),
  null_sd = round(apply(null_lg_kw, 2, sd), 4),
  ecdf_pvalue = round(pvals_lg_kw, 4)
) |>
  arrange(ecdf_pvalue)

results_lg
```

### BH FDR Correction Across All Tests

```{r fdr_correction}
all_results <- bind_rows(results_gr, results_lg)

# Joint FDR across both gene sets
n_gr <- length(pvals_gr_kw)
n_lg <- length(pvals_lg_kw)
all_kw_pvals <- c(pvals_gr_kw, pvals_lg_kw)
all_kw_fdr <- p.adjust(all_kw_pvals, method = "BH")
kw_fdr_gr <- setNames(all_kw_fdr[1:n_gr], names(pvals_gr_kw))
kw_fdr_lg <- setNames(all_kw_fdr[(n_gr + 1):(n_gr + n_lg)], names(pvals_lg_kw))

# kOut FDR
n_gr_out <- length(pvals_gr_out)
n_lg_out <- length(pvals_lg_out)
all_kout_pvals <- c(pvals_gr_out, pvals_lg_out)
all_kout_fdr <- p.adjust(all_kout_pvals, method = "BH")
kout_fdr_gr <- setNames(all_kout_fdr[1:n_gr_out], names(pvals_gr_out))
kout_fdr_lg <- setNames(all_kout_fdr[(n_gr_out + 1):(n_gr_out + n_lg_out)],
                         names(pvals_lg_out))

all_results <- all_results |>
  mutate(
    fdr = c(kw_fdr_gr[results_gr$module], kw_fdr_lg[results_lg$module]),
    sig = ifelse(fdr < 0.05, "*", "")
  ) |>
  arrange(ecdf_pvalue)

cat("Significant at FDR < 0.05:\n")
all_results |>
  filter(fdr < 0.05) |>
  dplyr::select(module, gene_set, n_genes, observed_mean_delta_k,
                ecdf_pvalue, fdr) |>
  print()

cat("\nAll leaf-gradient modules:\n")
all_results |>
  filter(gene_set == "Leaf Gradient Hubs") |>
  dplyr::select(module, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()
```

## Save Permutation Results

```{r save_results}
OUTPUT_DIR <- file.path(GR_RUN, "06_preservation")

write_csv(all_results,
          file.path(OUTPUT_DIR,
                    "permutation_delta_kWithin_results.csv"))
saveRDS(
  list(
    observed_gr_kw = observed_gr_kw,
    observed_lg_kw = observed_lg_kw,
    observed_gr_out = observed_gr_out,
    observed_lg_out = observed_lg_out,
    null_gr_kw = null_gr_kw,
    null_lg_kw = null_lg_kw,
    null_gr_out = null_gr_out,
    null_lg_out = null_lg_out,
    pvals_gr_kw = pvals_gr_kw,
    pvals_lg_kw = pvals_lg_kw,
    pvals_gr_out = pvals_gr_out,
    pvals_lg_out = pvals_lg_out,
    gr_colors = gr_colors,
    lg_colors = lg_colors
  ),
  file.path(OUTPUT_DIR, "permutation_delta_kWithin.rds")
)
```

# Visualization

## Z-score Ridge Plots (kWithin)

Two-panel figure comparing connectivity disruption in genotype-response
modules versus leaf-gradient modules. Z-scores are computed per module as
$(observed - mean_{null}) / sd_{null}$. Red lines mark observed values.

```{r zscore_kwithin, fig.width=10, fig.height=7}
# Z-score genotype-response null distributions
gr_null_mean <- colMeans(null_gr_kw)
gr_null_sd   <- apply(null_gr_kw, 2, sd)
gr_null_z    <- null_gr_kw
for (m in colnames(gr_null_z)) {
  gr_null_z[, m] <- (gr_null_z[, m] - gr_null_mean[m]) / gr_null_sd[m]
}
gr_obs_z <- (observed_gr_kw - gr_null_mean) / gr_null_sd

# Z-score leaf-gradient null distributions
lg_null_mean <- colMeans(null_lg_kw)
lg_null_sd   <- apply(null_lg_kw, 2, sd)
lg_null_z    <- null_lg_kw
for (m in colnames(lg_null_z)) {
  lg_null_z[, m] <- (lg_null_z[, m] - lg_null_mean[m]) / lg_null_sd[m]
}
lg_obs_z <- (observed_lg_kw - lg_null_mean) / lg_null_sd

# Module sizes for y-axis labels
gr_mod_sizes <- sapply(colnames(gr_null_z), function(m)
  sum(gr_colors == m, na.rm = TRUE))
lg_mod_sizes <- sapply(colnames(lg_null_z), function(m)
  sum(lg_colors == m))

# Build combined density data
gr_density <- map_dfr(colnames(gr_null_z), function(m) {
  tibble(module = sprintf("%s (n=%d)", m, gr_mod_sizes[m]),
         fill_hex = m, z_value = gr_null_z[, m],
         gene_set = "Genotype Response DEGs")
})
lg_density <- map_dfr(colnames(lg_null_z), function(m) {
  tibble(module = sprintf("%s (n=%d)", m, lg_mod_sizes[m]),
         fill_hex = lg_color_hex[m], z_value = lg_null_z[, m],
         gene_set = "Leaf Gradient Hubs")
})
density_kw <- bind_rows(gr_density, lg_density)

# Build combined points data
gr_points <- tibble(
  module = sprintf("%s (n=%d)", names(gr_obs_z), gr_mod_sizes[names(gr_obs_z)]),
  fill_hex = names(gr_obs_z),
  z_observed = gr_obs_z, fdr = kw_fdr_gr[names(gr_obs_z)],
  gene_set = "Genotype Response DEGs"
)
lg_points <- tibble(
  module = sprintf("%s (n=%d)", names(lg_obs_z), lg_mod_sizes[names(lg_obs_z)]),
  fill_hex = lg_color_hex[names(lg_obs_z)],
  z_observed = lg_obs_z, fdr = kw_fdr_lg[names(lg_obs_z)],
  gene_set = "Leaf Gradient Hubs"
)
points_kw <- bind_rows(gr_points, lg_points)

# Order modules by z_observed WITHIN each gene_set, compute per-facet rank
points_kw <- points_kw |>
  group_by(gene_set) |>
  arrange(z_observed, .by_group = TRUE) |>
  mutate(facet_rank = row_number()) |>
  ungroup()

# Set factor levels per facet ordering
mod_order <- points_kw |> arrange(gene_set, facet_rank) |> pull(module)
density_kw$module <- factor(density_kw$module, levels = mod_order)
points_kw$module  <- factor(points_kw$module, levels = mod_order)

# Propagate facet_rank to density data
density_kw <- density_kw |>
  left_join(points_kw |> dplyr::select(module, gene_set, facet_rank) |> distinct(),
            by = c("module", "gene_set"))

# Build fill color map (module label -> hex)
fill_map <- setNames(points_kw$fill_hex, as.character(points_kw$module))

# GO term annotations
gr_go_df <- if (!is.null(gr_go_terms)) {
  tibble(
    module = sprintf("%s (n=%d)", names(gr_go_terms),
                     gr_mod_sizes[names(gr_go_terms)]),
    go_desc = unname(gr_go_terms), gene_set = "Genotype Response DEGs"
  ) |> filter(module %in% levels(density_kw$module))
}
lg_go_df <- if (!is.null(lg_go_terms)) {
  tibble(
    module = sprintf("%s (n=%d)", names(lg_go_terms),
                     lg_mod_sizes[names(lg_go_terms)]),
    go_desc = unname(lg_go_terms), gene_set = "Leaf Gradient Hubs"
  ) |> filter(module %in% levels(density_kw$module))
}
go_df <- bind_rows(gr_go_df, lg_go_df) |>
  mutate(module = factor(module, levels = levels(density_kw$module))) |>
  left_join(points_kw |> dplyr::select(module, gene_set, facet_rank) |> distinct(),
            by = c("module", "gene_set")) |>
  mutate(y_pos = facet_rank + 0.35)

p_kw <- ggplot(density_kw, aes(x = z_value, y = module, fill = module)) +
  geom_density_ridges(color = "grey40", alpha = 0.7,
                      scale = 0.9, rel_min_height = 0.01) +
  scale_fill_manual(values = fill_map) +
  guides(fill = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_segment(
    data = points_kw,
    aes(x = z_observed, xend = z_observed,
        y = facet_rank, yend = facet_rank + 0.8),
    color = "red", linewidth = 0.8, inherit.aes = FALSE
  ) +
  geom_text(
    data = points_kw |> filter(fdr < 0.05),
    aes(x = z_observed - 0.6, y = facet_rank + 0.35,
        label = sprintf("FDR=%.3f", fdr)),
    color = "red", size = 3, hjust = 0.5, inherit.aes = FALSE
  ) +
  geom_text(
    data = go_df,
    aes(x = 2.5, y = y_pos, label = go_desc),
    hjust = 0, size = 14 * 0.8 / .pt, fontface = "italic",
    inherit.aes = FALSE
  ) +
  scale_x_continuous(limits = c(NA, 5), breaks = seq(-10, 10, 2)) +
  facet_grid(gene_set ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 14) +
  theme(panel.spacing.y = unit(1, "lines"),
        strip.text.y = element_text(size = 14)) +
  labs(y = NULL, x = expression(Delta * k[Within] ~ "(Inv4m - CTRL, Z-score)"))

print(p_kw)

ggsave(file.path(OUTPUT_DIR, "zscore_density_kWithin.pdf"),
       p_kw, width = 10, height = 7)
ggsave(file.path(OUTPUT_DIR, "zscore_density_kWithin.png"),
       p_kw, width = 10, height = 10, dpi = 300)
```

## Z-score Ridge Plots (kOut)

Same approach for inter-modular connectivity ($\Delta k_{Out}$).

```{r zscore_kout, fig.width=10, fig.height=7}
# Z-score kOut
gr_null_mean_out <- colMeans(null_gr_out)
gr_null_sd_out   <- apply(null_gr_out, 2, sd)
gr_null_z_out    <- null_gr_out
for (m in colnames(gr_null_z_out)) {
  gr_null_z_out[, m] <- (gr_null_z_out[, m] - gr_null_mean_out[m]) / gr_null_sd_out[m]
}
gr_obs_z_out <- (observed_gr_out - gr_null_mean_out) / gr_null_sd_out

lg_null_mean_out <- colMeans(null_lg_out)
lg_null_sd_out   <- apply(null_lg_out, 2, sd)
lg_null_z_out    <- null_lg_out
for (m in colnames(lg_null_z_out)) {
  lg_null_z_out[, m] <- (lg_null_z_out[, m] - lg_null_mean_out[m]) / lg_null_sd_out[m]
}
lg_obs_z_out <- (observed_lg_out - lg_null_mean_out) / lg_null_sd_out

# Build combined density data
gr_density_out <- map_dfr(colnames(gr_null_z_out), function(m) {
  tibble(module = sprintf("%s (n=%d)", m, gr_mod_sizes[m]),
         fill_hex = m, z_value = gr_null_z_out[, m],
         gene_set = "Genotype Response DEGs")
})
lg_density_out <- map_dfr(colnames(lg_null_z_out), function(m) {
  tibble(module = sprintf("%s (n=%d)", m, lg_mod_sizes[m]),
         fill_hex = lg_color_hex[m], z_value = lg_null_z_out[, m],
         gene_set = "Leaf Gradient Hubs")
})
density_kout <- bind_rows(gr_density_out, lg_density_out)

# Build combined points data
gr_points_out <- tibble(
  module = sprintf("%s (n=%d)", names(gr_obs_z_out), gr_mod_sizes[names(gr_obs_z_out)]),
  fill_hex = names(gr_obs_z_out),
  z_observed = gr_obs_z_out, fdr = kout_fdr_gr[names(gr_obs_z_out)],
  gene_set = "Genotype Response DEGs"
)
lg_points_out <- tibble(
  module = sprintf("%s (n=%d)", names(lg_obs_z_out), lg_mod_sizes[names(lg_obs_z_out)]),
  fill_hex = lg_color_hex[names(lg_obs_z_out)],
  z_observed = lg_obs_z_out, fdr = kout_fdr_lg[names(lg_obs_z_out)],
  gene_set = "Leaf Gradient Hubs"
)
points_kout <- bind_rows(gr_points_out, lg_points_out)

# Order modules by z_observed WITHIN each gene_set, compute per-facet rank
points_kout <- points_kout |>
  group_by(gene_set) |>
  arrange(z_observed, .by_group = TRUE) |>
  mutate(facet_rank = row_number()) |>
  ungroup()

mod_order_out <- points_kout |> arrange(gene_set, facet_rank) |> pull(module)
density_kout$module <- factor(density_kout$module, levels = mod_order_out)
points_kout$module  <- factor(points_kout$module, levels = mod_order_out)

density_kout <- density_kout |>
  left_join(points_kout |> dplyr::select(module, gene_set, facet_rank) |> distinct(),
            by = c("module", "gene_set"))

fill_map_out <- setNames(points_kout$fill_hex, as.character(points_kout$module))

# GO annotations with per-facet rank
go_df_out <- bind_rows(gr_go_df, lg_go_df) |>
  mutate(module = factor(module, levels = levels(density_kout$module))) |>
  left_join(points_kout |> dplyr::select(module, gene_set, facet_rank) |> distinct(),
            by = c("module", "gene_set")) |>
  mutate(y_pos = facet_rank + 0.35)

p_kout <- ggplot(density_kout, aes(x = z_value, y = module, fill = module)) +
  geom_density_ridges(color = "grey40", alpha = 0.7,
                      scale = 0.9, rel_min_height = 0.01) +
  scale_fill_manual(values = fill_map_out) +
  guides(fill = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_segment(
    data = points_kout,
    aes(x = z_observed, xend = z_observed,
        y = facet_rank, yend = facet_rank + 0.8),
    color = "red", linewidth = 0.8, inherit.aes = FALSE
  ) +
  geom_text(
    data = points_kout |> filter(fdr < 0.05),
    aes(x = z_observed - 0.6, y = facet_rank + 0.35,
        label = sprintf("FDR=%.3f", fdr)),
    color = "red", size = 3, hjust = 0.5, inherit.aes = FALSE
  ) +
  geom_text(
    data = go_df_out,
    aes(x = 2.5, y = y_pos, label = go_desc),
    hjust = 0, size = 14 * 0.8 / .pt, fontface = "italic",
    inherit.aes = FALSE
  ) +
  scale_x_continuous(limits = c(NA, 5), breaks = seq(-10, 10, 2)) +
  facet_grid(gene_set ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 14) +
  theme(panel.spacing.y = unit(1, "lines"),
        strip.text.y = element_text(size = 14)) +
  labs(y = NULL, x = expression(Delta * k[Out] ~ "(Inv4m - CTRL, Z-score)"))

print(p_kout)

ggsave(file.path(OUTPUT_DIR, "zscore_density_kOut.pdf"),
       p_kout, width = 10, height = 7)
ggsave(file.path(OUTPUT_DIR, "zscore_density_kOut.png"),
       p_kout, width = 10, height = 7, dpi = 300)
```

## kOut: CTRL vs Inv4m Gene-level Scatter

Each point is a gene; axes are kOut in CTRL vs kOut in Inv4m. Points
above the diagonal gained inter-modular connectivity. Colored by module.

```{r kout_scatter, fig.width=8, fig.height=8}
# Gene-level kOut from genotype-response adjacencies
adj_ctrl_gr  <- adjacency(datExpr_gr[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_gr <- adjacency(datExpr_gr[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_gr  <- intramodularConnectivity(adj_ctrl_gr, gr_colors)
k_inv4m_gr <- intramodularConnectivity(adj_inv4m_gr, gr_colors)

kout_scatter <- tibble(
  gene = rownames(k_ctrl_gr),
  module = gr_colors[rownames(k_ctrl_gr)],
  kOut_CTRL = k_ctrl_gr$kOut,
  kOut_Inv4m = k_inv4m_gr$kOut
) |>
  filter(module != "grey", !is.na(module))

p_kout_scatter <- ggplot(
  kout_scatter,
  aes(x = kOut_CTRL, y = kOut_Inv4m, color = module)
) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50") +
  scale_color_identity() +
  theme_minimal(base_size = 14) +
  labs(
    x = expression(k[Out] ~ "(CTRL)"),
    y = expression(k[Out] ~ "(Inv4m)"),
    title = "Inter-modular connectivity: CTRL vs Inv4m"
  ) +
  coord_equal()

print(p_kout_scatter)

cat("Genes above diagonal (kOut gain):",
    sum(kout_scatter$kOut_Inv4m > kout_scatter$kOut_CTRL),
    "of", nrow(kout_scatter), "\n")
wt_kout <- wilcox.test(kout_scatter$kOut_Inv4m,
                        kout_scatter$kOut_CTRL,
                        paired = TRUE)
cat("Paired Wilcoxon p:", format.pval(wt_kout$p.value), "\n")

ggsave(file.path(OUTPUT_DIR, "kOut_CTRL_vs_Inv4m_scatter.pdf"),
       p_kout_scatter, width = 8, height = 8)
ggsave(file.path(OUTPUT_DIR, "kOut_CTRL_vs_Inv4m_scatter.png"),
       p_kout_scatter, width = 8, height = 8, dpi = 300)
```

## |logFC| vs Delta kWithin

Scatterplot of Inv4m expression effect size against connectivity change,
for genotype-response and leaf-gradient genes separately.

```{r logfc_vs_delta_k, fig.width=14, fig.height=4}
# Load DEG effects
effects <- read_csv(
  file.path(paths$intermediate, "predictor_effects_leaf_interaction_model.csv"),
  show_col_types = FALSE
)
inv4m_effects <- effects |>
  filter(grepl("Inv4m", predictor), !grepl(":", predictor)) |>
  dplyr::select(gene, logFC, adj.P.Val)

# Gene-level delta kWithin for genotype-response genes
adj_ctrl_gr  <- adjacency(datExpr_gr[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_gr <- adjacency(datExpr_gr[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_gr  <- intramodularConnectivity(adj_ctrl_gr, gr_colors)
k_inv4m_gr <- intramodularConnectivity(adj_inv4m_gr, gr_colors)

gr_dk <- tibble(
  gene = rownames(k_ctrl_gr),
  delta_kWithin = k_inv4m_gr$kWithin - k_ctrl_gr$kWithin,
  group = "Genotype Response DEGs"
)

# Gene-level delta kWithin for leaf-gradient genes
adj_ctrl_lg  <- adjacency(datExpr_lg[ctrl_idx, ], power = SOFT_POWER, type = "signed")
adj_inv4m_lg <- adjacency(datExpr_lg[inv4m_idx, ], power = SOFT_POWER, type = "signed")
k_ctrl_lg  <- intramodularConnectivity(adj_ctrl_lg, lg_colors)
k_inv4m_lg <- intramodularConnectivity(adj_inv4m_lg, lg_colors)

lg_dk <- tibble(
  gene = names(lg_colors),
  delta_kWithin = k_inv4m_lg$kWithin - k_ctrl_lg$kWithin,
  group = "Leaf Gradient Hubs"
)

# Merge with logFC
scatter_df <- bind_rows(gr_dk, lg_dk) |>
  left_join(inv4m_effects |> dplyr::select(gene, logFC, adj.P.Val), by = "gene") |>
  mutate(
    abs_logFC = abs(logFC),
    neg_log10p = -log10(adj.P.Val)
  ) |>
  filter(!is.na(logFC))

# Correlations per group
cor_logfc <- scatter_df |>
  group_by(group) |>
  summarise(rho = cor(abs_logFC, delta_kWithin, method = "spearman"),
            .groups = "drop")
cat("Spearman correlations (|logFC| vs delta_k):\n")
print(cor_logfc)

# Panel A: |logFC| vs delta kWithin
p_logfc <- ggplot(scatter_df, aes(x = abs_logFC, y = delta_kWithin, color = group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Genotype Response DEGs" = "grey40",
                                "Leaf Gradient Hubs" = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("|" * log[2] * "FC" * "| (Inv4m vs CTRL)"),
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)"),
    color = NULL
  ) +
  theme(legend.position = c(0.2, 0.15))

# Panel B: -log10P vs delta kWithin
p_pval <- ggplot(scatter_df, aes(x = neg_log10p, y = delta_kWithin, color = group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Genotype Response DEGs" = "grey40",
                                "Leaf Gradient Hubs" = "steelblue")) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("-" * log[10] * "(FDR)"),
    y = expression(Delta ~ k[Within] ~ "(Inv4m - CTRL)"),
    color = NULL
  ) +
  theme(legend.position = "none")

p_combined <- p_logfc + p_pval + plot_annotation(tag_levels = "A")
print(p_combined)

ggsave(file.path(OUTPUT_DIR, "logFC_pval_vs_delta_kWithin.pdf"),
       p_combined, width = 14, height = 4)
ggsave(file.path(OUTPUT_DIR, "logFC_pval_vs_delta_kWithin.png"),
       p_combined, width = 14, height = 6, dpi = 300)
```

### Linear model: does expression effect size predict connectivity loss?

```{r lm_logfc_delta_k}
# Genotype-response only
gr_only <- scatter_df |> filter(group == "Genotype Response DEGs")
fit_gr <- lm(delta_kWithin ~ abs_logFC, data = gr_only)

cat("=== Genotype-response genes: lm(delta_kWithin ~ |logFC|) ===\n")
summary(fit_gr)
cat(sprintf("\nR2 = %.4f  (|logFC| explains %.1f%% of variance in delta_k)\n",
            summary(fit_gr)$r.squared,
            summary(fit_gr)$r.squared * 100))

# Leaf-gradient only
lg_only <- scatter_df |> filter(group == "Leaf Gradient Hubs")
fit_lg <- lm(delta_kWithin ~ abs_logFC, data = lg_only)

cat("\n=== Leaf-gradient genes: lm(delta_kWithin ~ |logFC|) ===\n")
summary(fit_lg)
cat(sprintf("\nR2 = %.4f  (|logFC| explains %.1f%% of variance in delta_k)\n",
            summary(fit_lg)$r.squared,
            summary(fit_lg)$r.squared * 100))

# Combined model with interaction
fit_interaction <- lm(delta_kWithin ~ abs_logFC * group, data = scatter_df)

cat("\n=== Combined: lm(delta_kWithin ~ |logFC| * group) ===\n")
summary(fit_interaction)
cat(sprintf("\nR2 = %.4f\n", summary(fit_interaction)$r.squared))
```

# Combined Statistics Table

```{r combined_table}
# Genotype-response table
gr_pres <- gr_preservation |>
  filter(module_label != "gold", module_label != "grey") |>
  dplyr::select(module = module_label, Zsummary, Zdensity, Zconnectivity)

gr_table <- gr_pres |>
  left_join(
    tibble(
      module = names(observed_gr_kw),
      delta_kWithin = round(observed_gr_kw, 3),
      kWithin_FDR = round(kw_fdr_gr[names(observed_gr_kw)], 4),
      delta_kOut = round(observed_gr_out, 3),
      kOut_FDR = round(kout_fdr_gr[names(observed_gr_out)], 4)
    ),
    by = "module"
  ) |>
  mutate(
    gene_set = "Genotype Response DEGs",
    n_genes = sapply(module, function(m)
      sum(gr_colors == m, na.rm = TRUE)),
    LFP = round(gr_lfp$mean_lfp[match(module, gr_lfp$color)], 3),
    Zsummary = round(Zsummary, 2),
    Zdensity = round(Zdensity, 2),
    Zconnectivity = round(Zconnectivity, 2)
  )

# Leaf-gradient table
lg_pres <- lg_preservation |>
  filter(module_label != "gold", module_label != "grey") |>
  dplyr::select(module = module_label, Zsummary, Zdensity, Zconnectivity)

# Map leaf-gradient module colors through the remap
lg_pres$module <- lg_color_remap[lg_pres$module]

lg_table <- lg_pres |>
  left_join(
    tibble(
      module = names(observed_lg_kw),
      delta_kWithin = round(observed_lg_kw, 3),
      kWithin_FDR = round(kw_fdr_lg[names(observed_lg_kw)], 4),
      delta_kOut = round(observed_lg_out, 3),
      kOut_FDR = round(kout_fdr_lg[names(observed_lg_out)], 4)
    ),
    by = "module"
  ) |>
  mutate(
    gene_set = "Leaf Gradient Hubs",
    n_genes = sapply(module, function(m)
      sum(lg_colors == m)),
    LFP = round(lg_lfp$mean_lfp[match(
      names(lg_color_remap)[match(module, lg_color_remap)],
      lg_lfp$color
    )], 3),
    Zsummary = round(Zsummary, 2),
    Zdensity = round(Zdensity, 2),
    Zconnectivity = round(Zconnectivity, 2)
  )

# Combine
combined_table <- bind_rows(gr_table, lg_table) |>
  dplyr::select(
    gene_set, module, n_genes, LFP,
    Zsummary, Zdensity, Zconnectivity,
    delta_kWithin, kWithin_FDR,
    delta_kOut, kOut_FDR
  ) |>
  arrange(gene_set, kWithin_FDR)

cat("=== COMBINED MODULE STATISTICS ===\n\n")
print(combined_table, n = 30, width = Inf)

write_csv(combined_table,
          file.path(OUTPUT_DIR, "combined_module_stats_table.csv"))
cat("\nSaved: combined_module_stats_table.csv\n")
```

# Summary

```{r summary}
n_gr_sig <- sum(all_results$gene_set == "Genotype Response DEGs" & all_results$fdr < 0.05)
n_lg_sig <- sum(all_results$gene_set == "Leaf Gradient Hubs" & all_results$fdr < 0.05)

cat("=== PERMUTATION TEST SUMMARY ===

Genotype-response modules significant (FDR < 0.05):", n_gr_sig, "of",
    sum(all_results$gene_set == "Genotype Response DEGs"), "
Leaf-gradient modules significant (FDR < 0.05):", n_lg_sig, "of",
    sum(all_results$gene_set == "Leaf Gradient Hubs"), "

Significant genotype-response modules:\n")
all_results |>
  filter(gene_set == "Genotype Response DEGs", fdr < 0.05) |>
  dplyr::select(module, n_genes, observed_mean_delta_k, ecdf_pvalue, fdr) |>
  print()

n_gr_sig_out <- sum(kout_fdr_gr < 0.05)
n_lg_sig_out <- sum(kout_fdr_lg < 0.05)

cat("
=== kOut (inter-modular) ===
Genotype-response modules with significant kOut gain (FDR < 0.05):",
    n_gr_sig_out, "of", length(kout_fdr_gr), "
Leaf-gradient modules with significant kOut gain (FDR < 0.05):",
    n_lg_sig_out, "of", length(kout_fdr_lg), "

=== CONCLUSIONS ===

1. kWithin: Genotype-response modules show significant intramodular
   connectivity loss beyond the permutation null.
   Leaf-gradient modules do not.

2. kOut: No significant gain in inter-modular connectivity
   for genotype-response or leaf-gradient modules. The perturbation
   degrades within-module coexpression structure without causing
   genes to gain spurious cross-module connections.

3. This asymmetry (kWithin loss without kOut gain) argues
   against a global artifact of mean expression shifts:
   if the connectivity change were driven by correlated
   mean shifts inflating all pairwise differences, both
   kWithin and kOut would be affected. Instead, only
   within-module relationships are disrupted, consistent
   with degradation of specific coexpression programs.

4. The linear model confirms this: |logFC| does not
   predict delta kWithin, ruling out differential expression
   magnitude as a driver of connectivity loss.
")
```

# Session Info

```{r session_info}
sessionInfo()
```
