---
title: "Zeal panel: Inv4m effect on flowering time (NC2025)"
author: "Fausto Rodríguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("docs", "inversion_paper"),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4, dpi = 150)
library(here)
source(here("scripts", "utils", "setup_paths.R"))
paths <- setup_project_paths("inversion_paper")

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggpubr)        # theme_classic2, stat_pvalue_manual
  library(ggbeeswarm)    # geom_quasirandom
  library(ggtext)        # element_markdown for *Inv4m* italics
  library(cowplot)
  library(lme4)
  library(lmerTest)
  library(knitr)
})

set.seed(42)
```

## Goal

Test whether the *Inv4m* introgression affects flowering time (DTS, DTA)
in the BZea NIL panel grown at NC2025 (raw data: `data/CLY25_ZEAL.csv`,
provided by Hannah Pil; spatial correction applied upstream).

`Inv4m` is recoded `recurrent` (B73 haplotype) vs `donor` (teosinte
haplotype) so the fixed-effect coefficient reads as the *donor − recurrent*
contrast.

## Data prep

```{r data}
corr <- read.csv(file.path(paths$data, "CLY25_ZEAL.csv"),  stringsAsFactors = FALSE)
meta <- read.csv(file.path(paths$data, "Bzea_metadata.csv"), stringsAsFactors = FALSE)

OUTLIER_PLOTS <- c(2756, 453, 304)   # Hannah's outlier list

df_full <- corr |>
  filter(!Plot %in% OUTLIER_PLOTS) |>
  rename(NIL = Genotype) |>
  filter(!NIL %in% c("B73", "Purple Check"), !is.na(NIL)) |>
  filter(inv4m_introgression %in% c("yes", "no")) |>
  mutate(
    Inv4m     = factor(
      ifelse(inv4m_introgression == "yes", "donor", "recurrent"),
      levels = c("recurrent", "donor")
    ),
    accession = substr(NIL, 1,  7),
    f1        = substr(NIL, 1, 10),
    bc1       = substr(NIL, 1, 13),
    bc2       = substr(NIL, 1, 16),
    taxon_L1  = case_when(
      grepl("^Zea mays",          meta$species[match(accession, meta$donor_id)]) ~ "Z_mays",
      grepl("^Zea luxurians",     meta$species[match(accession, meta$donor_id)]) ~ "Z_luxurians",
      grepl("^Zea diploperennis", meta$species[match(accession, meta$donor_id)]) ~ "Z_diploperennis"
    ),
    taxon_L2  = meta$species[match(accession, meta$donor_id)]
  ) |>
  select(-species) |>
  left_join(meta |> select(donor_id, ancestry = race),
            by = c("accession" = "donor_id"))

both_arms_bc1 <- df_full |>
  distinct(bc1, Inv4m) |>
  count(bc1) |>
  filter(n == 2) |>
  pull(bc1)
df <- df_full |> filter(bc1 %in% both_arms_bc1)

cat(sprintf("plots: %d  |  bc1 families: %d  |  bc2 families: %d  |  NILs: %d\n",
            nrow(df),
            dplyr::n_distinct(df$bc1),
            dplyr::n_distinct(df$bc2),
            dplyr::n_distinct(df$NIL)))
```

## Model selection

A note on terminology: the `Bzea_metadata.csv` file stores the maize
germplasm grouping in a column named `race`. Throughout this notebook
that column is renamed to `ancestry` on join, following current
convention in human and increasingly in plant genetics literature where
"race" is avoided. The two terms refer to the same field; the rename is
cosmetic.

The candidate hierarchy (top → bottom):

```
taxon_L1 (Zea species)
  └─ taxon_L2 (subspecies / species)
       └─ ancestry (race column)
            └─ accession (donor germplasm)
                 └─ f1
                      └─ bc1
                           └─ bc2
                                └─ NIL (full pedigree string)
                                     └─ replicate plots
```

A first decision is which levels carry enough variance contrast to be
identifiable. Two checks: per-level group counts (a level with <5 groups
won't yield a credible variance estimate) and how many groups have **both**
inv4m arms (without that, the level can't be separated from the Inv4m
fixed effect).

### Per-level diagnostic

```{r level-diagnostic}
levels_to_check <- c("taxon_L1", "taxon_L2", "ancestry",
                     "accession", "f1", "bc1", "bc2", "NIL")

summarise_level <- function(col) {
  tab <- table(df[[col]])
  arms <- df |>
    distinct(.data[[col]], Inv4m) |>
    count(.data[[col]]) |>
    pull(n)
  data.frame(
    level         = col,
    n_groups      = length(tab),
    plots_min     = min(tab),
    plots_med     = median(tab),
    plots_max     = max(tab),
    groups_w_both_arms = sum(arms == 2)
  )
}
diag_tbl <- do.call(rbind, lapply(levels_to_check, summarise_level))
kable(diag_tbl, caption = "Per-level group counts and inv4m arm coverage")
```

**Findings:**

- **`taxon_L1` collapses to a single level** (`Z_mays`). The both-arms-at-BC1
  filter strips parviglumis, luxurians, and diploperennis donor lines because
  those lineages don't carry the *Inv4m* mexicana haplotype to introgress.
  Drop it.
- **`taxon_L2` has only 2 levels** (Z. mays mexicana + Z. mays huehuetenangensis).
  Two groups is too few for a credible random-effect variance.
  Drop it (or treat as fixed, but it's nested in `ancestry` anyway).
- **`ancestry` (5 levels)** is the highest taxonomic level the data actually supports.
- **`accession` / `f1` / `bc1`** (18 / 21 / 35 levels) are the candidate
  intermediate pedigree levels; whether they carry independent variance is
  an empirical question — see next section.
- **`bc2` (100 groups)** and **`NIL` (218)** are fine; both are required to
  capture the within-family-vs-between-NIL structure.

### Full slash-nested fit (6 random effects)

Fit the full nested structure first to see which intermediates carry
identifiable variance.

```{r full-fit}
fit_one <- function(trait, formula_str, label) {
  f <- as.formula(paste(trait, "~", formula_str))
  m <- suppressMessages(lmerTest::lmer(f, data = df))
  fe <- summary(m)$coefficients["Inv4mdonor", ]
  ci <- confint(m, parm = "Inv4mdonor", method = "Wald")
  list(model = m, trait = trait, label = label,
       estimate = unname(fe["Estimate"]),
       se       = unname(fe["Std. Error"]),
       df       = unname(fe["df"]),
       t        = unname(fe["t value"]),
       p        = unname(fe["Pr(>|t|)"]),
       ci_lo    = ci[1, 1],
       ci_hi    = ci[1, 2],
       singular = lme4::isSingular(m),
       vc       = as.data.frame(VarCorr(m)))
}

TRAITS <- c("DTA_corr", "DTS_corr", "PH_corr")

full_re <- "Inv4m + (1 | ancestry / accession / f1 / bc1 / bc2 / NIL)"
full_fits <- lapply(TRAITS, fit_one,
                    formula_str = full_re, label = "full 6-RE slash")
names(full_fits) <- TRAITS

for (tr in names(full_fits)) {
  f <- full_fits[[tr]]
  cat(sprintf("\n%s   isSingular = %s\n", tr, f$singular))
  vc <- f$vc; vc$pct <- round(100 * vc$vcov / sum(vc$vcov), 1)
  print(vc[, c("grp", "vcov", "sdcor", "pct")], row.names = FALSE)
}
```

**Result:** `accession`, `f1`, and `bc1` collapse to **0 variance** for both
traits. The model is singular but the Inv4m fixed effect is unbiased — the
intermediates are simply unidentifiable beyond what `ancestry` (above) and
`bc2` (below) already absorb. Mathematically this is expected because the
levels are *strictly nested by construction* (each `bc2` is a unique
substring that uniquely determines its `bc1`, `f1`, `accession`), so any
correlation among siblings sharing a higher level is already captured by
the lower level.

### Parsimonious model

Drop the three zero-variance levels:

```{r parsim-fit}
parsim_re <- "Inv4m + (1 | ancestry / bc2 / NIL)"
parsim_fits <- lapply(TRAITS, fit_one,
                      formula_str = parsim_re, label = "parsimonious 3-RE slash")
names(parsim_fits) <- TRAITS

for (tr in names(parsim_fits)) {
  f <- parsim_fits[[tr]]
  cat(sprintf("\n%s   isSingular = %s\n", tr, f$singular))
  vc <- f$vc; vc$pct <- round(100 * vc$vcov / sum(vc$vcov), 1)
  print(vc[, c("grp", "vcov", "sdcor", "pct")], row.names = FALSE)
  cat(sprintf("Inv4m (donor - recurrent) = %+.3f d   SE = %.3f   95%% CI [%+.3f, %+.3f]   p = %.4g\n",
              f$estimate, f$se, f$ci_lo, f$ci_hi, f$p))
}
```

The parsimonious model converges cleanly (`isSingular = FALSE`) with
**identical** Inv4m estimates and SEs to the singular full fit — the
dropped REs were carrying exactly zero information.

### Sanity check: equivalent flat formulation

`(1 | ancestry / bc2 / NIL)` is mathematically identical to
`(1|ancestry) + (1|bc2) + (1|bc2:NIL)` because the IDs are globally unique
(each `bc2` belongs to one `ancestry`; each `NIL` belongs to one `bc2`). The
slash form is preferred because it makes the nesting visible in the
formula itself.

```{r equiv-check}
flat_re   <- "Inv4m + (1|ancestry) + (1|bc2) + (1|bc2:NIL)"
flat_fits <- lapply(TRAITS, fit_one,
                    formula_str = flat_re, label = "parsim flat")
names(flat_fits) <- TRAITS

cmp <- bind_rows(
  lapply(TRAITS, function(tr)
    tibble(trait = tr,
           form  = c("full slash (6 RE)",
                     "parsim slash (3 RE)",
                     "parsim flat (3 RE)"),
           est   = c(full_fits[[tr]]$estimate, parsim_fits[[tr]]$estimate,
                     flat_fits[[tr]]$estimate),
           se    = c(full_fits[[tr]]$se,       parsim_fits[[tr]]$se,
                     flat_fits[[tr]]$se),
           singular = c(full_fits[[tr]]$singular, parsim_fits[[tr]]$singular,
                        flat_fits[[tr]]$singular)))
)
kable(cmp, digits = 4, caption = "Inv4m fixed effect across three formulations")
```

**Canonical model (used downstream):**

$$
\texttt{trait} \sim \text{Inv4m} + (1 \mid \text{ancestry} / \text{bc2} / \text{NIL})
$$

```{r canonical-handle}
fits <- parsim_fits   # downstream chunks reference `fits`
```

### Where does the variance live?

A natural follow-up: are `bc1`/`bc2` effects "absorbed into ancestry"? To
check, refit the parsimonious model dropping either `ancestry` or `bc2`. If
they were fighting for the same variance, the survivor would grow to
compensate.

```{r variance-partitioning}
vc_table <- function(formula_str, label) {
  m <- suppressMessages(lmerTest::lmer(
    as.formula(paste("DTS_corr ~", formula_str)), data = df))
  vc <- as.data.frame(VarCorr(m))
  vc$pct <- round(100 * vc$vcov / sum(vc$vcov), 1)
  vc$model <- label
  vc[, c("model", "grp", "vcov", "pct")]
}

bind_rows(
  vc_table("Inv4m + (1 | ancestry / bc2 / NIL)",  "ancestry + bc2 + NIL  (canonical)"),
  vc_table("Inv4m + (1 | ancestry) + (1 | NIL)",  "ancestry + NIL  (no bc2)"),
  vc_table("Inv4m + (1 | bc2 / NIL)",         "bc2 + NIL   (no ancestry)")
) |>
  pivot_wider(id_cols = grp, names_from = model, values_from = pct,
              values_fill = 0) |>
  kable(caption = "Variance components (% of total) for DTS_corr under three nestings")
```

**Reading the table:** when `bc2` is dropped, `ancestry` doesn't grow — the
`bc2` variance flows down into `NIL` and `Residual` (the next-finer
levels), not up into `ancestry`. Likewise dropping `ancestry` doesn't inflate
`bc2`. The two REs are **independent partitions** of different sources:
`ancestry` = between-ancestry differences in donor genetic background;
`bc2` = within-ancestry between-family differences (genetic noise within a
ancestry). The dropped levels (`accession`, `f1`, `bc1`) had 0 *additional*
variance because every signal they could carry was already attributable
to the level above (ancestry) or below (bc2) — the structure was overspecified,
not the variance was hiding.

## Permutation test

Permute the `Inv4m` label within each `bc2` family (the finest pedigree
level in the model) so the null preserves the nested structure.
1000 iterations per trait.

```{r perm, cache=TRUE}
permute_within_bc2 <- function(d) {
  d |> group_by(bc2) |> mutate(Inv4m = sample(Inv4m)) |> ungroup()
}

perm_test <- function(trait, n_perm = 1000) {
  observed <- fits[[trait]]$estimate
  f <- as.formula(paste(trait, "~ Inv4m + (1 | ancestry / bc2 / NIL)"))
  null_coefs <- numeric(n_perm)
  for (i in seq_len(n_perm)) {
    d_perm <- permute_within_bc2(df)
    m_perm <- suppressMessages(suppressWarnings(lme4::lmer(f, data = d_perm)))
    null_coefs[i] <- fixef(m_perm)["Inv4mdonor"]
  }
  p_perm <- (sum(abs(null_coefs) >= abs(observed)) + 1) / (n_perm + 1)
  list(observed = observed, null = null_coefs, p_perm = p_perm)
}

perms <- lapply(names(fits), perm_test, n_perm = 1000)
names(perms) <- names(fits)

for (tr in names(perms)) {
  cat(sprintf("%s: permutation p (1000 within-bc2 perms) = %.4g\n",
              tr, perms[[tr]]$p_perm))
}
```

```{r perm-null-plot}
null_df <- bind_rows(lapply(names(perms), function(tr) {
  tibble(trait    = tr,
         null     = perms[[tr]]$null,
         observed = perms[[tr]]$observed)
})) |>
  mutate(trait_label = recode(trait,
                              DTS_corr = "Days to silking",
                              DTA_corr = "Days to anthesis"))

p_null <- ggplot(null_df, aes(x = null)) +
  geom_histogram(bins = 40, fill = "grey75", color = "white") +
  geom_vline(aes(xintercept = observed), color = "#551a8b", linewidth = 1) +
  facet_wrap(~ trait_label, scales = "free") +
  labs(x = "Inv4m coefficient under within-BC2 permutation",
       y = "count",
       title = "Permutation null vs observed Inv4m effect",
       subtitle = "Purple line = observed coefficient (real data); 1000 within-BC2 permutations") +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
print(p_null)
```

## Boxplot (Figure 2 style)

Two-trait boxplot pooling all NILs by Inv4m arm, styled to match Figure 2
(panels A/B) so this can drop into the main figure or supplement without
visual jarring. Significance star reflects the LMM Satterthwaite p (the
permutation p agrees within rounding).

```{r boxplot-fig2-style, fig.width=7.2, fig.height=4.6}
results_tbl <- tibble(
  trait    = names(fits),
  estimate = sapply(fits, `[[`, "estimate"),
  se       = sapply(fits, `[[`, "se"),
  ci_lo    = sapply(fits, `[[`, "ci_lo"),
  ci_hi    = sapply(fits, `[[`, "ci_hi"),
  p_lmm    = sapply(fits, `[[`, "p"),
  p_perm   = sapply(perms, `[[`, "p_perm")
) |>
  mutate(star = case_when(p_lmm < 0.001 ~ "***",
                          p_lmm < 0.01  ~ "**",
                          p_lmm < 0.05  ~ "*",
                          TRUE          ~ "ns"))

# Figure 2's shared theme (verbatim where possible)
pheno_theme_panel <- ggpubr::theme_classic2(base_size = 14) +
  theme(
    plot.title       = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.margin      = margin(t = 30, r = 5, b = 5, l = 5, unit = "pt"),
    axis.title.y     = element_text(size = 12),
    axis.title.x     = element_blank(),
    axis.text.x      = ggtext::element_markdown(size = 12, face = "bold",
                                                color = "black"),
    axis.text.y      = element_text(size = 10),
    legend.position  = "none"
  )

zeal_long <- df |>
  pivot_longer(cols = c(DTA_corr, DTS_corr, PH_corr),
               names_to = "trait", values_to = "value") |>
  filter(!is.na(value))

trait_lookup <- c(DTA_corr = "DTA",
                  DTS_corr = "DTS",
                  PH_corr  = "PH")
y_lab_pool   <- c(DTA_corr = "Days to anthesis",
                  DTS_corr = "Days to silking",
                  PH_corr  = "Plant height (cm)")
y_breaks_pool <- list(
  DTA_corr = scales::breaks_width(1),
  DTS_corr = scales::breaks_width(1),
  PH_corr  = scales::breaks_width(20)
)
star_lookup  <- setNames(
  results_tbl$star[match(TRAITS, results_tbl$trait)], TRAITS)

make_zeal_boxplot <- function(tr) {
  d <- zeal_long |> filter(trait == tr)
  ymax <- max(d$value, na.rm = TRUE)
  yrng <- diff(range(d$value, na.rm = TRUE))
  bracket_df <- tibble(group1 = "recurrent", group2 = "donor",
                       y.position = ymax + 0.08 * yrng,
                       label = star_lookup[[tr]])
  ggplot(d, aes(x = Inv4m, y = value, color = Inv4m)) +
    geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0,
                 outlier.shape = NA) +
    ggbeeswarm::geom_quasirandom(size = 0.8, alpha = 0.5) +
    scale_color_manual(values = c("recurrent" = "gold",
                                  "donor"     = "purple4")) +
    scale_x_discrete(labels = c("recurrent" = "CTRL",
                                "donor"     = "*Inv4m*")) +
    ggpubr::stat_pvalue_manual(bracket_df, tip.length = 0.01, size = 5,
                               bracket.size = 0.6, inherit.aes = FALSE) +
    scale_y_continuous(breaks = y_breaks_pool[[tr]],
                       expand = expansion(mult = c(0.05, 0.15))) +
    labs(title = trait_lookup[[tr]], y = y_lab_pool[[tr]], x = NULL) +
    pheno_theme_panel
}

box_plots <- lapply(TRAITS, make_zeal_boxplot)
panel_box <- cowplot::plot_grid(plotlist = box_plots, nrow = 1, ncol = 3,
                                align = "hv", axis = "tblr")
panel_box
```

## Per-ancestry breakdown

The pooled Inv4m effect averages across 5 races. Which races drive it?
Promote `ancestry` to a fixed-effect interaction to extract per-ancestry contrasts:

$$
\texttt{trait} \sim \text{Inv4m} \times \text{ancestry} + (1 \mid \text{bc2} / \text{NIL})
$$

```{r per-ancestry-fit}
suppressPackageStartupMessages(library(emmeans))

per_race <- function(trait) {
  f <- as.formula(paste(trait, "~ Inv4m * ancestry + (1 | bc2/NIL)"))
  m <- suppressMessages(lmerTest::lmer(f, data = df |> mutate(ancestry = factor(ancestry))))
  aov <- anova(m, type = "III")
  em  <- emmeans(m, ~ Inv4m | ancestry, lmer.df = "satterthwaite")
  ct  <- as.data.frame(contrast(em, method = "revpairwise"))
  ct$p_adj_BH <- p.adjust(ct$p.value, method = "BH")
  ct$trait    <- trait
  list(model = m, anova = aov, contrast = ct |> arrange(estimate))
}

per_race_fits <- lapply(TRAITS, per_race)
names(per_race_fits) <- TRAITS

cat("Inv4m × ancestry interaction (does the effect vary by ancestry?):\n")
for (tr in names(per_race_fits)) {
  pp <- per_race_fits[[tr]]$anova["Inv4m:ancestry", "Pr(>F)"]
  cat(sprintf("  %s: p = %.3f\n", tr, pp))
}

per_race_tbl <- bind_rows(lapply(per_race_fits, `[[`, "contrast")) |>
  mutate(trait_label = recode(trait,
                              DTA_corr = "Days to anthesis",
                              DTS_corr = "Days to silking",
                              PH_corr  = "Plant height (cm)"),
         star = case_when(p_adj_BH < 0.001 ~ "***",
                          p_adj_BH < 0.01  ~ "**",
                          p_adj_BH < 0.05  ~ "*",
                          TRUE             ~ "ns"))

per_race_tbl |>
  select(trait_label, ancestry, estimate, SE, p.value, p_adj_BH, star) |>
  kable(digits = 3,
        caption = "Per-ancestry donor − recurrent contrast (BH-FDR over 5 races within each trait)")
```

The Inv4m × ancestry interaction is **not significant** for either trait,
so formally the data don't reject "all races have the same Inv4m
effect" — the per-ancestry table below is descriptive, not a hypothesis
test the paper relies on.

### Per-ancestry boxplots (Figure 2 style)

```{r per-ancestry-boxplot, fig.width=10, fig.height=10}
race_n <- df |>
  group_by(ancestry, Inv4m) |>
  summarise(n = sum(!is.na(DTS_corr) | !is.na(DTA_corr) | !is.na(PH_corr)),
            .groups = "drop") |>
  pivot_wider(names_from = Inv4m, values_from = n, values_fill = 0) |>
  mutate(ancestry_lab = sprintf("%s (n=%d/%d)", ancestry, recurrent, donor))
ancestry_lab_lookup <- setNames(race_n$ancestry_lab, race_n$ancestry)

zeal_long_race <- df |>
  pivot_longer(cols = c(DTA_corr, DTS_corr, PH_corr),
               names_to = "trait", values_to = "value") |>
  filter(!is.na(value)) |>
  mutate(ancestry_lab = factor(ancestry_lab_lookup[as.character(ancestry)],
                           levels = ancestry_lab_lookup[
                             c("Durango", "Nobogame", "Mesa-Central",
                               "Chalco", "Huehuetenango")]),
         trait_label = factor(recode(trait,
                                     DTA_corr = "Days to anthesis",
                                     DTS_corr = "Days to silking",
                                     PH_corr  = "Plant height (cm)"),
                              levels = c("Days to anthesis",
                                         "Days to silking",
                                         "Plant height (cm)")))

# star table keyed by ancestry × trait_label
star_df <- per_race_tbl |>
  mutate(ancestry_lab = ancestry_lab_lookup[as.character(ancestry)]) |>
  select(ancestry_lab, trait_label, star)

# y-position for star bracket per facet
bracket_df <- zeal_long_race |>
  group_by(ancestry_lab, trait_label) |>
  summarise(ymax = max(value, na.rm = TRUE),
            yrng = diff(range(value, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(y.position = ymax + 0.10 * yrng,
         group1 = "recurrent", group2 = "donor") |>
  left_join(star_df, by = c("ancestry_lab", "trait_label")) |>
  rename(label = star)

p_per_race <- ggplot(zeal_long_race,
                     aes(x = Inv4m, y = value, color = Inv4m)) +
  geom_boxplot(width = 0.4, linewidth = 0.8, alpha = 0,
               outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 0.7, alpha = 0.5) +
  scale_color_manual(values = c("recurrent" = "gold",
                                "donor"     = "purple4")) +
  scale_x_discrete(labels = c("recurrent" = "CTRL",
                              "donor"     = "*Inv4m*")) +
  ggpubr::stat_pvalue_manual(bracket_df, tip.length = 0.01, size = 4,
                             bracket.size = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
  facet_grid(trait_label ~ ancestry_lab, scales = "free_y", switch = "y") +
  labs(x = NULL, y = NULL,
       title    = "Inv4m effect by donor ancestry",
       subtitle = "NC2025 BZea NIL panel; FDR adjusted per ancestry emmeans contrast (5 ancestry family per trait)") +
  theme_bw(base_size = 12) +
  theme(plot.title       = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey92", color = NA),
        strip.text       = element_text(face = "bold"),
        axis.text.x      = ggtext::element_markdown(face = "bold"),
        legend.position  = "none",
        strip.placement  = "outside")
print(p_per_race)
```

```{r per-ancestry-save}
ggsave(file.path(paths$figures, "Zeal_Inv4m_per_race_boxplot.png"),
       p_per_race, width = 10, height = 8, dpi = 300)
ggsave(file.path(paths$figures, "Zeal_Inv4m_per_race_boxplot.svg"),
       p_per_race, width = 10, height = 8, fix_text_size = TRUE)
write.csv(per_race_tbl,
          file.path(paths$intermediate, "Zeal_Inv4m_per_race_contrasts.csv"),
          row.names = FALSE)
```

### Ancestry centered boxplot

The pooled boxplot above looks flat because `ancestry` carries ~20% of the
total variance — mexicana races differ in baseline flowering by several
days, which dwarfs the ~0.9 day Inv4m shift on the raw scale. Subtract
each ancestry's mean and the within ancestry contrast becomes visible:

$$
\tilde{y}_{ij} = y_{ij} - \overline{y}_{\text{ancestry}=j}
$$

```{r boxplot-ancestry-resid, fig.width=7.2, fig.height=4.6}
zeal_resid <- df |>
  group_by(ancestry) |>
  mutate(DTA_resid = DTA_corr - mean(DTA_corr, na.rm = TRUE),
         DTS_resid = DTS_corr - mean(DTS_corr, na.rm = TRUE),
         PH_resid  = PH_corr  - mean(PH_corr,  na.rm = TRUE)) |>
  ungroup() |>
  pivot_longer(cols = c(DTA_resid, DTS_resid, PH_resid),
               names_to = "trait", values_to = "value") |>
  filter(!is.na(value))

resid_lookup <- c(DTA_resid = "DTA",
                  DTS_resid = "DTS",
                  PH_resid  = "PH")
y_lab_resid <- c(DTA_resid = "Days to anthesis (ancestry centered)",
                 DTS_resid = "Days to silking (ancestry centered)",
                 PH_resid  = "Plant height, cm (ancestry centered)")
y_breaks_resid <- list(
  DTA_resid = scales::breaks_width(1),
  DTS_resid = scales::breaks_width(1),
  PH_resid  = scales::breaks_width(20)
)
star_lookup_resid <- setNames(
  results_tbl$star[match(TRAITS, results_tbl$trait)],
  c("DTA_resid", "DTS_resid", "PH_resid"))

make_resid_boxplot <- function(tr) {
  d <- zeal_resid |> filter(trait == tr)
  ymax <- max(d$value, na.rm = TRUE)
  yrng <- diff(range(d$value, na.rm = TRUE))
  bracket_df <- tibble(group1 = "recurrent", group2 = "donor",
                       y.position = ymax + 0.08 * yrng,
                       label = star_lookup_resid[[tr]])
  ggplot(d, aes(x = Inv4m, y = value, color = Inv4m)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0,
                 outlier.shape = NA) +
    ggbeeswarm::geom_quasirandom(size = 0.8, alpha = 0.5) +
    scale_color_manual(values = c("recurrent" = "gold",
                                  "donor"     = "purple4")) +
    scale_x_discrete(labels = c("recurrent" = "CTRL",
                                "donor"     = "*Inv4m*")) +
    ggpubr::stat_pvalue_manual(bracket_df, tip.length = 0.01, size = 5,
                               bracket.size = 0.6, inherit.aes = FALSE) +
    scale_y_continuous(breaks = y_breaks_resid[[tr]],
                       expand = expansion(mult = c(0.05, 0.15))) +
    labs(title = resid_lookup[[tr]], y = y_lab_resid[[tr]],
         x = NULL) +
    pheno_theme_panel
}

resid_plots <- lapply(c("DTA_resid", "DTS_resid", "PH_resid"),
                      make_resid_boxplot)
panel_resid <- cowplot::plot_grid(plotlist = resid_plots,
                                  nrow = 1, ncol = 3,
                                  align = "hv", axis = "tblr")
panel_resid

ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_boxplot_ancResid.png"),
       panel_resid, width = 7.2, height = 4.6, dpi = 300)
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_boxplot_ancResid.svg"),
       panel_resid, width = 7.2, height = 4.6, fix_text_size = TRUE)
```

The *Inv4m* boxes now sit visibly below the B73 boxes — same data, same
LMM contrast (−0.88 d / −0.61 d), but the ancestry stratification is no
longer masking it. The dashed line is the within-ancestry mean (= zero by
construction).

### Lineage corrected boxplot (ancestry + bc2 BLUPs subtracted, NIL preserved)

Subtract the upper random effect BLUPs (ancestry + bc2) but **keep** the
NIL random effect, so each NIL line still contributes its own deviation
to the visual spread:

$$
\tilde{y}_{ij} = y_{ij} - \hat{u}_{\text{ancestry}} - \hat{u}_{\text{bc2}}
              = X\beta + \hat{u}_{\text{NIL}} + \varepsilon
$$

This matters because Inv4m is perfectly nested within NIL (each NIL
carries one Inv4m haplotype, never both). Subtracting the NIL BLUP would
absorb the line level variation the LMM uses for inference, leaving
only plot to plot replication noise — the boxes would look artificially
tight. Keeping the NIL BLUP makes the visual spread honest: each Inv4m
arm shows the distribution of NIL line means around its arm mean, plus
within NIL plot noise. The donor − recurrent gap is still exactly the
LMM coefficient.

```{r boxplot-lmm-resid, fig.width=7.2, fig.height=4.6}
m_dts <- parsim_fits[["DTS_corr"]]$model
m_dta <- parsim_fits[["DTA_corr"]]$model
m_ph  <- parsim_fits[["PH_corr"]]$model

# IMPORTANT: predict()/residuals() return values only for rows the model
# actually fitted (NA-omitted). Align by the original row indices stored
# in model.frame() to avoid silent recycling.
#
# Lineage corrected = subtract ancestry + bc2 BLUPs; KEEP NIL BLUP.
#   y - ancestry_BLUP - bc2_BLUP
# = X*beta + NIL_BLUP + residual
# = predict(m, re.form = ~(1 | NIL:bc2:ancestry)) + residuals(m)
# (the deepest random effect grouping name from the slash form)
add_lmm_resid <- function(d, model, target_col, out_col) {
  rows_used <- as.integer(rownames(model.frame(model)))
  d[[out_col]] <- NA_real_
  d[[out_col]][rows_used] <-
    predict(model, re.form = ~(1 | NIL:bc2:ancestry)) + residuals(model)
  d
}

df_lmm <- df |>
  add_lmm_resid(m_dts, "DTS_corr", "DTS_lmm_resid") |>
  add_lmm_resid(m_dta, "DTA_corr", "DTA_lmm_resid") |>
  add_lmm_resid(m_ph,  "PH_corr",  "PH_lmm_resid")

# sanity check: donor - recurrent mean of residuals should equal the LMM beta
sanity <- df_lmm |>
  group_by(Inv4m) |>
  summarise(DTA_mean = mean(DTA_lmm_resid, na.rm = TRUE),
            DTS_mean = mean(DTS_lmm_resid, na.rm = TRUE),
            PH_mean  = mean(PH_lmm_resid,  na.rm = TRUE),
            .groups = "drop")
cat("residualized mean per arm:\n"); print(sanity)
cat(sprintf("residualized donor - recurrent: DTA = %+.3f (beta %+.3f), DTS = %+.3f (beta %+.3f), PH = %+.3f (beta %+.3f)\n",
            diff(sanity$DTA_mean), fixef(m_dta)["Inv4mdonor"],
            diff(sanity$DTS_mean), fixef(m_dts)["Inv4mdonor"],
            diff(sanity$PH_mean),  fixef(m_ph)["Inv4mdonor"]))

zeal_lmm <- df_lmm |>
  pivot_longer(cols = c(DTA_lmm_resid, DTS_lmm_resid, PH_lmm_resid),
               names_to = "trait", values_to = "value") |>
  filter(!is.na(value))

lmm_lookup <- c(DTA_lmm_resid = "DTA",
                DTS_lmm_resid = "DTS",
                PH_lmm_resid  = "PH")
y_lab_lookup <- c(DTA_lmm_resid = "Days to anthesis (lineage corrected)",
                  DTS_lmm_resid = "Days to silking (lineage corrected)",
                  PH_lmm_resid  = "Plant height, cm (lineage corrected)")
y_breaks_lookup <- list(
  DTA_lmm_resid = scales::breaks_width(1),
  DTS_lmm_resid = scales::breaks_width(1),
  PH_lmm_resid  = scales::breaks_width(20)
)
star_lookup_lmm <- setNames(
  results_tbl$star[match(c("DTA_corr", "DTS_corr", "PH_corr"), results_tbl$trait)],
  c("DTA_lmm_resid", "DTS_lmm_resid", "PH_lmm_resid"))

make_lmm_boxplot <- function(tr) {
  d <- zeal_lmm |> filter(trait == tr)
  ymax <- max(d$value, na.rm = TRUE)
  yrng <- diff(range(d$value, na.rm = TRUE))
  bracket_df <- tibble(group1 = "recurrent", group2 = "donor",
                       y.position = ymax + 0.08 * yrng,
                       label = star_lookup_lmm[[tr]])
  ggplot(d, aes(x = Inv4m, y = value, color = Inv4m)) +
    geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0,
                 outlier.shape = NA) +
    ggbeeswarm::geom_quasirandom(size = 0.8, alpha = 0.5) +
    scale_color_manual(values = c("recurrent" = "gold",
                                  "donor"     = "purple4")) +
    scale_x_discrete(labels = c("recurrent" = "CTRL",
                                "donor"     = "*Inv4m*")) +
    ggpubr::stat_pvalue_manual(bracket_df, tip.length = 0.01, size = 5,
                               bracket.size = 0.6, inherit.aes = FALSE) +
    scale_y_continuous(breaks = y_breaks_lookup[[tr]],
                       expand = expansion(mult = c(0.05, 0.15))) +
    labs(title = lmm_lookup[[tr]], y = y_lab_lookup[[tr]],
         x = NULL) +
    pheno_theme_panel
}

lmm_plots <- lapply(c("DTA_lmm_resid", "DTS_lmm_resid", "PH_lmm_resid"),
                    make_lmm_boxplot)
panel_lmm <- cowplot::plot_grid(plotlist = lmm_plots,
                                nrow = 1, ncol = 3,
                                align = "hv", axis = "tblr")
panel_lmm

ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_boxplot_lmmResid.png"),
       panel_lmm, width = 7.2, height = 4.6, dpi = 300)
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_boxplot_lmmResid.svg"),
       panel_lmm, width = 7.2, height = 4.6, fix_text_size = TRUE)

# write the per-plot residualized values alongside raw + ancestry-centered
out <- df_lmm |>
  group_by(ancestry) |>
  mutate(DTS_anc_centered = DTS_corr - mean(DTS_corr, na.rm = TRUE),
         DTA_anc_centered = DTA_corr - mean(DTA_corr, na.rm = TRUE),
         PH_anc_centered  = PH_corr  - mean(PH_corr,  na.rm = TRUE)) |>
  ungroup() |>
  select(Plot, NIL, ancestry, bc2, Inv4m,
         DTA_corr, DTS_corr, PH_corr,
         DTA_anc_centered, DTS_anc_centered, PH_anc_centered,
         DTA_lmm_resid, DTS_lmm_resid, PH_lmm_resid)
write.csv(out,
          file.path(paths$intermediate, "Zeal_phenotypes_race_corrected.csv"),
          row.names = FALSE)
```

Each Inv4m arm now shows the spread of NIL line means around its arm
mean. The donor − recurrent gap matches the LMM coefficient (within a
small drift from BLUP shrinkage), and the within arm SD reflects the
line to line variation the LMM weighs — not the artificially tight
plot to plot noise that subtracting the NIL BLUP would produce.

## Forest plot

```{r forest, fig.width=8, fig.height=4.8}
unit_lookup <- c(DTA_corr = "d", DTS_corr = "d", PH_corr = "cm")
results_tbl <- results_tbl |>
  mutate(trait_label = recode(trait,
                              DTA_corr = "Days to anthesis",
                              DTS_corr = "Days to silking",
                              PH_corr  = "Plant height (cm)"),
         unit  = unit_lookup[trait],
         label = sprintf("%+.2f %s   p=%.3g %s", estimate, unit, p_lmm, star))

print(results_tbl |> select(trait, estimate, se, ci_lo, ci_hi, p_lmm, p_perm))

# faceted forest, free x scale per trait so days and cm coexist honestly
p_forest <- ggplot(results_tbl,
                   aes(x = estimate, y = trait_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.15, linewidth = 0.8) +
  geom_point(size = 4, color = "purple4") +
  geom_text(aes(label = label), nudge_y = 0.25, size = 4.2) +
  facet_wrap(~ trait_label, scales = "free", ncol = 1) +
  labs(
    x = "Inv4m effect (donor − recurrent, native units)",
    y = NULL,
    title    = "Inv4m accelerates flowering in the BZea NIL panel",
    subtitle = "NC2025 field; LMM with ancestry / bc2 / NIL nesting; 95% Wald CI"
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title       = element_text(face = "bold"),
        strip.background = element_blank(),
        strip.text       = element_blank())
print(p_forest)
```

## Supplementary figure assembly

4 panel composite: (A) flowering boxplot with DTA + DTS as nested facets
sharing a single days y axis, (B) flowering forest of LMM Inv4m effect
(DTA + DTS in days), (C) PH boxplot alone (cm), (D) PH forest alone
(cm). Mixed unit traits are kept on separate panels rather than forced
onto a shared axis.

```{r supp-figure, fig.width=11, fig.height=4.6}
suppressPackageStartupMessages(library(ggh4x))

forest_df <- results_tbl |>
  mutate(trait_short = factor(recode(trait,
                                     DTA_corr = "Anthesis",
                                     DTS_corr = "Silking",
                                     PH_corr  = "Plant height"),
                              levels = c("Anthesis", "Silking", "Plant height")))

## Panel A — flowering boxplot (anthesis + silking nested, shared y axis)
flow_long <- df_lmm |>
  pivot_longer(cols = c(DTA_lmm_resid, DTS_lmm_resid),
               names_to = "trait", values_to = "value") |>
  filter(!is.na(value)) |>
  mutate(trait_short = factor(recode(trait,
                                     DTA_lmm_resid = "Anthesis",
                                     DTS_lmm_resid = "Silking"),
                              levels = c("Anthesis", "Silking")))

flow_stars <- forest_df |> filter(trait %in% c("DTA_corr", "DTS_corr")) |>
  transmute(trait_short = factor(recode(trait,
                                        DTA_corr = "Anthesis",
                                        DTS_corr = "Silking"),
                                 levels = c("Anthesis", "Silking")),
            star = star)
flow_brackets <- flow_long |>
  group_by(trait_short) |>
  summarise(ymax = max(value, na.rm = TRUE),
            yrng = diff(range(value, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(group1 = "recurrent", group2 = "donor",
         y.position = ymax + 0.08 * yrng) |>
  left_join(flow_stars, by = "trait_short") |>
  rename(label = star)

p_flow_box <- ggplot(flow_long,
                     aes(x = Inv4m, y = value, color = Inv4m)) +
  geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0,
               outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 0.8, alpha = 0.5) +
  scale_color_manual(values = c("recurrent" = "gold",
                                "donor"     = "purple4")) +
  scale_x_discrete(labels = c("recurrent" = "CTRL",
                              "donor"     = "*Inv4m*")) +
  ggpubr::stat_pvalue_manual(flow_brackets, tip.length = 0.01, size = 5,
                             bracket.size = 0.6, inherit.aes = FALSE) +
  scale_y_continuous(breaks = scales::breaks_width(1),
                     expand = expansion(mult = c(0.05, 0.15))) +
  ggh4x::facet_nested(~ trait_short, scales = "fixed") +
  labs(x = NULL, y = "Days (lineage corrected)") +
  pheno_theme_panel +
  theme(strip.background  = element_blank(),
        strip.text        = element_text(size = 13, face = "bold"),
        ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.5))

## Panel B — flowering forest (DTA + DTS, days)
forest_flow <- forest_df |> filter(trait %in% c("DTA_corr", "DTS_corr"))
p_forest_flow <- ggplot(forest_flow,
                        aes(x = trait_short, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.15, linewidth = 0.8) +
  geom_point(size = 4, color = "purple4") +
  scale_y_continuous(breaks = scales::breaks_width(0.5)) +
  labs(x = NULL, y = "*Inv4m* effect (days)") +
  pheno_theme_panel +
  theme(axis.title.y = ggtext::element_markdown(size = 12))

## Panel C — PH boxplot alone (cm)
ph_long <- df_lmm |>
  filter(!is.na(PH_lmm_resid)) |>
  rename(value = PH_lmm_resid)
ph_star <- forest_df |> filter(trait == "PH_corr") |> pull(star)
ph_bracket <- tibble(group1 = "recurrent", group2 = "donor",
                     y.position = max(ph_long$value, na.rm = TRUE) +
                                  0.08 * diff(range(ph_long$value, na.rm = TRUE)),
                     label = ph_star)
p_ph_box <- ggplot(ph_long, aes(x = Inv4m, y = value, color = Inv4m)) +
  geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0,
               outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 0.8, alpha = 0.5) +
  scale_color_manual(values = c("recurrent" = "gold",
                                "donor"     = "purple4")) +
  scale_x_discrete(labels = c("recurrent" = "CTRL",
                              "donor"     = "*Inv4m*")) +
  ggpubr::stat_pvalue_manual(ph_bracket, tip.length = 0.01, size = 5,
                             bracket.size = 0.6, inherit.aes = FALSE) +
  scale_y_continuous(breaks = scales::breaks_width(20),
                     expand = expansion(mult = c(0.05, 0.15))) +
  labs(title = "Plant height", x = NULL,
       y = "Plant height, cm (lineage corrected)") +
  pheno_theme_panel

## Panel D — PH forest alone (cm)
forest_ph <- forest_df |> filter(trait == "PH_corr")
p_forest_ph <- ggplot(forest_ph,
                      aes(x = trait_short, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.15, linewidth = 0.8) +
  geom_point(size = 4, color = "purple4") +
  labs(x = NULL, y = "*Inv4m* effect (cm)") +
  pheno_theme_panel +
  theme(axis.title.y = ggtext::element_markdown(size = 12))

panel_supp <- cowplot::plot_grid(p_flow_box, p_forest_flow,
                                 p_ph_box,   p_forest_ph,
                                 nrow = 1, ncol = 4,
                                 align = "hv", axis = "tblr",
                                 rel_widths = c(1.6, 1, 1, 0.9),
                                 labels = c("A", "B", "C", "D"),
                                 label_size = 16,
                                 label_fontface = "bold")
panel_supp

ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_supp.png"),
       panel_supp, width = 11, height = 4.6, dpi = 300)
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_supp.svg"),
       panel_supp, width = 11, height = 4.6, fix_text_size = TRUE)
```

## Save outputs

```{r save}
write.csv(results_tbl,
          file.path(paths$intermediate, "Zeal_Inv4m_flowering_lmm_results.csv"),
          row.names = FALSE)
write.csv(null_df,
          file.path(paths$intermediate, "Zeal_Inv4m_flowering_lmm_perm_null.csv"),
          row.names = FALSE)
write.csv(diag_tbl,
          file.path(paths$intermediate, "Zeal_Inv4m_flowering_lmm_level_diagnostic.csv"),
          row.names = FALSE)

# Figure 2-style boxplot (drop-in)
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_boxplot.png"),
       panel_box, width = 4.8, height = 4.6, dpi = 300)
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_boxplot.svg"),
       panel_box, width = 4.8, height = 4.6, fix_text_size = TRUE)

# Forest summary
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_forest.png"),
       p_forest, width = 7, height = 3.2, dpi = 300)
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_forest.svg"),
       p_forest, width = 7, height = 3.2, fix_text_size = TRUE)

# Permutation null
ggsave(file.path(paths$figures, "Zeal_Inv4m_flowering_perm_null.png"),
       p_null, width = 9, height = 3.6, dpi = 300)
```

## Session info

```{r session}
sessionInfo()
```
